### 2016-05-06 Paper_v1.tex
First commit

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses
\usepackage{ifthen}  % Conditional
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort,comma,numbers,sort&compress]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{fullpage}
\usepackage{comment}
\usepackage{lineno}
%\usepackage{chicago}
\usepackage{textcomp}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e}
\DeclareMathOperator*{\argmin}{\arg\!\min}

\usepackage{algorithm2e}
\usepackage{algpseudocode}
%\usepackage[space]{cite}
\urlstyle{rm}

%\textwidth = 6.50 in
%\textheight = 9.5 in
%\oddsidemargin =  0.0 in
%\evensidemargin = 0.0 in
%\topmargin = -0.50 in
%\headheight = 0.0 in
%\headsep = 0.25 in
%\parskip = 0.15in
%\linespread{1.75}
\doublespace

%\bibliographystyle{chicago}
\bibliographystyle{plos2009}

\makeatletter
\renewcommand\subsection{\@startsection
	{subsection}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\bfseries}}
\renewcommand\subsubsection{\@startsection
	{subsubsection}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\itshape}}
\renewcommand\section{\@startsection
	{subsection}{2}{0mm}
	{-0.2in}
	{0.05\baselineskip}
	{\normalfont\large\bfseries}}
\renewcommand\paragraph{\@startsection
	{paragraph}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\itshape}}
\makeatother

%Review style settings
%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}

%Publication style settings

% Single space'd bib -
\setlength\bibsep{0pt}

\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

% Change the number format in the ref list -
\renewcommand{\bibnumfmt}[1]{#1.}

% Change Figure to Fig.
\renewcommand{\figurename}{Fig.}

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell Free Protein Synthesis in \emph{Escherichia~coli}}}}
\vspace{0.05in}
{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, James Swartz$^{1}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering {$^{1}$School of Chemical Engineering}}
{\par \centering {Stanford University, Stanford, California 94305}}
{\par \centering \textbf{Running Title:}~Dynamic model of cell free protein synthesis}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Scientific~Reports}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Fill me in.

%We further tested the predictive power of the coagulation model parameters against data not used in training, and found good agreement between simulations and experimental measurements. Lastly, we tested the performance of DOPS on commonly used test functions for global optimization and on published biochemical parameter estimation benchmark problems. For the wide range of problems that we considered, DOPS outperformed other meta-heuristic approaches despite a limited number of function evaluations.

\vspace{0.1in}
{\noindent \textbf{Keywords:}~Biochemical engineering, systems biology, cell free protein synthesis}

% Extra abstract
% Mathematical modeling of biological systems with multiple feed back loops is one such area where parameter estimation is a difficult non-linear optimization problem. This difficulty is further compounded when dealing with parameter vectors of high dimensions.

%In this study, we present the dynamically dimensioned particle
%a novel meta-heuristic approach that combines a variant of particle swarm optimization (PSO) along with dynamically dimensioned search (DDS) to obtain near optimal solutions of high dimensional biochemical networks within a relatively few function evaluations.
%We use a particle swarm optimization technique that uses multi-swarms to generate candidate vectors which are then greedily updated using DDS by dynamically varying the perturbed parameter dimensions. We tested this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and also against DDS alone. The new algorithm outperforms all the other meta-heuristics on the coagulation model. The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data. We also performed this comparison on commonly used test functions (Ackley and Rosenbrock) for global optimization and found the same behavior. Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach. We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.

\pagebreak

\setcounter{page}{1}

% Uncomment in production -
%\linenumbers

\section*{Introduction}
The introduction has four paragraphs (introduction no longer than 3 pages). Follow the cell free paper from last year:
\begin{enumerate}
	\item{\textbf{First~paragraph}: Introduce mathematical modeling, and its role in biochemical engineering. }
	\item{\textbf{Second~paragraph}: Contrast current static metabolic modeling approaches e.g., FBA with dynamic models.}
	\item{\textbf{Third~paragraph}: Introduce cell free protein synthesis.}
	\item{\textbf{Fourth~paragraph}: In this study, [Repeat the abstract with some additional detail]. Taken together, [killer statement].}
\end{enumerate}

% Extra text from the introduction -
%With the capacity to construct very large models using various mathematical formalisms \cite{chen2009input,tasseff2011modeling,luan2007computationally,mo2007genome,orth2011comprehensive,karr2012whole,buchel2013path2models,smallbone2010towards}, more often than not,

%this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics  The new algorithm outperforms all the other meta-heuristics on the coagulation model. The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data. We also performed this comparison on  and found the same behavior. Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach. We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.
%took into cognizance the fact that it may not be necessary to obtain the exact solution for high dimensional biological systems and that good enough solutions can be quickly obtained without expending a lot of objective function evaluations. Hence our current approach uses the power of population based heuristics along with DDS to obtain near optimal or good solutions. Though in theory we can combine any meta-heuristic with DDS, for the current purpose we used Particle Swarm Optimization (PSO). PSO unlike Genetic Algorithm (GA) or Differential Evolution (DE) does not have complex operations like cross over, mutation or recombination. It is simple to use and does not have a lot of parameters associated with other heuristics. However PSO is known to rapidly converge to a local optimum and thus several variants \cite{peer2003using,zhan2009adaptive,li2007fast} have been developed of which Multi swarm particle swarm optimization approaches (MLSPSO) \cite{zhao2008dynamic,liang2005dynamic} are one of the effective ones. We used MLSPSO to preclude bad regions of search and thereafter search within these regions using DDS.

\clearpage

\section*{Results}

The results are presented in \textbf{past~tense}. Each paragraph starts with a statement of the result in that paragraph in active voice.
Each results paragraph ends with a Taken together type statement followed by a link statement e.g., Next we considered etc. When referring to figures, state what the figures shows (Fig. ZZ).

\begin{enumerate}
	\item{\textbf{First~section}:Description of the model biology}
	\item{\textbf{Second~section}:Estimation of the model parameters, and refinement of the model structure (inclusion of the AA degradation pathways)}
	\item{\textbf{Third~section}:Analysis of the flux distribution (over the ensemble?), sensitivity results (first parameters, then AA)}
\end{enumerate}

\clearpage

%As the dimensionality of  increases, the search region gets widened and thus the problem becomes more challenging.
%

\section*{Discussion}

The discussion has three (sometimes four) paragraphs:
\begin{enumerate}
	\item{\textbf{First~paragraph}: Present a modified version of the last paragraph of the introduction. In this study, [...]. Taken together, [killer statement]}
	\item{\textbf{Second~paragraph}: Contrast the key findings of the study with other computational/experimental studies}
	\item{\textbf{Third~paragraph}: Present future directions. If you had more time, what would like to do? Highlight the key shortcomings of the approach and how will we address them in the future.
	In this case, we will have a scaling issue if we extend to genome scale. We should extend to dynamic cases, and we need to experimentally validate the findings. }
\end{enumerate}

\clearpage

\section*{Materials and Methods}

Follow the cell free paper from last year:
\begin{enumerate}
	\item{\textbf{Model~formulation}: Present the spatial flux balance approach, and enzyme balances for the (un)channeled case. Outline any bounds formulation, and state all parameter assumptions.}
\end{enumerate}

\clearpage

%Need to talk more about biochemical benefits and importance for of biochemical problems

\section*{Acknowledgements}
This study was supported by an award from [FILL ME IN].
\clearpage

\bibliography{References_v1}

\clearpage

% Figures and captions go here ...
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.00\textwidth]{./figs/<Filename>.pdf}
%\caption{Captiontext goes here}
%}\label{fig:<label_name>}
%\end{figure}

\clearpage

% Supplemental figures -
% Set the S-
\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{T\arabic{table}}
\renewcommand\thepage{S-\arabic{page}}
\renewcommand\theequation{S\arabic{equation}}

% Reset the counters -
\setcounter{equation}{0}
\setcounter{table}{0}
\setcounter{figure}{0}
\setcounter{page}{1}

% Supplemental figures go here ...
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.00\textwidth]{./figs/<Filename>.pdf}
%\caption{Captiontext goes here}
%}\label{fig:<label_name>}
%\end{figure}

\end{document}

### 2016-05-17 Paper_v1.tex
Added some text

@@
 \documentclass[12pt]{article}
+
 % Load packages
@@
 \section*{Introduction}
+
+Mathematical modeling has long contributed to our understanding of metabolism.
+Decades before the genomics revolution, mechanistically, structured metabolic models arose from the desire to predict microbial phenotypes resulting from changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}.
+The single cell \textit{E.
+coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple, regulated catabolic and anabolic pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}.
+Shuler and coworkers generated many single cell kinetic models, including single cell models of eukaryotes \citep{1989_steinmeyer_shuler_ChemEngSci,1992_wu_shuler_AnnNYAcadSci}, minimal cell architectures \citep{2004_castellanos_shuler_PNAS}, as well as DNA sequence based whole-cell models of \textit{E.
+coli} \citep{2008_atlas_shuler_IETSysBio}.
+Conversely, highly abstracted kinetic frameworks, such as the cybernetic framework, represented a paradigm shift, viewing cells as growth-optimizing strategists \citep{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}.
+Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}.
+Unfortunately, traditional, fully structured cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously.
+However, recent cybernetic formulations from Ramkrishna and colleagues have successfully treated this identifiability challenge through elementary mode reduction~\cite{2009_song_ramkrishna_BiotechBioeng,Song:2011aa}.
+
+In the post genomics world, large-scale stoichiometric reconstructions of microbial metabolism popularized by static, constraint-based modeling techniques such as flux balance analysis (FBA) have become standard tools \citep{2012_lewis_palsson_NatRevMicrobio}.
+Since the first genome-scale stoichiometric model of \textit{E.
+coli}, developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, well over 100 organisms, including industrially important prokaryotes such as \textit{E.
+coli} \citep{Feist:2007aa} or \textit{B.
+subtilis} \citep{Oh:2007aa}, are now available \citep{2009_feist_palsson_NatRevMicrobio}.
+Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome-scale kinetic models to an underdetermined linear algebraic system, which can be solved efficiently even for large systems.
+Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
+Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \citep{2002_ibarra_edwards_palsson_Nat}, although other metabolic objectives have also been estimated \citep{2007_schuetz_sauer_MolSysBio}.
+Recent advances in constraint-based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \citep{2013_hyduke_lewis_palsson_MolBioSys}.
+Thus, modern constraint-based approaches have proven extremely useful in the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
+However, genome-scale kinetic models of industrial important organisms such as \textit{E. coli} have yet to be constructed.
+
+Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
+Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall.
+This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
+Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
+Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \citep{Jewett:2008aa}.
+However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}.
+Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}.
+Interestingly, many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system.
+For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth.
+Thus, cell-free operation holds several significant advantages for model development, identification and validation.
+Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms,
+such as \textit{E. coli} or \textit{B.~subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
+
+In this study, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
+The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulation with traditional kinetic pathway modeling.
+This integration allows the description of complex regulatory interactions, such as time-dependent allosteric regulation of enzyme activity, in the absence of specific mechanistic information.
+The regulatory rules are easy to understand, easy to formulate and do not rely on overarching theoretical abstractions or restrictive assumptions.
+We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks.
+In particular, we tested whether our effective modeling approach could describe classically expected enzyme kinetic behavior, and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, from synthetic experimental data.
+Toward these questions, we explored five hypothetical cell-free networks.
+Each network shared the same enzymatic connectivity, but had different allosteric regulatory connectivity.
+We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information.
+Second, when integrated into network models, these rules captured classical regulatory patterns such as product-induced feedback inhibition.
+Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data
+starting from an unbiased collection of possible allosteric structures using particle swarm optimization.
+However, when starting with an initial population that was heavily enriched with incorrect structures, our particle swarm approach could converge to an incorrect structure.
+While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
+
 The introduction has four paragraphs (introduction no longer than 3 pages). Follow the cell free paper from last year:
@@
 
-Follow the cell free paper from last year:
-\begin{enumerate}
-	\item{\textbf{Model~formulation}: Present the spatial flux balance approach, and enzyme balances for the (un)channeled case. Outline any bounds formulation, and state all parameter assumptions.}
-\end{enumerate}
+\subsection*{Formulation and Solution of the Model Equations}
+We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell-free metabolic networks:
+\begin{eqnarray}
+	\frac{dx_{i}}{dt} & = & \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
+	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i = 1,2,\hdots,\mathcal{E}}
+\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of enzymes in the model.
+The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$.
+Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$).
+The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$.
+If $\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$.
+Conversely, if $\sigma_{ij}<0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates metabolite $i$ is not connected with reaction $j$.
+Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant.
+The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right) = \mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).
+
+The reaction rate was written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$), $r_{j}\left(\mathbf{x},\mathbf{k}\right) = \bar{r}_{j}v_{j}$.
+In this study, we used either saturation or mass action kinetics.
+The control term $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
+For each rate, we used a rule-based approach to select from competing control factors.
+If rate j was influenced by $1,\dots,m$ factors, we modeled this relationship as
+$v_{j} = \mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
+where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
+The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
+variable. Each regulatory transfer function took the form:
+\begin{equation}\label{eqn:control-factor}
+	f_{ij}\left(\mathcal{Z}_{i},k_{ij},\eta_{ij}\right) = k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}/\left({1 + k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}}\right)
+\end{equation}where $\mathcal{Z}_{i}$ denotes the abundance factor $i$, $k_{ij}$ denotes a gain parameter, and $\eta_{ij}$ denotes a cooperativity parameter.
+In this study, we used $\mathcal{I}_{j}\in\left\{\max,\min\right\}$ \cite{pr3010138}. If a process has no modifying factors, $v_{j} = 1$.
+We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
+\begin{equation}\label{eqn:rate-bar}
+	\bar{r}_{j} = k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
+\end{equation}
+where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
+$K_{js}$ denotes the saturation constant for species $s$ in reaction $j$.
+The product in Equation~\eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$).
+
+\subsection*{Generation of model ensemble}
+Fill me in.
+
+\subsection*{Global and local sensitivity analysis}
+We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which parameters controlled the performance of the reduced order model \citep{SOBOL_METHOD}.
+We computed the total sensitivity index of each parameter relative to two performance objectives, the peak thrombin time and the area under the thrombin curve (thrombin exposure).
+We established the sampling bounds for each parameter from the minimum and maximum value of that parameter in the parameter set ensemble.
+We used the sampling method of Saltelli \textit{et al.} \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ parameter sets which obeyed our parameter ranges,
+where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, $N$ = 10,000 and $d$ = 22, so the total sensitivity indices were computed from
+460,000 model evaluations. The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \citep{SALIB}.
 

### 2016-05-18 Paper_v1.tex
May17

@@
 \usepackage{amsmath}
-\usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e}
+%\usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e}
 \DeclareMathOperator*{\argmin}{\arg\!\min}
 
-\usepackage{algorithm2e}
-\usepackage{algpseudocode}
+%\usepackage{algorithm2e}
+%\usepackage{algpseudocode}
 %\usepackage[space]{cite}
@@
 \begin{eqnarray}
-	\frac{dx_{i}}{dt} & = & \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
-	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i = 1,2,\hdots,\mathcal{E}}
+	\frac{dx_{i}}{dt} & = & \sum_{j=1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
+	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i=1,2,\hdots,\mathcal{E}}
 \end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of enzymes in the model.
@@
 If $\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$.
-Conversely, if $\sigma_{ij}<0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates metabolite $i$ is not connected with reaction $j$.
+Conversely, if $\sigma_{ij}<0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij}=0$ indicates metabolite $i$ is not connected with reaction $j$.
 Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant.
-The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right) = \mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).
+The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right)=\mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right)=\mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).
 
-The reaction rate was written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$), $r_{j}\left(\mathbf{x},\mathbf{k}\right) = \bar{r}_{j}v_{j}$.
+The reaction rate was written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$), $r_{j}\left(\mathbf{x},\mathbf{k}\right)=\bar{r}_{j}v_{j}$.
 In this study, we used either saturation or mass action kinetics.
@@
 If rate j was influenced by $1,\dots,m$ factors, we modeled this relationship as
-$v_{j} = \mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
+$v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
 where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
@@
 \begin{equation}\label{eqn:control-factor}
-	f_{ij}\left(\mathcal{Z}_{i},k_{ij},\eta_{ij}\right) = k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}/\left({1 + k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}}\right)
+	f_{ij}\left(\mathcal{Z}_{i},k_{ij},\eta_{ij}\right)=k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}/\left({1 + k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}}\right)
 \end{equation}where $\mathcal{Z}_{i}$ denotes the abundance factor $i$, $k_{ij}$ denotes a gain parameter, and $\eta_{ij}$ denotes a cooperativity parameter.
-In this study, we used $\mathcal{I}_{j}\in\left\{\max,\min\right\}$ \cite{pr3010138}. If a process has no modifying factors, $v_{j} = 1$.
+In this study, we used $\mathcal{I}_{j}\in\left\{mean\right\}$ \cite{pr3010138}. If a process has no modifying factors, $v_{j}=1$.
 We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
 \begin{equation}\label{eqn:rate-bar}
-	\bar{r}_{j} = k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
+	\bar{r}_{j}=k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
 \end{equation}
@@
 \subsection*{Generation of model ensemble}
-Fill me in.
+We generated an ensemble of 18,000 parameter sets via a downhill-only random walk Monte Carlo method \cite{?}.
+Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-absolute-error between experimental data and model predicitons:
+\begin{equation}\label{eqn:cost-function}
+    cost=\sum_{i=1}^{\mathcal{D}}\left(w_i\sum_{j=1}^{\mathcal{T}}abs\left(x_{ij}^{data}-x_{i}^{sim}|_{t(j)}\right)\right)
+\end{equation}
+where $\mathcal{D}$ denotes the number of datasets, $w_i$ denotes a weight, equal to 5 for the glucose, CAT, pyruvate, lactate, acetate, succinate, and malate datasets, and 1 elsewhere, $\mathcal{T}$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ dneotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
+We then perturbed model parameters:
+\begin{equation}\label{eqn:parameter-perturbation}
+    k_i^{new}=k_i*exp(a\,r_i)\qquad{i=1,2,\hdots,\mathcal{P}}
+\end{equation}
+where $\mathcal{P}$ denotes the number of parameters, equal to 652, which includes 163 rate constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i$th parameter, $k_i$ denotes the current value of the $i$th parameter, $a$ denotes a distribution variance, set to 0.03, and $r$ denotes a random sample from the normal distribution.
+We stored the parameter set and calculated its cost; if it was less than the previous cost, we used the new parameter set to generate the following set.
+After generating 180,000 sets we defined the 18,000 sets with the lowest cost values as our ensemble, and the set with the lowest cost value as our best-fit set.
 
 \subsection*{Global and local sensitivity analysis}
-We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which parameters controlled the performance of the reduced order model \citep{SOBOL_METHOD}.
-We computed the total sensitivity index of each parameter relative to two performance objectives, the peak thrombin time and the area under the thrombin curve (thrombin exposure).
-We established the sampling bounds for each parameter from the minimum and maximum value of that parameter in the parameter set ensemble.
-We used the sampling method of Saltelli \textit{et al.} \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ parameter sets which obeyed our parameter ranges,
-where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, $N$ = 10,000 and $d$ = 22, so the total sensitivity indices were computed from
-460,000 model evaluations. The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \citep{SALIB}.
+We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which of the experimentally controllable parameters affected the performance of the reduced order model \citep{SOBOL_METHOD}.
+This included the initial conditions of glucose, oxygen, amino acids, and enzymes.
+We computed the total sensitivity index of each parameter relative to a performance objective of area under the CAT curve (CAT production).
+We established the sampling bounds for each parameter from the value of that parameter in the set used to generate the ensemble.
+We used the sampling method of Saltelli \textit{et al.} \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ sets which obeyed our parameter ranges,
+where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, $N$ = 300 and $d$ = 185, so the total sensitivity indices were computed from
+111,600 model evaluations. The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \citep{SALIB}.
+We conducted a local sensitivity analysis to estimate which of the other model parameters affected performance.
+This included the same parameters that were varied in the ensemble: rate constants, saturation constants, and control parameters.
+The local sensitivity for each parameter was calculated across a sub-ensemble of 180 parameter sets, randomly chosen from the ensemble of 18,000 sets:
+\begin{equation}\label{eqn:local-sensitivity}
+\begin{split}
+    S_{ij}=\frac{p_{ij}}{AUC(p_{ij})}\frac{AUC(p_{ij}+\Delta p_{ij})-AUC(p_{ij})}{\Delta p_{ij}}\qquad{i=1,2,\hdots,\mathcal{E}}\qquad{j=1,2,\hdots,\mathcal{P}} \\
+    \Delta p_{ij}=0.001\;p_{ij}\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad
+\end{split}
+\end{equation}
+where $\mathcal{E}$ denotes the number of parameter sets in the sub-ensemble, equal to 180, $\mathcal{P}$ denotes the number of parameters, equal to 652, $S_{ij}$ denotes the sensitivity of the $j$th parameter for the $i$th parameter set, $p_{ij}$ denotes the value of the $j$th parameter for the $i$th parameter set, $\Delta p_{ij}$ denotes the perturbation of the $j$th parameter for the $i$th parameter set, equal to 0.1\% of the parameter value, and $AUC()$ denotes the area under the CAT curve.
+We then calculated the mean and
+
+\subsection*{Calculation of CAT yield}
+The theoritical carbon yield of CAT was calculated using flux balance anaylsis (FBA) with a sequence-based analysis on CAT.
+The sequence specific FBA \cite{2002_allen_palsson} problem was formulated as:
+\begin{equation}\label{eqn:FBA-yield}
+\begin{split}
+	\max_{\boldsymbol{v}}{} \! \left( v_{obj}=\mathbf{\boldsymbol{c}}^T \boldsymbol{v} \right) \qquad \alpha_i \leq v_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R} \\
+	\mathrm{Subject \; to:}	\; \; \mathbf{S}\mathbf{v}=\mathbf{0}\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad
+\end{split}
+\end{equation}
+where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{v}$ denotes the unknown flux vector, $\boldsymbol{c}$ denotes the objective selection vector, and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $v_{i}$, respectively.
+The stoichiometric matrix was expanded to include the transcription and translation reactions for producing CAT. 
+The objective $v_{obj}$ was to maximize the specific rate of CAT formation.
+The specific glucose, amino acids and oxygen uptake rates were constrained to allow a maximum flux of 10 mM/hr.
+The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
+The solution flux vector was used to calculate the theoritical carbon yield of CAT based on carbon consumed (glucose and amino acids).
+The carbon yield was formulated as:
+\begin{equation}\nonumber
+	Yield=\frac{C_{CAT}v_{CAT}}{\sum_{i=1}^{\mathcal{R}}C_{i}v_{i}}
+\end{equation}
+where $C_{CAT}$ and $C_{i}$ denote the carbon number of CAT and substrate $i$, respectively, $v_{CAT}$ and $v_{i}$ denote the flux of CAT and substrate $i$, respectively,
+and $\mathcal{R}$ denotes the number of substrates consumed.
 
@@
 % Figures and captions go here ...
-%\begin{figure}[ht]
-%\centering
-%\includegraphics[width=1.00\textwidth]{./figs/<Filename>.pdf}
-%\caption{Captiontext goes here}
-%}\label{fig:<label_name>}
-%\end{figure}
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.75\textwidth]{./Figures/Network.pdf}
+\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and NADPH/NADH transfer.
+Mean ± standard error for the reaction flux at 1.5 hrs (top) and 3 hrs (bottom), normalized to CAT synthesis flux.}
+\label{fig:Network}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+%\includegraphics[width=1.00\textwidth]{./Figures/BothCarbon.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/Carbon.pdf}
+\includegraphics[width=0.05\textwidth]{./Figures/null.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/CarbonNoControl.pdf}
+\caption{A: Central carbon metabolism in the presence of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species.
+Best-fit parameter set (orange line) versus experimental data (points).
+95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 18,000 sets.
+B: Central carbon metabolism in the absence of allosteric control.}
+\label{fig:BothCarbon}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/Energy.pdf}
+\caption{Energy species and energy totals by base in the presence of allosteric control.
+Best-fit parameter set (orange line) versus experimental data (points).
+95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 18,000 sets.}
+\label{fig:Energy}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/Amino.pdf}
+\caption{Amino acids in the presence of allosteric control.
+Best-fit parameter set (orange line) versus experimental data (points).
+95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 18,000 sets.}
+\label{fig:Amino}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+%\includegraphics[width=1.00\textwidth]{./Figures/SensitivitiesHalfTall.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/Sensitivity.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/Metab_sensitivity.pdf}
+\caption{Total-order global sensitivities for experimentally controllable initial conditions, including glucose, oxygen, amino acids, and enzymes.}
+\label{fig:SensitivitiesHalfTall}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.94\textwidth]{./Figures/SensAll.png}
+\caption{Mean and standard error of local sensitivities of rate constants (top), saturation constants (middle), and control parameters (bottom).}
+\label{fig:SensAll}
+\end{figure}
 

### 2016-05-19 Paper_v1.tex
May18

@@
 \vspace{0.05in}
-{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, James Swartz$^{1}$ and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, Joseph Wayman, Kara Calhoun$^{1}$, James Swartz$^{1}$ and Jeffrey D. Varner$^{*}$}}
 \vspace{0.10in}
@@
 
-The results are presented in \textbf{past~tense}. Each paragraph starts with a statement of the result in that paragraph in active voice.
-Each results paragraph ends with a Taken together type statement followed by a link statement e.g., Next we considered etc. When referring to figures, state what the figures shows (Fig. ZZ).
+%The results are presented in \textbf{past~tense}. Each paragraph starts with a statement of the result in that paragraph in active voice.
+%Each results paragraph ends with a Taken together type statement followed by a link statement e.g., Next we considered etc. When referring to figures, state what the figures shows (Fig. ZZ).
 
-\begin{enumerate}
-	\item{\textbf{First~section}:Description of the model biology}
-	\item{\textbf{Second~section}:Estimation of the model parameters, and refinement of the model structure (inclusion of the AA degradation pathways)}
-	\item{\textbf{Third~section}:Analysis of the flux distribution (over the ensemble?), sensitivity results (first parameters, then AA)}
-\end{enumerate}
+%\begin{enumerate}
+%	\item{\textbf{First~section}:Description of the model biology}
+%	\item{\textbf{Second~section}:Estimation of the model parameters, and refinement of the model structure (inclusion of the AA degradation pathways)}
+%	\item{\textbf{Third~section}:Analysis of the flux distribution (over the ensemble?), sensitivity results (first parameters, then AA)}
+%\end{enumerate}
+
+\subsection*{Estimation of an ensmeble of models}
+We constructed an initial model of cell-free \textit{E.
+coli} metabolism by removing the growth-associated processes from the genome-scale model of Palsson and coworkers \cite{?} and adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset from Swartz and coworkers \cite{?}.
+This initial model captured the core metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle), as well as the synthesis of energy species, amino acids, and CAT.
+We generated mass balance equations around metabolites and enzymes in the network, modeled as ordinary differential equations with reaction rates equal to the product of a kinetic term and a control term.
+We used multiple saturation kinetics to model metabolic fluxes and mass action kinetics to model enzyme degradation.
+We modeled the control term using a rule-based approach in which each control factor had a regulatory transfer function, modeled as a Hill function \cite{?}, and the control term was calculated as the mean of all transfer functions.
+%The model was trained against the Swartz dataset, which included measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (ATP, ADP, AMP, GTP, GDP, GMP, CTP, CDP, CMP, UTP, UDP, UMP), and 18 of the 20 proteinogenic amino acids.
+The model was trained against the Swartz dataset, which included measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (AXP, GXP, CXP, UXP), and 18 of the 20 proteinogenic amino acids.
+We generated an ensemble of 18,000 parameter sets by minimizing the error between the Swartz dataset and the metabolite concentrations predicted by the model.
+Out of these 18,000 sets, we defined the set with the lowest error value as our best-fit set.
+The ensemble captures the central carbon metabolism, including glucose uptake, CAT production, and the dynamics of the organic acid intermediates (Fig.~\ref{fig:BothCarbon}, top).
+Allosteric control is important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:BothCarbon}, bottom).
+The ensemble also captures the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:BothCarbon}, top) and the totals by base (Fig.~\ref{fig:Energy}).
+The ensemble and the best-fit set also predict some of the amino acid measurements, while failing to predict others (Fig.~\ref{fig:Amino}).
+This is likely due to a structural deficiency in the model; in some cases, the consumption of an amino acid through CAT synthesis is not enough to explain the decrease shown in the data, and there are no other reactions that consume it.
+Thus, a more comprehensive biological description is needed to fully explain amino acid dynamics.
+
+\subsection*{Sensitivity analysis}
+We conducted a local sensitivity analysis to determine which of the kinetic and control parameters affected model performance.
+We calculated performance as area under the CAT curve, which was directly related to CAT synthesis rate, as the culture time was fixed and no CAT degradation was modeled.
+We randomly chose 180 sets of the 18,000 sets in the ensemble and defined these as our sub-ensemble; for each set in the sub-ensemble, we varied the rate constant and saturation constants of each metabolic flux and measured the resulting change in CAT production to estimate the sensitivity of model performance to that parameter.
+We did the same for the control parameters, both the order (Hill coefficient) and gain (related to the the dissociation constant).
+This allowed us to estimate the relative importance of the kinetic and control parameters to CAT production across the ensemble.
+Of the rate constants, those with the highest positive sensitivities were CAT synthesis, GTP synthesis, GMP synthesis, glutamine synthesis, and aspartate synthesis (Fig.~\ref{fig:LocalSensitivity}, top).
+This is explained by GTP and the amino acids being reactants for CAT synthesis.
+The rate constants with the largest negative sensitivities were GTP degradation, arginine synthesis, and UMP synthesis.
+While GTP degradation is obvious, the others can be explained in that they consume ATP as well as several amino acids, all of which are reactants for CAT synthesis.
+Of the saturation constants, the reverse is seen: the largest positive sensitivities are those associated with arginine synthesis, while the largest negative sensitivities are those associated with CAT synthesis, GTP synthesis, and GMP synthesis (Fig.~\ref{fig:LocalSensitivity}, middle).
+This is because an increase in saturation constant causes a decrease in the corresponding reaction rate \cite{?}.
+The control parameters were seen to be the least significant and the most uncertain (Fig.~\ref{fig:LocalSensitivity}, bottom).
+Only two had a small standard error across the ensemble, relative to the ensemble mean sensitivity: the gain and order for pyruvate acting as an inhibitor on the pdh reaction.
+This could be because pdh consumes pyruvate and diverts carbon away from the pathways that ultimately contribute to CAT production.
+Taken together with the lack of change in glucose uptake and CAT production when control is removed (Fig.~\ref{fig:BothCarbon}), this suggests that allosteric control is relatively unimportant to CAT production and metabolism as a whole compared to kinetic effects.
+
+We conducted a global sensitivity analysis on the parameters that could be controlled experimentally: the initial conditions of glucose, oxygen, amino acids, and enzymes.
+We used the variance-based method of Sobol, and the same objective function of area under the CAT curve.
+Using a parameter set of relatively good fit against data, we defined parameter bounds and generated a Sobol sequence of 111,600 parameter values that fit within those bounds.
+We then calculated the total-order sensitivity and confidence interval for each of the experimentally controllable initial conditions.
+As the sensitivities were total-order, they were guaranteed to be non-negative.
+The largest sensitivities belonged to the initial conditions of the CAT macromolecular synthesis machinery, GTP synthase, GTP degradation, some amino acids such as phenylalanine, proline, and leucine, and some amino acid synthases (Fig.~\ref{fig:GlobalSensitivity}).
+This is all explained by GTP and amino acids being reactants for CAT synthesis.
+While some of the amino acids and amino acid synthases were among the highest in sensitivity, theirs were also very uncertain relative to the CAT macromolecular synthesis machinery and GTP synthase.
+The initial conditions of glucose and oxygen were among the least important according to the global sensitivity analysis, suggesting that the model predicts that CAT production can be sustained by consuming initial stores or can be powered by other means.
+
+\subsection*{Calculation of CAT yield}
+We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
+Arginine and glutamate were excluded due to not being present in the Swartz dataset.
+The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
+We then used sequence-specific FBA to calculate a theoretical maximum CAT yield of 0.1942.
+Thus, we showed that our experimental dataset and best-fit parameter set were producing CAT at 44.5\% and 44.9\% of the theoretical maximum, respectively.
+This allowed us to quantify the amount of carbon being diverted to byproducts, and suggests that there is potential for a doubling of CAT production by reducing this diversion of carbon.
 
 \clearpage
-
-%As the dimensionality of  increases, the search region gets widened and thus the problem becomes more challenging.
-%
 
@@
 	\item{\textbf{Third~paragraph}: Present future directions. If you had more time, what would like to do? Highlight the key shortcomings of the approach and how will we address them in the future.
-	In this case, we will have a scaling issue if we extend to genome scale. We should extend to dynamic cases, and we need to experimentally validate the findings. }
+	In this case, we will have a scaling issue if we extend to genome scale. We should extend to dynamic cases, and we need to experimentally validate the findings.}
 \end{enumerate}
@@
 
-\subsection*{Formulation and Solution of the Model Equations}
+\subsection*{Formulation and solution of the model equations}
 We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell-free metabolic networks:
@@
 We generated an ensemble of 18,000 parameter sets via a downhill-only random walk Monte Carlo method \cite{?}.
-Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-absolute-error between experimental data and model predicitons:
+Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-absolute-error between experimental data and model predictions:
 \begin{equation}\label{eqn:cost-function}
@@
 where $\mathcal{E}$ denotes the number of parameter sets in the sub-ensemble, equal to 180, $\mathcal{P}$ denotes the number of parameters, equal to 652, $S_{ij}$ denotes the sensitivity of the $j$th parameter for the $i$th parameter set, $p_{ij}$ denotes the value of the $j$th parameter for the $i$th parameter set, $\Delta p_{ij}$ denotes the perturbation of the $j$th parameter for the $i$th parameter set, equal to 0.1\% of the parameter value, and $AUC()$ denotes the area under the CAT curve.
-We then calculated the mean and
+We then calculated the mean and standard error of each local sensitivity across the sub-ensemble of 180 sets.
 
 \subsection*{Calculation of CAT yield}
-The theoritical carbon yield of CAT was calculated using flux balance anaylsis (FBA) with a sequence-based analysis on CAT.
+The yield on CAT production was calculated for three cases: the experimental data, the best-fit parameter set, and a theoretical maximum yield.
+In each case the yield was formulated as a ratio of carbon produced as CAT to carbon consumed as reactants (glucose and amino acids):
+\begin{equation}\label{eqn:yield-definition}
+	Yield=\frac{\Delta CAT\;C_{CAT}}{\sum_{i=1}^{\mathcal{R}}\max(\Delta m_{i},0)\;C_{m_i}}
+\end{equation}
+where $\Delta CAT$ denotes the amount of CAT produced, $C_{CAT}$ denotes carbon number of CAT, $\mathcal{R}$ denotes the number of reactants, $\Delta m_{i}$ denotes the amount of the $i$th reactant consumed, never allowed to be negative, and $C_{m_i}$ denotes the carbon number of the $i$th reactant.
+Because no data was available for arginine or glutamate, these reactants were left out of all three calculations.
+In the experimental case and the best-fit set case, yield was calculated by setting $\Delta CAT$ equal to the final minus the initial CAT concentration and setting $\Delta m_{i}$ equal to the initial minus the final reactant concentration.
+The theoretical yield was calculated using flux balance anaylsis (FBA) with a sequence-based analysis on CAT.
 The sequence specific FBA \cite{2002_allen_palsson} problem was formulated as:
-\begin{equation}\label{eqn:FBA-yield}
+\begin{equation}\label{eqn:FBA}
 \begin{split}
@@
 The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
-The solution flux vector was used to calculate the theoritical carbon yield of CAT based on carbon consumed (glucose and amino acids).
-The carbon yield was formulated as:
-\begin{equation}\nonumber
-	Yield=\frac{C_{CAT}v_{CAT}}{\sum_{i=1}^{\mathcal{R}}C_{i}v_{i}}
+The solution flux vector was used to calculate the theoretical carbon yield of CAT, reformulated in terms of flux:
+\begin{equation}\label{eqn:yield}
+	Yield=\frac{v_{CAT}\;C_{CAT}}{\sum_{i=1}^{\mathcal{R}}\max(v_{i},0)\;C_{i}}
 \end{equation}
-where $C_{CAT}$ and $C_{i}$ denote the carbon number of CAT and substrate $i$, respectively, $v_{CAT}$ and $v_{i}$ denote the flux of CAT and substrate $i$, respectively,
-and $\mathcal{R}$ denotes the number of substrates consumed.
+where $v_{CAT}$ denotes the CAT export flux and $v_{i}$ denotes the import flux of the $i$th substrate.
 
@@
 \includegraphics[width=1.00\textwidth]{./Figures/CarbonNoControl.pdf}
-\caption{A: Central carbon metabolism in the presence of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species.
+\caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species.
 Best-fit parameter set (orange line) versus experimental data (points).
-95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 18,000 sets.
-B: Central carbon metabolism in the absence of allosteric control.}
+95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 18,000 sets.}
 \label{fig:BothCarbon}
@@
 \centering
+\includegraphics[width=0.94\textwidth]{./Figures/SensAll.png}
+\caption{Mean and standard error of local sensitivities of rate constants (top), saturation constants (middle), and control parameters (bottom).}
+\label{fig:LocalSensitivity}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
 %\includegraphics[width=1.00\textwidth]{./Figures/SensitivitiesHalfTall.pdf}
@@
 \caption{Total-order global sensitivities for experimentally controllable initial conditions, including glucose, oxygen, amino acids, and enzymes.}
-\label{fig:SensitivitiesHalfTall}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=0.94\textwidth]{./Figures/SensAll.png}
-\caption{Mean and standard error of local sensitivities of rate constants (top), saturation constants (middle), and control parameters (bottom).}
-\label{fig:SensAll}
+\label{fig:GlobalSensitivity}
 \end{figure}

### 2016-05-20 Paper_v1.tex
May19

@@
 This is explained by GTP and the amino acids being reactants for CAT synthesis.
+Also, GMP synthesis increases the total amount of guanosine, allowing for more GTP production.
 The rate constants with the largest negative sensitivities were GTP degradation, arginine synthesis, and UMP synthesis.
@@
 This could be because pdh consumes pyruvate and diverts carbon away from the pathways that ultimately contribute to CAT production.
-Taken together with the lack of change in glucose uptake and CAT production when control is removed (Fig.~\ref{fig:BothCarbon}), this suggests that allosteric control is relatively unimportant to CAT production and metabolism as a whole compared to kinetic effects.
+Taken together with the lack of change in glucose uptake and CAT production when control is removed (Fig.~\ref{fig:BothCarbon}), this suggests that allosteric control is not the limiting factor to CAT production.
 
@@
 We then used sequence-specific FBA to calculate a theoretical maximum CAT yield of 0.1942.
-Thus, we showed that our experimental dataset and best-fit parameter set were producing CAT at 44.5\% and 44.9\% of the theoretical maximum, respectively.
+Thus, we showed that our experimental dataset and best-fit parameter set were each producing CAT at 45\% of the theoretical maximum.
 This allowed us to quantify the amount of carbon being diverted to byproducts, and suggests that there is potential for a doubling of CAT production by reducing this diversion of carbon.
@@
 
-The discussion has three (sometimes four) paragraphs:
-\begin{enumerate}
-	\item{\textbf{First~paragraph}: Present a modified version of the last paragraph of the introduction. In this study, [...]. Taken together, [killer statement]}
-	\item{\textbf{Second~paragraph}: Contrast the key findings of the study with other computational/experimental studies}
-	\item{\textbf{Third~paragraph}: Present future directions. If you had more time, what would like to do? Highlight the key shortcomings of the approach and how will we address them in the future.
-	In this case, we will have a scaling issue if we extend to genome scale. We should extend to dynamic cases, and we need to experimentally validate the findings.}
-\end{enumerate}
+%The discussion has three (sometimes four) paragraphs:
+%\begin{enumerate}
+%	\item{\textbf{First~paragraph}: Present a modified version of the last paragraph of the introduction. In this study, [...]. Taken together, [killer statement]}
+%	\item{\textbf{Second~paragraph}: Contrast the key findings of the study with other computational/experimental studies}
+%	\item{\textbf{Third~paragraph}: Present future directions. If you had more time, what would like to do? Highlight the key shortcomings of the approach and how will we address them in the future.
+%	In this case, we will have a scaling issue if we extend to genome scale. We should extend to dynamic cases, and we need to experimentally validate the findings.}
+%\end{enumerate}
+
+We constructed a model of cell-free \textit{E. coli} metabolism with CAT production based on saturation kinetics and allosteric control.
+We trained the model against measurements of glucose, CAT, organic acid intermediates, energy species, and amino acids.
+We generated an ensemble of 18,000 parameter sets that predict the experimental dataset, with the exception of some amino acids.
+We calculated the carbon yield of CAT production at 45\% of the theoretical maximum obtained by flux balance analysis; this was in line with the CAT yield shown by the dataset.
+We conducted sensitivity analyses and determined that CAT production was most sensitive to the rate constants, saturation constants, and enzyme initial conditions associated with CAT synthesis, GTP synthesis, GMP synthesis, and amino acid synthesis, and to a lesser extent some amino acid initial conditions.
+CAT production was much less sensitive to control parameters, suggesting that while allosteric control affects some organic acid intermediates, it does not limit CAT production.
+The initial conditions of glucose and oxygen were even less important to CAT production, suggesting that it can be driven by other means.
+However, the low sensitivities to glucose and oxygen only apply in the specific case of the parameter sets that were studied.
+When other species' initial conditions and associated parameters are different, glucose and oxygen can become more important to CAT production.
+
+2nd paragraph: Compare to other studies
+This is the first study to model a cell-free \textit{E. coli} system at this scope, toward genome-scale.
+
+One important direction for the future is to expand the biological description, especially with regard to amino acids.
+Specifically, adding more reactions that consume amino acids would improve the model's ability to predict those that show a decrase in the experimental data.
+Also, including specific transcription and translation steps for CAT would allow us to more accurately model the complexity and the resource cost of protein synthesis.
+Another area for future work is to more thoroughly sample parameter space.
+For the metabolites in the dataset, initial conditions were fixed at the initial data values.
+All other parameters were varied in a manner so as to best fit the dataset.
+However, the resulting ensemble may not represent every biological or practical possibility.
+In a different region of parameter space, the system could behave differently, including the flux distribution through the network, the accuracy and spread of ensemble fits, the relative sensitivities, and the yield as a percentage of the theoretical maximum.
+Testing the model under a variety of conditions could strengthen or challenge the findings of this study.
+Further experimentatiion could also be used to gain a deeper understanding of model performance under a variety of conditions.
+Specifically, CAT production performed in the absence of amino acids could inform the system's ability to manufacture them, while experimentation in the absence of glucose or oxygen could shed light on how important they are to protein synthesis, and under which conditions.
+Finally, the approach should be extended to other protein products.
+CAT is only a test protein used for model identification; the modeling framework, and to some extent the parameter values, should be protein agnostic.
+An important extension of this study would be to apply its insights to other protein applications, where possible.
 
@@
 \caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and NADPH/NADH transfer.
-Mean ± standard error for the reaction flux at 1.5 hrs (top) and 3 hrs (bottom), normalized to CAT synthesis flux.}
+FBA flux value (top), and mean ± standard error across ensemble at 1.5 hrs (middle) and 3 hrs (bottom), normalized to CAT synthesis flux.}
 \label{fig:Network}

### 2016-05-20 Paper_v1.tex
May20

@@
 The objective $v_{obj}$ was to maximize the specific rate of CAT formation.
-The specific glucose, amino acids and oxygen uptake rates were constrained to allow a maximum flux of 10 mM/hr.
+The specific glucose uptake rate was constrained to allow a maximum flux of 10 mM/hr \cite{2002_Mahadevan_BiophysJ}; the amino acids and oxygen uptake rates were also bound to allow a maximum flux of 10 mM/hr, but did not reach this maximum flux.
+Glucose, oxygen, and amino acids were modeled as being imported into the system, whereas CAT synthesis was modeled through protein export.
+The rest of the network followed a psuedo steady-state asusmption where all other metabolites were not allowed to accumulate; thus, the network could be solved by linear programming. 
 The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.

### 2016-05-24 Paper_v1.tex
May23_2

@@
 The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulation with traditional kinetic pathway modeling.
-This integration allows the description of complex regulatory interactions, such as time-dependent allosteric regulation of enzyme activity, in the absence of specific mechanistic information.
+%This integration allows the description of complex regulatory interactions, such as time-dependent allosteric regulation of enzyme activity, in the absence of specific mechanistic information.
+This integration allows the description of complex regulatory interactions in the absence of specific mechanistic information.
 The regulatory rules are easy to understand, easy to formulate and do not rely on overarching theoretical abstractions or restrictive assumptions.
-We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks.
-In particular, we tested whether our effective modeling approach could describe classically expected enzyme kinetic behavior, and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, from synthetic experimental data.
-Toward these questions, we explored five hypothetical cell-free networks.
-Each network shared the same enzymatic connectivity, but had different allosteric regulatory connectivity.
-We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information.
+%We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks.
+%In particular, we tested whether our effective modeling approach could describe classically expected enzyme kinetic behavior, and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, from synthetic experimental data.
+%Toward these questions, we explored five hypothetical cell-free networks.
+%Each network shared the same enzymatic connectivity, but had different allosteric regulatory connectivity.
+%We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information.
 Second, when integrated into network models, these rules captured classical regulatory patterns such as product-induced feedback inhibition.
-Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data
-starting from an unbiased collection of possible allosteric structures using particle swarm optimization.
+%Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data
+%starting from an unbiased collection of possible allosteric structures using particle swarm optimization.
 However, when starting with an initial population that was heavily enriched with incorrect structures, our particle swarm approach could converge to an incorrect structure.
@@
 
-The introduction has four paragraphs (introduction no longer than 3 pages). Follow the cell free paper from last year:
-\begin{enumerate}
-	\item{\textbf{First~paragraph}: Introduce mathematical modeling, and its role in biochemical engineering. }
-	\item{\textbf{Second~paragraph}: Contrast current static metabolic modeling approaches e.g., FBA with dynamic models.}
-	\item{\textbf{Third~paragraph}: Introduce cell free protein synthesis.}
-	\item{\textbf{Fourth~paragraph}: In this study, [Repeat the abstract with some additional detail]. Taken together, [killer statement].}
-\end{enumerate}
+%The introduction has four paragraphs (introduction no longer than 3 pages). Follow the cell free paper from last year:
+%\begin{enumerate}
+%	\item{\textbf{First~paragraph}: Introduce mathematical modeling, and its role in biochemical engineering. }
+%	\item{\textbf{Second~paragraph}: Contrast current static metabolic modeling approaches e.g., FBA with dynamic models.}
+%	\item{\textbf{Third~paragraph}: Introduce cell free protein synthesis.}
+%	\item{\textbf{Fourth~paragraph}: In this study, [Repeat the abstract with some additional detail]. Taken together, [killer statement].}
+%\end{enumerate}
 
@@
 
-2nd paragraph: Compare to other studies
-This is the first study to model a cell-free \textit{E. coli} system at this scope, toward genome-scale.
+In this study, we present the first dynamic, cell-free model of \textit{E. coli} metabolism and biosynthesis at this scope.
+While the early models of Shuler and coworkers achieved large-scale, dynamic descriptions of single cells \citep{1984_domach_shuler_BiotechBioeng_01}, and the stoichiometric models associated with the FBA approach are computationally efficient amd widespread \citep{2009_feist_palsson_NatRevMicrobio}, none have yet been able to construct genome-scale kinetic models of cell-free \textit{E.
+coli} metabolism.
+We have harnessed the advantages of cell-free systems (no cell wall, no cell growth) and integrated traditional kinetics with allosteric control to build an ensemble of kinetic models that fit a comprehensive cell-free dataset, toward genome-scale.
+This study should provide a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} metabolism, toward industrial-scale biosynthetic production.
 

### 2016-05-26 Paper_v1.tex
May26

@@
 
+%PEP was modeled as an inhibitor for phosphofructokinase \cite{2011_cabrera_JBiolChem}.
+%PEP was modeled as an activator for fructose-biphosphatase \cite{2000_donahue_JBacteriol,2006_hines_JBiolChem,2007_hines_JBiolChem}.
+%FDP was modeled as an activator for pyruvate kinase \cite{2010_zhu_Biochimie}.
+%PEP was modeled as an inhibitor for PEP carboxykinase \cite{2003_ravanal_JProteinChem}.
+%FDP was modeled as an activator for PEP carboxylase \cite{1972_wohl_JBiolChem}.
+%Pyruvate was modeled as an inhibitor for pyruvate dehydrogenase \cite{2007_kale_JBiolChem,2002_arjunan_Biochemistry}.
+%PEP was modeled as an inhibitor for PEP synthetase \cite{1973_chulavatnatol_JBiolChem}.
+%AKG was modeled as an inhibitor for citrate synthase \cite{1994_pereira_JBiolChem,1983_robinson_FEBSLett}.
+%PEP was modeled as an inhibitor for isocitrate dehydrogenase \cite{2007_ogawa_JBacteriol}.
+%AKG was modeled as an inhibitor for isocitrate lyase \cite{1988_mackintosh_BiochemJ}.
+%PEP was modeled as an inhibitor for isocitrate lyase \cite{2007_ogawa_JBacteriol,1988_mackintosh_BiochemJ}.
+%3PG was modeled as an inhibitor for isocitrate lyase \cite{1988_mackintosh_BiochemJ}.
+%AKG was modeled as an inhibitor for malate synthase \cite{?}.
+%PEP was modeled as an inhibitor for malate synthase \cite{?}.
+%3PG was modeled as an inhibitor for malate synthase \cite{?}.
+%Acetyl CoA was modeled as an inhibitor for malate dehydrogenase \cite{?}.
+%Pyruvate was modeled as an activator for lactate dehydrogenase \cite{2008_okino_ApplMicrobiolBiotechnol}.
+
+We added regulation to the network as informed by literature, for a total of 17 interactions.
+PEP was modeled as an inhibitor for phosphofructokinase \cite{2011_cabrera_JBiolChem}, PEP carboxykinase \cite{2003_ravanal_JProteinChem}, PEP synthetase \cite{1973_chulavatnatol_JBiolChem}, isocitrate dehydrogenase \cite{2007_ogawa_JBacteriol}, isocitrate lyase \cite{2007_ogawa_JBacteriol,1988_mackintosh_BiochemJ}, and malate synthase \cite{?}, and as an activator for fructose-biphosphatase \cite{2000_donahue_JBacteriol,2006_hines_JBiolChem,2007_hines_JBiolChem}.
+AKG was modeled as an inhibitor for citrate synthase \cite{1994_pereira_JBiolChem,1983_robinson_FEBSLett}, isocitrate lyase \cite{1988_mackintosh_BiochemJ}, and malate synthase \cite{?}.
+3PG was modeled as an inhibitor for isocitrate lyase \cite{1988_mackintosh_BiochemJ} and malate synthase \cite{?}.
+FDP was modeled as an activator for pyruvate kinase \cite{2010_zhu_Biochimie} and PEP carboxylase \cite{1972_wohl_JBiolChem}.
+Pyruvate was modeled as an inhibitor for pyruvate dehydrogenase \cite{2007_kale_JBiolChem,2002_arjunan_Biochemistry} and as an activator for lactate dehydrogenase \cite{2008_okino_ApplMicrobiolBiotechnol}.
+Acetyl CoA was modeled as an inhibitor for malate dehydrogenase \cite{?}.
+
 \subsection*{Generation of model ensemble}

### 2016-05-26 Paper_v1.tex
May 26

@@
 We added regulation to the network as informed by literature, for a total of 17 interactions.
-PEP was modeled as an inhibitor for phosphofructokinase \cite{2011_cabrera_JBiolChem}, PEP carboxykinase \cite{2003_ravanal_JProteinChem}, PEP synthetase \cite{1973_chulavatnatol_JBiolChem}, isocitrate dehydrogenase \cite{2007_ogawa_JBacteriol}, isocitrate lyase \cite{2007_ogawa_JBacteriol,1988_mackintosh_BiochemJ}, and malate synthase \cite{?}, and as an activator for fructose-biphosphatase \cite{2000_donahue_JBacteriol,2006_hines_JBiolChem,2007_hines_JBiolChem}.
-AKG was modeled as an inhibitor for citrate synthase \cite{1994_pereira_JBiolChem,1983_robinson_FEBSLett}, isocitrate lyase \cite{1988_mackintosh_BiochemJ}, and malate synthase \cite{?}.
-3PG was modeled as an inhibitor for isocitrate lyase \cite{1988_mackintosh_BiochemJ} and malate synthase \cite{?}.
-FDP was modeled as an activator for pyruvate kinase \cite{2010_zhu_Biochimie} and PEP carboxylase \cite{1972_wohl_JBiolChem}.
-Pyruvate was modeled as an inhibitor for pyruvate dehydrogenase \cite{2007_kale_JBiolChem,2002_arjunan_Biochemistry} and as an activator for lactate dehydrogenase \cite{2008_okino_ApplMicrobiolBiotechnol}.
-Acetyl CoA was modeled as an inhibitor for malate dehydrogenase \cite{?}.
+PEP was modeled as an inhibitor for phosphofructokinase \cite{2010_kotte_MolSystBiol,2011_cabrera_JBiolChem}, PEP carboxykinase \cite{2010_kotte_MolSystBiol,2003_ravanal_JProteinChem}, PEP synthetase \cite{2010_kotte_MolSystBiol,1973_chulavatnatol_JBiolChem}, isocitrate dehydrogenase \cite{2010_kotte_MolSystBiol,2007_ogawa_JBacteriol}, isocitrate lyase \cite{2010_kotte_MolSystBiol,2007_ogawa_JBacteriol,1988_mackintosh_BiochemJ}, and malate synthase \cite{?}, and as an activator for fructose-biphosphatase \cite{2010_kotte_MolSystBiol,2000_donahue_JBacteriol,2006_hines_JBiolChem,2007_hines_JBiolChem}.
+AKG was modeled as an inhibitor for citrate synthase \cite{2010_kotte_MolSystBiol,1994_pereira_JBiolChem,1983_robinson_FEBSLett}, isocitrate lyase \cite{2010_kotte_MolSystBiol,1988_mackintosh_BiochemJ}, and malate synthase \cite{?}.
+3PG was modeled as an inhibitor for isocitrate lyase \cite{2010_kotte_MolSystBiol,1988_mackintosh_BiochemJ} and malate synthase \cite{?}.
+FDP was modeled as an activator for pyruvate kinase \cite{2010_kotte_MolSystBiol,2010_zhu_Biochimie} and PEP carboxylase \cite{2010_kotte_MolSystBiol,1972_wohl_JBiolChem}.
+Pyruvate was modeled as an inhibitor for pyruvate dehydrogenase \cite{2010_kotte_MolSystBiol,2007_kale_JBiolChem,2002_arjunan_Biochemistry} and as an activator for lactate dehydrogenase \cite{2008_okino_ApplMicrobiolBiotechnol}.
+Acetyl CoA was modeled as an inhibitor for malate dehydrogenase \cite{2010_kotte_MolSystBiol}.
 

### 2016-05-26 Paper_v1.tex
May 26_2

--- 2016-05-24 Paper_v1.tex
@@
 
+We added regulation to the network as informed by literature, for a total of 17 interactions.
+PEP was modeled as an inhibitor for phosphofructokinase \cite{2010_kotte_MolSystBiol,2011_cabrera_JBiolChem}, PEP carboxykinase \cite{2010_kotte_MolSystBiol}, PEP synthetase \cite{2010_kotte_MolSystBiol,1973_chulavatnatol_JBiolChem}, isocitrate dehydrogenase \cite{2010_kotte_MolSystBiol,2007_ogawa_JBacteriol}, and isocitrate lyase/malate synthase \cite{2010_kotte_MolSystBiol,2007_ogawa_JBacteriol,1988_mackintosh_BiochemJ}, and as an activator for fructose-biphosphatase \cite{2010_kotte_MolSystBiol,2000_donahue_JBacteriol,2006_hines_JBiolChem,2007_hines_JBiolChem}.
+AKG was modeled as an inhibitor for citrate synthase \cite{2010_kotte_MolSystBiol,1994_pereira_JBiolChem,1983_robinson_FEBSLett} and isocitrate lyase/malate synthase \cite{2010_kotte_MolSystBiol,1988_mackintosh_BiochemJ}.
+3PG was modeled as an inhibitor for isocitrate lyase/malate synthase \cite{2010_kotte_MolSystBiol,1988_mackintosh_BiochemJ}.
+FDP was modeled as an activator for pyruvate kinase \cite{2010_kotte_MolSystBiol,2010_zhu_Biochimie} and PEP carboxylase \cite{2010_kotte_MolSystBiol,1972_wohl_JBiolChem}.
+Pyruvate was modeled as an inhibitor for pyruvate dehydrogenase \cite{2010_kotte_MolSystBiol,2007_kale_JBiolChem,2002_arjunan_Biochemistry} and as an activator for lactate dehydrogenase \cite{2008_okino_ApplMicrobiolBiotechnol}.
+Acetyl CoA was modeled as an inhibitor for malate dehydrogenase \cite{2010_kotte_MolSystBiol}.
+
 \subsection*{Generation of model ensemble}

### 2016-05-31 Paper_v1.tex
May31

@@
 We constructed an initial model of cell-free \textit{E.
-coli} metabolism by removing the growth-associated processes from the genome-scale model of Palsson and coworkers \cite{?} and adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset from Swartz and coworkers \cite{?}.
+coli} metabolism by removing the growth-associated processes from the genome-scale model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS} and adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
 This initial model captured the core metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle), as well as the synthesis of energy species, amino acids, and CAT.
@@
 We used multiple saturation kinetics to model metabolic fluxes and mass action kinetics to model enzyme degradation.
-We modeled the control term using a rule-based approach in which each control factor had a regulatory transfer function, modeled as a Hill function \cite{?}, and the control term was calculated as the mean of all transfer functions.
-%The model was trained against the Swartz dataset, which included measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (ATP, ADP, AMP, GTP, GDP, GMP, CTP, CDP, CMP, UTP, UDP, UMP), and 18 of the 20 proteinogenic amino acids.
-The model was trained against the Swartz dataset, which included measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (AXP, GXP, CXP, UXP), and 18 of the 20 proteinogenic amino acids.
-We generated an ensemble of 18,000 parameter sets by minimizing the error between the Swartz dataset and the metabolite concentrations predicted by the model.
+We modeled the control term using a rule-based approach in which each control factor had a regulatory transfer function, modeled as a Hill function, and the control term was calculated as the mean of all transfer functions.
+%The model was trained against the training dataset, which included measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (ATP, ADP, AMP, GTP, GDP, GMP, CTP, CDP, CMP, UTP, UDP, UMP), and 18 of the 20 proteinogenic amino acids.
+The model was trained against the training dataset, which included measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (AXP, GXP, CXP, UXP), and 18 of the 20 proteinogenic amino acids.
+We generated an ensemble of 18,000 parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model.
 Out of these 18,000 sets, we defined the set with the lowest error value as our best-fit set.
@@
 Of the saturation constants, the reverse is seen: the largest positive sensitivities are those associated with arginine synthesis, while the largest negative sensitivities are those associated with CAT synthesis, GTP synthesis, and GMP synthesis (Fig.~\ref{fig:LocalSensitivity}, middle).
-This is because an increase in saturation constant causes a decrease in the corresponding reaction rate \cite{?}.
+This is because an increase in saturation constant causes a decrease in the corresponding reaction rate.
 The control parameters were seen to be the least significant and the most uncertain (Fig.~\ref{fig:LocalSensitivity}, bottom).
@@
 We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
-Arginine and glutamate were excluded due to not being present in the Swartz dataset.
+Arginine and glutamate were excluded due to not being present in the training dataset.
 The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
@@
 \subsection*{Generation of model ensemble}
-We generated an ensemble of 18,000 parameter sets via a downhill-only random walk Monte Carlo method \cite{?}.
+We generated an ensemble of 18,000 parameter sets via a downhill-only random walk Monte Carlo method.
 Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-absolute-error between experimental data and model predictions:

### 2016-06-01 Paper_v1.tex
may31_3

@@
 \section*{Abstract}
-Fill me in.
 
-%We further tested the predictive power of the coagulation model parameters against data not used in training, and found good agreement between simulations and experimental measurements.
-Lastly, we tested the performance of DOPS on commonly used test functions for global optimization and on published biochemical parameter estimation benchmark problems.
-For the wide range of problems that we considered, DOPS outperformed other meta-heuristic approaches despite a limited number of function evaluations.
+In this study, we present a framework for dynamic, cell-free metabolic modeling, integrating a simple logical description of regulation with traditional enzyme kinetics.
+Using this framework, we have constructed an ensemble of models for production of chloramphenicol acetyltransferase in a cell-free \textit{E. coli} system.
+Our ensemble fits measurements of glucose, chloramphenicol acetyltransferase, organic acids, and energy species, but fails to capture much of the amino acid dynamics in the dataset.
+Our model predicts a carbon yield of 0.0871 for chloramphenicol acetyltransferase synthesis, equal to 45\% of the theoretical maximum calculated with constraint-based modeling.
+It also predicts most carbon flux going through glycolysis and the TCA cycle, while the constraint-based approach predicts most carbon flux going through the Entner-Doudoroff pathway and only part of glycolysis.
+Our sensitivity analysis indicates that chloramphenicol acetyltransferase synthesis is most sensitive to the parameters associated with the synthesis of GTP, GMP, and amino acids, and relatively insensitive to glucose and oxygen levels and allosteric control.
+However, allosteric control is important for the model to accurately predict the dynamics of some organic acids.
+This modeling framework represents a significant achievement toward genome-scale, dynamic modeling of \textit{E. coli} cell-free protein synthesis.
 
@@
 {\noindent \textbf{Keywords:}~Biochemical engineering, systems biology, cell free protein synthesis}
-
-% Extra abstract
-% Mathematical modeling of biological systems with multiple feed back loops is one such area where parameter estimation is a difficult non-linear optimization problem.
-This difficulty is further compounded when dealing with parameter vectors of high dimensions.
-
-%In this study, we present the dynamically dimensioned particle
-%a novel meta-heuristic approach that combines a variant of particle swarm optimization (PSO) along with dynamically dimensioned search (DDS) to obtain near optimal solutions of high dimensional biochemical networks within a relatively few function evaluations.
-%We use a particle swarm optimization technique that uses multi-swarms to generate candidate vectors which are then greedily updated using DDS by dynamically varying the perturbed parameter dimensions.
-We tested this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and also against DDS alone.
-The new algorithm outperforms all the other meta-heuristics on the coagulation model.
-The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data.
-We also performed this comparison on commonly used test functions (Ackley and Rosenbrock) for global optimization and found the same behavior.
-Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach.
-We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.
 
@@
 The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulation with traditional kinetic pathway modeling.
-%This integration allows the description of complex regulatory interactions, such as time-dependent allosteric regulation of enzyme activity, in the absence of specific mechanistic information.
 This integration allows the description of complex regulatory interactions in the absence of specific mechanistic information.
 The regulatory rules are easy to understand, easy to formulate and do not rely on overarching theoretical abstractions or restrictive assumptions.
-%We tested our approach by modeling the time evolution of several hypothetical cell-free metabolic networks.
-%In particular, we tested whether our effective modeling approach could describe classically expected enzyme kinetic behavior, and second whether we could simultaneously estimate kinetic parameters and regulatory connectivity, in the absence of specific mechanistic knowledge, from synthetic experimental data.
-%Toward these questions, we explored five hypothetical cell-free networks.
-%Each network shared the same enzymatic connectivity, but had different allosteric regulatory connectivity.
-%We found that simple effective rules, when integrated with traditional enzyme kinetic expressions, captured complex allosteric patterns such as ultrasensitivity or non-competitive inhibition in the absence of mechanistic information.
-Second, when integrated into network models, these rules captured classical regulatory patterns such as product-induced feedback inhibition.
-%Lastly, we showed, at least for the network architectures considered here, that we could simultaneously estimate kinetic parameters and allosteric connectivity from synthetic data
-%starting from an unbiased collection of possible allosteric structures using particle swarm optimization.
-However, when starting with an initial population that was heavily enriched with incorrect structures, our particle swarm approach could converge to an incorrect structure.
+Using this approach, we have built an ensemble of models that predict an experimental dataset for CAT production in a cell-free \textit{E. coli} system.
+We calculated the CAT yield for the ensemble at 45\% of the theoretical maximum according to constraint-based modeling, and determined through sensitivity analysis that CAT production was most sensitive to the parameters associated with CAT synthesis and the synthesis of GTP, GMP, and amino acids.
 While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
-
-%The introduction has four paragraphs (introduction no longer than 3 pages). Follow the cell free paper from last year:
-%\begin{enumerate}
-%	\item{\textbf{First~paragraph}: Introduce mathematical modeling, and its role in biochemical engineering. }
-%	\item{\textbf{Second~paragraph}: Contrast current static metabolic modeling approaches e.g., FBA with dynamic models.}
-%	\item{\textbf{Third~paragraph}: Introduce cell free protein synthesis.}
-%	\item{\textbf{Fourth~paragraph}: In this study, [Repeat the abstract with some additional detail]. Taken together, [killer statement].}
-%\end{enumerate}
-
-% Extra text from the introduction -
-%With the capacity to construct very large models using various mathematical formalisms \cite{chen2009input,tasseff2011modeling,luan2007computationally,mo2007genome,orth2011comprehensive,karr2012whole,buchel2013path2models,smallbone2010towards}, more often than not,
-
-%this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics  The new algorithm outperforms all the other meta-heuristics on the coagulation model.
-The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data.
-We also performed this comparison on  and found the same behavior.
-Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach.
-We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.
-%took into cognizance the fact that it may not be necessary to obtain the exact solution for high dimensional biological systems and that good enough solutions can be quickly obtained without expending a lot of objective function evaluations.
-Hence our current approach uses the power of population based heuristics along with DDS to obtain near optimal or good solutions.
-Though in theory we can combine any meta-heuristic with DDS, for the current purpose we used Particle Swarm Optimization (PSO).
-PSO unlike Genetic Algorithm (GA) or Differential Evolution (DE) does not have complex operations like cross over, mutation or recombination.
-It is simple to use and does not have a lot of parameters associated with other heuristics.
-However PSO is known to rapidly converge to a local optimum and thus several variants \cite{peer2003using,zhan2009adaptive,li2007fast} have been developed of which Multi swarm particle swarm optimization approaches (MLSPSO) \cite{zhao2008dynamic,liang2005dynamic} are one of the effective ones.
-We used MLSPSO to preclude bad regions of search and thereafter search within these regions using DDS.
 
@@
 coli} metabolism by removing the growth-associated processes from the genome-scale model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS} and adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
-This initial model captured the core metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle), as well as the synthesis of energy species, amino acids, and CAT.
+This initial model captured the core metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle), as well as the synthesis of energy species, amino acids, and CAT  (Fig.~\ref{fig:Network}).
 We generated mass balance equations around metabolites and enzymes in the network, modeled as ordinary differential equations with reaction rates equal to the product of a kinetic term and a control term.
@@
 Out of these 18,000 sets, we defined the set with the lowest error value as our best-fit set.
+The ensemble predicts most carbon flux going through glycolysis and the TCA cycle, and to a lesser extent the Entner-Doudoroff pathway  (Fig.~\ref{fig:Network}, bottom two rows of gray boxes).
+Pentose phosphate is much less utilized.
+We compared this flux profile with the steady-state fluxes calculated by sequence-specific FBA (Fig.~\ref{fig:Network}, top row of gray boxes).
+In the steady-state case, carbon flux avoids the upper reactions of glycolysis in favor of the Entner-Doudoroff pathway.
+It also not does participate in the TCA cycle, nor the reactions of pentose phosphate that are not required for Entner-Doudoroff.
+
 The ensemble captures the central carbon metabolism, including glucose uptake, CAT production, and the dynamics of the organic acid intermediates (Fig.~\ref{fig:BothCarbon}, top).

### 2016-06-01 Paper_v1.tex
First set of changes

@@
 {\par \centering {$^{1}$School of Chemical Engineering}}
-{\par \centering {Stanford University, Stanford, California 94305}}
-{\par \centering \textbf{Running Title:}~Dynamic model of cell free protein synthesis}
+{\par \centering {Stanford University, Stanford, CA 94305}}
+{\par \centering \textbf{Running Title:}~Dynamic modeling of cell free protein synthesis}
 \vspace{0.1in}
@@
 \section*{Abstract}
-
-In this study, we present a framework for dynamic, cell-free metabolic modeling, integrating a simple logical description of regulation with traditional enzyme kinetics.
-Using this framework, we have constructed an ensemble of models for production of chloramphenicol acetyltransferase in a cell-free \textit{E. coli} system.
-Our ensemble fits measurements of glucose, chloramphenicol acetyltransferase, organic acids, and energy species, but fails to capture much of the amino acid dynamics in the dataset.
-Our model predicts a carbon yield of 0.0871 for chloramphenicol acetyltransferase synthesis, equal to 45\% of the theoretical maximum calculated with constraint-based modeling.
-It also predicts most carbon flux going through glycolysis and the TCA cycle, while the constraint-based approach predicts most carbon flux going through the Entner-Doudoroff pathway and only part of glycolysis.
-Our sensitivity analysis indicates that chloramphenicol acetyltransferase synthesis is most sensitive to the parameters associated with the synthesis of GTP, GMP, and amino acids, and relatively insensitive to glucose and oxygen levels and allosteric control.
-However, allosteric control is important for the model to accurately predict the dynamics of some organic acids.
-This modeling framework represents a significant achievement toward genome-scale, dynamic modeling of \textit{E. coli} cell-free protein synthesis.
+Cell-free systems offer many advantages for the study, manipulation and modeling of bacterial metabolism.
+In this study, we developed an ensemble of dynamic \textit{E. coli} cell free protein synthesis models.
+We trained the model using dynamic measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
+We generated an ensemble of N = 18,000 parameter sets that described the experimental data, with the exception of some of the amino acid dynamics.
+Analysis of the model ensemble predicted a carbon yield of 0.0871 for chloramphenicol acetyltransferase synthesis,
+equal to 45\% of the maximum theoretical yield calculated by sequence specific constraint based modeling.
+The ensemble of models captured the dynamic distribution of metabolic flux.
+The ensemble of dynamic models predicted that most of the carbon flux going through glycolysis and the TCA cycle,
+while the constraint based approach predicted most of the carbon flux went through the Entner-Doudoroff pathway and only part of glycolysis.
+Local and global sensitivity analysis indicated that chloramphenicol acetyltransferase production was most sensitive to parameters associated with the synthesis of GTP, GMP, and amino acids, and relatively insensitive to allosteric control mechanisms, and glucose and oxygen levels.
 
@@
 
+%In this study, we present a framework for dynamic, cell-free metabolic modeling, integrating a simple logical description of regulation with traditional enzyme kinetics.
+%Using this framework, we have constructed an ensemble of models for production of chloramphenicol acetyltransferase in a cell-free \textit{E. coli} system.
+%Our ensemble fits measurements of glucose, chloramphenicol acetyltransferase, organic acids, and energy species, but fails to capture much of the amino acid dynamics in the dataset.
+
 % Uncomment in production -
-%\linenumbers
+\linenumbers
 
 \section*{Introduction}
+
+Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
+Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall.
+This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
+Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
+Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \citep{Jewett:2008aa}.
+However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}.
+Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}.
+Interestingly, many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system.
+For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth.
+Thus, cell-free operation holds several significant advantages for model development, identification and validation.
+Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms,
+such as \textit{E. coli} or \textit{B.~subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
 
@@
 
-Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
-Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall.
-This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
-Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
-Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \citep{Jewett:2008aa}.
-However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}.
-Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}.
-Interestingly, many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system.
-For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth.
-Thus, cell-free operation holds several significant advantages for model development, identification and validation.
-Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms,
-such as \textit{E. coli} or \textit{B.~subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
-
 In this study, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
@@
 
-\subsection*{Estimation of an ensmeble of models}
-We constructed an initial model of cell-free \textit{E.
-coli} metabolism by removing the growth-associated processes from the genome-scale model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS} and adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
-This initial model captured the core metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle), as well as the synthesis of energy species, amino acids, and CAT  (Fig.~\ref{fig:Network}).
-We generated mass balance equations around metabolites and enzymes in the network, modeled as ordinary differential equations with reaction rates equal to the product of a kinetic term and a control term.
+%We generated mass balance equations around metabolites and enzymes in the network, modeled as ordinary differential equations with reaction rates equal to the product of a kinetic term and a control term.
 We used multiple saturation kinetics to model metabolic fluxes and mass action kinetics to model enzyme degradation.
-We modeled the control term using a rule-based approach in which each control factor had a regulatory transfer function, modeled as a Hill function, and the control term was calculated as the mean of all transfer functions.
+%We modeled the control term using a rule-based approach in which each control factor had a regulatory transfer function, modeled as a Hill function, and the control term was calculated as the mean of all transfer functions.
 %The model was trained against the training dataset, which included measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (ATP, ADP, AMP, GTP, GDP, GMP, CTP, CDP, CMP, UTP, UDP, UMP), and 18 of the 20 proteinogenic amino acids.
-The model was trained against the training dataset, which included measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (AXP, GXP, CXP, UXP), and 18 of the 20 proteinogenic amino acids.
-We generated an ensemble of 18,000 parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model.
-Out of these 18,000 sets, we defined the set with the lowest error value as our best-fit set.
-The ensemble predicts most carbon flux going through glycolysis and the TCA cycle, and to a lesser extent the Entner-Doudoroff pathway  (Fig.~\ref{fig:Network}, bottom two rows of gray boxes).
+%The ensemble predicts most carbon flux going through glycolysis and the TCA cycle, and to a lesser extent the Entner-Doudoroff pathway  (Fig.~\ref{fig:Network}, bottom two rows of gray boxes).
 Pentose phosphate is much less utilized.
-We compared this flux profile with the steady-state fluxes calculated by sequence-specific FBA (Fig.~\ref{fig:Network}, top row of gray boxes).
-In the steady-state case, carbon flux avoids the upper reactions of glycolysis in favor of the Entner-Doudoroff pathway.
-It also not does participate in the TCA cycle, nor the reactions of pentose phosphate that are not required for Entner-Doudoroff.
+%We compared this flux profile with the steady-state fluxes calculated by sequence-specific FBA (Fig.~\ref{fig:Network}, top row of gray boxes).
+%In the steady-state case, carbon flux avoids the upper reactions of glycolysis in favor of the Entner-Doudoroff pathway.
+%It also not does participate in the TCA cycle, nor the reactions of pentose phosphate that are not required for Entner-Doudoroff.
 
-The ensemble captures the central carbon metabolism, including glucose uptake, CAT production, and the dynamics of the organic acid intermediates (Fig.~\ref{fig:BothCarbon}, top).
+\subsection*{Estimation of an ensemble of cell free protein synthesis models.}
+We used the hybrid cell free modeling framework of Wayman et al. to simulate the production of a model protein [REFHERE].
+The cell-free \textit{E.
+coli} metabolic model was constructed by removing the growth-associated processes from the model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS}, and by adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
+Thus, the model described core central carbon metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle),
+as well as the synthesis of energy species, amino acids biosynthesis and degradation, and biosynthesis of the CAT protein.
+An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids.
+We generated an ensemble of N = 18,000 parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model.
+We defined the set with the lowest error value as the best-fit parameter set.
+[STATISTICS ON PARAMETERS].
+
+The ensemble of models captured the time evolution of cell free CAT biosynthesis (Fig.~\ref{fig:BothCarbon} - \ref{fig:Amino}).
+Glucose was exhausted with 3 hr [FILL ME IN].
+The ensemble also captured the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:BothCarbon}, top) and the totals by base .
+The ensemble and the best-fit set also predict some of the amino acid measurements, while failing to predict others (Fig.~\ref{fig:Amino}).
+the central carbon metabolism, including glucose uptake, CAT production, and the dynamics of the organic acid intermediates .
 Allosteric control is important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:BothCarbon}, bottom).
-The ensemble also captures the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:BothCarbon}, top) and the totals by base (Fig.~\ref{fig:Energy}).
-The ensemble and the best-fit set also predict some of the amino acid measurements, while failing to predict others (Fig.~\ref{fig:Amino}).
 This is likely due to a structural deficiency in the model; in some cases, the consumption of an amino acid through CAT synthesis is not enough to explain the decrease shown in the data, and there are no other reactions that consume it.
 Thus, a more comprehensive biological description is needed to fully explain amino acid dynamics.
+
+\subsection*{Maximum theoretical CAT yield showed CFPS can be optimized.}
+We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
+Arginine and glutamate were excluded due to not being present in the training dataset.
+The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
+We then used sequence-specific FBA to calculate a theoretical maximum CAT yield of 0.1942.
+Thus, we showed that our experimental dataset and best-fit parameter set were each producing CAT at 45\% of the theoretical maximum.
+This allowed us to quantify the amount of carbon being diverted to byproducts, and suggests that there is potential for a doubling of CAT production by reducing this diversion of carbon.
 
@@
 
-\subsection*{Calculation of CAT yield}
-We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
-Arginine and glutamate were excluded due to not being present in the training dataset.
-The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
-We then used sequence-specific FBA to calculate a theoretical maximum CAT yield of 0.1942.
-Thus, we showed that our experimental dataset and best-fit parameter set were each producing CAT at 45\% of the theoretical maximum.
-This allowed us to quantify the amount of carbon being diverted to byproducts, and suggests that there is potential for a doubling of CAT production by reducing this diversion of carbon.
-
 \clearpage
@@
 
-We constructed a model of cell-free \textit{E. coli} metabolism with CAT production based on saturation kinetics and allosteric control.
-We trained the model against measurements of glucose, CAT, organic acid intermediates, energy species, and amino acids.
-We generated an ensemble of 18,000 parameter sets that predict the experimental dataset, with the exception of some amino acids.
-We calculated the carbon yield of CAT production at 45\% of the theoretical maximum obtained by flux balance analysis; this was in line with the CAT yield shown by the dataset.
-We conducted sensitivity analyses and determined that CAT production was most sensitive to the rate constants, saturation constants, and enzyme initial conditions associated with CAT synthesis, GTP synthesis, GMP synthesis, and amino acid synthesis, and to a lesser extent some amino acid initial conditions.
-CAT production was much less sensitive to control parameters, suggesting that while allosteric control affects some organic acid intermediates, it does not limit CAT production.
-The initial conditions of glucose and oxygen were even less important to CAT production, suggesting that it can be driven by other means.
-However, the low sensitivities to glucose and oxygen only apply in the specific case of the parameter sets that were studied.
-When other species' initial conditions and associated parameters are different, glucose and oxygen can become more important to CAT production.
+In this study, we developed an ensemble of dynamic \textit{E. coli} cell free protein synthesis models.
+We trained the model using dynamic measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
+We generated an ensemble of N = 18,000 parameter sets which described the experimental measurements, with the exception of some of the amino acid dynamics.
+To gauge the performance of the cell free reaction mixture, we compared the observed carbon yield, with the maximum theoretical carbon yield calculated using sequence specific flux
+balance analysis. The observed CAT carbon yield was 45\% of the maximum theoretical yield.
+The predicted metabolic flux distributions between the dynamic model and constraints based calculation were significantly different.
+The ensemble of dynamic models predicted carbon flux going through glycolysis and the TCA cycle, while the constraint based approach predicted most of the carbon flux went through the Entner-Doudoroff pathway.
+We then conducted sensitivity analysis to determine which model parameters and experimentally controllable factors influenced reaction performance.
+CAT production was most sensitive to model parameters and initial conditions directly associated with CAT synthesis.
+CAT production was also sensitive to GTP/GMP synthesis, and amino acid synthesis, and to a lesser extent some amino acid initial conditions.
+On the other hand, CAT production was much less sensitive to allosteric control parameters,
+suggesting that while allosteric control affected some organic acid intermediates, it was not critical to CAT production.
+The initial conditions of glucose and oxygen were also less important to CAT production.
+Taken together, we presented the first dynamic model of \textit{E. coli} cell free protein synthesis.
+We integrated traditional kinetics with logical rule-based allosteric control to construct an ensemble of kinetic models that described
+a comprehensive cell-free dataset. This study should provide a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} metabolism, toward industrial-scale biosynthetic production.
 
-In this study, we present the first dynamic, cell-free model of \textit{E. coli} metabolism and biosynthesis at this scope.
-While the early models of Shuler and coworkers achieved large-scale, dynamic descriptions of single cells \citep{1984_domach_shuler_BiotechBioeng_01}, and the stoichiometric models associated with the FBA approach are computationally efficient amd widespread \citep{2009_feist_palsson_NatRevMicrobio}, none have yet been able to construct genome-scale kinetic models of cell-free \textit{E.
+%The ensemble of models captured the dynamic distribution of metabolic flux.
+%The ensemble of dynamic models predicted that most of the carbon flux going through glycolysis and the TCA cycle,
+%while the constraint based approach predicted most of the carbon flux went through the Entner-Doudoroff pathway and only part of glycolysis.
+
+% We present the first dynamic, cell-free model of \textit{E. coli} metabolism and biosynthesis at this scope.
+% While the early models of Shuler and coworkers achieved large-scale, dynamic descriptions of single cells \citep{1984_domach_shuler_BiotechBioeng_01}, and the stoichiometric models associated with the FBA approach are computationally efficient amd widespread \citep{2009_feist_palsson_NatRevMicrobio}, none have yet been able to construct genome-scale kinetic models of cell-free \textit{E.
 coli} metabolism.
-We have harnessed the advantages of cell-free systems (no cell wall, no cell growth) and integrated traditional kinetics with allosteric control to build an ensemble of kinetic models that fit a comprehensive cell-free dataset, toward genome-scale.
-This study should provide a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} metabolism, toward industrial-scale biosynthetic production.
+% This study should provide a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} metabolism, toward industrial-scale biosynthetic production.
+%However, the low sensitivities to glucose and oxygen only apply in the specific case of the parameter sets that were studied.
+%When other species' initial conditions and associated parameters are different, glucose and oxygen could become more important to CAT production.
 
@@
 Testing the model under a variety of conditions could strengthen or challenge the findings of this study.
-Further experimentatiion could also be used to gain a deeper understanding of model performance under a variety of conditions.
+Further experimentation could also be used to gain a deeper understanding of model performance under a variety of conditions.
 Specifically, CAT production performed in the absence of amino acids could inform the system's ability to manufacture them, while experimentation in the absence of glucose or oxygen could shed light on how important they are to protein synthesis, and under which conditions.
@@
 
-\section*{Acknowledgements}
-This study was supported by an award from [FILL ME IN].
+\section*{Funding}
+This study was supported by a National Science Foundation Graduate Research Fellowship (DGE-1333468) to N.H and by an award from
+the US Army and Systems Biology of Trauma Induced Coagulopathy (W911NF-10-1-0376) to J.V for the support of M.V.
+
 \clearpage

### 2016-06-04 Paper_v1.tex
Jun3

@@
 \section*{Abstract}
-Cell-free systems offer many advantages for the study, manipulation and modeling of bacterial metabolism.
-In this study, we developed an ensemble of dynamic \textit{E. coli} cell free protein synthesis models.
-We trained the model using dynamic measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
-We generated an ensemble of N = 18,000 parameter sets that described the experimental data, with the exception of some of the amino acid dynamics.
-Analysis of the model ensemble predicted a carbon yield of 0.0871 for chloramphenicol acetyltransferase synthesis,
-equal to 45\% of the maximum theoretical yield calculated by sequence specific constraint based modeling.
-The ensemble of models captured the dynamic distribution of metabolic flux.
-The ensemble of dynamic models predicted that most of the carbon flux going through glycolysis and the TCA cycle,
-while the constraint based approach predicted most of the carbon flux went through the Entner-Doudoroff pathway and only part of glycolysis.
-Local and global sensitivity analysis indicated that chloramphenicol acetyltransferase production was most sensitive to parameters associated with the synthesis of GTP, GMP, and amino acids, and relatively insensitive to allosteric control mechanisms, and glucose and oxygen levels.
+Cell free protein expression systems have become widely used in systems and synthetic biology.
+In this study, we developed an ensemble of dynamic \textit{E. coli} cell free protein synthesis (CFPS) models.
+Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
+The ensemble described the training data, with the exception of some of the amino acid dynamics.
+To gauge the performance of the cell free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
+balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
+The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
+The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
+while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
+Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
+GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
+On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
+Taken together, we presented the first dynamic model of \textit{E. coli} cell free protein synthesis.
+This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
 
@@
 
-In this study, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
-The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulation with traditional kinetic pathway modeling.
-This integration allows the description of complex regulatory interactions in the absence of specific mechanistic information.
-The regulatory rules are easy to understand, easy to formulate and do not rely on overarching theoretical abstractions or restrictive assumptions.
-Using this approach, we have built an ensemble of models that predict an experimental dataset for CAT production in a cell-free \textit{E. coli} system.
-We calculated the CAT yield for the ensemble at 45\% of the theoretical maximum according to constraint-based modeling, and determined through sensitivity analysis that CAT production was most sensitive to the parameters associated with CAT synthesis and the synthesis of GTP, GMP, and amino acids.
-While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
+In this study, we developed an ensemble of \textit{E. coli} cell free protein synthesis (CFPS) models using the hybrid cell free modeling approach of Wayman et al [REFHERE].
+Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
+The ensemble described the training data, with the exception of some of the amino acid dynamics.
+To gauge the performance of the cell free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
+balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
+The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
+The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
+while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
+Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
+GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
+On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
+Taken together, we presented the first dynamic model of \textit{E. coli} cell free protein synthesis.
+We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
+This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
+
+% In this study, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
+% The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulation with traditional kinetic pathway modeling.
+% This integration allows the description of complex regulatory interactions in the absence of specific mechanistic information.
+% The regulatory rules are easy to understand, easy to formulate and do not rely on overarching theoretical abstractions or restrictive assumptions.
+% Using this approach, we have built an ensemble of models that predict an experimental dataset for CAT production in a cell-free \textit{E. coli} system.
+% We calculated the CAT yield for the ensemble at 45\% of the theoretical maximum according to constraint-based modeling, and determined through sensitivity analysis that CAT production was most sensitive to the parameters associated with CAT synthesis and the synthesis of GTP, GMP, and amino acids.
+% While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
 
@@
 
-In this study, we developed an ensemble of dynamic \textit{E. coli} cell free protein synthesis models.
-We trained the model using dynamic measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
-We generated an ensemble of N = 18,000 parameter sets which described the experimental measurements, with the exception of some of the amino acid dynamics.
-To gauge the performance of the cell free reaction mixture, we compared the observed carbon yield, with the maximum theoretical carbon yield calculated using sequence specific flux
-balance analysis. The observed CAT carbon yield was 45\% of the maximum theoretical yield.
-The predicted metabolic flux distributions between the dynamic model and constraints based calculation were significantly different.
-The ensemble of dynamic models predicted carbon flux going through glycolysis and the TCA cycle, while the constraint based approach predicted most of the carbon flux went through the Entner-Doudoroff pathway.
-We then conducted sensitivity analysis to determine which model parameters and experimentally controllable factors influenced reaction performance.
-CAT production was most sensitive to model parameters and initial conditions directly associated with CAT synthesis.
-CAT production was also sensitive to GTP/GMP synthesis, and amino acid synthesis, and to a lesser extent some amino acid initial conditions.
-On the other hand, CAT production was much less sensitive to allosteric control parameters,
-suggesting that while allosteric control affected some organic acid intermediates, it was not critical to CAT production.
-The initial conditions of glucose and oxygen were also less important to CAT production.
-Taken together, we presented the first dynamic model of \textit{E. coli} cell free protein synthesis.
-We integrated traditional kinetics with logical rule-based allosteric control to construct an ensemble of kinetic models that described
-a comprehensive cell-free dataset. This study should provide a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} metabolism, toward industrial-scale biosynthetic production.
+In this study, we developed an ensemble of \textit{E. coli} cell free protein synthesis (CFPS) models using the hybrid cell free modeling approach of Wayman et al [REFHERE].
+Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
+The ensemble described the training data, with the exception of some of the amino acid dynamics.
+To gauge the performance of the cell free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
+balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
+The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
+The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
+while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
+Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
+GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
+On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
+% Taken together, we presented the first dynamic model of \textit{E. coli} cell free protein synthesis.
+% We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
+% This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
 
@@
 %while the constraint based approach predicted most of the carbon flux went through the Entner-Doudoroff pathway and only part of glycolysis.
-
 % We present the first dynamic, cell-free model of \textit{E. coli} metabolism and biosynthesis at this scope.
@@
 
-One important direction for the future is to expand the biological description, especially with regard to amino acids.
+The cell free model ensemble described the training data with the exception of some of the amino acids.
 Specifically, adding more reactions that consume amino acids would improve the model's ability to predict those that show a decrase in the experimental data.

### 2016-06-04 Code/Ensemble/src/Algorithm.jl

@@ -1,0 +1,142 @@
+include("DataDictionary.jl")
+include("SolveBalances.jl")
+include("CalcError.jl")
+using PyPlot
+
+# Create Ensemble directory if it doesn't already exist
+if ~isdir("Ensemble")
+    mkdir("Ensemble")
+end
+
+# Create Best directory if it doesn't already exist
+if ~isdir("Ensemble/Best")
+    mkdir("Ensemble/Best")
+end
+
+# Setup the timescale
+TSTART = 0.0
+TSTOP = 3.0
+Ts = 0.1
+
+# Create dictionary of parameters
+data_dictionary_best = DataDictionary(TSTART,TSTOP,Ts)
+experiment_data_dictionary = data_dictionary_best["EXPERIMENT_DATA_DICTIONARY"]
+
+#data_indices = Int64[]
+#for (key,value) in experiment_data_dictionary
+#    push!(data_indices,value.model_index)
+#end
+#sort!(data_indices)
+
+#init_indices = collect(1:103)
+#deleteat!(init_indices,data_indices)
+
+# Initialize number of parameters to be varied
+num_rate = 163
+#num_init = 66
+num_init = 0
+num_sat = 455
+num_cont = 34
+
+# Create or read the relevant information
+if ~isfile("Ensemble/Best/params")
+    # Take rate constants to be varied from data_dictionary_best
+    rate = data_dictionary_best["RATE_CONSTANT_ARRAY"][1:163]
+    # Take initial conditions to be varied from data_dictionary_best
+#    init = data_dictionary_best["INITIAL_CONDITION_ARRAY"][init_indices]
+    # Take saturation constants to be varied from data_dictionary_best
+    sat_vector = vec(data_dictionary_best["SATURATION_CONSTANT_ARRAY"])
+    sat = sat_vector[find(sat_vector.>0)]
+    # Take control parameters to be varied from data_dictionary_best
+    cont = vec(data_dictionary_best["CONTROL_PARAMETER_ARRAY"]')
+    # Concatenate into params vector
+#    params = [rate;init;sat;cont]
+    params_best = [rate;sat;cont]
+    # Solve the balance equations and calculate cost
+    Tsim_best,X_best = SolveBalances(TSTART,TSTOP,Ts,data_dictionary_best)
+    cost_best = CalcError(experiment_data_dictionary,Tsim_best,X_best)
+    # Save the relevant information
[…]

### 2016-06-04 Code/Ensemble/src/CalcArea.jl

@@ -1,0 +1,17 @@
+function CalcArea(Tsim,X)
+
+if !(length(Tsim)==length(X))
+    throw("Vectors must be of same length")
+end
+
+AUC = Float64[]
+
+i_range = 2:length(Tsim)
+for i in i_range
+    push!(AUC,(Tsim[i]-Tsim[i-1])*(X[i]+X[i-1])/2) # Trapezoid rule
+end
+
+AUC = sum(AUC)
+
+return AUC
+end

### 2016-06-04 Code/Ensemble/src/Figure1.jl

@@ -1,0 +1,754 @@
+# MassBalances.jl is set up to save the flux vector at all timepoints satisfying t < 1e-11 or 1.48 < t < 1.52 or t > 2.98
+# At each timepoint, a vector of length 327 is saved to disk: the first element is the timepoint, followed by a flux value for each of the 326 reactions
+# This is done for each set in the ensemble (there are 18000 such sets)
+
+# For each set in the ensemble, we read all flux vectors and merge them into an array
+# We then find the timepoint closest to 1.5 hr and save the corresponding flux vector, under the directory for that set, under the name "flux_1_5"
+# We then find the timepoint closest to 3 hr and save the corresponding flux vector, under the directory for that set, under the name "flux_3"
+num_dir = convert(Int64,readdlm("Ensemble/dir")[1])
+for i in 1:num_dir
+    # Read number of timepoints
+    num_tp = convert(Int64,readdlm("./Ensemble/$i/Flux/num_tp")[1])
+    # Read first flux
+    rate_vector = readdlm("./Ensemble/$i/Flux/1")
+    # Read all other fluxes; append to first flux
+    j_range = 2:num_tp
+    for j in j_range
+        r = readdlm("./Ensemble/$i/Flux/$j")
+        rate_vector = [rate_vector r]
+    end
+    # Save flux at t = 0
+    tp = rate_vector[1,:]
+    index = find(tp.==minimum(tp))[1]
+    writedlm("./Ensemble/$i/flux_0",rate_vector[2:end,index])
+    # Save flux at t = 1.5
+    tp_1_5 = abs(tp-1.5)
+    index = find(tp_1_5.==minimum(tp_1_5))[1]
+    writedlm("./Ensemble/$i/flux_1_5",rate_vector[2:end,index])
+    # Save flux at t = 3
+    tp_3 = abs(tp-3.0)
+    index = find(tp_3.==minimum(tp_3))[1]
+    writedlm("./Ensemble/$i/flux_3",rate_vector[2:end,index])
[…]

### 2016-06-04 Code/Ensemble/src/Figure5.jl

@@ -1,0 +1,117 @@
+using PyPlot
+using PyCall
+@pyimport numpy as np
+
+include("Local_sens_fxn.jl")
+
+num_iter = 180 # Should be set equal to number of samples in sub-ensemble
+for i in 1:num_iter
+    Local_sens_fxn(i)
+end
+
+Rate = readdlm("Sens/1/Rate")
+Sat = readdlm("Sens/1/Sat")
+Cont = readdlm("Sens/1/Cont")
+
+for i in 1:num_iter
+    rate = readdlm("Sens/$i/Rate")
+    sat = readdlm("Sens/$i/Sat")
+    cont = readdlm("Sens/$i/Cont")
+    Rate = [Rate rate]
+    Sat = [Sat sat]
+    Cont = [Cont cont]
+end
+
+writedlm("Sens/Rate",Rate)
+writedlm("Sens/Sat",Sat)
+writedlm("Sens/Cont",Cont)
+
+Rate_m = Float64[]
+Rate_s = Float64[]
+for i in 1:size(Rate)[1]
+    rate = Rate[i,:]
+    push!(Rate_m,mean(rate))
+    push!(Rate_s,std(rate))
+end
+
+Sat_m = Float64[]
+Sat_s = Float64[]
+for i in 1:size(Sat)[1]
+    sat = Sat[i,:]
+    push!(Sat_m,mean(sat))
+    push!(Sat_s,std(sat))
+end
+
+Cont_m = Float64[]
+Cont_s = Float64[]
+for i in 1:size(Cont)[1]
+    cont = Cont[i,:]
+    push!(Cont_m,mean(cont))
+    push!(Cont_s,std(cont))
+end
+
+writedlm("Sens/Rate_m",Rate_m)
+writedlm("Sens/Rate_s",Rate_s)
+writedlm("Sens/Sat_m",Sat_m)
+writedlm("Sens/Sat_s",Sat_s)
+writedlm("Sens/Cont_m",Cont_m)
+writedlm("Sens/Cont_s",Cont_s)
+
+PyCall.PyDict(matplotlib["rcParams"])["font.sans-serif"] = ["Helvetica"]
+
+Rate_m = readdlm("Sens/Rate_m")
+Rate_s = readdlm("Sens/Rate_s")/sqrt(180)
+width = 1
+a = [1:2:2*length(Rate_m);]
+threshold = 0.05*ones(a)
+PyPlot.matplotlib[:rc]("text",usetex=true)
[…]

### 2016-06-04 Code/Ensemble/src/Figures234.jl

@@ -1,0 +1,582 @@
+include("DataDictionary.jl")
+include("SolveBalances.jl")
+using PyPlot
+
+# Setup the timescale
+TSTART = 0.0
+TSTOP = 3.0
+Ts = 0.1
+
+# Create dictionary of parameters (best-fit set)
+data_dictionary = DataDictionary(TSTART,TSTOP,Ts)
+
+#-----------------------------------------------------------------------------
+# In order to simulate predictions in the absence of allosteric control,
+# comment the line "rate_vector = rate_vector.*control_vector;" in Control.jl
+#-----------------------------------------------------------------------------
+
+# Solve the balances to calculate predictions based on best-fit set
+Tsim,X = SolveBalances(TSTART,TSTOP,Ts,data_dictionary)
+
+# 
+experimental_data_dictionary = data_dictionary["EXPERIMENT_DATA_DICTIONARY"]
+
+# Store the glucose, CAT, and organic acid raw data by species
+glc = experimental_data_dictionary["M_glc_D_c"].raw_measurement_array
+pyr = experimental_data_dictionary["M_pyr_c"].raw_measurement_array
+lac = experimental_data_dictionary["M_lac_D_c"].raw_measurement_array
+ac = experimental_data_dictionary["M_ac_c"].raw_measurement_array
+succ = experimental_data_dictionary["M_succ_c"].raw_measurement_array
+mal = experimental_data_dictionary["M_mal_L_c"].raw_measurement_array
+CAT = experimental_data_dictionary["CAT"].raw_measurement_array
+
+# Store the energy species raw data by species
+atp = experimental_data_dictionary["M_atp_c"].raw_measurement_array
+adp = experimental_data_dictionary["M_adp_c"].raw_measurement_array
[…]

### 2016-06-04 Code/Ensemble/src/MassBalances.jl

@@ -1,0 +1,93 @@
+include("Kinetics.jl");
+include("Control.jl");
+
+# ----------------------------------------------------------------------------------- #
+# Copyright (c) 2016 Varnerlab
+# School of Chemical Engineering Purdue University
+# W. Lafayette IN 46907 USA
+
+# Permission is hereby granted, free of charge, to any person obtaining a copy
+# of this software and associated documentation files (the "Software"), to deal
+# in the Software without restriction, including without limitation the rights 
+# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell 
+# copies of the Software, and to permit persons to whom the Software is 
+# furnished to do so, subject to the following conditions:
+#
+# The above copyright notice and this permission notice shall be included in
+# all copies or substantial portions of the Software.
+#
+# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
+# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
+# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
+# OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
+# THE SOFTWARE.
+# ----------------------------------------------------------------------------------- #
+function MassBalances(t,x,dxdt_vector,data_dictionary)
+# ---------------------------------------------------------------------- #
+# MassBalances.jl was generated using the Kwatee code generation system.
+# Username: nicholas
+# Type: CFPS-JULIA
+# Version: 1.0
+# Generation timestamp: 05-11-2016 16:22:26
+# 
+# Arguments: 
+# t  - current time 
+# x  - state vector 
+# dxdt_vector - right hand side vector 
+# data_dictionary  - Data dictionary instance (holds model parameters) 
+# ---------------------------------------------------------------------- #
+
[…]

### 2016-06-04 Code/Ensemble/src/PlotSingle.jl

@@ -1,0 +1,498 @@
+using PyPlot
+function PlotSingle(Tsim,X,data_dictionary)
+
+experimental_data_dictionary = data_dictionary["EXPERIMENT_DATA_DICTIONARY"]
+
+glc = experimental_data_dictionary["M_glc_D_c"].raw_measurement_array
+pyr = experimental_data_dictionary["M_pyr_c"].raw_measurement_array
+lac = experimental_data_dictionary["M_lac_D_c"].raw_measurement_array
+ac = experimental_data_dictionary["M_ac_c"].raw_measurement_array
+succ = experimental_data_dictionary["M_succ_c"].raw_measurement_array
+mal = experimental_data_dictionary["M_mal_L_c"].raw_measurement_array
+CAT = experimental_data_dictionary["CAT"].raw_measurement_array
+
+atp = experimental_data_dictionary["M_atp_c"].raw_measurement_array
+adp = experimental_data_dictionary["M_adp_c"].raw_measurement_array
+amp = experimental_data_dictionary["M_amp_c"].raw_measurement_array
+axp = atp+adp+amp
+axp[:,1] = atp[:,1]
+gtp = experimental_data_dictionary["M_gtp_c"].raw_measurement_array
+gdp = experimental_data_dictionary["M_gdp_c"].raw_measurement_array
+gmp = experimental_data_dictionary["M_gmp_c"].raw_measurement_array
+gxp = gtp+gdp+gmp
+gxp[:,1] = gtp[:,1]
+ctp = experimental_data_dictionary["M_ctp_c"].raw_measurement_array
+cdp = experimental_data_dictionary["M_cdp_c"].raw_measurement_array
+cmp = experimental_data_dictionary["M_cmp_c"].raw_measurement_array
+cxp = ctp+cdp+cmp
+cxp[:,1] = ctp[:,1]
+utp = experimental_data_dictionary["M_utp_c"].raw_measurement_array
[…]

### 2016-06-04 Code/Ensemble/src/SolveBalances.jl

@@ -1,0 +1,56 @@
+include("MassBalances.jl")
+using Sundials;
+
+# ----------------------------------------------------------------------------------- #
+# Copyright (c) 2016 Varnerlab
+# School of Chemical Engineering Purdue University
+# W. Lafayette IN 46907 USA
+
+# Permission is hereby granted, free of charge, to any person obtaining a copy
+# of this software and associated documentation files (the "Software"), to deal
+# in the Software without restriction, including without limitation the rights 
+# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell 
+# copies of the Software, and to permit persons to whom the Software is 
+# furnished to do so, subject to the following conditions:
+#
+# The above copyright notice and this permission notice shall be included in
+# all copies or substantial portions of the Software.
+#
+# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
+# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
+# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
+# OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
+# THE SOFTWARE.
+# ----------------------------------------------------------------------------------- #
+function SolveBalances(TSTART,TSTOP,Ts,data_dictionary)
+# ----------------------------------------------------------------------------------- #
+# SolveBalances.jl was generated using the Kwatee code generation system.
+# SolveBalances: Solves model equations from TSTART to TSTOP given parameters in data_dictionary.
+# Username: nicholas
+# Type: CFPS-JULIA
+# Version: 1.0
+# Generation timestamp: 05-11-2016 16:22:26
+# 
+# Input arguments: 
+# TSTART  - Time start 
+# TSTOP  - Time stop 
+# Ts - Time step 
+# data_dictionary  - Data dictionary instance (holds model parameters) 
+# 
+# Return arguments: 
+# TSIM - Simulation time vector 
[…]

### 2016-06-04 Code/Sobol/0/src/CalcError.jl

@@ -1,0 +1,28 @@
+using Interpolations
+
+function CalcError(experiment_data_dictionary,Tsim,X)
+# Initialize array
+Error = Float64[]
+for (key,dataset) in experiment_data_dictionary
+    # Get raw data and timepoints
+    raw_array = dataset.raw_measurement_array
+    data_tps = raw_array[:,1]
+    data_vals = raw_array[:,2]
+    # Interpolate simulation array to fit dataset
+    index = dataset.model_index
+    sim = X[:,index]
+    itp = interpolate((Tsim,),sim,Gridded(Linear()))
+    sim_itp = itp[data_tps]
+    # Calculate error
+    error = sum(abs(sim_itp-data_vals))/maximum(data_vals)
+    # Weight certain datasets as more important
+    weight = (key == "M_glc_D_c" || key == "CAT" || key == "M_pyr_c" || key == "M_lac_D_c" || key == "M_ac_c" || key == "M_succ_c" || key == "M_mal_L_c")?5:1
+    error *= weight
+    # Push to array
+    push!(Error,error)
+end
+
+cost = sum(Error)
+
+return cost
+end

### 2016-06-04 Code/Sobol/0/src/CreateRunDir.jl

@@ -1,0 +1,9 @@
+# Import number of runs; if this is first run, set equal to 1
+if isfile("./Sobol/num_run")
+    num_run = convert(Int64,readdlm("./Sobol/num_run")[1])+1
+else
+    num_run = 1
+end
+# Create directory for next run and save number of runs
+mkdir("./Sobol/Run$num_run")
+writedlm("./Sobol/num_run",num_run)

### 2016-06-04 Code/Sobol/0/src/CreateSobolDir.jl

@@ -1,0 +1,4 @@
+# If no directory called Sobol exists, create one
+if ~isdir("Sobol")
+    mkdir("Sobol")
+end

### 2016-06-04 Code/Sobol/0/src/Figure6.jl

@@ -1,0 +1,42 @@
+using PyPlot
+using PyCall
+@pyimport numpy as np
+
+PyCall.PyDict(matplotlib["rcParams"])["font.sans-serif"] = ["Helvetica"]
+
+Sobol = readdlm("SA.txt")
+Sobol_metab = Sobol[1:22,:]
+
+Sens = Sobol[:,4] # total order sensitivities
+CI = Sobol[:,5] # total order sensitivity confidence intervals
+
+Sens_metab = Sobol_metab[:,4] # total order sensitivities
+CI_metab = Sobol_metab[:,5] # total order sensitivity confidence intervals
+
+width = 1
+
+PyPlot.matplotlib[:rc]("text",usetex=true)
+
+# Plot all indices
+alpha = length(Sens)
+x = [1:2:2*alpha;]
+threshold = 0.05*ones(x)
+fig1, ax1 = plt[:subplots](figsize=(15,7))
+b = ax1[:bar](x,Sens,width,align="center",yerr=CI,ecolor="black",color="black")
+ax1[:set_xticks]([])
+plt[:margins](0.002)
+plt[:ylim]([0,1])
+plt[:yticks]([0:.2:1],fontsize=20);
+plt[:tick_params](axis="x",which="both",bottom="off",top="off")
+
+# Plot metabolite indices only
+alpha = length(Sens_metab)
+x = [1:2:2*alpha;]
+threshold = 0.05*ones(x)
+fig2, ax2 = plt[:subplots](figsize=(15,7))
+b = ax2[:bar](x,Sens_metab,width,align="center",yerr=CI_metab,ecolor="black",color="black")
+ax2[:set_xticks]([])
+plt[:margins](0.002)
+plt[:ylim]([0,.2])
+plt[:yticks]([0:.05:.2],fontsize=20);
+plt[:tick_params](axis="x",which="both",bottom="off",top="off")

### 2016-06-04 Code/Sobol/0/src/MassBalances.jl

@@ -1,0 +1,93 @@
+include("Kinetics.jl");
+include("Control.jl");
+
+# ----------------------------------------------------------------------------------- #
+# Copyright (c) 2016 Varnerlab
+# School of Chemical Engineering Purdue University
+# W. Lafayette IN 46907 USA
+
+# Permission is hereby granted, free of charge, to any person obtaining a copy
+# of this software and associated documentation files (the "Software"), to deal
+# in the Software without restriction, including without limitation the rights 
+# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell 
+# copies of the Software, and to permit persons to whom the Software is 
+# furnished to do so, subject to the following conditions:
+#
+# The above copyright notice and this permission notice shall be included in
+# all copies or substantial portions of the Software.
+#
+# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
+# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
+# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
+# OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
+# THE SOFTWARE.
+# ----------------------------------------------------------------------------------- #
+function MassBalances(t,x,dxdt_vector,data_dictionary)
+# ---------------------------------------------------------------------- #
+# MassBalances.jl was generated using the Kwatee code generation system.
+# Username: nicholas
+# Type: CFPS-JULIA
+# Version: 1.0
+# Generation timestamp: 05-11-2016 16:22:26
+# 
+# Arguments: 
+# t  - current time 
+# x  - state vector 
+# dxdt_vector - right hand side vector 
+# data_dictionary  - Data dictionary instance (holds model parameters) 
+# ---------------------------------------------------------------------- #
+
[…]

### 2016-06-04 Code/Sobol/0/src/ModelOutput.jl

@@ -1,0 +1,15 @@
+# Import number of jobs
+num_jobs = convert(Int64,readdlm("num_jobs")[1])
+
+# Initialize total output array
+Output = Float64[]
+
+# For each job, add job output to total output array
+i_range = 0:num_jobs-1
+for i in i_range
+    output = readdlm("../../$i/src/model_output.txt")
+    Output = [Output;output]
+end
+
+# Save in src
+writedlm("model_output.txt",Output)

### 2016-06-04 Code/Sobol/0/src/RunModel.jl

@@ -1,0 +1,11 @@
+function RunModel(num_sample)
+include("SolveBalances.jl")
+
+# Insert sobol values
+IC_vector = vec(readdlm("../../model_input/$num_sample"))
+
+# Solve balance equations
+Tsim,X = SolveBalances(IC_vector)
+
+return Tsim,X
+end

### 2016-06-04 Code/Sobol/0/src/RunSobol.jl

@@ -1,0 +1,41 @@
+include("RunModel.jl")
+include("CalcArea.jl")
+
+# Import number of runs
+num_run = convert(Int64,readdlm("./Sobol/num_run")[1])
+
+# Import number of samples; if this is first sample, set equal to 1
+if isfile("./Sobol/Run$num_run/num_sample")
+    num_sample = convert(Int64,readdlm("./Sobol/Run$num_run/num_sample")[1])+1
+else
+    num_sample = 1
+end
+
+# Import partition size and number of iterations (per job)
+partition_size = convert(Int64,readdlm("partition_size")[1])
+iterations_per_job = convert(Int64,readdlm("iterations_per_job")[1])
+
+# Import job number
+pwd_str = pwd()
+str = pwd_str[end-4]
+job = parse("$str")
+
+# Solve partition_size number of samples
+for i = num_sample:num_sample+partition_size-1
+    # Create directory for next sample and save number of samples
+    mkdir("./Sobol/Run$num_run/Sample$num_sample")
+    writedlm("./Sobol/Run$num_run/num_sample",num_sample)
+
+    # Simulate
+    Tsim,X = RunModel(iterations_per_job*job+num_sample)
+
+    # Calculate area under CAT curve and save as objective function
+    obj_fun = CalcArea(Tsim,X[:,92])
+    writedlm("./Sobol/Run$num_run/Sample$num_sample/obj_fun.dat",obj_fun)
+
+    # Save timecourse and species
+    writedlm("./Sobol/Run$num_run/Sample$num_sample/Tsim.dat",Tsim)
+    writedlm("./Sobol/Run$num_run/Sample$num_sample/X.dat",X)
+
+    num_sample += 1
+end

### 2016-06-04 Code/Sobol/0/src/SA.jl

@@ -1,0 +1,11 @@
+# Read SobolAnalysis.txt (it should omit blank lines)
+SobolAnalysis = readdlm("SobolAnalysis.txt")
+
+# Find last row that contains sensitivities (not co-sensitivities)
+last_row = find(SobolAnalysis[:,1].=="Parameter_1")[1]-1
+
+# Remove co-sensitivities, as well as headings in first row
+SA = SobolAnalysis[2:last_row,:]
+
+# Write to disk
+writedlm("SA.txt",SA)

### 2016-06-04 Code/Sobol/0/src/SALib_analyze.sh

@@ -1,0 +1,9 @@
+julia ModelOutput.jl
+
+python -m SALib.analyze.sobol \
+     -p ./param.txt \
+     -Y model_output.txt \
+     -c 0 \
+     > SobolAnalysis.txt
+
+julia SA.jl

### 2016-06-04 Code/Sobol/0/src/SALib_saltelli.sh

@@ -1,0 +1,11 @@
+n=300
+echo $n > n
+
+julia SetBounds.jl
+
+python -m SALib.sample.saltelli \
+     -n $n \
+     -p ./param.txt \
+     -o model_input.txt \
+
+julia SplitModelInput.jl

### 2016-06-04 Code/Sobol/0/src/SetBounds.jl

@@ -1,0 +1,50 @@
+# Do Sobol on initial conditions of enzymes, amino acids, glucose, and oxygen
+
+# Import optimal initial conditions
+init_cond = vec(abs(float(open(readdlm,"initial_condition.dat"))))
+
+#	54	 id:M_glc_D_c	 symbol:M_glc_D_c
+#	93	 id:M_o2_c	 symbol:M_o2_c
+#	80	 id:M_ala_L_c	 symbol:M_ala_L_c
+#	43	 id:M_arg_L_c	 symbol:M_arg_L_c
+#	76	 id:M_asn_L_c	 symbol:M_asn_L_c
+#	73	 id:M_asp_L_c	 symbol:M_asp_L_c
+#	77	 id:M_cys_L_c	 symbol:M_cys_L_c
+#	91	 id:M_gln_L_c	 symbol:M_gln_L_c
+#	90	 id:M_glu_L_c	 symbol:M_glu_L_c
+#	74	 id:M_gly_L_c	 symbol:M_gly_L_c
+#	79	 id:M_his_L_c	 symbol:M_his_L_c
+#	75	 id:M_ile_L_c	 symbol:M_ile_L_c
+#	89	 id:M_leu_L_c	 symbol:M_leu_L_c
+#	78	 id:M_lys_L_c	 symbol:M_lys_L_c
+#	88	 id:M_met_L_c	 symbol:M_met_L_c
+#	81	 id:M_phe_L_c	 symbol:M_phe_L_c
+#	82	 id:M_pro_L_c	 symbol:M_pro_L_c
+#	83	 id:M_ser_L_c	 symbol:M_ser_L_c
+#	84	 id:M_thr_L_c	 symbol:M_thr_L_c
+#	85	 id:M_trp_L_c	 symbol:M_trp_L_c
+#	86	 id:M_tyr_L_c	 symbol:M_tyr_L_c
+#	87	 id:M_val_L_c	 symbol:M_val_L_c
+
+#	104	 id:E_R_PTS	 symbol:E_R_PTS
+#	105	 id:E_R_glk_atp	 symbol:E_R_glk_atp
+#           ...                 ...
+#	265	 id:E_R_ump_ppi	 symbol:E_R_ump_ppi
+#	266	 id:E_R_ump_pi	 symbol:E_R_ump_pi
+
+# Array of initial conditions
+#IC = init_cond[[54;93;80;43;76;73;77;91;90;74;79;75;89;78;88;81:87;104:266]]
[…]

### 2016-06-04 Code/Sobol/0/src/SolveBalances.jl

@@ -1,0 +1,102 @@
+include("MassBalances.jl")
+using Sundials;
+
+# ----------------------------------------------------------------------------------- #
+# Copyright (c) 2015 Varnerlab
+# School of Chemical Engineering Purdue University
+# W. Lafayette IN 46907 USA
+
+# Permission is hereby granted, free of charge, to any person obtaining a copy
+# of this software and associated documentation files (the "Software"), to deal
+# in the Software without restriction, including without limitation the rights 
+# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell 
+# copies of the Software, and to permit persons to whom the Software is 
+# furnished to do so, subject to the following conditions:
+#
+# The above copyright notice and this permission notice shall be included in
+# all copies or substantial portions of the Software.
+#
+# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
+# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
+# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
+# OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
+# THE SOFTWARE.
+# ----------------------------------------------------------------------------------- #
+function SolveBalances(IC_vector)
+# ----------------------------------------------------------------------------------- #
+# SolveBalances.jl was generated using the Kwatee code generation system.
+# SolveBalances: Solves model equations from TSTART to TSTOP given parameters in data_dictionary.
+# Username: nicholas
+# Type: CFPS-JULIA
+# Version: 1.0
+# Generation timestamp: 12-15-2015 14:24:33
+# 
+# Input arguments: 
+# TSTART  - Time start 
+# TSTOP  - Time stop 
+# Ts - Time step 
+# data_dictionary  - Data dictionary instance (holds model parameters) 
+# 
+# Return arguments: 
+# TSIM - Simulation time vector 
[…]

### 2016-06-04 Code/Sobol/0/src/SplitModelInput.jl

@@ -1,0 +1,11 @@
+# ReadSobol samples
+model_input = readdlm("model_input.txt")
+# Create directory, if necessary
+if ~isdir("../../model_input")
+    mkdir("../../model_input")
+end
+# Save each vector separately
+i_range = 1:size(model_input)[1]
+for i in i_range
+    writedlm("../../model_input/$i",model_input[i,:])
+end

### 2016-06-04 Code/Sobol/0/src/WriteModelOutput.jl

@@ -1,0 +1,16 @@
+# Import run number and number of samples
+num_run = convert(Int64,readdlm("./Sobol/num_run")[1])
+num_sample = convert(Int64,readdlm("./Sobol/Run$num_run/num_sample")[1])
+
+# Initialize objective function array
+ObjFun = Float64[]
+
+# For each sample, import objective function value and add to objective function array
+sample_range = 1:num_sample
+for sample in sample_range
+    push!(ObjFun,readdlm("./Sobol/Run$num_run/Sample$sample/obj_fun.dat")[1])
+end
+
+# Save in run folder and in src
+writedlm("./Sobol/Run$num_run/model_output.txt",ObjFun)
+writedlm("model_output.txt",ObjFun)

### 2016-06-04 Code/Sobol/0/src/runSALib.sh

@@ -1,0 +1,24 @@
+n=$(<n)
+partition_size=186
+num_jobs=10
+echo $partition_size > partition_size
+echo $num_jobs > num_jobs
+
+julia CreateSobolDir.jl
+julia CreateRunDir.jl
+
+param=($(wc param.txt))
+num_params=${param[0]}
+
+num_iterations=$((2*n*(num_params+1)))
+iterations_per_job=$((num_iterations/num_jobs))
+echo $iterations_per_job > iterations_per_job
+
+num_partitions=$((iterations_per_job/partition_size))
+
+for j in $(eval echo "{1..$num_partitions}")
+do
+    julia RunSobol.jl
+done
+
+julia WriteModelOutput.jl

### 2016-06-21 Paper_v1.tex
June 20th by CHS.
June 21st by Shih

@@
 
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/AAs_bounded.pdf}
+\caption{Total-order global sensitivities for limiting all controllable initial conditions}
+\label{fig:AAs_bounded}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/AAs_unbounded.pdf}
+\caption{Total-order global sensitivities for each amino acids in the case that each of them are unbounded respectively.}
+\label{fig:AAs_unbounded}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/C_Secretion.pdf}
+\caption{Global sensitivity analysis for alternative carbon sources while glutamic acid as the indication of amino acids.}
+\label{fig:C_Secretion.pdf.pdf}
+\end{figure}
+
 \clearpage
@@
 \end{document}
+\grid

### 2016-06-21 Paper_v1.tex
Jun. 21st afternoon

--- 2016-06-04 Paper_v1.tex
@@
 
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/AAs_bounded.pdf}
+\caption{Total-order global sensitivities of protein export flux to amino acid import fluxes, with flux upper bounds set to 0.5 mM/hr.}
+\label{fig:AAs_bounded}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/AAs_unbounded.pdf}
+\caption{Total-order global sensitivities of protein export flux to amino acid import fluxes, with each amino acid flux unbounded respectively.}
+\label{fig:AAs_unbounded}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/C_Secretion.pdf}
+\caption{Total-order global sensitivities of protein export flux to glutamic acid, oxygen and glucose import fluxes, and translation initiation rate, in the case of four organic acid export flux upper bounds set to 0.1 mM/hr, whereas organic acid export fluxes are unbounded in the control.}
+\label{fig:C_Secretion.pdf.pdf}
+\end{figure}
+
 \clearpage
@@
 \end{document}
+\grid
+\grid

### 2016-06-22 Paper_v1.tex
Jun22

--- 2016-06-04 Paper_v1.tex
@@
 
+\subsection*{Sensitivity analysis on FBA system}
+We conducted global sensitivity on certain fluxes' upper bounds within a constraint-based FBA system, with protein export rate as the objective function.
+
 \clearpage
@@
 %The ensemble of dynamic models predicted that most of the carbon flux going through glycolysis and the TCA cycle,
-%while the constraint based approach predicted most of the carbon flux went through the Entner-Doudoroff pathway and only part of glycolysis.
+%while the constraint-based approach predicted most of the carbon flux went through the Entner-Doudoroff pathway and only part of glycolysis.
 % We present the first dynamic, cell-free model of \textit{E. coli} metabolism and biosynthesis at this scope.
@@
 
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/AAs_bounded.pdf}
+\caption{Total-order global sensitivities of protein export flux to amino acid import flux upper bounds. Each upper bound was varied from 0 to 0.5 mM/hr.}
+\label{fig:AAs_bounded}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/AAs_unbounded.pdf}
+\caption{Total-order global sensitivities of protein export flux to amino acid import flux upper bounds.
+Each upper bound was varied from 0 to 0.5 mM/hr, except the upper bound with respect to which sensitivity was calculated; this was varied from 0 to 10 mM/hr.}
+\label{fig:AAs_unbounded}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/C_Secretion.pdf}
+\caption{Total-order global sensitivities of protein export flux to glutamic acid, oxygen and glucose import flux upper bounds, and translation initiation rate upper bound.
+Lactate, ethanol, acetate, and formate export fluxes were held constant at 0.1 mM/hr in the control case.
+One of these export fluxes was held constant at 10 mM/hr, while the others were held constant at 0.1 mM/hr, in each of the other four cases.}
+\label{fig:C_Secretion.pdf.pdf}
+\end{figure}
+
 \clearpage

### 2016-06-22 Code/FluxSobol/sensitivity/analyze_final.sh
Jun. 21st by Shih

@@ -1,0 +1,34 @@
+#!/bin/bash
+
+# Perform sensitivity analysis from the command line
+# To run: ./shell-example-analyze.sh
+# Note that "python -m" runs the __main__ module in the SALib.analyze package
+# The -B flag prevents the interpreter from compiling bytecode (.pyc files)
+
+python -m SALib.analyze.sobol \
+     -p ./Range.txt \
+     -Y performance.txt \
+     -c 0 \
+     --max-order=2 \
+     > SAout.txt \
+
+# Options:
+# -p, --paramfile: Your parameter range file (3 columns: parameter name, lower bound, upper bound)
+#
+# -Y, --model-output-file: File of model output values to analyze
+#
+# -c, --column (optional): Column of model output file to analyze.
+#                If the file only has one column, this argument will be ignored.
+#
+# --delimiter (optional): Model output file delimiter.
+#
+# --max-order (optional): Maximum order of indices to calculate.
+#               This must match the value chosen during sampling.
+#
+# -r, --resamples (optional): Number of bootstrap resamples used to calculate confidence intervals on indices. Default 1000.
+#
+# --parallel (optional): Flag to enable parallel execution with multiprocessing
+#
+# --processors (optional, int): Number of processors to be used with the parallel option
+#
+# Last line determines the file to output the result by Sobol analysis

### 2016-06-22 Code/FluxSobol/sensitivity/sample.sh

@@ -1,0 +1,30 @@
+#!/bin/bash
+
+# Generate sensitivity analysis samples from the command line
+# To run: ./shell-example-sample.sh
+# Note that "python -m" runs the __main__ module in the SALib.sample package
+# The -B flag prevents the interpreter from compiling bytecode (.pyc files)
+
+python -m SALib.sample.saltelli \
+     -n 5000 \
+	   -p ./Range.txt \
+	   -o ./Samples.txt \
+	   --delimiter=' ' \
+	   --precision=8 \
+
+## Options:
+# -p, --paramfile: Your parameter range file (3 columns: parameter name, lower bound, upper bound)
+#
+# -n, --samples: Sample size.
+#				 Number of model runs is N(2D + 2) if calculating second-order indices (default)
+#        or N(D + 2) otherwise.
+#
+# -o, --output: File to output your samples into.
+#
+# --delimiter (optional): Output file delimiter.
+#
+# --precision (optional): Digits of precision in the output file. Default is 8.
+#
+# --max-order (optional): Maximum order of indices to calculate. Choose 1 or 2, default is 2.
+#								   Choosing 1 will reduce total model runs from N(2D + 2) to N(D + 2)
+#								   Must use the same value (either 1 or 2) for both sampling and analys

### 2016-06-22 Code/FluxSobol/src/SampleBounds.jl

@@ -1,0 +1,99 @@
+# Script to sample the bounds file -
+
+# Script to max product -
+include("DataFile.jl")
+include("Solve.jl")
+include("Types.jl")
+
+# Load the data file with the default bounds -
+data_dictionary = DataFile(0,0,0);
+
+# Set the max -
+data_dictionary["MIN_FLAG"] = false;
+
+# Get the flux model dictionary -
+flux_model_dictionary = data_dictionary["FLUX_MODEL_DICTIONARY"]
+flux_model_dictionary["PROTEIN_export"].flux_obj_coeff = 1.0;
+
+# Reactions that we block ...
+blocked_reaction_set = Set{AbstractString}()
+push!(blocked_reaction_set,"R_PTS")
+push!(blocked_reaction_set,"R_nuo")
+push!(blocked_reaction_set,"R_ndh1")
+push!(blocked_reaction_set,"R_ndh2")
+push!(blocked_reaction_set,"M_lac_D_c_exchange_reverse")
+push!(blocked_reaction_set,"M_etoh_c_exchange_reverse")
+push!(blocked_reaction_set,"M_akg_c_exchange_reverse")
+push!(blocked_reaction_set,"M_fum_c_exchange_reverse")
+push!(blocked_reaction_set,"M_mal_L_c_exchange_reverse")
+push!(blocked_reaction_set,"M_pyr_c_exchange_reverse")
+push!(blocked_reaction_set,"M_ac_c_exchange_reverse")
+push!(blocked_reaction_set,"M_succ_D_c_exchange_reverse")
+data_dictionary["blocked_reaction_set"] = blocked_reaction_set
+
+
+# set the bound -
+# data_dictionary["M_o2_c_exchange_reverse"] = 100
+
+# Set the plasmid concentration -
+#gene_copy_number = (10e-6/1e9)*(6.02e23)*10.0;
+#data_dictionary["deGFP_gene_copies"] = gene_copy_number;
+
+# Load the bounds data -
+bounds_array = readdlm("../sensitivity/Samples.txt")
+(nr,nc) = size(bounds_array)
+
+# Setup performance array -
+performance_array = zeros(nr)
+
+# Setup bounds dictionary -
+bounds_dictionary = Dict()
+
+# Ok, we need to enumerate through this list, and calculate the performance of the model for each value -
+for index in collect(1:nr)
+
+  # Set the bounds array -
+  # The first 20 cols are AAs
[…]

### 2016-06-24 Paper_v1.tex
Jun. 21st by Shih
Jun24

@@
 \vspace{0.05in}
-{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, Joseph Wayman, Kara Calhoun$^{1}$, James Swartz$^{1}$ and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, Che-Hsiao Shih$^{1}$, Joseph Wayman, Kara Calhoun$^{2}$, James Swartz$^{2}$ and Jeffrey D. Varner$^{*}$}}
 \vspace{0.10in}
@@
 {\par \centering {$^{1}$School of Chemical Engineering}}
+{\par \centering {Purdue University, Lafayette, IN 47901}}
+\vspace{0.1in}
+{\par \centering {$^{2}$School of Chemical Engineering}}
 {\par \centering {Stanford University, Stanford, CA 94305}}
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/AAs_bounded.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/AA_bounded_bar.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/AA_bounded_scatter.pdf}
 \caption{Total-order global sensitivities of protein export flux to amino acid import flux upper bounds. Each upper bound was varied from 0 to 0.5 mM/hr.}
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/AAs_unbounded.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/AA_unbounded_bar.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/AA_unbounded_scatter.pdf}
 \caption{Total-order global sensitivities of protein export flux to amino acid import flux upper bounds.

### 2016-11-04 Paper_v1.tex
Nov4

@@
 
+%\begin{table}[ht]
+%\centering
+%\includegraphics[width=0.94\textwidth]{./Figures/AIC.pdf}
+%\caption{}
+%\label{fig:AIC}
+%\end{table}
+
 \begin{figure}[ht]
 \centering
-\includegraphics[width=0.94\textwidth]{./Figures/SensAll.png}
-\caption{Mean and standard error of local sensitivities of rate constants (top), saturation constants (middle), and control parameters (bottom).}
-\label{fig:LocalSensitivity}
+\includegraphics[width=0.94\textwidth]{./Figures/ClusterCAT.pdf}
+\caption{Normalized first-order and pairwise sensitivities of CAT production to rate constants.
+Dark pixels indicate the most sensitive indices.
+Dendrogram indicates high responders (dark gray background), medium responders (medium gray background), and low responders (light gray background).}
+\label{fig:ClusterCAT}
 \end{figure}
@@
 \centering
-%\includegraphics[width=1.00\textwidth]{./Figures/SensitivitiesHalfTall.pdf}
-\includegraphics[width=1.00\textwidth]{./Figures/Sensitivity.pdf}
-\includegraphics[width=1.00\textwidth]{./Figures/Metab_sensitivity.pdf}
-\caption{Total-order global sensitivities for experimentally controllable initial conditions, including glucose, oxygen, amino acids, and enzymes.}
-\label{fig:GlobalSensitivity}
+\includegraphics[width=1.00\textwidth]{./Figures/ClusterSystem.pdf}
+\caption{Normalized first-order and pairwise sensitivities of system state to rate constants.
+Dark pixels indicate the most sensitive indices.
+Dendrogram indicates high responders (dark gray background), medium responders (medium gray background), and low responders (light gray background).}
+\label{fig:ClusterSystem}
 \end{figure}
@@
 One of these export fluxes was held constant at 10 mM/hr, while the others were held constant at 0.1 mM/hr, in each of the other four cases.}
-\label{fig:C_Secretion.pdf.pdf}
+\label{fig:C_Secretion.pdf}
 \end{figure}
+
+%\begin{figure}[ht]
+%\centering
+%\includegraphics[width=1.00\textwidth]{./Figures/MikeYield.pdf}
+%\caption{}
+%\label{fig:MikeYield.pdf}
+%\end{figure}
 

### 2016-11-06 Paper_v1.tex
Nov6

@@
 \usepackage{graphicx}
+\usepackage{cite}
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{mathtools}
 \usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
@@
 We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
-Arginine and glutamate were excluded due to not being present in the training dataset.
+%Arginine and glutamate were excluded due to not being present in the training dataset.
 The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
-We then used sequence-specific FBA to calculate a theoretical maximum CAT yield of 0.1942.
-Thus, we showed that our experimental dataset and best-fit parameter set were each producing CAT at 45\% of the theoretical maximum.
+While the model ensemble described the experimental data, it was unclear whether the performance of the CFPS system was optimal.
+To address this question, we used ssFBA in combination with the cell-free metabolic network and a detailed promoter model under a T7 polymerase to compute the maximum theoretical carbon yield.
+However, we first validated the ssFBA approach by comparing an ensemble of simulated versus measured concentrations of CAT over a one hour period (Fig. ~\ref{fig:CATProdYield}A). 
+The ensemble of 100 sets captured the CAT concentration profile which was randomly generated by sampling RNA polymerase levels, ribosome levels and elongation rates in a physiological range.
+We then used sequence-specific FBA to calculate a theoretical maximum CAT yield under four different cases: unconstrained, limited oxidative phosphorylation, bounded by transcription/translation rates, and bounded by experimental data (Fig.
+~\ref{fig:CATProdYield}B).
+The theoretical maximum carbon yield of CAT was 0.349 for an unconstrained case and 0.194 for the transcription and translation constrained case.
+Thus, we showed that our experimental dataset and best-fit parameter set were each producing CAT at 25\% of the theoretical maximum and 45\% of a theoretical physiological case.
+Whereas, the case constrained by experimental data had a carbon yield of 0.062 $\pm$ 0.02, similar to the experimental yield. 
 This allowed us to quantify the amount of carbon being diverted to byproducts, and suggests that there is potential for a doubling of CAT production by reducing this diversion of carbon.
+In comparing the flux distributions between the unconstrained and constrained cases (Fig.
+~\ref{fig:Network}), both constrained cases heavily utilize the Entner–Doudoroff pathway which may be a first viable knockout to increase CAT yield.
 
@@
 
-\subsection*{Calculation of CAT yield}
-The yield on CAT production was calculated for three cases: the experimental data, the best-fit parameter set, and a theoretical maximum yield.
-In each case the yield was formulated as a ratio of carbon produced as CAT to carbon consumed as reactants (glucose and amino acids):
+\subsection*{Sequence specific FBA and calculation of CAT yield}
+The yield on CAT production was calculated for each case as a ratio of carbon produced as CAT to carbon consumed as reactants (glucose and amino acids):
 \begin{equation}\label{eqn:yield-definition}
@@
 In the experimental case and the best-fit set case, yield was calculated by setting $\Delta CAT$ equal to the final minus the initial CAT concentration and setting $\Delta m_{i}$ equal to the initial minus the final reactant concentration.
-The theoretical yield was calculated using flux balance anaylsis (FBA) with a sequence-based analysis on CAT.
+The theoretical yield was calculated using flux balance anaylsis (FBA) with a sequence-specific based analysis on CAT.
 The sequence specific FBA \cite{2002_allen_palsson} problem was formulated as:
-\begin{equation}\label{eqn:FBA}
-\begin{split}
-	\max_{\boldsymbol{v}}{} \! \left( v_{obj}=\mathbf{\boldsymbol{c}}^T \boldsymbol{v} \right) \qquad \alpha_i \leq v_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R} \\
-	\mathrm{Subject \; to:}	\; \; \mathbf{S}\mathbf{v}=\mathbf{0}\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad
-\end{split}
+\begin{equation}\nonumber
+ \begin{multlined}
+	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{obj} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
+	\mathrm{Subject \; to:}
+	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
+\alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
+ \end{multlined}
 \end{equation}
-where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{v}$ denotes the unknown flux vector, $\boldsymbol{c}$ denotes the objective selection vector, and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $v_{i}$, respectively.
-The stoichiometric matrix was expanded to include the transcription and translation reactions for producing CAT.
-The objective $v_{obj}$ was to maximize the specific rate of CAT formation.
-The specific glucose uptake rate was constrained to allow a maximum flux of 10 mM/hr \cite{2002_Mahadevan_BiophysJ}; the amino acids and oxygen uptake rates were also bound to allow a maximum flux of 10 mM/hr, but did not reach this maximum flux.
-Glucose, oxygen, and amino acids were modeled as being imported into the system, whereas CAT synthesis was modeled through protein export.
-The rest of the network followed a psuedo steady-state asusmption where all other metabolites were not allowed to accumulate; thus, the network could be solved by linear programming.
+where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
+and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
+The objective $w_{obj}$ was to maximize the specific rate of CAT formation.
+The specific glucose uptake rate was constrained to allow a maximum flux of 40 mM/hr according to literature data; the specific amino acid uptake rates were also bound to allow a maximum flux of 30 mM/hr, but did not reach this maximum flux.
+The transcription and translation template reactions were added to the metabolic network and are based off sequence specific analysis \cite{2002_allen_palsson} involving transcription initiaion, transcription, mRNA degradation, translation initiation, translation, and tRNA charging.
+The mRNA and protein sequence of each protein was determined from literature. 
+The transcription rate was constrained as:
+\begin{equation}\nonumber
+	w_{tx} = RNAP\frac{v_{RNAP}}{l_{mRNA}}\left(\frac{Gene}{km+Gene}\right)P
+\end{equation}
+where $RNAP$ is the concentration of RNA polymerase, $v_{RNAP}$ is the elongation rate (nucleotides/hr) by the RNA polymerase, $l_{mRNA}$ is the number of nucleotides in the mRNA, $Gene$ is the gene concentration, $km$ is the plasmid saturation coefficient, and $P$ is the promoter activity.
+
+The promoter activity was formulated following Moon et al. for synthetic circuits as:
+\begin{equation}\nonumber
+	P = \frac{K_{1}}{1 + K_{1}}
+\end{equation}
+where $K_{1}$ represents the state of T7 RNA polymerase binding.
+
+The translation rate was constrained as:
+ \begin{equation}\nonumber
+	w_{tl} = K_{P}Ribo\frac{v_{Ribo}}{l_{protein}}[mRNA_{ss}]
+\end{equation}
+where $K_{P}$ is the polysome amplification constant, $Ribo$ is the ribosome concentration,  $v_{Ribo}$ is the elongation rate (amino acids/hr) of the ribosome, $l_{protein}$ is the number of amino acids in the protein of interest, and $mRNA_{ss}$ is the mRNA concentration at steady state determined by the transcription rate divided by the degradation rate of mRNA.
+An ensemble of flux distributions was calculated for 100 sets by randonmly sampling. 
+Glucose, oxygen, and amino acids were modeled as being imported into the system, whereas CAT synthesis and metabolite byproduct formation was modeled as an export from the system.
+The rest of the network followed a pseudo steady-state asusmption where all other metabolites were not allowed to accumulate; thus, the network could be solved by linear programming.
 The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
-The solution flux vector was used to calculate the theoretical carbon yield of CAT, reformulated in terms of flux:
-\begin{equation}\label{eqn:yield}
-	Yield=\frac{v_{CAT}\;C_{CAT}}{\sum_{i=1}^{\mathcal{R}}\max(v_{i},0)\;C_{i}}
-\end{equation}
-where $v_{CAT}$ denotes the CAT export flux and $v_{i}$ denotes the import flux of the $i$th substrate.
-
+The solution flux vector was used to calculate the theoretical carbon yield of CAT for four different cases.
+%\begin{equation}\label{eqn:yield}
+%	Yield=\frac{w_{CAT}\;C_{CAT}}{\sum_{i=1}^{\mathcal{R}}\max(w_{i},0)\;C_{i}}
+%\end{equation}
+%where $w_{CAT}$ denotes the CAT export flux and $w_{i}$ denotes the import flux of the $i$th substrate.
+For the unconstrained case, all rates were left unbounded. 
+An ensemble of flux distributions was calculated by randonmly sampling the maximum specific glucose uptake rate from 30 to 40 mM/hr determined from experimental data.
+For the limited oxidative phosphorylation case, an ensemble of flux distributions was generated by randomly sampling the maximum specific oxygen uptake rate to from 0 to 10 mM/hr.
+For the TXTL case (constrained by transcription and translation rates), an ensemble was generated by randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature as well as the specific glucose uptake rate from 30 to 40 mM/hr.
+For the case where the flux was constrained by experimental data, the lower and upper bounds where data was avaible were randomly sampled in the physiological range of the data in addition to randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
 \clearpage
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./Figures/Network.pdf}
-\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and NADPH/NADH transfer.
-FBA flux value (top), and mean ± standard error across ensemble at 1.5 hrs (middle) and 3 hrs (bottom), normalized to CAT synthesis flux.}
+\includegraphics[width=0.75\textwidth]{./Figures/CAT_prod_yield.pdf}
+\caption{Sequence specific flux balance analysis of CAT production and yield.
+A.
+95\% confidence interval of the ensemble (light blue region) for CAT concentration versus time.
+B.
+Theoretical maximum carbon yield of CAT calcualted by ssFBA for four different cases: unconstrained (black), limited oxidative phosphorylation activity (gray), constrained by transcription and translation rates (light grey), and constrained by experimental data (white).
+Error bars represent standard deviation of the ensemble.
+}
+\label{fig:CATProdYield}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.8\textwidth]{./Figures/CAT_flux_final.pdf}
+\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, and oxidative phosphorylation.
+Sequence specific FBA flux value (mean ± standard deviation) across ensemble for 1 hr, normalized to glucose uptake flux.
+Unconstrained (top row), constrained by transcription and translation rates (second row), and constrained by experimental data (bottom row) shows the flux distribution throughout central carbon metabolism.}
 \label{fig:Network}
@@
 Best-fit parameter set (orange line) versus experimental data (points).
-95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 18,000 sets.}
+95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
 \label{fig:BothCarbon}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+%\includegraphics[width=1.00\textwidth]{./Figures/BothCarbon.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/Carbon.png}
+\includegraphics[width=0.05\textwidth]{./Figures/null.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/CarbonNoControl.png}
+\caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species.
+Best-fit parameter set (orange line) versus experimental data (points).
+95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
+\label{fig:BothCarbon_png}
 \end{figure}
@@
 Best-fit parameter set (orange line) versus experimental data (points).
-95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 18,000 sets.}
+95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
 \label{fig:Energy}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/Energy.png}
+\caption{Energy species and energy totals by base in the presence of allosteric control.
+Best-fit parameter set (orange line) versus experimental data (points).
+95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
+\label{fig:Energy_png}
 \end{figure}
@@
 Best-fit parameter set (orange line) versus experimental data (points).
-95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 18,000 sets.}
+95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
 \label{fig:Amino}
@@
 \centering
-\includegraphics[width=0.94\textwidth]{./Figures/ClusterCAT.pdf}
+\includegraphics[width=0.94\textwidth]{./Figures/ClusterCAT161.pdf}
 \caption{Normalized first-order and pairwise sensitivities of CAT production to rate constants.
@@
 \label{fig:ClusterCAT}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.94\textwidth]{./Figures/ClusterCAT161.png}
+\caption{Normalized first-order and pairwise sensitivities of CAT production to rate constants.
+Dark pixels indicate the most sensitive indices.
+Dendrogram indicates high responders (dark gray background), medium responders (medium gray background), and low responders (light gray background).}
+\label{fig:ClusterCAT_png}
 \end{figure}
@@
 
-%\begin{figure}[ht]
-%\centering
-%\includegraphics[width=1.00\textwidth]{./Figures/MikeYield.pdf}
-%\caption{}
-%\label{fig:MikeYield.pdf}
-%\end{figure}
-
 \clearpage

### 2016-11-06 Paper_v1.tex
Nov6_2

@@
 \subsection*{Generation of model ensemble}
-We generated an ensemble of 18,000 parameter sets via a downhill-only random walk Monte Carlo method.
-Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-absolute-error between experimental data and model predictions:
+We generated an ensemble of 100 diverse parameter sets via a Markov chain Monte Carlo random walk.
+Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-squared-error between experimental data and model predictions:
 \begin{equation}\label{eqn:cost-function}
-    cost=\sum_{i=1}^{\mathcal{D}}\left(w_i\sum_{j=1}^{\mathcal{T}}abs\left(x_{ij}^{data}-x_{i}^{sim}|_{t(j)}\right)\right)
+    cost=\sum_{i=1}^{\mathcal{D}}\left(w_i\sum_{j=1}^{\mathcal{T}_i}abs\left(x_{ij}^{data}-x_{i}^{sim}|_{t(j)}\right)\right)
 \end{equation}
-where $\mathcal{D}$ denotes the number of datasets, $w_i$ denotes a weight, equal to 5 for the glucose, CAT, pyruvate, lactate, acetate, succinate, and malate datasets, and 1 elsewhere, $\mathcal{T}$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ dneotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
+where $\mathcal{D}$ denotes the number of datasets, $w_i$ denotes the weight of the $i$th dataset, 
+%equal to 5 for the glucose, CAT, pyruvate, lactate, acetate, succinate, and malate datasets, and 1 elsewhere,
+$\mathcal{T}_i$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ dneotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
 We then perturbed model parameters:
@@
 \end{equation}
-where $\mathcal{P}$ denotes the number of parameters, equal to 652, which includes 163 rate constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i$th parameter, $k_i$ denotes the current value of the $i$th parameter, $a$ denotes a distribution variance, set to 0.03, and $r$ denotes a random sample from the normal distribution.
-We stored the parameter set and calculated its cost; if it was less than the previous cost, we used the new parameter set to generate the following set.
-After generating 180,000 sets we defined the 18,000 sets with the lowest cost values as our ensemble, and the set with the lowest cost value as our best-fit set.
+where $\mathcal{P}$ denotes the number of parameters, equal to 815, which includes 163 rate constants, 163 enzyme degradation rate constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i$th parameter, $k_i$ denotes the current value of the $i$th parameter, $a$ denotes a distribution variance,
+%set to 0.03, 
+and $r_i$ denotes a random sample from the normal distribution.
+For each newly generated parameter set, we re-solved the balance equations and calculated the cost function.
+All sets with a lower cost than the previous set, and some with higher cost, were added to the ensemble.
+After generating 12,437 sets, we selected 100 sets with minimal correlation to each other so as to avoid over-sampling any region of parameter space.
+The original 12,437-set ensemble had a [mean,median,maximum] Pearson correlation coefficient [REFERENCE NEEDED?] of [?] between pairs of sets; the 100-set ensemble had a [mean,median,maximum] Pearson correlation coefficient of [?] between pairs of sets.
 

### 2016-11-06 Paper_v1.tex
Nov6_3

@@
 
-\subsection*{Global and local sensitivity analysis}
-We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which of the experimentally controllable parameters affected the performance of the reduced order model \citep{SOBOL_METHOD}.
-This included the initial conditions of glucose, oxygen, amino acids, and enzymes.
-We computed the total sensitivity index of each parameter relative to a performance objective of area under the CAT curve (CAT production).
-We established the sampling bounds for each parameter from the value of that parameter in the set used to generate the ensemble.
-We used the sampling method of Saltelli \textit{et al.} \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ sets which obeyed our parameter ranges,
-where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, $N$ = 300 and $d$ = 185, so the total sensitivity indices were computed from
-111,600 model evaluations. The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \citep{SALIB}.
-We conducted a local sensitivity analysis to estimate which of the other model parameters affected performance.
-This included the same parameters that were varied in the ensemble: rate constants, saturation constants, and control parameters.
-The local sensitivity for each parameter was calculated across a sub-ensemble of 180 parameter sets, randomly chosen from the ensemble of 18,000 sets:
+\subsection*{Local sensitivity analysis}
+We determined the reactions most important to protein production by computing the local sensitivity of CAT concentration to each rate constant in the network.
+Each sensitivity index was formulated as:
 \begin{equation}\label{eqn:local-sensitivity}
 \begin{split}
-    S_{ij}=\frac{p_{ij}}{AUC(p_{ij})}\frac{AUC(p_{ij}+\Delta p_{ij})-AUC(p_{ij})}{\Delta p_{ij}}\qquad{i=1,2,\hdots,\mathcal{E}}\qquad{j=1,2,\hdots,\mathcal{P}} \\
-    \Delta p_{ij}=0.001\;p_{ij}\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad
+    S_{ij}=norm(CAT(p_i,p_j,t)-CAT(\alpha*p_i,\alpha*p_j,t))\qquad{i,j=1,2,\hdots\mathcal{P}}\\
 \end{split}
 \end{equation}
-where $\mathcal{E}$ denotes the number of parameter sets in the sub-ensemble, equal to 180, $\mathcal{P}$ denotes the number of parameters, equal to 652, $S_{ij}$ denotes the sensitivity of the $j$th parameter for the $i$th parameter set, $p_{ij}$ denotes the value of the $j$th parameter for the $i$th parameter set, $\Delta p_{ij}$ denotes the perturbation of the $j$th parameter for the $i$th parameter set, equal to 0.1\% of the parameter value, and $AUC()$ denotes the area under the CAT curve.
-We then calculated the mean and standard error of each local sensitivity across the sub-ensemble of 180 sets.
+where $S_{ij}$ denotes the sensitivity of CAT production to the $i$th and $j$th parameters, $CAT(p_i,p_j,t)$ denotes CAT concentration as a function of time and the $i$th and $j$th parameters, $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
+In calculating the pairwise sensitivities, each parameter was perturbed by 1\%;
 

### 2016-11-10 Paper_v1.tex
Nov7
Nov9

@@
 
-The ensemble of models captured the time evolution of cell free CAT biosynthesis (Fig.~\ref{fig:BothCarbon} - \ref{fig:Amino}).
+The ensemble of models captured the time evolution of cell free CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
 Glucose was exhausted with 3 hr [FILL ME IN].
-The ensemble also captured the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:BothCarbon}, top) and the totals by base .
+The ensemble also captured the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:CarbonBoth}, top) and the totals by base .
 The ensemble and the best-fit set also predict some of the amino acid measurements, while failing to predict others (Fig.~\ref{fig:Amino}).
 the central carbon metabolism, including glucose uptake, CAT production, and the dynamics of the organic acid intermediates .
-Allosteric control is important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:BothCarbon}, bottom).
+Allosteric control is important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).
 This is likely due to a structural deficiency in the model; in some cases, the consumption of an amino acid through CAT synthesis is not enough to explain the decrease shown in the data, and there are no other reactions that consume it.
 Thus, a more comprehensive biological description is needed to fully explain amino acid dynamics.
+
+\subsection*{Sensitivity analysis}
+We performed a local sensitivity analysis to determine the network reactions with the greatest effect on protein production and overall system state.
+CAT production was msot sensitive to the CAT synthesis reaction, oxidative phosphorylation activity, and alanine synthesis, as well as various reactions in glycolysis, the TCA cycle, and amino acid synthesis and degradation.
+
+%We conducted a local sensitivity analysis to determine which of the kinetic and control parameters affected model performance.
+%We calculated performance as area under the CAT curve, which was directly related to CAT synthesis rate, as the culture time was fixed and no CAT degradation was modeled.
+%We randomly chose 180 sets of the 18,000 sets in the ensemble and defined these as our sub-ensemble; for each set in the sub-ensemble, we varied the rate constant and saturation constants of each metabolic flux and measured the resulting change in CAT production to estimate the sensitivity of model performance to that parameter.
+%We did the same for the control parameters, both the order (Hill coefficient) and gain (related to the the dissociation constant).
+%This allowed us to estimate the relative importance of the kinetic and control parameters to CAT production across the ensemble.
+%Of the rate constants, those with the highest positive sensitivities were CAT synthesis, GTP synthesis, GMP synthesis, glutamine synthesis, and aspartate synthesis (Fig.~\ref{fig:LocalSensitivity}, top).
+%This is explained by GTP and the amino acids being reactants for CAT synthesis.
+%Also, GMP synthesis increases the total amount of guanosine, allowing for more GTP production.
+%The rate constants with the largest negative sensitivities were GTP degradation, arginine synthesis, and UMP synthesis.
+%While GTP degradation is obvious, the others can be explained in that they consume ATP as well as several amino acids, all of which are reactants for CAT synthesis.
+%Of the saturation constants, the reverse is seen: the largest positive sensitivities are those associated with arginine synthesis, while the largest negative sensitivities are those associated with CAT synthesis, GTP synthesis, and GMP synthesis (Fig.~\ref{fig:LocalSensitivity}, middle).
+%This is because an increase in saturation constant causes a decrease in the corresponding reaction rate.
+%The control parameters were seen to be the least significant and the most uncertain (Fig.~\ref{fig:LocalSensitivity}, bottom).
+%Only two had a small standard error across the ensemble, relative to the ensemble mean sensitivity: the gain and order for pyruvate acting as an inhibitor on the pdh reaction.
+%This could be because pdh consumes pyruvate and diverts carbon away from the pathways that ultimately contribute to CAT production.
+%Taken together with the lack of change in glucose uptake and CAT production when control is removed (Fig.~\ref{fig:CarbonBoth}), this suggests that allosteric control is not the limiting factor to CAT production.
+
+%We conducted a global sensitivity analysis on the parameters that could be controlled experimentally: the initial conditions of glucose, oxygen, amino acids, and enzymes.
+%We used the variance-based method of Sobol, and the same objective function of area under the CAT curve.
+%Using a parameter set of relatively good fit against data, we defined parameter bounds and generated a Sobol sequence of 111,600 parameter values that fit within those bounds.
+%We then calculated the total-order sensitivity and confidence interval for each of the experimentally controllable initial conditions.
+%As the sensitivities were total-order, they were guaranteed to be non-negative.
+%The largest sensitivities belonged to the initial conditions of the CAT macromolecular synthesis machinery, GTP synthase, GTP degradation, some amino acids such as phenylalanine, proline, and leucine, and some amino acid synthases (Fig.~\ref{fig:GlobalSensitivity}).
+%This is all explained by GTP and amino acids being reactants for CAT synthesis.
+%While some of the amino acids and amino acid synthases were among the highest in sensitivity, theirs were also very uncertain relative to the CAT macromolecular synthesis machinery and GTP synthase.
+%The initial conditions of glucose and oxygen were among the least important according to the global sensitivity analysis, suggesting that the model predicts that CAT production can be sustained by consuming initial stores or can be powered by other means.
 
@@
 ~\ref{fig:Network}), both constrained cases heavily utilize the Entner–Doudoroff pathway which may be a first viable knockout to increase CAT yield.
-
-\subsection*{Sensitivity analysis}
-We conducted a local sensitivity analysis to determine which of the kinetic and control parameters affected model performance.
-We calculated performance as area under the CAT curve, which was directly related to CAT synthesis rate, as the culture time was fixed and no CAT degradation was modeled.
-We randomly chose 180 sets of the 18,000 sets in the ensemble and defined these as our sub-ensemble; for each set in the sub-ensemble, we varied the rate constant and saturation constants of each metabolic flux and measured the resulting change in CAT production to estimate the sensitivity of model performance to that parameter.
-We did the same for the control parameters, both the order (Hill coefficient) and gain (related to the the dissociation constant).
-This allowed us to estimate the relative importance of the kinetic and control parameters to CAT production across the ensemble.
-Of the rate constants, those with the highest positive sensitivities were CAT synthesis, GTP synthesis, GMP synthesis, glutamine synthesis, and aspartate synthesis (Fig.~\ref{fig:LocalSensitivity}, top).
-This is explained by GTP and the amino acids being reactants for CAT synthesis.
-Also, GMP synthesis increases the total amount of guanosine, allowing for more GTP production.
-The rate constants with the largest negative sensitivities were GTP degradation, arginine synthesis, and UMP synthesis.
-While GTP degradation is obvious, the others can be explained in that they consume ATP as well as several amino acids, all of which are reactants for CAT synthesis.
-Of the saturation constants, the reverse is seen: the largest positive sensitivities are those associated with arginine synthesis, while the largest negative sensitivities are those associated with CAT synthesis, GTP synthesis, and GMP synthesis (Fig.~\ref{fig:LocalSensitivity}, middle).
-This is because an increase in saturation constant causes a decrease in the corresponding reaction rate.
-The control parameters were seen to be the least significant and the most uncertain (Fig.~\ref{fig:LocalSensitivity}, bottom).
-Only two had a small standard error across the ensemble, relative to the ensemble mean sensitivity: the gain and order for pyruvate acting as an inhibitor on the pdh reaction.
-This could be because pdh consumes pyruvate and diverts carbon away from the pathways that ultimately contribute to CAT production.
-Taken together with the lack of change in glucose uptake and CAT production when control is removed (Fig.~\ref{fig:BothCarbon}), this suggests that allosteric control is not the limiting factor to CAT production.
-
-We conducted a global sensitivity analysis on the parameters that could be controlled experimentally: the initial conditions of glucose, oxygen, amino acids, and enzymes.
-We used the variance-based method of Sobol, and the same objective function of area under the CAT curve.
-Using a parameter set of relatively good fit against data, we defined parameter bounds and generated a Sobol sequence of 111,600 parameter values that fit within those bounds.
-We then calculated the total-order sensitivity and confidence interval for each of the experimentally controllable initial conditions.
-As the sensitivities were total-order, they were guaranteed to be non-negative.
-The largest sensitivities belonged to the initial conditions of the CAT macromolecular synthesis machinery, GTP synthase, GTP degradation, some amino acids such as phenylalanine, proline, and leucine, and some amino acid synthases (Fig.~\ref{fig:GlobalSensitivity}).
-This is all explained by GTP and amino acids being reactants for CAT synthesis.
-While some of the amino acids and amino acid synthases were among the highest in sensitivity, theirs were also very uncertain relative to the CAT macromolecular synthesis machinery and GTP synthase.
-The initial conditions of glucose and oxygen were among the least important according to the global sensitivity analysis, suggesting that the model predicts that CAT production can be sustained by consuming initial stores or can be powered by other means.
 
@@
 
-\subsection*{Local sensitivity analysis}
+\subsection*{Sensitivity analysis}
 We determined the reactions most important to protein production by computing the local sensitivity of CAT concentration to each rate constant in the network.
@@
 where $S_{ij}$ denotes the sensitivity of CAT production to the $i$th and $j$th parameters, $CAT(p_i,p_j,t)$ denotes CAT concentration as a function of time and the $i$th and $j$th parameters, $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
-In calculating the pairwise sensitivities, each parameter was perturbed by 1\%;
+In calculating the pairwise sensitivities, each parameter was perturbed by 1\%; first-order sensitivities ($i$ = $j$) were subject to two 1\% perturbations, equivalent to a perturbation of 2.01\%.
 
@@
 \section*{Funding}
-This study was supported by a National Science Foundation Graduate Research Fellowship (DGE-1333468) to N.H and by an award from
-the US Army and Systems Biology of Trauma Induced Coagulopathy (W911NF-10-1-0376) to J.V for the support of M.V.
+This study was supported by a National Science Foundation Graduate Research Fellowship (DGE-1333468) to N.H and by an award from the US Army and Systems Biology of Trauma Induced Coagulopathy (W911NF-10-1-0376) to J.V for the support of M.V.
 
@@
 
-% Figures and captions go here ...
 \begin{figure}[ht]
 \centering
-\includegraphics[width=0.75\textwidth]{./Figures/CAT_prod_yield.pdf}
-\caption{Sequence specific flux balance analysis of CAT production and yield.
-A.
-95\% confidence interval of the ensemble (light blue region) for CAT concentration versus time.
-B.
-Theoretical maximum carbon yield of CAT calcualted by ssFBA for four different cases: unconstrained (black), limited oxidative phosphorylation activity (gray), constrained by transcription and translation rates (light grey), and constrained by experimental data (white).
-Error bars represent standard deviation of the ensemble.
-}
-\label{fig:CATProdYield}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=0.8\textwidth]{./Figures/CAT_flux_final.pdf}
-\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, and oxidative phosphorylation.
-Sequence specific FBA flux value (mean ± standard deviation) across ensemble for 1 hr, normalized to glucose uptake flux.
-Unconstrained (top row), constrained by transcription and translation rates (second row), and constrained by experimental data (bottom row) shows the flux distribution throughout central carbon metabolism.}
-\label{fig:Network}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-%\includegraphics[width=1.00\textwidth]{./Figures/BothCarbon.pdf}
-\includegraphics[width=1.00\textwidth]{./Figures/Carbon.pdf}
-\includegraphics[width=0.05\textwidth]{./Figures/null.pdf}
-\includegraphics[width=1.00\textwidth]{./Figures/CarbonNoControl.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/CarbonBoth.pdf}
 \caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species.
@@
 95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
-\label{fig:BothCarbon}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-%\includegraphics[width=1.00\textwidth]{./Figures/BothCarbon.pdf}
-\includegraphics[width=1.00\textwidth]{./Figures/Carbon.png}
-\includegraphics[width=0.05\textwidth]{./Figures/null.pdf}
-\includegraphics[width=1.00\textwidth]{./Figures/CarbonNoControl.png}
-\caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species.
-Best-fit parameter set (orange line) versus experimental data (points).
-95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
-\label{fig:BothCarbon_png}
+\label{fig:CarbonBoth}
 \end{figure}
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/Energy.png}
-\caption{Energy species and energy totals by base in the presence of allosteric control.
-Best-fit parameter set (orange line) versus experimental data (points).
-95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
-\label{fig:Energy_png}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
 \includegraphics[width=1.00\textwidth]{./Figures/Amino.pdf}
@@
 
-%\begin{table}[ht]
-%\centering
-%\includegraphics[width=0.94\textwidth]{./Figures/AIC.pdf}
-%\caption{}
-%\label{fig:AIC}
-%\end{table}
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/BoxPlot-KCFM.pdf}
+\caption{Log of cost function across 37 datasets for data-trained ensemble (blue) and randomly generated ensemble (red, gray background).
+Median (bars), interquartile range (boxes), range excluding outliers (dashed lines), and outliers (circles) for each dataset.
+Median across all datasets (large bar overlaid).}
+\label{fig:BoxPlot}
+\end{figure}
 
@@
 \centering
-\includegraphics[width=0.94\textwidth]{./Figures/ClusterCAT161.pdf}
-\caption{Normalized first-order and pairwise sensitivities of CAT production to rate constants.
-Dark pixels indicate the most sensitive indices.
-Dendrogram indicates high responders (dark gray background), medium responders (medium gray background), and low responders (light gray background).}
+\includegraphics[width=1.00\textwidth,trim=0 0 263 0,clip]{./Figures/ClusterCAT.pdf}
+\caption{Normalized first-order and pairwise sensitivities of CAT production to rate constants.}
 \label{fig:ClusterCAT}
@@
 \centering
-\includegraphics[width=0.94\textwidth]{./Figures/ClusterCAT161.png}
-\caption{Normalized first-order and pairwise sensitivities of CAT production to rate constants.
-Dark pixels indicate the most sensitive indices.
-Dendrogram indicates high responders (dark gray background), medium responders (medium gray background), and low responders (light gray background).}
-\label{fig:ClusterCAT_png}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=1.00\textwidth]{./Figures/ClusterSystem.pdf}
-\caption{Normalized first-order and pairwise sensitivities of system state to rate constants.
-Dark pixels indicate the most sensitive indices.
-Dendrogram indicates high responders (dark gray background), medium responders (medium gray background), and low responders (light gray background).}
+\includegraphics[width=1.15\textwidth,trim=0 0 263 0,clip]{./Figures/ClusterSystem.pdf}
+\caption{Normalized first-order and pairwise sensitivities of system state to rate constants. A (most sensitive parameters): hexokinase run by GTP, and lactate dehydrogenase forward reaction. B: }
 \label{fig:ClusterSystem}
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/AA_bounded_bar.pdf}
-\includegraphics[width=1.00\textwidth]{./Figures/AA_bounded_scatter.pdf}
-\caption{Total-order global sensitivities of protein export flux to amino acid import flux upper bounds. Each upper bound was varied from 0 to 0.5 mM/hr.}
-\label{fig:AAs_bounded}
+\includegraphics[width=0.75\textwidth]{./Figures/CAT_prod_yield.pdf}
+\caption{Sequence specific flux balance analysis of CAT production and yield.
+A.
+95\% confidence interval of the ensemble (light blue region) for CAT concentration versus time.
+B.
+Theoretical maximum carbon yield of CAT calcualted by ssFBA for four different cases: unconstrained (black), limited oxidative phosphorylation activity (gray), constrained by transcription and translation rates (light grey), and constrained by experimental data (white).
+Error bars represent standard deviation of the ensemble.}
+\label{fig:CATProdYield}
 \end{figure}
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/AA_unbounded_bar.pdf}
-\includegraphics[width=1.00\textwidth]{./Figures/AA_unbounded_scatter.pdf}
-\caption{Total-order global sensitivities of protein export flux to amino acid import flux upper bounds.
-Each upper bound was varied from 0 to 0.5 mM/hr, except the upper bound with respect to which sensitivity was calculated; this was varied from 0 to 10 mM/hr.}
-\label{fig:AAs_unbounded}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=1.00\textwidth]{./Figures/C_Secretion.pdf}
-\caption{Total-order global sensitivities of protein export flux to glutamic acid, oxygen and glucose import flux upper bounds, and translation initiation rate upper bound.
-Lactate, ethanol, acetate, and formate export fluxes were held constant at 0.1 mM/hr in the control case.
-One of these export fluxes was held constant at 10 mM/hr, while the others were held constant at 0.1 mM/hr, in each of the other four cases.}
-\label{fig:C_Secretion.pdf}
+\includegraphics[width=0.8\textwidth]{./Figures/CAT_flux_final.pdf}
+\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, and oxidative phosphorylation.
+Sequence specific FBA flux value (mean ± standard deviation) across ensemble for 1 hr, normalized to glucose uptake flux.
+Unconstrained (top row), constrained by transcription and translation rates (second row), and constrained by experimental data (bottom row) shows the flux distribution throughout central carbon metabolism.}
+\label{fig:Network}
 \end{figure}

### 2016-11-11 Paper_v1.tex
MVupdates2

@@
 \usepackage{amsmath}
+\usepackage{textgreek}
 %\usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e}
@@
 \begin{titlepage}
-{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell Free Protein Synthesis in \emph{Escherichia~coli}}}}
+{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell-Free Protein Synthesis in \emph{Escherichia~coli}}}}
 \vspace{0.05in}
-{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, Che-Hsiao Shih$^{1}$, Joseph Wayman, Kara Calhoun$^{2}$, James Swartz$^{2}$ and Jeffrey D. Varner$^{*}$}}
+{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, Joseph Wayman, Kara Calhoun$^{1}$, James Swartz$^{1}$ and Jeffrey D. Varner$^{*}$}}
 \vspace{0.10in}
-{\par \centering {School of Chemical and Biomolecular Engineering}}
+{\par \centering {Robert Frederick Smith School of Chemical and Biomolecular Engineering}}
 {\par \centering {Cornell University, Ithaca NY 14853}}
@@
 {\par \centering {$^{1}$School of Chemical Engineering}}
-{\par \centering {Purdue University, Lafayette, IN 47901}}
-\vspace{0.1in}
-{\par \centering {$^{2}$School of Chemical Engineering}}
 {\par \centering {Stanford University, Stanford, CA 94305}}
-{\par \centering \textbf{Running Title:}~Dynamic modeling of cell free protein synthesis}
+{\par \centering \textbf{Running Title:}~Dynamic modeling of cell-free protein synthesis}
 \vspace{0.1in}
@@
 {\par \centering Jeffrey D. Varner,}
-{\par \centering Professor, School of Chemical and Biomolecular Engineering,}
+{\par \centering Professor, Robert Frederick Smith School of Chemical and Biomolecular Engineering,}
 {\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
@@
 \section*{Abstract}
-Cell free protein expression systems have become widely used in systems and synthetic biology.
-In this study, we developed an ensemble of dynamic \textit{E. coli} cell free protein synthesis (CFPS) models.
+Cell-free protein expression systems have become widely used in systems and synthetic biology.
+In this study, we developed an ensemble of dynamic \textit{E. coli} cell-free protein synthesis (CFPS) models.
 Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
 The ensemble described the training data, with the exception of some of the amino acid dynamics.
-To gauge the performance of the cell free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
+To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
 balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
@@
 On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
-Taken together, we presented the first dynamic model of \textit{E. coli} cell free protein synthesis.
+Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
 This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
@@
 \vspace{0.1in}
-{\noindent \textbf{Keywords:}~Biochemical engineering, systems biology, cell free protein synthesis}
+{\noindent \textbf{Keywords:}~Biochemical engineering, systems biology, cell-free protein synthesis}
 
@@
 
-In this study, we developed an ensemble of \textit{E. coli} cell free protein synthesis (CFPS) models using the hybrid cell free modeling approach of Wayman et al [REFHERE].
+In this study, we developed an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models using the hybrid cell-free modeling approach of Wayman et al [REFHERE].
 Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
 The ensemble described the training data, with the exception of some of the amino acid dynamics.
-To gauge the performance of the cell free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
+To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
 balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
@@
 On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
-Taken together, we presented the first dynamic model of \textit{E. coli} cell free protein synthesis.
+Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
 We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
@@
 
-\subsection*{Estimation of an ensemble of cell free protein synthesis models.}
-We used the hybrid cell free modeling framework of Wayman et al. to simulate the production of a model protein [REFHERE].
+\subsection*{Estimation of an ensemble of cell-free protein synthesis models.}
+We used the hybrid cell-free modeling framework of Wayman et al. to simulate the production of a model protein [REFHERE].
 The cell-free \textit{E.
@@
 
-The ensemble of models captured the time evolution of cell free CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
+The ensemble of models captured the time evolution of cell-free CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
 Glucose was exhausted with 3 hr [FILL ME IN].
@@
 However, we first validated the ssFBA approach by comparing an ensemble of simulated versus measured concentrations of CAT over a one hour period (Fig. ~\ref{fig:CATProdYield}A). 
-The ensemble of 100 sets captured the CAT concentration profile which was randomly generated by sampling RNA polymerase levels, ribosome levels and elongation rates in a physiological range.
-We then used sequence-specific FBA to calculate a theoretical maximum CAT yield under four different cases: unconstrained, limited oxidative phosphorylation, bounded by transcription/translation rates, and bounded by experimental data (Fig.
+The ensemble of 100 sets captured the CAT concentration profile which was generated by sampling RNA polymerase levels, ribosome levels and elongation rates in a physiological range.
+We then used sequence-specific FBA to calculate a theoretical maximum CAT yield under three different cases: unconstrained, constrained by transcription/translation rates, and constrained by transcription/translation rates and measurements (Fig.
 ~\ref{fig:CATProdYield}B).
-The theoretical maximum carbon yield of CAT was 0.349 for an unconstrained case and 0.194 for the transcription and translation constrained case.
-Thus, we showed that our experimental dataset and best-fit parameter set were each producing CAT at 25\% of the theoretical maximum and 45\% of a theoretical physiological case.
+The theoretical maximum carbon yield of CAT was 0.35 $\pm$ 0.006 for an unconstrained case and 0.225 $\pm$ 0.03 for the transcription and translation constrained case.
+Thus, we showed that our experimental dataset and best-fit parameter set were each producing CAT at 25\% of the theoretical maximum and 38\% of a theoretical physiological case.
 Whereas, the case constrained by experimental data had a carbon yield of 0.062 $\pm$ 0.02, similar to the experimental yield. 
-This allowed us to quantify the amount of carbon being diverted to byproducts, and suggests that there is potential for a doubling of CAT production by reducing this diversion of carbon.
 In comparing the flux distributions between the unconstrained and constrained cases (Fig.
-~\ref{fig:Network}), both constrained cases heavily utilize the Entner–Doudoroff pathway which may be a first viable knockout to increase CAT yield.
-
-\subsection*{Sensitivity analysis on FBA system}
-We conducted global sensitivity on certain fluxes' upper bounds within a constraint-based FBA system, with protein export rate as the objective function.
+~\ref{fig:Network}), the constrained cases heavily utilize the first step in the pentose phosphate pathway to generate NADPH.
+The majority of the flux continues through the Entner–Doudoroff pathway whereas in the unconstrained case, the majority of flux travels through glycolysis.
+In all cases, the energy source comes from oxidative phosphorylation with some from the citric acid cycle.
+Whereas, in just the TX/TL case, there is a high flux through fumerate dehydrogenase from aspartic acid uptake. 
+In the unconstrained and most constrained case, we see a mixture of acetate and lactate accumulation.
+This shows the system is producing NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to supply enough NADH for oxidative phosphorylation.
+As a result, high oxidative phosphorylation activity relative to our cell free system leads to an acetate overflow.
+This suggests that there is potential for increasing CAT production by reducing this diversion of carbon.
+To simulate potential knockouts, we constrained the specific glucose and amino acid uptake rates to the same values as simulated with no knockouts.
+In an ssFBA simulation with constrained TX/TL rates, knocking out the \textit{gnd} reaction decreases flux of acetate production but increases flux through \textit{pntAB} which is responsible for regenerating NADPH.
+The simulation showed carbon was diverted towards lactate, however since CAT production is constrained by the translation rate, we expected there to be no increase in CAT production.
+The decrease in acetate production is a promising result to increase CAT yield. 
+A second simulation with a knockout of \textit{gnd} and phosphate acetyltransferase, showed carbon being diverted towards lactate and succinate, however it required a higher flux through oxidative phosphorylation and the TCA cycle to meet the energetic needs of the system.
 
@@
 
-In this study, we developed an ensemble of \textit{E. coli} cell free protein synthesis (CFPS) models using the hybrid cell free modeling approach of Wayman et al [REFHERE].
+In this study, we developed an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models using the hybrid cell-free modeling approach of Wayman et al [REFHERE].
 Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
 The ensemble described the training data, with the exception of some of the amino acid dynamics.
-To gauge the performance of the cell free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
-balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
+To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
+balance analysis. The observed CAT yield was 25\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
 The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
@@
 On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
-% Taken together, we presented the first dynamic model of \textit{E. coli} cell free protein synthesis.
+% Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
 % We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
 % This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
+
+In comparing the theoretical maximum carbon yield of CAT from ssFBA predictions to the kinetic model and experimental measurements, suggests that there is potential for increasing CAT yield in CFPS as well as CFPS performance.
+The theoretical maximum yield of CAT was 0.35 for an unconstrained case and 0.225 for a transcription/translation constrained case. 
+Knockouts of \textit{gnd} and phosphate acetyltransferase, show carbon can be diverted away from acetate and potentially towards CAT or other proteins of interest expressed in CFPS. 
+The other limitations to address in CFPS would be to enhance the transcription and translation rates since the protein of interest to be expressed is ultimately bounded by these kinetic rates.    
+Li et al.
+have increased productivity of firefly lucifease by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA and tRNAs \cite{2014_li_PlosOne}.
+Underwood et al.
+has also shown the increase in ribosome levels does not significantly increase protein yields and rates, however adding elongation factors increased yields by 23\% at 30 minutes\cite{2005_underwood_biotech}.
+In addition to improving CFPS performance, Jewett et al.
+has showed that oxidative phosphorylation operates in cell-free systems and knocking out oxidative phosphorylation reactions is detrimental to protein yield \cite{2008_jewett_molsysbiol}.
+However, it is inconclusive how much oxidative phosphorylation activity there is compared to \textit{in vivo} systems and both of our models suggest oxidative phosphorylation is vital to CAT production.
+Thus, this is a potential place for improvement to optimize CFPS for better performance and protein yield.
 
@@
 
-The cell free model ensemble described the training data with the exception of some of the amino acids.
+The cell-free model ensemble described the training data with the exception of some of the amino acids.
 Specifically, adding more reactions that consume amino acids would improve the model's ability to predict those that show a decrase in the experimental data.
@@
 The objective $w_{obj}$ was to maximize the specific rate of CAT formation.
-The specific glucose uptake rate was constrained to allow a maximum flux of 40 mM/hr according to literature data; the specific amino acid uptake rates were also bound to allow a maximum flux of 30 mM/hr, but did not reach this maximum flux.
+The specific glucose uptake rate was constrained to allow a maximum flux of 40 mM/hr according to experimental data; the specific amino acid uptake rates were bound to allow a maximum flux of 30 mM/hr, but did not reach this maximum flux.
 The transcription and translation template reactions were added to the metabolic network and are based off sequence specific analysis \cite{2002_allen_palsson} involving transcription initiaion, transcription, mRNA degradation, translation initiation, translation, and tRNA charging.
-The mRNA and protein sequence of each protein was determined from literature. 
+The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
+The solution flux vector was used to calculate the theoretical carbon yield of CAT.
+Glucose, oxygen, and amino acids were modeled as being imported into the system, whereas CAT synthesis and metabolite byproduct formation was modeled as an export from the system.
+The rest of the network followed a pseudo steady-state asusmption where all other metabolites were not allowed to accumulate; thus, the network could be solved by linear programming.
+
 The transcription rate was constrained as:
@@
 where $RNAP$ is the concentration of RNA polymerase, $v_{RNAP}$ is the elongation rate (nucleotides/hr) by the RNA polymerase, $l_{mRNA}$ is the number of nucleotides in the mRNA, $Gene$ is the gene concentration, $km$ is the plasmid saturation coefficient, and $P$ is the promoter activity.
+The mRNA and protein sequence of CAT was determined from literature.
 
@@
 where $K_{P}$ is the polysome amplification constant, $Ribo$ is the ribosome concentration,  $v_{Ribo}$ is the elongation rate (amino acids/hr) of the ribosome, $l_{protein}$ is the number of amino acids in the protein of interest, and $mRNA_{ss}$ is the mRNA concentration at steady state determined by the transcription rate divided by the degradation rate of mRNA.
-An ensemble of flux distributions was calculated for 100 sets by randonmly sampling. 
-Glucose, oxygen, and amino acids were modeled as being imported into the system, whereas CAT synthesis and metabolite byproduct formation was modeled as an export from the system.
-The rest of the network followed a pseudo steady-state asusmption where all other metabolites were not allowed to accumulate; thus, the network could be solved by linear programming.
-The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
-The solution flux vector was used to calculate the theoretical carbon yield of CAT for four different cases.
+
 %\begin{equation}\label{eqn:yield}
@@
 %where $w_{CAT}$ denotes the CAT export flux and $w_{i}$ denotes the import flux of the $i$th substrate.
-For the unconstrained case, all rates were left unbounded. 
+An ensemble of 100 sets of flux distributions was calculated for three different cases, unconstrained, constrained by transcription and translation (TX/TL) rates, and  constrained by TX/TL rates and experimental data.
+For the unconstrained case, all rates were left unbounded, except for the specific glucose uptake rate. 
 An ensemble of flux distributions was calculated by randonmly sampling the maximum specific glucose uptake rate from 30 to 40 mM/hr determined from experimental data.
-For the limited oxidative phosphorylation case, an ensemble of flux distributions was generated by randomly sampling the maximum specific oxygen uptake rate to from 0 to 10 mM/hr.
-For the TXTL case (constrained by transcription and translation rates), an ensemble was generated by randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature as well as the specific glucose uptake rate from 30 to 40 mM/hr.
-For the case where the flux was constrained by experimental data, the lower and upper bounds where data was avaible were randomly sampled in the physiological range of the data in addition to randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
+For the case constrained by TX/TL rates, an ensemble was generated by randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
+RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the elongation rate by RNA polymerase between 20 and 30 nts/sec, and the elongation rate by ribosomes between 1.5 and 3 AA/sec \cite{2005_underwood_biotech, 2016_garamella_acssynbio}.
+For the case constrained by TX/TL rates and experimental data, an ensemble was generated by randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
+The lower and upper bound constraints were randomly sampled in the physiological range of the experimental noise where data was available, except for CAT flux, which was determined from the transcription and translation rates.
 \clearpage
@@
 B.
-Theoretical maximum carbon yield of CAT calcualted by ssFBA for four different cases: unconstrained (black), limited oxidative phosphorylation activity (gray), constrained by transcription and translation rates (light grey), and constrained by experimental data (white).
+Theoretical maximum carbon yield of CAT calcualted by ssFBA for three different cases: unconstrained except for glucose uptake (black), constrained by transcription and translation rates (grey), and constrained by transcription, translation rates and experimental measurements where available (light grey).
 Error bars represent standard deviation of the ensemble.}
@@
 Sequence specific FBA flux value (mean ± standard deviation) across ensemble for 1 hr, normalized to glucose uptake flux.
-Unconstrained (top row), constrained by transcription and translation rates (second row), and constrained by experimental data (bottom row) shows the flux distribution throughout central carbon metabolism.}
+Flux distribution for three different cases: unconstrained except for glucose uptake (top row), constrained by transcription and translation rates (second row), and constrained by transcription, translation rates and experimental measurements where available (bottom row).}
 \label{fig:Network}

### 2016-11-11 Paper_v1.tex
Nov11

@@
 \usepackage{amsmath}
-\usepackage{textgreek}
+%\usepackage{textgreek}
 %\usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e}
@@
 \end{equation}
-where $\mathcal{D}$ denotes the number of datasets, $w_i$ denotes the weight of the $i$th dataset, 
-%equal to 5 for the glucose, CAT, pyruvate, lactate, acetate, succinate, and malate datasets, and 1 elsewhere,
-$\mathcal{T}_i$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ dneotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
+where $\mathcal{D}$ denotes the number of datasets, $w_i$ denotes the weight of the $i$th dataset, $\mathcal{T}_i$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ dneotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
 We then perturbed model parameters:
@@
 \end{equation}
-where $\mathcal{P}$ denotes the number of parameters, equal to 815, which includes 163 rate constants, 163 enzyme degradation rate constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i$th parameter, $k_i$ denotes the current value of the $i$th parameter, $a$ denotes a distribution variance,
-%set to 0.03, 
-and $r_i$ denotes a random sample from the normal distribution.
+where $\mathcal{P}$ denotes the number of parameters, equal to 815, which includes 163 rate constants, 163 enzyme degradation rate constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i$th parameter, $k_i$ denotes the current value of the $i$th parameter, $a$ denotes a distribution variance, and $r_i$ denotes a random sample from the normal distribution.
 For each newly generated parameter set, we re-solved the balance equations and calculated the cost function.
@@
 After generating 12,437 sets, we selected 100 sets with minimal correlation to each other so as to avoid over-sampling any region of parameter space.
-The original 12,437-set ensemble had a [mean,median,maximum] Pearson correlation coefficient [REFERENCE NEEDED?] of [?] between pairs of sets; the 100-set ensemble had a [mean,median,maximum] Pearson correlation coefficient of [?] between pairs of sets.
+The original 12,437-set ensemble had a mean Pearson correlation coefficient [REFERENCE NEEDED?] of 0.93 between pairs of sets; the 100-set ensemble had a mean Pearson correlation coefficient of 0.84 between pairs of sets.
 
@@
 \centering
-\includegraphics[width=1.00\textwidth,trim=0 0 263 0,clip]{./Figures/ClusterCAT.pdf}
+\includegraphics[width=0.8\textwidth,trim=0 28 615 40,clip]{./Figures/Sensitivity.pdf}
 \caption{Normalized first-order and pairwise sensitivities of CAT production to rate constants.}
-\label{fig:ClusterCAT}
+\label{fig:Sensitivity}
 \end{figure}
@@
 \centering
-\includegraphics[width=1.15\textwidth,trim=0 0 263 0,clip]{./Figures/ClusterSystem.pdf}
-\caption{Normalized first-order and pairwise sensitivities of system state to rate constants. A (most sensitive parameters): hexokinase run by GTP, and lactate dehydrogenase forward reaction. B: }
-\label{fig:ClusterSystem}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=0.75\textwidth]{./Figures/CAT_prod_yield.pdf}
+\includegraphics[width=0.95\textwidth]{./Figures/CAT_prod_yield.pdf}
 \caption{Sequence specific flux balance analysis of CAT production and yield.

### 2016-11-11 Paper_v1_.tex
Nov11

--- 2016-11-10 Paper_v1.tex
@@
 \begin{titlepage}
-{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell Free Protein Synthesis in \emph{Escherichia~coli}}}}
+{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell-Free Protein Synthesis in \emph{Escherichia~coli}}}}
 \vspace{0.05in}
@@
 {\par \centering {Stanford University, Stanford, CA 94305}}
-{\par \centering \textbf{Running Title:}~Dynamic modeling of cell free protein synthesis}
+{\par \centering \textbf{Running Title:}~Dynamic modeling of cell-free protein synthesis}
 \vspace{0.1in}
@@
 An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids.
-We generated an ensemble of N = 18,000 parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model.
-We defined the set with the lowest error value as the best-fit parameter set.
-[STATISTICS ON PARAMETERS].
+We generated an ensemble of N = 100 diverse parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model, and choosing sets so as to minimize the Pearson correlation coefficient between sets in the ensemble.
 
-The ensemble of models captured the time evolution of cell free CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
-Glucose was exhausted with 3 hr [FILL ME IN].
-The ensemble also captured the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:CarbonBoth}, top) and the totals by base .
+The ensemble of models captured the time evolution of cell-free CAT biosynthesis (Fig.~\ref{fig:BothCarbon} - \ref{fig:Amino}).
+%Glucose was exhausted with 3 hr [FILL ME IN].
+The ensemble also captured the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:BothCarbon}, top) and the totals by base (Fig.~\ref{fig:Energy}).
 The ensemble and the best-fit set also predict some of the amino acid measurements, while failing to predict others (Fig.~\ref{fig:Amino}).
 the central carbon metabolism, including glucose uptake, CAT production, and the dynamics of the organic acid intermediates .
-Allosteric control is important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).
+Allosteric control is important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:BothCarbon}, bottom).
 This is likely due to a structural deficiency in the model; in some cases, the consumption of an amino acid through CAT synthesis is not enough to explain the decrease shown in the data, and there are no other reactions that consume it.
 Thus, a more comprehensive biological description is needed to fully explain amino acid dynamics.
-
-\subsection*{Sensitivity analysis}
-We performed a local sensitivity analysis to determine the network reactions with the greatest effect on protein production and overall system state.
-CAT production was msot sensitive to the CAT synthesis reaction, oxidative phosphorylation activity, and alanine synthesis, as well as various reactions in glycolysis, the TCA cycle, and amino acid synthesis and degradation.
-
-%We conducted a local sensitivity analysis to determine which of the kinetic and control parameters affected model performance.
-%We calculated performance as area under the CAT curve, which was directly related to CAT synthesis rate, as the culture time was fixed and no CAT degradation was modeled.
-%We randomly chose 180 sets of the 18,000 sets in the ensemble and defined these as our sub-ensemble; for each set in the sub-ensemble, we varied the rate constant and saturation constants of each metabolic flux and measured the resulting change in CAT production to estimate the sensitivity of model performance to that parameter.
-%We did the same for the control parameters, both the order (Hill coefficient) and gain (related to the the dissociation constant).
-%This allowed us to estimate the relative importance of the kinetic and control parameters to CAT production across the ensemble.
-%Of the rate constants, those with the highest positive sensitivities were CAT synthesis, GTP synthesis, GMP synthesis, glutamine synthesis, and aspartate synthesis (Fig.~\ref{fig:LocalSensitivity}, top).
-%This is explained by GTP and the amino acids being reactants for CAT synthesis.
-%Also, GMP synthesis increases the total amount of guanosine, allowing for more GTP production.
-%The rate constants with the largest negative sensitivities were GTP degradation, arginine synthesis, and UMP synthesis.
-%While GTP degradation is obvious, the others can be explained in that they consume ATP as well as several amino acids, all of which are reactants for CAT synthesis.
-%Of the saturation constants, the reverse is seen: the largest positive sensitivities are those associated with arginine synthesis, while the largest negative sensitivities are those associated with CAT synthesis, GTP synthesis, and GMP synthesis (Fig.~\ref{fig:LocalSensitivity}, middle).
-%This is because an increase in saturation constant causes a decrease in the corresponding reaction rate.
-%The control parameters were seen to be the least significant and the most uncertain (Fig.~\ref{fig:LocalSensitivity}, bottom).
-%Only two had a small standard error across the ensemble, relative to the ensemble mean sensitivity: the gain and order for pyruvate acting as an inhibitor on the pdh reaction.
-%This could be because pdh consumes pyruvate and diverts carbon away from the pathways that ultimately contribute to CAT production.
-%Taken together with the lack of change in glucose uptake and CAT production when control is removed (Fig.~\ref{fig:CarbonBoth}), this suggests that allosteric control is not the limiting factor to CAT production.
-
-%We conducted a global sensitivity analysis on the parameters that could be controlled experimentally: the initial conditions of glucose, oxygen, amino acids, and enzymes.
-%We used the variance-based method of Sobol, and the same objective function of area under the CAT curve.
-%Using a parameter set of relatively good fit against data, we defined parameter bounds and generated a Sobol sequence of 111,600 parameter values that fit within those bounds.
-%We then calculated the total-order sensitivity and confidence interval for each of the experimentally controllable initial conditions.
-%As the sensitivities were total-order, they were guaranteed to be non-negative.
-%The largest sensitivities belonged to the initial conditions of the CAT macromolecular synthesis machinery, GTP synthase, GTP degradation, some amino acids such as phenylalanine, proline, and leucine, and some amino acid synthases (Fig.~\ref{fig:GlobalSensitivity}).
-%This is all explained by GTP and amino acids being reactants for CAT synthesis.
-%While some of the amino acids and amino acid synthases were among the highest in sensitivity, theirs were also very uncertain relative to the CAT macromolecular synthesis machinery and GTP synthase.
-%The initial conditions of glucose and oxygen were among the least important according to the global sensitivity analysis, suggesting that the model predicts that CAT production can be sustained by consuming initial stores or can be powered by other means.
 
@@
 
+\subsection*{Sensitivity analysis}
+We conducted a local sensitivity analysis to determine which of the kinetic and control parameters affected model performance.
+We calculated performance as area under the CAT curve, which was directly related to CAT synthesis rate, as the culture time was fixed and no CAT degradation was modeled.
+We randomly chose 180 sets of the 18,000 sets in the ensemble and defined these as our sub-ensemble; for each set in the sub-ensemble, we varied the rate constant and saturation constants of each metabolic flux and measured the resulting change in CAT production to estimate the sensitivity of model performance to that parameter.
+We did the same for the control parameters, both the order (Hill coefficient) and gain (related to the the dissociation constant).
+This allowed us to estimate the relative importance of the kinetic and control parameters to CAT production across the ensemble.
+Of the rate constants, those with the highest positive sensitivities were CAT synthesis, GTP synthesis, GMP synthesis, glutamine synthesis, and aspartate synthesis (Fig.~\ref{fig:LocalSensitivity}, top).
+This is explained by GTP and the amino acids being reactants for CAT synthesis.
+Also, GMP synthesis increases the total amount of guanosine, allowing for more GTP production.
+The rate constants with the largest negative sensitivities were GTP degradation, arginine synthesis, and UMP synthesis.
+While GTP degradation is obvious, the others can be explained in that they consume ATP as well as several amino acids, all of which are reactants for CAT synthesis.
+Of the saturation constants, the reverse is seen: the largest positive sensitivities are those associated with arginine synthesis, while the largest negative sensitivities are those associated with CAT synthesis, GTP synthesis, and GMP synthesis (Fig.~\ref{fig:LocalSensitivity}, middle).
+This is because an increase in saturation constant causes a decrease in the corresponding reaction rate.
+The control parameters were seen to be the least significant and the most uncertain (Fig.~\ref{fig:LocalSensitivity}, bottom).
+Only two had a small standard error across the ensemble, relative to the ensemble mean sensitivity: the gain and order for pyruvate acting as an inhibitor on the pdh reaction.
+This could be because pdh consumes pyruvate and diverts carbon away from the pathways that ultimately contribute to CAT production.
+Taken together with the lack of change in glucose uptake and CAT production when control is removed (Fig.~\ref{fig:BothCarbon}), this suggests that allosteric control is not the limiting factor to CAT production.
+
+We conducted a global sensitivity analysis on the parameters that could be controlled experimentally: the initial conditions of glucose, oxygen, amino acids, and enzymes.
+We used the variance-based method of Sobol, and the same objective function of area under the CAT curve.
+Using a parameter set of relatively good fit against data, we defined parameter bounds and generated a Sobol sequence of 111,600 parameter values that fit within those bounds.
+We then calculated the total-order sensitivity and confidence interval for each of the experimentally controllable initial conditions.
+As the sensitivities were total-order, they were guaranteed to be non-negative.
+The largest sensitivities belonged to the initial conditions of the CAT macromolecular synthesis machinery, GTP synthase, GTP degradation, some amino acids such as phenylalanine, proline, and leucine, and some amino acid synthases (Fig.~\ref{fig:GlobalSensitivity}).
+This is all explained by GTP and amino acids being reactants for CAT synthesis.
+While some of the amino acids and amino acid synthases were among the highest in sensitivity, theirs were also very uncertain relative to the CAT macromolecular synthesis machinery and GTP synthase.
+The initial conditions of glucose and oxygen were among the least important according to the global sensitivity analysis, suggesting that the model predicts that CAT production can be sustained by consuming initial stores or can be powered by other means.
+
 \subsection*{Sensitivity analysis on FBA system}
@@
 On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
-% Taken together, we presented the first dynamic model of \textit{E. coli} cell free protein synthesis.
+% Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
 % We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
@@
 \end{equation}
-where $\mathcal{D}$ denotes the number of datasets, $w_i$ denotes the weight of the $i$th dataset, 
+where $\mathcal{D}$ denotes the number of datasets, equal to 37, $w_i$ denotes the weight of the $i$th dataset, 
 %equal to 5 for the glucose, CAT, pyruvate, lactate, acetate, succinate, and malate datasets, and 1 elsewhere,
@@
 After generating 12,437 sets, we selected 100 sets with minimal correlation to each other so as to avoid over-sampling any region of parameter space.
-The original 12,437-set ensemble had a [mean,median,maximum] Pearson correlation coefficient [REFERENCE NEEDED?] of [?] between pairs of sets; the 100-set ensemble had a [mean,median,maximum] Pearson correlation coefficient of [?] between pairs of sets.
+The original 12,437-set ensemble had a mean Pearson correlation coefficient [REFERENCE NEEDED?] of 0.94 between pairs of sets, while the 100-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.
 
-\subsection*{Sensitivity analysis}
+\subsection*{Local sensitivity analysis}
 We determined the reactions most important to protein production by computing the local sensitivity of CAT concentration to each rate constant in the network.
@@
 In calculating the pairwise sensitivities, each parameter was perturbed by 1\%; first-order sensitivities ($i$ = $j$) were subject to two 1\% perturbations, equivalent to a perturbation of 2.01\%.
+The array of sensitivity indices was organized into a clustergram (Fig.~\ref{fig:ClusterCAT}) with x-axis corresponding to the 163 rate constants in order, and y-axis corresponding to the 163 rate constants rearranged from least sensitive (top) to most sensitive (bottom).
+A dendrogram (left) illustrates a hierarchy of rate constants with similar effect on CAT production, separating them into high responders (dark gray background), medium responders (medium gray background), and low responders (light gray background).
+
+Likewise, we determined the reactions most important to the system as a whole by computing the local sensitivity of the species for which data exist to each rate constant in the network.
+Each sensitivity index was formulated as:
+\begin{equation}\label{eqn:local-sensitivity}
+\begin{split}
+    S_{ij}=norm(X(p_i,p_j,t)-X(\alpha*p_i,\alpha*p_j,t))\qquad{i,j=1,2,\hdots\mathcal{P}}\\
+\end{split}
+\end{equation}
+where $S_{ij}$ denotes the sensitivity of the system state to the $i$th and $j$th parameters, and $X(p_i,p_j,t)$ denotes the system state, an array consisting of the concentration of every species for which data exist as a function of time and the $i$th and $j$th parameters.
+% $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
+This array of sensitivity indices was also organized into a clustergram (Fig.~\ref{fig:ClusterSystem}), with corresponding dendrogram (left).
 
@@
 \section*{Funding}
-This study was supported by a National Science Foundation Graduate Research Fellowship (DGE-1333468) to N.H and by an award from the US Army and Systems Biology of Trauma Induced Coagulopathy (W911NF-10-1-0376) to J.V for the support of M.V.
+This study was supported by a National Science Foundation Graduate Research Fellowship (DGE-1333468) to N.H and by an award from
+the US Army and Systems Biology of Trauma Induced Coagulopathy (W911NF-10-1-0376) to J.V for the support of M.V.
 
@@
 
+% Figures and captions go here ...
 \begin{figure}[ht]
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/CarbonBoth.pdf}
+\includegraphics[width=0.75\textwidth]{./Figures/CAT_prod_yield.pdf}
+\caption{Sequence specific flux balance analysis of CAT production and yield.
+A.
+95\% confidence interval of the ensemble (light blue region) for CAT concentration versus time.
+B.
+Theoretical maximum carbon yield of CAT calcualted by ssFBA for four different cases: unconstrained (black), limited oxidative phosphorylation activity (gray), constrained by transcription and translation rates (light grey), and constrained by experimental data (white).
+Error bars represent standard deviation of the ensemble.}
+\label{fig:CATProdYield}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.8\textwidth]{./Figures/CAT_flux_final.pdf}
+\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, and oxidative phosphorylation.
+Sequence specific FBA flux value (mean ± standard deviation) across ensemble for 1 hr, normalized to glucose uptake flux.
+Unconstrained (top row), constrained by transcription and translation rates (second row), and constrained by experimental data (bottom row) shows the flux distribution throughout central carbon metabolism.}
+\label{fig:Network}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/Carbon.pdf}
+\includegraphics[width=0.05\textwidth]{./Figures/null.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/CarbonNoControl.pdf}
 \caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species.
@@
 95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
-\label{fig:CarbonBoth}
+\label{fig:BothCarbon}
 \end{figure}
@@
 
-\begin{figure}[ht]
-\centering
-\includegraphics[width=1.00\textwidth]{./Figures/BoxPlot-KCFM.pdf}
-\caption{Log of cost function across 37 datasets for data-trained ensemble (blue) and randomly generated ensemble (red, gray background).
-Median (bars), interquartile range (boxes), range excluding outliers (dashed lines), and outliers (circles) for each dataset.
-Median across all datasets (large bar overlaid).}
-\label{fig:BoxPlot}
-\end{figure}
+%\begin{table}[ht]
+%\centering
+%\includegraphics[width=0.94\textwidth]{./Figures/AIC.pdf}
+%\caption{}
+%\label{fig:AIC}
+%\end{table}
 
@@
 \centering
-\includegraphics[width=1.00\textwidth,trim=0 0 263 0,clip]{./Figures/ClusterCAT.pdf}
-\caption{Normalized first-order and pairwise sensitivities of CAT production to rate constants.}
+\includegraphics[width=0.94\textwidth]{./Figures/ClusterCAT161.pdf}
+\caption{Normalized first-order and pairwise sensitivities of CAT production to rate constants.
+Dark pixels indicate the most sensitive indices.
+Dendrogram indicates high responders (dark gray background), medium responders (medium gray background), and low responders (light gray background).}
 \label{fig:ClusterCAT}
@@
 \centering
-\includegraphics[width=1.15\textwidth,trim=0 0 263 0,clip]{./Figures/ClusterSystem.pdf}
-\caption{Normalized first-order and pairwise sensitivities of system state to rate constants. A (most sensitive parameters): hexokinase run by GTP, and lactate dehydrogenase forward reaction. B: }
+\includegraphics[width=1.00\textwidth]{./Figures/ClusterSystem.pdf}
+\caption{Normalized first-order and pairwise sensitivities of system state to rate constants.
+Dark pixels indicate the most sensitive indices.
+Dendrogram indicates high responders (dark gray background), medium responders (medium gray background), and low responders (light gray background).}
 \label{fig:ClusterSystem}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./Figures/CAT_prod_yield.pdf}
-\caption{Sequence specific flux balance analysis of CAT production and yield.
-A.
-95\% confidence interval of the ensemble (light blue region) for CAT concentration versus time.
-B.
-Theoretical maximum carbon yield of CAT calcualted by ssFBA for four different cases: unconstrained (black), limited oxidative phosphorylation activity (gray), constrained by transcription and translation rates (light grey), and constrained by experimental data (white).
-Error bars represent standard deviation of the ensemble.}
-\label{fig:CATProdYield}
+\includegraphics[width=1.00\textwidth]{./Figures/AA_bounded_bar.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/AA_bounded_scatter.pdf}
+\caption{Total-order global sensitivities of protein export flux to amino acid import flux upper bounds. Each upper bound was varied from 0 to 0.5 mM/hr.}
+\label{fig:AAs_bounded}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.8\textwidth]{./Figures/CAT_flux_final.pdf}
-\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, and oxidative phosphorylation.
-Sequence specific FBA flux value (mean ± standard deviation) across ensemble for 1 hr, normalized to glucose uptake flux.
-Unconstrained (top row), constrained by transcription and translation rates (second row), and constrained by experimental data (bottom row) shows the flux distribution throughout central carbon metabolism.}
-\label{fig:Network}
+\includegraphics[width=1.00\textwidth]{./Figures/AA_unbounded_bar.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/AA_unbounded_scatter.pdf}
+\caption{Total-order global sensitivities of protein export flux to amino acid import flux upper bounds.
+Each upper bound was varied from 0 to 0.5 mM/hr, except the upper bound with respect to which sensitivity was calculated; this was varied from 0 to 10 mM/hr.}
+\label{fig:AAs_unbounded}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/C_Secretion.pdf}
+\caption{Total-order global sensitivities of protein export flux to glutamic acid, oxygen and glucose import flux upper bounds, and translation initiation rate upper bound.
+Lactate, ethanol, acetate, and formate export fluxes were held constant at 0.1 mM/hr in the control case.
+One of these export fluxes was held constant at 10 mM/hr, while the others were held constant at 0.1 mM/hr, in each of the other four cases.}
+\label{fig:C_Secretion.pdf}
 \end{figure}

### 2016-11-11 Paper_v1.tex
Nov11_3

@@
 An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids.
-We generated an ensemble of N = 18,000 parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model.
-We defined the set with the lowest error value as the best-fit parameter set.
-[STATISTICS ON PARAMETERS].
+We generated an ensemble of N = 100 diverse parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model, and choosing sets so as to minimize the Pearson correlation coefficient between sets in the ensemble.
 
 The ensemble of models captured the time evolution of cell-free CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
-Glucose was exhausted with 3 hr [FILL ME IN].
-The ensemble also captured the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:CarbonBoth}, top) and the totals by base .
-The ensemble and the best-fit set also predict some of the amino acid measurements, while failing to predict others (Fig.~\ref{fig:Amino}).
-the central carbon metabolism, including glucose uptake, CAT production, and the dynamics of the organic acid intermediates .
+The ensemble and the best-fit set also captured the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:CarbonBoth}, top) and the totals by base, as well as the amino acid measurements (Fig.~\ref{fig:Amino}).
 Allosteric control is important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).
-This is likely due to a structural deficiency in the model; in some cases, the consumption of an amino acid through CAT synthesis is not enough to explain the decrease shown in the data, and there are no other reactions that consume it.
-Thus, a more comprehensive biological description is needed to fully explain amino acid dynamics.
 
@@
 
-\subsection*{Sensitivity analysis}
+\subsection*{Local sensitivity analysis}
 We determined the reactions most important to protein production by computing the local sensitivity of CAT concentration to each rate constant in the network.
@@
 In calculating the pairwise sensitivities, each parameter was perturbed by 1\%; first-order sensitivities ($i$ = $j$) were subject to two 1\% perturbations, equivalent to a perturbation of 2.01\%.
+The array of sensitivity indices was organized into a clustergram (Fig.~\ref{fig:Sensitivity}, top) with both axes corresponding to the 163 rate constants rearranged from most sensitive (top, right) to least sensitive (bottom, left).
+
+Likewise, we determined the reactions most important to the system as a whole by computing the local sensitivity of the species for which data exist to each rate constant in the network.
+Each sensitivity index was formulated as:
+\begin{equation}\label{eqn:local-sensitivity}
+\begin{split}
+    S_{ij}=norm(X(p_i,p_j,t)-X(\alpha*p_i,\alpha*p_j,t))\qquad{i,j=1,2,\hdots\mathcal{P}}\\
+\end{split}
+\end{equation}
+where $S_{ij}$ denotes the sensitivity of the system state to the $i$th and $j$th parameters, and $X(p_i,p_j,t)$ denotes the system state, an array consisting of the concentration of every species for which data exist as a function of time and the $i$th and $j$th parameters.
+% $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
 
@@
 \includegraphics[width=0.8\textwidth,trim=0 28 615 40,clip]{./Figures/Sensitivity.pdf}
-\caption{Normalized first-order and pairwise sensitivities of CAT production to rate constants.}
+\caption{Normalized first-order and pairwise sensitivities of CAT production (top) and system state (bottom) to rate constants.}
 \label{fig:Sensitivity}

### 2016-11-11 Paper_v1_.tex
Nov11_4

--- 2016-11-11 Paper_v1.tex
@@
 Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
-The ensemble described the training data, with the exception of some of the amino acid dynamics.
-To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
-balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
-The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
-The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
-while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
-Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
-GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
+The ensemble described all of the training data, especially the central carbon metabolism.
+To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux balance analysis.
+The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
+%The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
+The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle, while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
+Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
 On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
@@
 The ensemble described the training data, with the exception of some of the amino acid dynamics.
-To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
-balance analysis. The observed CAT yield was 25\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
+To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield with the maximum theoretical CAT carbon yield calculated using sequence specific flux balance analysis.
+The observed CAT yield was 25\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
 The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
-The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
-while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
-Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
-GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
+The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle, while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
+Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
 On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
@@
 \end{equation}
-where $\mathcal{D}$ denotes the number of datasets, $w_i$ denotes the weight of the $i$th dataset, $\mathcal{T}_i$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ dneotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
+where $\mathcal{D}$ denotes the number of datasets, equal to 37, $w_i$ denotes the weight of the $i$th dataset, $\mathcal{T}_i$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ denotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
 We then perturbed model parameters:
@@
 After generating 12,437 sets, we selected 100 sets with minimal correlation to each other so as to avoid over-sampling any region of parameter space.
-The original 12,437-set ensemble had a mean Pearson correlation coefficient [REFERENCE NEEDED?] of 0.93 between pairs of sets; the 100-set ensemble had a mean Pearson correlation coefficient of 0.84 between pairs of sets.
+The original 12,437-set ensemble had a mean Pearson correlation coefficient [REFERENCE NEEDED?] of 0.94 between pairs of sets, while the 100-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.
 
-\subsection*{Sensitivity analysis}
+\subsection*{Local sensitivity analysis}
 We determined the reactions most important to protein production by computing the local sensitivity of CAT concentration to each rate constant in the network.
@@
 In calculating the pairwise sensitivities, each parameter was perturbed by 1\%; first-order sensitivities ($i$ = $j$) were subject to two 1\% perturbations, equivalent to a perturbation of 2.01\%.
+The array of sensitivity indices was organized into a clustergram (Fig.~\ref{fig:Sensitivity}, top) with both axes corresponding to the 163 rate constants rearranged from most sensitive (top, right) to least sensitive (bottom, left).
+
+Likewise, we determined the reactions most important to the system as a whole by computing the local sensitivity of the species for which data exist to each rate constant in the network.
+Each sensitivity index was formulated as:
+\begin{equation}\label{eqn:local-sensitivity}
+\begin{split}
+    S_{ij}=norm(X(p_i,p_j,t)-X(\alpha*p_i,\alpha*p_j,t))\qquad{i,j=1,2,\hdots\mathcal{P}}\\
+\end{split}
+\end{equation}
+where $S_{ij}$ denotes the sensitivity of the system state to the $i$th and $j$th parameters, and $X(p_i,p_j,t)$ denotes the system state, an array consisting of the concentration of every species for which data exist as a function of time and the $i$th and $j$th parameters.
+% $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
+This array of sensitivity indices was also organized into a clustergram (Fig.~\ref{fig:Sensitivity}, bottom).
 
@@
 
+% Figures and captions go here ...
 \begin{figure}[ht]

### 2016-11-12 Paper_v1.tex
Nov12

@@
 
-%The results are presented in \textbf{past~tense}. Each paragraph starts with a statement of the result in that paragraph in active voice.
-%Each results paragraph ends with a Taken together type statement followed by a link statement e.g., Next we considered etc. When referring to figures, state what the figures shows (Fig. ZZ).
-
-%\begin{enumerate}
-%	\item{\textbf{First~section}:Description of the model biology}
-%	\item{\textbf{Second~section}:Estimation of the model parameters, and refinement of the model structure (inclusion of the AA degradation pathways)}
-%	\item{\textbf{Third~section}:Analysis of the flux distribution (over the ensemble?), sensitivity results (first parameters, then AA)}
-%\end{enumerate}
-
-%We generated mass balance equations around metabolites and enzymes in the network, modeled as ordinary differential equations with reaction rates equal to the product of a kinetic term and a control term.
-We used multiple saturation kinetics to model metabolic fluxes and mass action kinetics to model enzyme degradation.
-%We modeled the control term using a rule-based approach in which each control factor had a regulatory transfer function, modeled as a Hill function, and the control term was calculated as the mean of all transfer functions.
-%The model was trained against the training dataset, which included measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (ATP, ADP, AMP, GTP, GDP, GMP, CTP, CDP, CMP, UTP, UDP, UMP), and 18 of the 20 proteinogenic amino acids.
-%The ensemble predicts most carbon flux going through glycolysis and the TCA cycle, and to a lesser extent the Entner-Doudoroff pathway  (Fig.~\ref{fig:Network}, bottom two rows of gray boxes).
-Pentose phosphate is much less utilized.
-%We compared this flux profile with the steady-state fluxes calculated by sequence-specific FBA (Fig.~\ref{fig:Network}, top row of gray boxes).
-%In the steady-state case, carbon flux avoids the upper reactions of glycolysis in favor of the Entner-Doudoroff pathway.
-%It also not does participate in the TCA cycle, nor the reactions of pentose phosphate that are not required for Entner-Doudoroff.
-
 \subsection*{Estimation of an ensemble of cell-free protein synthesis models.}
@@
 coli} metabolic model was constructed by removing the growth-associated processes from the model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS}, and by adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
-Thus, the model described core central carbon metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle),
-as well as the synthesis of energy species, amino acids biosynthesis and degradation, and biosynthesis of the CAT protein.
+Thus, the model described core central carbon metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle), as well as the synthesis of energy species, amino acids biosynthesis and degradation, and biosynthesis of the CAT protein.
 An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids.
 We generated an ensemble of N = 100 diverse parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model, and choosing sets so as to minimize the Pearson correlation coefficient between sets in the ensemble.
+We also defined the single set with lowest cost function as our best-fit set.
 
 The ensemble of models captured the time evolution of cell-free CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
-The ensemble and the best-fit set also captured the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:CarbonBoth}, top) and the totals by base, as well as the amino acid measurements (Fig.~\ref{fig:Amino}).
-Allosteric control is important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).
+The central carbon metabolism (Fig.~\ref{fig:CarbonBoth}, top), energy species (Fig.~\ref{fig:Energy}), and amino acids (Fig.~\ref{fig:Amino}) were all well captured by the ensemble and the best-fit set.
+Allosteric control was seen to be particularly important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).
 
@@
 We performed a local sensitivity analysis to determine the network reactions with the greatest effect on protein production and overall system state.
-CAT production was msot sensitive to the CAT synthesis reaction, oxidative phosphorylation activity, and alanine synthesis, as well as various reactions in glycolysis, the TCA cycle, and amino acid synthesis and degradation.
+%CAT production was most sensitive to the CAT synthesis reaction, oxidative phosphorylation activity, and alanine synthesis, as well as various reactions in glycolysis, pentose phosphate, the TCA cycle, and amino acid synthesis and degradation (Fig.~\ref{fig:Sensitivity}, top).
+CAT production was most sensitive to the CAT synthesis reaction, oxidative phosphorylation activity, and alanine synthesis (Fig.~\ref{fig:Sensitivity}, top, section A).
+The 16 next most important reactions to CAT production (section B) came from various pathways across the network: 4 each from glycolysis, the TCA cycle, and amino acid synthesis/degradation; 2 from pentose phosphate; and 1 each from the Entner-Doudoroff pathway and the energy species reactions.
+The pairwise sensitivities (off-diagonal elements) are often quite different from the corresponding first-order sensitivities (diagonal elements), and lead to some interesting outcomes.
+For example, glutamine synthesis and arginine degradation are both among the most important reactions to CAT production (they rank 5th and 10th, respectively).
+This is likely because they both affect the sensitive glutamine-glutamate balance; glutamine synthesis consumes glutamate, while arginine degradation produces it.
+However, when both are perturbed, their combined effect on the model is very low, as the respective consumption and production of glutamate cancel out.
 
-%We conducted a local sensitivity analysis to determine which of the kinetic and control parameters affected model performance.
-%We calculated performance as area under the CAT curve, which was directly related to CAT synthesis rate, as the culture time was fixed and no CAT degradation was modeled.
-%We randomly chose 180 sets of the 18,000 sets in the ensemble and defined these as our sub-ensemble; for each set in the sub-ensemble, we varied the rate constant and saturation constants of each metabolic flux and measured the resulting change in CAT production to estimate the sensitivity of model performance to that parameter.
-%We did the same for the control parameters, both the order (Hill coefficient) and gain (related to the the dissociation constant).
-%This allowed us to estimate the relative importance of the kinetic and control parameters to CAT production across the ensemble.
-%Of the rate constants, those with the highest positive sensitivities were CAT synthesis, GTP synthesis, GMP synthesis, glutamine synthesis, and aspartate synthesis (Fig.~\ref{fig:LocalSensitivity}, top).
-%This is explained by GTP and the amino acids being reactants for CAT synthesis.
-%Also, GMP synthesis increases the total amount of guanosine, allowing for more GTP production.
-%The rate constants with the largest negative sensitivities were GTP degradation, arginine synthesis, and UMP synthesis.
-%While GTP degradation is obvious, the others can be explained in that they consume ATP as well as several amino acids, all of which are reactants for CAT synthesis.
-%Of the saturation constants, the reverse is seen: the largest positive sensitivities are those associated with arginine synthesis, while the largest negative sensitivities are those associated with CAT synthesis, GTP synthesis, and GMP synthesis (Fig.~\ref{fig:LocalSensitivity}, middle).
-%This is because an increase in saturation constant causes a decrease in the corresponding reaction rate.
-%The control parameters were seen to be the least significant and the most uncertain (Fig.~\ref{fig:LocalSensitivity}, bottom).
-%Only two had a small standard error across the ensemble, relative to the ensemble mean sensitivity: the gain and order for pyruvate acting as an inhibitor on the pdh reaction.
-%This could be because pdh consumes pyruvate and diverts carbon away from the pathways that ultimately contribute to CAT production.
-%Taken together with the lack of change in glucose uptake and CAT production when control is removed (Fig.~\ref{fig:CarbonBoth}), this suggests that allosteric control is not the limiting factor to CAT production.
-
-%We conducted a global sensitivity analysis on the parameters that could be controlled experimentally: the initial conditions of glucose, oxygen, amino acids, and enzymes.
-%We used the variance-based method of Sobol, and the same objective function of area under the CAT curve.
-%Using a parameter set of relatively good fit against data, we defined parameter bounds and generated a Sobol sequence of 111,600 parameter values that fit within those bounds.
-%We then calculated the total-order sensitivity and confidence interval for each of the experimentally controllable initial conditions.
-%As the sensitivities were total-order, they were guaranteed to be non-negative.
-%The largest sensitivities belonged to the initial conditions of the CAT macromolecular synthesis machinery, GTP synthase, GTP degradation, some amino acids such as phenylalanine, proline, and leucine, and some amino acid synthases (Fig.~\ref{fig:GlobalSensitivity}).
-%This is all explained by GTP and amino acids being reactants for CAT synthesis.
-%While some of the amino acids and amino acid synthases were among the highest in sensitivity, theirs were also very uncertain relative to the CAT macromolecular synthesis machinery and GTP synthase.
-%The initial conditions of glucose and oxygen were among the least important according to the global sensitivity analysis, suggesting that the model predicts that CAT production can be sustained by consuming initial stores or can be powered by other means.
+The system state as a whole was most sensitive to glucose uptake via GTP and the forward reaction of lactate dehydrogenase (pyruvate being consumed to produce lactate) (Fig.~\ref{fig:Sensitivity}, bottom, section F).
+The 30 next most important reactions to the system state (section G) came from various pathways across the network: 8 from amino acid synthesis/degradation; 6 from glycolysis; 4 from TCA; 2 each from pentose phosphate, Entner-Doudoroff, energy species reactions, NAD species reactions, and small molecule transport; 1 from oxidative phosphorylation, and 1 pyrophosphatase.
+The system state has even more pairwise sensitivities that differ from the corresponding first-order sensitivities and stand out as significant.
+For example, the first-order effect of alanine synthesis is very high; it consumes both pruvate and glutamate, two key species in the network.
+In addition, a handful of alanine synthesis pairwise sensitivities are high.
+However, there are enough reactions that, when paired with alanine synthesis, have little effect on the model.
+Malic enzyme is one of these, as it produces the pyruvate that alanine synthesis consumes.
+Thus, the total-order alanine synthesis sensitivity is low, placing it at the very bottom of section I.
+Another interesting result of this analysis is seen in the intersection of sections F and G with section J.
+The 53 reactions in section J were turned off in the best-fit set (rate constants were set to 0); therefore, the perturbation of these reactions (multiplying the rate constant by 1.01) had no actual effect on the model.
+Thus, all pairwise sensitivities with reactions in section J can essentially be considered first-order sensitivities for the other reactions.
+Interestingly, the reactions in section F and several in section G showed most of thier highest sensitivities when paired with the "non-effects" of section J.
+Of these, three involved pyruvate, strengthening its role as a key metabolite; the others were glucose uptake via GTP and CTP, fumarate reductase, and SO4 transport.
+This suggests that these reactions' effects on the model were canceled out or lessened by most other reactions, but were of course not affected by the reactions in section J.
 
@@
 After generating 12,437 sets, we selected 100 sets with minimal correlation to each other so as to avoid over-sampling any region of parameter space.
-The original 12,437-set ensemble had a mean Pearson correlation coefficient [REFERENCE NEEDED?] of 0.93 between pairs of sets; the 100-set ensemble had a mean Pearson correlation coefficient of 0.84 between pairs of sets.
+The original 12,437-set ensemble had a mean Pearson correlation coefficient [REFERENCE NEEDED?] of 0.94 between pairs of sets, while the 100-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.
 
@@
 % $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
+This array of sensitivity indices was also organized into a clustergram (Fig.~\ref{fig:Sensitivity}, bottom).
 

### 2016-11-14 Paper_v1.tex
Nov14
Nov14_2

@@
 The system state as a whole was most sensitive to glucose uptake via GTP and the forward reaction of lactate dehydrogenase (pyruvate being consumed to produce lactate) (Fig.~\ref{fig:Sensitivity}, bottom, section F).
-The 30 next most important reactions to the system state (section G) came from various pathways across the network: 8 from amino acid synthesis/degradation; 6 from glycolysis; 4 from TCA; 2 each from pentose phosphate, Entner-Doudoroff, energy species reactions, NAD species reactions, and small molecule transport; 1 from oxidative phosphorylation, and 1 pyrophosphatase.
+The 30 next most important reactions to the system state (section G) came from various pathways across the network: 8 from amino acid synthesis/degradation; 6 from glycolysis; 4 from TCA; 2 each from pentose phosphate, Entner-Doudoroff, energy species reactions, NAD species reactions, and small molecule transport; 1 from oxidative phosphorylation; and 1 pyrophosphatase.
 The system state has even more pairwise sensitivities that differ from the corresponding first-order sensitivities and stand out as significant.
@@
 In addition, a handful of alanine synthesis pairwise sensitivities are high.
-However, there are enough reactions that, when paired with alanine synthesis, have little effect on the model.
-Malic enzyme is one of these, as it produces the pyruvate that alanine synthesis consumes.
+However, there are enough reactions that, when paired with alanine synthesis, have little effect on the model; malic enzyme is one of these, as it produces the pyruvate that alanine synthesis consumes.
 Thus, the total-order alanine synthesis sensitivity is low, placing it at the very bottom of section I.
@@
 Interestingly, the reactions in section F and several in section G showed most of thier highest sensitivities when paired with the "non-effects" of section J.
-Of these, three involved pyruvate, strengthening its role as a key metabolite; the others were glucose uptake via GTP and CTP, fumarate reductase, and SO4 transport.
+Of these, three involved pyruvate, strengthening its role as a key metabolite; the others were glucose uptake via GTP and CTP, fumarate reductase, and SO$_4$ transport.
 This suggests that these reactions' effects on the model were canceled out or lessened by most other reactions, but were of course not affected by the reactions in section J.
@@
 To address this question, we used ssFBA in combination with the cell-free metabolic network and a detailed promoter model under a T7 polymerase to compute the maximum theoretical carbon yield.
-However, we first validated the ssFBA approach by comparing an ensemble of simulated versus measured concentrations of CAT over a one hour period (Fig. ~\ref{fig:CATProdYield}A). 
+We first validated the ssFBA approach by comparing an ensemble of simulated versus measured concentrations of CAT over a one hour period (Fig.~\ref{fig:CATProdYield}A). 
 The ensemble of 100 sets captured the CAT concentration profile which was generated by sampling RNA polymerase levels, ribosome levels and elongation rates in a physiological range.
-We then used sequence-specific FBA to calculate a theoretical maximum CAT yield under three different cases: unconstrained, constrained by transcription/translation rates, and constrained by transcription/translation rates and measurements (Fig.
-~\ref{fig:CATProdYield}B).
+We then used sequence-specific FBA to calculate a theoretical maximum CAT yield under three different cases: unconstrained, constrained by transcription/translation rates, and constrained by transcription/translation rates and measurements (Fig.~\ref{fig:CATProdYield}B).
 The theoretical maximum carbon yield of CAT was 0.35 $\pm$ 0.006 for an unconstrained case and 0.225 $\pm$ 0.03 for the transcription and translation constrained case.
-Thus, we showed that our experimental dataset and best-fit parameter set were each producing CAT at 25\% of the theoretical maximum and 38\% of a theoretical physiological case.
-Whereas, the case constrained by experimental data had a carbon yield of 0.062 $\pm$ 0.02, similar to the experimental yield. 
-In comparing the flux distributions between the unconstrained and constrained cases (Fig.
-~\ref{fig:Network}), the constrained cases heavily utilize the first step in the pentose phosphate pathway to generate NADPH.
-The majority of the flux continues through the Entner–Doudoroff pathway whereas in the unconstrained case, the majority of flux travels through glycolysis.
-In all cases, the energy source comes from oxidative phosphorylation with some from the citric acid cycle.
-Whereas, in just the TX/TL case, there is a high flux through fumerate dehydrogenase from aspartic acid uptake. 
-In the unconstrained and most constrained case, we see a mixture of acetate and lactate accumulation.
-This shows the system is producing NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to supply enough NADH for oxidative phosphorylation.
-As a result, high oxidative phosphorylation activity relative to our cell free system leads to an acetate overflow.
+Thus, the experimental dataset and best-fit parameter set each produced CAT at 25\% of the theoretical maximum and 38\% of a theoretical physiological case.
+However, the case constrained by experimental data showed a carbon yield of 0.062 $\pm$ 0.02, similar to the experimental yield. 
+In comparing the flux distributions between the unconstrained and constrained cases (Fig.~\ref{fig:Network}), the constrained cases heavily utilized the first step in the pentose phosphate pathway to generate NADPH.
+In these cases the majority of the flux continued through the Entner–Doudoroff pathway, whereas in the unconstrained case the majority of flux traveled through glycolysis.
+In all cases, the energy source came primarily from oxidative phosphorylation, as well as partly from the TCA cycle.
+In the TX/TL constrained case, there was a high flux through fumerate dehydrogenase from aspartic acid uptake, whereas in the unconstrained and most constrained cases, acetate and lactate accumulation occurred.
+This shows that the system is producing NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to supply enough NADH for oxidative phosphorylation.
+As a result, high oxidative phosphorylation activity relative to our cell-free system leads to an acetate overflow.
 This suggests that there is potential for increasing CAT production by reducing this diversion of carbon.
 To simulate potential knockouts, we constrained the specific glucose and amino acid uptake rates to the same values as simulated with no knockouts.
-In an ssFBA simulation with constrained TX/TL rates, knocking out the \textit{gnd} reaction decreases flux of acetate production but increases flux through \textit{pntAB} which is responsible for regenerating NADPH.
-The simulation showed carbon was diverted towards lactate, however since CAT production is constrained by the translation rate, we expected there to be no increase in CAT production.
-The decrease in acetate production is a promising result to increase CAT yield. 
-A second simulation with a knockout of \textit{gnd} and phosphate acetyltransferase, showed carbon being diverted towards lactate and succinate, however it required a higher flux through oxidative phosphorylation and the TCA cycle to meet the energetic needs of the system.
+In an ssFBA simulation with constrained TX/TL rates, knocking out the \textit{gnd} reaction decreases flux of acetate production but increases flux through \textit{pntAB}, which is responsible for regenerating NADPH.
+The simulation showed carbon was diverted toward lactate; however, since CAT production is constrained by the translation rate, we expected no increase in CAT production.
+The decrease in acetate production is promising as a mechanism to increase CAT yield. 
+A second simulation with a knockout of \textit{gnd} and phosphate acetyltransferase showed carbon being diverted toward lactate and succinate; however, it required a higher flux through oxidative phosphorylation and the TCA cycle to meet the energetic needs of the system.
 
@@
 The theoretical maximum yield of CAT was 0.35 for an unconstrained case and 0.225 for a transcription/translation constrained case. 
-Knockouts of \textit{gnd} and phosphate acetyltransferase, show carbon can be diverted away from acetate and potentially towards CAT or other proteins of interest expressed in CFPS. 
+Knockouts of \textit{gnd} and phosphate acetyltransferase, show carbon can be diverted away from acetate and potentially toward CAT or other proteins of interest expressed in CFPS. 
 The other limitations to address in CFPS would be to enhance the transcription and translation rates since the protein of interest to be expressed is ultimately bounded by these kinetic rates.    
@@
 Underwood et al.
-has also shown the increase in ribosome levels does not significantly increase protein yields and rates, however adding elongation factors increased yields by 23\% at 30 minutes\cite{2005_underwood_biotech}.
+has also shown the increase in ribosome levels does not significantly increase protein yields and rates; however, adding elongation factors increased yields by 23\% at 30 minutes\cite{2005_underwood_biotech}.
 In addition to improving CFPS performance, Jewett et al.
@@
 Each sensitivity index was formulated as:
-\begin{equation}\label{eqn:local-sensitivity}
+\begin{equation}\label{eqn:CAT-sensitivity}
 \begin{split}
@@
 Each sensitivity index was formulated as:
-\begin{equation}\label{eqn:local-sensitivity}
+\begin{equation}\label{eqn:system-sensitivity}
 \begin{split}
@@
 \centering
+\includegraphics[width=1.00\textwidth]{./Figures/CarbonBoth_png.pdf}
+\caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species.
+Best-fit parameter set (orange line) versus experimental data (points).
+95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
+\label{fig:CarbonBoth_png}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
 \includegraphics[width=1.00\textwidth]{./Figures/Energy.pdf}
@@
 \centering
-\includegraphics[width=0.8\textwidth,trim=0 28 615 40,clip]{./Figures/Sensitivity.pdf}
+\includegraphics[width=0.76\textwidth,trim=0 28 615 40,clip]{./Figures/Sensitivity.pdf}
 \caption{Normalized first-order and pairwise sensitivities of CAT production (top) and system state (bottom) to rate constants.}

### 2016-11-14 Paper_v1.tex
Nov14_3

@@
 \urlstyle{rm}
+\hyphenation{dec-ades} %Latex does not know how to hyphenate this by default, so we add it to the list
 
@@
 In this study, we developed an ensemble of dynamic \textit{E. coli} cell-free protein synthesis (CFPS) models.
-Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
+Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids, and the protein product, chloramphenicol acetyltransferase (CAT).
 The ensemble described all of the training data, especially the central carbon metabolism.
-To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux balance analysis.
-The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
+To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence-specific flux balance analysis.
+The observed CAT yield was 25\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
 %The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
@@
 This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
-Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
+Second, cell-free systems allow us to study biological processes without the complications associated with cell growth.
 Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \citep{Jewett:2008aa}.
@@
 Mathematical modeling has long contributed to our understanding of metabolism.
-Decades before the genomics revolution, mechanistically, structured metabolic models arose from the desire to predict microbial phenotypes resulting from changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}.
+Decades before the genomics revolution, mechanistically structured metabolic models arose from the desire to predict microbial phenotypes resulting from changes in intracellular or extracellular states \citep{1976_fredrickson_BiotechBioeng}.
 The single cell \textit{E.
@@
 In this study, we developed an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models using the hybrid cell-free modeling approach of Wayman et al [REFHERE].
-Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
-The ensemble described the training data, with the exception of some of the amino acid dynamics.
-To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
-balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
+Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids, and the protein product, chloramphenicol acetyltransferase (CAT).
+The ensemble and the best-fit parameter set correctly described the training data.
+To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield with the maximum theoretical CAT carbon yield calculated using unconstrained sequence-specific flux balance analysis.
+The CAT yield of the model was 25\% of the maximum theoretical yield, suggesting that CAT production could be further optimized.
+We also calculated CAT carbon yield for a case with transcription and translation constrained by literature; the CAT yield of the model was 38\% of this theoretical physiological case.
 The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
-The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
-while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
-Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
-GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
-On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
-Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
-We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
+The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle, while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
+Sensitivity analysis of the dynamic model suggested that CAT production was most sensitive to CAT synthesis itself, as well as certain reactions in amino acid synthesis/degradation, glycolysis, and the TCA cycle, and to a lesser extent the pentose phosphate pathway and oxidative phosphorylation.
+Sensitivity analysis also showed that the system as a whole was most sensitive to these same parts of the network.
+%: mostly amino acid synthesis/degradation, glycolysis, and the TCA cycle, and to a lesser extent the pentose phosphate and Entner-Doudoroff pathways and the reactions involving the energy species and cofactors.
+CAT production was also robust to allosteric control, but other metabolites, specifically organic acid intermediates, were not.
+Taken together, we present the first dynamic model of \textit{E. coli} cell-free protein synthesis.
+We have integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
 This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
@@
 We generated an ensemble of N = 100 diverse parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model, and choosing sets so as to minimize the Pearson correlation coefficient between sets in the ensemble.
-We also defined the single set with lowest cost function as our best-fit set.
+We also defined the set with the lowest cost function as our best-fit set.
 
@@
 The central carbon metabolism (Fig.~\ref{fig:CarbonBoth}, top), energy species (Fig.~\ref{fig:Energy}), and amino acids (Fig.~\ref{fig:Amino}) were all well captured by the ensemble and the best-fit set.
-Allosteric control was seen to be particularly important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).
+Allosteric control was seen to be particularly important to the dynamics of the organic acid intermediates, as without it several of the measurements were not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).
 
@@
 This suggests that these reactions' effects on the model were canceled out or lessened by most other reactions, but were of course not affected by the reactions in section J.
+This is also likely the reason that reactions in section J rank above those in section K, despite having no effect themselves on the model.
 
@@
 However, the case constrained by experimental data showed a carbon yield of 0.062 $\pm$ 0.02, similar to the experimental yield.
+
 In comparing the flux distributions between the unconstrained and constrained cases (Fig.~\ref{fig:Network}), the constrained cases heavily utilized the first step in the pentose phosphate pathway to generate NADPH.
@@
 
-In this study, we developed an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models using the hybrid cell-free modeling approach of Wayman et al [REFHERE].
-Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
-The ensemble described the training data, with the exception of some of the amino acid dynamics.
-To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield with the maximum theoretical CAT carbon yield calculated using sequence specific flux balance analysis.
-The observed CAT yield was 25\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
-The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
-The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle, while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
-Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
-On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
-% Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
-% We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
-% This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
+In this study we present an ensemble of \textit{E.
+coli} cell-free protein synthesis (CFPS) models that accurately predict a comprehensive CFPS dataset of glucose, CAT, central carbon metabolites, energy species, and amino acid measurements.
+We used the hybrid cell-free modeling approach of Wayman et al [REFHERE], which integrates traditional kinetics and a logic-based allosteric description.
+We also show that the model produces CAT at 25\% of the theoretical maximum in terms of carbon yield, and at 38\% of a physiological case in which transcription and translation are constrained.
+The theoretical maximum and TX/TL-constrained case were obtained using FBA, which predicted a different flux distribution than the ensemble of dynamic models.
+The ensemble of dynamic models predicted most of the carbon flux going through glycolysis and the TCA cycle, while FBA predicted significant flux through the Entner-Doudoroff pathway.
+Sensitivity analysis of the dynamic model suggested that both CAT production and the entire metabolism were most sensitive to amino acid synthesis and degradation reactions, and reactions in glycolysis and the TCA cycle.
+CAT production was also very sensitive to the CAT synthesis reaction, unsurprisingly.
+The allosteric control component of the hybrid modeling approach was shown as important to central carbon metabolism, but not very important to CAT production.
+Taken together, this is the first dynamic model of \textit{E. coli} cell-free protein synthesis, and an important step toward a functional genome scale description.
 
-In comparing the theoretical maximum carbon yield of CAT from ssFBA predictions to the kinetic model and experimental measurements, suggests that there is potential for increasing CAT yield in CFPS as well as CFPS performance.
+In comparing the theoretical maximum carbon yield of CAT from ssFBA predictions to the kinetic model and experimental measurements, this suggests that there is potential for increasing CAT yield in CFPS as well as CFPS performance.
 The theoretical maximum yield of CAT was 0.35 for an unconstrained case and 0.225 for a transcription/translation constrained case. 
-Knockouts of \textit{gnd} and phosphate acetyltransferase, show carbon can be diverted away from acetate and potentially toward CAT or other proteins of interest expressed in CFPS. 
-The other limitations to address in CFPS would be to enhance the transcription and translation rates since the protein of interest to be expressed is ultimately bounded by these kinetic rates.    
+Knockouts of \textit{gnd} and phosphate acetyltransferase show carbon can be diverted away from acetate and potentially toward CAT or other proteins of interest expressed in CFPS. 
+Another limitation to be addressed in CFPS is the transcription and translation description, since the protein of interest to be expressed is ultimately bounded by these kinetic rates.
 Li et al.
-have increased productivity of firefly lucifease by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA and tRNAs \cite{2014_li_PlosOne}.
-Underwood et al.
-has also shown the increase in ribosome levels does not significantly increase protein yields and rates; however, adding elongation factors increased yields by 23\% at 30 minutes\cite{2005_underwood_biotech}.
-In addition to improving CFPS performance, Jewett et al.
-has showed that oxidative phosphorylation operates in cell-free systems and knocking out oxidative phosphorylation reactions is detrimental to protein yield \cite{2008_jewett_molsysbiol}.
-However, it is inconclusive how much oxidative phosphorylation activity there is compared to \textit{in vivo} systems and both of our models suggest oxidative phosphorylation is vital to CAT production.
-Thus, this is a potential place for improvement to optimize CFPS for better performance and protein yield.
+have increased productivity of firefly lucifease by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
+Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes\cite{2005_underwood_biotech}.
+In addition to improving CFPS performance, Jewett and coworkers have shown that oxidative phosphorylation operates in cell-free systems, and that knocking out these reactions is detrimental to protein yield \cite{2008_jewett_molsysbiol}.
+However, it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, and both of the modeling approaches we present suggest that oxidative phosphorylation is important to CAT production.
+Thus, this is a potential area for improvement of CFPS performance and protein yield.
 
@@
 
-The cell-free model ensemble described the training data with the exception of some of the amino acids.
-Specifically, adding more reactions that consume amino acids would improve the model's ability to predict those that show a decrase in the experimental data.
-Also, including specific transcription and translation steps for CAT would allow us to more accurately model the complexity and the resource cost of protein synthesis.
+%The cell-free model ensemble described the training data with the exception of some of the amino acids.
+%Specifically, adding more reactions that consume amino acids would improve the model's ability to predict those that show a decrase in the experimental data.
+
+A logical next step for this work would be sequence-specific dynamic modeling, as the kinetic modeling approach in this study used a single reaction to approximate CAT synthesis.
+Including specific transcription and translation steps for CAT would allow us to more accurately model the complexity and the resource cost of protein synthesis.
+In addition, sensitivity analysis could be performed on these new parameters to determine the r
+
 Another area for future work is to more thoroughly sample parameter space.
@@
 \begin{equation}\label{eqn:cost-function}
-    cost=\sum_{i=1}^{\mathcal{D}}\left(w_i\sum_{j=1}^{\mathcal{T}_i}abs\left(x_{ij}^{data}-x_{i}^{sim}|_{t(j)}\right)\right)
+    cost=\sum_{i=1}^{\mathcal{D}}\left(w_i\sum_{j=1}^{\mathcal{T}_i}\bigg(\frac{x_{ij}^{data}-x_{i}^{sim}|_{t(j)}}{\max\limits_{j}\left(x_{ij}^{data}\right)}\bigg)^2 \right)
 \end{equation}
@@
 
-\subsection*{Sequence specific FBA and calculation of CAT yield}
+\subsection*{Sequence-specific FBA and calculation of CAT yield}
 The yield on CAT production was calculated for each case as a ratio of carbon produced as CAT to carbon consumed as reactants (glucose and amino acids):
@@
 The theoretical yield was calculated using flux balance anaylsis (FBA) with a sequence-specific based analysis on CAT.
-The sequence specific FBA \cite{2002_allen_palsson} problem was formulated as:
+The sequence-specific FBA \cite{2002_allen_palsson} problem was formulated as:
 \begin{equation}\nonumber
@@
 The specific glucose uptake rate was constrained to allow a maximum flux of 40 mM/hr according to experimental data; the specific amino acid uptake rates were bound to allow a maximum flux of 30 mM/hr, but did not reach this maximum flux.
-The transcription and translation template reactions were added to the metabolic network and are based off sequence specific analysis \cite{2002_allen_palsson} involving transcription initiaion, transcription, mRNA degradation, translation initiation, translation, and tRNA charging.
+The transcription and translation template reactions were added to the metabolic network and are based off sequence-specific analysis \cite{2002_allen_palsson} involving transcription initiaion, transcription, mRNA degradation, translation initiation, translation, and tRNA charging.
 The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
@@
 \includegraphics[width=0.95\textwidth]{./Figures/CAT_prod_yield.pdf}
-\caption{Sequence specific flux balance analysis of CAT production and yield.
+\caption{Sequence-specific flux balance analysis of CAT production and yield.
 A.
@@
 \caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, and oxidative phosphorylation.
-Sequence specific FBA flux value (mean ± standard deviation) across ensemble for 1 hr, normalized to glucose uptake flux.
+Sequence-specific FBA flux value (mean ± standard deviation) across ensemble for 1 hr, normalized to glucose uptake flux.
 Flux distribution for three different cases: unconstrained except for glucose uptake (top row), constrained by transcription and translation rates (second row), and constrained by transcription, translation rates and experimental measurements where available (bottom row).}

### 2016-11-15 Paper_v1.tex
Nov14_5

@@
 The ensemble described all of the training data, especially the central carbon metabolism.
-To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence-specific flux balance analysis.
-The observed CAT yield was 25\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
-%The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
-The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle, while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
-Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
-On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
-Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
-This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
+The model predicted a carbon yield for CAT production that was equal to 25\% of the maximum theoretical yield, calculated using sequence-specific flux balance analysis.
+This suggests that CAT production could be further optimized.
+While the dynamic models predicted that the majority of carbon flux went through glycolysis and the TCA cycle, the flux balance analysis showed significant flux through the Entner-Doudoroff pathway.
+The dynamic modeling approach predicted that glycolysis, the TCA cycle, and amino acid synthesis and degradation were most important to both CAT production and the system as a whole, while CAT production alone depended heavily on the CAT synthesis reaction.
+Conversely, CAT production was robust to allosteric control, as was most of the network, with the exception of the organic acids in central carbon metabolism.
+This study is the first to model dynamic protein production in \textit{E. coli}, and should provide a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
 
@@
 
-In this study, we developed an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models using the hybrid cell-free modeling approach of Wayman et al [REFHERE].
+In this study, we developed an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models using the hybrid cell-free modeling approach of Wayman and coworkers [REFHERE].
 Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids, and the protein product, chloramphenicol acetyltransferase (CAT).
@@
 \subsection*{Estimation of an ensemble of cell-free protein synthesis models.}
-We used the hybrid cell-free modeling framework of Wayman et al. to simulate the production of a model protein [REFHERE].
+We used the hybrid cell-free modeling framework of Wayman and coworkers to simulate the production of a model protein [REFHERE].
 The cell-free \textit{E.
@@
 coli} cell-free protein synthesis (CFPS) models that accurately predict a comprehensive CFPS dataset of glucose, CAT, central carbon metabolites, energy species, and amino acid measurements.
-We used the hybrid cell-free modeling approach of Wayman et al [REFHERE], which integrates traditional kinetics and a logic-based allosteric description.
+We used the hybrid cell-free modeling approach of Wayman and coworkers [REFHERE], which integrates traditional kinetics and a logic-based allosteric description.
 We also show that the model produces CAT at 25\% of the theoretical maximum in terms of carbon yield, and at 38\% of a physiological case in which transcription and translation are constrained.
-The theoretical maximum and TX/TL-constrained case were obtained using FBA, which predicted a different flux distribution than the ensemble of dynamic models.
+The theoretical maximum and TX/TL constrained case were obtained using FBA, which predicted a different flux distribution than the ensemble of dynamic models.
 The ensemble of dynamic models predicted most of the carbon flux going through glycolysis and the TCA cycle, while FBA predicted significant flux through the Entner-Doudoroff pathway.
-Sensitivity analysis of the dynamic model suggested that both CAT production and the entire metabolism were most sensitive to amino acid synthesis and degradation reactions, and reactions in glycolysis and the TCA cycle.
+Sensitivity analysis of the dynamic model suggested that both CAT production and the entire metabolic network were most sensitive to amino acid synthesis and degradation reactions, and reactions in glycolysis and the TCA cycle.
 CAT production was also very sensitive to the CAT synthesis reaction, unsurprisingly.
@@
 have increased productivity of firefly lucifease by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
-Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes\cite{2005_underwood_biotech}.
+Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
 In addition to improving CFPS performance, Jewett and coworkers have shown that oxidative phosphorylation operates in cell-free systems, and that knocking out these reactions is detrimental to protein yield \cite{2008_jewett_molsysbiol}.
@@
 
-%The cell-free model ensemble described the training data with the exception of some of the amino acids.
-%Specifically, adding more reactions that consume amino acids would improve the model's ability to predict those that show a decrase in the experimental data.
-
 A logical next step for this work would be sequence-specific dynamic modeling, as the kinetic modeling approach in this study used a single reaction to approximate CAT synthesis.
-Including specific transcription and translation steps for CAT would allow us to more accurately model the complexity and the resource cost of protein synthesis.
-In addition, sensitivity analysis could be performed on these new parameters to determine the r
-
+Including specific transcription and translation steps for CAT would allow more accurate modeling of the complexity and the resource cost of protein synthesis.
+In addition, sensitivity analysis could be performed on these new parameters to determine the robustness of CAT synthesis to the processes of transcription and translation.
 Another area for future work is to more thoroughly sample parameter space.
-For the metabolites in the dataset, initial conditions were fixed at the initial data values.
-All other parameters were varied in a manner so as to best fit the dataset.
-However, the resulting ensemble may not represent every biological or practical possibility.
-In a different region of parameter space, the system could behave differently, including the flux distribution through the network, the accuracy and spread of ensemble fits, the relative sensitivities, and the yield as a percentage of the theoretical maximum.
+Parameters were varied so as to best fit the dataset; however, the resulting ensemble may not represent every biological possibility.
+In a different region of parameter space, the system may behave differently but still fit the experimental data.
+This could include the flux distribution through the network, the variation of predictions across the ensemble, and the relative sensitivity values.
 Testing the model under a variety of conditions could strengthen or challenge the findings of this study.
@@
 where $S_{ij}$ denotes the sensitivity of CAT production to the $i$th and $j$th parameters, $CAT(p_i,p_j,t)$ denotes CAT concentration as a function of time and the $i$th and $j$th parameters, $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
-In calculating the pairwise sensitivities, each parameter was perturbed by 1\%; first-order sensitivities ($i$ = $j$) were subject to two 1\% perturbations, equivalent to a perturbation of 2.01\%.
+In calculating the pairwise sensitivities, each parameter was perturbed by 1\%; first-order sensitivities ($i$ = $j$) were subject to two 1\% perturbations, equivalent to a perturbation of 2.01\%.. 
 The array of sensitivity indices was organized into a clustergram (Fig.~\ref{fig:Sensitivity}, top) with both axes corresponding to the 163 rate constants rearranged from most sensitive (top, right) to least sensitive (bottom, left).
@@
 where $\Delta CAT$ denotes the amount of CAT produced, $C_{CAT}$ denotes carbon number of CAT, $\mathcal{R}$ denotes the number of reactants, $\Delta m_{i}$ denotes the amount of the $i$th reactant consumed, never allowed to be negative, and $C_{m_i}$ denotes the carbon number of the $i$th reactant.
-Because no data was available for arginine or glutamate, these reactants were left out of all three calculations.
-In the experimental case and the best-fit set case, yield was calculated by setting $\Delta CAT$ equal to the final minus the initial CAT concentration and setting $\Delta m_{i}$ equal to the initial minus the final reactant concentration.
-The theoretical yield was calculated using flux balance anaylsis (FBA) with a sequence-specific based analysis on CAT.
-The sequence-specific FBA \cite{2002_allen_palsson} problem was formulated as:
+Because no data was available for arginine or glutamate, these reactants were left out of all yield calculations.
+Yield of the best-fit parameter set and the experimental data were calculated by setting $\Delta CAT$ equal to the final minus the initial CAT concentration and setting $\Delta m_{i}$ equal to the initial minus the final reactant concentration.
+CAT carbon yields for three cases discussed below were calculated using flux balance anaylsis (FBA) with a sequence-specific description of CAT synthesis.
+This sequence-specific FBA \cite{2002_allen_palsson} problem was formulated as:
 \begin{equation}\nonumber
@@
 The specific glucose uptake rate was constrained to allow a maximum flux of 40 mM/hr according to experimental data; the specific amino acid uptake rates were bound to allow a maximum flux of 30 mM/hr, but did not reach this maximum flux.
-The transcription and translation template reactions were added to the metabolic network and are based off sequence-specific analysis \cite{2002_allen_palsson} involving transcription initiaion, transcription, mRNA degradation, translation initiation, translation, and tRNA charging.
+The transcription and translation template reactions come from sequence-specific analysis \cite{2002_allen_palsson}, and include transcription initiation, transcription, mRNA degradation, translation initiation, translation, and tRNA charging.
 The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
-The solution flux vector was used to calculate the theoretical carbon yield of CAT.
-Glucose, oxygen, and amino acids were modeled as being imported into the system, whereas CAT synthesis and metabolite byproduct formation was modeled as an export from the system.
-The rest of the network followed a pseudo steady-state asusmption where all other metabolites were not allowed to accumulate; thus, the network could be solved by linear programming.
+The solution flux vector was used to calculate the carbon yield of CAT for the three FBA cases.
+Glucose, oxygen, and amino acids were modeled as being imported into the system, while CAT synthesis and metabolite byproduct formation were modeled as export from the system.
+The rest of the network followed a pseudo steady-state assumption where metabolites were not allowed to accumulate; thus, the network could be solved by linear programming instead of solving differential equations.
 
-The transcription rate was constrained as:
+The transcription rate was formulated as:
 \begin{equation}\nonumber
-	w_{tx} = RNAP\frac{v_{RNAP}}{l_{mRNA}}\left(\frac{Gene}{km+Gene}\right)P
+	w_{TX} = RNAP\left(\frac{v_{RNAP}}{l_{mRNA}}\right)\left(\frac{Gene}{k_m+Gene}\right)P
 \end{equation}
-where $RNAP$ is the concentration of RNA polymerase, $v_{RNAP}$ is the elongation rate (nucleotides/hr) by the RNA polymerase, $l_{mRNA}$ is the number of nucleotides in the mRNA, $Gene$ is the gene concentration, $km$ is the plasmid saturation coefficient, and $P$ is the promoter activity.
+where $w_{TX}$ denotes the transcription rate, $RNAP$ denotes the concentration of RNA polymerase, $v_{RNAP}$ denotes the elongation rate by the RNA polymerase in nucleotides per hour, $l_{mRNA}$ denotes the mRNA length in nucleotides, $Gene$ denotes the gene concentration, $k_m$ denotes the plasmid saturation coefficient, and $P$ denotes the promoter activity.
 The mRNA and protein sequence of CAT was determined from literature.
-
 The promoter activity was formulated following Moon et al. for synthetic circuits as:
@@
 \end{equation}
-where $K_{1}$ represents the state of T7 RNA polymerase binding.
+where $K_{1}$ denotes the state of T7 RNA polymerase binding. 
+The translation rate was formulated as:
+ \begin{equation}\nonumber
+	w_{TL} = K_{P}\ Ribo\left(\frac{v_{Ribo}}{l_{Protein}}\right)mRNA_{SS}
+\end{equation}
+where $K_{P}$ denotes the polysome amplification constant, $Ribo$ denotes the ribosome concentration, $v_{Ribo}$ denotes the elongation rate of the ribosome in amino acids per hour, $l_{Protein}$ denotes the number of amino acids in the protein of interest, and $mRNA_{SS}$ denotes the mRNA concentration at steady state, equal to the transcription rate divided by the degradation rate of mRNA.
 
-The translation rate was constrained as:
- \begin{equation}\nonumber
-	w_{tl} = K_{P}Ribo\frac{v_{Ribo}}{l_{protein}}[mRNA_{ss}]
-\end{equation}
-where $K_{P}$ is the polysome amplification constant, $Ribo$ is the ribosome concentration,  $v_{Ribo}$ is the elongation rate (amino acids/hr) of the ribosome, $l_{protein}$ is the number of amino acids in the protein of interest, and $mRNA_{ss}$ is the mRNA concentration at steady state determined by the transcription rate divided by the degradation rate of mRNA.
+An ensemble of 100 sets of flux distributions was calculated for three different cases: unconstrained, constrained by transcription and translation (TX/TL) rates, and  constrained by TX/TL rates and experimental data.
+For the unconstrained case, all rates were left unbounded, except the specific glucose uptake rate. 
+An ensemble of flux distributions was then calculated by randomly sampling the maximum specific glucose uptake rate from within a range of 30 to 40 mM/hr, determined from experimental data.
+For the case constrained by TX/TL rates, an ensemble was generated by randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
+RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/sec \cite{2005_underwood_biotech, 2016_garamella_acssynbio}.
+For the case constrained by TX/TL rates and experimental data, the lower and upper bounds on the fluxes for the data-informed metabolites were sampled within the range given by the experimental noise.
+This included the data for glucose, organic acids, energy species, and amino acids; CAT was not constrained by experimental data, but by the TX/TL rates as stated above.
 
-%\begin{equation}\label{eqn:yield}
-%	Yield=\frac{w_{CAT}\;C_{CAT}}{\sum_{i=1}^{\mathcal{R}}\max(w_{i},0)\;C_{i}}
-%\end{equation}
-%where $w_{CAT}$ denotes the CAT export flux and $w_{i}$ denotes the import flux of the $i$th substrate.
-An ensemble of 100 sets of flux distributions was calculated for three different cases, unconstrained, constrained by transcription and translation (TX/TL) rates, and  constrained by TX/TL rates and experimental data.
-For the unconstrained case, all rates were left unbounded, except for the specific glucose uptake rate. 
-An ensemble of flux distributions was calculated by randonmly sampling the maximum specific glucose uptake rate from 30 to 40 mM/hr determined from experimental data.
-For the case constrained by TX/TL rates, an ensemble was generated by randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
-RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the elongation rate by RNA polymerase between 20 and 30 nts/sec, and the elongation rate by ribosomes between 1.5 and 3 AA/sec \cite{2005_underwood_biotech, 2016_garamella_acssynbio}.
-For the case constrained by TX/TL rates and experimental data, an ensemble was generated by randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
-The lower and upper bound constraints were randomly sampled in the physiological range of the experimental noise where data was available, except for CAT flux, which was determined from the transcription and translation rates.
 \clearpage
 
-%Need to talk more about biochemical benefits and importance for of biochemical problems
+%Need to talk more about biochemical benefits and importance for biochemical problems
 

### 2016-11-15 Paper_v1.tex
Some JV changes

@@
 Conversely, CAT production was robust to allosteric control, as was most of the network, with the exception of the organic acids in central carbon metabolism.
-This study is the first to model dynamic protein production in \textit{E. coli}, and should provide a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
+This study is the first to model dynamic protein production in \textit{E. coli}, and should provide a foundation for genome-scale,
+dynamic modeling of cell-free \textit{E. coli} protein synthesis.
 
@@
 
+% Interestingly, many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system.
+% For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth.
+% Thus, cell-free operation holds several significant advantages for model development, identification and validation.
+Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms,
+% such as \textit{E. coli} or \textit{B.~subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
+% Conversely, highly abstracted kinetic frameworks, such as the cybernetic framework, represented a paradigm shift, viewing cells as growth-optimizing strategists \citep{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}.
+% Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}.
+% Unfortunately, traditional, fully structured cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously.
+% However, recent cybernetic formulations from Ramkrishna and colleagues have successfully treated this identifiability challenge through elementary mode reduction~\cite{2009_song_ramkrishna_BiotechBioeng,Song:2011aa}.
+
 \section*{Introduction}
@@
 Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
-Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall.
-This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
-Second, cell-free systems allow us to study biological processes without the complications associated with cell growth.
+Central amongst these, is direct access to metabolites and the biosynthetic machinery without the interference of a cell wall, or complications associated with cell growth.
+This allows us to interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
 Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \citep{Jewett:2008aa}.
-However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}.
-Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}.
-Interestingly, many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system.
-For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth.
-Thus, cell-free operation holds several significant advantages for model development, identification and validation.
-Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms,
-such as \textit{E. coli} or \textit{B.~subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
+However, CFPS is not new; CFPS in crude \textit{E.~coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}.
+Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa,Pardee:2016aa}.
+However, if CFPS is to become a mainstream technology for applications such as point of care manufacturing, we must first understand the performance limits of these systems.
+One tool to address this question is mathematical modeling.
 
@@
 coli} \citep{2008_atlas_shuler_IETSysBio}.
-Conversely, highly abstracted kinetic frameworks, such as the cybernetic framework, represented a paradigm shift, viewing cells as growth-optimizing strategists \citep{1985_dhurjati_ramkrishna_tsao_BiotechBioeng}.
-Cybernetic models have been highly successful at predicting metabolic choice behavior, e.g., diauxie behavior \citep{1986_kompala_ramkrishna_tsao_BiotechBioeng}, steady-state multiplicity \citep{2012_kim_ramkrishna_BiotechProg}, as well as the cellular response to metabolic engineering modifications \citep{1999_varner_ramkrishna_MetaEng}.
-Unfortunately, traditional, fully structured cybernetic models also suffer from an identifiability challenge, as both the kinetic parameters and an abstracted model of cellular objectives must be estimated simultaneously.
-However, recent cybernetic formulations from Ramkrishna and colleagues have successfully treated this identifiability challenge through elementary mode reduction~\cite{2009_song_ramkrishna_BiotechBioeng,Song:2011aa}.
-
-In the post genomics world, large-scale stoichiometric reconstructions of microbial metabolism popularized by static, constraint-based modeling techniques such as flux balance analysis (FBA) have become standard tools \citep{2012_lewis_palsson_NatRevMicrobio}.
+In the post genomics world, large-scale stoichiometric reconstructions of microbial metabolism popularized by techniques such as flux balance analysis (FBA) have become a standard approach \citep{2012_lewis_palsson_NatRevMicrobio}.
 Since the first genome-scale stoichiometric model of \textit{E.
-coli}, developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, well over 100 organisms, including industrially important prokaryotes such as \textit{E.
-coli} \citep{Feist:2007aa} or \textit{B.
-subtilis} \citep{Oh:2007aa}, are now available \citep{2009_feist_palsson_NatRevMicrobio}.
+coli}, developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, well over 100 organisms, including industrially important prokaryotes are now available \citep{2009_feist_palsson_NatRevMicrobio,Feist:2007aa,Oh:2007aa}.
 Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome-scale kinetic models to an underdetermined linear algebraic system, which can be solved efficiently even for large systems.
@@
 
-In this study, we developed an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models using the hybrid cell-free modeling approach of Wayman and coworkers [REFHERE].
+In this study, we developed an ensemble of kinetic cell-free protein synthesis (CFPS) models using dynamic metabolite measurements in an \textit{E. coli} cell free extract.
 Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids, and the protein product, chloramphenicol acetyltransferase (CAT).
-The ensemble and the best-fit parameter set correctly described the training data.
-To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield with the maximum theoretical CAT carbon yield calculated using unconstrained sequence-specific flux balance analysis.
-The CAT yield of the model was 25\% of the maximum theoretical yield, suggesting that CAT production could be further optimized.
-We also calculated CAT carbon yield for a case with transcription and translation constrained by literature; the CAT yield of the model was 38\% of this theoretical physiological case.
+Characteristic values for model parameters and initial conditions, estimated from literature, were used to constrain the parameter estimation problem.
+The ensemble of parameter sets described the training data with a median cost that was greater than two orders of magnitude smaller than random sets constructed using the literature parameter constraints.
+We then used the ensemble of kinetic models to analyze the CFPS reaction.
+First, sensitivity analysis of the dynamic model suggested that CAT production was most sensitive to CAT synthesis parameters, as well as reactions in amino acid synthesis/degradation, glycolysis, and the TCA cycle, and to a lesser extent the pentose phosphate pathway and oxidative phosphorylation.
+Sensitivity analysis also showed that the system as a whole was most sensitive to these same parts of the network.
+CAT production was also robust to allosteric control, but other metabolites, specifically organic acid intermediates such as pyruvate, were not.
+Next, to gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield with the maximum theoretical CAT carbon yield calculated using sequence-specific flux balance analysis.
+The CAT yield estimated from the kinetic model was 25\% of the maximum theoretical yield, but 38\% of the theoretical yield when physiologically realistic constraints were used.
 The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
 The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle, while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
-Sensitivity analysis of the dynamic model suggested that CAT production was most sensitive to CAT synthesis itself, as well as certain reactions in amino acid synthesis/degradation, glycolysis, and the TCA cycle, and to a lesser extent the pentose phosphate pathway and oxidative phosphorylation.
-Sensitivity analysis also showed that the system as a whole was most sensitive to these same parts of the network.
-%: mostly amino acid synthesis/degradation, glycolysis, and the TCA cycle, and to a lesser extent the pentose phosphate and Entner-Doudoroff pathways and the reactions involving the energy species and cofactors.
-CAT production was also robust to allosteric control, but other metabolites, specifically organic acid intermediates, were not.
-Taken together, we present the first dynamic model of \textit{E. coli} cell-free protein synthesis.
-We have integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
+Taken together, we have integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
 This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
@@
 
-\subsection*{Estimation of an ensemble of cell-free protein synthesis models.}
-We used the hybrid cell-free modeling framework of Wayman and coworkers to simulate the production of a model protein [REFHERE].
+The ensemble of kinetic CFPS models captured the time evolution of CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
+The cell-free \textit{E.~coli} metabolic network was constructed by removing growth associated reactions from the iAF1260 reconstruction \cite{Feist:2007aa},
+and by adding reactions describing chloramphenicol acetyltransferase (CAT) biosynthesis, a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
+The CFPS model equations were formulated using the hybrid cell-free modeling framework of Wayman et al. \cite{pr3010138}.
+An ensemble of probable model parameters (N $>$ 10,000) was estimated from the time-series metabolite measurements using a constrained Markov Chain Monte Carlo (MCMC) approach;
+the MCMC algorithm minimized the error between the training data and model simulations starting from an initial parameter estimate assembled from literature and inspection.
+We selected parameter sets for the ensemble based upon their error value and the Pearson correlation coefficient between sets in the ensemble.
+The parameter set with the lowest error value was defined as the best-fit set.
+The central carbon metabolism (Fig.~\ref{fig:CarbonBoth}, top), energy species (Fig.~\ref{fig:Energy}), and amino acids (Fig.~\ref{fig:Amino}) were all captured by the ensemble and the best-fit set.
+Allosteric control was important to the dynamics of the organic acid intermediates, as without it several of the measurements were not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).
+However,
+
+We used the hybrid cell-free modeling framework of Wayman and coworkers to simulate the production of CAT \citep{pr3010138}.
 The cell-free \textit{E.
-coli} metabolic model was constructed by removing the growth-associated processes from the model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS}, and by adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
+coli} metabolic network was constructed by removing the growth-associated processes from the model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS}, and by adding reactions for the synthesis of
 Thus, the model described core central carbon metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle), as well as the synthesis of energy species, amino acids biosynthesis and degradation, and biosynthesis of the CAT protein.
-An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids.
-We generated an ensemble of N = 100 diverse parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model, and choosing sets so as to minimize the Pearson correlation coefficient between sets in the ensemble.
-We also defined the set with the lowest cost function as our best-fit set.
-
-The ensemble of models captured the time evolution of cell-free CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
-The central carbon metabolism (Fig.~\ref{fig:CarbonBoth}, top), energy species (Fig.~\ref{fig:Energy}), and amino acids (Fig.~\ref{fig:Amino}) were all well captured by the ensemble and the best-fit set.
-Allosteric control was seen to be particularly important to the dynamics of the organic acid intermediates, as without it several of the measurements were not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).
+An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids by minimizing the difference between the experimental dataset and metabolite simulations subject to literature constraints on parameter values and initial conditions.
 
@@
 coli} cell-free protein synthesis (CFPS) models that accurately predict a comprehensive CFPS dataset of glucose, CAT, central carbon metabolites, energy species, and amino acid measurements.
-We used the hybrid cell-free modeling approach of Wayman and coworkers [REFHERE], which integrates traditional kinetics and a logic-based allosteric description.
-We also show that the model produces CAT at 25\% of the theoretical maximum in terms of carbon yield, and at 38\% of a physiological case in which transcription and translation are constrained.
+We used the hybrid cell-free modeling approach of Wayman and coworkers, which integrates traditional kinetic modeling with a logic-based description of allosteric regulation.
+We showed that the model produces CAT at 25\% of the theoretical maximum in terms of carbon yield, and at 38\% of a physiological case in which transcription and translation are constrained.
 The theoretical maximum and TX/TL constrained case were obtained using FBA, which predicted a different flux distribution than the ensemble of dynamic models.
@@
 The reaction rate was written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$), $r_{j}\left(\mathbf{x},\mathbf{k}\right)=\bar{r}_{j}v_{j}$.
-In this study, we used either saturation or mass action kinetics.
-The control term $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
+We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
+\begin{equation}\label{eqn:rate-bar}
+	\bar{r}_{j}=V_{j}^{max}\epsilon_{i}\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}
+\end{equation}
+where $V_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$,
+$K_{js}$ denotes the saturation constant for species $s$ in reaction $j$ and $m_{j}^{-}$ denotes the set of \textit{reactants} for reaction $j$.
+On the other hand, the control term $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
 For each rate, we used a rule-based approach to select from competing control factors.
@@
 $v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
-where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
+where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a transfer function quantifying the influence of factor $i$ on rate $j$.
 The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
-variable. Each regulatory transfer function took the form:
-\begin{equation}\label{eqn:control-factor}
-	f_{ij}\left(\mathcal{Z}_{i},k_{ij},\eta_{ij}\right)=k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}/\left({1 + k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}}\right)
-\end{equation}where $\mathcal{Z}_{i}$ denotes the abundance factor $i$, $k_{ij}$ denotes a gain parameter, and $\eta_{ij}$ denotes a cooperativity parameter.
-In this study, we used $\mathcal{I}_{j}\in\left\{mean\right\}$ \cite{pr3010138}. If a process has no modifying factors, $v_{j}=1$.
-We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
-\begin{equation}\label{eqn:rate-bar}
-	\bar{r}_{j}=k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
-\end{equation}
-where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
-$K_{js}$ denotes the saturation constant for species $s$ in reaction $j$.
-The product in Equation~\eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$).
+variable. We used hill-like transfer functions and $\mathcal{I}_{j}\in\left\{min,max\right\}$ in this study \cite{pr3010138}.
 
-We added regulation to the network as informed by literature, for a total of 17 interactions.
+% The reaction rate was written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$), $r_{j}\left(\mathbf{x},\mathbf{k}\right)=\bar{r}_{j}v_{j}$.
+% In this study, we used either saturation or mass action kinetics.
+% The control term $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
+% For each rate, we used a rule-based approach to select from competing control factors.
+% If rate j was influenced by $1,\dots,m$ factors, we modeled this relationship as
+% $v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
+% where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
+% The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
+% variable. Each regulatory transfer function took the form:
+% \begin{equation}\label{eqn:control-factor}
+% 	f_{ij}\left(\mathcal{Z}_{i},k_{ij},\eta_{ij}\right)=k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}/\left({1 + k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}}\right)
+% \end{equation}where $\mathcal{Z}_{i}$ denotes the abundance factor $i$, $k_{ij}$ denotes a gain parameter, and $\eta_{ij}$ denotes a cooperativity parameter.
+% In this study, we used $\mathcal{I}_{j}\in\left\{mean\right\}$ \cite{pr3010138}. If a process has no modifying factors, $v_{j}=1$.
+% We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
+% \begin{equation}\label{eqn:rate-bar}
+% 	\bar{r}_{j}=k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
+% \end{equation}
+% where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
+% $K_{js}$ denotes the saturation constant for species $s$ in reaction $j$.
+% The product in Equation~\eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$).
+
+We included 17 allosteric regulation terms, taken from literature, to the CFPS model.
 PEP was modeled as an inhibitor for phosphofructokinase \cite{2010_kotte_MolSystBiol,2011_cabrera_JBiolChem}, PEP carboxykinase \cite{2010_kotte_MolSystBiol}, PEP synthetase \cite{2010_kotte_MolSystBiol,1973_chulavatnatol_JBiolChem}, isocitrate dehydrogenase \cite{2010_kotte_MolSystBiol,2007_ogawa_JBacteriol}, and isocitrate lyase/malate synthase \cite{2010_kotte_MolSystBiol,2007_ogawa_JBacteriol,1988_mackintosh_BiochemJ}, and as an activator for fructose-biphosphatase \cite{2010_kotte_MolSystBiol,2000_donahue_JBacteriol,2006_hines_JBiolChem,2007_hines_JBiolChem}.
@@
 
-\subsection*{Generation of model ensemble}
-We generated an ensemble of 100 diverse parameter sets via a Markov chain Monte Carlo random walk.
-Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-squared-error between experimental data and model predictions:
+\subsection*{Estimation of kinetic model parameters.}
+We generated an ensemble of diverse parameter sets using a constrained Markov Chain Monte Carlo (MCMC) random walk strategy.
+Starting from a single best fit parameter set estimated by inspection and literature,
+we calculated the cost function, equal to the sum-squared-error between experimental data and model predictions:
 \begin{equation}\label{eqn:cost-function}
-    cost=\sum_{i=1}^{\mathcal{D}}\left(w_i\sum_{j=1}^{\mathcal{T}_i}\bigg(\frac{x_{ij}^{data}-x_{i}^{sim}|_{t(j)}}{\max\limits_{j}\left(x_{ij}^{data}\right)}\bigg)^2 \right)
+    \texttt{cost}=\sum_{i=1}^{\mathcal{D}}\left[\frac{w_i}{\mathcal{Y}_{i}^2}\sum_{j=1}^{\mathcal{T}_i}\bigg(y_{ij}-x_{i}|_{t(j)}\bigg)^2 \right]
 \end{equation}
-where $\mathcal{D}$ denotes the number of datasets, equal to 37, $w_i$ denotes the weight of the $i$th dataset, $\mathcal{T}_i$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ denotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
-We then perturbed model parameters:
+where $\mathcal{D}$ denotes the number of datasets ($\mathcal{D}$~=~37),
+$w_i$ denotes the weight of the $i^{th}$ dataset,
+$\mathcal{T}_i$ denotes the number of timepoints in the $i^{th}$ dataset,
+$t(j)$ denotes the $j^{th}$ timepoint, $y_{ij}$ denotes the measurement value of the $i^{th}$ dataset at the $j^{th}$ timepoint,
+and $x_{i}|_{t(j)}$ denotes the simulated value of the metabolite corresponding to the $i^{th}$ dataset, interpolated to the $j^{th}$ timepoint.
+Lastly, the cost calculation was scaled by the maximum experimental value in the $i^{th}$ dataset, $\mathcal{Y}_{i}=\max_{j}\left(y_{ij}\right)$.
+We then perturbed each model parameter:
 \begin{equation}\label{eqn:parameter-perturbation}
-    k_i^{new}=k_i*exp(a\,r_i)\qquad{i=1,2,\hdots,\mathcal{P}}
+    k_i^{new}=k_i \cdot \exp(a\,r_i)\qquad{i=1,2,\hdots,\mathcal{P}}
 \end{equation}
-where $\mathcal{P}$ denotes the number of parameters, equal to 815, which includes 163 rate constants, 163 enzyme degradation rate constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i$th parameter, $k_i$ denotes the current value of the $i$th parameter, $a$ denotes a distribution variance, and $r_i$ denotes a random sample from the normal distribution.
+where $\mathcal{P}$ denotes the number of parameters ($\mathcal{P}$~=~815),
+which includes 163 rate constants, 163 enzyme activity decay constants, 455 saturation constants, and 34 control parameters,
+$k_i^{new}$ denotes the new value of the $i^{th}$ parameter,
+$k_i$ denotes the current value of the $i^{th}$ parameter,
+$a$ denotes a distribution variance, and $r_i$ denotes a random sample from the normal distribution.
 For each newly generated parameter set, we re-solved the balance equations and calculated the cost function.
 All sets with a lower cost than the previous set, and some with higher cost, were added to the ensemble.
-After generating 12,437 sets, we selected 100 sets with minimal correlation to each other so as to avoid over-sampling any region of parameter space.
-The original 12,437-set ensemble had a mean Pearson correlation coefficient [REFERENCE NEEDED?] of 0.94 between pairs of sets, while the 100-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.
+After generating 12,437 sets, we selected 100 sets with minimal set to set correlation to avoid over-sampling any region of parameter space.
+The original 12,437-set ensemble had a mean Pearson correlation coefficient of 0.94 between pairs of sets,
+while the 100-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.
 
-\subsection*{Local sensitivity analysis}
-We determined the reactions most important to protein production by computing the local sensitivity of CAT concentration to each rate constant in the network.
-Each sensitivity index was formulated as:
+\subsection*{Sensitivity analysis of the CFPS model.}
+We determined the reactions most important to protein production by computing the local sensitivity of CAT concentration (denoted as $\texttt{CAT}$)
+to each individual rate constant, and each pair of rate constants in the network.
+The sensitivity index was formulated as:
 \begin{equation}\label{eqn:CAT-sensitivity}
-\begin{split}
-    S_{ij}=norm(CAT(p_i,p_j,t)-CAT(\alpha*p_i,\alpha*p_j,t))\qquad{i,j=1,2,\hdots\mathcal{P}}\\
-\end{split}
+    \mathcal{S}_{ij}^{\texttt{CAT}} = \norm{\texttt{CAT}(p_i,p_j,t)-\texttt{CAT}(\alpha \cdot p_i,\alpha \cdot p_j,t)}_{2}\qquad{i,j=1,2,\hdots\mathcal{P}}
 \end{equation}
-where $S_{ij}$ denotes the sensitivity of CAT production to the $i$th and $j$th parameters, $CAT(p_i,p_j,t)$ denotes CAT concentration as a function of time and the $i$th and $j$th parameters, $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
-In calculating the pairwise sensitivities, each parameter was perturbed by 1\%; first-order sensitivities ($i$ = $j$) were subject to two 1\% perturbations, equivalent to a perturbation of 2.01\%.. 
-The array of sensitivity indices was organized into a clustergram (Fig.~\ref{fig:Sensitivity}, top) with both axes corresponding to the 163 rate constants rearranged from most sensitive (top, right) to least sensitive (bottom, left).
+where $\mathcal{S}_{ij}^{\texttt{CAT}}$ denotes the sensitivity of CAT production to the $i^{th}$ and $j^{th}$ parameters,
+$\texttt{CAT}(p_i,p_j,t)$ denotes CAT concentration as a function of time and the $i^{th}$ and $j^{th}$ parameters,
+$\alpha$ denotes the perturbation factor, and $\mathcal{P}$ denotes the number of rate constants ($\mathcal{P}$ = 163).
+In calculating the pairwise sensitivities, each parameter was perturbed by 1\%; first-order sensitivities ($i$ = $j$) were subject to two 1\% perturbations.
+% The array of sensitivity indices was organized into a clustergram with both axes corresponding to the 163 rate constants
+% rearranged from most sensitive (top, right) to least sensitive (bottom, left).
 
-Likewise, we determined the reactions most important to the system as a whole by computing the local sensitivity of the species for which data exist to each rate constant in the network.
-Each sensitivity index was formulated as:
+Likewise, we determined the reactions most important to global system performance by computing the sensitivity of all species for which data exists (denoted as $\texttt{X}$)
+to each rate constant in the network.
+In this case, each sensitivity index was formulated as:
 \begin{equation}\label{eqn:system-sensitivity}
-\begin{split}
-    S_{ij}=norm(X(p_i,p_j,t)-X(\alpha*p_i,\alpha*p_j,t))\qquad{i,j=1,2,\hdots\mathcal{P}}\\
-\end{split}
+    \mathcal{S}_{ij}^{\texttt{X}}=\norm{\texttt{X}(p_i,p_j,t)-\texttt{X}(\alpha*p_i,\alpha*p_j,t)}\qquad{i,j=1,2,\hdots\mathcal{P}}
 \end{equation}
-where $S_{ij}$ denotes the sensitivity of the system state to the $i$th and $j$th parameters, and $X(p_i,p_j,t)$ denotes the system state, an array consisting of the concentration of every species for which data exist as a function of time and the $i$th and $j$th parameters.
-% $\alpha$ denotes the perturbation factor, equal to 1.01, and $\mathcal{P}$ denotes the number of rate constants, equal to 163.
-This array of sensitivity indices was also organized into a clustergram (Fig.~\ref{fig:Sensitivity}, bottom).
+where $\mathcal{S}_{ij}^{\texttt{X}}$ denotes the sensitivity of the system state to the $i^{th}$ and $j^{th}$ parameters,
+and $\texttt{X}(p_i,p_j,t)$ denotes the system state, an array consisting of the concentration of every species for
+which data exists as a function of time and the $i^{th}$ and $j^{th}$ parameters.
+
+% This array of sensitivity indices was also organized into a clustergram (Fig.~\ref{fig:Sensitivity}, bottom).
 
@@
 \end{equation}
-where $\Delta CAT$ denotes the amount of CAT produced, $C_{CAT}$ denotes carbon number of CAT, $\mathcal{R}$ denotes the number of reactants, $\Delta m_{i}$ denotes the amount of the $i$th reactant consumed, never allowed to be negative, and $C_{m_i}$ denotes the carbon number of the $i$th reactant.
+where $\Delta CAT$ denotes the amount of CAT produced, $C_{CAT}$ denotes carbon number of CAT, $\mathcal{R}$ denotes the number of reactants, $\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed, never allowed to be negative, and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
 Because no data was available for arginine or glutamate, these reactants were left out of all yield calculations.
@@
 CAT carbon yields for three cases discussed below were calculated using flux balance anaylsis (FBA) with a sequence-specific description of CAT synthesis.
-This sequence-specific FBA \cite{2002_allen_palsson} problem was formulated as:
+This sequence-specific FBA \citep{Allen:2003aa} problem was formulated as:
 \begin{equation}\nonumber
@@
 The specific glucose uptake rate was constrained to allow a maximum flux of 40 mM/hr according to experimental data; the specific amino acid uptake rates were bound to allow a maximum flux of 30 mM/hr, but did not reach this maximum flux.
-The transcription and translation template reactions come from sequence-specific analysis \cite{2002_allen_palsson}, and include transcription initiation, transcription, mRNA degradation, translation initiation, translation, and tRNA charging.
+The transcription and translation template reactions come from sequence-specific analysis \cite{Allen:2003aa}, and include transcription initiation, transcription, mRNA degradation, translation initiation, translation, and tRNA charging.
 The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
@@
 
+%Need to talk more about biochemical benefits and importance for biochemical problems
+
 \clearpage
 
-%Need to talk more about biochemical benefits and importance for biochemical problems
+\section*{Competing interests}
+The authors declare that they have no competing interests.
+
+\section*{Author's contributions}
+J.V and A.Y directed the study. R.T, H.J and J.C conducted the cell culture measurements.
+J.V and W.D developed the reduced order HL-60 models and the parameter ensemble.
+W.D analyzed the model ensemble, and generated figures for the manuscript.
+The manuscript was prepared and edited for publication by W.D, A.Y and J.V.
+
+\section*{Acknowledgements}
+We gratefully acknowledge the suggestions from the anonymous reviewers to improve this manuscript.
 

### 2016-11-15 Paper_v1.tex
Some updates

@@
 Sensitivity analysis also showed that the system as a whole was most sensitive to these same parts of the network.
-CAT production was also robust to allosteric control, but other metabolites, specifically organic acid intermediates such as pyruvate, were not.
+CAT production and other metabolites, specifically organic acid intermediates such as pyruvate, were sensitive to the presence of allosteric control mechanisms.
 Next, to gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield with the maximum theoretical CAT carbon yield calculated using sequence-specific flux balance analysis.
@@
 The CFPS model equations were formulated using the hybrid cell-free modeling framework of Wayman et al. \cite{pr3010138}.
-An ensemble of probable model parameters (N $>$ 10,000) was estimated from the time-series metabolite measurements using a constrained Markov Chain Monte Carlo (MCMC) approach;
-the MCMC algorithm minimized the error between the training data and model simulations starting from an initial parameter estimate assembled from literature and inspection.
-We selected parameter sets for the ensemble based upon their error value and the Pearson correlation coefficient between sets in the ensemble.
+An ensemble of model parameters (N $>$ 10,000) was estimated from measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids using a constrained Markov Chain Monte Carlo (MCMC) approach.
+The MCMC algorithm minimized the error between the training data and model simulations starting from an initial parameter set assembled from literature and inspection.
+Parameter sets were selected for the ensemble based upon their error, and the Pearson correlation coefficient between the candidate and existing sets the ensemble.
 The parameter set with the lowest error value was defined as the best-fit set.
-The central carbon metabolism (Fig.~\ref{fig:CarbonBoth}, top), energy species (Fig.~\ref{fig:Energy}), and amino acids (Fig.~\ref{fig:Amino}) were all captured by the ensemble and the best-fit set.
-Allosteric control was important to the dynamics of the organic acid intermediates, as without it several of the measurements were not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).
-However,
+Central carbon metabolism (Fig.~\ref{fig:CarbonBoth}, top), energy species (Fig.~\ref{fig:Energy}), and amino acids (Fig.~\ref{fig:Amino}) were captured by the ensemble and the best-fit set.
+The constrained MCMC approach estimated parameter sets with a median error greater than two-order of magnitude less than random parameter sets generated within
+the same parameter bounds (Fig. \ref{fig:BoxPlot}); thus, we have confidence in the predictive capability of the estimated parameters.
+Allosteric control was important to the dynamics of the organic acid intermediates and CAT biosynthesis (Fig.~\ref{fig:CarbonBoth}, bottom).
+The acetate, lactate, pyruvate, succinate, malate and CAT trajectories were qualitatively different in the absence of allosteric control following glucose exhaustion.
+In particular, the rate of CAT biosynthesis and lactate accumulation and subsequent consumption decreased following glucose exhaustion in the absence of control.
 
-We used the hybrid cell-free modeling framework of Wayman and coworkers to simulate the production of CAT \citep{pr3010138}.
-The cell-free \textit{E.
+% We used the hybrid cell-free modeling framework of Wayman and coworkers to simulate the production of CAT \citep{pr3010138}.
+% The cell-free \textit{E.
 coli} metabolic network was constructed by removing the growth-associated processes from the model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS}, and by adding reactions for the synthesis of
-Thus, the model described core central carbon metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle), as well as the synthesis of energy species, amino acids biosynthesis and degradation, and biosynthesis of the CAT protein.
-An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids by minimizing the difference between the experimental dataset and metabolite simulations subject to literature constraints on parameter values and initial conditions.
+% Thus, the model described core central carbon metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle), as well as the synthesis of energy species, amino acids biosynthesis and degradation, and biosynthesis of the CAT protein.
+% An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids by minimizing the difference between the experimental dataset and metabolite simulations subject to literature constraints on parameter values and initial conditions.
+% We performed a local sensitivity analysis to determine the network reactions with the greatest effect on protein production and overall system state.
 
-\subsection*{Sensitivity analysis}
-We performed a local sensitivity analysis to determine the network reactions with the greatest effect on protein production and overall system state.
-%CAT production was most sensitive to the CAT synthesis reaction, oxidative phosphorylation activity, and alanine synthesis, as well as various reactions in glycolysis, pentose phosphate, the TCA cycle, and amino acid synthesis and degradation (Fig.~\ref{fig:Sensitivity}, top).
+To better understand which parameters, and parameter combinations influenced model performance we performed sensitivity analysis (Fig.~\ref{fig:Sensitivity}).
 CAT production was most sensitive to the CAT synthesis reaction, oxidative phosphorylation activity, and alanine synthesis (Fig.~\ref{fig:Sensitivity}, top, section A).
-The 16 next most important reactions to CAT production (section B) came from various pathways across the network: 4 each from glycolysis, the TCA cycle, and amino acid synthesis/degradation; 2 from pentose phosphate; and 1 each from the Entner-Doudoroff pathway and the energy species reactions.
-The pairwise sensitivities (off-diagonal elements) are often quite different from the corresponding first-order sensitivities (diagonal elements), and lead to some interesting outcomes.
-For example, glutamine synthesis and arginine degradation are both among the most important reactions to CAT production (they rank 5th and 10th, respectively).
-This is likely because they both affect the sensitive glutamine-glutamate balance; glutamine synthesis consumes glutamate, while arginine degradation produces it.
-However, when both are perturbed, their combined effect on the model is very low, as the respective consumption and production of glutamate cancel out.
+The 16 next most important reactions to CAT production (section B) came from various pathways across the network: four each from glycolysis, the TCA cycle, and amino acid synthesis/degradation; two from the pentose phosphate pathway; and one each from the Entner-Doudoroff pathway and the energy species reactions.
+The pairwise sensitivities (off-diagonal elements) were different from the corresponding first-order sensitivities (diagonal elements), and led to interesting outcomes.
+For example, glutamine synthesis and arginine degradation were both among the most important reactions to CAT production (they rank 5th and 10th, respectively).
+This was likely because they both affect the sensitive glutamine-glutamate balance; glutamine synthesis consumes glutamate, while arginine degradation produces it.
+However, when both were perturbed, their combined effect on the model was low, as the respective contributions to consumption and production of glutamate cancelled.
+The system state as a whole was most sensitive to glucose uptake via GTP and the forward reaction of lactate dehydrogenase (Fig.~\ref{fig:Sensitivity}, bottom, section F).
+The 30 next most important reactions to the system state (section G) came from various pathways across the network: eight from amino acid synthesis/degradation; six from glycolysis; four from the TCA cycle; and two each from the pentose phosphate pathway, Entner-Doudoroff, energy/reducing pathways, and small molecule transport; one from oxidative phosphorylation; and one from pyrophosphatase consumption.
+The system state had more pairwise sensitivities that differ from the corresponding first-order sensitivities and stand out as significant.
+For example, the first-order effect of alanine synthesis was large; it consumes both pyruvate and glutamate, two key species in the network.
+In addition, a handful of alanine synthesis pairwise sensitivities were also large.
+However, there were enough reactions that, when paired with alanine synthesis, had little effect on the model; malic enzyme is one of these, as it produces the pyruvate that alanine synthesis consumes.
+Thus, the total-order alanine synthesis sensitivity was low, placing it at the very bottom of section I.
+Another interesting result was the intersection of sections F and G with section J.
+The 53 reactions in section J were turned off in the best-fit set (rate constants were set to 0);
+therefore, the perturbation of these reactions had no effect on the model.
+Thus, all pairwise sensitivities with reactions in section J were pseudo first-order sensitivities for the other reactions.
+Interestingly, many reactions in section F and several in section G showed their highest sensitivities when paired with the "non-effects" of section J.
+Of these, three involved pyruvate, strengthening its role as a key metabolite; the others were glucose consumption via GTP/CTP-specific hexokinases,
+fumarate reductase, and SO$_4$ utilization.
+This suggested that these reactions' effects on the model were canceled out or lessened by most other reactions, but were of course not affected by the reactions in section J.
+This was also likely the reason that reactions in section J rank above those in section K, despite having no effect themselves on the model.
+Taken together, sensitivity analysis identified blocks of parameters that either individually, or in combination influenced model performance.
+However, the sensitivity analysis did not establish what the maximum performance of the system was.
+To answer that question we performed sequence-specific flux balance analysis of CAT production.
 
-The system state as a whole was most sensitive to glucose uptake via GTP and the forward reaction of lactate dehydrogenase (pyruvate being consumed to produce lactate) (Fig.~\ref{fig:Sensitivity}, bottom, section F).
-The 30 next most important reactions to the system state (section G) came from various pathways across the network: 8 from amino acid synthesis/degradation; 6 from glycolysis; 4 from TCA; 2 each from pentose phosphate, Entner-Doudoroff, energy species reactions, NAD species reactions, and small molecule transport; 1 from oxidative phosphorylation; and 1 pyrophosphatase.
-The system state has even more pairwise sensitivities that differ from the corresponding first-order sensitivities and stand out as significant.
-For example, the first-order effect of alanine synthesis is very high; it consumes both pruvate and glutamate, two key species in the network.
-In addition, a handful of alanine synthesis pairwise sensitivities are high.
-However, there are enough reactions that, when paired with alanine synthesis, have little effect on the model; malic enzyme is one of these, as it produces the pyruvate that alanine synthesis consumes.
-Thus, the total-order alanine synthesis sensitivity is low, placing it at the very bottom of section I.
-Another interesting result of this analysis is seen in the intersection of sections F and G with section J.
-The 53 reactions in section J were turned off in the best-fit set (rate constants were set to 0); therefore, the perturbation of these reactions (multiplying the rate constant by 1.01) had no actual effect on the model.
-Thus, all pairwise sensitivities with reactions in section J can essentially be considered first-order sensitivities for the other reactions.
-Interestingly, the reactions in section F and several in section G showed most of thier highest sensitivities when paired with the "non-effects" of section J.
-Of these, three involved pyruvate, strengthening its role as a key metabolite; the others were glucose uptake via GTP and CTP, fumarate reductase, and SO$_4$ transport.
-This suggests that these reactions' effects on the model were canceled out or lessened by most other reactions, but were of course not affected by the reactions in section J.
-This is also likely the reason that reactions in section J rank above those in section K, despite having no effect themselves on the model.
-
-\subsection*{Maximum theoretical CAT yield showed CFPS can be optimized.}
-We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
-%Arginine and glutamate were excluded due to not being present in the training dataset.
-The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
-While the model ensemble described the experimental data, it was unclear whether the performance of the CFPS system was optimal.
-To address this question, we used ssFBA in combination with the cell-free metabolic network and a detailed promoter model under a T7 polymerase to compute the maximum theoretical carbon yield.
-We first validated the ssFBA approach by comparing an ensemble of simulated versus measured concentrations of CAT over a one hour period (Fig.~\ref{fig:CATProdYield}A).
-The ensemble of 100 sets captured the CAT concentration profile which was generated by sampling RNA polymerase levels, ribosome levels and elongation rates in a physiological range.
-We then used sequence-specific FBA to calculate a theoretical maximum CAT yield under three different cases: unconstrained, constrained by transcription/translation rates, and constrained by transcription/translation rates and measurements (Fig.~\ref{fig:CATProdYield}B).
-The theoretical maximum carbon yield of CAT was 0.35 $\pm$ 0.006 for an unconstrained case and 0.225 $\pm$ 0.03 for the transcription and translation constrained case.
+We used sequence-specific flux balance analysis (ssFBA) to calculate the theoretical maximum CAT carbon yield for different constraint values (Fig.~\ref{fig:CATProdYield}).
+The experimental CAT carbon yield was 0.0865, while the best-fit parameter set had a carbon yield of 0.0871.
+Thus, although the kinetic model described the experimental data including the yield, it was unclear whether the performance of the CFPS system was optimal.
+To address this question, we used ssFBA in combination with the cell-free metabolic network and a T7 promoter model to estimate the maximum theoretical CAT carbon yield.
+Toward this, we first validated the ssFBA approach by comparing the simulated versus measured concentrations of CAT over the first hour of the CFPS reaction (Fig.~\ref{fig:CATProdYield}A).
+We sampled different RNA polymerase/ribosome levels and elongation rates in the physiological range to establish the uncertainty in the ssFBA simulation.
+The ssFBA estimate of the CAT abundance was consistent with the measured values. Next, we calculate the CAT carbon yield for three classes of constraints:
+(i) theoretical max glucose, amino acid and oxygen upper bounds, and no transcriptional/translational constraints; (ii) theoretical maximum glucose, amino acid and oxygen upper bounds,
+and realistic transcriptional/translational constraints and (iii) metabolite values constrained by the data, and realistic transcriptional/translational constraints.
+The unconstrained theoretical maximum CAT carbon yield was 0.35 $\pm$ 0.01 (Fig.~\ref{fig:CATProdYield}B, left);
+this case had no upper bound on the transcription and translation reactions, and was only constrained by glucose, oxygen and amino acid consumption rates.
+On the other hand, for realistic constraints on transcription and translation, the CAT carbon yield was 0.225 $\pm$ 0.03 (Fig.~\ref{fig:CATProdYield}B, middle).
+Lastly, when using realistic metabolite and transcription and translation constraints the predicted carbon yield was 0.062 $\pm$ 0.02, similar to the experimental yield Fig.~\ref{fig:CATProdYield}B, end).
 Thus, the experimental dataset and best-fit parameter set each produced CAT at 25\% of the theoretical maximum and 38\% of a theoretical physiological case.
-However, the case constrained by experimental data showed a carbon yield of 0.062 $\pm$ 0.02, similar to the experimental yield.
 
@@
 
-\subsection*{Formulation and solution of the model equations}
+\subsection*{Formulation and solution of the model equations.}
 We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell-free metabolic networks:
@@
 Conversely, if $\sigma_{ij}<0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij}=0$ indicates metabolite $i$ is not connected with reaction $j$.
-Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant.
+Lastly, $\lambda_{i}$ denotes the scaled enzyme activity decay constant.
 The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right)=\mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right)=\mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).
@@
 
-We included 17 allosteric regulation terms, taken from literature, to the CFPS model.
+We included 17 allosteric regulation terms, taken from literature, in the CFPS model.
 PEP was modeled as an inhibitor for phosphofructokinase \cite{2010_kotte_MolSystBiol,2011_cabrera_JBiolChem}, PEP carboxykinase \cite{2010_kotte_MolSystBiol}, PEP synthetase \cite{2010_kotte_MolSystBiol,1973_chulavatnatol_JBiolChem}, isocitrate dehydrogenase \cite{2010_kotte_MolSystBiol,2007_ogawa_JBacteriol}, and isocitrate lyase/malate synthase \cite{2010_kotte_MolSystBiol,2007_ogawa_JBacteriol,1988_mackintosh_BiochemJ}, and as an activator for fructose-biphosphatase \cite{2010_kotte_MolSystBiol,2000_donahue_JBacteriol,2006_hines_JBiolChem,2007_hines_JBiolChem}.
@@
 For the case constrained by TX/TL rates, an ensemble was generated by randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
-RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/sec \cite{2005_underwood_biotech, 2016_garamella_acssynbio}.
+RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/sec \cite{2005_underwood_biotech, Garamella:2016aa}.
 For the case constrained by TX/TL rates and experimental data, the lower and upper bounds on the fluxes for the data-informed metabolites were sampled within the range given by the experimental noise.
@@
 % Figures and captions go here ...
-\begin{figure}[ht]
-\centering
-\includegraphics[width=1.00\textwidth]{./Figures/CarbonBoth.pdf}
-\caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species.
+% \begin{figure}[ht]
+% \centering
+% \includegraphics[width=1.00\textwidth]{./Figures/CarbonBoth.pdf}
+% \caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species.
 Best-fit parameter set (orange line) versus experimental data (points).
 95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
-\label{fig:CarbonBoth}
-\end{figure}
+% \label{fig:CarbonBoth}
+% \end{figure}
 
@@
 95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
-\label{fig:CarbonBoth_png}
+\label{fig:CarbonBoth}
 \end{figure}

### 2016-11-15 Paper_v1.tex
MM update

@@
 \begin{equation}\label{eqn:yield-definition}
-	Yield=\frac{\Delta CAT\;C_{CAT}}{\sum_{i=1}^{\mathcal{R}}\max(\Delta m_{i},0)\;C_{m_i}}
+	Yield=\frac{\Delta m_{CAT}\;C_{CAT}}{\sum_{i=1}^{\mathcal{R}}\max(\Delta m_{i},0)\;C_{m_i}}
 \end{equation}
-where $\Delta CAT$ denotes the amount of CAT produced, $C_{CAT}$ denotes carbon number of CAT, $\mathcal{R}$ denotes the number of reactants, $\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed, never allowed to be negative, and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
-Because no data was available for arginine or glutamate, these reactants were left out of all yield calculations.
+where $\Delta m_{CAT}$ denotes the amount of CAT produced, $C_{CAT}$ denotes carbon number of CAT, $\mathcal{R}$ denotes the number of reactants, $\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed, never allowed to be negative, and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
+Because no data was available for arginine or glutamate, these reactants were left out of the experimental yield calculation.
 Yield of the best-fit parameter set and the experimental data were calculated by setting $\Delta CAT$ equal to the final minus the initial CAT concentration and setting $\Delta m_{i}$ equal to the initial minus the final reactant concentration.
-CAT carbon yields for three cases discussed below were calculated using flux balance anaylsis (FBA) with a sequence-specific description of CAT synthesis.
+Theoretical maximum CAT carbon yields for three cases discussed below were calculated using flux balance anaylsis (FBA) with a sequence-specific description of CAT synthesis, where $\Delta m_{i}$ denotes the flux of the $i^{th}$ species.
 This sequence-specific FBA \citep{Allen:2003aa} problem was formulated as:

### 2016-11-16 Paper_v1.tex
Nov16

@@
 The MCMC algorithm minimized the error between the training data and model simulations starting from an initial parameter set assembled from literature and inspection.
-Parameter sets were selected for the ensemble based upon their error, and the Pearson correlation coefficient between the candidate and existing sets the ensemble.
+Parameter sets were selected for the ensemble based upon their error, and the Pearson correlation coefficient between the candidate and existing sets in the ensemble.
 The parameter set with the lowest error value was defined as the best-fit set.
@@
 
-To better understand which parameters, and parameter combinations influenced model performance we performed sensitivity analysis (Fig.~\ref{fig:Sensitivity}).
+To better understand which parameters and parameter combinations influenced model performance we performed sensitivity analysis (Fig.~\ref{fig:Sensitivity}).
 CAT production was most sensitive to the CAT synthesis reaction, oxidative phosphorylation activity, and alanine synthesis (Fig.~\ref{fig:Sensitivity}, top, section A).
@@
 \end{equation}
-where $\mathcal{P}$ denotes the number of parameters ($\mathcal{P}$~=~815),
-which includes 163 rate constants, 163 enzyme activity decay constants, 455 saturation constants, and 34 control parameters,
-$k_i^{new}$ denotes the new value of the $i^{th}$ parameter,
-$k_i$ denotes the current value of the $i^{th}$ parameter,
-$a$ denotes a distribution variance, and $r_i$ denotes a random sample from the normal distribution.
+where $\mathcal{P}$ denotes the number of parameters ($\mathcal{P}$~=~815), which includes 163 rate constants, 163 enzyme activity decay constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i^{th}$ parameter, $k_i$ denotes the current value of the $i^{th}$ parameter, $a$ denotes a distribution variance, and $r_i$ denotes a random sample from the normal distribution.
+Parameters were kept between an upper and lower bound that varied by parameter type:
+\begin{equation}\label{eqn:parameter-bounds}
+    k_i^{new,bounded}=\min\left(\max\left(k_i^{new},l_i\right),u_i\right)
+\end{equation}
+where $k_i^{new,bounded}$ denotes the bounded parameter value, $l_i$ denotes the lower bound for that parameter type, and $u_i$ denotes the upper bound for that parameter type.
+Rate constants were bounded between 0 and 500000 mM/h [REFERENCE NEEDED]; this resulted in a mean of 324.6 mM/h and a standard deviation of 4311.2 mM/h for rate constants, giving a coefficient of variation (CV) of 13.3.
+Enzyme degradation rates were bounded between 0 and 1 h$^{-1}$ [REFERENCE NEEDED]; $\mu$ = 0.19 h$^{-1}$, $\sigma$ = 0.32 h$^{-1}$, CV = 1.64.
+Saturation constants were bounded between 0.001 and 10 mM [REFERENCE NEEDED]; $\mu$ = 0.76 mM, $\sigma$ = 1.71 mM, CV = 2.25.
+Control parameters (gains and orders) were left unbounded.
+Control gains: $\mu$ = 0.51, $\sigma$ = 1.21, CV = 2.35.
+Control orders: $\mu$ = 14.6, $\sigma$ = 66.0, CV = 4.53.
 For each newly generated parameter set, we re-solved the balance equations and calculated the cost function.
@@
 After generating 12,437 sets, we selected 100 sets with minimal set to set correlation to avoid over-sampling any region of parameter space.
-The original 12,437-set ensemble had a mean Pearson correlation coefficient of 0.94 between pairs of sets,
-while the 100-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.
+The original 12,437-set ensemble had a mean Pearson correlation coefficient of 0.94 between pairs of sets, while the 100-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.
 
@@
 The objective $w_{obj}$ was to maximize the specific rate of CAT formation.
-The specific glucose uptake rate was constrained to allow a maximum flux of 40 mM/hr according to experimental data; the specific amino acid uptake rates were bound to allow a maximum flux of 30 mM/hr, but did not reach this maximum flux.
+The specific glucose uptake rate was constrained to allow a maximum flux of 40 mM/h according to experimental data; the specific amino acid uptake rates were bound to allow a maximum flux of 30 mM/h, but did not reach this maximum flux.
 The transcription and translation template reactions come from sequence-specific analysis \cite{Allen:2003aa}, and include transcription initiation, transcription, mRNA degradation, translation initiation, translation, and tRNA charging.
@@
 For the unconstrained case, all rates were left unbounded, except the specific glucose uptake rate.
-An ensemble of flux distributions was then calculated by randomly sampling the maximum specific glucose uptake rate from within a range of 30 to 40 mM/hr, determined from experimental data.
+An ensemble of flux distributions was then calculated by randomly sampling the maximum specific glucose uptake rate from within a range of 30 to 40 mM/h, determined from experimental data.
 For the case constrained by TX/TL rates, an ensemble was generated by randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.

### 2016-11-18 Paper_v1.tex
Discussion update

@@
 The ensemble described all of the training data, especially the central carbon metabolism.
-The model predicted a carbon yield for CAT production that was equal to 25\% of the maximum theoretical yield, calculated using sequence-specific flux balance analysis.
+The model predicted a carbon yield for CAT production that was equal to 23\% of the maximum theoretical yield, calculated using sequence-specific flux balance analysis.
 This suggests that CAT production could be further optimized.
-While the dynamic models predicted that the majority of carbon flux went through glycolysis and the TCA cycle, the flux balance analysis showed significant flux through the Entner-Doudoroff pathway.
-The dynamic modeling approach predicted that glycolysis, the TCA cycle, and amino acid synthesis and degradation were most important to both CAT production and the system as a whole, while CAT production alone depended heavily on the CAT synthesis reaction.
+The dynamic modeling approach predicted that substrate consumption of glucose and pyruvate and oxidative phosphorylation were most important to both CAT production and the system as a whole, while CAT production alone depended heavily on the CAT synthesis reaction.
 Conversely, CAT production was robust to allosteric control, as was most of the network, with the exception of the organic acids in central carbon metabolism.
@@
 We then used the ensemble of kinetic models to analyze the CFPS reaction.
-First, sensitivity analysis of the dynamic model suggested that CAT production was most sensitive to CAT synthesis parameters, as well as reactions in amino acid synthesis/degradation, glycolysis, and the TCA cycle, and to a lesser extent the pentose phosphate pathway and oxidative phosphorylation.
-Sensitivity analysis also showed that the system as a whole was most sensitive to these same parts of the network.
+First, sensitivity analysis of the dynamic model suggested that CAT production was most sensitive to CAT synthesis parameters, as well as reactions in oxidative phosphorylation and pyruvate consumption.
+Sensitivity analysis also showed that the system as a whole was most sensitive to these same parts of the network and glucose consumption.
 CAT production and other metabolites, specifically organic acid intermediates such as pyruvate, were sensitive to the presence of allosteric control mechanisms.
 Next, to gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield with the maximum theoretical CAT carbon yield calculated using sequence-specific flux balance analysis.
-The CAT yield estimated from the kinetic model was 25\% of the maximum theoretical yield, but 38\% of the theoretical yield when physiologically realistic constraints were used.
-The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
-The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle, while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
+The CAT yield estimated from the kinetic model was 23\% of the maximum theoretical yield, but 36\% of the theoretical yield when physiologically realistic constraints were used.
 Taken together, we have integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
@@
 We used sequence-specific flux balance analysis (ssFBA) to calculate the theoretical maximum CAT carbon yield for different constraint values (Fig.~\ref{fig:CATProdYield}).
-The experimental CAT carbon yield was 0.0865, while the best-fit parameter set had a carbon yield of 0.0871.
+The experimental CAT carbon yield was 0.0821, while the best-fit parameter set had a carbon yield of 0.086 $\pm$ 0.004.
 Thus, although the kinetic model described the experimental data including the yield, it was unclear whether the performance of the CFPS system was optimal.
@@
 and realistic transcriptional/translational constraints and (iii) metabolite values constrained by the data, and realistic transcriptional/translational constraints.
-The unconstrained theoretical maximum CAT carbon yield was 0.35 $\pm$ 0.01 (Fig.~\ref{fig:CATProdYield}B, left);
+The unconstrained theoretical maximum CAT carbon yield was 0.363 $\pm$ 0.02 (Fig.~\ref{fig:CATProdYield}B, left);
 this case had no upper bound on the transcription and translation reactions, and was only constrained by glucose, oxygen and amino acid consumption rates.
-On the other hand, for realistic constraints on transcription and translation, the CAT carbon yield was 0.225 $\pm$ 0.03 (Fig.~\ref{fig:CATProdYield}B, middle).
+On the other hand, for realistic constraints on transcription and translation, the CAT carbon yield was 0.226 $\pm$ 0.03 (Fig.~\ref{fig:CATProdYield}B, middle).
 Lastly, when using realistic metabolite and transcription and translation constraints the predicted carbon yield was 0.062 $\pm$ 0.02, similar to the experimental yield Fig.~\ref{fig:CATProdYield}B, end).
-Thus, the experimental dataset and best-fit parameter set each produced CAT at 25\% of the theoretical maximum and 38\% of a theoretical physiological case.
+Thus, the experimental dataset and best-fit parameter set each produced CAT at 23\% of the theoretical maximum and 36\% of a theoretical physiological case.
 
@@
 In all cases, the energy source came primarily from oxidative phosphorylation, as well as partly from the TCA cycle.
-In the TX/TL constrained case, there was a high flux through fumerate dehydrogenase from aspartic acid uptake, whereas in the unconstrained and most constrained cases, acetate and lactate accumulation occurred.
+In the transcription/translation constrained case, there was a high flux through fumerate dehydrogenase from aspartic acid uptake, whereas in the unconstrained and most constrained cases, acetate and lactate accumulation occurred.
 This shows that the system is producing NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to supply enough NADH for oxidative phosphorylation.
@@
 To simulate potential knockouts, we constrained the specific glucose and amino acid uptake rates to the same values as simulated with no knockouts.
-In an ssFBA simulation with constrained TX/TL rates, knocking out the \textit{gnd} reaction decreases flux of acetate production but increases flux through \textit{pntAB}, which is responsible for regenerating NADPH.
+In an ssFBA simulation with constrained transcription/translation rates, knocking out the \textit{gnd} reaction decreases flux of acetate production but increases flux through \textit{pntAB}, which is responsible for regenerating NADPH.
 The simulation showed carbon was diverted toward lactate; however, since CAT production is constrained by the translation rate, we expected no increase in CAT production.
@@
 We used the hybrid cell-free modeling approach of Wayman and coworkers, which integrates traditional kinetic modeling with a logic-based description of allosteric regulation.
-We showed that the model produces CAT at 25\% of the theoretical maximum in terms of carbon yield, and at 38\% of a physiological case in which transcription and translation are constrained.
-The theoretical maximum and TX/TL constrained case were obtained using FBA, which predicted a different flux distribution than the ensemble of dynamic models.
-The ensemble of dynamic models predicted most of the carbon flux going through glycolysis and the TCA cycle, while FBA predicted significant flux through the Entner-Doudoroff pathway.
+We showed that the model produces CAT at 23\% of the theoretical maximum in terms of carbon yield, and at 36\% of a physiological case in which transcription and translation are constrained.
+The theoretical maximum and transcription/translation constrained case were obtained using FBA, which predicted a different flux distribution for each case.
+The unconstrained case predicted most of the carbon flux going through glycolysis, while the constrained cases predicted significant flux through the Entner-Doudoroff pathway and some through TCA cycle.
+However, all cases relied on oxidative phosphorylation for the system's energetic needs as well as for tRNA charging for CAT synthesis.
 Sensitivity analysis of the dynamic model suggested that both CAT production and the entire metabolic network were most sensitive to amino acid synthesis and degradation reactions, and reactions in glycolysis and the TCA cycle.
@@
 
+We present an ensemble of models that quantitatively describes the system behavior of cell-free metabolism and production of CAT.
+Experimental observations of the metabolites and cometabolites validate the structure of the model and the estimation of kinetic parameters. 
+This is important in applying metabolic engineering principles to rationally design cell-free production processes and predict the redirection of carbon fluxes to product forming pathways.
+The most sensitivite parameters for model performance are the uptake of glucose followed by pyruvate.
+This outstanding control on model performace was expected as these metabolites are responsible for driving CFPS and represent the first step in the model network.
+Nevertheless there are further reactions that excercise considerable sensitivity to model performance. 
+Oxidative phosphorylation activity shows to be vital to model performance, since it provides most of the energetic demands of CFPS according to the model.  
+In examining oxidative phosphorylation activity, knockouts in the electron transport pathways disrupt model performance and show CAT carbon yield dropping from 8.6\% to 2.7\%, consistent with literature observations\cite{2008_jewett_molsysbiol}.
+Additional reactions with considerable sensitivity are responsible for the consumption of glucose and pyruvate followed by the accumulation of pyruvate which allows it to be consumed during the second phase of CFPS.
+Following these reactions, the first step in pentose phosphate pathway is important as it degrades G6P to 6PG and produces NADPH.
+The interconversion of NADPH to NADH shows to be the next sensitive reaction, which helps to supply NADH to oxidative phosphorylation reactions to meet the energetic needs of the system. 
+The insignificant reactions lead to the accumulation of unwanted byproducts such as ethanol and formate.
+
+In analyzing the model sensitive parameters for CAT production, CAT synthesis followed by oxidative phosphorylation are the most sensitive.
+Following these reactions, the same theme of significance is seen as in the sensitivity of overall model performace. 
+The most important are glucose and pyruvate consuming reactions followed by downstream reactions from their respective products which is necessary to drive CFPS.
+In addition, the conversion of ATP to GTP shows significance since it aides in the translation of CAT.
+Thus to improve the efficiency of CAT production, supplementation with additional GTP may be of some benefit. 
+The overall sensitivity results show the substrates driving CFPS are vital to CFPS performance and CAT production, this can be seen by the biphasic operation of CFPS. 
+CFPS has biphasic formation of CAT, with the first phase operating on glucose and the second phase operating on pyruvate. 
+During the first phase, there is an accumulation of metabolite byproducts from central carbon with the majority of flux going towards acetate and some towards pyruvate, lactate, and succinate but these are consumed in the second phase, except for acetate.
+CFPS operating under glucose consumption has a higher production flux of CAT than that of pyruvate/lactate consumption. 
 In comparing the theoretical maximum carbon yield of CAT from ssFBA predictions to the kinetic model and experimental measurements, this suggests that there is potential for increasing CAT yield in CFPS as well as CFPS performance.
-The theoretical maximum yield of CAT was 0.35 for an unconstrained case and 0.225 for a transcription/translation constrained case.
+%Yield performance can be potentially improved by diverting carbon flux away from acetate. 
+The theoretical maximum yield of CAT was 36\% for an unconstrained case and 22.6\% for a transcription/translation constrained case.
 Knockouts of \textit{gnd} and phosphate acetyltransferase show carbon can be diverted away from acetate and potentially toward CAT or other proteins of interest expressed in CFPS.
@@
 Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
-In addition to improving CFPS performance, Jewett and coworkers have shown that oxidative phosphorylation operates in cell-free systems, and that knocking out these reactions is detrimental to protein yield \cite{2008_jewett_molsysbiol}.
+In addition to improving CFPS performance, Jewett and coworkers have shown that oxidative phosphorylation operates in cell-free systems, and that knocking out these reactions is detrimental to protein yield\cite{2008_jewett_molsysbiol}.
 However, it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, and both of the modeling approaches we present suggest that oxidative phosphorylation is important to CAT production.
@@
 Lastly, the cost calculation was scaled by the maximum experimental value in the $i^{th}$ dataset, $\mathcal{Y}_{i}=\max_{j}\left(y_{ij}\right)$.
-We then perturbed each model parameter:
+We then perturbed each model parameter between an upper and lower bound that varied by parameter type:
 \begin{equation}\label{eqn:parameter-perturbation}
-    k_i^{new}=k_i \cdot \exp(a\,r_i)\qquad{i=1,2,\hdots,\mathcal{P}}
+    k_i^{new}=\min\left(\max\left(k_i \cdot \exp(a \cdot r_i),l_i\right),u_i\right)\qquad{i=1,2,\hdots,\mathcal{P}}
 \end{equation}
-where $\mathcal{P}$ denotes the number of parameters ($\mathcal{P}$~=~815), which includes 163 rate constants, 163 enzyme activity decay constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i^{th}$ parameter, $k_i$ denotes the current value of the $i^{th}$ parameter, $a$ denotes a distribution variance, and $r_i$ denotes a random sample from the normal distribution.
-Parameters were kept between an upper and lower bound that varied by parameter type:
-\begin{equation}\label{eqn:parameter-bounds}
-    k_i^{new,bounded}=\min\left(\max\left(k_i^{new},l_i\right),u_i\right)
-\end{equation}
-where $k_i^{new,bounded}$ denotes the bounded parameter value, $l_i$ denotes the lower bound for that parameter type, and $u_i$ denotes the upper bound for that parameter type.
-Rate constants were bounded between 0 and 500000 mM/h [REFERENCE NEEDED]; this resulted in a mean of 324.6 mM/h and a standard deviation of 4311.2 mM/h for rate constants, giving a coefficient of variation (CV) of 13.3.
+where $\mathcal{P}$ denotes the number of parameters ($\mathcal{P}$~=~815), which includes 163 rate constants, 163 enzyme activity decay constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i^{th}$ parameter, $k_i$ denotes the current value of the $i^{th}$ parameter, $a$ denotes a distribution variance, $r_i$ denotes a random sample from the normal distribution, $l_i$ denotes the lower bound for that parameter type, and $u_i$ denotes the upper bound for that parameter type.
+Rate constants were bounded between 0 and 500,000 mM/h [REFERENCE NEEDED]; this resulted in a mean of 324.6 mM/h and a standard deviation of 4311.2 mM/h for rate constants, giving a coefficient of variation (CV) of 13.3.
 Enzyme degradation rates were bounded between 0 and 1 h$^{-1}$ [REFERENCE NEEDED]; $\mu$ = 0.19 h$^{-1}$, $\sigma$ = 0.32 h$^{-1}$, CV = 1.64.
@@
 \begin{equation}\label{eqn:system-sensitivity}
-    \mathcal{S}_{ij}^{\texttt{X}}=\norm{\texttt{X}(p_i,p_j,t)-\texttt{X}(\alpha*p_i,\alpha*p_j,t)}\qquad{i,j=1,2,\hdots\mathcal{P}}
+    \mathcal{S}_{ij}^{\texttt{X}}=\norm{\texttt{X}(p_i,p_j,t)-\texttt{X}(\alpha \cdot p_i,\alpha \cdot p_j,t)}\qquad{i,j=1,2,\hdots\mathcal{P}}
 \end{equation}
@@
 where $\Delta m_{CAT}$ denotes the amount of CAT produced, $C_{CAT}$ denotes carbon number of CAT, $\mathcal{R}$ denotes the number of reactants, $\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed, never allowed to be negative, and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
-Because no data was available for arginine or glutamate, these reactants were left out of the experimental yield calculation.
+Because no data was available for arginine or glutamate, these reactants were left out of all yield calculations.
 Yield of the best-fit parameter set and the experimental data were calculated by setting $\Delta CAT$ equal to the final minus the initial CAT concentration and setting $\Delta m_{i}$ equal to the initial minus the final reactant concentration.
@@
 This sequence-specific FBA \citep{Allen:2003aa} problem was formulated as:
-\begin{equation}\nonumber
+\begin{equation}
  \begin{multlined}
@@
 The transcription rate was formulated as:
-\begin{equation}\nonumber
+\begin{equation}
 	w_{TX} = RNAP\left(\frac{v_{RNAP}}{l_{mRNA}}\right)\left(\frac{Gene}{k_m+Gene}\right)P
@@
 The promoter activity was formulated following Moon et al. for synthetic circuits as:
-\begin{equation}\nonumber
+\begin{equation}
 	P = \frac{K_{1}}{1 + K_{1}}
@@
 The translation rate was formulated as:
- \begin{equation}\nonumber
+ \begin{equation}
 	w_{TL} = K_{P}\ Ribo\left(\frac{v_{Ribo}}{l_{Protein}}\right)mRNA_{SS}
@@
 
-An ensemble of 100 sets of flux distributions was calculated for three different cases: unconstrained, constrained by transcription and translation (TX/TL) rates, and  constrained by TX/TL rates and experimental data.
+An ensemble of 100 sets of flux distributions was calculated for three different cases: unconstrained, constrained by transcription/translation rates, and constrained by transcription/translation rates and experimental measurements.
 For the unconstrained case, all rates were left unbounded, except the specific glucose uptake rate.
 An ensemble of flux distributions was then calculated by randomly sampling the maximum specific glucose uptake rate from within a range of 30 to 40 mM/h, determined from experimental data.
-For the case constrained by TX/TL rates, an ensemble was generated by randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
+For the case constrained by transcription/translation rates, an ensemble was generated by randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
 RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/sec \cite{2005_underwood_biotech, Garamella:2016aa}.
-For the case constrained by TX/TL rates and experimental data, the lower and upper bounds on the fluxes for the data-informed metabolites were sampled within the range given by the experimental noise.
-This included the data for glucose, organic acids, energy species, and amino acids; CAT was not constrained by experimental data, but by the TX/TL rates as stated above.
+For the case constrained by transcription/translation rates and experimental measurements, the lower and upper bounds on the fluxes for the data-informed metabolites were sampled within the range given by the experimental noise.
+This included the data for glucose, organic acids, energy species, and amino acids; CAT was not constrained by experimental data, but by the transcription/translation rates as stated above.
 
@@
 B.
-Theoretical maximum carbon yield of CAT calcualted by ssFBA for three different cases: unconstrained except for glucose uptake (black), constrained by transcription and translation rates (grey), and constrained by transcription, translation rates and experimental measurements where available (light grey).
+Theoretical maximum carbon yield of CAT calcualted by ssFBA for three different cases: unconstrained except for glucose uptake (black), constrained by transcription/translation (TX/TL) rates (grey), and constrained by transcription/translation (TX/TL) rates and experimental measurements where available (light grey).
 Error bars represent standard deviation of the ensemble.}

### 2016-11-18 Paper_v1.tex
Nov18

@@
 This is important in applying metabolic engineering principles to rationally design cell-free production processes and predict the redirection of carbon fluxes to product forming pathways.
-The most sensitivite parameters for model performance are the uptake of glucose followed by pyruvate.
+The most sensitive parameter for the model as a whole is the uptake of glucose, followed by pyruvate.
 This outstanding control on model performace was expected as these metabolites are responsible for driving CFPS and represent the first step in the model network.
-Nevertheless there are further reactions that excercise considerable sensitivity to model performance. 
-Oxidative phosphorylation activity shows to be vital to model performance, since it provides most of the energetic demands of CFPS according to the model.  
-In examining oxidative phosphorylation activity, knockouts in the electron transport pathways disrupt model performance and show CAT carbon yield dropping from 8.6\% to 2.7\%, consistent with literature observations\cite{2008_jewett_molsysbiol}.
-Additional reactions with considerable sensitivity are responsible for the consumption of glucose and pyruvate followed by the accumulation of pyruvate which allows it to be consumed during the second phase of CFPS.
-Following these reactions, the first step in pentose phosphate pathway is important as it degrades G6P to 6PG and produces NADPH.
-The interconversion of NADPH to NADH shows to be the next sensitive reaction, which helps to supply NADH to oxidative phosphorylation reactions to meet the energetic needs of the system. 
-The insignificant reactions lead to the accumulation of unwanted byproducts such as ethanol and formate.
+Nevertheless, there are further reactions that excercise considerable sensitivity to model performance. 
+Oxidative phosphorylation activity is vital, since it provides most of the energetic needs of CFPS.
+In examining oxidative phosphorylation activity, knockouts in the electron transport pathways disrupt metabolism across the network and show CAT carbon yield dropping from 8.6\% to 2.7\%, consistent with literature observations \cite{2008_jewett_molsysbiol}.
+Jewett and coworkers also saw a decrease in CAT yield, ranging from 1.5-fold to 4-fold, when knocking out oxidative phosphorylation.
+Other reactions with considerable sensitivity are those responsible for the consumption of glucose and the consumption and production of pyruvate, which is consumed during the second phase of CFPS. 
+Also important is the first step in pentose phosphate pathway, as it converts G6P to 6PG and produces NADPH.
+The next most sensitive reaction is the conversion of NADPH to NADH, which helps to supply NADH to oxidative phosphorylation reactions to meet the energetic needs of the system. 
+Many of the insignificant reactions lead to the accumulation of unwanted byproducts such as ethanol and formate.
 
-In analyzing the model sensitive parameters for CAT production, CAT synthesis followed by oxidative phosphorylation are the most sensitive.
-Following these reactions, the same theme of significance is seen as in the sensitivity of overall model performace. 
-The most important are glucose and pyruvate consuming reactions followed by downstream reactions from their respective products which is necessary to drive CFPS.
-In addition, the conversion of ATP to GTP shows significance since it aides in the translation of CAT.
-Thus to improve the efficiency of CAT production, supplementation with additional GTP may be of some benefit. 
-The overall sensitivity results show the substrates driving CFPS are vital to CFPS performance and CAT production, this can be seen by the biphasic operation of CFPS. 
-CFPS has biphasic formation of CAT, with the first phase operating on glucose and the second phase operating on pyruvate. 
-During the first phase, there is an accumulation of metabolite byproducts from central carbon with the majority of flux going towards acetate and some towards pyruvate, lactate, and succinate but these are consumed in the second phase, except for acetate.
-CFPS operating under glucose consumption has a higher production flux of CAT than that of pyruvate/lactate consumption. 
-In comparing the theoretical maximum carbon yield of CAT from ssFBA predictions to the kinetic model and experimental measurements, this suggests that there is potential for increasing CAT yield in CFPS as well as CFPS performance.
+In analyzing the model parameters' effect on CAT production, CAT synthesis is the most important, followed by oxidative phosphorylation and the glucose and pyruvate consuming reactions, as well as the downstream reactions which are necessary to drive CFPS.
+In addition, the conversion of ATP to GTP shows significance since it facilitates CAT synthesis.
+Thus, supplementation with additional GTP may improve the efficiency of CAT production.
+A similar theme is seen in the sensitivity of overall model performace; the most important reactions are glucose and pyruvate consuming reactions and downstream reactions.
+The overall sensitivity results show that the substrates driving CFPS are vital to system performance and CAT production.
+This can be seen in the biphasic operation of CFPS, with the first phase operating on glucose and the second phase operating on pyruvate. 
+During the first phase, there is an accumulation of byproducts from central carbon with the majority of flux going toward acetate and some toward pyruvate, lactate, and succinate, but with the exception of acetate these are all consumed in the second phase.
+CFPS operating under glucose consumption has higher CAT production than that of pyruvate/lactate consumption, as can be seen in the experimental data (\texttildelow 10 $\mu$M/h vs \texttildelow 5 $\mu$M/h).
+
+Comparing the theoretical maximum carbon yield of CAT from ssFBA predictions to those of the kinetic model and experimental measurements suggests that there is potential for increasing CAT yield as well as CFPS performance.
 %Yield performance can be potentially improved by diverting carbon flux away from acetate. 
-The theoretical maximum yield of CAT was 36\% for an unconstrained case and 22.6\% for a transcription/translation constrained case.
-Knockouts of \textit{gnd} and phosphate acetyltransferase show carbon can be diverted away from acetate and potentially toward CAT or other proteins of interest expressed in CFPS.
-Another limitation to be addressed in CFPS is the transcription and translation description, since the protein of interest to be expressed is ultimately bounded by these kinetic rates.
+The theoretical maximum CAT yield was 36\% for an unconstrained case and 23\% for a transcription/translation constrained case.
+Knockouts of \textit{gnd} and phosphate acetyltransferase show that carbon can be diverted away from acetate and potentially toward CAT or other proteins of interest expressed in CFPS.
+Another limitation to be addressed in CFPS is the transcription and translation description, since protein production is ultimately bounded by these kinetic rates.
 Li et al.
@@
 Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
-In addition to improving CFPS performance, Jewett and coworkers have shown that oxidative phosphorylation operates in cell-free systems, and that knocking out these reactions is detrimental to protein yield\cite{2008_jewett_molsysbiol}.
-However, it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, and both of the modeling approaches we present suggest that oxidative phosphorylation is important to CAT production.
-Thus, this is a potential area for improvement of CFPS performance and protein yield.
+In addition to improving CFPS performance, Jewett and coworkers have shown that oxidative phosphorylation operates in cell-free systems, and that knocking out these reactions is detrimental to protein yield \cite{2008_jewett_molsysbiol}.
+However, it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, and both of the modeling approaches we present suggest its importance to CAT production.
+Thus, oxidative phosphorylation is a potential area for improvement of CFPS performance and protein yield.
 

### 2016-11-18 Paper_v1.tex
Nov18_2

@@
 
-% We used the hybrid cell-free modeling framework of Wayman and coworkers to simulate the production of CAT \citep{pr3010138}.
-% The cell-free \textit{E.
-coli} metabolic network was constructed by removing the growth-associated processes from the model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS}, and by adding reactions for the synthesis of
-% Thus, the model described core central carbon metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle), as well as the synthesis of energy species, amino acids biosynthesis and degradation, and biosynthesis of the CAT protein.
-% An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids by minimizing the difference between the experimental dataset and metabolite simulations subject to literature constraints on parameter values and initial conditions.
-% We performed a local sensitivity analysis to determine the network reactions with the greatest effect on protein production and overall system state.
+We used sequence-specific flux balance analysis (ssFBA) to calculate the theoretical maximum CAT carbon yield for different constraint values (Fig.~\ref{fig:CATProdYield}).
+The experimental CAT carbon yield was 0.0821, while the best-fit parameter set had a carbon yield of 0.086 $\pm$ 0.004.
+Thus, although the kinetic model described the experimental data including the yield, it was unclear whether the performance of the CFPS system was optimal.
+To address this question, we used ssFBA in combination with the cell-free metabolic network and a T7 promoter model to estimate the maximum theoretical CAT carbon yield.
+Toward this, we first validated the ssFBA approach by comparing the simulated versus measured concentrations of CAT over the first hour of the CFPS reaction (Fig.~\ref{fig:CATProdYield}A).
+We sampled different RNA polymerase/ribosome levels and elongation rates in the physiological range to establish the uncertainty in the ssFBA simulation.
+The ssFBA estimate of the CAT abundance was consistent with the measured values. Next, we calculate the CAT carbon yield for three classes of constraints:
+(i) theoretical max glucose, amino acid and oxygen upper bounds, and no transcriptional/translational constraints; (ii) theoretical maximum glucose, amino acid and oxygen upper bounds,
+and realistic transcriptional/translational constraints and (iii) metabolite values constrained by the data, and realistic transcriptional/translational constraints.
+The unconstrained theoretical maximum CAT carbon yield was 0.363 $\pm$ 0.02 (Fig.~\ref{fig:CATProdYield}B, left);
+this case had no upper bound on the transcription and translation reactions, and was only constrained by glucose, oxygen and amino acid consumption rates.
+On the other hand, for realistic constraints on transcription and translation, the CAT carbon yield was 0.226 $\pm$ 0.03 (Fig.~\ref{fig:CATProdYield}B, middle).
+Lastly, when using realistic metabolite and transcription and translation constraints the predicted carbon yield was 0.062 $\pm$ 0.02, similar to the experimental yield Fig.~\ref{fig:CATProdYield}B, end).
+Thus, the experimental dataset and best-fit parameter set each produced CAT at 23\% of the theoretical maximum and 36\% of a theoretical physiological case.
 
@@
 Another interesting result was the intersection of sections F and G with section J.
-The 53 reactions in section J were turned off in the best-fit set (rate constants were set to 0);
+The 53 reactions in section J were turned off in the best-fit set ($V^{max}=0$);
 therefore, the perturbation of these reactions had no effect on the model.
@@
 To answer that question we performed sequence-specific flux balance analysis of CAT production.
-
-We used sequence-specific flux balance analysis (ssFBA) to calculate the theoretical maximum CAT carbon yield for different constraint values (Fig.~\ref{fig:CATProdYield}).
-The experimental CAT carbon yield was 0.0821, while the best-fit parameter set had a carbon yield of 0.086 $\pm$ 0.004.
-Thus, although the kinetic model described the experimental data including the yield, it was unclear whether the performance of the CFPS system was optimal.
-To address this question, we used ssFBA in combination with the cell-free metabolic network and a T7 promoter model to estimate the maximum theoretical CAT carbon yield.
-Toward this, we first validated the ssFBA approach by comparing the simulated versus measured concentrations of CAT over the first hour of the CFPS reaction (Fig.~\ref{fig:CATProdYield}A).
-We sampled different RNA polymerase/ribosome levels and elongation rates in the physiological range to establish the uncertainty in the ssFBA simulation.
-The ssFBA estimate of the CAT abundance was consistent with the measured values. Next, we calculate the CAT carbon yield for three classes of constraints:
-(i) theoretical max glucose, amino acid and oxygen upper bounds, and no transcriptional/translational constraints; (ii) theoretical maximum glucose, amino acid and oxygen upper bounds,
-and realistic transcriptional/translational constraints and (iii) metabolite values constrained by the data, and realistic transcriptional/translational constraints.
-The unconstrained theoretical maximum CAT carbon yield was 0.363 $\pm$ 0.02 (Fig.~\ref{fig:CATProdYield}B, left);
-this case had no upper bound on the transcription and translation reactions, and was only constrained by glucose, oxygen and amino acid consumption rates.
-On the other hand, for realistic constraints on transcription and translation, the CAT carbon yield was 0.226 $\pm$ 0.03 (Fig.~\ref{fig:CATProdYield}B, middle).
-Lastly, when using realistic metabolite and transcription and translation constraints the predicted carbon yield was 0.062 $\pm$ 0.02, similar to the experimental yield Fig.~\ref{fig:CATProdYield}B, end).
-Thus, the experimental dataset and best-fit parameter set each produced CAT at 23\% of the theoretical maximum and 36\% of a theoretical physiological case.
 
@@
 \end{equation}
-where $\mathcal{P}$ denotes the number of parameters ($\mathcal{P}$~=~815), which includes 163 rate constants, 163 enzyme activity decay constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i^{th}$ parameter, $k_i$ denotes the current value of the $i^{th}$ parameter, $a$ denotes a distribution variance, $r_i$ denotes a random sample from the normal distribution, $l_i$ denotes the lower bound for that parameter type, and $u_i$ denotes the upper bound for that parameter type.
-Rate constants were bounded between 0 and 500,000 mM/h [REFERENCE NEEDED]; this resulted in a mean of 324.6 mM/h and a standard deviation of 4311.2 mM/h for rate constants, giving a coefficient of variation (CV) of 13.3.
-Enzyme degradation rates were bounded between 0 and 1 h$^{-1}$ [REFERENCE NEEDED]; $\mu$ = 0.19 h$^{-1}$, $\sigma$ = 0.32 h$^{-1}$, CV = 1.64.
-Saturation constants were bounded between 0.001 and 10 mM [REFERENCE NEEDED]; $\mu$ = 0.76 mM, $\sigma$ = 1.71 mM, CV = 2.25.
-Control parameters (gains and orders) were left unbounded.
-Control gains: $\mu$ = 0.51, $\sigma$ = 1.21, CV = 2.35.
-Control orders: $\mu$ = 14.6, $\sigma$ = 66.0, CV = 4.53.
+where $\mathcal{P}$ denotes the number of parameters ($\mathcal{P}$~=~815), which includes 163 maximum reaction rates ($V^{max}$), 163 enzyme activity decay constants, 455 saturation constants ($K_{js}$), and 34 control parameters, $k_i^{new}$ denotes the new value of the $i^{th}$ parameter, $k_i$ denotes the current value of the $i^{th}$ parameter, $a$ denotes a distribution variance, $r_i$ denotes a random sample from the normal distribution, $l_i$ denotes the lower bound for that parameter type, and $u_i$ denotes the upper bound for that parameter type.
+Maximum reaction rates were bounded between 0 and 500,000 mM/h \cite{Keseler:2013aa}.
+Assuming a total enzyme concentration of 50 $\mu$M, this corresponds to catalytic rate bounds of 0 and 2778 s$^{-1}$.
+These bounds resulted in a median catalytic rate of 0.016 s$^{-1}$ across the ensemble.
+Enzyme activity decay constants were bounded between 0 and 1 h$^{-1}$, corresponding to half lives of 42 minutes and infinity; median = 25 min.
+Saturation constants were bounded between 0.001 and 10 mM; median = 0.16 mM.
+Control parameters (gains and orders) were left unbounded; gain median = 0.076, order median = 0.69.
 For each newly generated parameter set, we re-solved the balance equations and calculated the cost function.
@@
 \subsection*{Sensitivity analysis of the CFPS model.}
-We determined the reactions most important to protein production by computing the local sensitivity of CAT concentration (denoted as $\texttt{CAT}$)
-to each individual rate constant, and each pair of rate constants in the network.
+We determined the reactions most important to protein production by computing the local sensitivity of CAT concentration (denoted as $\texttt{CAT}$) to each individual maximum reaction rate, and each pair of maximum reaction rates in the network.
 The sensitivity index was formulated as:
@@
 $\texttt{CAT}(p_i,p_j,t)$ denotes CAT concentration as a function of time and the $i^{th}$ and $j^{th}$ parameters,
-$\alpha$ denotes the perturbation factor, and $\mathcal{P}$ denotes the number of rate constants ($\mathcal{P}$ = 163).
+$\alpha$ denotes the perturbation factor, and $\mathcal{P}$ denotes the number of maximum reaction rates ($\mathcal{P}$ = 163).
 In calculating the pairwise sensitivities, each parameter was perturbed by 1\%; first-order sensitivities ($i$ = $j$) were subject to two 1\% perturbations.
-% The array of sensitivity indices was organized into a clustergram with both axes corresponding to the 163 rate constants
+% The array of sensitivity indices was organized into a clustergram with both axes corresponding to the 163 maximum reaction rates
 % rearranged from most sensitive (top, right) to least sensitive (bottom, left).
@@
 Likewise, we determined the reactions most important to global system performance by computing the sensitivity of all species for which data exists (denoted as $\texttt{X}$)
-to each rate constant in the network.
+to each maximum reaction rate in the network.
 In this case, each sensitivity index was formulated as:
@@
 \includegraphics[width=0.76\textwidth,trim=0 28 615 40,clip]{./Figures/Sensitivity.pdf}
-\caption{Normalized first-order and pairwise sensitivities of CAT production (top) and system state (bottom) to rate constants.}
+\caption{Normalized first-order and pairwise sensitivities of CAT production (top) and system state (bottom) to maximum reaction rates.}
 \label{fig:Sensitivity}

### 2016-11-21 Paper_v1.tex
some changes on materials and methods

@@
 The ensemble of kinetic CFPS models captured the time evolution of CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
-The cell-free \textit{E.~coli} metabolic network was constructed by removing growth associated reactions from the iAF1260 reconstruction \cite{Feist:2007aa},
+The cell-free \textit{E.~coli} metabolic network was constructed by removing growth associated reactions from the MG1655 reconstruction \cite{Feist121},
 and by adding reactions describing chloramphenicol acetyltransferase (CAT) biosynthesis, a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
@@
 the same parameter bounds (Fig. \ref{fig:BoxPlot}); thus, we have confidence in the predictive capability of the estimated parameters.
-Allosteric control was important to the dynamics of the organic acid intermediates and CAT biosynthesis (Fig.~\ref{fig:CarbonBoth}, bottom).
-The acetate, lactate, pyruvate, succinate, malate and CAT trajectories were qualitatively different in the absence of allosteric control following glucose exhaustion.
-In particular, the rate of CAT biosynthesis and lactate accumulation and subsequent consumption decreased following glucose exhaustion in the absence of control.
+The model captures the biphasic nature of CFPS shown by the experimental data: in the first hour glucose consumption powers the system, and CAT is produced at \texttildelow 10 $\mu$M/h; subsequently, pyruvate and lactate reserves are consumed to power metabolism, and CAT is produced less efficiently at \texttildelow 5 $\mu$M/h.
+Allosteric control was important to this biphasic operation, as without it, both phases are sped up, and after 1.5 hours CAT synthesis has ceased (Fig.~\ref{fig:CarbonBoth}, bottom).
+In addition, acetate accumulation also stops after 1.5 hours since there is no more substrate to power the system.
+Interestingly, malate captures the experimental measurements during the glucose consumption phase, but increases sharply during the pyruvate consumption phase.
 
-We used sequence-specific flux balance analysis (ssFBA) to calculate the theoretical maximum CAT carbon yield for different constraint values (Fig.~\ref{fig:CATProdYield}).
-The experimental CAT carbon yield was 0.0821, while the best-fit parameter set had a carbon yield of 0.086 $\pm$ 0.004.
-Thus, although the kinetic model described the experimental data including the yield, it was unclear whether the performance of the CFPS system was optimal.
-To address this question, we used ssFBA in combination with the cell-free metabolic network and a T7 promoter model to estimate the maximum theoretical CAT carbon yield.
-Toward this, we first validated the ssFBA approach by comparing the simulated versus measured concentrations of CAT over the first hour of the CFPS reaction (Fig.~\ref{fig:CATProdYield}A).
-We sampled different RNA polymerase/ribosome levels and elongation rates in the physiological range to establish the uncertainty in the ssFBA simulation.
-The ssFBA estimate of the CAT abundance was consistent with the measured values. Next, we calculate the CAT carbon yield for three classes of constraints:
-(i) theoretical max glucose, amino acid and oxygen upper bounds, and no transcriptional/translational constraints; (ii) theoretical maximum glucose, amino acid and oxygen upper bounds,
-and realistic transcriptional/translational constraints and (iii) metabolite values constrained by the data, and realistic transcriptional/translational constraints.
-The unconstrained theoretical maximum CAT carbon yield was 0.363 $\pm$ 0.02 (Fig.~\ref{fig:CATProdYield}B, left);
-this case had no upper bound on the transcription and translation reactions, and was only constrained by glucose, oxygen and amino acid consumption rates.
-On the other hand, for realistic constraints on transcription and translation, the CAT carbon yield was 0.226 $\pm$ 0.03 (Fig.~\ref{fig:CATProdYield}B, middle).
-Lastly, when using realistic metabolite and transcription and translation constraints the predicted carbon yield was 0.062 $\pm$ 0.02, similar to the experimental yield Fig.~\ref{fig:CATProdYield}B, end).
-Thus, the experimental dataset and best-fit parameter set each produced CAT at 23\% of the theoretical maximum and 36\% of a theoretical physiological case.
+While the model ensemble decribed the experimental data, it was unclear whether the performance of CFPS was optimal.
+To address this question, we used sequence-specific flux balance analysis in combination with a T7 promoter model to estimate the theoretical maximum performance of the system.
+We first validated the ssFBA approach by comparing simulated and measured concentrations of CAT for the first hour under glucose consumption (Fig.~\ref{fig:CATProdYield}A).
+The ssFBA ensemble was able to predict the CAT dataset despite having no adjustable parameters.
+Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates, or the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
+Thus, the metabolic network and molecular description of transcription and translation were consistent with experimental measurements.
+To gauge the performance of CFPS, we next calculated the CAT carbon yield for three classes of constraints: (i) theoretical maximum glucose, amino acid and oxygen upper bounds, and no transcriptional/translational constraints; (ii) theoretical maximum glucose, amino acid and oxygen upper bounds, and realistic transcriptional/translational constraints; and (iii) metabolite fluxes constrained by the data, and realistic transcriptional/translational constraints (Fig.~\ref{fig:CATProdYield}B).
+The unconstrained theoretical maximum CAT carbon yield (i) was 0.363 $\pm$ 0.02 (Fig.~\ref{fig:CATProdYield}B, left);
+On the other hand, for realistic constraints on transcription and translation (ii), the CAT carbon yield was 0.226 $\pm$ 0.03 (Fig.~\ref{fig:CATProdYield}B, middle).
+Lastly, when using realistic metabolite and transcription and translation constraints (iii), the predicted carbon yield was 0.062 $\pm$ 0.02.
+By comparison, the best-fit parameter set had a CAT carbon yield of 0.086 $\pm$ 0.004, equivalent to 23\% of the theoretical maximum (i) and 36\% of the physiological case (ii).
+The experimental dataset had a CAT carbon yield of 0.0821, similar to both the kinetic model and the experimentally constrained case (iii).
 
 To better understand which parameters and parameter combinations influenced model performance we performed sensitivity analysis (Fig.~\ref{fig:Sensitivity}).
-CAT production was most sensitive to the CAT synthesis reaction, oxidative phosphorylation activity, and alanine synthesis (Fig.~\ref{fig:Sensitivity}, top, section A).
-The 16 next most important reactions to CAT production (section B) came from various pathways across the network: four each from glycolysis, the TCA cycle, and amino acid synthesis/degradation; two from the pentose phosphate pathway; and one each from the Entner-Doudoroff pathway and the energy species reactions.
+According to the sensitivity results, CAT production was most dependent on coenzymes and pyruvate.
+CAT production was most sensitive to the CAT synthesis reaction, oxidative phosphorylation, and the pyruvate-consuming alanine synthesis reaction (Fig.~\ref{fig:Sensitivity}, top, section A).
+Taking into account these three reactions as well as the next 16 most important reactions (section B), we see a common theme of reactions that involve the coenzymes ATP, GTP, NADH, and NADPH as well as the metabolites pyruvate and glutamate.
+Glutamate is important as a reagant for the synthesis of all other amino acids which are required to assemble the CAT protein.
+Meanwhile, the coenzymes are needed to provide energy to power the synthesis of CAT, and pyruvate becomes important for energy supply after glucose is exhausted.
+In addition, pyruvate is a key metabolite required for the synthesis of several amino acids.
 The pairwise sensitivities (off-diagonal elements) were different from the corresponding first-order sensitivities (diagonal elements), and led to interesting outcomes.
+The combination of certain reactions had a greater effect on CAT production than that of the reactions by themselves.
 For example, glutamine synthesis and arginine degradation were both among the most important reactions to CAT production (they rank 5th and 10th, respectively).
@@
 However, when both were perturbed, their combined effect on the model was low, as the respective contributions to consumption and production of glutamate cancelled.
-The system state as a whole was most sensitive to glucose uptake via GTP and the forward reaction of lactate dehydrogenase (Fig.~\ref{fig:Sensitivity}, bottom, section F).
-The 30 next most important reactions to the system state (section G) came from various pathways across the network: eight from amino acid synthesis/degradation; six from glycolysis; four from the TCA cycle; and two each from the pentose phosphate pathway, Entner-Doudoroff, energy/reducing pathways, and small molecule transport; one from oxidative phosphorylation; and one from pyrophosphatase consumption.
-The system state had more pairwise sensitivities that differ from the corresponding first-order sensitivities and stand out as significant.
+
+The system state was also most dependent on coenzymes and substrates; however, instead of pyruvate and glutamate, the substrates driving metabolism were glucose and pyruvate.
+The metabolism as a whole was most sensitive to glucose uptake via GTP and the forward reaction of lactate dehydrogenase, consuming pyruvate (Fig.~\ref{fig:Sensitivity}, bottom, section F).
+These two and the next 30 most important reactions (section G) largely involved conenzymes, especially ATP and NADPH, as well as substrate-consuming reactions and oxidative phosphorylation.
+The system state had even more pairwise sensitivities that differed from the corresponding first-order sensitivities and stood out as significant.
 For example, the first-order effect of alanine synthesis was large; it consumes both pyruvate and glutamate, two key species in the network.
-In addition, a handful of alanine synthesis pairwise sensitivities were also large.
+%In addition, a handful of alanine synthesis pairwise sensitivities were also large.
 However, there were enough reactions that, when paired with alanine synthesis, had little effect on the model; malic enzyme is one of these, as it produces the pyruvate that alanine synthesis consumes.
@@
 therefore, the perturbation of these reactions had no effect on the model.
-Thus, all pairwise sensitivities with reactions in section J were pseudo first-order sensitivities for the other reactions.
+As a result, all pairwise sensitivities with reactions in section J were pseudo first-order sensitivities for the other reactions.
 Interestingly, many reactions in section F and several in section G showed their highest sensitivities when paired with the "non-effects" of section J.
@@
 This suggested that these reactions' effects on the model were canceled out or lessened by most other reactions, but were of course not affected by the reactions in section J.
-This was also likely the reason that reactions in section J rank above those in section K, despite having no effect themselves on the model.
+This was also likely the reason that reactions in section J rank above those in section K, despite having no effect on the model themselves.
 Taken together, sensitivity analysis identified blocks of parameters that either individually, or in combination influenced model performance.
-However, the sensitivity analysis did not establish what the maximum performance of the system was.
-To answer that question we performed sequence-specific flux balance analysis of CAT production.
 
-In comparing the flux distributions between the unconstrained and constrained cases (Fig.~\ref{fig:Network}), the constrained cases heavily utilized the first step in the pentose phosphate pathway to generate NADPH.
-In these cases the majority of the flux continued through the Entner–Doudoroff pathway, whereas in the unconstrained case the majority of flux traveled through glycolysis.
-In all cases, the energy source came primarily from oxidative phosphorylation, as well as partly from the TCA cycle.
-In the transcription/translation constrained case, there was a high flux through fumerate dehydrogenase from aspartic acid uptake, whereas in the unconstrained and most constrained cases, acetate and lactate accumulation occurred.
-This shows that the system is producing NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to supply enough NADH for oxidative phosphorylation.
-As a result, high oxidative phosphorylation activity relative to our cell-free system leads to an acetate overflow.
-This suggests that there is potential for increasing CAT production by reducing this diversion of carbon.
-To simulate potential knockouts, we constrained the specific glucose and amino acid uptake rates to the same values as simulated with no knockouts.
-In an ssFBA simulation with constrained transcription/translation rates, knocking out the \textit{gnd} reaction decreases flux of acetate production but increases flux through \textit{pntAB}, which is responsible for regenerating NADPH.
+The sensitivity analysis showed oxidative phosphorylation was significant for CAT production as well as the system state.
+To further investigate this, we knocked out key reactions in oxidative phosphorylation to examine its effect on glucose uptake and CAT production (Fig.~\ref{fig:oxphos_ko}).
+A single knockout of the reaction \textit{cyd} was detrimental to model performance and CAT production, reducing the CAT carbon yield from 8.6\% to 2.8\%.
+In addition, the glucose uptake rate was reduced compared to that of the control (no knockouts).
+A knockout of \textit{nuo} showed a less drastic effect, reducing the CAT carbon yield to 6.8\%; however, the glucose uptake rate remained similar to that of the control.
+Knocking out \textit{app} showed a CAT yield to 8.8\%, but was not statistically different from that of the control.
+With all three reactions knocked out, CAT yield was 2.7\%, not statistically different from the \textit{cyd} knockout.
+Thus, the model suggests that the majority of the energetic needs of the system as well as the production of CAT are met by oxidative phosphorylation, specifically by \textit{cyd}.
+%The \textit{cyd} enzyme is responsible for transferring electrons with oxygen as an electron acceptor allowing the release of energy to form ATP.
+
+To investigate the difference in yields between the unconstrained and constrained cases, we compared the flux distributions from the ssFBA simulations (Fig.~\ref{fig:Network}).
+The constrained cases (ii \& iii) heavily utilized the first step in the pentose phosphate pathway to generate NADPH; the carbon flux continued through the Entner–Doudoroff pathway toward pyruvate, a key metabolite as shown by sensitivity analysis.
+For case ii, the majority of the flux proceeded toward acetate accumulation, whereas in case iii, the flux accumulated as pyruvate, lactate, and acetate with some going through the TCA cycle.
+In comparison, the unconstrained case (i) showed the majority of flux traveling through glycolysis towards pyruvate, leading to an accumulation of lactate, acetate and malate.
+%In the second case (ii), there was a high flux through fumarate dehydrogenase from aspartic acid uptake, whereas in cases i and iii, acetate and lactate accumulation occurred.
+In all cases the energy source was primarily oxidative phosphorylation, and to a lesser extent the TCA cycle.
+%However, oxidative phosphorylation is not operating at its highest efficiency, since there is a mixture of acetate and lactate accumulation.
+However, the accumulation of acetate and lactate signifies that the system is not operating at its highest efficiency.
+The system produces NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to power oxidative phosphorylation.
+Oxidative phosphorylation leads to a high redox ratio contributing to the accumulation of acetate overflow and diverting flux away from the TCA cycle.
+This suggests that there is potential to increase CAT production by reducing the accumulation of acetate and lactate.
+To investigate this further, we simulated potential knockouts with constrained transcription/translation rates, but constrained the specific glucose and amino acid uptake rates to the same values as simulated with no knockouts.
+Knocking out the \textit{gnd} reaction decreased flux of acetate production but increased flux through \textit{pntAB}, which is responsible for regenerating NADPH.
 The simulation showed carbon was diverted toward lactate; however, since CAT production is constrained by the translation rate, we expected no increase in CAT production.
-The decrease in acetate production is promising as a mechanism to increase CAT yield.
 A second simulation with a knockout of \textit{gnd} and phosphate acetyltransferase showed carbon being diverted toward lactate and succinate; however, it required a higher flux through oxidative phosphorylation and the TCA cycle to meet the energetic needs of the system.
+The decrease in acetate production and the diversion of carbon flux shows a promising mechanism to increase CAT yield.
 
@@
 We used the hybrid cell-free modeling approach of Wayman and coworkers, which integrates traditional kinetic modeling with a logic-based description of allosteric regulation.
+CFPS is seen to be biphasic relying on glucose during the first hour and pyruvate and lactate afterward.
+Allosteric control was essential to the maintainence of the network and production of CAT, as without it, central carbon metabolism is exhuasted within 1.5 hours leading to low CAT production.
+Having captured the experimental data, we investigated if CAT yield and CFPS performance could be further improved.
 We showed that the model produces CAT at 23\% of the theoretical maximum in terms of carbon yield, and at 36\% of a physiological case in which transcription and translation are constrained.
-The theoretical maximum and transcription/translation constrained case were obtained using FBA, which predicted a different flux distribution for each case.
-The unconstrained case predicted most of the carbon flux going through glycolysis, while the constrained cases predicted significant flux through the Entner-Doudoroff pathway and some through TCA cycle.
-However, all cases relied on oxidative phosphorylation for the system's energetic needs as well as for tRNA charging for CAT synthesis.
-Sensitivity analysis of the dynamic model suggested that both CAT production and the entire metabolic network were most sensitive to amino acid synthesis and degradation reactions, and reactions in glycolysis and the TCA cycle.
-CAT production was also very sensitive to the CAT synthesis reaction, unsurprisingly.
-The allosteric control component of the hybrid modeling approach was shown as important to central carbon metabolism, but not very important to CAT production.
-Taken together, this is the first dynamic model of \textit{E. coli} cell-free protein synthesis, and an important step toward a functional genome scale description.
+The accumulation of waste byproducts, especially acetate, is responsible for this sub-optimal yield.
+Sensitivity analysis showed that certain substrates and energy species are instrumental to CAT production and overall metabolism.
+The system heavily relied on oxidative phosphorylation for the system's energetic needs as well as for CAT synthesis.
+A single knockout in oxidative phosphorylation reduced the CAT carbon yield \texttildelow 3-fold, as well as disrupting the system state showing its crucial role in CFPS.
+In comparing flux distributions between low and high yield cases, carbon flux could be potentially diverted toward CAT by reducing acetate overflow and minimizing flux through the Entner-Doudoroff pathway.
+Taken together, these findings represent the first dynamic model of \textit{E. coli} cell-free protein synthesis, and an important step toward a functional genome scale description.
 
@@
 This is important in applying metabolic engineering principles to rationally design cell-free production processes and predict the redirection of carbon fluxes to product forming pathways.
-The most sensitive parameter for the model as a whole is the uptake of glucose, followed by pyruvate.
+In analyzing the model parameters' effect on CAT production, CAT synthesis is the most important, followed by oxidative phosphorylation and the glutamate and pyruvate consuming reactions, as well as coenzyme reactions which are necessary to drive CAT synthesis.
+For example, the conversion of ATP to GTP shows significance since it is necessary for CAT synthesis.
+While Jewett and coworkers have shown that ATP may be at saturation in CFPS \cite{2008_jewett_molsysbiol}, GTP is also required for CAT synthesis and may be a limiting reactant.
+Thus, supplementation with additional GTP may improve the efficiency of CAT production.
+A similar theme is seen in the sensitivity of overall model state, where the most important reactions are glucose and pyruvate consuming reactions and coenzyme reactions which are vital to drive CFPS.
+This can be seen in the biphasic operation of CFPS, with the first phase operating on glucose and the second phase operating on pyruvate.
+During the first phase, there is an accumulation of byproducts from central carbon with the majority of flux going toward acetate and some toward pyruvate, lactate, and succinate; with the exception of acetate, these are all consumed in the second phase.
+This shows that CAT production can be sustained by pyruvate and glutamate in the absence of glucose, which provides alternative strategies to optimize CFPS performance.
+This is in accordance with literature, which showed pyruvate provided a relatively slow but continuous supply of ATP \cite{swartz_nature2001}.
+Taken together, this shows CFPS can be designed towards a specified application either requiring a slow stable energy source or faster production.
 This outstanding control on model performace was expected as these metabolites are responsible for driving CFPS and represent the first step in the model network.
-Nevertheless, there are further reactions that excercise considerable sensitivity to model performance. 
-Oxidative phosphorylation activity is vital, since it provides most of the energetic needs of CFPS.
-In examining oxidative phosphorylation activity, knockouts in the electron transport pathways disrupt metabolism across the network and show CAT carbon yield dropping from 8.6\% to 2.7\%, consistent with literature observations \cite{2008_jewett_molsysbiol}.
-Jewett and coworkers also saw a decrease in CAT yield, ranging from 1.5-fold to 4-fold, when knocking out oxidative phosphorylation.
-Other reactions with considerable sensitivity are those responsible for the consumption of glucose and the consumption and production of pyruvate, which is consumed during the second phase of CFPS. 
-Also important is the first step in pentose phosphate pathway, as it converts G6P to 6PG and produces NADPH.
-The next most sensitive reaction is the conversion of NADPH to NADH, which helps to supply NADH to oxidative phosphorylation reactions to meet the energetic needs of the system. 
-Many of the insignificant reactions lead to the accumulation of unwanted byproducts such as ethanol and formate.
-
-In analyzing the model parameters' effect on CAT production, CAT synthesis is the most important, followed by oxidative phosphorylation and the glucose and pyruvate consuming reactions, as well as the downstream reactions which are necessary to drive CFPS.
-In addition, the conversion of ATP to GTP shows significance since it facilitates CAT synthesis.
-Thus, supplementation with additional GTP may improve the efficiency of CAT production.
-A similar theme is seen in the sensitivity of overall model performace; the most important reactions are glucose and pyruvate consuming reactions and downstream reactions.
-The overall sensitivity results show that the substrates driving CFPS are vital to system performance and CAT production.
-This can be seen in the biphasic operation of CFPS, with the first phase operating on glucose and the second phase operating on pyruvate. 
-During the first phase, there is an accumulation of byproducts from central carbon with the majority of flux going toward acetate and some toward pyruvate, lactate, and succinate, but with the exception of acetate these are all consumed in the second phase.
-CFPS operating under glucose consumption has higher CAT production than that of pyruvate/lactate consumption, as can be seen in the experimental data (\texttildelow 10 $\mu$M/h vs \texttildelow 5 $\mu$M/h).
-
+Nevertheless, there are further reactions with considerable impact on model performance.
+In examining oxidative phosphorylation activity, knockouts in the electron transport pathways disrupt metabolism across the network and show CAT carbon yield dropping from 8.6\% to 2.7\%; Jewett and coworkers also saw a decrease in CAT yield, ranging from 1.5-fold to 4-fold, when knocking out oxidative phosphorylation reactions\cite{2008_jewett_molsysbiol}.
+Oxidative phosphorylation is vital, since it provides most of the energetic needs of CFPS.
+However, it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, and both of our modeling approaches suggest its importance to CAT production and CFPS.
+Thus, oxidative phosphorylation is a potential area for improvement for CFPS performance and protein yield.
+%Also important is the first step in pentose phosphate pathway, as it converts G6P to 6PG and produces NADPH.
+%The next most sensitive reaction is the conversion of NADPH to NADH, which helps to supply NADH to oxidative phosphorylation reactions.
+%Many of the insignificant reactions lead to the accumulation of unwanted byproducts such as ethanol and formate.
 Comparing the theoretical maximum carbon yield of CAT from ssFBA predictions to those of the kinetic model and experimental measurements suggests that there is potential for increasing CAT yield as well as CFPS performance.
 %Yield performance can be potentially improved by diverting carbon flux away from acetate.
-The theoretical maximum CAT yield was 36\% for an unconstrained case and 23\% for a transcription/translation constrained case.
+The model and experimental yields were 36\% of the theoretical maximum and 23\% of a physiologically constrained case.
 Knockouts of \textit{gnd} and phosphate acetyltransferase show that carbon can be diverted away from acetate and potentially toward CAT or other proteins of interest expressed in CFPS.
@@
 Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
-In addition to improving CFPS performance, Jewett and coworkers have shown that oxidative phosphorylation operates in cell-free systems, and that knocking out these reactions is detrimental to protein yield \cite{2008_jewett_molsysbiol}.
-However, it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, and both of the modeling approaches we present suggest its importance to CAT production.
-Thus, oxidative phosphorylation is a potential area for improvement of CFPS performance and protein yield.
 
@@
 \subsection*{Estimation of kinetic model parameters.}
-We generated an ensemble of diverse parameter sets using a constrained Markov Chain Monte Carlo (MCMC) random walk strategy.
+We estimated an ensemble of diverse parameter sets using a constrained Markov Chain Monte Carlo (MCMC) random walk strategy.
 Starting from a single best fit parameter set estimated by inspection and literature,
@@
 For each newly generated parameter set, we re-solved the balance equations and calculated the cost function.
-All sets with a lower cost than the previous set, and some with higher cost, were added to the ensemble.
-After generating 12,437 sets, we selected 100 sets with minimal set to set correlation to avoid over-sampling any region of parameter space.
-The original 12,437-set ensemble had a mean Pearson correlation coefficient of 0.94 between pairs of sets, while the 100-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.
+All sets with a lower cost (and some with higher cost) were accepted into the ensemble.
+After generating greater that 10,000 sets, we selected N = 100 sets with minimal set to set correlation to avoid over-sampling any region of parameter space.
+% The original 12,437-set ensemble had a mean Pearson correlation coefficient of 0.94 between pairs of sets, while the 100-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.
 
-\subsection*{Sensitivity analysis of the CFPS model.}
+\subsection*{Sensitivity analysis of the kinetic CFPS model.}
 We determined the reactions most important to protein production by computing the local sensitivity of CAT concentration (denoted as $\texttt{CAT}$) to each individual maximum reaction rate, and each pair of maximum reaction rates in the network.
@@
 In calculating the pairwise sensitivities, each parameter was perturbed by 1\%; first-order sensitivities ($i$ = $j$) were subject to two 1\% perturbations.
+Parameters and parameter combinations were stratified into five degrees of importance, from least to most sensitive.
+% This was achieved by assigning them a value of 0, 0.25, 0.5, 0.75, or 1.
 % The array of sensitivity indices was organized into a clustergram with both axes corresponding to the 163 maximum reaction rates
@@
 
-Likewise, we determined the reactions most important to global system performance by computing the sensitivity of all species for which data exists (denoted as $\texttt{X}$)
+Likewise, we determined which reactions were most important to global system performance by computing the sensitivity of all species for which data exists (denoted as $\texttt{X}$)
 to each maximum reaction rate in the network.
@@
 \begin{equation}\label{eqn:system-sensitivity}
-    \mathcal{S}_{ij}^{\texttt{X}}=\norm{\texttt{X}(p_i,p_j,t)-\texttt{X}(\alpha \cdot p_i,\alpha \cdot p_j,t)}\qquad{i,j=1,2,\hdots\mathcal{P}}
+    \mathcal{S}_{ij}^{\texttt{X}}=\norm{\texttt{X}(p_i,p_j,t)-\texttt{X}(\alpha \cdot p_i,\alpha \cdot p_j,t)}_{2}\qquad{i,j=1,2,\hdots\mathcal{P}}
 \end{equation}
@@
 which data exists as a function of time and the $i^{th}$ and $j^{th}$ parameters.
+The parameter sensitivities were stratified into five degrees of importance, from least to most sensitive, as above.
 
 % This array of sensitivity indices was also organized into a clustergram (Fig.~\ref{fig:Sensitivity}, bottom).
+% Theoretical maximum CAT carbon yields for three cases discussed below were calculated using flux balance anaylsis (FBA) with a sequence-specific description of CAT synthesis, where $\Delta m_{i}$ denotes the flux of the $i^{th}$ species.
+% The transcription and translation template reactions come from sequence-specific analysis \cite{Allen:2003aa}, and include transcription initiation, transcription, mRNA degradation, translation initiation, translation, and tRNA charging.
+% The rest of the network followed a pseudo steady-state assumption where metabolites were not allowed to accumulate; thus, the network could be solved by linear programming instead of solving differential equations.
+% \begin{eqnarray*}
+% G+R_{1} &\longrightarrow& G^{*}\\
+% G^{*}+151 \cdot gtp+144 \cdot ctp+189 \cdot utp+176 \cdot atp &\longrightarrow& mRNA+G+R_{1}+1320 \cdot Pi\\
+% mRNA &\longrightarrow& 151 \cdot gmp+144 \cdot cmp+189 \cdot ump+176 \cdot amp\\
+% \end{eqnarray*}
 
-\subsection*{Sequence-specific FBA and calculation of CAT yield}
-The yield on CAT production was calculated for each case as a ratio of carbon produced as CAT to carbon consumed as reactants (glucose and amino acids):
+\subsection*{Sequence specific calculation of carbon yield.}
+We estimated the theoretical maximum carbon yield of CAT using sequence-specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
+The CAT carbon yield ($Y_{C}^{CAT}$) was calculated as the ratio of carbon produced as CAT dived by the carbon consumed as reactants (glucose and amino acids):
 \begin{equation}\label{eqn:yield-definition}
-	Yield=\frac{\Delta m_{CAT}\;C_{CAT}}{\sum_{i=1}^{\mathcal{R}}\max(\Delta m_{i},0)\;C_{m_i}}
+	Y_{C}^{CAT}=\frac{\Delta\texttt{CAT}\cdot C_{CAT}}{\displaystyle\sum_{i=1}^{\mathcal{R}}\max(\Delta m_{i},0)\cdot C_{m_i}}
 \end{equation}
-where $\Delta m_{CAT}$ denotes the amount of CAT produced, $C_{CAT}$ denotes carbon number of CAT, $\mathcal{R}$ denotes the number of reactants, $\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed, never allowed to be negative, and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
-Because no data was available for arginine or glutamate, these reactants were left out of all yield calculations.
-Yield of the best-fit parameter set and the experimental data were calculated by setting $\Delta CAT$ equal to the final minus the initial CAT concentration and setting $\Delta m_{i}$ equal to the initial minus the final reactant concentration.
-Theoretical maximum CAT carbon yields for three cases discussed below were calculated using flux balance anaylsis (FBA) with a sequence-specific description of CAT synthesis, where $\Delta m_{i}$ denotes the flux of the $i^{th}$ species.
-This sequence-specific FBA \citep{Allen:2003aa} problem was formulated as:
+where $\Delta\texttt{CAT}$ denotes the abundance of CAT produced, $C_{CAT}$ denotes carbon number of CAT, $\mathcal{R}$ denotes the number of reactants,
+$\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed (never allowed to be negative), and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
+Arginine or glutamate were not considered in the yield calculations, as no experimental measurements were available for these amino acids.
+Yield of the best-fit parameter set and the experimental data were calculated by setting $\Delta\texttt{CAT}$ equal to the final minus the initial CAT concentration,
+and setting $\Delta m_{i}$ equal to the initial minus the final reactant concentration.
+
+The sequence specific flux balance analysis problem was formulated as:
 \begin{equation}
@@
 and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
-The objective $w_{obj}$ was to maximize the specific rate of CAT formation.
-The specific glucose uptake rate was constrained to allow a maximum flux of 40 mM/h according to experimental data; the specific amino acid uptake rates were bound to allow a maximum flux of 30 mM/h, but did not reach this maximum flux.
-The transcription and translation template reactions come from sequence-specific analysis \cite{Allen:2003aa}, and include transcription initiation, transcription, mRNA degradation, translation initiation, translation, and tRNA charging.
-The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
-The solution flux vector was used to calculate the carbon yield of CAT for the three FBA cases.
-Glucose, oxygen, and amino acids were modeled as being imported into the system, while CAT synthesis and metabolite byproduct formation were modeled as export from the system.
-The rest of the network followed a pseudo steady-state assumption where metabolites were not allowed to accumulate; thus, the network could be solved by linear programming instead of solving differential equations.
+The objective was to maximize the CAT translation rate.
+The glucose uptake rate was bounded in the range [0,40 mM/h] according to experimental data;
+while the amino acid uptake rates were bounded by [0,30 mM/h], but did not reach the maximum flux.
+The transcription and translation rates were modeled using template reactions
 
-The transcription rate was formulated as:
+The transcription rate ($w_{TX}$) was fixed in the ssFBA calculation as:
 \begin{equation}
-	w_{TX} = RNAP\left(\frac{v_{RNAP}}{l_{mRNA}}\right)\left(\frac{Gene}{k_m+Gene}\right)P
+	w_{TX} = \left[R_{1}\left(\frac{K_{T7}}{1 + K_{T7}}\right)\left(\frac{v_{RNAP}}{l_{mRNA}}\right)\right]\left(\frac{G}{K_{TX}+G}\right)
 \end{equation}
-where $w_{TX}$ denotes the transcription rate, $RNAP$ denotes the concentration of RNA polymerase, $v_{RNAP}$ denotes the elongation rate by the RNA polymerase in nucleotides per hour, $l_{mRNA}$ denotes the mRNA length in nucleotides, $Gene$ denotes the gene concentration, $k_m$ denotes the plasmid saturation coefficient, and $P$ denotes the promoter activity.
-The mRNA and protein sequence of CAT was determined from literature.
-The promoter activity was formulated following Moon et al. for synthetic circuits as:
-\begin{equation}
-	P = \frac{K_{1}}{1 + K_{1}}
-\end{equation}
-where $K_{1}$ denotes the state of T7 RNA polymerase binding.
+where $R_{1}$ denotes the concentration of RNA polymerase, $v_{TX}$ denotes RNA polymerase elongation rate (nt/hr),
+$l_{G}$ denotes the gene length in nucleotides, $G$ denotes the gene concentration and $K_{TX}$ denotes the plasmid saturation coefficient.
+The gene and protein sequence for CAT was determined from literature.
+The last term in the $w_{TX}$ expression describes T7 promoter activity, where $K_{T7}$ quantifies T7 RNA polymerase binding [REFHERE].
 The translation rate was formulated as:
@@
 This included the data for glucose, organic acids, energy species, and amino acids; CAT was not constrained by experimental data, but by the transcription/translation rates as stated above.
+The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
 
+% The solution flux vector was used to calculate the carbon yield of CAT for the three FBA cases.
+% Glucose, oxygen, and amino acids were modeled as being imported into the system, while CAT synthesis and metabolite byproduct formation were modeled as export from the system.
 %Need to talk more about biochemical benefits and importance for biochemical problems
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/CarbonBoth_png.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/allostericControl_noindex.pdf}
 \caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species.
@@
 \centering
-\includegraphics[width=0.95\textwidth]{./Figures/CAT_prod_yield.pdf}
+\includegraphics[width=1\textwidth]{./Figures/oxphos_ko.pdf}
+\caption{The effect of oxidative phosphorylation on glucose uptake, CAT production and CAT carbon yield.
+A.
+95\% confidence interval of an ensemble for glucose concentration versus time for no knockouts (blue shaded region), \textit{cyd} knockout (green shaded region), and \textit{nuo} knockout (red shaded region).
+B.
+95\% confidence interval of an ensemble for CAT concentration versus time for no knockouts (blue shaded region), \textit{cyd} knockout (green shaded region), and \textit{nuo} knockout (red shaded region).
+C.
+CAT carbon yield for 5 different cases of oxidative phosphorylation: no knockouts (blue), \textit{cyd} knockout (green), \textit{nuo} knockout (red), \textit{app} knockout (light grey), and a combination of \textit{cyd}, \textit{nuo}, \textit{app} knockouts (dark grey).}
+\label{fig:oxphos_ko}
+\end{figure}
+\clearpage
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1\textwidth]{./Figures/CAT_prod_yield.pdf}
 \caption{Sequence-specific flux balance analysis of CAT production and yield.
@@
 \end{figure}
+\clearpage
 

### 2016-11-21 Paper_v1.tex
some changes on materials and methods

@@
 \begin{equation}
-	w_{TX} = \left[R_{1}\left(\frac{K_{T7}}{1 + K_{T7}}\right)\left(\frac{v_{RNAP}}{l_{mRNA}}\right)\right]\left(\frac{G}{K_{TX}+G}\right)
+	w_{TX} = \left[R_{1}\left(\frac{K_{T7}}{1 + K_{T7}}\right)\left(\frac{v_{TX}}{l_{G}}\right)\right]\left(\frac{G}{K_{TX}+G}\right)
 \end{equation}
@@
 The last term in the $w_{TX}$ expression describes T7 promoter activity, where $K_{T7}$ quantifies T7 RNA polymerase binding [REFHERE].
-The translation rate was formulated as:
+The translation rate ($w_{TL}$) was bounded by:
  \begin{equation}
-	w_{TL} = K_{P}\ Ribo\left(\frac{v_{Ribo}}{l_{Protein}}\right)mRNA_{SS}
+	0\leq w_{TL} \leq \left[K_{P} R_{2}\left(\frac{v_{TL}}{l_{P}}\right)\right]\left(\frac{mRNA_{SS}}{K_{TL}+mRNA_{SS}}\right)
 \end{equation}
-where $K_{P}$ denotes the polysome amplification constant, $Ribo$ denotes the ribosome concentration, $v_{Ribo}$ denotes the elongation rate of the ribosome in amino acids per hour, $l_{Protein}$ denotes the number of amino acids in the protein of interest, and $mRNA_{SS}$ denotes the mRNA concentration at steady state, equal to the transcription rate divided by the degradation rate of mRNA.
+where $K_{P}$ denotes the polysome amplification constant, $R_{2}$ denotes the ribosome concentration, $v_{TL}$ denotes the ribosome elongation rate (amino acids per hour),
+$l_{P}$ denotes the number of amino acids in the protein of interest, and $mRNA_{SS}$ denotes the steady-state mRNA concentration:
+\begin{equation}
+	mRNA_{SS}\simeq\frac{w_{TX}}{\lambda}
+\end{equation}where $\lambda$ denotes the mRNA degradation rate.
 

### 2016-11-21 Paper_v1.tex
more changes in ssFBA MM

@@
 Maximum reaction rates were bounded between 0 and 500,000 mM/h \cite{Keseler:2013aa}.
-Assuming a total enzyme concentration of 50 $\mu$M, this corresponds to catalytic rate bounds of 0 and 2778 s$^{-1}$.
-These bounds resulted in a median catalytic rate of 0.016 s$^{-1}$ across the ensemble.
-Enzyme activity decay constants were bounded between 0 and 1 h$^{-1}$, corresponding to half lives of 42 minutes and infinity; median = 25 min.
+Assuming a total enzyme concentration of 5.0 $\mu$M, this corresponds to catalytic rate bounds of 0 and 27,780 s$^{-1}$.
+These bounds resulted in a median catalytic rate of 0.16 s$^{-1}$ across the ensemble.
+Enzyme activity decay constants were bounded between 0 and 1 h$^{-1}$, corresponding to half lives of 42 minutes and infinity; median = 25 h.
 Saturation constants were bounded between 0.001 and 10 mM; median = 0.16 mM.
@@
 \subsection*{Sequence specific calculation of carbon yield.}
-We estimated the theoretical maximum carbon yield of CAT using sequence-specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
+We estimated the theoretical maximum CAT carbon yield using sequence-specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
+The sequence specific flux balance analysis problem was formulated as:
+\begin{equation}
+ \begin{multlined}
+	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{TL} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
+	\mathrm{Subject \; to:}
+	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
+\alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
+ \end{multlined}
+\end{equation}
+where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
+and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
+The glucose uptake rate was bounded in the range [0,40 mM/h] according to experimental data;
+while the amino acid uptake rates were bounded by [0,30 mM/h], but did not reach the maximum flux.
+The transcription and translation rates were modeled using the template reactions: [FILL IN OR REFERENCE TEMPLATE TABLE].
+The transcription rate ($w_{TX}$) was fixed in the ssFBA calculation as:
+\begin{equation}
+	w_{TX} = \left[R_{1}\left(\frac{v_{TX}}{l_{G}}\right)\left(\frac{K_{T7}}{1 + K_{T7}}\right)\right]\left(\frac{G}{K_{TX}+G}\right)
+\end{equation}
+The first term in the brackets denotes the maximum rate of transcription, where $R_{1}$ denotes the concentration of RNA polymerase, $v_{TX}$ denotes the RNA polymerase elongation rate (nt/hr), $l_{G}$ denotes the gene length in nucleotides, and the last term describes T7 promoter activity, where $K_{T7}$ quantifies T7 RNA polymerase binding [REF-HERE].
+The second term denotes the gene saturation term, where $G$ denotes the gene concentration and $K_{TX}$ denotes a saturation coefficient.
+The CAT gene and protein sequences were determined from literature.
+The translation rate ($w_{TL}$) was bounded by:
+ \begin{equation}
+	0\leq w_{TL} \leq \left[K_{P} R_{2}\left(\frac{v_{TL}}{l_{P}}\right)\right]\left(\frac{mRNA_{SS}}{K_{TL}+mRNA_{SS}}\right)
+\end{equation}
+The first term in brackets denotes the maximum translation rate, where $K_{P}$ denotes the polysome amplification constant, $R_{2}$ denotes the ribosome concentration, $v_{TL}$ denotes the ribosome elongation rate (amino acids per hour), $l_{P}$ denotes the number of amino acids in the protein of interest, and $mRNA_{SS}$ denotes the steady-state mRNA concentration:
+\begin{equation}
+	mRNA_{SS}\simeq\frac{w_{TX}}{\lambda}
+\end{equation}where $\lambda$ denotes the rate constant controlling the mRNA degradation rate.
+
 The CAT carbon yield ($Y_{C}^{CAT}$) was calculated as the ratio of carbon produced as CAT dived by the carbon consumed as reactants (glucose and amino acids):
@@
 and setting $\Delta m_{i}$ equal to the initial minus the final reactant concentration.
-
-The sequence specific flux balance analysis problem was formulated as:
-\begin{equation}
- \begin{multlined}
-	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{obj} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
-	\mathrm{Subject \; to:}
-	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
-\alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
- \end{multlined}
-\end{equation}
-where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
-and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
-The objective was to maximize the CAT translation rate.
-The glucose uptake rate was bounded in the range [0,40 mM/h] according to experimental data;
-while the amino acid uptake rates were bounded by [0,30 mM/h], but did not reach the maximum flux.
-The transcription and translation rates were modeled using template reactions
-
-The transcription rate ($w_{TX}$) was fixed in the ssFBA calculation as:
-\begin{equation}
-	w_{TX} = \left[R_{1}\left(\frac{K_{T7}}{1 + K_{T7}}\right)\left(\frac{v_{TX}}{l_{G}}\right)\right]\left(\frac{G}{K_{TX}+G}\right)
-\end{equation}
-where $R_{1}$ denotes the concentration of RNA polymerase, $v_{TX}$ denotes RNA polymerase elongation rate (nt/hr),
-$l_{G}$ denotes the gene length in nucleotides, $G$ denotes the gene concentration and $K_{TX}$ denotes the plasmid saturation coefficient.
-The gene and protein sequence for CAT was determined from literature.
-The last term in the $w_{TX}$ expression describes T7 promoter activity, where $K_{T7}$ quantifies T7 RNA polymerase binding [REFHERE].
-The translation rate ($w_{TL}$) was bounded by:
- \begin{equation}
-	0\leq w_{TL} \leq \left[K_{P} R_{2}\left(\frac{v_{TL}}{l_{P}}\right)\right]\left(\frac{mRNA_{SS}}{K_{TL}+mRNA_{SS}}\right)
-\end{equation}
-where $K_{P}$ denotes the polysome amplification constant, $R_{2}$ denotes the ribosome concentration, $v_{TL}$ denotes the ribosome elongation rate (amino acids per hour),
-$l_{P}$ denotes the number of amino acids in the protein of interest, and $mRNA_{SS}$ denotes the steady-state mRNA concentration:
-\begin{equation}
-	mRNA_{SS}\simeq\frac{w_{TX}}{\lambda}
-\end{equation}where $\lambda$ denotes the mRNA degradation rate.
 

### 2016-11-21 Paper_v1.tex
Template Reactions add

@@
 while the amino acid uptake rates were bounded by [0,30 mM/h], but did not reach the maximum flux.
-The transcription and translation rates were modeled using the template reactions: [FILL IN OR REFERENCE TEMPLATE TABLE].
+The transcription and translation rates were modeled using the template reactions:
+
+\begin{eqnarray*}
+G_{p}+R_{1} &\longrightarrow& G_{p}^{*} \\
+G_{p}^{*}+\sum_{i}\eta_{i}iTP &\longrightarrow& mRNA+G_{p}+R_{1}+2\sum_{i}\eta_{i}P \qquad{i=A,C,G,U}\\
+mRNA &\longrightarrow& \sum_{i}\eta_{i}iMP \\
+mRNA+R_{2} &\longrightarrow& R_{2}^{*} \\
+\alpha_{j}X_{j}+\alpha_{j}tRNA+\alpha_{j}ATP &\longrightarrow& \alpha_{j}XtRNA_{j}+\alpha_{j}AMP+2\alpha_{j}P \qquad{j=1,2,\hdots,20}\\
+R_{2}^{*}+\sum_{j}\alpha_{j}\Big(XtRNA_{j}+2GTP\Big) &\longrightarrow& Protein+R_{2}+mRNA\\
+&& \quad +\sum_{j}\alpha_{j}\Big(tRNA+2GDP+2P\Big)
+\end{eqnarray*}
+where $G_{p}$ denotes the gene of protein $p$, $R_{1}$ denotes the concentration of RNA polymerase, $G_{p}^{*}$ denotes the gene bounded by the RNA polymerase, $\eta_{i}$ and $ \alpha_{j}$ denotes the coefficient for species $i$ and $j$ respectively, $P$ denotes phosphate, $R_{2}$ denotes the ribosome concentration, $R_{2}^{*}$ denotes bounded ribosome, and $X_{j}$ denotes the amino acid species $j$.
+
 The transcription rate ($w_{TX}$) was fixed in the ssFBA calculation as:
@@
 \end{equation}
-The first term in the brackets denotes the maximum rate of transcription, where $R_{1}$ denotes the concentration of RNA polymerase, $v_{TX}$ denotes the RNA polymerase elongation rate (nt/hr), $l_{G}$ denotes the gene length in nucleotides, and the last term describes T7 promoter activity, where $K_{T7}$ quantifies T7 RNA polymerase binding [REF-HERE].
+The first term in the brackets denotes the maximum rate of transcription, $v_{TX}$ denotes the RNA polymerase elongation rate (nt/hr), $l_{G}$ denotes the gene length in nucleotides, and the last term describes T7 promoter activity, where $K_{T7}$ quantifies T7 RNA polymerase binding \cite{Moon:2012aa}.
 The second term denotes the gene saturation term, where $G$ denotes the gene concentration and $K_{TX}$ denotes a saturation coefficient.
@@
 \end{equation}
-The first term in brackets denotes the maximum translation rate, where $K_{P}$ denotes the polysome amplification constant, $R_{2}$ denotes the ribosome concentration, $v_{TL}$ denotes the ribosome elongation rate (amino acids per hour), $l_{P}$ denotes the number of amino acids in the protein of interest, and $mRNA_{SS}$ denotes the steady-state mRNA concentration:
+The first term in brackets denotes the maximum translation rate, where $K_{P}$ denotes the polysome amplification constant, $v_{TL}$ denotes the ribosome elongation rate (amino acids per hour), $l_{P}$ denotes the number of amino acids in the protein of interest, and $mRNA_{SS}$ denotes the steady-state mRNA concentration:
 \begin{equation}

### 2016-11-21 Paper_v1.tex
Some changes in MM

@@
 and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
-The glucose uptake rate was bounded in the range [0,40 mM/h] according to experimental data;
+The objective of the calculation is the maximization of CAT translation rate, $w_{TL}$.
+The total glucose uptake rate was bounded by [0,40 mM/h] according to experimental data;
 while the amino acid uptake rates were bounded by [0,30 mM/h], but did not reach the maximum flux.
 The transcription and translation rates were modeled using the template reactions:
-
 \begin{eqnarray*}
 G_{p}+R_{1} &\longrightarrow& G_{p}^{*} \\
-G_{p}^{*}+\sum_{i}\eta_{i}iTP &\longrightarrow& mRNA+G_{p}+R_{1}+2\sum_{i}\eta_{i}P \qquad{i=A,C,G,U}\\
-mRNA &\longrightarrow& \sum_{i}\eta_{i}iMP \\
+G_{p}^{*}+\sum_{k}\eta_{k}\cdot kTP &\longrightarrow& mRNA+G_{p}+R_{1}+2\sum_{k}\eta_{k}\cdot Pi \qquad{k=A,C,G,U}\\
+mRNA &\longrightarrow& \sum_{k}\eta_{k}\cdot kMP \\
 mRNA+R_{2} &\longrightarrow& R_{2}^{*} \\
-\alpha_{j}X_{j}+\alpha_{j}tRNA+\alpha_{j}ATP &\longrightarrow& \alpha_{j}XtRNA_{j}+\alpha_{j}AMP+2\alpha_{j}P \qquad{j=1,2,\hdots,20}\\
-R_{2}^{*}+\sum_{j}\alpha_{j}\Big(XtRNA_{j}+2GTP\Big) &\longrightarrow& Protein+R_{2}+mRNA\\
-&& \quad +\sum_{j}\alpha_{j}\Big(tRNA+2GDP+2P\Big)
+\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP &\longrightarrow& \alpha_{j}\cdot AA_{j}:tRNA_{j}+\alpha_{j}\cdot AMP+2\alpha_{j}\cdot Pi \qquad{j=1,2,\hdots,20}\\
+R_{2}^{*}+\sum_{j}\alpha_{j}\cdot \Big(AA_{j}:tRNA_{j}+2\cdot GTP\Big) &\longrightarrow& \mathcal{P}+R_{2}+mRNA+\\
+&& \quad +\sum_{j}\alpha_{j}\Big(tRNA+2\cdot GDP+2\cdot Pi\Big)
 \end{eqnarray*}
-where $G_{p}$ denotes the gene of protein $p$, $R_{1}$ denotes the concentration of RNA polymerase, $G_{p}^{*}$ denotes the gene bounded by the RNA polymerase, $\eta_{i}$ and $ \alpha_{j}$ denotes the coefficient for species $i$ and $j$ respectively, $P$ denotes phosphate, $R_{2}$ denotes the ribosome concentration, $R_{2}^{*}$ denotes bounded ribosome, and $X_{j}$ denotes the amino acid species $j$.
+where $G_{p}$ denotes the gene of protein $p$,
+$R_{1}$ denotes the concentration of RNA polymerase,
+$G_{p}^{*}$ denotes the gene bounded by the RNA polymerase,
+$\eta_{i}$ and $ \alpha_{j}$ denote the coefficients for species $i$ and $j$ respectively,
+$P_{i}$ denotes inorganic phosphate, $R_{2}$ denotes the ribosome concentration, $R_{2}^{*}$ denotes bounded ribosome, and $X_{j}$ denotes the amino acid species $j$.
 

### 2016-11-21 Paper_v1.tex
Nov21

@@
 For example, the conversion of ATP to GTP shows significance since it is necessary for CAT synthesis.
-While Jewett and coworkers have shown that ATP may be at saturation in CFPS \cite{2008_jewett_molsysbiol}, GTP is also required for CAT synthesis and may be a limiting reactant.
+While Jewett and coworkers have shown that ATP may be at saturation in CFPS \cite{Jewett:2008aa}, GTP is also required for CAT synthesis and may be a limiting reactant.
 Thus, supplementation with additional GTP may improve the efficiency of CAT production.
@@
 Nevertheless, there are further reactions with considerable impact on model performance.
-In examining oxidative phosphorylation activity, knockouts in the electron transport pathways disrupt metabolism across the network and show CAT carbon yield dropping from 8.6\% to 2.7\%; Jewett and coworkers also saw a decrease in CAT yield, ranging from 1.5-fold to 4-fold, when knocking out oxidative phosphorylation reactions\cite{2008_jewett_molsysbiol}.
+In examining oxidative phosphorylation activity, knockouts in the electron transport pathways disrupt metabolism across the network and show CAT carbon yield dropping from 8.6\% to 2.7\%; Jewett and coworkers also saw a decrease in CAT yield, ranging from 1.5-fold to 4-fold, when knocking out oxidative phosphorylation reactions\cite{Jewett:2008aa}.
 Oxidative phosphorylation is vital, since it provides most of the energetic needs of CFPS.
@@
 where $\mathcal{P}$ denotes the number of parameters ($\mathcal{P}$~=~815), which includes 163 maximum reaction rates ($V^{max}$), 163 enzyme activity decay constants, 455 saturation constants ($K_{js}$), and 34 control parameters, $k_i^{new}$ denotes the new value of the $i^{th}$ parameter, $k_i$ denotes the current value of the $i^{th}$ parameter, $a$ denotes a distribution variance, $r_i$ denotes a random sample from the normal distribution, $l_i$ denotes the lower bound for that parameter type, and $u_i$ denotes the upper bound for that parameter type.
-Maximum reaction rates were bounded between 0 and 500,000 mM/h \cite{Keseler:2013aa}.
+Maximum reaction rates were bounded between 0 and 500,000 mM/h \cite{Milo:2009aa}.
 Assuming a total enzyme concentration of 5.0 $\mu$M, this corresponds to catalytic rate bounds of 0 and 27,780 s$^{-1}$.

### 2016-11-22 Paper_v1.tex
Working my way through the results

@@
 \usepackage{amsmath}
+\usepackage{color}
 %\usepackage{textgreek}
@@
 % While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
+% In the second case (ii), there was a high flux through fumarate dehydrogenase from aspartic acid uptake, whereas in cases i and iii, acetate and lactate accumulation occurred.
+% However, oxidative phosphorylation is not operating at its highest efficiency, since there is a mixture of acetate and lactate accumulation.
+% In addition, a handful of alanine synthesis pairwise sensitivities were also large.
 
@@
 The ensemble of kinetic CFPS models captured the time evolution of CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
-The cell-free \textit{E.~coli} metabolic network was constructed by removing growth associated reactions from the MG1655 reconstruction \cite{Feist121},
-and by adding reactions describing chloramphenicol acetyltransferase (CAT) biosynthesis, a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
+The cell-free \textit{E.~coli} metabolic network was constructed by removing growth associated reactions from the MG1655 reconstruction \cite{Feist:2007aa},
+and by adding reactions describing chloramphenicol acetyltransferase (CAT) biosynthesis, a model protein for which there exists a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
 The CFPS model equations were formulated using the hybrid cell-free modeling framework of Wayman et al. \cite{pr3010138}.
@@
 the same parameter bounds (Fig. \ref{fig:BoxPlot}); thus, we have confidence in the predictive capability of the estimated parameters.
-The model captures the biphasic nature of CFPS shown by the experimental data: in the first hour glucose consumption powers the system, and CAT is produced at \texttildelow 10 $\mu$M/h; subsequently, pyruvate and lactate reserves are consumed to power metabolism, and CAT is produced less efficiently at \texttildelow 5 $\mu$M/h.
-Allosteric control was important to this biphasic operation, as without it, both phases are sped up, and after 1.5 hours CAT synthesis has ceased (Fig.~\ref{fig:CarbonBoth}, bottom).
-In addition, acetate accumulation also stops after 1.5 hours since there is no more substrate to power the system.
-Interestingly, malate captures the experimental measurements during the glucose consumption phase, but increases sharply during the pyruvate consumption phase.
+The model captured the biphasic CAT production: during the first hour glucose powers production, and CAT is produced at \texttildelow 10 $\mu$M/h; subsequently, pyruvate and lactate reserves are consumed to power metabolism, and CAT is produced less efficiently at \texttildelow 5 $\mu$M/h.
+Allosteric control was important to biphasic CAT production; without control, the CAT production rate increased and then ceased after 1.5 hr (Fig.~\ref{fig:CarbonBoth}, bottom).
+In addition, acetate no longer accumulated after 1.5 hours, in the absence of allosteric control.
+Interestingly, the simulated malate abundance tracked the experimental measurements during the glucose consumption phase, but increased sharply during the pyruvate consumption phase,
+without allosteric control. Taken together, we produced an ensemble of kinetic models that was consistent with time series measurements of the production of a model protein.
+However, while the ensemble described the experimental data, it was unclear which kinetic parameters most influenced CAT production, and whether the performance of the CFPS reaction was optimal.
 
-While the model ensemble decribed the experimental data, it was unclear whether the performance of CFPS was optimal.
-To address this question, we used sequence-specific flux balance analysis in combination with a T7 promoter model to estimate the theoretical maximum performance of the system.
-We first validated the ssFBA approach by comparing simulated and measured concentrations of CAT for the first hour under glucose consumption (Fig.~\ref{fig:CATProdYield}A).
-The ssFBA ensemble was able to predict the CAT dataset despite having no adjustable parameters.
-Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates, or the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
-Thus, the metabolic network and molecular description of transcription and translation were consistent with experimental measurements.
-To gauge the performance of CFPS, we next calculated the CAT carbon yield for three classes of constraints: (i) theoretical maximum glucose, amino acid and oxygen upper bounds, and no transcriptional/translational constraints; (ii) theoretical maximum glucose, amino acid and oxygen upper bounds, and realistic transcriptional/translational constraints; and (iii) metabolite fluxes constrained by the data, and realistic transcriptional/translational constraints (Fig.~\ref{fig:CATProdYield}B).
-The unconstrained theoretical maximum CAT carbon yield (i) was 0.363 $\pm$ 0.02 (Fig.~\ref{fig:CATProdYield}B, left);
-On the other hand, for realistic constraints on transcription and translation (ii), the CAT carbon yield was 0.226 $\pm$ 0.03 (Fig.~\ref{fig:CATProdYield}B, middle).
-Lastly, when using realistic metabolite and transcription and translation constraints (iii), the predicted carbon yield was 0.062 $\pm$ 0.02.
-By comparison, the best-fit parameter set had a CAT carbon yield of 0.086 $\pm$ 0.004, equivalent to 23\% of the theoretical maximum (i) and 36\% of the physiological case (ii).
-The experimental dataset had a CAT carbon yield of 0.0821, similar to both the kinetic model and the experimentally constrained case (iii).
+To better understand which parameters and parameter combinations influenced the performance of the kinetic model, we performed sensitivity analysis (Fig.~\ref{fig:Sensitivity}).
+We perturbed each $V^{max}$ parameter, either individually or in pairwise combinations and measured the change in either CAT production or the overall system state.
+CAT production was most dependent upon the abundance of cofactors such as ATP, GTP, NADH and NADPH and the metabolite pyruvate.
+CAT production was most sensitive to the CAT synthesis reaction, oxidative phosphorylation, and the pyruvate-consuming alanine synthesis reaction (Fig.~\ref{fig:Sensitivity}, top, section A).
+Taking into account these three reactions, and the next 16 most important reactions (section B), we saw a common theme of reactions involving the cofactors ATP, GTP, NADH, and NADPH
+as well as the metabolites pyruvate and glutamate.
+Glutamate was important as a precursor for the synthesis of other amino acids required by CAT production.
+Meanwhile, the cofactors provided energy to power CAT synthesis, while pyruvate was important for energy generation following glucose depletion.
+In addition, pyruvate was required for the synthesis of several amino acids.
+The pairwise sensitivities (off-diagonal elements) were different from the corresponding first-order sensitivities (diagonal elements), and led to interesting outcomes.
+The combination of certain reactions had a greater effect on CAT production than that of the individual reactions by themselves.
+For example, glutamine synthesis and arginine degradation were both among the most important reactions to CAT production (they ranked 5th and 10th, respectively).
+This was likely because they both affected the sensitive glutamine-glutamate balance; glutamine synthesis consumes glutamate, while arginine degradation produces it.
+However, when both were perturbed, their combined effect on the model was low, as the respective contributions to consumption and production of glutamate cancelled.
+\textcolor{red}{NICK: ARE THERE EXAMPLES OF POSITIVE SYNERGY?}
 
-To better understand which parameters and parameter combinations influenced model performance we performed sensitivity analysis (Fig.~\ref{fig:Sensitivity}).
-According to the sensitivity results, CAT production was most dependent on coenzymes and pyruvate.
-CAT production was most sensitive to the CAT synthesis reaction, oxidative phosphorylation, and the pyruvate-consuming alanine synthesis reaction (Fig.~\ref{fig:Sensitivity}, top, section A).
-Taking into account these three reactions as well as the next 16 most important reactions (section B), we see a common theme of reactions that involve the coenzymes ATP, GTP, NADH, and NADPH as well as the metabolites pyruvate and glutamate.
-Glutamate is important as a reagant for the synthesis of all other amino acids which are required to assemble the CAT protein.
-Meanwhile, the coenzymes are needed to provide energy to power the synthesis of CAT, and pyruvate becomes important for energy supply after glucose is exhausted.
-In addition, pyruvate is a key metabolite required for the synthesis of several amino acids.
-The pairwise sensitivities (off-diagonal elements) were different from the corresponding first-order sensitivities (diagonal elements), and led to interesting outcomes.
-The combination of certain reactions had a greater effect on CAT production than that of the reactions by themselves.
-For example, glutamine synthesis and arginine degradation were both among the most important reactions to CAT production (they rank 5th and 10th, respectively).
-This was likely because they both affect the sensitive glutamine-glutamate balance; glutamine synthesis consumes glutamate, while arginine degradation produces it.
-However, when both were perturbed, their combined effect on the model was low, as the respective contributions to consumption and production of glutamate cancelled.
-
-The system state was also most dependent on coenzymes and substrates; however, instead of pyruvate and glutamate, the substrates driving metabolism were glucose and pyruvate.
+The overall system state was also sensitive to cofactors and substrates; however, instead of pyruvate and glutamate, the substrates driving metabolism were glucose and pyruvate.
 The metabolism as a whole was most sensitive to glucose uptake via GTP and the forward reaction of lactate dehydrogenase, consuming pyruvate (Fig.~\ref{fig:Sensitivity}, bottom, section F).
-These two and the next 30 most important reactions (section G) largely involved conenzymes, especially ATP and NADPH, as well as substrate-consuming reactions and oxidative phosphorylation.
+These two and the next 30 most important reactions (section G) largely involved cofactors, especially ATP and NADPH, as well as substrate-consuming reactions and oxidative phosphorylation.
 The system state had even more pairwise sensitivities that differed from the corresponding first-order sensitivities and stood out as significant.
-For example, the first-order effect of alanine synthesis was large; it consumes both pyruvate and glutamate, two key species in the network.
-%In addition, a handful of alanine synthesis pairwise sensitivities were also large.
-However, there were enough reactions that, when paired with alanine synthesis, had little effect on the model; malic enzyme is one of these, as it produces the pyruvate that alanine synthesis consumes.
+For example, the first-order effect of alanine synthesis was large; it consumed both pyruvate and glutamate, two key species in the network.
+However, there were enough reactions that, when paired with alanine synthesis, had little effect on the model;
+for example, malic enzyme produces pyruvate that can be consumed by alanine synthesis.
 Thus, the total-order alanine synthesis sensitivity was low, placing it at the very bottom of section I.
@@
 Interestingly, many reactions in section F and several in section G showed their highest sensitivities when paired with the "non-effects" of section J.
+\textcolor{red}{NICK: DOES THIS MAKE SENSE? WHY ARE THESE DIFFERENT THAN THE SINGLE PERTURBATIONS?}
 Of these, three involved pyruvate, strengthening its role as a key metabolite; the others were glucose consumption via GTP/CTP-specific hexokinases,
@@
 This suggested that these reactions' effects on the model were canceled out or lessened by most other reactions, but were of course not affected by the reactions in section J.
-This was also likely the reason that reactions in section J rank above those in section K, despite having no effect on the model themselves.
+This was also likely the reason that section J ranked above section K, despite having no effect on the model themselves.
 Taken together, sensitivity analysis identified blocks of parameters that either individually, or in combination influenced model performance.
 
-The sensitivity analysis showed oxidative phosphorylation was significant for CAT production as well as the system state.
-To further investigate this, we knocked out key reactions in oxidative phosphorylation to examine its effect on glucose uptake and CAT production (Fig.~\ref{fig:oxphos_ko}).
-A single knockout of the reaction \textit{cyd} was detrimental to model performance and CAT production, reducing the CAT carbon yield from 8.6\% to 2.8\%.
+Gene knockouts in the electron transport chain significantly reduced the performance of the CFPS reaction (Fig.~\ref{fig:oxphos_ko}).
+A key finding of both the CAT and overall system state sensitivity analysis was the importance of oxidative phosphorylation.
+To investigate this further, we knocked out key oxidative phosphorylation reactions in the ensemble of kinetic models to examine the effect on glucose uptake and CAT production.
+A single \textit{cyd} knockout reduced the CAT carbon yield from 8.6\% to 2.8\%.
 In addition, the glucose uptake rate was reduced compared to that of the control (no knockouts).
-A knockout of \textit{nuo} showed a less drastic effect, reducing the CAT carbon yield to 6.8\%; however, the glucose uptake rate remained similar to that of the control.
-Knocking out \textit{app} showed a CAT yield to 8.8\%, but was not statistically different from that of the control.
-With all three reactions knocked out, CAT yield was 2.7\%, not statistically different from the \textit{cyd} knockout.
-Thus, the model suggests that the majority of the energetic needs of the system as well as the production of CAT are met by oxidative phosphorylation, specifically by \textit{cyd}.
+On the other hand, a \textit{nuo} knockout showed a less dramatic decrease in yield,
+reducing the CAT carbon yield to 6.8\%; however, the glucose uptake rate remained similar to that of the control.
+Knocking out \textit{app} increased CAT yield to 8.8\%, but this increase was not statistically significantly different from that of the control.
+Lastly, knocking out all three reactions reduced the CAT yield to 2.7\%, but this was not statistically significantly different from the \textit{cyd} knockout alone.
+Thus, the model suggested the key step in oxidative phosphorylation was catalyzed by the gene product of \textit{cyd}.
+However, while disruption of \textit{cyd} significantly reduced the CAT carbon yield, it did not eliminate the ability of CFPS reaction to produce CAT.
+This suggested there was a mixture of energy sources supporting CAT production, with the most significant being oxidative phosphorylation.
+
 %The \textit{cyd} enzyme is responsible for transferring electrons with oxygen as an electron acceptor allowing the release of energy to form ATP.
+%The sensitivity analysis showed oxidative phosphorylation was significant for CAT production as well as the system state.
 
-To investigate the difference in yields between the unconstrained and constrained cases, we compared the flux distributions from the ssFBA simulations (Fig.~\ref{fig:Network}).
-The constrained cases (ii \& iii) heavily utilized the first step in the pentose phosphate pathway to generate NADPH; the carbon flux continued through the Entner–Doudoroff pathway toward pyruvate, a key metabolite as shown by sensitivity analysis.
-For case ii, the majority of the flux proceeded toward acetate accumulation, whereas in case iii, the flux accumulated as pyruvate, lactate, and acetate with some going through the TCA cycle.
-In comparison, the unconstrained case (i) showed the majority of flux traveling through glycolysis towards pyruvate, leading to an accumulation of lactate, acetate and malate.
-%In the second case (ii), there was a high flux through fumarate dehydrogenase from aspartic acid uptake, whereas in cases i and iii, acetate and lactate accumulation occurred.
-In all cases the energy source was primarily oxidative phosphorylation, and to a lesser extent the TCA cycle.
-%However, oxidative phosphorylation is not operating at its highest efficiency, since there is a mixture of acetate and lactate accumulation.
+Sequence-specific flux balance analysis (ssFBA) predicted optimal CAT yields with no adjustable parameters (Fig.~\ref{fig:CATProdYield}).
+Before exploring CFPS optimality, we first validated the ssFBA approach by comparing simulated and measured concentrations of CAT for the first hour of glucose consumption.
+We chose this time window (during the first phase of CAT production) because it was approximately linear in both glucose consumption and by-/production formation.
+The ssFBA calculation had no adjustable parameters; bounds on transcription and translation rates, and biochemical fluxes were either estimated from data, or
+from mechanistic models parameterized from literature.
+Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates, or the upper bounds for oxygen and glucose consumption rates was addressed
+by sampling plausible ranges for these parameters.
+The ensemble of ssFBA simulations predicted CAT formation as a function of time during the first hour of production when constrained by the experimental metabolite data (Fig.~\ref{fig:CATProdYield}A).
+Thus, the molecular description of transcription and translation were consistent with experimental measurements.
+Next, to gauge the performance of the CFPS reaction, we next calculated the CAT carbon yield for three classes of constraints:
+(i) theoretical maximum glucose, amino acid and oxygen upper bounds, and no transcriptional/translational constraints;
+(ii) theoretical maximum glucose, amino acid and oxygen upper bounds, and realistic transcriptional/translational constraints; and
+(iii) metabolite fluxes constrained by the CAT data, and realistic transcriptional/translational constraints (Fig.~\ref{fig:CATProdYield}B).
+The unconstrained theoretical maximum CAT carbon yield (case i) was 36.3\% $\pm$ 2.0\% (Fig.~\ref{fig:CATProdYield}B, left);
+this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, and transcription and translation were unbounded.
+On the other hand, for realistic transcription and translation constraints (case ii), the optimal CAT carbon yield decreased to 22.6\% $\pm$ 3.0\% (Fig.~\ref{fig:CATProdYield}B, middle).
+Lastly, when both realistic metabolite and transcription/translation constraints were applied (case iii), the predicted carbon yield was 6.2\% $\pm$ 2.0\%
+By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield of 8.6\% $\pm$ 0.4\%, equivalent to 23\% of the theoretical maximum (i)
+and 36\% of the physiological case (ii). The experimental dataset had a CAT carbon yield of 8.2\%,
+similar to both the kinetic model and the experimentally constrained ssFBA calculation (case iii).
+Thus, while the CFPS reaction was not optimal, the ssFBA calculations suggested that an approximately three-fold increase in carbon yield was theoretically possible.
+
+To investigate the differences in carbon yields, we compared the flux distributions predicted by ssFBA simulations for the different constraint cases  (Fig.~\ref{fig:Network}).
+The constrained cases (ii \& iii) heavily utilized the first step in the pentose phosphate pathway to generate NADPH;
+the carbon flux then continued through the Entner–Doudoroff pathway toward pyruvate.
+For case ii, the majority of the flux proceeded toward acetate accumulation, whereas in case iii, the flux accumulated as pyruvate, lactate, and acetate with some flux through the TCA cycle.
+In comparison, the unconstrained case (i) showed the majority of flux traveling through glycolysis towards pyruvate, leading to accumulation of lactate, acetate and malate.
+In all cases, the energy source was primarily oxidative phosphorylation, and to a lesser extent the TCA cycle.
 However, the accumulation of acetate and lactate signifies that the system is not operating at its highest efficiency.
-The system produces NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to power oxidative phosphorylation.
-Oxidative phosphorylation leads to a high redox ratio contributing to the accumulation of acetate overflow and diverting flux away from the TCA cycle.
-This suggests that there is potential to increase CAT production by reducing the accumulation of acetate and lactate.
-To investigate this further, we simulated potential knockouts with constrained transcription/translation rates, but constrained the specific glucose and amino acid uptake rates to the same values as simulated with no knockouts.
-Knocking out the \textit{gnd} reaction decreased flux of acetate production but increased flux through \textit{pntAB}, which is responsible for regenerating NADPH.
-The simulation showed carbon was diverted toward lactate; however, since CAT production is constrained by the translation rate, we expected no increase in CAT production.
-A second simulation with a knockout of \textit{gnd} and phosphate acetyltransferase showed carbon being diverted toward lactate and succinate; however, it required a higher flux through oxidative phosphorylation and the TCA cycle to meet the energetic needs of the system.
-The decrease in acetate production and the diversion of carbon flux shows a promising mechanism to increase CAT yield.
+The system produced NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to power oxidative phosphorylation.
+Oxidative phosphorylation lead to a high redox ratio contributing to the accumulation of acetate overflow and diverting flux away from the TCA cycle.
+This suggested CAT production could be increased by reducing the accumulation of acetate and lactate.
+To investigate this further, we simulated potential knockouts with constrained transcription/translation rates.
+Knocking out the \textit{gnd} reaction decreased acetate flux by more than half.
+In addition, less uptake of amino acids were required which increased the carbon yield of CAT by 2\% (up to approximately 8\%) compared to the control (no knockouts).
+The simulation showed an increase in oxidative phosphorylation flux and the majority of flux traveling through the pentose phosphate pathway.
+A second simulation with both \textit{gnd} and phosphate acetyltransferase knocked out,
+showed less carbon was needed to meet the translation bound, which increased the carbon yield by 6\% compared to the control (up to approximately 14\%).
+In the dual knockout, flux towards acetate was mostly coming from amino acid degradation.
+Sixty percent of the carbon flux traveled through glycolysis, while the remaining traveled through the pentose phosphate pathway.
+This was similar to the flux distribution of the unconstrained case (case i).
+Taken together, the dual knockout decreased acetate production and required less amino acid consumption, thus it is a promising strategy to increase the CAT carbon yield.
+\textcolor{red}{MIKE: CAN WE GET A FIGURE THAT SHOWS THE KO FLUXES -and- WHAT ARE THE ERRORS ON THE YIELDS?}
+
+% To investigate this further, we simulated potential knockouts with constrained transcription/translation rates,
+% and constrained the specific glucose and amino acid rates to the same simulated without knockouts.
+% Knocking out \textit{gnd} decreased acetate production but increased flux through \textit{pntAB}, which is responsible for regenerating NADPH.
+% The simulation showed carbon was diverted toward lactate; however, since CAT production was constrained by the translation rate, we expected no increase in CAT production.
+% A second simulation with \textit{gnd} and phosphate acetyltransferase knockouts showed carbon being diverted toward lactate and succinate;
+% however, it required a higher flux through oxidative phosphorylation and the TCA cycle to meet the energetic needs of CAT production.
+% The decrease in acetate production and the diversion of carbon flux shows a promising mechanism to increase CAT yield.
 
@@
 CFPS is seen to be biphasic relying on glucose during the first hour and pyruvate and lactate afterward.
-Allosteric control was essential to the maintainence of the network and production of CAT, as without it, central carbon metabolism is exhuasted within 1.5 hours leading to low CAT production.
+Allosteric control was essential to the maintenance of the network and production of CAT, as without it, central carbon metabolism is exhuasted within 1.5 hours leading to low CAT production.
 Having captured the experimental data, we investigated if CAT yield and CFPS performance could be further improved.
@@
 This is important in applying metabolic engineering principles to rationally design cell-free production processes and predict the redirection of carbon fluxes to product forming pathways.
-In analyzing the model parameters' effect on CAT production, CAT synthesis is the most important, followed by oxidative phosphorylation and the glutamate and pyruvate consuming reactions, as well as coenzyme reactions which are necessary to drive CAT synthesis.
+In analyzing the model parameters' effect on CAT production, CAT synthesis is the most important, followed by oxidative phosphorylation and the glutamate and pyruvate consuming reactions, as well as cofactor reactions which are necessary to drive CAT synthesis.
 For example, the conversion of ATP to GTP shows significance since it is necessary for CAT synthesis.
@@
 Thus, supplementation with additional GTP may improve the efficiency of CAT production.
-A similar theme is seen in the sensitivity of overall model state, where the most important reactions are glucose and pyruvate consuming reactions and coenzyme reactions which are vital to drive CFPS.
+A similar theme is seen in the sensitivity of overall model state, where the most important reactions are glucose and pyruvate consuming reactions and cofactor reactions which are vital to drive CFPS.
 This can be seen in the biphasic operation of CFPS, with the first phase operating on glucose and the second phase operating on pyruvate.
@@
 G_{p}+R_{1} &\longrightarrow& G_{p}^{*} \\
-G_{p}^{*}+\sum_{k}\eta_{k}\cdot kTP &\longrightarrow& mRNA+G_{p}+R_{1}+2\sum_{k}\eta_{k}\cdot Pi \qquad{k=A,C,G,U}\\
+G_{p}^{*}+\sum_{k}\eta_{k}\cdot kTP &\longrightarrow& mRNA+G_{p}+R_{1}+\\
+&& \qquad 2\sum_{k}\eta_{k}\cdot Pi \qquad{k=A,C,G,U}\\
 mRNA &\longrightarrow& \sum_{k}\eta_{k}\cdot kMP \\
 mRNA+R_{2} &\longrightarrow& R_{2}^{*} \\
-\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP &\longrightarrow& \alpha_{j}\cdot AA_{j}:tRNA_{j}+\alpha_{j}\cdot AMP+2\alpha_{j}\cdot Pi \qquad{j=1,2,\hdots,20}\\
+\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP &\longrightarrow& \alpha_{j}\cdot AA_{j}:tRNA_{j}+\\
+&& \qquad \alpha_{j}\cdot AMP+2\alpha_{j}\cdot Pi \qquad{j=1,2,\hdots,20}\\
 R_{2}^{*}+\sum_{j}\alpha_{j}\cdot \Big(AA_{j}:tRNA_{j}+2\cdot GTP\Big) &\longrightarrow& \mathcal{P}+R_{2}+mRNA+\\
@@
 The manuscript was prepared and edited for publication by W.D, A.Y and J.V.
+\textcolor{red}{JV: UPDATE THIS TEXT}
 

### 2016-11-28 Paper_v1.tex
Updates to MM

@@
 %\bibliographystyle{chicago}
-\bibliographystyle{plos2009}
+\bibliographystyle{plos2015}
 
@@
 We estimated the theoretical maximum CAT carbon yield using sequence-specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
-The sequence specific flux balance analysis problem was formulated as:
+The sequence specific flux balance analysis problem was formulated as a linear program:
 \begin{equation}
@@
 and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
-The objective of the calculation is the maximization of CAT translation rate, $w_{TL}$.
+The stoichiometry of the kinetic model was used for the ssFBA calculations, with the execpetion of the transcription and translation rates.
+The transcription (TX) and translation (TL) stoichiometry was modeled using the template reactions taken from Allen and Palsson \cite{Allen:2003aa}:
+\begin{eqnarray*}
+\mathrm{{G}_{\mathcal{P}}}+\mathrm{{R}_{1}} &\longrightarrow& \mathrm{G_{\mathcal{P}}^{*}} \\
+\mathrm{G_{\mathcal{P}}^{*}}+\sum_{\mathrm{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \mathrm{\left\{k\right\}TP} &\overset{\rm TX}{\longrightarrow}& \mathrm{mRNA}+\mathrm{G_{\mathcal{P}}}+\mathrm{R_{1}}+ \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}2\eta_{k}\cdot Pi}\\
+\mathrm{mRNA} &\longrightarrow& \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}\eta_{k}\cdot \left\{k\right\}MP} \\
+\mathrm{mRNA}+\mathrm{R_{2}} &\longrightarrow& \mathrm{R_{2}^{*}} \\
+\mathrm{\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP} &\longrightarrow& \mathrm{\alpha_{j}\cdot AA_{j}-tRNA_{j}+}\\
+&& \mathrm{\qquad \alpha_{j}\cdot AMP+2\alpha_{j}\cdot Pi} \qquad{j=1,2,\hdots,20}\\
+\mathrm{R_{2}^{*}+\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot \Big(AA_{j}-tRNA_{j}+2\cdot GTP\Big)} &\stackrel{\rm TL}{\longrightarrow}& \mathcal{P}+\mathrm{R_{2}+mRNA}+\\
+&& \mathrm{\quad +\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot\Big(tRNA+2\cdot GDP+2\cdot Pi\Big)}
+\end{eqnarray*}
+where $G_{\mathcal{P}}$ denotes the gene encoding protein product $\mathcal{P}$,
+$\rm R_{1}$ denotes the concentration of RNA polymerase,
+$G_{\mathcal{P}}^{*}$ denotes the gene bounded by the RNA polymerase,
+$\eta_{i}$ and $ \alpha_{j}$ denote the stoichiometric coefficients for nucleotide and amino acid, respectively,
+$\rm Pi$ denotes inorganic phosphate,
+$\rm R_{2}$ denotes the ribosome concentration,
+$\rm R_{2}^{*}$ denotes bounded ribosome,
+and $AA_{j}$ denotes $j^{th}$ amino acid.
+
+The transcription rate ($w_{TX}$) was fixed in the ssFBA calculation at:
+\begin{equation}
+	w_{TX} = V_{TX}^{max}\left(\frac{G}{K_{TX}+G}\right)
+\end{equation}
+where $G$ denotes the gene concentration, and $K_{TX}$ denotes a transcription saturation coefficient.
+The maximum rate of transcription $V_{TX}^{max}$ was formulated as:
+\begin{equation}
+	V_{TX}^{max} \equiv \left[R_{1}\left(\frac{v_{TX}}{l_{G}}\right)\left(\frac{K_{T7}}{1 + K_{T7}}\right)\right]
+\end{equation}
+The term $R_{1}$ denotes the RNA polymerase abundance,
+$v_{TX}$ denotes the RNA polymerase elongation rate (nt/hr),
+$l_{G}$ denotes the gene length in nucleotides, and the last term describes T7 promoter activity,
+where $K_{T7}$ denotes a T7 RNA polymerase binding constant \cite{Moon:2012aa}.
+On the other hand, the translation rate ($w_{TL}$) was bounded by:
+ \begin{equation}
+	0\leq w_{TL} \leq V_{TL}^{max}\left(\frac{\rm mRNA_{SS}}{K_{TL}+\rm mRNA_{SS}}\right)
+\end{equation}
+where $\rm mRNA_{SS}$ denotes the steady state mRNA abundance, and $K_{TL}$ denotes the translation saturation constant.
+The maximum translation rate $V_{TL}^{max}$ was formulated as:
+\begin{equation}
+	V_{TL}^{max} \equiv \left[K_{P} R_{2}\left(\frac{v_{TL}}{l_{P}}\right)\right]
+\end{equation}
+The term $K_{P}$ denotes the polysome amplification constant,
+$v_{TL}$ denotes the ribosome elongation rate (amino acids per hour),
+$l_{P}$ denotes the number of amino acids in the protein of interest,
+and $\rm mRNA_{SS}$ denotes the steady-state mRNA concentration:
+\begin{equation}
+	\rm mRNA_{SS}\simeq\frac{w_{TX}}{\lambda}
+\end{equation}where $\lambda$ denotes the rate constant controlling the mRNA degradation rate.
+
+The objective of the sequence specific flux balance calculation was to maximize the rate of CAT translation, $w_{TL}$.
 The total glucose uptake rate was bounded by [0,40 mM/h] according to experimental data;
 while the amino acid uptake rates were bounded by [0,30 mM/h], but did not reach the maximum flux.
-The transcription and translation rates were modeled using the template reactions:
-\begin{eqnarray*}
-G_{p}+R_{1} &\longrightarrow& G_{p}^{*} \\
-G_{p}^{*}+\sum_{k}\eta_{k}\cdot kTP &\longrightarrow& mRNA+G_{p}+R_{1}+\\
-&& \qquad 2\sum_{k}\eta_{k}\cdot Pi \qquad{k=A,C,G,U}\\
-mRNA &\longrightarrow& \sum_{k}\eta_{k}\cdot kMP \\
-mRNA+R_{2} &\longrightarrow& R_{2}^{*} \\
-\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP &\longrightarrow& \alpha_{j}\cdot AA_{j}:tRNA_{j}+\\
-&& \qquad \alpha_{j}\cdot AMP+2\alpha_{j}\cdot Pi \qquad{j=1,2,\hdots,20}\\
-R_{2}^{*}+\sum_{j}\alpha_{j}\cdot \Big(AA_{j}:tRNA_{j}+2\cdot GTP\Big) &\longrightarrow& \mathcal{P}+R_{2}+mRNA+\\
-&& \quad +\sum_{j}\alpha_{j}\Big(tRNA+2\cdot GDP+2\cdot Pi\Big)
-\end{eqnarray*}
-where $G_{p}$ denotes the gene of protein $p$,
-$R_{1}$ denotes the concentration of RNA polymerase,
-$G_{p}^{*}$ denotes the gene bounded by the RNA polymerase,
-$\eta_{i}$ and $ \alpha_{j}$ denote the coefficients for species $i$ and $j$ respectively,
-$P_{i}$ denotes inorganic phosphate, $R_{2}$ denotes the ribosome concentration, $R_{2}^{*}$ denotes bounded ribosome, and $X_{j}$ denotes the amino acid species $j$.
+The CAT gene and protein sequences were taken from literature.
+The sequence specific flux balance linear program was solved using the GNU Linear Programming Kit (GLPK) v4.52 \cite{GLPK}.
 
-The transcription rate ($w_{TX}$) was fixed in the ssFBA calculation as:
-\begin{equation}
-	w_{TX} = \left[R_{1}\left(\frac{v_{TX}}{l_{G}}\right)\left(\frac{K_{T7}}{1 + K_{T7}}\right)\right]\left(\frac{G}{K_{TX}+G}\right)
-\end{equation}
-The first term in the brackets denotes the maximum rate of transcription, $v_{TX}$ denotes the RNA polymerase elongation rate (nt/hr), $l_{G}$ denotes the gene length in nucleotides, and the last term describes T7 promoter activity, where $K_{T7}$ quantifies T7 RNA polymerase binding \cite{Moon:2012aa}.
-The second term denotes the gene saturation term, where $G$ denotes the gene concentration and $K_{TX}$ denotes a saturation coefficient.
-The CAT gene and protein sequences were determined from literature.
-The translation rate ($w_{TL}$) was bounded by:
- \begin{equation}
-	0\leq w_{TL} \leq \left[K_{P} R_{2}\left(\frac{v_{TL}}{l_{P}}\right)\right]\left(\frac{mRNA_{SS}}{K_{TL}+mRNA_{SS}}\right)
-\end{equation}
-The first term in brackets denotes the maximum translation rate, where $K_{P}$ denotes the polysome amplification constant, $v_{TL}$ denotes the ribosome elongation rate (amino acids per hour), $l_{P}$ denotes the number of amino acids in the protein of interest, and $mRNA_{SS}$ denotes the steady-state mRNA concentration:
-\begin{equation}
-	mRNA_{SS}\simeq\frac{w_{TX}}{\lambda}
-\end{equation}where $\lambda$ denotes the rate constant controlling the mRNA degradation rate.
-
+\subsubsection*{Calculation of the carbon yield.}
 The CAT carbon yield ($Y_{C}^{CAT}$) was calculated as the ratio of carbon produced as CAT dived by the carbon consumed as reactants (glucose and amino acids):
@@
 
+\subsubsection*{Quantification of uncertainty.}
 An ensemble of 100 sets of flux distributions was calculated for three different cases: unconstrained, constrained by transcription/translation rates, and constrained by transcription/translation rates and experimental measurements.
@@
 This included the data for glucose, organic acids, energy species, and amino acids; CAT was not constrained by experimental data, but by the transcription/translation rates as stated above.
-The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
+
+%The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52)
 
@@
 \section*{Author's contributions}
-J.V and A.Y directed the study. R.T, H.J and J.C conducted the cell culture measurements.
-J.V and W.D developed the reduced order HL-60 models and the parameter ensemble.
-W.D analyzed the model ensemble, and generated figures for the manuscript.
-The manuscript was prepared and edited for publication by W.D, A.Y and J.V.
-\textcolor{red}{JV: UPDATE THIS TEXT}
+J.V directed the modeling study.
+K.C and J.S conducted the cell free protein synthesis experiments.
+J.V, J.W, and N.H developed the cell free protein synthesis mathematical model, and parameter ensemble.
+J.V and M.V performed the sequence specific flux balance analysis calculations.
+The manuscript was prepared and edited for publication by J.S, N.H, M.V, J.W and J.V.
 
@@
 \section*{Funding}
-This study was supported by a National Science Foundation Graduate Research Fellowship (DGE-1333468) to N.H and by an award from the US Army and Systems Biology of Trauma Induced Coagulopathy (W911NF-10-1-0376) to J.V for the support of M.V.
+This study was supported by a National Science Foundation Graduate Research Fellowship (DGE-1333468) to N.H.
+Research reported in this publication was also supported by the Systems Biology Coagulopathy of Trauma Program
+with support from the US Army Medical Research and Materiel Command under award number W911NF-10-1-0376.
 

### 2016-11-29 Paper_v1.tex
flux KO

@@
 This suggested CAT production could be increased by reducing the accumulation of acetate and lactate.
-To investigate this further, we simulated potential knockouts with constrained transcription/translation rates.
-Knocking out the \textit{gnd} reaction decreased acetate flux by more than half.
-In addition, less uptake of amino acids were required which increased the carbon yield of CAT by 2\% (up to approximately 8\%) compared to the control (no knockouts).
-The simulation showed an increase in oxidative phosphorylation flux and the majority of flux traveling through the pentose phosphate pathway.
+To investigate this further, we simulated potential knockouts with constrained transcription/translation rates  (Fig.~\ref{fig:flux_ko}).
+Knocking out the \textit{gnd} reaction, the first step in Entner-Doudoroff pathway, decreased acetate flux by about half.
+In addition, less uptake of amino acids were required which increased the carbon yield of CAT by 2.7\% (up to approximately 25.3\% $\pm$ 3.6\%) compared to the control (no knockouts).
+The simulation showed an increase in oxidative phosphorylation flux and the flux splitting between glycolysis and pentose phosphate pathway.
 A second simulation with both \textit{gnd} and phosphate acetyltransferase knocked out,
-showed less carbon was needed to meet the translation bound, which increased the carbon yield by 6\% compared to the control (up to approximately 14\%).
-In the dual knockout, flux towards acetate was mostly coming from amino acid degradation.
-Sixty percent of the carbon flux traveled through glycolysis, while the remaining traveled through the pentose phosphate pathway.
+showed less carbon was needed to meet the translation bound, which increased the carbon yield by 4\% compared to the control (up to approximately 26.7\% $\pm$ 3.4\%).
+In the dual knockout, flux towards acetate was almost negligible with some coming from amino acid degradation.
+Sixty-one percent of the carbon flux traveled through glycolysis, while the remaining traveled through pentose phosphate pathway.
 This was similar to the flux distribution of the unconstrained case (case i).
 Taken together, the dual knockout decreased acetate production and required less amino acid consumption, thus it is a promising strategy to increase the CAT carbon yield.
-\textcolor{red}{MIKE: CAN WE GET A FIGURE THAT SHOWS THE KO FLUXES -and- WHAT ARE THE ERRORS ON THE YIELDS?}
 
@@
 The model and experimental yields were 36\% of the theoretical maximum and 23\% of a physiologically constrained case.
-Knockouts of \textit{gnd} and phosphate acetyltransferase show that carbon can be diverted away from acetate and potentially toward CAT or other proteins of interest expressed in CFPS.
+Knockouts of \textit{gnd} and phosphate acetyltransferase show that carbon can be diverted away from acetate and toward CAT or other proteins of interest expressed in CFPS.
 Another limitation to be addressed in CFPS is the transcription and translation description, since protein production is ultimately bounded by these kinetic rates.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{./Figures/CAT_flux_final.pdf}
+\includegraphics[width=0.75\textwidth]{./Figures/CAT_flux_final.pdf}
 \caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, and oxidative phosphorylation.
@@
 \label{fig:Network}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.75\textwidth]{./Figures/CAT_flux_ko.pdf}
+\caption{Flux profile with knockouts for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, and oxidative phosphorylation.
+Sequence-specific FBA flux value (mean ± standard deviation) across ensemble for 1 hr, normalized to glucose uptake flux.
+Flux distribution for three different cases: constrained by transcription and translation rates (top row), constrained by transcription and translation rates with \textit{gnd} knocked out (second row), and constrained by transcription and translation rates with \textit{gnd} and phosphate acetyltransferase knocked out.}
+\label{fig:flux_ko}
 \end{figure}

### 2016-11-30 Paper_v1.tex
Nov30

@@
 We perturbed each $V^{max}$ parameter, either individually or in pairwise combinations and measured the change in either CAT production or the overall system state.
-CAT production was most dependent upon the abundance of cofactors such as ATP, GTP, NADH and NADPH and the metabolite pyruvate.
 CAT production was most sensitive to the CAT synthesis reaction, oxidative phosphorylation, and the pyruvate-consuming alanine synthesis reaction (Fig.~\ref{fig:Sensitivity}, top, section A).
-Taking into account these three reactions, and the next 16 most important reactions (section B), we saw a common theme of reactions involving the cofactors ATP, GTP, NADH, and NADPH
-as well as the metabolites pyruvate and glutamate.
-Glutamate was important as a precursor for the synthesis of other amino acids required by CAT production.
+%Taking into account these three reactions, and the next 16 most important reactions (section B), we saw a common theme of reactions involving the cofactors ATP, NADH, NADPH, and coenzyme A, as well as the metabolites pyruvate and glutamate.
+We saw a common theme of the most important reactions producing or consuming the cofactors ATP, NADH, NADPH, and coenzyme A, as well as the metabolites pyruvate and glutamate.
+Of the 25 reactions to which CAT production was most sensitive, 9 produced or consumed ATP, making it the most represented in these top reactions (with the exception of hydrogen and phosphate ion).
+The next most represented were pyruvate, glutamate, and ADP with 7 reactions each, followed by coenzyme A, $\alpha$-ketoglutarate, NAD/NADH, and NADP/NADPH, with 6 reactions each.
+This makes sense, as glutamate was an important precursor for the synthesis of other amino acids required by CAT production.
 Meanwhile, the cofactors provided energy to power CAT synthesis, while pyruvate was important for energy generation following glucose depletion.
@@
 The pairwise sensitivities (off-diagonal elements) were different from the corresponding first-order sensitivities (diagonal elements), and led to interesting outcomes.
-The combination of certain reactions had a greater effect on CAT production than that of the individual reactions by themselves.
+The combination of certain reactions had a much greater or lesser effect on CAT production than that of the individual reactions by themselves.
 For example, glutamine synthesis and arginine degradation were both among the most important reactions to CAT production (they ranked 5th and 10th, respectively).
@@
 However, when both were perturbed, their combined effect on the model was low, as the respective contributions to consumption and production of glutamate cancelled.
-\textcolor{red}{NICK: ARE THERE EXAMPLES OF POSITIVE SYNERGY?}
+An example of positive synergy can be seen in cometabolite interconversion.
+Pyridine nucleotide transhydrogenase catalyzes two reactions: one converts NAD and NADPH into NADH and NADP, while the other does the reverse and also generates a proton gradient.
+Increasing one or the other has little effect on the model, as the reaction is hampered by the loss of reactants.
+Increasing both, however, allows all cometabolites to be conserved and the reactions to continue unhindered.
+Thus, the pairwise sensitivity is much higher than the sum of the first-order sensitivities.
 
-The overall system state was also sensitive to cofactors and substrates; however, instead of pyruvate and glutamate, the substrates driving metabolism were glucose and pyruvate.
-The metabolism as a whole was most sensitive to glucose uptake via GTP and the forward reaction of lactate dehydrogenase, consuming pyruvate (Fig.~\ref{fig:Sensitivity}, bottom, section F).
-These two and the next 30 most important reactions (section G) largely involved cofactors, especially ATP and NADPH, as well as substrate-consuming reactions and oxidative phosphorylation.
-The system state had even more pairwise sensitivities that differed from the corresponding first-order sensitivities and stood out as significant.
-For example, the first-order effect of alanine synthesis was large; it consumed both pyruvate and glutamate, two key species in the network.
-However, there were enough reactions that, when paired with alanine synthesis, had little effect on the model;
-for example, malic enzyme produces pyruvate that can be consumed by alanine synthesis.
-Thus, the total-order alanine synthesis sensitivity was low, placing it at the very bottom of section I.
-Another interesting result was the intersection of sections F and G with section J.
-The 53 reactions in section J were turned off in the best-fit set ($V^{max}=0$);
-therefore, the perturbation of these reactions had no effect on the model.
-As a result, all pairwise sensitivities with reactions in section J were pseudo first-order sensitivities for the other reactions.
-Interestingly, many reactions in section F and several in section G showed their highest sensitivities when paired with the "non-effects" of section J.
-\textcolor{red}{NICK: DOES THIS MAKE SENSE? WHY ARE THESE DIFFERENT THAN THE SINGLE PERTURBATIONS?}
-Of these, three involved pyruvate, strengthening its role as a key metabolite; the others were glucose consumption via GTP/CTP-specific hexokinases,
-fumarate reductase, and SO$_4$ utilization.
-This suggested that these reactions' effects on the model were canceled out or lessened by most other reactions, but were of course not affected by the reactions in section J.
-This was also likely the reason that section J ranked above section K, despite having no effect on the model themselves.
+The overall system state was also sensitive to cofactors and substrates; however, instead of pyruvate and glutamate, the substrates driving metabolism were pyruvate and G3P.
+The system was most sensitive to an oxidative phosphorylation reaction (cytochrome oxidase), which converted ubiquinol to ubiquinone while generating a proton gradient.
+The next 4 most important reactions were all consumers of pyruvate: lactate dehydrogenase, pyruvate formate lyase, alanine synthesis, and PEP synthase.
+Of the 25 reactions to which the system was most sensitive, 7 produced or consumed pyruvate and 7 participated in NAD/NADH exchange.
+With the exception of hydrogen, these were the most represented species in the top 25 reactions.
+Also important were coenzyme A (5 reactions) and acetyl coenzyme A (4 reactions), as well as ubiquinone/ubiquinol, NADP/NADPH, G3P, and ATP  (4 reactions each).
+%We saw a common theme of the most important reactions producing or consuming the cofactors ATP, NADH, NADPH, ubiquinone/ubiquinol, and coenzyme A, as well as the metabolites pyruvate and G3P.
+%These two and the next 30 most important reactions (section G) largely involved cofactors, especially ATP and NADPH, as well as substrate-consuming reactions and oxidative phosphorylation.
+The system state also had pairwise sensitivities that differed from the corresponding first-order sensitivities and stood out as significant.
+For example, alanine degradation was among the most important reactions, as it produced pyruvate; GMP synthesis was also moderately important, as it produced glutamate.
+However, when both reactions were increased, the combined effect on the model was almost zero.
+%This can be understood by also considering a third reaction: alanine synthesis from pyruvate and glutamate.
+This can be understood by considering the reactions that involve both pyruvate and glutamate: they all either produced both of these substrates or consumed both.
+When alanine degradation was increased, the excess pyruvate stimulated these reactions to consume both pyruvate and glutamate; the amounts of both of these substrates could not be conserved.
+But when GMP synthesis was perturbed as well, the glutamate deficiency was corrected, cancelling much of the effect on the system.
+One of the pyruvate- and glutamate-consuming reactions was alanine synthesis; in this case, the levels of pyruvate, glutamate, and alanine were all virtually unchanged.
+%An example of positive synergy can be seen in the interplay of amino acid synthesis and the TCA cycle: aspartate synthesis and the reverse reaction of succinyl coenzyme A synthetase are both relatively unimportant to the system.
+An example of positive synergy can be seen in histidine synthesis, one of the least influential reactions that consumes three units of ATP.
+When perturbed in combination with the reverse reaction of succinyl coenzyme A synthetase, another reaction with little overall effect on the model, the combined effect on the system is much greater.
+This may be because the reverse reaction of succinyl coenzyme A synthetase produced ATP, which further stimulated histidine synthesis.
+%Another interesting result was the intersection of sections F and G with section J.
+%The 53 reactions in section J were turned off in the best-fit set ($V^{max}=0$);
+%therefore, the perturbation of these reactions had no effect on the model.
+%As a result, all pairwise sensitivities with reactions in section J were pseudo first-order sensitivities for the other reactions.
+%Interestingly, many reactions in section F and several in section G showed their highest sensitivities when paired with the "non-effects" of section J.
+%\textcolor{red}{NICK: DOES THIS MAKE SENSE? WHY ARE THESE DIFFERENT THAN THE SINGLE PERTURBATIONS?}
+%Of these, three involved pyruvate, strengthening its role as a key metabolite; the others were glucose consumption via GTP/CTP-specific hexokinases,
+%fumarate reductase, and SO$_4$ utilization.
+%This suggested that these reactions' effects on the model were canceled out or lessened by most other reactions, but were of course not affected by the reactions in section J.
+%This was also likely the reason that section J ranked above section K, despite having no effect on the model themselves.
 Taken together, sensitivity analysis identified blocks of parameters that either individually, or in combination influenced model performance.

### 2016-11-30 Paper_v1.tex
Nov30_2

@@
 This suggests that CAT production could be further optimized.
-The dynamic modeling approach predicted that substrate consumption of glucose and pyruvate and oxidative phosphorylation were most important to both CAT production and the system as a whole, while CAT production alone depended heavily on the CAT synthesis reaction.
+The dynamic modeling approach predicted that substrate consumption and oxidative phosphorylation were most important to both CAT production and the system as a whole, while CAT production alone depended heavily on the CAT synthesis reaction.
 Conversely, CAT production was robust to allosteric control, as was most of the network, with the exception of the organic acids in central carbon metabolism.

### 2016-12-02 Paper_v1.tex
Dec2

@@
 \usepackage{color}
+\usepackage{booktabs}
+\setlength\heavyrulewidth{0.4ex}
+\setlength\lightrulewidth{0.25ex}
 %\usepackage{textgreek}
@@
 The ensemble described all of the training data, especially the central carbon metabolism.
-The model predicted a carbon yield for CAT production that was equal to 23\% of the maximum theoretical yield, calculated using sequence-specific flux balance analysis.
+The model predicted a carbon yield for CAT production that was equal to 35\% of yield for a physiologically realistic case, calculated using sequence-specific flux balance analysis.
 This suggests that CAT production could be further optimized.
@@
 Next, to gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield with the maximum theoretical CAT carbon yield calculated using sequence-specific flux balance analysis.
-The CAT yield estimated from the kinetic model was 23\% of the maximum theoretical yield, but 36\% of the theoretical yield when physiologically realistic constraints were used.
+The CAT yield estimated from the kinetic model was 35\% of the theoretical yield when physiologically realistic constraints were used.
 Taken together, we have integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
@@
 The ensemble of kinetic CFPS models captured the time evolution of CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
-The cell-free \textit{E.~coli} metabolic network was constructed by removing growth associated reactions from the MG1655 reconstruction \cite{Feist:2007aa},
-and by adding reactions describing chloramphenicol acetyltransferase (CAT) biosynthesis, a model protein for which there exists a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
+The cell-free \textit{E.~coli} metabolic network was constructed by removing growth associated reactions from the MG1655 reconstruction \cite{Feist:2007aa}, and by adding reactions describing chloramphenicol acetyltransferase (CAT) biosynthesis, a model protein for which there exists a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
+In addition, reactions that were knocked out from the cell extract preparation were removed from the network ($\Delta$speA, $\Delta$tnaA, $\Delta$sdaA, $\Delta$sdaB, $\Delta$gshA, $\Delta$tonA, $\Delta$endA).
 The CFPS model equations were formulated using the hybrid cell-free modeling framework of Wayman et al. \cite{pr3010138}.
@@
 To investigate this further, we knocked out key oxidative phosphorylation reactions in the ensemble of kinetic models to examine the effect on glucose uptake and CAT production.
-A single \textit{cyd} knockout reduced the CAT carbon yield from 8.6\% to 2.8\%.
+%A single \textit{cyd} knockout reduced the CAT carbon yield from 7.9\% to 2.6\% (Table~\ref{tbl:yield_breakdown}).
+A single \textit{cyd} knockout reduced the CAT carbon yield from 2.7\% to 0.9\% (Table~\ref{tbl:yield_breakdown}).
 In addition, the glucose uptake rate was reduced compared to that of the control (no knockouts).
+%On the other hand, a \textit{nuo} knockout showed a less dramatic decrease in yield,
+%reducing the CAT carbon yield to 6.9\%; however, the glucose uptake rate remained similar to that of the control.
 On the other hand, a \textit{nuo} knockout showed a less dramatic decrease in yield,
-reducing the CAT carbon yield to 6.8\%; however, the glucose uptake rate remained similar to that of the control.
-Knocking out \textit{app} increased CAT yield to 8.8\%, but this increase was not statistically significantly different from that of the control.
-Lastly, knocking out all three reactions reduced the CAT yield to 2.7\%, but this was not statistically significantly different from the \textit{cyd} knockout alone.
+reducing the CAT carbon yield to 2.3\%; however, the glucose uptake rate remained similar to that of the control.
+%Knocking out \textit{app} increased CAT yield to 8.3\%, but this increase was not statistically significantly different from that of the control.
+Knocking out \textit{app} did not change the CAT yield (it remained at 2.7\%).
+%Lastly, knocking out all three reactions reduced the CAT yield to 3.0\%, but this was not statistically significantly different from the \textit{cyd} knockout alone.
+Lastly, knocking out all three reactions reduced the CAT yield to 0.7\%, similar to knocking out the \textit{cyd} alone.
 Thus, the model suggested the key step in oxidative phosphorylation was catalyzed by the gene product of \textit{cyd}.
@@
 This suggested there was a mixture of energy sources supporting CAT production, with the most significant being oxidative phosphorylation.
-
 %The \textit{cyd} enzyme is responsible for transferring electrons with oxygen as an electron acceptor allowing the release of energy to form ATP.
 %The sensitivity analysis showed oxidative phosphorylation was significant for CAT production as well as the system state.
+A similar distribution of the carbon contribution to CAT yield was seen across the best-fit set and all knockouts.
+In all cases, glutamate was responsible for about two-thirds of carbon consumption toward CAT.
+This was due to its role in generating other amino acids and as a possible energy source for metabolism.
+The much greater consumption of this one amino acid was made possible by its much larger initial condition, as it was present in the cell media in the form of magnesium glutamate, ammonium glutamate, and potassium glutamate [CITE?
+THESIS NOT CURRENTLY IN BIBLIOGRAPHY].
+Glucose made the next largest contribution, between 20\% and 30\%, dwarfing all amino acids other than glutamate.
+This makes sense, as glucose powers the entire metabolism during the first hour, including energy species synthesis and amino acid synthesis.
+
+Furthermore, we examined the proportion of amino acid consumption that was due to CAT synthesis in the best-fit set (Table~\ref{tbl:AA_breakdown}).
+Eight of the 20 amino acids had a percentage of 100\% or higher.
+For six of these (isoleucine, leucine, methionine, phenylalanine, tryptophan, valine), synthesis was inactive or negligible, meaning that CAT synthesis was the only reaction that consumed these species.
+This is why methionine, and to a lesser extent isoleucine, leucine, and valine do not fit the experimental data as well as some other amino acids; consumption via CAT synthesis was not enough to fit the data, and no other reactions exist in the network to consume these species.
+For histidine and tyrosine, synthesis was active, allowing the percentage to be higher than 100\%.
+Another ten amino acids registered percentages lower than 100\%, meaning that they were consumed by other pathways in the network.
+The most extreme of these is glutamate, with a percentage of 0.2\%, because it was consumed by numerous reactions for the synthesis of other amino acids and as an energy substrate.
+Likewise, arginine, asparagine, glycine, and proline are consumed in the production of other amino acids, while cysteine and lysine are consumed to form other metabolites.
+%ARG R_arg_deg1-4 go to GLU
+%ASN R_asn_deg goes to ASP
+%ASP R_thr, maybe negligible
+%CYS cys_deg goes to pyr
+%GLY R_gar goes to GLU
+%LYS lys_deg goes to cadav
+%PRO pro_deg goes to GLU
+%SER R_glyA, maybe negligible
+%THR R_thr_deg4, maybe negligible
+Finally, alanine and glutamine increased during the timecourse, meaning that synthesis outweighed all consumption (including consumption toward CAT).
 
@@
 (iii) metabolite fluxes constrained by the CAT data, and realistic transcriptional/translational constraints (Fig.~\ref{fig:CATProdYield}B).
-The unconstrained theoretical maximum CAT carbon yield (case i) was 36.3\% $\pm$ 2.0\% (Fig.~\ref{fig:CATProdYield}B, left);
-this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, and transcription and translation were unbounded.
+The unconstrained theoretical maximum CAT carbon yield (case i) was 36.3\% $\pm$ 2.0\% (Fig.~\ref{fig:CATProdYield}B, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, and transcription and translation were unbounded.
 On the other hand, for realistic transcription and translation constraints (case ii), the optimal CAT carbon yield decreased to 22.6\% $\pm$ 3.0\% (Fig.~\ref{fig:CATProdYield}B, middle).
 Lastly, when both realistic metabolite and transcription/translation constraints were applied (case iii), the predicted carbon yield was 6.2\% $\pm$ 2.0\%
-By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield of 8.6\% $\pm$ 0.4\%, equivalent to 23\% of the theoretical maximum (i)
-and 36\% of the physiological case (ii). The experimental dataset had a CAT carbon yield of 8.2\%,
-similar to both the kinetic model and the experimentally constrained ssFBA calculation (case iii).
+%By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield of 8.6\% $\pm$ 0.4\%, equivalent to 23\% of the theoretical maximum (i) and 35\% of the physiological case (ii).
+By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield (without arginine and glutamate) of 7.9\%, equivalent to 35\% of the physiological case (ii).
+The experimental dataset had a CAT carbon yield of 8.2\%, similar to both the kinetic model and the experimentally constrained ssFBA calculation (case iii).
 Thus, while the CFPS reaction was not optimal, the ssFBA calculations suggested that an approximately three-fold increase in carbon yield was theoretically possible.
@@
 Having captured the experimental data, we investigated if CAT yield and CFPS performance could be further improved.
-We showed that the model produces CAT at 23\% of the theoretical maximum in terms of carbon yield, and at 36\% of a physiological case in which transcription and translation are constrained.
+We showed that the model produces CAT with a carbon yield equal to 35\% of that of a physiological case in which transcription and translation are constrained.
 The accumulation of waste byproducts, especially acetate, is responsible for this sub-optimal yield.
@@
 %Many of the insignificant reactions lead to the accumulation of unwanted byproducts such as ethanol and formate.
-Comparing the theoretical maximum carbon yield of CAT from ssFBA predictions to those of the kinetic model and experimental measurements suggests that there is potential for increasing CAT yield as well as CFPS performance.
+%Comparing the theoretical maximum carbon yield of CAT from ssFBA predictions to those of the kinetic model and experimental measurements suggests that there is potential for increasing CAT yield as well as CFPS performance.
 %Yield performance can be potentially improved by diverting carbon flux away from acetate.
-The model and experimental yields were 36\% of the theoretical maximum and 23\% of a physiologically constrained case.
+Comparing the physiologically realistic carbon yield of CAT from ssFBA predictions to those of the kinetic model and experimental measurements suggests that there is potential for increasing CAT yield as well as CFPS performance.
+%The model and experimental yields were 35\% of the physiologically constrained case.
 Knockouts of \textit{gnd} and phosphate acetyltransferase show that carbon can be diverted away from acetate and toward CAT or other proteins of interest expressed in CFPS.
@@
 \subsubsection*{Calculation of the carbon yield.}
-The CAT carbon yield ($Y_{C}^{CAT}$) was calculated as the ratio of carbon produced as CAT dived by the carbon consumed as reactants (glucose and amino acids):
+The CAT carbon yield ($Y_{C}^{CAT}$) was calculated as the ratio of carbon produced as CAT divided by the carbon consumed as reactants (glucose and amino acids):
 \begin{equation}\label{eqn:yield-definition}
@@
 $\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed (never allowed to be negative), and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
-Arginine or glutamate were not considered in the yield calculations, as no experimental measurements were available for these amino acids.
+For flux balance analysis, arginine and glutamate were excluded from the yield calculations, as no experimental measurements were available for these amino acids.
 Yield of the best-fit parameter set and the experimental data were calculated by setting $\Delta\texttt{CAT}$ equal to the final minus the initial CAT concentration,
 and setting $\Delta m_{i}$ equal to the initial minus the final reactant concentration.
+The individual CAT production and substrate consumption terms for the best-fit set, kinetic models with knockouts, and experimental data are shown in Table~\ref{tbl:yield_breakdown}.
+Consumption of amino acids via CAT synthesis was calculated as a percentage of total net consumption for the best-fit set (Table~\ref{tbl:AA_breakdown}).
+Consumption toward CAT was calculated as CAT production times the stoichiometric coefficient for that amino acid in the CAT synthesis reaction.
+Total net consumption was calculated as amino acid concentration at 3 hours minus concentration at 0 hours, and was excluded if negative.
 
@@
 
+%\begin{center}
+%  \captionof{table}{Your caption here}
+%    \footnotesize
+%%    \small
+%    \begin{tabular}{|c|c|c|c|c|c|c|}
+%      \hline
+%        & Best-fit & $\Delta$cyd & $\Delta$nuo & $\Delta$app & $\Delta$cyd $\Delta$nuo $\Delta$app & Data \\ \hline \hline
+%        \multicolumn{7}{|c|}{Carbon Produced (mM)} \\ \hline
+%        CAT & 20.9 & 6.5 & 18.1 & 21.4 & 5.1 & 21.6 \\ \hline \hline
+%        \multicolumn{7}{|c|}{Carbon Consumed (mM)} \\ \hline
+%        GLC & 215.4 & 215.4 & 215.4 & 215.4 & 159.8 & 215.4 \\ \hline
+%        ALA & 0.0 & 0.0 & 1.7 & 0.0 & 0.0 & 0.0 \\ \hline
+%        ARG & 10.2 & 0.9 & 1.1 & 9.9 & 1.3 & - \\ \hline
+%        ASN & 6.2 & 6.3 & 6.2 & 6.2 & 6.3 & 6.3 \\ \hline
+%        ASP & 7.5 & 0.0 & 3.9 & 7.5 & 0.0 & 9.6 \\ \hline
+%        CYS & 3.0 & 2.9 & 3.0 & 3.1 & 2.9 & 3.7 \\ \hline
+%        GLN & 0.0 & 1.8 & 0.0 & 0.0 & 2.7 & 0.0 \\ \hline
+%        GLU & 492.6 & 505.1 & 528.2 & 505.6 & 501.8 & - \\ \hline
+%        GLY & 3.1 & 1.1 & 2.6 & 3.1 & 0.9 & 1.5 \\ \hline
+%        HIS & 0.2 & 0.4 & 1.1 & 0.2 & 0.3 & 0.0 \\ \hline
+%        ILE & 1.0 & 0.3 & 0.8 & 1.0 & 0.2 & 1.7 \\ \hline
+%        LEU & 1.4 & 0.4 & 1.2 & 1.4 & 0.3 & 2 \\ \hline
+%        LYS & 10.7 & 13.2 & 13.1 & 10.7 & 13.2 & 8.3 \\ \hline
+%        MET & 0.8 & 0.2 & 0.7 & 0.8 & 0.2 & 2.9 \\ \hline
+%        PHE & 3.2 & 1.0 & 2.8 & 3.3 & 0.8 & 1.6 \\ \hline
+%        PRO & 2.4 & 0.2 & 0.7 & 2.4 & 0.2 & 1.9 \\ \hline
+%        SER & 2.5 & 2.1 & 2.4 & 2.5 & 2.1 & 1.8 \\ \hline
+%        THR & 3.4 & 2.9 & 3.3 & 3.4 & 2.8 & 2.8 \\ \hline
+%        TRP & 1.0 & 0.3 & 0.8 & 1.0 & 0.2 & 1.2 \\ \hline
+%        TYR & 1.1 & 0.4 & 1.1 & 1.1 & 0.4 & 0.6 \\ \hline
+%        VAL & 1.4 & 0.4 & 1.2 & 1.5 & 0.4 & 2.4 \\ \hline
+%        $\Sigma$ & 767.1 & 755.3 & 791.3 & 780.1 & 696.8 & - \\ \hline
+%        $\Sigma$ w/o ARG,GLU & 264.3 & 249.3 & 262.0 & 264.6 & 193.7 & 263.7 \\ \hline \hline
+%        Yield & 2.7\% & 0.9\% & 2.3\% & 2.7\% & 0.7\% & - \\ \hline
+%        Yield w/o ARG, GLU & 7.9\% & 2.6\% & 6.9\% & 8.1\% & 2.7\% & 8.2\% \\ \hline
+%    \end{tabular}
+%\end{center}
+
+%\noindent% Insures there's no paragraph indent
+%\begin{minipage}{\textwidth}% Minipage has width exactly the same as the text block
+%  \centering% Centers the contents of the minipage
+%  This is an interesting table.% Your caption goes here.
+%  \bigskip
+%  \bigskip
+%  \footnotesize
+%  \begin{tabular}{|c|c|c|c|c|c|c|}
+%      \hline
+%        & Best-fit & $\Delta$cyd & $\Delta$nuo & $\Delta$app & $\Delta$cyd $\Delta$nuo $\Delta$app & Data \\ \hline \hline
+%        \multicolumn{7}{|c|}{Carbon Produced (mM)} \\ \hline
+%        CAT & 20.9 & 6.5 & 18.1 & 21.4 & 5.1 & 21.6 \\ \hline \hline
+%        \multicolumn{7}{|c|}{Carbon Consumed (mM)} \\ \hline
+%        GLC & 215.4 & 215.4 & 215.4 & 215.4 & 159.8 & 215.4 \\ \hline
+%        ALA & 0.0 & 0.0 & 1.7 & 0.0 & 0.0 & 0.0 \\ \hline
+%        ARG & 10.2 & 0.9 & 1.1 & 9.9 & 1.3 & - \\ \hline
+%        ASN & 6.2 & 6.3 & 6.2 & 6.2 & 6.3 & 6.3 \\ \hline
+%        ASP & 7.5 & 0.0 & 3.9 & 7.5 & 0.0 & 9.6 \\ \hline
+%        CYS & 3.0 & 2.9 & 3.0 & 3.1 & 2.9 & 3.7 \\ \hline
+%        GLN & 0.0 & 1.8 & 0.0 & 0.0 & 2.7 & 0.0 \\ \hline
+%        GLU & 492.6 & 505.1 & 528.2 & 505.6 & 501.8 & - \\ \hline
+%        GLY & 3.1 & 1.1 & 2.6 & 3.1 & 0.9 & 1.5 \\ \hline
+%        HIS & 0.2 & 0.4 & 1.1 & 0.2 & 0.3 & 0.0 \\ \hline
+%        ILE & 1.0 & 0.3 & 0.8 & 1.0 & 0.2 & 1.7 \\ \hline
+%        LEU & 1.4 & 0.4 & 1.2 & 1.4 & 0.3 & 2 \\ \hline
+%        LYS & 10.7 & 13.2 & 13.1 & 10.7 & 13.2 & 8.3 \\ \hline
+%        MET & 0.8 & 0.2 & 0.7 & 0.8 & 0.2 & 2.9 \\ \hline
+%        PHE & 3.2 & 1.0 & 2.8 & 3.3 & 0.8 & 1.6 \\ \hline
+%        PRO & 2.4 & 0.2 & 0.7 & 2.4 & 0.2 & 1.9 \\ \hline
+%        SER & 2.5 & 2.1 & 2.4 & 2.5 & 2.1 & 1.8 \\ \hline
+%        THR & 3.4 & 2.9 & 3.3 & 3.4 & 2.8 & 2.8 \\ \hline
+%        TRP & 1.0 & 0.3 & 0.8 & 1.0 & 0.2 & 1.2 \\ \hline
+%        TYR & 1.1 & 0.4 & 1.1 & 1.1 & 0.4 & 0.6 \\ \hline
+%        VAL & 1.4 & 0.4 & 1.2 & 1.5 & 0.4 & 2.4 \\ \hline
+%        $\Sigma$ & 767.1 & 755.3 & 791.3 & 780.1 & 696.8 & - \\ \hline
+%        $\Sigma$ w/o ARG,GLU & 264.3 & 249.3 & 262.0 & 264.6 & 193.7 & 263.7 \\ \hline \hline
+%        Yield & 2.7\% & 0.9\% & 2.3\% & 2.7\% & 0.7\% & - \\ \hline
+%        Yield w/o ARG, GLU & 7.9\% & 2.6\% & 6.9\% & 8.1\% & 2.7\% & 8.2\% \\ \hline
+%    \end{tabular}
+%\end{minipage}
+
+\begin{table}
+\centering
+    \caption{CAT carbon yield breakdown for best-fit set, knockouts, and experimental data.
+Carbon produced as CAT, carbon consumed as glucose and each amino acid, sum of consumed species (with and without arginine and glutamate), and yield  (with and without arginine and glutamate).}
+    \renewcommand{\arraystretch}{1.3}
+    \begin{tabular}{lrrrrrr} \toprule
+        Carbon Produced (mM) & Best-fit & $\Delta$cyd & $\Delta$nuo & $\Delta$app & $\Delta$cyd $\Delta$nuo $\Delta$app & Data \\ \hline
+        CAT & 20.9 & 6.5 & 18.1 & 21.4 & 5.1 & 21.6 \\ \midrule
+        Carbon Consumed (mM) \\ \hline
+        GLC & 215.4 & 215.4 & 215.4 & 215.4 & 159.8 & 215.4 \\ \hline
+        ALA & 0.0 & 0.0 & 1.7 & 0.0 & 0.0 & 0.0 \\ \hline
+        ARG & 10.2 & 0.9 & 1.1 & 9.9 & 1.3 & - \\ \hline
+        ASN & 6.2 & 6.3 & 6.2 & 6.2 & 6.3 & 6.3 \\ \hline
+        ASP & 7.5 & 0.0 & 3.9 & 7.5 & 0.0 & 9.6 \\ \hline
+        CYS & 3.0 & 2.9 & 3.0 & 3.1 & 2.9 & 3.7 \\ \hline
+        GLN & 0.0 & 1.8 & 0.0 & 0.0 & 2.7 & 0.0 \\ \hline
+        GLU & 492.6 & 505.1 & 528.2 & 505.6 & 501.8 & - \\ \hline
+        GLY & 3.1 & 1.1 & 2.6 & 3.1 & 0.9 & 1.5 \\ \hline
+        HIS & 0.2 & 0.4 & 1.1 & 0.2 & 0.3 & 0.0 \\ \hline
+        ILE & 1.0 & 0.3 & 0.8 & 1.0 & 0.2 & 1.7 \\ \hline
+        LEU & 1.4 & 0.4 & 1.2 & 1.4 & 0.3 & 2 \\ \hline
+        LYS & 10.7 & 13.2 & 13.1 & 10.7 & 13.2 & 8.3 \\ \hline
+        MET & 0.8 & 0.2 & 0.7 & 0.8 & 0.2 & 2.9 \\ \hline
+        PHE & 3.2 & 1.0 & 2.8 & 3.3 & 0.8 & 1.6 \\ \hline
+        PRO & 2.4 & 0.2 & 0.7 & 2.4 & 0.2 & 1.9 \\ \hline
+        SER & 2.5 & 2.1 & 2.4 & 2.5 & 2.1 & 1.8 \\ \hline
+        THR & 3.4 & 2.9 & 3.3 & 3.4 & 2.8 & 2.8 \\ \hline
+        TRP & 1.0 & 0.3 & 0.8 & 1.0 & 0.2 & 1.2 \\ \hline
+        TYR & 1.1 & 0.4 & 1.1 & 1.1 & 0.4 & 0.6 \\ \hline
+        VAL & 1.4 & 0.4 & 1.2 & 1.5 & 0.4 & 2.4 \\ \midrule
+        Sum & 767.1 & 755.3 & 791.3 & 780.1 & 696.8 & - \\ \hline
+        Sum w/o ARG,GLU & 264.3 & 249.3 & 262.0 & 264.6 & 193.7 & 263.7 \\ \midrule
+        Yield & 2.7\% & 0.9\% & 2.3\% & 2.7\% & 0.7\% & - \\ \hline
+        Yield w/o ARG, GLU & 7.9\% & 2.6\% & 6.9\% & 8.1\% & 2.7\% & 8.2\% \\ \bottomrule
+    \end{tabular}
+\label{tbl:yield_breakdown}
+\end{table}
+
+\begin{table}
+\centering
+    \caption{Amino acid consumption toward CAT as a percentage of total amino acid consumption, for the best-fit set.
+Total consumption defined as decrease in amino acid concentration over 3 hours, such that for amino acids with synthesis active, percentage may be greater than 100\%.
+Alanine and glutamine did not decrease over the timecourse.}
+    \renewcommand{\arraystretch}{1.3}
+    \begin{tabular}{lrrr} \toprule
+        Species & Total Consumption & Consumption Toward CAT & Percentage \\ \hline
+%        ALA  & -3.852  & 0.267  & -6.9\% \\ \hline
+        ALA  & - & 0.27 & - \\ \hline
+%        ARG  & 1.697  & 0.178  & 10.5\% \\ \hline
+%        ASN  & 1.547  & 0.089  & 5.8\% \\ \hline
+%        ASP  & 1.875  & 0.214  & 11.4\% \\ \hline
+%        CYS  & 1.016  & 0.089  & 8.8\% \\ \hline
+        ARG  & 1.70  & 0.18  & 11\% \\ \hline
+        ASN  & 1.55  & 0.09  & 6\% \\ \hline
+        ASP  & 1.87  & 0.21  & 11\% \\ \hline
+        CYS  & 1.02  & 0.09  & 9\% \\ \hline
+%        GLN  & -2.278  & 0.214  & -9.4\% \\ \hline
+        GLN  & - & 0.21 & - \\ \hline
+%        GLU  & 98.526  & 0.232  & 0.2\% \\ \hline
+%        GLY  & 1.545  & 0.178  & 11.5\% \\ \hline
+%        HIS  & 0.036  & 0.214  & 587.9\% \\ \hline
+%        ILE  & 0.16  & 0.16  & 100\% \\ \hline
+%        LEU  & 0.231  & 0.232  & 100.2\% \\ \hline
+%        LYS  & 1.785  & 0.214  & 12.0\% \\ \hline
+%        MET  & 0.16  & 0.16  & 100\% \\ \hline
+%        PHE  & 0.356  & 0.356  & 100\% \\ \hline
+%        PRO  & 0.481  & 0.125  & 26\% \\ \hline
+%        SER  & 0.83  & 0.178  & 21.5\% \\ \hline
+%        THR  & 0.857  & 0.232  & 27\% \\ \hline
+%        TRP  & 0.089  & 0.089  & 100\% \\ \hline
+%        TYR  & 0.12  & 0.196  & 164\% \\ \hline
+%        VAL  & 0.285  & 0.285  & 100\% \\ \hline
+        GLU  & 98.53  & 0.23  & 0.2\% \\ \hline
+        GLY  & 1.55  & 0.18  & 12\% \\ \hline
+        HIS  & 0.04  & 0.21  & 588\% \\ \hline
+        ILE  & 0.16  & 0.16  & 100\% \\ \hline
+        LEU  & 0.23  & 0.23  & 100\% \\ \hline
+        LYS  & 1.79  & 0.21  & 12\% \\ \hline
+        MET  & 0.16  & 0.16  & 100\% \\ \hline
+        PHE  & 0.36  & 0.36  & 100\% \\ \hline
+        PRO  & 0.48  & 0.12  & 26\% \\ \hline
+        SER  & 0.83  & 0.18  & 21\% \\ \hline
+        THR  & 0.86  & 0.23  & 27\% \\ \hline
+        TRP  & 0.09  & 0.09  & 100\% \\ \hline
+        TYR  & 0.12  & 0.20  & 164\% \\ \hline
+        VAL  & 0.29  & 0.29  & 100\% \\ \bottomrule
+    \end{tabular}
+\label{tbl:AA_breakdown}
+\end{table}
+
 \clearpage

### 2016-12-03 Paper_v1.tex
Dec 2 updates

@@
 Next, to gauge the performance of the CFPS reaction, we next calculated the CAT carbon yield for three classes of constraints:
-(i) theoretical maximum glucose, amino acid and oxygen upper bounds, and no transcriptional/translational constraints;
-(ii) theoretical maximum glucose, amino acid and oxygen upper bounds, and realistic transcriptional/translational constraints; and
-(iii) metabolite fluxes constrained by the CAT data, and realistic transcriptional/translational constraints (Fig.~\ref{fig:CATProdYield}B).
-The unconstrained theoretical maximum CAT carbon yield (case i) was 36.3\% $\pm$ 2.0\% (Fig.~\ref{fig:CATProdYield}B, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, and transcription and translation were unbounded.
-On the other hand, for realistic transcription and translation constraints (case ii), the optimal CAT carbon yield decreased to 22.6\% $\pm$ 3.0\% (Fig.~\ref{fig:CATProdYield}B, middle).
-Lastly, when both realistic metabolite and transcription/translation constraints were applied (case iii), the predicted carbon yield was 6.2\% $\pm$ 2.0\%
+(i) theoretical maximum glucose, amino acid and oxygen upper bounds, and realistic transcriptional/translational constraints;
+(ii) theoretical maximum glucose, amino acid and oxygen upper bounds, realistic transcriptional/translational constraints and knockouts of amino acid synthesis reactions of amino acids supplemented in the \textit{E.
+coli} extract preparation.
+(iii) metabolite fluxes constrained by the CAT data, and realistic transcriptional/translational constraints and knockouts of amino acid synthesis reactions of amino acids supplemented in the \textit{E.
+coli} extract preparation (Fig.~\ref{fig:CATProdYield}B).
+The physiological theoretical maximum CAT carbon yield (case i) was 21.8\% $\pm$ 2.8\% (Fig.~\ref{fig:CATProdYield}B, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, with bounded transcription and translation rates (96\% without glucose contribution).
+On the other hand,(case ii) the optimal CAT carbon yield without certain amino acid synthesis reactions decreased to 18.2\% $\pm$ 3.0\% (Fig.~\ref{fig:CATProdYield}B, middle).
+Lastly, when transcription/translation constraints were applied with experimental measurements of metabolite measurements (case iii), the predicted carbon yield was 6.0\% $\pm$ 2.0\%
 %By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield of 8.6\% $\pm$ 0.4\%, equivalent to 23\% of the theoretical maximum (i) and 35\% of the physiological case (ii).
@@
 To investigate the differences in carbon yields, we compared the flux distributions predicted by ssFBA simulations for the different constraint cases  (Fig.~\ref{fig:Network}).
-The constrained cases (ii \& iii) heavily utilized the first step in the pentose phosphate pathway to generate NADPH;
+All cases heavily utilized the first step in the pentose phosphate pathway to generate NADPH;
 the carbon flux then continued through the Entner–Doudoroff pathway toward pyruvate.
-For case ii, the majority of the flux proceeded toward acetate accumulation, whereas in case iii, the flux accumulated as pyruvate, lactate, and acetate with some flux through the TCA cycle.
-In comparison, the unconstrained case (i) showed the majority of flux traveling through glycolysis towards pyruvate, leading to accumulation of lactate, acetate and malate.
+The majority of the flux proceeded toward acetate accumulation, whereas in case ii, the flux accumulated as lactate and acetate, meanwhile for case iii the flux was distributed between pyruvate, lactate, and acetate.
 In all cases, the energy source was primarily oxidative phosphorylation, and to a lesser extent the TCA cycle.
-However, the accumulation of acetate and lactate signifies that the system is not operating at its highest efficiency.
+However, the accumulation of pyruvate, lacatate, and acetate signifies that the system is not operating at its highest efficiency (case iii).
 The system produced NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to power oxidative phosphorylation.
@@
 To investigate this further, we simulated potential knockouts with constrained transcription/translation rates  (Fig.~\ref{fig:flux_ko}).
-Knocking out the \textit{gnd} reaction, the first step in Entner-Doudoroff pathway, decreased acetate flux by about half.
-In addition, less uptake of amino acids were required which increased the carbon yield of CAT by 2.7\% (up to approximately 25.3\% $\pm$ 3.6\%) compared to the control (no knockouts).
-The simulation showed an increase in oxidative phosphorylation flux and the flux splitting between glycolysis and pentose phosphate pathway.
+Knocking out the \textit{gnd} reaction, the first step in Entner-Doudoroff pathway, decreased acetate flux by about twenty percent
+In addition, less uptake of amino acids were required which increased the carbon yield of CAT by 1.1\% (up to approximately 22.9\% $\pm$ 2.4\%) compared to the control (no knockouts) for case i.
+The simulation showed an increase in oxidative phosphorylation flux and the flux heavily utilizing glycolysis instead of pentose phosphate pathway.
 A second simulation with both \textit{gnd} and phosphate acetyltransferase knocked out,
-showed less carbon was needed to meet the translation bound, which increased the carbon yield by 4\% compared to the control (up to approximately 26.7\% $\pm$ 3.4\%).
+showed very similar results as the first single knockout.
 In the dual knockout, flux towards acetate was almost negligible with some coming from amino acid degradation.
-Sixty-one percent of the carbon flux traveled through glycolysis, while the remaining traveled through pentose phosphate pathway.
-This was similar to the flux distribution of the unconstrained case (case i).
-Taken together, the dual knockout decreased acetate production and required less amino acid consumption, thus it is a promising strategy to increase the CAT carbon yield.
+Taken together, a \textit{gnd} knockout decreased acetate production and required less amino acid consumption, thus it is a promising strategy to increase the CAT carbon yield.
 
@@
 %The model and experimental yields were 35\% of the physiologically constrained case.
-Knockouts of \textit{gnd} and phosphate acetyltransferase show that carbon can be diverted away from acetate and toward CAT or other proteins of interest expressed in CFPS.
+A knockout of \textit{gnd} and shows that carbon can be diverted away from acetate and toward CAT or other proteins of interest expressed in CFPS.
 Another limitation to be addressed in CFPS is the transcription and translation description, since protein production is ultimately bounded by these kinetic rates.
@@
 $\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed (never allowed to be negative), and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
-For flux balance analysis, arginine and glutamate were excluded from the yield calculations, as no experimental measurements were available for these amino acids.
+Arginine and glutamate were excluded from the yield calculations for the experimental yield calculation, as no experimental measurements were available for these amino acids.
 Yield of the best-fit parameter set and the experimental data were calculated by setting $\Delta\texttt{CAT}$ equal to the final minus the initial CAT concentration,
@@
 \label{fig:Network}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=0.75\textwidth]{./Figures/CAT_flux_ko.pdf}
-\caption{Flux profile with knockouts for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, and oxidative phosphorylation.
-Sequence-specific FBA flux value (mean ± standard deviation) across ensemble for 1 hr, normalized to glucose uptake flux.
-Flux distribution for three different cases: constrained by transcription and translation rates (top row), constrained by transcription and translation rates with \textit{gnd} knocked out (second row), and constrained by transcription and translation rates with \textit{gnd} and phosphate acetyltransferase knocked out.}
-\label{fig:flux_ko}
 \end{figure}

### 2016-12-07 Paper_v1.tex
Dec6

--- 2016-12-02 Paper_v1.tex
@@
 This makes sense, as glucose powers the entire metabolism during the first hour, including energy species synthesis and amino acid synthesis.
+The knockouts in oxidative phosphorylation also reduced the energy efficiency of CAT production  (Fig.~\ref{fig:oxphos_ko}).
+The best-fit set, $\Delta$app, and experimental dataset had efficiencies of 1.8-1.9\%, $\Delta$nuo had an efficiency of 1.6\%, and $\Delta$cyd and $\Delta$cyd$\Delta$nuo$\Delta$app had efficiencies of 0.6\%.
+This was in line with the effect of knockouts on yield, which in turn was more or less proportional to the total amount of CAT produced in each case.
+Taken together, these results show that oxidative phosphorylation, particularly \textit{cyd}, is important to efficient CAT production.
 
@@
 Total net consumption was calculated as amino acid concentration at 3 hours minus concentration at 0 hours, and was excluded if negative.
+
+Energy efficiency was also calculated for best-fit set, knockouts, and experimental data, formulated as:
+\begin{equation}\label{eqn:energy-efficiency-definition}
+	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2*\rm {ATP}_{CAT}+\rm {GTP}_{CAT}\right)}{\Delta\texttt{GLC}\cdot \rm {ATP}_{GLC}}
+\end{equation}
+where $\rm Efficiency$ denotes the energy efficiency of CAT production, $\rm {ATP}_{CAT}$ denotes the stoichiometric coefficient of ATP in CAT synthesis (multiplied by 2 because AMP, rather than ADP, is a product of CAT synthesis), $\rm {GTP}_{CAT}$ denotes the stoichiometric coefficient of GTP in CAT synthesis, and $\rm {ATP}_{GLC}$ denotes the number of ATP molecules ideally produced from one molecule of glucose.
+$\rm {ATP}_{CAT}$ = 219, $\rm {GTP}_{CAT}$ = 438, $\rm {ATP}_{GLC}$ = 24.
 
@@
     \caption{CAT carbon yield breakdown for best-fit set, knockouts, and experimental data.
-Carbon produced as CAT, carbon consumed as glucose and each amino acid, sum of consumed species (with and without arginine and glutamate), and yield  (with and without arginine and glutamate).}
-    \renewcommand{\arraystretch}{1.3}
+Carbon produced as CAT, carbon consumed as glucose and each amino acid, sum of consumed species (with and without arginine and glutamate), and yield  (with and without arginine and glutamate).
+Energy efficiency for best-fit set, knockouts, and experimental data.}
+    \renewcommand{\arraystretch}{1.25}
     \begin{tabular}{lrrrrrr} \toprule
@@
         Yield & 2.7\% & 0.9\% & 2.3\% & 2.7\% & 0.7\% & - \\ \hline
-        Yield w/o ARG, GLU & 7.9\% & 2.6\% & 6.9\% & 8.1\% & 2.7\% & 8.2\% \\ \bottomrule
+        Yield w/o ARG, GLU & 7.9\% & 2.6\% & 6.9\% & 8.1\% & 2.7\% & 8.2\% \\ \midrule
+        Energy Efficiency & 1.8\% & 0.6\% & 1.6\% & 1.8\% & 0.6\% & 1.9\% \\ \bottomrule
+%        Efficiencies based on ATP number of 24: 1.811977 & 0.558704 & 1.565814 & 1.849871 & 0.600006 & 1.873067
     \end{tabular}
@@
 Alanine and glutamine did not decrease over the timecourse.}
-    \renewcommand{\arraystretch}{1.3}
+    \renewcommand{\arraystretch}{1.25}
     \begin{tabular}{lrrr} \toprule

### 2016-12-07 Paper_v1.tex
Dec6_4

@@
 The knockouts in oxidative phosphorylation also reduced the energy efficiency of CAT production  (Fig.~\ref{fig:oxphos_ko}).
-The best-fit set, $\Delta$app, and experimental dataset had efficiencies of 1.8-1.9\%, $\Delta$nuo had an efficiency of 1.6\%, and $\Delta$cyd and $\Delta$cyd$\Delta$nuo$\Delta$app had efficiencies of 0.6\%.
+The best-fit set, $\Delta$app, and experimental dataset had efficiencies of 1.3\%, $\Delta$nuo had an efficiency of 1.1\%, and $\Delta$cyd and $\Delta$cyd$\Delta$nuo$\Delta$app had efficiencies of 0.4\%.
 This was in line with the effect of knockouts on yield, which in turn was more or less proportional to the total amount of CAT produced in each case.
@@
         Yield w/o ARG, GLU & 7.9\% & 2.6\% & 6.9\% & 8.1\% & 2.7\% & 8.2\% \\ \midrule
-        Energy Efficiency & 1.8\% & 0.6\% & 1.6\% & 1.8\% & 0.6\% & 1.9\% \\ \bottomrule
+%        Energy Efficiency & 1.8\% & 0.6\% & 1.6\% & 1.8\% & 0.6\% & 1.9\% \\ \bottomrule
+        Energy Efficiency & 1.3\% & 0.4\% & 1.1\% & 1.3\% & 0.4\% & 1.3\% \\ \bottomrule
 %        Efficiencies based on ATP number of 24: 1.811977 & 0.558704 & 1.565814 & 1.849871 & 0.600006 & 1.873067
+%        Efficiencies based on ATP number of 24: 1.28348 & 0.395749 & 1.10912 & 1.31033 & 0.425004 & 1.32676
     \end{tabular}

### 2016-12-07 Paper_v1.tex
Dec6_5

@@
 coli} extract preparation (Fig.~\ref{fig:CATProdYield}B).
-The physiological theoretical maximum CAT carbon yield (case i) was 21.8\% $\pm$ 2.8\% (Fig.~\ref{fig:CATProdYield}B, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, with bounded transcription and translation rates (96\% without glucose contribution).
-On the other hand,(case ii) the optimal CAT carbon yield without certain amino acid synthesis reactions decreased to 18.2\% $\pm$ 3.0\% (Fig.~\ref{fig:CATProdYield}B, middle).
-Lastly, when transcription/translation constraints were applied with experimental measurements of metabolite measurements (case iii), the predicted carbon yield was 6.0\% $\pm$ 2.0\%
+The physiological theoretical maximum CAT carbon yield (case i) was 21.8\% $\pm$ 2.8\% (Fig.~\ref{fig:CATProdYield}B, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, with bounded transcription and translation rates (96\% without glucose contribution in the carbon yield calculation).
+For case ii, the optimal CAT carbon yield decreased to 18.2\% $\pm$ 3.0\% (Fig.~\ref{fig:CATProdYield}B, middle).
+Lastly, when metabolite constraints were applied with experimental measurements of measurements (case iii), the predicted carbon yield was 6.0\% $\pm$ 2.0\%
 %By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield of 8.6\% $\pm$ 0.4\%, equivalent to 23\% of the theoretical maximum (i) and 35\% of the physiological case (ii).
-By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield (without arginine and glutamate) of 7.9\%, equivalent to 35\% of the physiological case (ii).
+By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield (without arginine and glutamate) of 7.9\%, equivalent to 36\% of the optimal case (i).
 The experimental dataset had a CAT carbon yield of 8.2\%, similar to both the kinetic model and the experimentally constrained ssFBA calculation (case iii).
+However, the energy efficiency for case i was 18.1\% and 15.5\% for the second case, while for the experimentally constrained case it was 3.9\%.  
 Thus, while the CFPS reaction was not optimal, the ssFBA calculations suggested that an approximately three-fold increase in carbon yield was theoretically possible.
@@
 
-\subsubsection*{Calculation of the carbon yield.}
+\subsubsection*{Quantification of uncertainty.}
+An ensemble of 100 sets of flux distributions was calculated for three different cases:  constrained by transcription/translation rates, constrained by transcription/translation rates without amino acid synthesis reactions, and constrained by transcription/translation rates and experimental measurements without amino acid synthesis reactions.
+For the first case, all rates were left unbounded, except the specific glucose uptake rate, transcription and translation rate.
+An ensemble of flux distributions was then calculated by randomly sampling the maximum specific glucose uptake rate from within a range of 30 to 40 mM/h, determined from experimental data and randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature..
+For the second case, an ensemble was generated by randomly sampling the same parameters as the first case, however certain amino acid synthesis reactions were removed from the network.
+This included all the amino acids that were present in the preparation of the \textit{E.
+coli} extract (alanine, arginine, aspartate, cysteine, glutamate, glutamine and serine were excluded from the media), thus reactions producing the excluded amino acids were left in the network.
+RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/sec \cite{2005_underwood_biotech, Garamella:2016aa}.
+For the third case, the ensemble was generated as in the second case, in addition to the lower and upper bounds on the fluxes for the data-informed metabolites were sampled within the range given by the experimental noise.
+This included the data for glucose, organic acids, energy species, and amino acids; CAT was not constrained by experimental data, but by the transcription/translation rates as stated above.
+
+\subsection*{Calculation of the carbon yield.}
 The CAT carbon yield ($Y_{C}^{CAT}$) was calculated as the ratio of carbon produced as CAT divided by the carbon consumed as reactants (glucose and amino acids):
@@
 
+\subsection*{Calculation of energy efficiency.}
 Energy efficiency was also calculated for best-fit set, knockouts, and experimental data, formulated as:
 \begin{equation}\label{eqn:energy-efficiency-definition}
-	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2*\rm {ATP}_{CAT}+\rm {GTP}_{CAT}\right)}{\Delta\texttt{GLC}\cdot \rm {ATP}_{GLC}}
+	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2\cdot\rm {ATP}_{CAT}+\rm {GTP}_{CAT}\right)}{\Delta\texttt{GLC}\cdot \rm {ATP}_{GLC}}
 \end{equation}
 where $\rm Efficiency$ denotes the energy efficiency of CAT production, $\rm {ATP}_{CAT}$ denotes the stoichiometric coefficient of ATP in CAT synthesis (multiplied by 2 because AMP, rather than ADP, is a product of CAT synthesis), $\rm {GTP}_{CAT}$ denotes the stoichiometric coefficient of GTP in CAT synthesis, and $\rm {ATP}_{GLC}$ denotes the number of ATP molecules ideally produced from one molecule of glucose.
-$\rm {ATP}_{CAT}$ = 219, $\rm {GTP}_{CAT}$ = 438, $\rm {ATP}_{GLC}$ = 24.
-
-\subsubsection*{Quantification of uncertainty.}
-An ensemble of 100 sets of flux distributions was calculated for three different cases: unconstrained, constrained by transcription/translation rates, and constrained by transcription/translation rates and experimental measurements.
-For the unconstrained case, all rates were left unbounded, except the specific glucose uptake rate.
-An ensemble of flux distributions was then calculated by randomly sampling the maximum specific glucose uptake rate from within a range of 30 to 40 mM/h, determined from experimental data.
-For the case constrained by transcription/translation rates, an ensemble was generated by randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
-RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/sec \cite{2005_underwood_biotech, Garamella:2016aa}.
-For the case constrained by transcription/translation rates and experimental measurements, the lower and upper bounds on the fluxes for the data-informed metabolites were sampled within the range given by the experimental noise.
-This included the data for glucose, organic acids, energy species, and amino acids; CAT was not constrained by experimental data, but by the transcription/translation rates as stated above.
-
+$\rm {ATP}_{CAT}$ = 219, $\rm {GTP}_{CAT}$ = 438, $\rm {ATP}_{GLC}$ = 17.
+
+Energy efficiency was also calculated for sequence specific flux balance analysis, formulated as:
+\begin{equation}\label{eqn:energy-efficiency-definition}
+	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2\left({ATP}_{TX}+\rm {CTP}_{TX}+\rm {GTP}_{TX}+\rm {UTP}_{TX}\right)+\rm2\cdot\rm {ATP}_{TL}+\rm {GTP}_{TL}\right)}{\Delta\texttt{GLC}\cdot \rm {ATP}_{GLC}}
+\end{equation}
+where $\rm Efficiency$ denotes the energy efficiency of CAT production, $\rm {ATP}_{TX}$, $\rm {CTP}_{TX}$, $\rm {GTP}_{TX}$, $\rm {UTP}_{TX}$ denote the stoichiometric coefficient of each energy species for CAT transcription, $\rm {ATP}_{TL}$ and $\rm {GTP}_{TL}$ denote the stoichiometric coefficient of ATP and GTP, respectively, in CAT translation, and $\rm {ATP}_{GLC}$ denotes the number of ATP molecules produced from one molecule of glucose.
+$\rm {ATP}_{TX}$ = 176, $\rm {CTP}_{TX}$ = 144, $\rm {GTP}_{TX}$ = 151, $\rm {UTP}_{TX}$ = 189, $\rm {ATP}_{TL}$ = 219, $\rm {GTP}_{TL}$ = 438, $\rm {ATP}_{GLC}$ = 17.
 %The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52)
@@
 B.
-Theoretical maximum carbon yield of CAT calcualted by ssFBA for three different cases: unconstrained except for glucose uptake (black), constrained by transcription/translation (TX/TL) rates (grey), and constrained by transcription/translation (TX/TL) rates and experimental measurements where available (light grey).
+Theoretical maximum carbon yield of CAT calcualted by ssFBA for three different cases: constrained by transcription/translation (TX/TL) rates (black), same as previous but without amino acid synthesis reactions (grey), and same as previous but constrained by experimental measurements where available (light grey).
 Error bars represent standard deviation of the ensemble.}
@@
 Sequence-specific FBA flux value (mean ± standard deviation) across ensemble for 1 hr, normalized to glucose uptake flux.
-Flux distribution for three different cases: unconstrained except for glucose uptake (top row), constrained by transcription and translation rates (second row), and constrained by transcription, translation rates and experimental measurements where available (bottom row).}
+Flux distribution for three different cases: constrained by transcription and translation rates (top row), same as previous but without amino acid synthesis reactions  (second row), and same as previous but constrained by experimental measurements where available (bottom row).}
 \label{fig:Network}

### 2016-12-07 Paper_v1.tex
Dec7

@@
 The knockouts in oxidative phosphorylation also reduced the energy efficiency of CAT production  (Fig.~\ref{fig:oxphos_ko}).
-The best-fit set, $\Delta$app, and experimental dataset had efficiencies of 1.3\%, $\Delta$nuo had an efficiency of 1.1\%, and $\Delta$cyd and $\Delta$cyd$\Delta$nuo$\Delta$app had efficiencies of 0.4\%.
+The best-fit set, $\Delta$app, and experimental dataset had efficiencies of 2.6\%, $\Delta$nuo had an efficiency of 2.2\%, and $\Delta$cyd and $\Delta$cyd$\Delta$nuo$\Delta$app had efficiencies of 0.8\%.
 This was in line with the effect of knockouts on yield, which in turn was more or less proportional to the total amount of CAT produced in each case.
@@
 %        Energy Efficiency & 1.8\% & 0.6\% & 1.6\% & 1.8\% & 0.6\% & 1.9\% \\ \bottomrule
-        Energy Efficiency & 1.3\% & 0.4\% & 1.1\% & 1.3\% & 0.4\% & 1.3\% \\ \bottomrule
-%        Efficiencies based on ATP number of 24: 1.811977 & 0.558704 & 1.565814 & 1.849871 & 0.600006 & 1.873067
-%        Efficiencies based on ATP number of 24: 1.28348 & 0.395749 & 1.10912 & 1.31033 & 0.425004 & 1.32676
+        Energy Efficiency & 2.6\% & 0.8\% & 2.2\% & 2.6\% & 0.8\% & 2.6\% \\ \bottomrule
+%        Efficiencies based on ATP number of 17: 2.558085 & 0.788758 & 2.210560 & 2.611583 & 0.847068 & 2.644330
+(CAT[end]-CAT[1])*(2*219+438)/((glc[1]-glc[end])*17)*100
+
     \end{tabular}

### 2016-12-07 Paper_v1.tex
Dec7_2

@@
 This makes sense, as glucose powers the entire metabolism during the first hour, including energy species synthesis and amino acid synthesis.
-The knockouts in oxidative phosphorylation also reduced the energy efficiency of CAT production  (Fig.~\ref{fig:oxphos_ko}).
-The best-fit set, $\Delta$app, and experimental dataset had efficiencies of 2.6\%, $\Delta$nuo had an efficiency of 2.2\%, and $\Delta$cyd and $\Delta$cyd$\Delta$nuo$\Delta$app had efficiencies of 0.8\%.
-This was in line with the effect of knockouts on yield, which in turn was more or less proportional to the total amount of CAT produced in each case.
+We also investigated the energy efficiency of CAT production across the kinetic models and experimental dataset.
+%The knockouts in oxidative phosphorylation also reduced the energy efficiency of CAT production  (Fig.~\ref{fig:oxphos_ko}).
+The best-fit set, $\Delta$app, and experimental data had energy efficiencies of 2.6\%, $\Delta$nuo had an efficiency of 2.2\%, and $\Delta$cyd and $\Delta$cyd$\Delta$nuo$\Delta$app had efficiencies of 0.8\%.
+The best-fit set, $\Delta$app, and experimental data had energy efficiencies of 2.6\%, calculated as a ratio of CAT production to glucose consumption, each converted into an equivalent number of ATP molecules.
+$\Delta$nuo had an efficiency of 2.2\%, and $\Delta$cyd and $\Delta$cyd$\Delta$nuo$\Delta$app had efficiencies of 0.8\%.
+%This was in line with the effect of knockouts on yield, which in turn was more or less proportional to the total amount of CAT produced in each case.
+The energy efficiencies across the different cases were more or less proportional to the carbon yields, and in turn the total amount of CAT synthesis.
 Taken together, these results show that oxidative phosphorylation, particularly \textit{cyd}, is important to efficient CAT production.
 
-Furthermore, we examined the proportion of amino acid consumption that was due to CAT synthesis in the best-fit set (Table~\ref{tbl:AA_breakdown}).
-Eight of the 20 amino acids had a percentage of 100\% or higher.
+%Furthermore, we examined the proportion of amino acid consumption that was due to CAT synthesis in the best-fit set (Table~\ref{tbl:AA_breakdown}).
+To better understand where carbon flux goes besides toward CAT, we examined the proportion of amino acid consumption that was due to CAT synthesis in the best-fit set (Table~\ref{tbl:AA_breakdown}).
+Eight of the 20 amino acids had a flux-toward-CAT percentage of 100\% or higher.
 For six of these (isoleucine, leucine, methionine, phenylalanine, tryptophan, valine), synthesis was inactive or negligible, meaning that CAT synthesis was the only reaction that consumed these species.
@@
 The experimental dataset had a CAT carbon yield of 8.2\%, similar to both the kinetic model and the experimentally constrained ssFBA calculation (case iii).
-However, the energy efficiency for case i was 18.1\% and 15.5\% for the second case, while for the experimentally constrained case it was 3.9\%.  
-Thus, while the CFPS reaction was not optimal, the ssFBA calculations suggested that an approximately three-fold increase in carbon yield was theoretically possible.
+However, the energy efficiency was 18.1\% for case i and 15.5\% for case ii, while for case iii it was 3.9\%.
+This dramatic decrease in efficiency when fluxes are constrained to data makes sense, as the network is forced toward a multitude of pathways that may not contribute to CAT production.
+However, the case iii efficiency was still about 50\% higher than that of the dataset itself and the best-fit kinetic model (2.6\%).
+This is likely because of the steady-state assumption and the ability to choose the optimum from a variety of flux distributions; meanwhile, the kinetic model diverts flux toward the accumulation of sub-optimal metabolites.
+Thus, while the CFPS reaction was not optimal, the ssFBA calculations suggested that an approximately three-fold increase in carbon yield and a seven-fold increase in energy efficiency were theoretically possible.
 
@@
 %        Efficiencies based on ATP number of 17: 2.558085 & 0.788758 & 2.210560 & 2.611583 & 0.847068 & 2.644330
-(CAT[end]-CAT[1])*(2*219+438)/((glc[1]-glc[end])*17)*100
-
+%(CAT[end]-CAT[1])*(2*219+438)/((glc[1]-glc[end])*17)*100
     \end{tabular}

### 2016-12-08 Paper_v1.tex
Dec8

--- 2016-12-07 Paper_v1.tex
@@
 This makes sense, as glucose powers the entire metabolism during the first hour, including energy species synthesis and amino acid synthesis.
-The knockouts in oxidative phosphorylation also reduced the energy efficiency of CAT production  (Fig.~\ref{fig:oxphos_ko}).
-The best-fit set, $\Delta$app, and experimental dataset had efficiencies of 2.6\%, $\Delta$nuo had an efficiency of 2.2\%, and $\Delta$cyd and $\Delta$cyd$\Delta$nuo$\Delta$app had efficiencies of 0.8\%.
-This was in line with the effect of knockouts on yield, which in turn was more or less proportional to the total amount of CAT produced in each case.
+%We also investigated the energy efficiency of CAT production across the kinetic models and experimental dataset.
+%The best-fit set, $\Delta$app, and experimental data had energy efficiencies of 2.6\%, $\Delta$nuo had an efficiency of 2.2\%, and $\Delta$cyd and $\Delta$cyd$\Delta$nuo$\Delta$app had efficiencies of 0.8\%.
+%The best-fit set, $\Delta$app, and experimental data had energy efficiencies of 2.6\%, calculated as a ratio of CAT production to glucose consumption, each converted into an equivalent number of ATP molecules.
+%$\Delta$nuo had an efficiency of 2.2\%, and $\Delta$cyd and $\Delta$cyd$\Delta$nuo$\Delta$app had efficiencies of 0.8\%.
+%The energy efficiencies across the different cases were more or less proportional to the carbon yields, and in turn the total amount of CAT synthesis.
 Taken together, these results show that oxidative phosphorylation, particularly \textit{cyd}, is important to efficient CAT production.
 
-Furthermore, we examined the proportion of amino acid consumption that was due to CAT synthesis in the best-fit set (Table~\ref{tbl:AA_breakdown}).
-Eight of the 20 amino acids had a percentage of 100\% or higher.
+%Furthermore, we examined the proportion of amino acid consumption that was due to CAT synthesis in the best-fit set (Table~\ref{tbl:AA_breakdown}).
+To better understand where carbon flux goes besides toward CAT, we examined the proportion of amino acid consumption that was due to CAT synthesis in the best-fit set (Table~\ref{tbl:AA_breakdown}).
+Eight of the 20 amino acids had a flux-toward-CAT percentage of 100\% or higher.
 For six of these (isoleucine, leucine, methionine, phenylalanine, tryptophan, valine), synthesis was inactive or negligible, meaning that CAT synthesis was the only reaction that consumed these species.
@@
 For case ii, the optimal CAT carbon yield decreased to 18.2\% $\pm$ 3.0\% (Fig.~\ref{fig:CATProdYield}B, middle).
-Lastly, when metabolite constraints were applied with experimental measurements of measurements (case iii), the predicted carbon yield was 6.0\% $\pm$ 2.0\%
-%By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield of 8.6\% $\pm$ 0.4\%, equivalent to 23\% of the theoretical maximum (i) and 35\% of the physiological case (ii).
+Lastly, when metabolite constraints were applied with experimental measurements of measurements (case iii), the predicted carbon yield was 6.0\% $\pm$ 2.0\%.
 By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield (without arginine and glutamate) of 7.9\%, equivalent to 36\% of the optimal case (i).
 The experimental dataset had a CAT carbon yield of 8.2\%, similar to both the kinetic model and the experimentally constrained ssFBA calculation (case iii).
-However, the energy efficiency for case i was 18.1\% and 15.5\% for the second case, while for the experimentally constrained case it was 3.9\%.  
-Thus, while the CFPS reaction was not optimal, the ssFBA calculations suggested that an approximately three-fold increase in carbon yield was theoretically possible.
+We also investigated the energy efficiency of CAT production for the best-fit set and the three sequence-specific cases, based on the amount of CAT production versus ATP production.
+For cases i and ii the energy efficiencies were 99.8\% and 94.4\%, respectively, while for case iii it was 31.9\%.
+This dramatic decrease in efficiency when fluxes are constrained to data makes sense, as the network is forced toward a multitude of pathways that may not contribute to CAT production.
+However, the case iii efficiency was still much higher than that of the best-fit kinetic model (9.2\%).
+This is likely because of the steady-state assumption and the ability to choose the optimum from a variety of flux distributions; meanwhile, the kinetic model diverts flux toward the accumulation of sub-optimal metabolites.
+Thus, the CFPS reaction was not optimal; the ssFBA calculations suggested that an approximately three-fold increase in carbon yield and an eleven-fold increase in energy efficiency were theoretically possible.
 
@@
 
-\subsection*{Sequence specific calculation of carbon yield.}
+\subsection*{Sequence-specific calculation of carbon yield.}
 We estimated the theoretical maximum CAT carbon yield using sequence-specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
-The sequence specific flux balance analysis problem was formulated as a linear program:
+The sequence-specific flux balance analysis problem was formulated as a linear program:
 \begin{equation}
@@
 
-The objective of the sequence specific flux balance calculation was to maximize the rate of CAT translation, $w_{TL}$.
+The objective of the sequence-specific flux balance calculation was to maximize the rate of CAT translation, $w_{TL}$.
 The total glucose uptake rate was bounded by [0,40 mM/h] according to experimental data;
@@
 The CAT gene and protein sequences were taken from literature.
-The sequence specific flux balance linear program was solved using the GNU Linear Programming Kit (GLPK) v4.52 \cite{GLPK}.
+The sequence-specific flux balance linear program was solved using the GNU Linear Programming Kit (GLPK) v4.52 \cite{GLPK}.
 
@@
 The individual CAT production and substrate consumption terms for the best-fit set, kinetic models with knockouts, and experimental data are shown in Table~\ref{tbl:yield_breakdown}.
-Consumption of amino acids via CAT synthesis was calculated as a percentage of total net consumption for the best-fit set (Table~\ref{tbl:AA_breakdown}).
-Consumption toward CAT was calculated as CAT production times the stoichiometric coefficient for that amino acid in the CAT synthesis reaction.
-Total net consumption was calculated as amino acid concentration at 3 hours minus concentration at 0 hours, and was excluded if negative.
+Total net consumption of amino acids and amino acid consumption via CAT synthesis were calculated for the best-fit set (Table~\ref{tbl:AA_breakdown}).
+Total net consumption was calculated as amino acid concentration at 0 hours minus concentration at 3 hours; it was negative if synthesis outweighed consumption.
+Consumption toward CAT was calculated as CAT concentration at 3 hours minus concentration at 0 hours, times the stoichiometric coefficient for that amino acid in the CAT synthesis reaction.
+The difference between these was defined as other consumption, equal to consumption from reactions other than CAT synthesis minus amino acid production.
 
 \subsection*{Calculation of energy efficiency.}
-Energy efficiency was also calculated for best-fit set, knockouts, and experimental data, formulated as:
+%Energy efficiency was also calculated for best-fit set, knockouts, and experimental data, formulated as:
+%\begin{equation}\label{eqn:energy-efficiency-definition}
+%	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2\cdot\rm {ATP}_{CAT}+\rm {GTP}_{CAT}\right)}{\Delta\texttt{GLC}\cdot \rm {ATP}_{GLC}}
+%\end{equation}
+%where $\rm Efficiency$ denotes the energy efficiency of CAT production, $\rm {ATP}_{CAT}$ denotes the stoichiometric coefficient of ATP in CAT synthesis (multiplied by 2 because AMP, rather than ADP, is a product of CAT synthesis), $\rm {GTP}_{CAT}$ denotes the stoichiometric coefficient of GTP in CAT synthesis, and $\rm {ATP}_{GLC}$ denotes the number of ATP molecules ideally produced from one molecule of glucose.
+%$\rm {ATP}_{CAT}$ = 219, $\rm {GTP}_{CAT}$ = 438, $\rm {ATP}_{GLC}$ = 17.
+
+Energy efficiency was also calculated for the best-fit set:
 \begin{equation}\label{eqn:energy-efficiency-definition}
-	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2\cdot\rm {ATP}_{CAT}+\rm {GTP}_{CAT}\right)}{\Delta\texttt{GLC}\cdot \rm {ATP}_{GLC}}
-\end{equation}
-where $\rm Efficiency$ denotes the energy efficiency of CAT production, $\rm {ATP}_{CAT}$ denotes the stoichiometric coefficient of ATP in CAT synthesis (multiplied by 2 because AMP, rather than ADP, is a product of CAT synthesis), $\rm {GTP}_{CAT}$ denotes the stoichiometric coefficient of GTP in CAT synthesis, and $\rm {ATP}_{GLC}$ denotes the number of ATP molecules ideally produced from one molecule of glucose.
-$\rm {ATP}_{CAT}$ = 219, $\rm {GTP}_{CAT}$ = 438, $\rm {ATP}_{GLC}$ = 17.
-
-Energy efficiency was also calculated for sequence specific flux balance analysis, formulated as:
+	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2\cdot\rm {ATP}_{CAT}+\rm {GTP}_{CAT}\right)}{\Delta\texttt{ATP}}
+\end{equation}
+where $\rm Efficiency$ denotes the energy efficiency of CAT production, $\rm {ATP}_{CAT}$ denotes the stoichiometric coefficient of ATP in CAT synthesis (multiplied by 2 because AMP, rather than ADP, is a product of CAT synthesis), $\rm {GTP}_{CAT}$ denotes the stoichiometric coefficient of GTP in CAT synthesis, and $\rm \Delta ATP$ denotes the amount of ATP production in 3 hours, equal to the sum of ATP-producing fluxes integrated over the timecourse.
+$\rm {ATP}_{CAT}$ = 219, $\rm {GTP}_{CAT}$ = 438.
+%, $\rm \Delta CAT$ = 0.018 mM/h, $\rm \Delta ATP$ = 170 mM/h.
+
+%Energy efficiency was also calculated for sequence-specific flux balance analysis, formulated as:
+%\begin{equation}\label{eqn:energy-efficiency-definition}
+%	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2\left({ATP}_{TX}+\rm {CTP}_{TX}+\rm {GTP}_{TX}+\rm {UTP}_{TX}\right)+\rm2\cdot\rm {ATP}_{TL}+\rm {GTP}_{TL}\right)}{\Delta\texttt{GLC}\cdot \rm {ATP}_{GLC}}
+%\end{equation}
+%where $\rm Efficiency$ denotes the energy efficiency of CAT production, $\rm {ATP}_{TX}$, $\rm {CTP}_{TX}$, $\rm {GTP}_{TX}$, $\rm {UTP}_{TX}$ denote the stoichiometric coefficient of each energy species for CAT transcription, $\rm {ATP}_{TL}$ and $\rm {GTP}_{TL}$ denote the stoichiometric coefficient of ATP and GTP, respectively, in CAT translation, and $\rm {ATP}_{GLC}$ denotes the number of ATP molecules produced from one molecule of glucose.
+%$\rm {ATP}_{TX}$ = 176, $\rm {CTP}_{TX}$ = 144, $\rm {GTP}_{TX}$ = 151, $\rm {UTP}_{TX}$ = 189, $\rm {ATP}_{TL}$ = 219, $\rm {GTP}_{TL}$ = 438, $\rm {ATP}_{GLC}$ = 17.
+
+For sequence-specific flux balance analysis, with a more in-depth biological description of CAT synthesis, energy efficiency was calculated slightly differently:
 \begin{equation}\label{eqn:energy-efficiency-definition}
-	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2\left({ATP}_{TX}+\rm {CTP}_{TX}+\rm {GTP}_{TX}+\rm {UTP}_{TX}\right)+\rm2\cdot\rm {ATP}_{TL}+\rm {GTP}_{TL}\right)}{\Delta\texttt{GLC}\cdot \rm {ATP}_{GLC}}
-\end{equation}
-where $\rm Efficiency$ denotes the energy efficiency of CAT production, $\rm {ATP}_{TX}$, $\rm {CTP}_{TX}$, $\rm {GTP}_{TX}$, $\rm {UTP}_{TX}$ denote the stoichiometric coefficient of each energy species for CAT transcription, $\rm {ATP}_{TL}$ and $\rm {GTP}_{TL}$ denote the stoichiometric coefficient of ATP and GTP, respectively, in CAT translation, and $\rm {ATP}_{GLC}$ denotes the number of ATP molecules produced from one molecule of glucose.
-$\rm {ATP}_{TX}$ = 176, $\rm {CTP}_{TX}$ = 144, $\rm {GTP}_{TX}$ = 151, $\rm {UTP}_{TX}$ = 189, $\rm {ATP}_{TL}$ = 219, $\rm {GTP}_{TL}$ = 438, $\rm {ATP}_{GLC}$ = 17.
-%The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52)
-
-% The solution flux vector was used to calculate the carbon yield of CAT for the three FBA cases.
-% Glucose, oxygen, and amino acids were modeled as being imported into the system, while CAT synthesis and metabolite byproduct formation were modeled as export from the system.
-%Need to talk more about biochemical benefits and importance for biochemical problems
+	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2\left({ATP}_{TX}+\rm {CTP}_{TX}+\rm {GTP}_{TX}+\rm {UTP}_{TX}\right)+\rm2\cdot\rm {ATP}_{TL}+\rm {GTP}_{TL}\right)}{\Delta\texttt{ATP}}
+\end{equation}
+where $\rm {ATP}_{TX}$, $\rm {CTP}_{TX}$, $\rm {GTP}_{TX}$, $\rm {UTP}_{TX}$ denote the stoichiometric coefficients of each energy species for CAT transcription, and $\rm {ATP}_{TL}$, $\rm {GTP}_{TL}$ denote the stoichiometric coefficients of ATP and GTP for CAT translation.
+$\rm {ATP}_{TX}$ = 176, $\rm {CTP}_{TX}$ = 144, $\rm {GTP}_{TX}$ = 151, $\rm {UTP}_{TX}$ = 189, $\rm {ATP}_{TL}$ = 219, $\rm {GTP}_{TL}$ = 438.
 
@@
 J.V, J.W, and N.H developed the cell free protein synthesis mathematical model, and parameter ensemble.
-J.V and M.V performed the sequence specific flux balance analysis calculations.
+J.V and M.V performed the sequence-specific flux balance analysis calculations.
 The manuscript was prepared and edited for publication by J.S, N.H, M.V, J.W and J.V.
@@
 Energy efficiency for best-fit set, knockouts, and experimental data.}
-    \renewcommand{\arraystretch}{1.25}
+    \renewcommand{\arraystretch}{1.3}
     \begin{tabular}{lrrrrrr} \toprule
-        Carbon Produced (mM) & Best-fit & $\Delta$cyd & $\Delta$nuo & $\Delta$app & $\Delta$cyd $\Delta$nuo $\Delta$app & Data \\ \hline
+        Carbon Produced (C-mM) & Best-fit & $\Delta$cyd & $\Delta$nuo & $\Delta$app & $\Delta$cyd $\Delta$nuo $\Delta$app & Data \\ \hline
         CAT & 20.9 & 6.5 & 18.1 & 21.4 & 5.1 & 21.6 \\ \midrule
-        Carbon Consumed (mM) \\ \hline
+        Carbon Consumed (C-mM) \\ \hline
         GLC & 215.4 & 215.4 & 215.4 & 215.4 & 159.8 & 215.4 \\ \hline
@@
         Yield & 2.7\% & 0.9\% & 2.3\% & 2.7\% & 0.7\% & - \\ \hline
-        Yield w/o ARG, GLU & 7.9\% & 2.6\% & 6.9\% & 8.1\% & 2.7\% & 8.2\% \\ \midrule
+        Yield w/o ARG, GLU & 7.9\% & 2.6\% & 6.9\% & 8.1\% & 2.7\% & 8.2\% \\ \bottomrule
 %        Energy Efficiency & 1.8\% & 0.6\% & 1.6\% & 1.8\% & 0.6\% & 1.9\% \\ \bottomrule
-        Energy Efficiency & 2.6\% & 0.8\% & 2.2\% & 2.6\% & 0.8\% & 2.6\% \\ \bottomrule
+%        Energy Efficiency & 2.6\% & 0.8\% & 2.2\% & 2.6\% & 0.8\% & 2.6\% \\ \bottomrule
 %        Efficiencies based on ATP number of 17: 2.558085 & 0.788758 & 2.210560 & 2.611583 & 0.847068 & 2.644330
-(CAT[end]-CAT[1])*(2*219+438)/((glc[1]-glc[end])*17)*100
-
     \end{tabular}
@@
 
+%\begin{table}
+%\centering
+%    \caption{Amino acid consumption toward CAT as a percentage of total amino acid consumption, for the best-fit set.
+Total consumption defined as decrease in amino acid concentration over 3 hours, such that for amino acids with synthesis active, percentage may be greater than 100\%.
+Alanine and glutamine did not decrease over the timecourse.}
+%    \renewcommand{\arraystretch}{1.25}
+%    \begin{tabular}{lrrr} \toprule
+%        Species & Total Consumption (mM) & Consumption Toward CAT (mM) & Percentage \\ \hline
+%%        ALA  & -3.852  & 0.267  & -6.9\% \\ \hline
+%        ALA  & - & 0.27 & - \\ \hline
+%%        ARG  & 1.697  & 0.178  & 10.5\% \\ \hline
+%%        ASN  & 1.547  & 0.089  & 5.8\% \\ \hline
+%%        ASP  & 1.875  & 0.214  & 11.4\% \\ \hline
+%%        CYS  & 1.016  & 0.089  & 8.8\% \\ \hline
+%        ARG  & 1.70  & 0.18  & 11\% \\ \hline
+%        ASN  & 1.55  & 0.09  & 6\% \\ \hline
+%        ASP  & 1.87  & 0.21  & 11\% \\ \hline
+%        CYS  & 1.02  & 0.09  & 9\% \\ \hline
+%%        GLN  & -2.278  & 0.214  & -9.4\% \\ \hline
+%        GLN  & - & 0.21 & - \\ \hline
+%%        GLU  & 98.526  & 0.232  & 0.2\% \\ \hline
+%%        GLY  & 1.545  & 0.178  & 11.5\% \\ \hline
+%%        HIS  & 0.036  & 0.214  & 587.9\% \\ \hline
+%%        ILE  & 0.16  & 0.16  & 100\% \\ \hline
+%%        LEU  & 0.231  & 0.232  & 100.2\% \\ \hline
+%%        LYS  & 1.785  & 0.214  & 12.0\% \\ \hline
+%%        MET  & 0.16  & 0.16  & 100\% \\ \hline
+%%        PHE  & 0.356  & 0.356  & 100\% \\ \hline
+%%        PRO  & 0.481  & 0.125  & 26\% \\ \hline
+%%        SER  & 0.83  & 0.178  & 21.5\% \\ \hline
+%%        THR  & 0.857  & 0.232  & 27\% \\ \hline
+%%        TRP  & 0.089  & 0.089  & 100\% \\ \hline
+%%        TYR  & 0.12  & 0.196  & 164\% \\ \hline
+%%        VAL  & 0.285  & 0.285  & 100\% \\ \hline
+%        GLU  & 98.53  & 0.23  & 0.2\% \\ \hline
+%        GLY  & 1.55  & 0.18  & 12\% \\ \hline
+%        HIS  & 0.04  & 0.21  & 588\% \\ \hline
+%        ILE  & 0.16  & 0.16  & 100\% \\ \hline
+%        LEU  & 0.23  & 0.23  & 100\% \\ \hline
+%        LYS  & 1.79  & 0.21  & 12\% \\ \hline
+%        MET  & 0.16  & 0.16  & 100\% \\ \hline
+%        PHE  & 0.36  & 0.36  & 100\% \\ \hline
+%        PRO  & 0.48  & 0.12  & 26\% \\ \hline
+%        SER  & 0.83  & 0.18  & 21\% \\ \hline
+%        THR  & 0.86  & 0.23  & 27\% \\ \hline
+%        TRP  & 0.09  & 0.09  & 100\% \\ \hline
+%        TYR  & 0.12  & 0.20  & 164\% \\ \hline
+%        VAL  & 0.29  & 0.29  & 100\% \\ \bottomrule
+%    \end{tabular}
+%\label{tbl:AA_breakdown}
+%\end{table}
+
 \begin{table}
 \centering
-    \caption{Amino acid consumption toward CAT as a percentage of total amino acid consumption, for the best-fit set.
-Total consumption defined as decrease in amino acid concentration over 3 hours, such that for amino acids with synthesis active, percentage may be greater than 100\%.
-Alanine and glutamine did not decrease over the timecourse.}
-    \renewcommand{\arraystretch}{1.25}
+    \caption{Amino acid total net consumption, consumption toward CAT, and net consumption other than toward CAT for the best-fit set.
+Total net consumption defined as decrease in amino acid concentration over 3 hours.
+Consumption toward CAT defined as increase in CAT concentration over 3 hours, multiplied by stoichiometric coefficient for given amino acid in CAT synthesis.}
+    \renewcommand{\arraystretch}{1.3}
     \begin{tabular}{lrrr} \toprule
-        Species & Total Consumption & Consumption Toward CAT & Percentage \\ \hline
-%        ALA  & -3.852  & 0.267  & -6.9\% \\ \hline
-        ALA  & - & 0.27 & - \\ \hline
-%        ARG  & 1.697  & 0.178  & 10.5\% \\ \hline
-%        ASN  & 1.547  & 0.089  & 5.8\% \\ \hline
-%        ASP  & 1.875  & 0.214  & 11.4\% \\ \hline
-%        CYS  & 1.016  & 0.089  & 8.8\% \\ \hline
-        ARG  & 1.70  & 0.18  & 11\% \\ \hline
-        ASN  & 1.55  & 0.09  & 6\% \\ \hline
-        ASP  & 1.87  & 0.21  & 11\% \\ \hline
-        CYS  & 1.02  & 0.09  & 9\% \\ \hline
-%        GLN  & -2.278  & 0.214  & -9.4\% \\ \hline
-        GLN  & - & 0.21 & - \\ \hline
-%        GLU  & 98.526  & 0.232  & 0.2\% \\ \hline
-%        GLY  & 1.545  & 0.178  & 11.5\% \\ \hline
-%        HIS  & 0.036  & 0.214  & 587.9\% \\ \hline
-%        ILE  & 0.16  & 0.16  & 100\% \\ \hline
-%        LEU  & 0.231  & 0.232  & 100.2\% \\ \hline
-%        LYS  & 1.785  & 0.214  & 12.0\% \\ \hline
-%        MET  & 0.16  & 0.16  & 100\% \\ \hline
-%        PHE  & 0.356  & 0.356  & 100\% \\ \hline
-%        PRO  & 0.481  & 0.125  & 26\% \\ \hline
-%        SER  & 0.83  & 0.178  & 21.5\% \\ \hline
-%        THR  & 0.857  & 0.232  & 27\% \\ \hline
-%        TRP  & 0.089  & 0.089  & 100\% \\ \hline
-%        TYR  & 0.12  & 0.196  & 164\% \\ \hline
-%        VAL  & 0.285  & 0.285  & 100\% \\ \hline
-        GLU  & 98.53  & 0.23  & 0.2\% \\ \hline
-        GLY  & 1.55  & 0.18  & 12\% \\ \hline
-        HIS  & 0.04  & 0.21  & 588\% \\ \hline
-        ILE  & 0.16  & 0.16  & 100\% \\ \hline
-        LEU  & 0.23  & 0.23  & 100\% \\ \hline
-        LYS  & 1.79  & 0.21  & 12\% \\ \hline
-        MET  & 0.16  & 0.16  & 100\% \\ \hline
-        PHE  & 0.36  & 0.36  & 100\% \\ \hline
-        PRO  & 0.48  & 0.12  & 26\% \\ \hline
-        SER  & 0.83  & 0.18  & 21\% \\ \hline
-        THR  & 0.86  & 0.23  & 27\% \\ \hline
-        TRP  & 0.09  & 0.09  & 100\% \\ \hline
-        TYR  & 0.12  & 0.20  & 164\% \\ \hline
-        VAL  & 0.29  & 0.29  & 100\% \\ \bottomrule
+        & \parbox{3.6cm}{Total Net \phantom{abcdefgh} Consumption (mM)} & \parbox{3.4cm}{Consumption \phantom{abc} Toward CAT (mM)} & \parbox{3.6cm}{Other \phantom{abcdefgh} Consumption (mM)} \\ \hline
+        ALA  & -3.85 & 0.27 & -4.12 \\ \hline
+        ARG  & 1.70  & 0.18  & 1.52 \\ \hline
+        ASN  & 1.55  & 0.09  & 1.46 \\ \hline
+        ASP  & 1.87  & 0.21  & 1.66 \\ \hline
+        CYS  & 1.02  & 0.09  & 0.93 \\ \hline
+        GLN  & -2.28 & 0.21 & -2.49 \\ \hline
+        GLU  & 98.53  & 0.23  & 98.30 \\ \hline
+        GLY  & 1.55  & 0.18  & 1.37 \\ \hline
+        HIS  & 0.04  & 0.21  & -0.17 \\ \hline
+        ILE  & 0.16  & 0.16  & 0.00 \\ \hline
+        LEU  & 0.23  & 0.23  & 0.00 \\ \hline
+        LYS  & 1.79  & 0.21  & 1.58 \\ \hline
+        MET  & 0.16  & 0.16  & 0.00 \\ \hline
+        PHE  & 0.36  & 0.36  & 0.00 \\ \hline
+        PRO  & 0.48  & 0.12  & 0.36 \\ \hline
+        SER  & 0.83  & 0.18  & 0.65 \\ \hline
+        THR  & 0.86  & 0.23  & 0.63 \\ \hline
+        TRP  & 0.09  & 0.09  & 0.00 \\ \hline
+        TYR  & 0.12  & 0.20  & -0.08 \\ \hline
+        VAL  & 0.29  & 0.29  & 0.00 \\ \bottomrule
     \end{tabular}

### 2016-12-08 Paper_v1.tex
Dec8_2

@@
 The ensemble described all of the training data, especially the central carbon metabolism.
-The model predicted a carbon yield for CAT production that was equal to 35\% of yield for a physiologically realistic case, calculated using sequence-specific flux balance analysis.
+The model predicted a carbon yield for CAT production that was equal to 36\% of yield for a physiologically realistic case, and an energy efficiency equal to 9\% of the physiologically realistic case, calculated using sequence-specific flux balance analysis.
 This suggests that CAT production could be further optimized.
@@
 Next, to gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield with the maximum theoretical CAT carbon yield calculated using sequence-specific flux balance analysis.
-The CAT yield estimated from the kinetic model was 35\% of the theoretical yield when physiologically realistic constraints were used.
+The CAT yield estimated from the kinetic model was 36\% of the theoretical yield when physiologically realistic constraints were used.
 Taken together, we have integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
@@
 This was due to its role in generating other amino acids and as a possible energy source for metabolism.
-The much greater consumption of this one amino acid was made possible by its much larger initial condition, as it was present in the cell media in the form of magnesium glutamate, ammonium glutamate, and potassium glutamate [CITE?
-THESIS NOT CURRENTLY IN BIBLIOGRAPHY].
+The much greater consumption of this one amino acid was made possible by its much larger initial condition, as it was present in the cell media in the form of magnesium glutamate, ammonium glutamate, and potassium glutamate.
+% [CITE? THESIS NOT CURRENTLY IN BIBLIOGRAPHY].
 Glucose made the next largest contribution, between 20\% and 30\%, dwarfing all amino acids other than glutamate.
@@
 Having captured the experimental data, we investigated if CAT yield and CFPS performance could be further improved.
-We showed that the model produces CAT with a carbon yield equal to 35\% of that of a physiological case in which transcription and translation are constrained.
+We showed that the model produces CAT with a carbon yield equal to 36\%, and an energy efficiency equal to 9\%, of that of a physiological case in which transcription and translation are constrained.
 The accumulation of waste byproducts, especially acetate, is responsible for this sub-optimal yield.
@@
 Comparing the physiologically realistic carbon yield of CAT from ssFBA predictions to those of the kinetic model and experimental measurements suggests that there is potential for increasing CAT yield as well as CFPS performance.
-%The model and experimental yields were 35\% of the physiologically constrained case.
+%The model and experimental yields were 36\% of the physiologically constrained case.
 A knockout of \textit{gnd} and shows that carbon can be diverted away from acetate and toward CAT or other proteins of interest expressed in CFPS.

### 2017-01-26 Code/Ensemble/src/Algorithm.jl
Jan26

@@ -1,142 +1,0 @@
-include("DataDictionary.jl")
-include("SolveBalances.jl")
-include("CalcError.jl")
-using PyPlot
-
-# Create Ensemble directory if it doesn't already exist
-if ~isdir("Ensemble")
-    mkdir("Ensemble")
-end
-
-# Create Best directory if it doesn't already exist
-if ~isdir("Ensemble/Best")
-    mkdir("Ensemble/Best")
-end
-
-# Setup the timescale
-TSTART = 0.0
-TSTOP = 3.0
-Ts = 0.1
-
-# Create dictionary of parameters
-data_dictionary_best = DataDictionary(TSTART,TSTOP,Ts)
-experiment_data_dictionary = data_dictionary_best["EXPERIMENT_DATA_DICTIONARY"]
-
-#data_indices = Int64[]
-#for (key,value) in experiment_data_dictionary
-#    push!(data_indices,value.model_index)
-#end
-#sort!(data_indices)
-
-#init_indices = collect(1:103)
-#deleteat!(init_indices,data_indices)
-
-# Initialize number of parameters to be varied
-num_rate = 163
-#num_init = 66
-num_init = 0
-num_sat = 455
-num_cont = 34
-
-# Create or read the relevant information
-if ~isfile("Ensemble/Best/params")
-    # Take rate constants to be varied from data_dictionary_best
-    rate = data_dictionary_best["RATE_CONSTANT_ARRAY"][1:163]
-    # Take initial conditions to be varied from data_dictionary_best
-#    init = data_dictionary_best["INITIAL_CONDITION_ARRAY"][init_indices]
-    # Take saturation constants to be varied from data_dictionary_best
-    sat_vector = vec(data_dictionary_best["SATURATION_CONSTANT_ARRAY"])
-    sat = sat_vector[find(sat_vector.>0)]
-    # Take control parameters to be varied from data_dictionary_best
-    cont = vec(data_dictionary_best["CONTROL_PARAMETER_ARRAY"]')
-    # Concatenate into params vector
-#    params = [rate;init;sat;cont]
-    params_best = [rate;sat;cont]
-    # Solve the balance equations and calculate cost
-    Tsim_best,X_best = SolveBalances(TSTART,TSTOP,Ts,data_dictionary_best)
-    cost_best = CalcError(experiment_data_dictionary,Tsim_best,X_best)
-    # Save the relevant information
[…]

### 2017-01-26 Code/Ensemble/src/CalcArea.jl

@@ -1,17 +1,0 @@
-function CalcArea(Tsim,X)
-
-if !(length(Tsim)==length(X))
-    throw("Vectors must be of same length")
-end
-
-AUC = Float64[]
-
-i_range = 2:length(Tsim)
-for i in i_range
-    push!(AUC,(Tsim[i]-Tsim[i-1])*(X[i]+X[i-1])/2) # Trapezoid rule
-end
-
-AUC = sum(AUC)
-
-return AUC
-end

### 2017-01-26 Code/Ensemble/src/CalcError.jl

@@ -1,28 +1,0 @@
-using Interpolations
-
-function CalcError(experiment_data_dictionary,Tsim,X)
-# Initialize array
-Error = Float64[]
-for (key,dataset) in experiment_data_dictionary
-    # Get raw data and timepoints
-    raw_array = dataset.raw_measurement_array
-    data_tps = raw_array[:,1]
-    data_vals = raw_array[:,2]
-    # Interpolate simulation array to fit dataset
-    index = dataset.model_index
-    sim = X[:,index]
-    itp = interpolate((Tsim,),sim,Gridded(Linear()))
-    sim_itp = itp[data_tps]
-    # Calculate error
-    error = sum(abs(sim_itp-data_vals))/maximum(data_vals)
-    # Weight certain datasets as more important
-    weight = (key == "M_glc_D_c" || key == "CAT" || key == "M_pyr_c" || key == "M_lac_D_c" || key == "M_ac_c" || key == "M_succ_c" || key == "M_mal_L_c")?5:1
-    error *= weight
-    # Push to array
-    push!(Error,error)
-end
-
-cost = sum(Error)
-
-return cost
-end

### 2017-01-26 Code/Ensemble/src/Figure1.jl

@@ -1,754 +1,0 @@
-# MassBalances.jl is set up to save the flux vector at all timepoints satisfying t < 1e-11 or 1.48 < t < 1.52 or t > 2.98
-# At each timepoint, a vector of length 327 is saved to disk: the first element is the timepoint, followed by a flux value for each of the 326 reactions
-# This is done for each set in the ensemble (there are 18000 such sets)
-
-# For each set in the ensemble, we read all flux vectors and merge them into an array
-# We then find the timepoint closest to 1.5 hr and save the corresponding flux vector, under the directory for that set, under the name "flux_1_5"
-# We then find the timepoint closest to 3 hr and save the corresponding flux vector, under the directory for that set, under the name "flux_3"
-num_dir = convert(Int64,readdlm("Ensemble/dir")[1])
-for i in 1:num_dir
-    # Read number of timepoints
-    num_tp = convert(Int64,readdlm("./Ensemble/$i/Flux/num_tp")[1])
-    # Read first flux
-    rate_vector = readdlm("./Ensemble/$i/Flux/1")
-    # Read all other fluxes; append to first flux
-    j_range = 2:num_tp
-    for j in j_range
-        r = readdlm("./Ensemble/$i/Flux/$j")
-        rate_vector = [rate_vector r]
-    end
-    # Save flux at t = 0
-    tp = rate_vector[1,:]
-    index = find(tp.==minimum(tp))[1]
-    writedlm("./Ensemble/$i/flux_0",rate_vector[2:end,index])
-    # Save flux at t = 1.5
-    tp_1_5 = abs(tp-1.5)
-    index = find(tp_1_5.==minimum(tp_1_5))[1]
-    writedlm("./Ensemble/$i/flux_1_5",rate_vector[2:end,index])
-    # Save flux at t = 3
-    tp_3 = abs(tp-3.0)
-    index = find(tp_3.==minimum(tp_3))[1]
-    writedlm("./Ensemble/$i/flux_3",rate_vector[2:end,index])
[…]

### 2017-01-26 Code/Ensemble/src/Figure5.jl

@@ -1,117 +1,0 @@
-using PyPlot
-using PyCall
-@pyimport numpy as np
-
-include("Local_sens_fxn.jl")
-
-num_iter = 180 # Should be set equal to number of samples in sub-ensemble
-for i in 1:num_iter
-    Local_sens_fxn(i)
-end
-
-Rate = readdlm("Sens/1/Rate")
-Sat = readdlm("Sens/1/Sat")
-Cont = readdlm("Sens/1/Cont")
-
-for i in 1:num_iter
-    rate = readdlm("Sens/$i/Rate")
-    sat = readdlm("Sens/$i/Sat")
-    cont = readdlm("Sens/$i/Cont")
-    Rate = [Rate rate]
-    Sat = [Sat sat]
-    Cont = [Cont cont]
-end
-
-writedlm("Sens/Rate",Rate)
-writedlm("Sens/Sat",Sat)
-writedlm("Sens/Cont",Cont)
-
-Rate_m = Float64[]
-Rate_s = Float64[]
-for i in 1:size(Rate)[1]
-    rate = Rate[i,:]
-    push!(Rate_m,mean(rate))
-    push!(Rate_s,std(rate))
-end
-
-Sat_m = Float64[]
-Sat_s = Float64[]
-for i in 1:size(Sat)[1]
-    sat = Sat[i,:]
-    push!(Sat_m,mean(sat))
-    push!(Sat_s,std(sat))
-end
-
-Cont_m = Float64[]
-Cont_s = Float64[]
-for i in 1:size(Cont)[1]
-    cont = Cont[i,:]
-    push!(Cont_m,mean(cont))
-    push!(Cont_s,std(cont))
-end
-
-writedlm("Sens/Rate_m",Rate_m)
-writedlm("Sens/Rate_s",Rate_s)
-writedlm("Sens/Sat_m",Sat_m)
-writedlm("Sens/Sat_s",Sat_s)
-writedlm("Sens/Cont_m",Cont_m)
-writedlm("Sens/Cont_s",Cont_s)
-
-PyCall.PyDict(matplotlib["rcParams"])["font.sans-serif"] = ["Helvetica"]
-
-Rate_m = readdlm("Sens/Rate_m")
-Rate_s = readdlm("Sens/Rate_s")/sqrt(180)
-width = 1
-a = [1:2:2*length(Rate_m);]
-threshold = 0.05*ones(a)
-PyPlot.matplotlib[:rc]("text",usetex=true)
[…]

### 2017-01-26 Code/Ensemble/src/Figures234.jl

@@ -1,582 +1,0 @@
-include("DataDictionary.jl")
-include("SolveBalances.jl")
-using PyPlot
-
-# Setup the timescale
-TSTART = 0.0
-TSTOP = 3.0
-Ts = 0.1
-
-# Create dictionary of parameters (best-fit set)
-data_dictionary = DataDictionary(TSTART,TSTOP,Ts)
-
-#-----------------------------------------------------------------------------
-# In order to simulate predictions in the absence of allosteric control,
-# comment the line "rate_vector = rate_vector.*control_vector;" in Control.jl
-#-----------------------------------------------------------------------------
-
-# Solve the balances to calculate predictions based on best-fit set
-Tsim,X = SolveBalances(TSTART,TSTOP,Ts,data_dictionary)
-
-# 
-experimental_data_dictionary = data_dictionary["EXPERIMENT_DATA_DICTIONARY"]
-
-# Store the glucose, CAT, and organic acid raw data by species
-glc = experimental_data_dictionary["M_glc_D_c"].raw_measurement_array
-pyr = experimental_data_dictionary["M_pyr_c"].raw_measurement_array
-lac = experimental_data_dictionary["M_lac_D_c"].raw_measurement_array
-ac = experimental_data_dictionary["M_ac_c"].raw_measurement_array
-succ = experimental_data_dictionary["M_succ_c"].raw_measurement_array
-mal = experimental_data_dictionary["M_mal_L_c"].raw_measurement_array
-CAT = experimental_data_dictionary["CAT"].raw_measurement_array
-
-# Store the energy species raw data by species
-atp = experimental_data_dictionary["M_atp_c"].raw_measurement_array
-adp = experimental_data_dictionary["M_adp_c"].raw_measurement_array
[…]

### 2017-01-26 Code/Ensemble/src/MassBalances.jl

@@ -1,93 +1,0 @@
-include("Kinetics.jl");
-include("Control.jl");
-
-# ----------------------------------------------------------------------------------- #
-# Copyright (c) 2016 Varnerlab
-# School of Chemical Engineering Purdue University
-# W. Lafayette IN 46907 USA
-
-# Permission is hereby granted, free of charge, to any person obtaining a copy
-# of this software and associated documentation files (the "Software"), to deal
-# in the Software without restriction, including without limitation the rights 
-# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell 
-# copies of the Software, and to permit persons to whom the Software is 
-# furnished to do so, subject to the following conditions:
-#
-# The above copyright notice and this permission notice shall be included in
-# all copies or substantial portions of the Software.
-#
-# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
-# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
-# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
-# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
-# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
-# OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
-# THE SOFTWARE.
-# ----------------------------------------------------------------------------------- #
-function MassBalances(t,x,dxdt_vector,data_dictionary)
-# ---------------------------------------------------------------------- #
-# MassBalances.jl was generated using the Kwatee code generation system.
-# Username: nicholas
-# Type: CFPS-JULIA
-# Version: 1.0
-# Generation timestamp: 05-11-2016 16:22:26
-# 
-# Arguments: 
-# t  - current time 
-# x  - state vector 
-# dxdt_vector - right hand side vector 
-# data_dictionary  - Data dictionary instance (holds model parameters) 
-# ---------------------------------------------------------------------- #
-
[…]

### 2017-01-26 Code/Ensemble/src/PlotSingle.jl

@@ -1,498 +1,0 @@
-using PyPlot
-function PlotSingle(Tsim,X,data_dictionary)
-
-experimental_data_dictionary = data_dictionary["EXPERIMENT_DATA_DICTIONARY"]
-
-glc = experimental_data_dictionary["M_glc_D_c"].raw_measurement_array
-pyr = experimental_data_dictionary["M_pyr_c"].raw_measurement_array
-lac = experimental_data_dictionary["M_lac_D_c"].raw_measurement_array
-ac = experimental_data_dictionary["M_ac_c"].raw_measurement_array
-succ = experimental_data_dictionary["M_succ_c"].raw_measurement_array
-mal = experimental_data_dictionary["M_mal_L_c"].raw_measurement_array
-CAT = experimental_data_dictionary["CAT"].raw_measurement_array
-
-atp = experimental_data_dictionary["M_atp_c"].raw_measurement_array
-adp = experimental_data_dictionary["M_adp_c"].raw_measurement_array
-amp = experimental_data_dictionary["M_amp_c"].raw_measurement_array
-axp = atp+adp+amp
-axp[:,1] = atp[:,1]
-gtp = experimental_data_dictionary["M_gtp_c"].raw_measurement_array
-gdp = experimental_data_dictionary["M_gdp_c"].raw_measurement_array
-gmp = experimental_data_dictionary["M_gmp_c"].raw_measurement_array
-gxp = gtp+gdp+gmp
-gxp[:,1] = gtp[:,1]
-ctp = experimental_data_dictionary["M_ctp_c"].raw_measurement_array
-cdp = experimental_data_dictionary["M_cdp_c"].raw_measurement_array
-cmp = experimental_data_dictionary["M_cmp_c"].raw_measurement_array
-cxp = ctp+cdp+cmp
-cxp[:,1] = ctp[:,1]
-utp = experimental_data_dictionary["M_utp_c"].raw_measurement_array
[…]

### 2017-01-26 Code/Ensemble/src/SolveBalances.jl

@@ -1,56 +1,0 @@
-include("MassBalances.jl")
-using Sundials;
-
-# ----------------------------------------------------------------------------------- #
-# Copyright (c) 2016 Varnerlab
-# School of Chemical Engineering Purdue University
-# W. Lafayette IN 46907 USA
-
-# Permission is hereby granted, free of charge, to any person obtaining a copy
-# of this software and associated documentation files (the "Software"), to deal
-# in the Software without restriction, including without limitation the rights 
-# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell 
-# copies of the Software, and to permit persons to whom the Software is 
-# furnished to do so, subject to the following conditions:
-#
-# The above copyright notice and this permission notice shall be included in
-# all copies or substantial portions of the Software.
-#
-# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
-# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
-# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
-# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
-# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
-# OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
-# THE SOFTWARE.
-# ----------------------------------------------------------------------------------- #
-function SolveBalances(TSTART,TSTOP,Ts,data_dictionary)
-# ----------------------------------------------------------------------------------- #
-# SolveBalances.jl was generated using the Kwatee code generation system.
-# SolveBalances: Solves model equations from TSTART to TSTOP given parameters in data_dictionary.
-# Username: nicholas
-# Type: CFPS-JULIA
-# Version: 1.0
-# Generation timestamp: 05-11-2016 16:22:26
-# 
-# Input arguments: 
-# TSTART  - Time start 
-# TSTOP  - Time stop 
-# Ts - Time step 
-# data_dictionary  - Data dictionary instance (holds model parameters) 
-# 
-# Return arguments: 
-# TSIM - Simulation time vector 
[…]

### 2017-01-26 Code/FluxSobol/sensitivity/analyze_final.sh

@@ -1,34 +1,0 @@
-#!/bin/bash
-
-# Perform sensitivity analysis from the command line
-# To run: ./shell-example-analyze.sh
-# Note that "python -m" runs the __main__ module in the SALib.analyze package
-# The -B flag prevents the interpreter from compiling bytecode (.pyc files)
-
-python -m SALib.analyze.sobol \
-     -p ./Range.txt \
-     -Y performance.txt \
-     -c 0 \
-     --max-order=2 \
-     > SAout.txt \
-
-# Options:
-# -p, --paramfile: Your parameter range file (3 columns: parameter name, lower bound, upper bound)
-#
-# -Y, --model-output-file: File of model output values to analyze
-#
-# -c, --column (optional): Column of model output file to analyze.
-#                If the file only has one column, this argument will be ignored.
-#
-# --delimiter (optional): Model output file delimiter.
-#
-# --max-order (optional): Maximum order of indices to calculate.
-#               This must match the value chosen during sampling.
-#
-# -r, --resamples (optional): Number of bootstrap resamples used to calculate confidence intervals on indices. Default 1000.
-#
-# --parallel (optional): Flag to enable parallel execution with multiprocessing
-#
-# --processors (optional, int): Number of processors to be used with the parallel option
-#
-# Last line determines the file to output the result by Sobol analysis

### 2017-01-26 Code/FluxSobol/sensitivity/sample.sh

@@ -1,30 +1,0 @@
-#!/bin/bash
-
-# Generate sensitivity analysis samples from the command line
-# To run: ./shell-example-sample.sh
-# Note that "python -m" runs the __main__ module in the SALib.sample package
-# The -B flag prevents the interpreter from compiling bytecode (.pyc files)
-
-python -m SALib.sample.saltelli \
-     -n 5000 \
-	   -p ./Range.txt \
-	   -o ./Samples.txt \
-	   --delimiter=' ' \
-	   --precision=8 \
-
-## Options:
-# -p, --paramfile: Your parameter range file (3 columns: parameter name, lower bound, upper bound)
-#
-# -n, --samples: Sample size.
-#				 Number of model runs is N(2D + 2) if calculating second-order indices (default)
-#        or N(D + 2) otherwise.
-#
-# -o, --output: File to output your samples into.
-#
-# --delimiter (optional): Output file delimiter.
-#
-# --precision (optional): Digits of precision in the output file. Default is 8.
-#
-# --max-order (optional): Maximum order of indices to calculate. Choose 1 or 2, default is 2.
-#								   Choosing 1 will reduce total model runs from N(2D + 2) to N(D + 2)
-#								   Must use the same value (either 1 or 2) for both sampling and analys

### 2017-01-26 Code/FluxSobol/src/Bounds.jl

@@ -1,514 +1,0 @@
-include("Types.jl")
-using GLPK
-# ----------------------------------------------------------------------------------- #
-# Copyright (c) 2016 Varnerlab
-# School of Chemical Engineering Purdue University
-# W. Lafayette IN 46907 USA
-
-# Permission is hereby granted, free of charge, to any person obtaining a copy
-# of this software and associated documentation files (the "Software"), to deal
-# in the Software without restriction, including without limitation the rights
-# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
-# copies of the Software, and to permit persons to whom the Software is
-# furnished to do so, subject to the following conditions:
-#
-# The above copyright notice and this permission notice shall be included in
-# all copies or substantial portions of the Software.
-#
-# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
-# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
-# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
-# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
[…]

### 2017-01-26 Code/FluxSobol/src/Control.jl

@@ -1,174 +1,0 @@
-# ----------------------------------------------------------------------------------- #
-# Copyright (c) 2016 Varnerlab
-# School of Chemical Engineering Purdue University
-# W. Lafayette IN 46907 USA
-
-# Permission is hereby granted, free of charge, to any person obtaining a copy
-# of this software and associated documentation files (the "Software"), to deal
-# in the Software without restriction, including without limitation the rights 
-# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell 
-# copies of the Software, and to permit persons to whom the Software is 
-# furnished to do so, subject to the following conditions:
-#
-# The above copyright notice and this permission notice shall be included in
-# all copies or substantial portions of the Software.
-#
-# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
-# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
-# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
-# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
-# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
-# OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
-# THE SOFTWARE.
-# ----------------------------------------------------------------------------------- #
-function Control(t,x,data_dictionary)
-# ---------------------------------------------------------------------- #
-# Control.jl was generated using the Kwatee code generation system.
-# Username: jeffreyvarner
[…]

### 2017-01-26 Code/FluxSobol/src/FluxDriver.jl

@@ -1,187 +1,0 @@
-include("DataFile.jl")
-include("Control.jl")
-using GLPK
-
-# ----------------------------------------------------------------------------------- #
-# Copyright (c) 2016 Varnerlab
-# School of Chemical Engineering Purdue University
-# W. Lafayette IN 46907 USA
-
-# Permission is hereby granted, free of charge, to any person obtaining a copy
-# of this software and associated documentation files (the "Software"), to deal
-# in the Software without restriction, including without limitation the rights 
-# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell 
-# copies of the Software, and to permit persons to whom the Software is 
-# furnished to do so, subject to the following conditions:
-#
-# The above copyright notice and this permission notice shall be included in
-# all copies or substantial portions of the Software.
-#
-# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
-# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
-# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
-# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
-# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
-# OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
-# THE SOFTWARE.
-# ----------------------------------------------------------------------------------- #
-function FluxDriver(time_index, species_abundance_array, specific_growth_rate, cellmass, dsa, step_size, data_dictionary; steady_state_flag=true)
-# ----------------------------------------------------------------------------------- #
-# FluxDriver.jl was generated using the Kwatee code generation system.
-# FluxDriver: Solves the flux balance analysis problem from TSTART to TSTOP given the model encoded in data_dictionary.
-# Username: jeffreyvarner
-# Type: NFBA-JULIA
-# Version: 1.0
-# Generation timestamp: 03-03-2016 21:13:51
-# 
-# Input arguments: 
-# data_dictionary  - Data dictionary instance (holds species and flux models etc) 
-# 
-# Return arguments: 
[…]

### 2017-01-26 Code/FluxSobol/src/SampleBounds.jl

@@ -1,99 +1,0 @@
-# Script to sample the bounds file -
-
-# Script to max product -
-include("DataFile.jl")
-include("Solve.jl")
-include("Types.jl")
-
-# Load the data file with the default bounds -
-data_dictionary = DataFile(0,0,0);
-
-# Set the max -
-data_dictionary["MIN_FLAG"] = false;
-
-# Get the flux model dictionary -
-flux_model_dictionary = data_dictionary["FLUX_MODEL_DICTIONARY"]
-flux_model_dictionary["PROTEIN_export"].flux_obj_coeff = 1.0;
-
-# Reactions that we block ...
-blocked_reaction_set = Set{AbstractString}()
-push!(blocked_reaction_set,"R_PTS")
-push!(blocked_reaction_set,"R_nuo")
-push!(blocked_reaction_set,"R_ndh1")
-push!(blocked_reaction_set,"R_ndh2")
-push!(blocked_reaction_set,"M_lac_D_c_exchange_reverse")
-push!(blocked_reaction_set,"M_etoh_c_exchange_reverse")
-push!(blocked_reaction_set,"M_akg_c_exchange_reverse")
-push!(blocked_reaction_set,"M_fum_c_exchange_reverse")
-push!(blocked_reaction_set,"M_mal_L_c_exchange_reverse")
-push!(blocked_reaction_set,"M_pyr_c_exchange_reverse")
-push!(blocked_reaction_set,"M_ac_c_exchange_reverse")
-push!(blocked_reaction_set,"M_succ_D_c_exchange_reverse")
-data_dictionary["blocked_reaction_set"] = blocked_reaction_set
-
-
-# set the bound -
-# data_dictionary["M_o2_c_exchange_reverse"] = 100
-
-# Set the plasmid concentration -
-#gene_copy_number = (10e-6/1e9)*(6.02e23)*10.0;
-#data_dictionary["deGFP_gene_copies"] = gene_copy_number;
-
-# Load the bounds data -
-bounds_array = readdlm("../sensitivity/Samples.txt")
-(nr,nc) = size(bounds_array)
-
-# Setup performance array -
-performance_array = zeros(nr)
-
-# Setup bounds dictionary -
-bounds_dictionary = Dict()
-
-# Ok, we need to enumerate through this list, and calculate the performance of the model for each value -
-for index in collect(1:nr)
-
-  # Set the bounds array -
-  # The first 20 cols are AAs
[…]

### 2017-01-26 Code/FluxSobol/src/Solve.jl

@@ -1,198 +1,0 @@
-# Simple script to testthe Kwatee LFBA code -
-
-# includes -
-include("DataFile.jl");
-include("FluxDriver.jl");
-include("Utility.jl")
-
-function Solve(data_dictionary)
-
-  # is the data_dictionary empty?
-  if (isempty(data_dictionary) == true)
-
-    # Load the data_dictionary -
-    data_dictionary = DataFile(0,0,0);
-  end
-
-  # Split the stm into internal and external blocks -
-  (intracellular_index_array, extracellular_index_array, intracellular_block,extracellular_block) = splitStoichiometricMatrixIntoInternalExternalBlocks(data_dictionary);
-
-  # Build state mapping index -
-  state_mapping_index_array = calculateStateMappingIndexArray([intracellular_index_array ; extracellular_index_array]);
-
-  # Get model dictionaries -
-  species_model_dictionary = data_dictionary["SPECIES_MODEL_DICTIONARY"];
-  dsa = calculateDilutionSelectionArray(species_model_dictionary);
-  tau_array = calculateTimeDelayParameterArray(species_model_dictionary);
-
-  # Get system dimension -
-  number_of_species = data_dictionary["NUMBER_OF_SPECIES"];
-  number_of_fluxes = data_dictionary["NUMBER_OF_FLUXES"];
-
-  # Setup the growth rate -
-  specific_growth_rate = 0.0;
-  TSTEP = 0.1;
-  state_array = calculateInitialConditionArray(species_model_dictionary);
-  intracellular_state_array = state_array[intracellular_index_array];
-  extracellular_state_array = state_array[extracellular_index_array];
-  cellmass = 0.01;
-
-  # Growth coefficients -
-  gca = calculateGrowthCoefficientArray(species_model_dictionary);
-
-  # Call the flux driver with a SS = true
-  (objective_value, flux_array, dual_array, uptake_array, exit_flag) = FluxDriver(1, state_array, specific_growth_rate, cellmass, dsa, TSTEP, data_dictionary, steady_state_flag=true);
-
-
-  # return -
-  return (objective_value, flux_array, dual_array, uptake_array, exit_flag);
-end
-
-
-function Solve(TSTART,TSTOP,TSTEP,data_dictionary)
-
-# is the data_dictionary empty?
-if (isempty(data_dictionary) == true)
-
-  # Load the data_dictionary -
-  data_dictionary = DataFile(TSTART,TSTOP,TSTEP);
-end
-
[…]

### 2017-01-26 Code/FluxSobol/src/Types.jl

@@ -1,67 +1,0 @@
-# ----------------------------------------------------------------------------------- #
-# Copyright (c) 2016 Varnerlab
-# School of Chemical Engineering Purdue University
-# W. Lafayette IN 46907 USA
-
-# Permission is hereby granted, free of charge, to any person obtaining a copy
-# of this software and associated documentation files (the "Software"), to deal
-# in the Software without restriction, including without limitation the rights 
-# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell 
-# copies of the Software, and to permit persons to whom the Software is 
-# furnished to do so, subject to the following conditions:
-#
-# The above copyright notice and this permission notice shall be included in
-# all copies or substantial portions of the Software.
-#
-# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
-# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
-# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
-# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
-# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
-# OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
-# THE SOFTWARE.
-# ----------------------------------------------------------------------------------- #
-
-# Define the custom species model type - 
-type SpeciesModel
-
-	# Model instance variables - 
-	species_index::Int
-	species_symbol::AbstractString
-	species_lower_bound::Float64
-	species_upper_bound::Float64
-	is_species_measured::Bool
-	species_measurement_array::Array{Float64,2}
-	species_constraint_type::Int32
-	species_initial_condition::Float64
-	is_biomass_precursor::Bool
-	biomass_precursor_coefficient::Float64
-	species_time_constant::Float64
-	is_species_diluted::Bool
-	is_species_extracellular::Bool
-
-	# Constructor - 
-	function SpeciesModel()
-		this = new();
-	end
-end
-
-# Define the custom flux model type - 
-type FluxModel
-
-	# Model instance variables - 
[…]

### 2017-01-26 Code/FluxSobol/src/Utility.jl

@@ -1,144 +1,0 @@
-include("Types.jl")
-
-# ================================================================================== #
-# Utility.jl
-# Library for setup functions for NFBA calculations for code generated by Kwatee
-#
-# Author: Varnerlab
-# Chemical Engineering, Purdue University
-# ================================================================================== #
-
-function calculateDilutionSelectionArray(species_model_dictionary::Dict{AbstractString,SpeciesModel})
-
-  # Method variables -
-  number_of_species = length(species_model_dictionary);
-  dilution_selection_array = ones(Float64,number_of_species);
-
-  # Iterate through the species models, and formulate the dsa -
-  for (key,species_model::SpeciesModel) in species_model_dictionary
-
-      # index -
-      species_index = species_model.species_index;
-
-      # Query the dilution status for this species -
-      if (species_model.is_species_diluted == false)
-        dilution_selection_array[species_index] = 0.0;
-      end
-  end
-
-  return dilution_selection_array;
-end
-
-function calculateGrowthCoefficientArray(species_model_dictionary::Dict{AbstractString,SpeciesModel})
-
-  # Method variables -
-  number_of_species = length(species_model_dictionary);
-  growth_coeffient_array = zeros(Float64, number_of_species);
-
-  # Iterate through the species models, and formulate the dsa -
-  for (key,species_model::SpeciesModel) in species_model_dictionary
-
-      # index -
-      species_index = species_model.species_index;
-      if (species_model.is_biomass_precursor == true)
-
-        precursor_coeff = species_model.biomass_precursor_coefficient;
-        growth_coeffient_array[species_index] = precursor_coeff;
-
-      end
-  end
-
-  return growth_coeffient_array;
-end
-
-function calculateInitialConditionArray(species_model_dictionary::Dict{AbstractString,SpeciesModel})
-
-  # Method variables -
-  number_of_species = length(species_model_dictionary);
-  initial_condition_array = zeros(Float64,number_of_species);
-
-  # Iterate through the species models, and formulate the dsa -
-  for (key,species_model::SpeciesModel) in species_model_dictionary
-
-      # index -
-      species_index = species_model.species_index;
-      initial_condition = species_model.species_initial_condition;
-
-      # add to array -
-      initial_condition_array[species_index] = initial_condition;
-  end
-
[…]

### 2017-01-26 Code/Sobol/0/src/CreateRunDir.jl

@@ -1,9 +1,0 @@
-# Import number of runs; if this is first run, set equal to 1
-if isfile("./Sobol/num_run")
-    num_run = convert(Int64,readdlm("./Sobol/num_run")[1])+1
-else
-    num_run = 1
-end
-# Create directory for next run and save number of runs
-mkdir("./Sobol/Run$num_run")
-writedlm("./Sobol/num_run",num_run)

### 2017-01-26 Code/Sobol/0/src/CreateSobolDir.jl

@@ -1,4 +1,0 @@
-# If no directory called Sobol exists, create one
-if ~isdir("Sobol")
-    mkdir("Sobol")
-end

### 2017-01-26 Code/Sobol/0/src/Figure6.jl

@@ -1,42 +1,0 @@
-using PyPlot
-using PyCall
-@pyimport numpy as np
-
-PyCall.PyDict(matplotlib["rcParams"])["font.sans-serif"] = ["Helvetica"]
-
-Sobol = readdlm("SA.txt")
-Sobol_metab = Sobol[1:22,:]
-
-Sens = Sobol[:,4] # total order sensitivities
-CI = Sobol[:,5] # total order sensitivity confidence intervals
-
-Sens_metab = Sobol_metab[:,4] # total order sensitivities
-CI_metab = Sobol_metab[:,5] # total order sensitivity confidence intervals
-
-width = 1
-
-PyPlot.matplotlib[:rc]("text",usetex=true)
-
-# Plot all indices
-alpha = length(Sens)
-x = [1:2:2*alpha;]
-threshold = 0.05*ones(x)
-fig1, ax1 = plt[:subplots](figsize=(15,7))
-b = ax1[:bar](x,Sens,width,align="center",yerr=CI,ecolor="black",color="black")
-ax1[:set_xticks]([])
-plt[:margins](0.002)
-plt[:ylim]([0,1])
-plt[:yticks]([0:.2:1],fontsize=20);
-plt[:tick_params](axis="x",which="both",bottom="off",top="off")
-
-# Plot metabolite indices only
-alpha = length(Sens_metab)
-x = [1:2:2*alpha;]
-threshold = 0.05*ones(x)
-fig2, ax2 = plt[:subplots](figsize=(15,7))
-b = ax2[:bar](x,Sens_metab,width,align="center",yerr=CI_metab,ecolor="black",color="black")
-ax2[:set_xticks]([])
-plt[:margins](0.002)
-plt[:ylim]([0,.2])
-plt[:yticks]([0:.05:.2],fontsize=20);
-plt[:tick_params](axis="x",which="both",bottom="off",top="off")

### 2017-01-26 Code/Sobol/0/src/MassBalances.jl

@@ -1,93 +1,0 @@
-include("Kinetics.jl");
-include("Control.jl");
-
-# ----------------------------------------------------------------------------------- #
-# Copyright (c) 2016 Varnerlab
-# School of Chemical Engineering Purdue University
-# W. Lafayette IN 46907 USA
-
-# Permission is hereby granted, free of charge, to any person obtaining a copy
-# of this software and associated documentation files (the "Software"), to deal
-# in the Software without restriction, including without limitation the rights 
-# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell 
-# copies of the Software, and to permit persons to whom the Software is 
-# furnished to do so, subject to the following conditions:
-#
-# The above copyright notice and this permission notice shall be included in
-# all copies or substantial portions of the Software.
-#
-# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
-# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
-# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
-# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
-# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
-# OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
-# THE SOFTWARE.
-# ----------------------------------------------------------------------------------- #
-function MassBalances(t,x,dxdt_vector,data_dictionary)
-# ---------------------------------------------------------------------- #
-# MassBalances.jl was generated using the Kwatee code generation system.
-# Username: nicholas
-# Type: CFPS-JULIA
-# Version: 1.0
-# Generation timestamp: 05-11-2016 16:22:26
-# 
-# Arguments: 
-# t  - current time 
-# x  - state vector 
-# dxdt_vector - right hand side vector 
-# data_dictionary  - Data dictionary instance (holds model parameters) 
-# ---------------------------------------------------------------------- #
-
[…]

### 2017-01-26 Code/Sobol/0/src/ModelOutput.jl

@@ -1,15 +1,0 @@
-# Import number of jobs
-num_jobs = convert(Int64,readdlm("num_jobs")[1])
-
-# Initialize total output array
-Output = Float64[]
-
-# For each job, add job output to total output array
-i_range = 0:num_jobs-1
-for i in i_range
-    output = readdlm("../../$i/src/model_output.txt")
-    Output = [Output;output]
-end
-
-# Save in src
-writedlm("model_output.txt",Output)

### 2017-01-26 Code/Sobol/0/src/RunModel.jl

@@ -1,11 +1,0 @@
-function RunModel(num_sample)
-include("SolveBalances.jl")
-
-# Insert sobol values
-IC_vector = vec(readdlm("../../model_input/$num_sample"))
-
-# Solve balance equations
-Tsim,X = SolveBalances(IC_vector)
-
-return Tsim,X
-end

### 2017-01-26 Code/Sobol/0/src/RunSobol.jl

@@ -1,41 +1,0 @@
-include("RunModel.jl")
-include("CalcArea.jl")
-
-# Import number of runs
-num_run = convert(Int64,readdlm("./Sobol/num_run")[1])
-
-# Import number of samples; if this is first sample, set equal to 1
-if isfile("./Sobol/Run$num_run/num_sample")
-    num_sample = convert(Int64,readdlm("./Sobol/Run$num_run/num_sample")[1])+1
-else
-    num_sample = 1
-end
-
-# Import partition size and number of iterations (per job)
-partition_size = convert(Int64,readdlm("partition_size")[1])
-iterations_per_job = convert(Int64,readdlm("iterations_per_job")[1])
-
-# Import job number
-pwd_str = pwd()
-str = pwd_str[end-4]
-job = parse("$str")
-
-# Solve partition_size number of samples
-for i = num_sample:num_sample+partition_size-1
-    # Create directory for next sample and save number of samples
-    mkdir("./Sobol/Run$num_run/Sample$num_sample")
-    writedlm("./Sobol/Run$num_run/num_sample",num_sample)
-
-    # Simulate
-    Tsim,X = RunModel(iterations_per_job*job+num_sample)
-
-    # Calculate area under CAT curve and save as objective function
-    obj_fun = CalcArea(Tsim,X[:,92])
-    writedlm("./Sobol/Run$num_run/Sample$num_sample/obj_fun.dat",obj_fun)
-
-    # Save timecourse and species
-    writedlm("./Sobol/Run$num_run/Sample$num_sample/Tsim.dat",Tsim)
-    writedlm("./Sobol/Run$num_run/Sample$num_sample/X.dat",X)
-
-    num_sample += 1
-end

### 2017-01-26 Code/Sobol/0/src/SA.jl

@@ -1,11 +1,0 @@
-# Read SobolAnalysis.txt (it should omit blank lines)
-SobolAnalysis = readdlm("SobolAnalysis.txt")
-
-# Find last row that contains sensitivities (not co-sensitivities)
-last_row = find(SobolAnalysis[:,1].=="Parameter_1")[1]-1
-
-# Remove co-sensitivities, as well as headings in first row
-SA = SobolAnalysis[2:last_row,:]
-
-# Write to disk
-writedlm("SA.txt",SA)

### 2017-01-26 Code/Sobol/0/src/SALib_analyze.sh

@@ -1,9 +1,0 @@
-julia ModelOutput.jl
-
-python -m SALib.analyze.sobol \
-     -p ./param.txt \
-     -Y model_output.txt \
-     -c 0 \
-     > SobolAnalysis.txt
-
-julia SA.jl

### 2017-01-26 Code/Sobol/0/src/SALib_saltelli.sh

@@ -1,11 +1,0 @@
-n=300
-echo $n > n
-
-julia SetBounds.jl
-
-python -m SALib.sample.saltelli \
-     -n $n \
-     -p ./param.txt \
-     -o model_input.txt \
-
-julia SplitModelInput.jl

### 2017-01-26 Code/Sobol/0/src/SetBounds.jl

@@ -1,50 +1,0 @@
-# Do Sobol on initial conditions of enzymes, amino acids, glucose, and oxygen
-
-# Import optimal initial conditions
-init_cond = vec(abs(float(open(readdlm,"initial_condition.dat"))))
-
-#	54	 id:M_glc_D_c	 symbol:M_glc_D_c
-#	93	 id:M_o2_c	 symbol:M_o2_c
-#	80	 id:M_ala_L_c	 symbol:M_ala_L_c
-#	43	 id:M_arg_L_c	 symbol:M_arg_L_c
-#	76	 id:M_asn_L_c	 symbol:M_asn_L_c
-#	73	 id:M_asp_L_c	 symbol:M_asp_L_c
-#	77	 id:M_cys_L_c	 symbol:M_cys_L_c
-#	91	 id:M_gln_L_c	 symbol:M_gln_L_c
-#	90	 id:M_glu_L_c	 symbol:M_glu_L_c
-#	74	 id:M_gly_L_c	 symbol:M_gly_L_c
-#	79	 id:M_his_L_c	 symbol:M_his_L_c
-#	75	 id:M_ile_L_c	 symbol:M_ile_L_c
-#	89	 id:M_leu_L_c	 symbol:M_leu_L_c
-#	78	 id:M_lys_L_c	 symbol:M_lys_L_c
-#	88	 id:M_met_L_c	 symbol:M_met_L_c
-#	81	 id:M_phe_L_c	 symbol:M_phe_L_c
-#	82	 id:M_pro_L_c	 symbol:M_pro_L_c
-#	83	 id:M_ser_L_c	 symbol:M_ser_L_c
-#	84	 id:M_thr_L_c	 symbol:M_thr_L_c
-#	85	 id:M_trp_L_c	 symbol:M_trp_L_c
-#	86	 id:M_tyr_L_c	 symbol:M_tyr_L_c
-#	87	 id:M_val_L_c	 symbol:M_val_L_c
-
-#	104	 id:E_R_PTS	 symbol:E_R_PTS
-#	105	 id:E_R_glk_atp	 symbol:E_R_glk_atp
-#           ...                 ...
-#	265	 id:E_R_ump_ppi	 symbol:E_R_ump_ppi
-#	266	 id:E_R_ump_pi	 symbol:E_R_ump_pi
-
-# Array of initial conditions
-#IC = init_cond[[54;93;80;43;76;73;77;91;90;74;79;75;89;78;88;81:87;104:266]]
[…]

### 2017-01-26 Code/Sobol/0/src/SolveBalances.jl

@@ -1,102 +1,0 @@
-include("MassBalances.jl")
-using Sundials;
-
-# ----------------------------------------------------------------------------------- #
-# Copyright (c) 2015 Varnerlab
-# School of Chemical Engineering Purdue University
-# W. Lafayette IN 46907 USA
-
-# Permission is hereby granted, free of charge, to any person obtaining a copy
-# of this software and associated documentation files (the "Software"), to deal
-# in the Software without restriction, including without limitation the rights 
-# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell 
-# copies of the Software, and to permit persons to whom the Software is 
-# furnished to do so, subject to the following conditions:
-#
-# The above copyright notice and this permission notice shall be included in
-# all copies or substantial portions of the Software.
-#
-# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
-# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
-# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
-# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
-# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
-# OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
-# THE SOFTWARE.
-# ----------------------------------------------------------------------------------- #
-function SolveBalances(IC_vector)
-# ----------------------------------------------------------------------------------- #
-# SolveBalances.jl was generated using the Kwatee code generation system.
-# SolveBalances: Solves model equations from TSTART to TSTOP given parameters in data_dictionary.
-# Username: nicholas
-# Type: CFPS-JULIA
-# Version: 1.0
-# Generation timestamp: 12-15-2015 14:24:33
-# 
-# Input arguments: 
-# TSTART  - Time start 
-# TSTOP  - Time stop 
-# Ts - Time step 
-# data_dictionary  - Data dictionary instance (holds model parameters) 
-# 
-# Return arguments: 
-# TSIM - Simulation time vector 
[…]

### 2017-01-26 Code/Sobol/0/src/SplitModelInput.jl

@@ -1,11 +1,0 @@
-# ReadSobol samples
-model_input = readdlm("model_input.txt")
-# Create directory, if necessary
-if ~isdir("../../model_input")
-    mkdir("../../model_input")
-end
-# Save each vector separately
-i_range = 1:size(model_input)[1]
-for i in i_range
-    writedlm("../../model_input/$i",model_input[i,:])
-end

### 2017-01-26 Code/Sobol/0/src/WriteModelOutput.jl

@@ -1,16 +1,0 @@
-# Import run number and number of samples
-num_run = convert(Int64,readdlm("./Sobol/num_run")[1])
-num_sample = convert(Int64,readdlm("./Sobol/Run$num_run/num_sample")[1])
-
-# Initialize objective function array
-ObjFun = Float64[]
-
-# For each sample, import objective function value and add to objective function array
-sample_range = 1:num_sample
-for sample in sample_range
-    push!(ObjFun,readdlm("./Sobol/Run$num_run/Sample$sample/obj_fun.dat")[1])
-end
-
-# Save in run folder and in src
-writedlm("./Sobol/Run$num_run/model_output.txt",ObjFun)
-writedlm("model_output.txt",ObjFun)

### 2017-01-26 Code/Sobol/0/src/runSALib.sh

@@ -1,24 +1,0 @@
-n=$(<n)
-partition_size=186
-num_jobs=10
-echo $partition_size > partition_size
-echo $num_jobs > num_jobs
-
-julia CreateSobolDir.jl
-julia CreateRunDir.jl
-
-param=($(wc param.txt))
-num_params=${param[0]}
-
-num_iterations=$((2*n*(num_params+1)))
-iterations_per_job=$((num_iterations/num_jobs))
-echo $iterations_per_job > iterations_per_job
-
-num_partitions=$((iterations_per_job/partition_size))
-
-for j in $(eval echo "{1..$num_partitions}")
-do
-    julia RunSobol.jl
-done
-
-julia WriteModelOutput.jl

### 2017-01-26 Code/src/MassBalances.jl

@@ -1,0 +1,74 @@
+include("Kinetics.jl");
+include("Control.jl");
+
+# ----------------------------------------------------------------------------------- #
+# Copyright (c) 2016 Varnerlab
+# School of Chemical Engineering Purdue University
+# W. Lafayette IN 46907 USA
+
+# Permission is hereby granted, free of charge, to any person obtaining a copy
+# of this software and associated documentation files (the "Software"), to deal
+# in the Software without restriction, including without limitation the rights 
+# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell 
+# copies of the Software, and to permit persons to whom the Software is 
+# furnished to do so, subject to the following conditions:
+#
+# The above copyright notice and this permission notice shall be included in
+# all copies or substantial portions of the Software.
+#
+# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
+# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
+# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
+# OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
+# THE SOFTWARE.
+# ----------------------------------------------------------------------------------- #
+function MassBalances(t,x,dxdt_vector,data_dictionary,TSTOP)
+# ---------------------------------------------------------------------- #
+# MassBalances.jl was generated using the Kwatee code generation system.
+# Username: nicholas
+# Type: CFPS-JULIA
+# Version: 1.0
+# Generation timestamp: 05-11-2016 16:22:26
+# 
+# Arguments: 
+# t  - current time 
+# x  - state vector 
+# dxdt_vector - right hand side vector 
+# data_dictionary  - Data dictionary instance (holds model parameters) 
+# ---------------------------------------------------------------------- #
+
+# Correct nagative x's = throws errors in control even if small - 
[…]

### 2017-01-26 Code/src/PlotEnsemble.jl

@@ -1,0 +1,557 @@
+include("DataDictionary.jl")
+using PyPlot
+
+# Setup the timescale
+TSTART = 0.0
+TSTOP = 3.0
+Ts = 0.1
+
+# Create dictionary of parameters
+data_dictionary = DataDictionary(TSTART,TSTOP,Ts)
+
+experiment_data_dictionary = data_dictionary["EXPERIMENT_DATA_DICTIONARY"]
+
+glc = experiment_data_dictionary["M_glc_D_c"].raw_measurement_array
+pyr = experiment_data_dictionary["M_pyr_c"].raw_measurement_array
+lac = experiment_data_dictionary["M_lac_D_c"].raw_measurement_array
+ac = experiment_data_dictionary["M_ac_c"].raw_measurement_array
+succ = experiment_data_dictionary["M_succ_c"].raw_measurement_array
+mal = experiment_data_dictionary["M_mal_L_c"].raw_measurement_array
+CAT = experiment_data_dictionary["CAT"].raw_measurement_array
+
+atp = experiment_data_dictionary["M_atp_c"].raw_measurement_array
+adp = experiment_data_dictionary["M_adp_c"].raw_measurement_array
+amp = experiment_data_dictionary["M_amp_c"].raw_measurement_array
+axp = atp+adp+amp
+axp[:,1] = atp[:,1]
+gtp = experiment_data_dictionary["M_gtp_c"].raw_measurement_array
+gdp = experiment_data_dictionary["M_gdp_c"].raw_measurement_array
+gmp = experiment_data_dictionary["M_gmp_c"].raw_measurement_array
+gxp = gtp+gdp+gmp
+gxp[:,1] = gtp[:,1]
+ctp = experiment_data_dictionary["M_ctp_c"].raw_measurement_array
+cdp = experiment_data_dictionary["M_cdp_c"].raw_measurement_array
+cmp = experiment_data_dictionary["M_cmp_c"].raw_measurement_array
+cxp = ctp+cdp+cmp
+cxp[:,1] = ctp[:,1]
[…]

### 2017-01-26 Code/src/PlotSingle.jl

@@ -1,0 +1,589 @@
+include("DataDictionary.jl")
+include("SolveBalances.jl")
+include("CalcError.jl")
+using PyPlot
+
+# Setup the timescale
+TSTART = 0.0
+TSTOP = 3.0
+Ts = 0.1
+
+# Create dictionary of parameters
+data_dictionary = DataDictionary(TSTART,TSTOP,Ts)
+
+# Initialize number of parameters to be varied
+num_rate = 326
+num_sat = 455
+num_cont = 34
+
+params_best = readdlm("params_best") # This is the best-fit set
+
+# Check that params is of correct length
+if ~(num_rate+num_sat+num_cont==length(params_best) || num_rate+num_sat==length(params_best))
+    throw("Wrong number of parameters")
+end
+
+# Add rate constants to data_dictionary
+rate = params_best[1:num_rate]
+#rate[71] = 0 # Uncomment this for cyd knockout
+#rate[72] = 0 # Uncomment this for app knockout
+#rate[74] = 0 # Uncomment this for nuo knockout
+data_dictionary["RATE_CONSTANT_ARRAY"] = rate
+
+# Add saturation constants to data_dictionary
+sat = params_best[num_rate+1:num_rate+num_sat]
+sat_vector = vec(data_dictionary["SATURATION_CONSTANT_ARRAY"])
+sat_vector[find(sat_vector.>0)] = sat
+data_dictionary["SATURATION_CONSTANT_ARRAY"] = reshape(sat_vector,size(data_dictionary["SATURATION_CONSTANT_ARRAY"]))
+
+# Add control parameters to data_dictionary
+cont = params_best[num_rate+num_sat+1:num_rate+num_sat+num_cont]
+data_dictionary["CONTROL_PARAMETER_ARRAY"] = reshape(cont,2,convert(Int64,length(cont)/2))'
+
[…]

### 2017-01-28 Paper_v1.tex
Jan27

@@
 
-Gene knockouts in the electron transport chain significantly reduced the performance of the CFPS reaction (Fig.~\ref{fig:oxphos_ko}).
+Gene knockouts in the electron transport chain significantly reduced the performance of the CFPS reaction (Fig.~\ref{fig:oxKO_ssFBA_yields}).
 A key finding of both the CAT and overall system state sensitivity analysis was the importance of oxidative phosphorylation.
@@
 A single \textit{cyd} knockout reduced the CAT carbon yield from 2.7\% to 0.9\% (Table~\ref{tbl:yield_breakdown}).
-In addition, the glucose uptake rate was reduced compared to that of the control (no knockouts).
+%In addition, the glucose uptake rate was reduced compared to that of the control (no knockouts).
 %On the other hand, a \textit{nuo} knockout showed a less dramatic decrease in yield,
@@
 On the other hand, a \textit{nuo} knockout showed a less dramatic decrease in yield,
-reducing the CAT carbon yield to 2.3\%; however, the glucose uptake rate remained similar to that of the control.
+reducing the CAT carbon yield to 2.3\%.
+%; however, the glucose uptake rate remained similar to that of the control.
 %Knocking out \textit{app} increased CAT yield to 8.3\%, but this increase was not statistically significantly different from that of the control.
@@
 %The sensitivity analysis showed oxidative phosphorylation was significant for CAT production as well as the system state.
-A similar distribution of the carbon contribution to CAT yield was seen across the best-fit set and all knockouts.
+Sequence-specific flux balance analysis (ssFBA) predicted optimal CAT yields with no adjustable parameters (Fig.~\ref{fig:oxKO_ssFBA_yields}).
+Before exploring CFPS optimality, we first validated the ssFBA approach by comparing simulated and measured concentrations of CAT for the first hour of glucose consumption.
+We chose this time window (during the first phase of CAT production) because it was approximately linear in both glucose consumption and by-/production formation.
+The ssFBA calculation had no adjustable parameters; bounds on transcription and translation rates, and biochemical fluxes were either estimated from data, or
+from mechanistic models parameterized from literature.
+Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates, or the upper bounds for oxygen and glucose consumption rates was addressed
+by sampling plausible ranges for these parameters.
+The ensemble of ssFBA simulations predicted CAT formation as a function of time during the first hour of production when constrained by the experimental metabolite data (Fig.~\ref{fig:oxKO_ssFBA_yields}A).
+Thus, the molecular description of transcription and translation were consistent with experimental measurements.
+Next, to gauge the performance of the CFPS reaction, we next calculated the CAT carbon yield for three classes of constraints:
+(i) theoretical maximum glucose, amino acid and oxygen upper bounds, and realistic transcriptional/translational constraints;
+(ii) theoretical maximum glucose, amino acid and oxygen upper bounds, realistic transcriptional/translational constraints and knockouts of amino acid synthesis reactions of amino acids supplemented in the \textit{E.
+coli} extract preparation.
+(iii) metabolite fluxes constrained by the CAT data, and realistic transcriptional/translational constraints and knockouts of amino acid synthesis reactions of amino acids supplemented in the \textit{E.
+coli} extract preparation (Fig.~\ref{fig:oxKO_ssFBA_yields}B).
+The physiological theoretical maximum CAT carbon yield (case i) was 21.8\% $\pm$ 2.8\% (Fig.~\ref{fig:oxKO_ssFBA_yields}B, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, with bounded transcription and translation rates (96\% without glucose contribution in the carbon yield calculation).
+For case ii, the optimal CAT carbon yield decreased to 18.2\% $\pm$ 3.0\% (Fig.~\ref{fig:oxKO_ssFBA_yields}B, middle).
+Lastly, when metabolite constraints were applied with experimental measurements of measurements (case iii), the predicted carbon yield was 6.0\% $\pm$ 2.0\%.
+By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield (without arginine and glutamate) of 7.9\%, equivalent to 36\% of the optimal case (i).
+The experimental dataset had a CAT carbon yield of 8.2\%, similar to both the kinetic model and the experimentally constrained ssFBA calculation (case iii).
+We also investigated the energy efficiency of CAT production for the best-fit set and the three sequence-specific cases, based on the amount of CAT production versus ATP production.
+For cases i and ii the energy efficiencies were 99.8\% and 94.4\%, respectively, while for case iii it was 31.9\%.
+This dramatic decrease in efficiency when fluxes are constrained to data makes sense, as the network is forced toward a multitude of pathways that may not contribute to CAT production.
+However, the case iii efficiency was still much higher than that of the best-fit kinetic model (9.2\%).
+This is likely because of the steady-state assumption and the ability to choose the optimum from a variety of flux distributions; meanwhile, the kinetic model diverts flux toward the accumulation of sub-optimal metabolites.
+Thus, the CFPS reaction was not optimal; the ssFBA calculations suggested that an approximately three-fold increase in carbon yield and an eleven-fold increase in energy efficiency were theoretically possible.
+
+A similar distribution of the carbon contribution to CAT yield was seen across the best-fit set and all knockouts (Table~\ref{tbl:yield_breakdown}).
 In all cases, glutamate was responsible for about two-thirds of carbon consumption toward CAT.
@@
 
-%Furthermore, we examined the proportion of amino acid consumption that was due to CAT synthesis in the best-fit set (Table~\ref{tbl:AA_breakdown}).
-To better understand where carbon flux goes besides toward CAT, we examined the proportion of amino acid consumption that was due to CAT synthesis in the best-fit set (Table~\ref{tbl:AA_breakdown}).
-Eight of the 20 amino acids had a flux-toward-CAT percentage of 100\% or higher.
-For six of these (isoleucine, leucine, methionine, phenylalanine, tryptophan, valine), synthesis was inactive or negligible, meaning that CAT synthesis was the only reaction that consumed these species.
-This is why methionine, and to a lesser extent isoleucine, leucine, and valine do not fit the experimental data as well as some other amino acids; consumption via CAT synthesis was not enough to fit the data, and no other reactions exist in the network to consume these species.
-For histidine and tyrosine, synthesis was active, allowing the percentage to be higher than 100\%.
-Another ten amino acids registered percentages lower than 100\%, meaning that they were consumed by other pathways in the network.
-The most extreme of these is glutamate, with a percentage of 0.2\%, because it was consumed by numerous reactions for the synthesis of other amino acids and as an energy substrate.
-Likewise, arginine, asparagine, glycine, and proline are consumed in the production of other amino acids, while cysteine and lysine are consumed to form other metabolites.
+Furthermore, we performed a carbon balance on the best-fit set to determine the proportion of carbon consumption going toward CAT synthesis (Fig.~\ref{fig:Carbon_Balance}).
+%To better understand where carbon flux goes besides toward CAT, we examined the proportion of amino acid consumption that was due to CAT synthesis in the best-fit set (Table~\ref{tbl:AA_breakdown}).
+%Eight of the 20 amino acids had a flux-toward-CAT percentage of 100\% or higher.
+%For six of these (isoleucine, leucine, methionine, phenylalanine, tryptophan, valine), synthesis was inactive or negligible, meaning that CAT synthesis was the only reaction that consumed these species.
+%This is why methionine, and to a lesser extent isoleucine, leucine, and valine do not fit the experimental data as well as some other amino acids; consumption via CAT synthesis was not enough to fit the data, and no other reactions exist in the network to consume these species.
+%For histidine and tyrosine, synthesis was active, allowing the percentage to be higher than 100\%.
+%Another ten amino acids registered percentages lower than 100\%, meaning that they were consumed by other pathways in the network.
+%The most extreme of these is glutamate, with a percentage of 0.2\%, because it was consumed by numerous reactions for the synthesis of other amino acids and as an energy substrate.
+%Likewise, arginine, asparagine, glycine, and proline are consumed in the production of other amino acids, while cysteine and lysine are consumed to form other metabolites.
 %ARG R_arg_deg1-4 go to GLU
@@
 %THR R_thr_deg4, maybe negligible
-Finally, alanine and glutamine increased during the timecourse, meaning that synthesis outweighed all consumption (including consumption toward CAT).
-
-Sequence-specific flux balance analysis (ssFBA) predicted optimal CAT yields with no adjustable parameters (Fig.~\ref{fig:CATProdYield}).
-Before exploring CFPS optimality, we first validated the ssFBA approach by comparing simulated and measured concentrations of CAT for the first hour of glucose consumption.
-We chose this time window (during the first phase of CAT production) because it was approximately linear in both glucose consumption and by-/production formation.
-The ssFBA calculation had no adjustable parameters; bounds on transcription and translation rates, and biochemical fluxes were either estimated from data, or
-from mechanistic models parameterized from literature.
-Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates, or the upper bounds for oxygen and glucose consumption rates was addressed
-by sampling plausible ranges for these parameters.
-The ensemble of ssFBA simulations predicted CAT formation as a function of time during the first hour of production when constrained by the experimental metabolite data (Fig.~\ref{fig:CATProdYield}A).
-Thus, the molecular description of transcription and translation were consistent with experimental measurements.
-Next, to gauge the performance of the CFPS reaction, we next calculated the CAT carbon yield for three classes of constraints:
-(i) theoretical maximum glucose, amino acid and oxygen upper bounds, and realistic transcriptional/translational constraints;
-(ii) theoretical maximum glucose, amino acid and oxygen upper bounds, realistic transcriptional/translational constraints and knockouts of amino acid synthesis reactions of amino acids supplemented in the \textit{E.
-coli} extract preparation.
-(iii) metabolite fluxes constrained by the CAT data, and realistic transcriptional/translational constraints and knockouts of amino acid synthesis reactions of amino acids supplemented in the \textit{E.
-coli} extract preparation (Fig.~\ref{fig:CATProdYield}B).
-The physiological theoretical maximum CAT carbon yield (case i) was 21.8\% $\pm$ 2.8\% (Fig.~\ref{fig:CATProdYield}B, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, with bounded transcription and translation rates (96\% without glucose contribution in the carbon yield calculation).
-For case ii, the optimal CAT carbon yield decreased to 18.2\% $\pm$ 3.0\% (Fig.~\ref{fig:CATProdYield}B, middle).
-Lastly, when metabolite constraints were applied with experimental measurements of measurements (case iii), the predicted carbon yield was 6.0\% $\pm$ 2.0\%.
-By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield (without arginine and glutamate) of 7.9\%, equivalent to 36\% of the optimal case (i).
-The experimental dataset had a CAT carbon yield of 8.2\%, similar to both the kinetic model and the experimentally constrained ssFBA calculation (case iii).
-We also investigated the energy efficiency of CAT production for the best-fit set and the three sequence-specific cases, based on the amount of CAT production versus ATP production.
-For cases i and ii the energy efficiencies were 99.8\% and 94.4\%, respectively, while for case iii it was 31.9\%.
-This dramatic decrease in efficiency when fluxes are constrained to data makes sense, as the network is forced toward a multitude of pathways that may not contribute to CAT production.
-However, the case iii efficiency was still much higher than that of the best-fit kinetic model (9.2\%).
-This is likely because of the steady-state assumption and the ability to choose the optimum from a variety of flux distributions; meanwhile, the kinetic model diverts flux toward the accumulation of sub-optimal metabolites.
-Thus, the CFPS reaction was not optimal; the ssFBA calculations suggested that an approximately three-fold increase in carbon yield and an eleven-fold increase in energy efficiency were theoretically possible.
+%Finally, alanine and glutamine increased during the timecourse, meaning that synthesis outweighed all consumption (including consumption toward CAT).
 
@@
 
-\begin{figure}[ht]
-\centering
-\includegraphics[width=1\textwidth]{./Figures/oxphos_ko.pdf}
-\caption{The effect of oxidative phosphorylation on glucose uptake, CAT production and CAT carbon yield.
+%\begin{figure}[ht]
+%\centering
+%\includegraphics[width=1\textwidth]{./Figures/oxphos_ko.pdf}
+%\caption{The effect of oxidative phosphorylation on glucose uptake, CAT production and CAT carbon yield.
 A.
@@
 C.
-CAT carbon yield for 5 different cases of oxidative phosphorylation: no knockouts (blue), \textit{cyd} knockout (green), \textit{nuo} knockout (red), \textit{app} knockout (light grey), and a combination of \textit{cyd}, \textit{nuo}, \textit{app} knockouts (dark grey).}
-\label{fig:oxphos_ko}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=1\textwidth]{./Figures/CAT_prod_yield.pdf}
-\caption{Sequence-specific flux balance analysis of CAT production and yield.
+CAT carbon yield for 5 different cases of oxidative phosphorylation: no knockouts (blue), \textit{cyd} knockout (green), \textit{nuo} knockout (red), \textit{app} knockout (light gray), and a combination of \textit{cyd}, \textit{nuo}, \textit{app} knockouts (dark gray).}
+%\label{fig:oxphos_ko}
+%\end{figure}
+%\clearpage
+
+%\begin{figure}[ht]
+%\centering
+%\includegraphics[width=1\textwidth]{./Figures/CAT_prod_yield.pdf}
+%\caption{Sequence-specific flux balance analysis of CAT production and yield.
 A.
@@
 B.
-Theoretical maximum carbon yield of CAT calcualted by ssFBA for three different cases: constrained by transcription/translation (TX/TL) rates (black), same as previous but without amino acid synthesis reactions (grey), and same as previous but constrained by experimental measurements where available (light grey).
+Theoretical maximum carbon yield of CAT calcualted by ssFBA for three different cases: constrained by transcription/translation (TX/TL) rates (black), same as previous but without amino acid synthesis reactions (gray), and same as previous but constrained by experimental measurements where available (light gray).
 Error bars represent standard deviation of the ensemble.}
-\label{fig:CATProdYield}
+%\label{fig:CATProdYield}
+%\end{figure}
+%\clearpage
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1\textwidth]{./Figures/oxKO_ssFBA_yields.pdf}
+\caption{The effects of oxidative phosphorylation and amino acid synthesis pathways on CAT production and carbon yield.
+A.
+95\% confidence interval of the ensemble of kinetic models for CAT concentration versus time, for the best-fit set with no knockouts (black shaded region and dashed line), \textit{nuo} knockout (medium gray), and \textit{cyd} knockout (light gray).
+B.
+CAT carbon yield of the ensemble of kinetic models for no knockouts (black), \textit{app} knockout (dark gray), \textit{nuo} knockout (medium gray), \textit{cyd} knockout (light gray), and all three knockouts (white).
+Error bars represent standard deviation of the ensemble.
+C.
+95\% confidence interval of the ensemble of ssFBA simulations (light blue region) of CAT concentration over time, against experimental data (black).
+D.
+Theoretical maximum carbon yield of CAT production, calcualted by ssFBA for three different cases: constrained by transcription/translation (TX/TL) rates (black), same as previous but without amino acid synthesis reactions (medium gray), and same as previous but constrained by experimental measurements where available (light gray).
+Error bars represent standard deviation of the ensemble.}
+\label{fig:oxKO_ssFBA_yields}
 \end{figure}
@@
 
-%\begin{center}
-%  \captionof{table}{Your caption here}
-%    \footnotesize
-%%    \small
-%    \begin{tabular}{|c|c|c|c|c|c|c|}
-%      \hline
-%        & Best-fit & $\Delta$cyd & $\Delta$nuo & $\Delta$app & $\Delta$cyd $\Delta$nuo $\Delta$app & Data \\ \hline \hline
-%        \multicolumn{7}{|c|}{Carbon Produced (mM)} \\ \hline
-%        CAT & 20.9 & 6.5 & 18.1 & 21.4 & 5.1 & 21.6 \\ \hline \hline
-%        \multicolumn{7}{|c|}{Carbon Consumed (mM)} \\ \hline
-%        GLC & 215.4 & 215.4 & 215.4 & 215.4 & 159.8 & 215.4 \\ \hline
-%        ALA & 0.0 & 0.0 & 1.7 & 0.0 & 0.0 & 0.0 \\ \hline
-%        ARG & 10.2 & 0.9 & 1.1 & 9.9 & 1.3 & - \\ \hline
-%        ASN & 6.2 & 6.3 & 6.2 & 6.2 & 6.3 & 6.3 \\ \hline
-%        ASP & 7.5 & 0.0 & 3.9 & 7.5 & 0.0 & 9.6 \\ \hline
-%        CYS & 3.0 & 2.9 & 3.0 & 3.1 & 2.9 & 3.7 \\ \hline
-%        GLN & 0.0 & 1.8 & 0.0 & 0.0 & 2.7 & 0.0 \\ \hline
-%        GLU & 492.6 & 505.1 & 528.2 & 505.6 & 501.8 & - \\ \hline
-%        GLY & 3.1 & 1.1 & 2.6 & 3.1 & 0.9 & 1.5 \\ \hline
-%        HIS & 0.2 & 0.4 & 1.1 & 0.2 & 0.3 & 0.0 \\ \hline
-%        ILE & 1.0 & 0.3 & 0.8 & 1.0 & 0.2 & 1.7 \\ \hline
-%        LEU & 1.4 & 0.4 & 1.2 & 1.4 & 0.3 & 2 \\ \hline
-%        LYS & 10.7 & 13.2 & 13.1 & 10.7 & 13.2 & 8.3 \\ \hline
-%        MET & 0.8 & 0.2 & 0.7 & 0.8 & 0.2 & 2.9 \\ \hline
-%        PHE & 3.2 & 1.0 & 2.8 & 3.3 & 0.8 & 1.6 \\ \hline
-%        PRO & 2.4 & 0.2 & 0.7 & 2.4 & 0.2 & 1.9 \\ \hline
-%        SER & 2.5 & 2.1 & 2.4 & 2.5 & 2.1 & 1.8 \\ \hline
-%        THR & 3.4 & 2.9 & 3.3 & 3.4 & 2.8 & 2.8 \\ \hline
-%        TRP & 1.0 & 0.3 & 0.8 & 1.0 & 0.2 & 1.2 \\ \hline
-%        TYR & 1.1 & 0.4 & 1.1 & 1.1 & 0.4 & 0.6 \\ \hline
-%        VAL & 1.4 & 0.4 & 1.2 & 1.5 & 0.4 & 2.4 \\ \hline
-%        $\Sigma$ & 767.1 & 755.3 & 791.3 & 780.1 & 696.8 & - \\ \hline
-%        $\Sigma$ w/o ARG,GLU & 264.3 & 249.3 & 262.0 & 264.6 & 193.7 & 263.7 \\ \hline \hline
-%        Yield & 2.7\% & 0.9\% & 2.3\% & 2.7\% & 0.7\% & - \\ \hline
-%        Yield w/o ARG, GLU & 7.9\% & 2.6\% & 6.9\% & 8.1\% & 2.7\% & 8.2\% \\ \hline
-%    \end{tabular}
-%\end{center}
-
-%\noindent% Insures there's no paragraph indent
-%\begin{minipage}{\textwidth}% Minipage has width exactly the same as the text block
-%  \centering% Centers the contents of the minipage
-%  This is an interesting table.% Your caption goes here.
-%  \bigskip
-%  \bigskip
-%  \footnotesize
-%  \begin{tabular}{|c|c|c|c|c|c|c|}
-%      \hline
-%        & Best-fit & $\Delta$cyd & $\Delta$nuo & $\Delta$app & $\Delta$cyd $\Delta$nuo $\Delta$app & Data \\ \hline \hline
-%        \multicolumn{7}{|c|}{Carbon Produced (mM)} \\ \hline
-%        CAT & 20.9 & 6.5 & 18.1 & 21.4 & 5.1 & 21.6 \\ \hline \hline
-%        \multicolumn{7}{|c|}{Carbon Consumed (mM)} \\ \hline
-%        GLC & 215.4 & 215.4 & 215.4 & 215.4 & 159.8 & 215.4 \\ \hline
-%        ALA & 0.0 & 0.0 & 1.7 & 0.0 & 0.0 & 0.0 \\ \hline
-%        ARG & 10.2 & 0.9 & 1.1 & 9.9 & 1.3 & - \\ \hline
-%        ASN & 6.2 & 6.3 & 6.2 & 6.2 & 6.3 & 6.3 \\ \hline
-%        ASP & 7.5 & 0.0 & 3.9 & 7.5 & 0.0 & 9.6 \\ \hline
-%        CYS & 3.0 & 2.9 & 3.0 & 3.1 & 2.9 & 3.7 \\ \hline
-%        GLN & 0.0 & 1.8 & 0.0 & 0.0 & 2.7 & 0.0 \\ \hline
-%        GLU & 492.6 & 505.1 & 528.2 & 505.6 & 501.8 & - \\ \hline
-%        GLY & 3.1 & 1.1 & 2.6 & 3.1 & 0.9 & 1.5 \\ \hline
-%        HIS & 0.2 & 0.4 & 1.1 & 0.2 & 0.3 & 0.0 \\ \hline
-%        ILE & 1.0 & 0.3 & 0.8 & 1.0 & 0.2 & 1.7 \\ \hline
-%        LEU & 1.4 & 0.4 & 1.2 & 1.4 & 0.3 & 2 \\ \hline
-%        LYS & 10.7 & 13.2 & 13.1 & 10.7 & 13.2 & 8.3 \\ \hline
-%        MET & 0.8 & 0.2 & 0.7 & 0.8 & 0.2 & 2.9 \\ \hline
-%        PHE & 3.2 & 1.0 & 2.8 & 3.3 & 0.8 & 1.6 \\ \hline
-%        PRO & 2.4 & 0.2 & 0.7 & 2.4 & 0.2 & 1.9 \\ \hline
-%        SER & 2.5 & 2.1 & 2.4 & 2.5 & 2.1 & 1.8 \\ \hline
-%        THR & 3.4 & 2.9 & 3.3 & 3.4 & 2.8 & 2.8 \\ \hline
-%        TRP & 1.0 & 0.3 & 0.8 & 1.0 & 0.2 & 1.2 \\ \hline
-%        TYR & 1.1 & 0.4 & 1.1 & 1.1 & 0.4 & 0.6 \\ \hline
-%        VAL & 1.4 & 0.4 & 1.2 & 1.5 & 0.4 & 2.4 \\ \hline
-%        $\Sigma$ & 767.1 & 755.3 & 791.3 & 780.1 & 696.8 & - \\ \hline
-%        $\Sigma$ w/o ARG,GLU & 264.3 & 249.3 & 262.0 & 264.6 & 193.7 & 263.7 \\ \hline \hline
-%        Yield & 2.7\% & 0.9\% & 2.3\% & 2.7\% & 0.7\% & - \\ \hline
-%        Yield w/o ARG, GLU & 7.9\% & 2.6\% & 6.9\% & 8.1\% & 2.7\% & 8.2\% \\ \hline
-%    \end{tabular}
-%\end{minipage}
-
 \begin{table}
 \centering
-    \caption{CAT carbon yield breakdown for best-fit set, knockouts, and experimental data.
-Carbon produced as CAT, carbon consumed as glucose and each amino acid, sum of consumed species (with and without arginine and glutamate), and yield  (with and without arginine and glutamate).
-Energy efficiency for best-fit set, knockouts, and experimental data.}
+    \caption{CAT carbon yield breakdown for best-fit set, knockouts, and experimental data. Carbon produced as CAT, carbon consumed as glucose and each amino acid, sum of consumed species, and yield.}
     \renewcommand{\arraystretch}{1.3}
     \begin{tabular}{lrrrrrr} \toprule
-        Carbon Produced (C-mM) & Best-fit & $\Delta$cyd & $\Delta$nuo & $\Delta$app & $\Delta$cyd $\Delta$nuo $\Delta$app & Data \\ \hline
+        Carbon Produced (C-mM) & Best-fit & $\Delta$cyd & $\Delta$nuo & $\Delta$app & \parbox{1 cm}{$\Delta$cyd $\Delta$nuo $\Delta$app} & Data \\ \hline
         CAT & 20.9 & 6.5 & 18.1 & 21.4 & 5.1 & 21.6 \\ \midrule
@@
         ALA & 0.0 & 0.0 & 1.7 & 0.0 & 0.0 & 0.0 \\ \hline
-        ARG & 10.2 & 0.9 & 1.1 & 9.9 & 1.3 & - \\ \hline
+%        ARG & 10.2 & 0.9 & 1.1 & 9.9 & 1.3 & - \\ \hline
         ASN & 6.2 & 6.3 & 6.2 & 6.2 & 6.3 & 6.3 \\ \hline
@@
         GLN & 0.0 & 1.8 & 0.0 & 0.0 & 2.7 & 0.0 \\ \hline
-        GLU & 492.6 & 505.1 & 528.2 & 505.6 & 501.8 & - \\ \hline
+%        GLU & 492.6 & 505.1 & 528.2 & 505.6 & 501.8 & - \\ \hline
         GLY & 3.1 & 1.1 & 2.6 & 3.1 & 0.9 & 1.5 \\ \hline
@@
         VAL & 1.4 & 0.4 & 1.2 & 1.5 & 0.4 & 2.4 \\ \midrule
-        Sum & 767.1 & 755.3 & 791.3 & 780.1 & 696.8 & - \\ \hline
-        Sum w/o ARG,GLU & 264.3 & 249.3 & 262.0 & 264.6 & 193.7 & 263.7 \\ \midrule
-        Yield & 2.7\% & 0.9\% & 2.3\% & 2.7\% & 0.7\% & - \\ \hline
-        Yield w/o ARG, GLU & 7.9\% & 2.6\% & 6.9\% & 8.1\% & 2.7\% & 8.2\% \\ \bottomrule
-%        Energy Efficiency & 1.8\% & 0.6\% & 1.6\% & 1.8\% & 0.6\% & 1.9\% \\ \bottomrule
-%        Energy Efficiency & 2.6\% & 0.8\% & 2.2\% & 2.6\% & 0.8\% & 2.6\% \\ \bottomrule
-%        Efficiencies based on ATP number of 17: 2.558085 & 0.788758 & 2.210560 & 2.611583 & 0.847068 & 2.644330
+%        Sum & 767.1 & 755.3 & 791.3 & 780.1 & 696.8 & - \\ \hline
+        Sum & 264.3 & 249.3 & 262.0 & 264.6 & 193.7 & 263.7 \\ \midrule
+%        Yield & 2.7\% & 0.9\% & 2.3\% & 2.7\% & 0.7\% & - \\ \hline
+        Yield & 7.9\% & 2.6\% & 6.9\% & 8.1\% & 2.7\% & 8.2\% \\ \bottomrule
     \end{tabular}
@@
 
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.75\textwidth]{./Figures/Carbon-Balance-BFS.pdf}
+\caption{Carbon balance for the best-fit set.
+Carbon moles produced as CAT, amino acids (alanine and glutamine), organic acids (lactate, acetate, succinate, and malate), and other byproducts including carbon dioxide, as percentages of total carbon consumption (glucose and all other amino acids).}
+\label{fig:Carbon_Balance}
+\end{figure}
+
 %\begin{table}
 %\centering
-%    \caption{Amino acid consumption toward CAT as a percentage of total amino acid consumption, for the best-fit set.
-Total consumption defined as decrease in amino acid concentration over 3 hours, such that for amino acids with synthesis active, percentage may be greater than 100\%.
-Alanine and glutamine did not decrease over the timecourse.}
-%    \renewcommand{\arraystretch}{1.25}
+%    \caption{Amino acid total net consumption, consumption toward CAT, and net consumption other than toward CAT for the best-fit set.
+Total net consumption defined as decrease in amino acid concentration over 3 hours.
+Consumption toward CAT defined as increase in CAT concentration over 3 hours, multiplied by stoichiometric coefficient for given amino acid in CAT synthesis.}
+%    \renewcommand{\arraystretch}{1.3}
 %    \begin{tabular}{lrrr} \toprule
-%        Species & Total Consumption (mM) & Consumption Toward CAT (mM) & Percentage \\ \hline
-%%        ALA  & -3.852  & 0.267  & -6.9\% \\ \hline
-%        ALA  & - & 0.27 & - \\ \hline
-%%        ARG  & 1.697  & 0.178  & 10.5\% \\ \hline
-%%        ASN  & 1.547  & 0.089  & 5.8\% \\ \hline
-%%        ASP  & 1.875  & 0.214  & 11.4\% \\ \hline
-%%        CYS  & 1.016  & 0.089  & 8.8\% \\ \hline
-%        ARG  & 1.70  & 0.18  & 11\% \\ \hline
-%        ASN  & 1.55  & 0.09  & 6\% \\ \hline
-%        ASP  & 1.87  & 0.21  & 11\% \\ \hline
-%        CYS  & 1.02  & 0.09  & 9\% \\ \hline
-%%        GLN  & -2.278  & 0.214  & -9.4\% \\ \hline
-%        GLN  & - & 0.21 & - \\ \hline
-%%        GLU  & 98.526  & 0.232  & 0.2\% \\ \hline
-%%        GLY  & 1.545  & 0.178  & 11.5\% \\ \hline
-%%        HIS  & 0.036  & 0.214  & 587.9\% \\ \hline
-%%        ILE  & 0.16  & 0.16  & 100\% \\ \hline
-%%        LEU  & 0.231  & 0.232  & 100.2\% \\ \hline
-%%        LYS  & 1.785  & 0.214  & 12.0\% \\ \hline
-%%        MET  & 0.16  & 0.16  & 100\% \\ \hline
-%%        PHE  & 0.356  & 0.356  & 100\% \\ \hline
-%%        PRO  & 0.481  & 0.125  & 26\% \\ \hline
-%%        SER  & 0.83  & 0.178  & 21.5\% \\ \hline
-%%        THR  & 0.857  & 0.232  & 27\% \\ \hline
-%%        TRP  & 0.089  & 0.089  & 100\% \\ \hline
-%%        TYR  & 0.12  & 0.196  & 164\% \\ \hline
-%%        VAL  & 0.285  & 0.285  & 100\% \\ \hline
-%        GLU  & 98.53  & 0.23  & 0.2\% \\ \hline
-%        GLY  & 1.55  & 0.18  & 12\% \\ \hline
-%        HIS  & 0.04  & 0.21  & 588\% \\ \hline
-%        ILE  & 0.16  & 0.16  & 100\% \\ \hline
-%        LEU  & 0.23  & 0.23  & 100\% \\ \hline
-%        LYS  & 1.79  & 0.21  & 12\% \\ \hline
-%        MET  & 0.16  & 0.16  & 100\% \\ \hline
-%        PHE  & 0.36  & 0.36  & 100\% \\ \hline
-%        PRO  & 0.48  & 0.12  & 26\% \\ \hline
-%        SER  & 0.83  & 0.18  & 21\% \\ \hline
-%        THR  & 0.86  & 0.23  & 27\% \\ \hline
-%        TRP  & 0.09  & 0.09  & 100\% \\ \hline
-%        TYR  & 0.12  & 0.20  & 164\% \\ \hline
-%        VAL  & 0.29  & 0.29  & 100\% \\ \bottomrule
+%        & \parbox{3.6cm}{Total Net \phantom{abcdefgh} Consumption (mM)} & \parbox{3.4cm}{Consumption \phantom{abc} Toward CAT (mM)} & \parbox{3.6cm}{Other \phantom{abcdefgh} Consumption (mM)} \\ \hline
+%        ALA  & -3.85 & 0.27 & -4.12 \\ \hline
+%        ARG  & 1.70  & 0.18  & 1.52 \\ \hline
+%        ASN  & 1.55  & 0.09  & 1.46 \\ \hline
+%        ASP  & 1.87  & 0.21  & 1.66 \\ \hline
+%        CYS  & 1.02  & 0.09  & 0.93 \\ \hline
+%        GLN  & -2.28 & 0.21 & -2.49 \\ \hline
+%        GLU  & 98.53  & 0.23  & 98.30 \\ \hline
+%        GLY  & 1.55  & 0.18  & 1.37 \\ \hline
+%        HIS  & 0.04  & 0.21  & -0.17 \\ \hline
+%        ILE  & 0.16  & 0.16  & 0.00 \\ \hline
+%        LEU  & 0.23  & 0.23  & 0.00 \\ \hline
+%        LYS  & 1.79  & 0.21  & 1.58 \\ \hline
+%        MET  & 0.16  & 0.16  & 0.00 \\ \hline
+%        PHE  & 0.36  & 0.36  & 0.00 \\ \hline
+%        PRO  & 0.48  & 0.12  & 0.36 \\ \hline
+%        SER  & 0.83  & 0.18  & 0.65 \\ \hline
+%        THR  & 0.86  & 0.23  & 0.63 \\ \hline
+%        TRP  & 0.09  & 0.09  & 0.00 \\ \hline
+%        TYR  & 0.12  & 0.20  & -0.08 \\ \hline
+%        VAL  & 0.29  & 0.29  & 0.00 \\ \bottomrule
 %    \end{tabular}
@@
 %\end{table}
-
-\begin{table}
-\centering
-    \caption{Amino acid total net consumption, consumption toward CAT, and net consumption other than toward CAT for the best-fit set.
-Total net consumption defined as decrease in amino acid concentration over 3 hours.
-Consumption toward CAT defined as increase in CAT concentration over 3 hours, multiplied by stoichiometric coefficient for given amino acid in CAT synthesis.}
-    \renewcommand{\arraystretch}{1.3}
-    \begin{tabular}{lrrr} \toprule
-        & \parbox{3.6cm}{Total Net \phantom{abcdefgh} Consumption (mM)} & \parbox{3.4cm}{Consumption \phantom{abc} Toward CAT (mM)} & \parbox{3.6cm}{Other \phantom{abcdefgh} Consumption (mM)} \\ \hline
-        ALA  & -3.85 & 0.27 & -4.12 \\ \hline
-        ARG  & 1.70  & 0.18  & 1.52 \\ \hline
-        ASN  & 1.55  & 0.09  & 1.46 \\ \hline
-        ASP  & 1.87  & 0.21  & 1.66 \\ \hline
-        CYS  & 1.02  & 0.09  & 0.93 \\ \hline
-        GLN  & -2.28 & 0.21 & -2.49 \\ \hline
-        GLU  & 98.53  & 0.23  & 98.30 \\ \hline
-        GLY  & 1.55  & 0.18  & 1.37 \\ \hline
-        HIS  & 0.04  & 0.21  & -0.17 \\ \hline
-        ILE  & 0.16  & 0.16  & 0.00 \\ \hline
-        LEU  & 0.23  & 0.23  & 0.00 \\ \hline
-        LYS  & 1.79  & 0.21  & 1.58 \\ \hline
-        MET  & 0.16  & 0.16  & 0.00 \\ \hline
-        PHE  & 0.36  & 0.36  & 0.00 \\ \hline
-        PRO  & 0.48  & 0.12  & 0.36 \\ \hline
-        SER  & 0.83  & 0.18  & 0.65 \\ \hline
-        THR  & 0.86  & 0.23  & 0.63 \\ \hline
-        TRP  & 0.09  & 0.09  & 0.00 \\ \hline
-        TYR  & 0.12  & 0.20  & -0.08 \\ \hline
-        VAL  & 0.29  & 0.29  & 0.00 \\ \bottomrule
-    \end{tabular}
-\label{tbl:AA_breakdown}
-\end{table}
 

### 2017-02-03 Paper_v1.tex
Feb3

@@
 Pyridine nucleotide transhydrogenase catalyzes two reactions: one converts NAD and NADPH into NADH and NADP, while the other does the reverse and also generates a proton gradient.
-Increasing one or the other has little effect on the model, as the reaction is hampered by the loss of reactants.
-Increasing both, however, allows all cometabolites to be conserved and the reactions to continue unhindered.
+Increasing one or the other has little effect on the model, as any increase in reaction flux is hampered by the imbalance created.
+Increasing both, however, allows all cometabolites to balance, so that the reaction fluxes increase unhindered.
 Thus, the pairwise sensitivity is much higher than the sum of the first-order sensitivities.
@@
 
-Gene knockouts in the electron transport chain significantly reduced the performance of the CFPS reaction (Fig.~\ref{fig:oxKO_ssFBA_yields}).
+To understand whether the CFPS performance was optimal, we calculated the carbon yield and energy efficiency of CAT production (Fig.~\ref{fig:?????}).
+The best-fit parameter set for the kinetic model predicted a CAT carbon yield of 7.9\%, while the experimental dataset had a CAT carbon yield of 8.2\%.
+This was calculated as the increase in CAT concentration times the CAT carbon number, divided by the sum of the consumption terms for glucose and all amino acids excpet arginine and glutamate, as no data were available for these, weighted by their respective carbon numbers.
+To explain where the remainder of carbon was going, we performed a carbon balance for the best-fit set.
+Of the other 92\% of carbon, 35\% accumulated as organic acids (lactate, acetate, succinate and malate) and 9\% accumulated as amino acids (alanine and glutamine).
+The remaining 48\% went to the net accumulation of all other metabolites, particularly carbon dioxide.
+The best-fit set produced CAT with an energy efficiency of 6.4\%.
+This was calculated as the increase in CAT concentration times the CAT number of equivalent ATP molecules, divided by glucose consumption times times the number of equivalent ATP molecules for glucose, equal to 15 in the optimal case.
+An additional 62\% of energy went to the accumulation of glycolysis metabolites, 6.4\% to TCA cycle metabolites, 0.2\% to pentose phosphate metabolites, and 25\% to acetate and lactate.
+This shows that there is much room for improvement of the efficiency of CFPS.
+Gene knockouts in the electron transport chain further reduced the performance of the CFPS reaction (Fig.~\ref{fig:oxKO_ssFBA_yields}).
 A key finding of both the CAT and overall system state sensitivity analysis was the importance of oxidative phosphorylation.
-To investigate this further, we knocked out key oxidative phosphorylation reactions in the ensemble of kinetic models to examine the effect on glucose uptake and CAT production.
-%A single \textit{cyd} knockout reduced the CAT carbon yield from 7.9\% to 2.6\% (Table~\ref{tbl:yield_breakdown}).
-A single \textit{cyd} knockout reduced the CAT carbon yield from 2.7\% to 0.9\% (Table~\ref{tbl:yield_breakdown}).
-%In addition, the glucose uptake rate was reduced compared to that of the control (no knockouts).
-%On the other hand, a \textit{nuo} knockout showed a less dramatic decrease in yield,
-%reducing the CAT carbon yield to 6.9\%; however, the glucose uptake rate remained similar to that of the control.
+To investigate this further, we knocked out key oxidative phosphorylation reactions in the ensemble of kinetic models to examine the effect on CAT production and carbon yield.
+A single \textit{cyd} knockout reduced the CAT carbon yield from 7.9\% to 2.6\% (Table~\ref{tbl:yield_breakdown}).
 On the other hand, a \textit{nuo} knockout showed a less dramatic decrease in yield,
-reducing the CAT carbon yield to 2.3\%.
-%; however, the glucose uptake rate remained similar to that of the control.
-%Knocking out \textit{app} increased CAT yield to 8.3\%, but this increase was not statistically significantly different from that of the control.
-Knocking out \textit{app} did not change the CAT yield (it remained at 2.7\%).
-%Lastly, knocking out all three reactions reduced the CAT yield to 3.0\%, but this was not statistically significantly different from the \textit{cyd} knockout alone.
+reducing the CAT carbon yield to 6.9\%.
+Knocking out \textit{app} increased CAT yield to 8.1\%, but this increase was not statistically significantly different from that of the control.
 Lastly, knocking out all three reactions reduced the CAT yield to 0.7\%, similar to knocking out the \textit{cyd} alone.
 Thus, the model suggested the key step in oxidative phosphorylation was catalyzed by the gene product of \textit{cyd}.
-However, while disruption of \textit{cyd} significantly reduced the CAT carbon yield, it did not eliminate the ability of CFPS reaction to produce CAT.
+However, while disruption of \textit{cyd} significantly reduced the CAT carbon yield, it did not completely eliminate the production of CAT.
 This suggested there was a mixture of energy sources supporting CAT production, with the most significant being oxidative phosphorylation.
 %The \textit{cyd} enzyme is responsible for transferring electrons with oxygen as an electron acceptor allowing the release of energy to form ATP.
-%The sensitivity analysis showed oxidative phosphorylation was significant for CAT production as well as the system state.
+
 Sequence-specific flux balance analysis (ssFBA) predicted optimal CAT yields with no adjustable parameters (Fig.~\ref{fig:oxKO_ssFBA_yields}).
 Before exploring CFPS optimality, we first validated the ssFBA approach by comparing simulated and measured concentrations of CAT for the first hour of glucose consumption.
-We chose this time window (during the first phase of CAT production) because it was approximately linear in both glucose consumption and by-/production formation.
-The ssFBA calculation had no adjustable parameters; bounds on transcription and translation rates, and biochemical fluxes were either estimated from data, or
-from mechanistic models parameterized from literature.
+We chose this time window (during the first phase of CAT production) because it was approximately linear both in glucose consumption and in the accumulation of most organic acids.
+As the ssFBA calculation had no adjustable parameters, bounds on transcription and translation rates and biochemical fluxes were either estimated from data or from mechanistic models parameterized from literature.
 Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates, or the upper bounds for oxygen and glucose consumption rates was addressed
 by sampling plausible ranges for these parameters.
-The ensemble of ssFBA simulations predicted CAT formation as a function of time during the first hour of production when constrained by the experimental metabolite data (Fig.~\ref{fig:oxKO_ssFBA_yields}A).
+The ensemble of ssFBA simulations predicted CAT formation as a function of time during the first hour of production when constrained by the experimental metabolite data (Fig.~\ref{fig:oxKO_ssFBA_yields}C).
 Thus, the molecular description of transcription and translation were consistent with experimental measurements.
@@
 (iii) metabolite fluxes constrained by the CAT data, and realistic transcriptional/translational constraints and knockouts of amino acid synthesis reactions of amino acids supplemented in the \textit{E.
-coli} extract preparation (Fig.~\ref{fig:oxKO_ssFBA_yields}B).
-The physiological theoretical maximum CAT carbon yield (case i) was 21.8\% $\pm$ 2.8\% (Fig.~\ref{fig:oxKO_ssFBA_yields}B, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, with bounded transcription and translation rates (96\% without glucose contribution in the carbon yield calculation).
-For case ii, the optimal CAT carbon yield decreased to 18.2\% $\pm$ 3.0\% (Fig.~\ref{fig:oxKO_ssFBA_yields}B, middle).
-Lastly, when metabolite constraints were applied with experimental measurements of measurements (case iii), the predicted carbon yield was 6.0\% $\pm$ 2.0\%.
-By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield (without arginine and glutamate) of 7.9\%, equivalent to 36\% of the optimal case (i).
+coli} extract preparation (Fig.~\ref{fig:oxKO_ssFBA_yields}D).
+The physiological theoretical maximum CAT carbon yield (case i) was 21.8\% $\pm$ 2.8\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, with bounded transcription and translation rates (96\% without glucose contribution in the carbon yield calculation).
+For case ii, the optimal CAT carbon yield decreased to 18.2\% $\pm$ 3.0\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, middle).
+Lastly, when metabolite constraints were applied with experimental measurements of measurements (case iii), the predicted carbon yield was 6.0\% $\pm$ 2.0\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, right).
+%By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield (without arginine and glutamate) of 7.9\%, equivalent to 36\% of the optimal case (i).
+By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield of 7.9\%, equivalent to 36\% of the optimal case (i).
 The experimental dataset had a CAT carbon yield of 8.2\%, similar to both the kinetic model and the experimentally constrained ssFBA calculation (case iii).
@@
 
-A similar distribution of the carbon contribution to CAT yield was seen across the best-fit set and all knockouts (Table~\ref{tbl:yield_breakdown}).
-In all cases, glutamate was responsible for about two-thirds of carbon consumption toward CAT.
-This was due to its role in generating other amino acids and as a possible energy source for metabolism.
-The much greater consumption of this one amino acid was made possible by its much larger initial condition, as it was present in the cell media in the form of magnesium glutamate, ammonium glutamate, and potassium glutamate.
-% [CITE? THESIS NOT CURRENTLY IN BIBLIOGRAPHY].
-Glucose made the next largest contribution, between 20\% and 30\%, dwarfing all amino acids other than glutamate.
-This makes sense, as glucose powers the entire metabolism during the first hour, including energy species synthesis and amino acid synthesis.
+%A similar distribution of the carbon contribution to CAT yield was seen across the best-fit set and all knockouts (Table~\ref{tbl:yield_breakdown}).
+%In all cases, glutamate was responsible for about two-thirds of carbon consumption toward CAT.
+%This was due to its role in generating other amino acids and as a possible energy source for metabolism.
+%The much greater consumption of this one amino acid was made possible by its much larger initial condition, as it was present in the cell media in the form of magnesium glutamate, ammonium glutamate, and potassium glutamate.
+%% [CITE? THESIS NOT CURRENTLY IN BIBLIOGRAPHY].
+%Glucose made the next largest contribution, between 20\% and 30\%, dwarfing all amino acids other than glutamate.
+%This makes sense, as glucose powers the entire metabolism during the first hour, including energy species synthesis and amino acid synthesis.
 %We also investigated the energy efficiency of CAT production across the kinetic models and experimental dataset.
@@
 %The energy efficiencies across the different cases were more or less proportional to the carbon yields, and in turn the total amount of CAT synthesis.
-Taken together, these results show that oxidative phosphorylation, particularly \textit{cyd}, is important to efficient CAT production.
+%Taken together, these results show that oxidative phosphorylation, particularly \textit{cyd}, is important to efficient CAT production.
 
 Furthermore, we performed a carbon balance on the best-fit set to determine the proportion of carbon consumption going toward CAT synthesis (Fig.~\ref{fig:Carbon_Balance}).
+
 %To better understand where carbon flux goes besides toward CAT, we examined the proportion of amino acid consumption that was due to CAT synthesis in the best-fit set (Table~\ref{tbl:AA_breakdown}).
@@
     \begin{tabular}{lrrrrrr} \toprule
-        Carbon Produced (C-mM) & Best-fit & $\Delta$cyd & $\Delta$nuo & $\Delta$app & \parbox{1 cm}{$\Delta$cyd $\Delta$nuo $\Delta$app} & Data \\ \hline
-        CAT & 20.9 & 6.5 & 18.1 & 21.4 & 5.1 & 21.6 \\ \midrule
+        Carbon Produced (C-mM) & Best-fit & $\Delta$app & $\Delta$nuo & $\Delta$cyd & \parbox{1 cm}{$\Delta$app $\Delta$nuo $\Delta$cyd} & Data \\ \hline
+        CAT & 20.9 & 21.4 & 18.1 & 6.5 & 5.1 & 21.6 \\ \midrule
         Carbon Consumed (C-mM) \\ \hline
@@
         ALA & 0.0 & 0.0 & 1.7 & 0.0 & 0.0 & 0.0 \\ \hline
-%        ARG & 10.2 & 0.9 & 1.1 & 9.9 & 1.3 & - \\ \hline
-        ASN & 6.2 & 6.3 & 6.2 & 6.2 & 6.3 & 6.3 \\ \hline
-        ASP & 7.5 & 0.0 & 3.9 & 7.5 & 0.0 & 9.6 \\ \hline
-        CYS & 3.0 & 2.9 & 3.0 & 3.1 & 2.9 & 3.7 \\ \hline
-        GLN & 0.0 & 1.8 & 0.0 & 0.0 & 2.7 & 0.0 \\ \hline
-%        GLU & 492.6 & 505.1 & 528.2 & 505.6 & 501.8 & - \\ \hline
-        GLY & 3.1 & 1.1 & 2.6 & 3.1 & 0.9 & 1.5 \\ \hline
-        HIS & 0.2 & 0.4 & 1.1 & 0.2 & 0.3 & 0.0 \\ \hline
-        ILE & 1.0 & 0.3 & 0.8 & 1.0 & 0.2 & 1.7 \\ \hline
-        LEU & 1.4 & 0.4 & 1.2 & 1.4 & 0.3 & 2 \\ \hline
-        LYS & 10.7 & 13.2 & 13.1 & 10.7 & 13.2 & 8.3 \\ \hline
-        MET & 0.8 & 0.2 & 0.7 & 0.8 & 0.2 & 2.9 \\ \hline
-        PHE & 3.2 & 1.0 & 2.8 & 3.3 & 0.8 & 1.6 \\ \hline
-        PRO & 2.4 & 0.2 & 0.7 & 2.4 & 0.2 & 1.9 \\ \hline
-        SER & 2.5 & 2.1 & 2.4 & 2.5 & 2.1 & 1.8 \\ \hline
-        THR & 3.4 & 2.9 & 3.3 & 3.4 & 2.8 & 2.8 \\ \hline
-        TRP & 1.0 & 0.3 & 0.8 & 1.0 & 0.2 & 1.2 \\ \hline
-        TYR & 1.1 & 0.4 & 1.1 & 1.1 & 0.4 & 0.6 \\ \hline
-        VAL & 1.4 & 0.4 & 1.2 & 1.5 & 0.4 & 2.4 \\ \midrule
-%        Sum & 767.1 & 755.3 & 791.3 & 780.1 & 696.8 & - \\ \hline
-        Sum & 264.3 & 249.3 & 262.0 & 264.6 & 193.7 & 263.7 \\ \midrule
-%        Yield & 2.7\% & 0.9\% & 2.3\% & 2.7\% & 0.7\% & - \\ \hline
-        Yield & 7.9\% & 2.6\% & 6.9\% & 8.1\% & 2.7\% & 8.2\% \\ \bottomrule
+%        ARG & 10.2 & 9.9 & 1.1 & 0.9 & 1.3 & - \\ \hline
+        ASN & 6.2 & 6.2 & 6.2 & 6.3 & 6.3 & 6.3 \\ \hline
+        ASP & 7.5 & 7.5 & 3.9 & 0.0 & 0.0 & 9.6 \\ \hline
+        CYS & 3.0 & 3.1 & 3.0 & 2.9 & 2.9 & 3.7 \\ \hline
+        GLN & 0.0 & 0.0 & 0.0 & 1.8 & 2.7 & 0.0 \\ \hline
+%        GLU & 492.6 & 505.6 & 528.2 & 505.1 & 501.8 & - \\ \hline
+        GLY & 3.1 & 3.1 & 2.6 & 1.1 & 0.9 & 1.5 \\ \hline
+        HIS & 0.2 & 0.2 & 1.1 & 0.4 & 0.3 & 0.0 \\ \hline
+        ILE & 1.0 & 1.0 & 0.8 & 0.3 & 0.2 & 1.7 \\ \hline
+        LEU & 1.4 & 1.4 & 1.2 & 0.4 & 0.3 & 2.0 \\ \hline
+        LYS & 10.7 & 10.7 & 13.1 & 13.2 & 13.2 & 8.3 \\ \hline
+        MET & 0.8 & 0.8 & 0.7 & 0.2 & 0.2 & 2.9 \\ \hline
+        PHE & 3.2 & 3.3 & 2.8 & 1.0 & 0.8 & 1.6 \\ \hline
+        PRO & 2.4 & 2.4 & 0.7 & 0.2 & 0.2 & 1.9 \\ \hline
+        SER & 2.5 & 2.5 & 2.4 & 2.1 & 2.1 & 1.8 \\ \hline
+        THR & 3.4 & 3.4 & 3.3 & 2.9 & 2.8 & 2.8 \\ \hline
+        TRP & 1.0 & 1.0 & 0.8 & 0.3 & 0.2 & 1.2 \\ \hline
+        TYR & 1.1 & 1.1 & 1.1 & 0.4 & 0.4 & 0.6 \\ \hline
+        VAL & 1.4 & 1.5 & 1.2 & 0.4 & 0.4 & 2.4 \\ \midrule
+%        Sum & 767.1 & 780.1 & 791.3 & 755.3 & 696.8 & - \\ \hline
+        Sum & 264.3 & 264.6 & 262.0 & 249.3 & 193.7 & 263.7 \\ \midrule
+%        Yield & 2.7\% & 2.7\% & 2.3\% & 0.9\% & 0.7\% & - \\ \hline
+        Yield & 7.9\% & 8.1\% & 6.9\% & 2.6\% & 2.7\% & 8.2\% \\ \bottomrule
     \end{tabular}

### 2017-02-06 Paper_v1.tex
Feb6

@@
 This was calculated as the increase in CAT concentration times the CAT carbon number, divided by the sum of the consumption terms for glucose and all amino acids excpet arginine and glutamate, as no data were available for these, weighted by their respective carbon numbers.
-To explain where the remainder of carbon was going, we performed a carbon balance for the best-fit set.
+To explain where the remainder of carbon was going, we performed a carbon balance for the best-fit set (Fig.~\ref{fig:Carbon_Balance}).
 Of the other 92\% of carbon, 35\% accumulated as organic acids (lactate, acetate, succinate and malate) and 9\% accumulated as amino acids (alanine and glutamine).
@@
 coli} extract preparation (Fig.~\ref{fig:oxKO_ssFBA_yields}D).
-The physiological theoretical maximum CAT carbon yield (case i) was 21.8\% $\pm$ 2.8\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, with bounded transcription and translation rates (96\% without glucose contribution in the carbon yield calculation).
-For case ii, the optimal CAT carbon yield decreased to 18.2\% $\pm$ 3.0\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, middle).
-Lastly, when metabolite constraints were applied with experimental measurements of measurements (case iii), the predicted carbon yield was 6.0\% $\pm$ 2.0\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, right).
-%By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield (without arginine and glutamate) of 7.9\%, equivalent to 36\% of the optimal case (i).
-By comparison, the best-fit parameter set for the kinetic model predicted a CAT carbon yield of 7.9\%, equivalent to 36\% of the optimal case (i).
-The experimental dataset had a CAT carbon yield of 8.2\%, similar to both the kinetic model and the experimentally constrained ssFBA calculation (case iii).
-We also investigated the energy efficiency of CAT production for the best-fit set and the three sequence-specific cases, based on the amount of CAT production versus ATP production.
-For cases i and ii the energy efficiencies were 99.8\% and 94.4\%, respectively, while for case iii it was 31.9\%.
+%The physiological theoretical maximum CAT carbon yield (case i) was 21.8\% $\pm$ 2.8\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, with bounded transcription and translation rates (96\% without glucose contribution in the carbon yield calculation).
+%For case ii, the optimal CAT carbon yield decreased to 18.2\% $\pm$ 3.0\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, middle).
+%Lastly, when metabolite constraints were applied with experimental measurements (case iii), the predicted carbon yield was 6.0\% $\pm$ 2.0\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, right).
+The physiological theoretical maximum CAT carbon yield (case i) was 49.3\% $\pm$ 3.5\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, with bounded transcription and translation rates (96\% without glucose contribution in the carbon yield calculation).
+For case ii, the optimal CAT carbon yield decreased to 48.9\% $\pm$ 3.5\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, middle).
+Lastly, when metabolite constraints based on experimental measurements were applied (case iii), the predicted carbon yield was 6.4\% $\pm$ 2.9\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, right).
+Unsurprisingly, this range of carbon yield encompasses both the best-fit set (7.9\%) and the experimental dataset (8.2\%).
+For cases i and ii the energy efficiencies were 72.1\% and 71.2\%, respectively, while for case iii it was only 5.1\%.
 This dramatic decrease in efficiency when fluxes are constrained to data makes sense, as the network is forced toward a multitude of pathways that may not contribute to CAT production.
-However, the case iii efficiency was still much higher than that of the best-fit kinetic model (9.2\%).
-This is likely because of the steady-state assumption and the ability to choose the optimum from a variety of flux distributions; meanwhile, the kinetic model diverts flux toward the accumulation of sub-optimal metabolites.
-Thus, the CFPS reaction was not optimal; the ssFBA calculations suggested that an approximately three-fold increase in carbon yield and an eleven-fold increase in energy efficiency were theoretically possible.
-
-%A similar distribution of the carbon contribution to CAT yield was seen across the best-fit set and all knockouts (Table~\ref{tbl:yield_breakdown}).
-%In all cases, glutamate was responsible for about two-thirds of carbon consumption toward CAT.
-%This was due to its role in generating other amino acids and as a possible energy source for metabolism.
-%The much greater consumption of this one amino acid was made possible by its much larger initial condition, as it was present in the cell media in the form of magnesium glutamate, ammonium glutamate, and potassium glutamate.
-%% [CITE? THESIS NOT CURRENTLY IN BIBLIOGRAPHY].
-%Glucose made the next largest contribution, between 20\% and 30\%, dwarfing all amino acids other than glutamate.
-%This makes sense, as glucose powers the entire metabolism during the first hour, including energy species synthesis and amino acid synthesis.
-%We also investigated the energy efficiency of CAT production across the kinetic models and experimental dataset.
-%The best-fit set, $\Delta$app, and experimental data had energy efficiencies of 2.6\%, $\Delta$nuo had an efficiency of 2.2\%, and $\Delta$cyd and $\Delta$cyd$\Delta$nuo$\Delta$app had efficiencies of 0.8\%.
-%The best-fit set, $\Delta$app, and experimental data had energy efficiencies of 2.6\%, calculated as a ratio of CAT production to glucose consumption, each converted into an equivalent number of ATP molecules.
-%$\Delta$nuo had an efficiency of 2.2\%, and $\Delta$cyd and $\Delta$cyd$\Delta$nuo$\Delta$app had efficiencies of 0.8\%.
-%The energy efficiencies across the different cases were more or less proportional to the carbon yields, and in turn the total amount of CAT synthesis.
-%Taken together, these results show that oxidative phosphorylation, particularly \textit{cyd}, is important to efficient CAT production.
-
-Furthermore, we performed a carbon balance on the best-fit set to determine the proportion of carbon consumption going toward CAT synthesis (Fig.~\ref{fig:Carbon_Balance}).
+However, %the case iii efficiency was still much higher than that of the best-fit kinetic model (9.2\%).
+%This is likely because of the steady-state assumption and the ability to choose the optimum from a variety of flux distributions; meanwhile, the kinetic model diverts flux toward the accumulation of sub-optimal metabolites.
+%Thus, the CFPS reaction was not optimal; the ssFBA calculations suggested that an approximately three-fold increase in carbon yield and an eleven-fold increase in energy efficiency were theoretically possible.
 
@@
 
-To investigate the differences in carbon yields, we compared the flux distributions predicted by ssFBA simulations for the different constraint cases  (Fig.~\ref{fig:Network}).
-All cases heavily utilized the first step in the pentose phosphate pathway to generate NADPH;
-the carbon flux then continued through the Entner–Doudoroff pathway toward pyruvate.
-The majority of the flux proceeded toward acetate accumulation, whereas in case ii, the flux accumulated as lactate and acetate, meanwhile for case iii the flux was distributed between pyruvate, lactate, and acetate.
-In all cases, the energy source was primarily oxidative phosphorylation, and to a lesser extent the TCA cycle.
-However, the accumulation of pyruvate, lacatate, and acetate signifies that the system is not operating at its highest efficiency (case iii).
-The system produced NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to power oxidative phosphorylation.
-Oxidative phosphorylation lead to a high redox ratio contributing to the accumulation of acetate overflow and diverting flux away from the TCA cycle.
-This suggested CAT production could be increased by reducing the accumulation of acetate and lactate.
-To investigate this further, we simulated potential knockouts with constrained transcription/translation rates  (Fig.~\ref{fig:flux_ko}).
-Knocking out the \textit{gnd} reaction, the first step in Entner-Doudoroff pathway, decreased acetate flux by about twenty percent
-In addition, less uptake of amino acids were required which increased the carbon yield of CAT by 1.1\% (up to approximately 22.9\% $\pm$ 2.4\%) compared to the control (no knockouts) for case i.
-The simulation showed an increase in oxidative phosphorylation flux and the flux heavily utilizing glycolysis instead of pentose phosphate pathway.
-A second simulation with both \textit{gnd} and phosphate acetyltransferase knocked out,
-showed very similar results as the first single knockout.
-In the dual knockout, flux towards acetate was almost negligible with some coming from amino acid degradation.
+To investigate %the differences in carbon yields, we compared the flux distributions predicted by ssFBA simulations for the different constraint cases (Fig.~\ref{fig:Network}).
+%All cases heavily utilized the first step in the pentose phosphate pathway to generate NADPH;
+%the carbon flux then continued through the Entner–Doudoroff pathway toward pyruvate.
+%The majority of the flux proceeded toward acetate accumulation, whereas in case ii, the flux accumulated as lactate and acetate, meanwhile for case iii the flux was distributed between pyruvate, lactate, and acetate.
+%In all cases, the energy source was primarily oxidative phosphorylation, and to a lesser extent the TCA cycle.
+%However, the accumulation of pyruvate, lacatate, and acetate signifies that the system is not operating at its highest efficiency (case iii).
+%The system produced NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to power oxidative phosphorylation.
+%Oxidative phosphorylation lead to a high redox ratio contributing to the accumulation of acetate overflow and diverting flux away from the TCA cycle.
+%This suggested CAT production could be increased by reducing the accumulation of acetate and lactate.
+%To investigate this further, we simulated potential knockouts with constrained transcription/translation rates  (Fig.~\ref{fig:flux_ko}).
+%Knocking out the \textit{gnd} reaction, the first step in Entner-Doudoroff pathway, decreased acetate flux by about twenty percent
+%In addition, less uptake of amino acids were required which increased the carbon yield of CAT by 1.1\% (up to approximately 22.9\% $\pm$ 2.4\%) compared to the control (no knockouts) for case i.
+%The simulation showed an increase in oxidative phosphorylation flux and the flux heavily utilizing glycolysis instead of pentose phosphate pathway.
+%A second simulation with both \textit{gnd} and phosphate acetyltransferase knocked out,
+%showed very similar results as the first single knockout.
+%In the dual knockout, flux towards acetate was almost negligible with some coming from amino acid degradation.
 Taken together, a \textit{gnd} knockout decreased acetate production and required less amino acid consumption, thus it is a promising strategy to increase the CAT carbon yield.

### 2017-02-07 Paper_v1.tex
Feb6_2

@@
 The ensemble described all of the training data, especially the central carbon metabolism.
-The model predicted a carbon yield for CAT production that was equal to 36\% of yield for a physiologically realistic case, and an energy efficiency equal to 9\% of the physiologically realistic case, calculated using sequence-specific flux balance analysis.
+The model predicted CAT production with a carbon yield equal to 16\%, and an energy efficiency equal to 9\%, of that of a physiologically realistic case calculated using sequence-specific flux balance analysis.
 This suggests that CAT production could be further optimized.
@@
 Next, to gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield with the maximum theoretical CAT carbon yield calculated using sequence-specific flux balance analysis.
-The CAT yield estimated from the kinetic model was 36\% of the theoretical yield when physiologically realistic constraints were used.
+The CAT yield estimated from the kinetic model was equal to 16\% of the theoretical yield when physiologically realistic constraints were used.
 Taken together, we have integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
@@
 The remaining 48\% went to the net accumulation of all other metabolites, particularly carbon dioxide.
-The best-fit set produced CAT with an energy efficiency of 6.4\%.
+The best-fit set and the experimental dataset both produced CAT with an energy efficiency of 7\%.
 This was calculated as the increase in CAT concentration times the CAT number of equivalent ATP molecules, divided by glucose consumption times times the number of equivalent ATP molecules for glucose, equal to 15 in the optimal case.
-An additional 62\% of energy went to the accumulation of glycolysis metabolites, 6.4\% to TCA cycle metabolites, 0.2\% to pentose phosphate metabolites, and 25\% to acetate and lactate.
+An additional 62\% of energy went to the accumulation of glycolysis metabolites, 6\% to TCA cycle metabolites, and 25\% to acetate and lactate.
 This shows that there is much room for improvement of the efficiency of CFPS.
@@
 The physiological theoretical maximum CAT carbon yield (case i) was 49.3\% $\pm$ 3.5\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, with bounded transcription and translation rates (96\% without glucose contribution in the carbon yield calculation).
-For case ii, the optimal CAT carbon yield decreased to 48.9\% $\pm$ 3.5\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, middle).
+For case ii, the optimal CAT carbon yield was 48.9\% $\pm$ 3.5\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, middle).
 Lastly, when metabolite constraints based on experimental measurements were applied (case iii), the predicted carbon yield was 6.4\% $\pm$ 2.9\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, right).
 Unsurprisingly, this range of carbon yield encompasses both the best-fit set (7.9\%) and the experimental dataset (8.2\%).
-For cases i and ii the energy efficiencies were 72.1\% and 71.2\%, respectively, while for case iii it was only 5.1\%.
+For cases i and ii the energy efficiencies were 72.1\% $\pm$ 9.5\% and 71.2\% $\pm$ 9.6\%, respectively, while for case iii it was only 5.1\% $\pm$ 2.4\%.
 This dramatic decrease in efficiency when fluxes are constrained to data makes sense, as the network is forced toward a multitude of pathways that may not contribute to CAT production.
-However, %the case iii efficiency was still much higher than that of the best-fit kinetic model (9.2\%).
+This range of energy efficiency encompasses both the best-fit set and the experimental dataset (7\%).
+However, the model and the experimental dataset fall short of the optimum (cases i and ii).
+This suggests there is much room for improvement, primarily by reducing byproduct formation.
 %This is likely because of the steady-state assumption and the ability to choose the optimum from a variety of flux distributions; meanwhile, the kinetic model diverts flux toward the accumulation of sub-optimal metabolites.
@@
 
-%To better understand where carbon flux goes besides toward CAT, we examined the proportion of amino acid consumption that was due to CAT synthesis in the best-fit set (Table~\ref{tbl:AA_breakdown}).
-%Eight of the 20 amino acids had a flux-toward-CAT percentage of 100\% or higher.
-%For six of these (isoleucine, leucine, methionine, phenylalanine, tryptophan, valine), synthesis was inactive or negligible, meaning that CAT synthesis was the only reaction that consumed these species.
-%This is why methionine, and to a lesser extent isoleucine, leucine, and valine do not fit the experimental data as well as some other amino acids; consumption via CAT synthesis was not enough to fit the data, and no other reactions exist in the network to consume these species.
-%For histidine and tyrosine, synthesis was active, allowing the percentage to be higher than 100\%.
-%Another ten amino acids registered percentages lower than 100\%, meaning that they were consumed by other pathways in the network.
-%The most extreme of these is glutamate, with a percentage of 0.2\%, because it was consumed by numerous reactions for the synthesis of other amino acids and as an energy substrate.
-%Likewise, arginine, asparagine, glycine, and proline are consumed in the production of other amino acids, while cysteine and lysine are consumed to form other metabolites.
-%ARG R_arg_deg1-4 go to GLU
-%ASN R_asn_deg goes to ASP
-%ASP R_thr, maybe negligible
-%CYS cys_deg goes to pyr
-%GLY R_gar goes to GLU
-%LYS lys_deg goes to cadav
-%PRO pro_deg goes to GLU
-%SER R_glyA, maybe negligible
-%THR R_thr_deg4, maybe negligible
-%Finally, alanine and glutamine increased during the timecourse, meaning that synthesis outweighed all consumption (including consumption toward CAT).
-
-To investigate %the differences in carbon yields, we compared the flux distributions predicted by ssFBA simulations for the different constraint cases (Fig.~\ref{fig:Network}).
-%All cases heavily utilized the first step in the pentose phosphate pathway to generate NADPH;
-%the carbon flux then continued through the Entner–Doudoroff pathway toward pyruvate.
-%The majority of the flux proceeded toward acetate accumulation, whereas in case ii, the flux accumulated as lactate and acetate, meanwhile for case iii the flux was distributed between pyruvate, lactate, and acetate.
-%In all cases, the energy source was primarily oxidative phosphorylation, and to a lesser extent the TCA cycle.
-%However, the accumulation of pyruvate, lacatate, and acetate signifies that the system is not operating at its highest efficiency (case iii).
-%The system produced NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to power oxidative phosphorylation.
-%Oxidative phosphorylation lead to a high redox ratio contributing to the accumulation of acetate overflow and diverting flux away from the TCA cycle.
-%This suggested CAT production could be increased by reducing the accumulation of acetate and lactate.
-%To investigate this further, we simulated potential knockouts with constrained transcription/translation rates  (Fig.~\ref{fig:flux_ko}).
-%Knocking out the \textit{gnd} reaction, the first step in Entner-Doudoroff pathway, decreased acetate flux by about twenty percent
-%In addition, less uptake of amino acids were required which increased the carbon yield of CAT by 1.1\% (up to approximately 22.9\% $\pm$ 2.4\%) compared to the control (no knockouts) for case i.
-%The simulation showed an increase in oxidative phosphorylation flux and the flux heavily utilizing glycolysis instead of pentose phosphate pathway.
-%A second simulation with both \textit{gnd} and phosphate acetyltransferase knocked out,
-%showed very similar results as the first single knockout.
-%In the dual knockout, flux towards acetate was almost negligible with some coming from amino acid degradation.
-Taken together, a \textit{gnd} knockout decreased acetate production and required less amino acid consumption, thus it is a promising strategy to increase the CAT carbon yield.
-
-% To investigate this further, we simulated potential knockouts with constrained transcription/translation rates,
-% and constrained the specific glucose and amino acid rates to the same simulated without knockouts.
-% Knocking out \textit{gnd} decreased acetate production but increased flux through \textit{pntAB}, which is responsible for regenerating NADPH.
-% The simulation showed carbon was diverted toward lactate; however, since CAT production was constrained by the translation rate, we expected no increase in CAT production.
-% A second simulation with \textit{gnd} and phosphate acetyltransferase knockouts showed carbon being diverted toward lactate and succinate;
-% however, it required a higher flux through oxidative phosphorylation and the TCA cycle to meet the energetic needs of CAT production.
-% The decrease in acetate production and the diversion of carbon flux shows a promising mechanism to increase CAT yield.
+To investigate the differences in carbon yield and energy efficiency, we compared the flux distributions predicted by ssFBA simulations for the different constraint cases (Fig.~\ref{fig:NetworkFlux}).
+%All cases heavily utilized the first step in the pentose phosphate pathway to generate NADPH; the carbon flux then continued through the Entner–Doudoroff pathway toward pyruvate.
+In cases i and ii, glycolysis was used for energy generation, and most of the carbon flux accumulated as acetate.
+The system produced NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to power oxidative phosphorylation.
+Case iii heavily utilized the first step in the pentose phosphate pathway to generate NADPH; the carbon flux then continued primarily through the Entner–Doudoroff pathway toward pyruvate.
+Case iii also predicted the accumulation of pyruvate, lacatate, acetate, and carbon dioxide, which contributed to the much lower carbon yield and energy efficiency.
+In all cases, the energy source was primarily oxidative phosphorylation converting cofactors generated throughout the network into ATP.
+%Oxidative phosphorylation lead to a high redox ratio contributing to the accumulation of acetate overflow.
+Taken together, this suggested CAT production could be increased by reducing the accumulation of acetate, lactate, and carbon dioxide.
 
@@
 Having captured the experimental data, we investigated if CAT yield and CFPS performance could be further improved.
-We showed that the model produces CAT with a carbon yield equal to 36\%, and an energy efficiency equal to 9\%, of that of a physiological case in which transcription and translation are constrained.
-The accumulation of waste byproducts, especially acetate, is responsible for this sub-optimal yield.
+We showed that the model produces CAT with a carbon yield equal to 16\% of that of a physiological case in which transcription and translation are constrained, and an energy efficiency equal to 9\% of that of the physiological case.
+The accumulation of waste byproducts, especially acetate and carbon dioxide, is responsible for this sub-optimal performance.
 Sensitivity analysis showed that certain substrates and energy species are instrumental to CAT production and overall metabolism.
 The system heavily relied on oxidative phosphorylation for the system's energetic needs as well as for CAT synthesis.
-A single knockout in oxidative phosphorylation reduced the CAT carbon yield \texttildelow 3-fold, as well as disrupting the system state showing its crucial role in CFPS.
+A single knockout in oxidative phosphorylation reduced the CAT carbon yield \texttildelow 3-fold, as well as disrupting the system state, showing its crucial role in CFPS.
 In comparing flux distributions between low and high yield cases, carbon flux could be potentially diverted toward CAT by reducing acetate overflow and minimizing flux through the Entner-Doudoroff pathway.
@@
 Nevertheless, there are further reactions with considerable impact on model performance.
-In examining oxidative phosphorylation activity, knockouts in the electron transport pathways disrupt metabolism across the network and show CAT carbon yield dropping from 8.6\% to 2.7\%; Jewett and coworkers also saw a decrease in CAT yield, ranging from 1.5-fold to 4-fold, when knocking out oxidative phosphorylation reactions\cite{Jewett:2008aa}.
+In examining oxidative phosphorylation activity, knockouts in the electron transport pathways disrupt metabolism across the network and drop CAT carbon yield from 7.9\% to 2.6\%; Jewett and coworkers also saw a similar decrease in CAT yield with pyruvate as the substrate, ranging from 1.5-fold to 4-fold, when knocking out oxidative phosphorylation reactions \cite{Jewett:2008aa}.
 Oxidative phosphorylation is vital, since it provides most of the energetic needs of CFPS.
-However, it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, and both of our modeling approaches suggest its importance to CAT production and CFPS.
-Thus, oxidative phosphorylation is a potential area for improvement for CFPS performance and protein yield.
+While it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, both of our modeling approaches suggest its importance to improving CFPS performance and protein yield.
 %Also important is the first step in pentose phosphate pathway, as it converts G6P to 6PG and produces NADPH.
@@
 %Many of the insignificant reactions lead to the accumulation of unwanted byproducts such as ethanol and formate.
-%Comparing the theoretical maximum carbon yield of CAT from ssFBA predictions to those of the kinetic model and experimental measurements suggests that there is potential for increasing CAT yield as well as CFPS performance.
-%Yield performance can be potentially improved by diverting carbon flux away from acetate.
-Comparing the physiologically realistic carbon yield of CAT from ssFBA predictions to those of the kinetic model and experimental measurements suggests that there is potential for increasing CAT yield as well as CFPS performance.
-%The model and experimental yields were 36\% of the physiologically constrained case.
-A knockout of \textit{gnd} and shows that carbon can be diverted away from acetate and toward CAT or other proteins of interest expressed in CFPS.
-Another limitation to be addressed in CFPS is the transcription and translation description, since protein production is ultimately bounded by these kinetic rates.
-Li et al.
-have increased productivity of firefly lucifease by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
-Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
 
@@
 Including specific transcription and translation steps for CAT would allow more accurate modeling of the complexity and the resource cost of protein synthesis.
-In addition, sensitivity analysis could be performed on these new parameters to determine the robustness of CAT synthesis to the processes of transcription and translation.
-Another area for future work is to more thoroughly sample parameter space.
-Parameters were varied so as to best fit the dataset; however, the resulting ensemble may not represent every biological possibility.
-In a different region of parameter space, the system may behave differently but still fit the experimental data.
-This could include the flux distribution through the network, the variation of predictions across the ensemble, and the relative sensitivity values.
-Testing the model under a variety of conditions could strengthen or challenge the findings of this study.
-Further experimentation could also be used to gain a deeper understanding of model performance under a variety of conditions.
-Specifically, CAT production performed in the absence of amino acids could inform the system's ability to manufacture them, while experimentation in the absence of glucose or oxygen could shed light on how important they are to protein synthesis, and under which conditions.
+Sensitivity analysis could then be performed on these new parameters to determine the robustness of CAT synthesis to the processes of transcription and translation.
+%Another area for future work is to more thoroughly sample parameter space.
+%Parameters were varied so as to best fit the dataset; however, the resulting ensemble may not represent every biological possibility.
+%In a different region of parameter space, the system may behave differently but still fit the experimental data.
+%This could include the flux distribution through the network, the variation of predictions across the ensemble, and the relative sensitivity values.
+%Testing the model under a variety of conditions could strengthen or challenge the findings of this study.
+In addition, further experimentation could also be used to gain a deeper understanding of model performance under a variety of conditions.
+Specifically, CAT production performed in the absence of amino acids could inform the system's ability to manufacture them, while experimentation in the absence of glucose or oxygen could shed light on the importance of those substrates.
+%how important they are to protein synthesis, and under which conditions.
 Finally, the approach should be extended to other protein products.
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./Figures/CAT_flux_final.pdf}
-\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, and oxidative phosphorylation.
-Sequence-specific FBA flux value (mean ± standard deviation) across ensemble for 1 hr, normalized to glucose uptake flux.
+\includegraphics[width=1.00\textwidth]{./Figures/NetworkFlux.pdf}
+\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, oxidative phosphorylation, and protein synthesis.
+Thicker and redder arrows signify larger flux values (see colorbar).
+A.
+ssFBA constrained by transcription and translation rates, with or without amino acid synthesis reactions.
+B.
+ssFBA constrained by transcription and translation rates and experimental measurements where available.}
+%Sequence-specific FBA flux value (mean ± standard deviation) across ensemble for 1 hr, normalized to glucose uptake flux.
 Flux distribution for three different cases: constrained by transcription and translation rates (top row), same as previous but without amino acid synthesis reactions  (second row), and same as previous but constrained by experimental measurements where available (bottom row).}
-\label{fig:Network}
+\label{fig:NetworkFlux}
 \end{figure}

### 2017-02-07 Paper_v1.tex
Feb7

@@
 
-To understand whether the CFPS performance was optimal, we calculated the carbon yield and energy efficiency of CAT production (Fig.~\ref{fig:?????}).
+To understand whether the CFPS performance was optimal, we calculated the carbon yield and energy efficiency of CAT production (Fig.~\ref{CAT_balances}).
 The best-fit parameter set for the kinetic model predicted a CAT carbon yield of 7.9\%, while the experimental dataset had a CAT carbon yield of 8.2\%.
 This was calculated as the increase in CAT concentration times the CAT carbon number, divided by the sum of the consumption terms for glucose and all amino acids excpet arginine and glutamate, as no data were available for these, weighted by their respective carbon numbers.
-To explain where the remainder of carbon was going, we performed a carbon balance for the best-fit set (Fig.~\ref{fig:Carbon_Balance}).
+To explain where the remainder of carbon was going, we performed a carbon balance for the best-fit set (Fig.~\ref{fig:CAT_balances}A).
 Of the other 92\% of carbon, 35\% accumulated as organic acids (lactate, acetate, succinate and malate) and 9\% accumulated as amino acids (alanine and glutamine).
 The remaining 48\% went to the net accumulation of all other metabolites, particularly carbon dioxide.
-The best-fit set and the experimental dataset both produced CAT with an energy efficiency of 7\%.
+The best-fit set and the experimental dataset both produced CAT with an energy efficiency of 7\% (Fig.~\ref{fig:CAT_balances}B).
 This was calculated as the increase in CAT concentration times the CAT number of equivalent ATP molecules, divided by glucose consumption times times the number of equivalent ATP molecules for glucose, equal to 15 in the optimal case.
-An additional 62\% of energy went to the accumulation of glycolysis metabolites, 6\% to TCA cycle metabolites, and 25\% to acetate and lactate.
+%An additional 62\% of energy went to the accumulation of glycolysis metabolites, 6\% to TCA cycle metabolites, and 25\% to acetate and lactate.
+An additional 62\% of energy went to the accumulation of glycolysis metabolites, and 31\% to organic acids (lactate, acetate, succinate and malate).
 This shows that there is much room for improvement of the efficiency of CFPS.
@@
 $\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed (never allowed to be negative), and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
-Arginine and glutamate were excluded from the yield calculations for the experimental yield calculation, as no experimental measurements were available for these amino acids.
+Arginine and glutamate were excluded from the yield calculations, as no experimental measurements were available for these amino acids.
 Yield of the best-fit parameter set and the experimental data were calculated by setting $\Delta\texttt{CAT}$ equal to the final minus the initial CAT concentration,
@@
 \subsection*{Calculation of energy efficiency.}
-%Energy efficiency was also calculated for best-fit set, knockouts, and experimental data, formulated as:
-%\begin{equation}\label{eqn:energy-efficiency-definition}
-%	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2\cdot\rm {ATP}_{CAT}+\rm {GTP}_{CAT}\right)}{\Delta\texttt{GLC}\cdot \rm {ATP}_{GLC}}
-%\end{equation}
-%where $\rm Efficiency$ denotes the energy efficiency of CAT production, $\rm {ATP}_{CAT}$ denotes the stoichiometric coefficient of ATP in CAT synthesis (multiplied by 2 because AMP, rather than ADP, is a product of CAT synthesis), $\rm {GTP}_{CAT}$ denotes the stoichiometric coefficient of GTP in CAT synthesis, and $\rm {ATP}_{GLC}$ denotes the number of ATP molecules ideally produced from one molecule of glucose.
-%$\rm {ATP}_{CAT}$ = 219, $\rm {GTP}_{CAT}$ = 438, $\rm {ATP}_{GLC}$ = 17.
-
-Energy efficiency was also calculated for the best-fit set:
+Energy efficiency was calculated as the ratio of CAT production to glucose consumption, both in terms of equivalent ATP molecules:
 \begin{equation}\label{eqn:energy-efficiency-definition}
-	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2\cdot\rm {ATP}_{CAT}+\rm {GTP}_{CAT}\right)}{\Delta\texttt{ATP}}
+	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2\left({ATP}_{TX}+\rm {CTP}_{TX}+\rm {GTP}_{TX}+\rm {UTP}_{TX}\right)+\rm2\cdot\rm {ATP}_{TL}+\rm {GTP}_{TL}\right)}{\Delta\texttt{GLC}\cdot \rm {ATP}_{GLC}}
 \end{equation}
-where $\rm Efficiency$ denotes the energy efficiency of CAT production, $\rm {ATP}_{CAT}$ denotes the stoichiometric coefficient of ATP in CAT synthesis (multiplied by 2 because AMP, rather than ADP, is a product of CAT synthesis), $\rm {GTP}_{CAT}$ denotes the stoichiometric coefficient of GTP in CAT synthesis, and $\rm \Delta ATP$ denotes the amount of ATP production in 3 hours, equal to the sum of ATP-producing fluxes integrated over the timecourse.
-$\rm {ATP}_{CAT}$ = 219, $\rm {GTP}_{CAT}$ = 438.
-%, $\rm \Delta CAT$ = 0.018 mM/h, $\rm \Delta ATP$ = 170 mM/h.
-
-%Energy efficiency was also calculated for sequence-specific flux balance analysis, formulated as:
-%\begin{equation}\label{eqn:energy-efficiency-definition}
-%	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2\left({ATP}_{TX}+\rm {CTP}_{TX}+\rm {GTP}_{TX}+\rm {UTP}_{TX}\right)+\rm2\cdot\rm {ATP}_{TL}+\rm {GTP}_{TL}\right)}{\Delta\texttt{GLC}\cdot \rm {ATP}_{GLC}}
-%\end{equation}
-%where $\rm Efficiency$ denotes the energy efficiency of CAT production, $\rm {ATP}_{TX}$, $\rm {CTP}_{TX}$, $\rm {GTP}_{TX}$, $\rm {UTP}_{TX}$ denote the stoichiometric coefficient of each energy species for CAT transcription, $\rm {ATP}_{TL}$ and $\rm {GTP}_{TL}$ denote the stoichiometric coefficient of ATP and GTP, respectively, in CAT translation, and $\rm {ATP}_{GLC}$ denotes the number of ATP molecules produced from one molecule of glucose.
-%$\rm {ATP}_{TX}$ = 176, $\rm {CTP}_{TX}$ = 144, $\rm {GTP}_{TX}$ = 151, $\rm {UTP}_{TX}$ = 189, $\rm {ATP}_{TL}$ = 219, $\rm {GTP}_{TL}$ = 438, $\rm {ATP}_{GLC}$ = 17.
-
-For sequence-specific flux balance analysis, with a more in-depth biological description of CAT synthesis, energy efficiency was calculated slightly differently:
-\begin{equation}\label{eqn:energy-efficiency-definition}
-	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2\left({ATP}_{TX}+\rm {CTP}_{TX}+\rm {GTP}_{TX}+\rm {UTP}_{TX}\right)+\rm2\cdot\rm {ATP}_{TL}+\rm {GTP}_{TL}\right)}{\Delta\texttt{ATP}}
-\end{equation}
-where $\rm {ATP}_{TX}$, $\rm {CTP}_{TX}$, $\rm {GTP}_{TX}$, $\rm {UTP}_{TX}$ denote the stoichiometric coefficients of each energy species for CAT transcription, and $\rm {ATP}_{TL}$, $\rm {GTP}_{TL}$ denote the stoichiometric coefficients of ATP and GTP for CAT translation.
-$\rm {ATP}_{TX}$ = 176, $\rm {CTP}_{TX}$ = 144, $\rm {GTP}_{TX}$ = 151, $\rm {UTP}_{TX}$ = 189, $\rm {ATP}_{TL}$ = 219, $\rm {GTP}_{TL}$ = 438.
+where $\rm {ATP}_{TX}$, $\rm {CTP}_{TX}$, $\rm {GTP}_{TX}$, $\rm {UTP}_{TX}$ denote the stoichiometric coefficients of each energy species for CAT transcription, $\rm {ATP}_{TL}$, $\rm {GTP}_{TL}$ denote the stoichiometric coefficients of ATP and GTP for CAT translation, $\Delta\texttt{GLC}$ denotes the glucose consumption, equal to the initial minus the final glucose concentration, and $\rm {ATP}_{GLC}$ denotes the equivalent ATP number for glucose.
+$\rm {ATP}_{TX}$ = 176, $\rm {CTP}_{TX}$ = 144, $\rm {GTP}_{TX}$ = 151, $\rm {UTP}_{TX}$ = 189, $\rm {ATP}_{TL}$ = 219, $\rm {GTP}_{TL}$ = 438, $\rm {ATP}_{GLC}$ = 15.
 
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./Figures/Carbon-Balance-BFS.pdf}
-\caption{Carbon balance for the best-fit set.
-Carbon moles produced as CAT, amino acids (alanine and glutamine), organic acids (lactate, acetate, succinate, and malate), and other byproducts including carbon dioxide, as percentages of total carbon consumption (glucose and all other amino acids).}
-\label{fig:Carbon_Balance}
+\includegraphics[width=1.00\textwidth]{./Figures/CAT_balances.pdf}
+\caption{Carbon and energy balances for the best-fit set.
+A.
+Carbon moles produced as CAT, amino acids (alanine and glutamine), organic acids (lactate, acetate, succinate, and malate), and other byproducts including carbon dioxide, as percentages of total carbon consumption (glucose and all other amino acids).
+B.
+Energy cost of CAT production, accumulation of organic acids (lactate, acetate, succinate, and malate), and glycolytic and pentose phosphate metabolites, as percentages of total energy utilization from glucose.
+Energy costs calculated in terms of equivalent ATP molecules.}
+\label{fig:CAT_balances}
 \end{figure}

### 2017-02-09 Paper_v1.tex
Feb9

@@
 The ensemble of kinetic CFPS models captured the time evolution of CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
-The cell-free \textit{E.~coli} metabolic network was constructed by removing growth associated reactions from the MG1655 reconstruction \cite{Feist:2007aa}, and by adding reactions describing chloramphenicol acetyltransferase (CAT) biosynthesis, a model protein for which there exists a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
+The cell-free \textit{E.~coli} metabolic network was constructed by removing growth associated reactions from the \textit{i}AF1260 reconstruction of K-12 MG1655 \textit{E.~coli} \cite{Feist:2007aa}, and by adding reactions describing chloramphenicol acetyltransferase (CAT) biosynthesis, a model protein for which there exists a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
 In addition, reactions that were knocked out from the cell extract preparation were removed from the network ($\Delta$speA, $\Delta$tnaA, $\Delta$sdaA, $\Delta$sdaB, $\Delta$gshA, $\Delta$tonA, $\Delta$endA).
@@
 \centering
-    \caption{CAT carbon yield breakdown for best-fit set, knockouts, and experimental data. Carbon produced as CAT, carbon consumed as glucose and each amino acid, sum of consumed species, and yield.}
+    \caption{CAT carbon yield breakdown for best-fit set, knockouts, and experimental data.
+Carbon produced as CAT, carbon consumed as glucose and each amino acid, sum of consumed species, and yield.
+Accumulation of alanine and glutamine (negative consumption terms) was not considered in yield calculation.}
     \renewcommand{\arraystretch}{1.3}
@@
         GLC & 215.4 & 215.4 & 215.4 & 215.4 & 159.8 & 215.4 \\ \hline
-        ALA & 0.0 & 0.0 & 1.7 & 0.0 & 0.0 & 0.0 \\ \hline
+        ALA & -11.6 & -11.4 & 1.7 & -3.8 & -3.2 & -12.1 \\ \hline
 %        ARG & 10.2 & 9.9 & 1.1 & 0.9 & 1.3 & - \\ \hline
@@
         CYS & 3.0 & 3.1 & 3.0 & 2.9 & 2.9 & 3.7 \\ \hline
-        GLN & 0.0 & 0.0 & 0.0 & 1.8 & 2.7 & 0.0 \\ \hline
+        GLN & -11.4 & -11.3 & -4.0 & 1.8 & 2.7 & -11.7 \\ \hline
 %        GLU & 492.6 & 505.6 & 528.2 & 505.1 & 501.8 & - \\ \hline

### 2017-02-16 Paper_v1.tex
Feb16

@@
 Conversely, CAT production was robust to allosteric control, as was most of the network, with the exception of the organic acids in central carbon metabolism.
-This study is the first to model dynamic protein production in \textit{E. coli}, and should provide a foundation for genome-scale,
-dynamic modeling of cell-free \textit{E. coli} protein synthesis.
+This study is the first to model dynamic protein production in \textit{E. coli}, and should provide a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
 
@@
 
-To better understand which parameters and parameter combinations influenced the performance of the kinetic model, we performed sensitivity analysis (Fig.~\ref{fig:Sensitivity}).
+To better understand which parameters and parameter combinations influenced the performance of the kinetic model, we performed sensitivity analysis (Fig.~\ref{fig:Sensitivity-CAT}).
 We perturbed each $V^{max}$ parameter, either individually or in pairwise combinations and measured the change in either CAT production or the overall system state.
-CAT production was most sensitive to the CAT synthesis reaction, oxidative phosphorylation, and the pyruvate-consuming alanine synthesis reaction (Fig.~\ref{fig:Sensitivity}, top, section A).
-%Taking into account these three reactions, and the next 16 most important reactions (section B), we saw a common theme of reactions involving the cofactors ATP, NADH, NADPH, and coenzyme A, as well as the metabolites pyruvate and glutamate.
-We saw a common theme of the most important reactions producing or consuming the cofactors ATP, NADH, NADPH, and coenzyme A, as well as the metabolites pyruvate and glutamate.
-Of the 25 reactions to which CAT production was most sensitive, 9 produced or consumed ATP, making it the most represented in these top reactions (with the exception of hydrogen and phosphate ion).
-The next most represented were pyruvate, glutamate, and ADP with 7 reactions each, followed by coenzyme A, $\alpha$-ketoglutarate, NAD/NADH, and NADP/NADPH, with 6 reactions each.
-This makes sense, as glutamate was an important precursor for the synthesis of other amino acids required by CAT production.
-Meanwhile, the cofactors provided energy to power CAT synthesis, while pyruvate was important for energy generation following glucose depletion.
-In addition, pyruvate was required for the synthesis of several amino acids.
-The pairwise sensitivities (off-diagonal elements) were different from the corresponding first-order sensitivities (diagonal elements), and led to interesting outcomes.
-The combination of certain reactions had a much greater or lesser effect on CAT production than that of the individual reactions by themselves.
-For example, glutamine synthesis and arginine degradation were both among the most important reactions to CAT production (they ranked 5th and 10th, respectively).
-This was likely because they both affected the sensitive glutamine-glutamate balance; glutamine synthesis consumes glutamate, while arginine degradation produces it.
-However, when both were perturbed, their combined effect on the model was low, as the respective contributions to consumption and production of glutamate cancelled.
-An example of positive synergy can be seen in cometabolite interconversion.
-Pyridine nucleotide transhydrogenase catalyzes two reactions: one converts NAD and NADPH into NADH and NADP, while the other does the reverse and also generates a proton gradient.
-Increasing one or the other has little effect on the model, as any increase in reaction flux is hampered by the imbalance created.
-Increasing both, however, allows all cometabolites to balance, so that the reaction fluxes increase unhindered.
-Thus, the pairwise sensitivity is much higher than the sum of the first-order sensitivities.
+The eigen decomposition of the sensitivity shows that CAT synthesis and oxidative phosphorylation are the most important to overall CAT production, followed by the pyruvate-consuming alanine synthesis reaction.
+Among the top 20 reactions, we saw a common theme of the cofactors ATP, NADH, NADPH, and coenzyme A, as well as the metabolites pyruvate, glutamate, and $\alpha$-ketoglutarate.
+ATP appears 7 times in these 20 reactions, while pyruvate, glutamate, and coenzyme A appear 6 times each.
+NADH, NADPH, and $\alpha$-ketoglutarate each appear 5 times.
+This result makes sense, as the high energy cost of protein synthesis means that energy cofactors played a crucial role.
+Also, pyruvate served as the primary substrate after glucose ran out, and pyruvate, glutamate, and $\alpha$-ketoglutarate were all important precursors for the synthesis of amino acids required by CAT production.
+%Meanwhile, the cofactors provided energy to power CAT synthesis, while pyruvate was important for energy generation following glucose depletion.
+We performed the same eigen decomposition on the sensitivity of the overall system state to network reactions (Fig.~\ref{fig:Sensitivity-System}).
+Cytochrome oxidase, part of oxidative phosphorylation, was seen to have the greatest effect on the system state.
+Next most influential was the forward reaction of lactate dehydrogenase, followed by NADH:ubiquinone oxidoreductase, another oxidative phosphorylation reaction.
+The overall system state was also sensitive to cofactors and substrates, specifically NADH and pyruvate.
+%The system was most sensitive to an oxidative phosphorylation reaction (cytochrome oxidase), which converted ubiquinol to ubiquinone while generating a proton gradient.
+%The next 4 most important reactions were all consumers of pyruvate: lactate dehydrogenase, pyruvate formate lyase, alanine synthesis, and PEP synthase.
+Among the top 20 reactions, NADH appears in 8 reactions, pyruvate in 6, coenzyme A in 5, and ATP in 4.
+Glutamate, $\alpha$-ketoglutarate, G3P, and ubiquinone/ubiquinol appear 3 times each.
+Taken together, sensitivity analysis identified that substrates and energy cofactors, specifically those around oxidative phosphorylation, most influenced model performance.
 
-The overall system state was also sensitive to cofactors and substrates; however, instead of pyruvate and glutamate, the substrates driving metabolism were pyruvate and G3P.
-The system was most sensitive to an oxidative phosphorylation reaction (cytochrome oxidase), which converted ubiquinol to ubiquinone while generating a proton gradient.
-The next 4 most important reactions were all consumers of pyruvate: lactate dehydrogenase, pyruvate formate lyase, alanine synthesis, and PEP synthase.
-Of the 25 reactions to which the system was most sensitive, 7 produced or consumed pyruvate and 7 participated in NAD/NADH exchange.
-With the exception of hydrogen, these were the most represented species in the top 25 reactions.
-Also important were coenzyme A (5 reactions) and acetyl coenzyme A (4 reactions), as well as ubiquinone/ubiquinol, NADP/NADPH, G3P, and ATP  (4 reactions each).
-%We saw a common theme of the most important reactions producing or consuming the cofactors ATP, NADH, NADPH, ubiquinone/ubiquinol, and coenzyme A, as well as the metabolites pyruvate and G3P.
-%These two and the next 30 most important reactions (section G) largely involved cofactors, especially ATP and NADPH, as well as substrate-consuming reactions and oxidative phosphorylation.
-The system state also had pairwise sensitivities that differed from the corresponding first-order sensitivities and stood out as significant.
-For example, alanine degradation was among the most important reactions, as it produced pyruvate; GMP synthesis was also moderately important, as it produced glutamate.
-However, when both reactions were increased, the combined effect on the model was almost zero.
-%This can be understood by also considering a third reaction: alanine synthesis from pyruvate and glutamate.
-This can be understood by considering the reactions that involve both pyruvate and glutamate: they all either produced both of these substrates or consumed both.
-When alanine degradation was increased, the excess pyruvate stimulated these reactions to consume both pyruvate and glutamate; the amounts of both of these substrates could not be conserved.
-But when GMP synthesis was perturbed as well, the glutamate deficiency was corrected, cancelling much of the effect on the system.
-One of the pyruvate- and glutamate-consuming reactions was alanine synthesis; in this case, the levels of pyruvate, glutamate, and alanine were all virtually unchanged.
-%An example of positive synergy can be seen in the interplay of amino acid synthesis and the TCA cycle: aspartate synthesis and the reverse reaction of succinyl coenzyme A synthetase are both relatively unimportant to the system.
-An example of positive synergy can be seen in histidine synthesis, one of the least influential reactions that consumes three units of ATP.
-When perturbed in combination with the reverse reaction of succinyl coenzyme A synthetase, another reaction with little overall effect on the model, the combined effect on the system is much greater.
-This may be because the reverse reaction of succinyl coenzyme A synthetase produced ATP, which further stimulated histidine synthesis.
-%Another interesting result was the intersection of sections F and G with section J.
-%The 53 reactions in section J were turned off in the best-fit set ($V^{max}=0$);
-%therefore, the perturbation of these reactions had no effect on the model.
-%As a result, all pairwise sensitivities with reactions in section J were pseudo first-order sensitivities for the other reactions.
-%Interestingly, many reactions in section F and several in section G showed their highest sensitivities when paired with the "non-effects" of section J.
-%\textcolor{red}{NICK: DOES THIS MAKE SENSE? WHY ARE THESE DIFFERENT THAN THE SINGLE PERTURBATIONS?}
-%Of these, three involved pyruvate, strengthening its role as a key metabolite; the others were glucose consumption via GTP/CTP-specific hexokinases,
-%fumarate reductase, and SO$_4$ utilization.
-%This suggested that these reactions' effects on the model were canceled out or lessened by most other reactions, but were of course not affected by the reactions in section J.
-%This was also likely the reason that section J ranked above section K, despite having no effect on the model themselves.
-Taken together, sensitivity analysis identified blocks of parameters that either individually, or in combination influenced model performance.
-
-To understand whether the CFPS performance was optimal, we calculated the carbon yield and energy efficiency of CAT production (Fig.~\ref{CAT_balances}).
+To understand whether the CFPS performance was optimal, we calculated the carbon yield and energy efficiency of CAT production (Fig.~\ref{fig:CAT_balances}).
 The best-fit parameter set for the kinetic model predicted a CAT carbon yield of 7.9\%, while the experimental dataset had a CAT carbon yield of 8.2\%.
@@
 
-% This array of sensitivity indices was also organized into a clustergram (Fig.~\ref{fig:Sensitivity}, bottom).
+% This array of sensitivity indices was also organized into a clustergram (Fig.~\ref{fig:Sensitivity-CAT}, bottom).
 % Theoretical maximum CAT carbon yields for three cases discussed below were calculated using flux balance anaylsis (FBA) with a sequence-specific description of CAT synthesis, where $\Delta m_{i}$ denotes the flux of the $i^{th}$ species.
@@
 \centering
-\includegraphics[width=0.76\textwidth,trim=0 28 615 40,clip]{./Figures/Sensitivity.pdf}
-\caption{Normalized first-order and pairwise sensitivities of CAT production (top) and system state (bottom) to maximum reaction rates.}
-\label{fig:Sensitivity}
+\includegraphics[width=0.95\textwidth]{./Figures/Sensitivity-CAT.pdf}
+\caption{Sensitivity of CAT production to model parameters, decomposed into eigenmodes.
+Sensitivity of top 10 eigenvectors  (columns, left to right) to top 20 most influential parameters (rows, bottom to top).
+Relative magnitudes of eigenvalues expressed as percentages.}
+\label{fig:Sensitivity-CAT}
 \end{figure}
-
-%\begin{figure}[ht]
-%\centering
-%\includegraphics[width=1\textwidth]{./Figures/oxphos_ko.pdf}
-%\caption{The effect of oxidative phosphorylation on glucose uptake, CAT production and CAT carbon yield.
-A.
-95\% confidence interval of an ensemble for glucose concentration versus time for no knockouts (blue shaded region), \textit{cyd} knockout (green shaded region), and \textit{nuo} knockout (red shaded region).
-B.
-95\% confidence interval of an ensemble for CAT concentration versus time for no knockouts (blue shaded region), \textit{cyd} knockout (green shaded region), and \textit{nuo} knockout (red shaded region).
-C.
-CAT carbon yield for 5 different cases of oxidative phosphorylation: no knockouts (blue), \textit{cyd} knockout (green), \textit{nuo} knockout (red), \textit{app} knockout (light gray), and a combination of \textit{cyd}, \textit{nuo}, \textit{app} knockouts (dark gray).}
-%\label{fig:oxphos_ko}
-%\end{figure}
-%\clearpage
-
-%\begin{figure}[ht]
-%\centering
-%\includegraphics[width=1\textwidth]{./Figures/CAT_prod_yield.pdf}
-%\caption{Sequence-specific flux balance analysis of CAT production and yield.
-A.
-95\% confidence interval of the ensemble (light blue region) for CAT concentration versus time.
-B.
-Theoretical maximum carbon yield of CAT calcualted by ssFBA for three different cases: constrained by transcription/translation (TX/TL) rates (black), same as previous but without amino acid synthesis reactions (gray), and same as previous but constrained by experimental measurements where available (light gray).
-Error bars represent standard deviation of the ensemble.}
-%\label{fig:CATProdYield}
-%\end{figure}
-%\clearpage
 
@@
 
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.9\textwidth]{./Figures/Sensitivity-System.pdf}
+\caption{Sensitivity of system state to model parameters, decomposed into eigenmodes.
+Sensitivity of top 10 eigenvectors  (columns, left to right) to top 20 most influential parameters (rows, bottom to top).
+Relative magnitudes of eigenvalues expressed as percentages.}
+\label{fig:Sensitivity-System}
+\end{figure}
+
 \end{document}

### 2017-04-18 Paper_v1.tex
Apr14
Apr18

@@
 
-The ensemble of kinetic CFPS models captured the time evolution of CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
+The ensemble of kinetic CFPS models captured the time evolution of CAT biosynthesis (Fig.~\ref{fig:Carbon} - \ref{fig:Amino}).
 The cell-free \textit{E.~coli} metabolic network was constructed by removing growth associated reactions from the \textit{i}AF1260 reconstruction of K-12 MG1655 \textit{E.~coli} \cite{Feist:2007aa}, and by adding reactions describing chloramphenicol acetyltransferase (CAT) biosynthesis, a model protein for which there exists a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
@@
 The CFPS model equations were formulated using the hybrid cell-free modeling framework of Wayman et al. \cite{pr3010138}.
-An ensemble of model parameters (N $>$ 10,000) was estimated from measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids using a constrained Markov Chain Monte Carlo (MCMC) approach.
+An initial ensemble of model parameter sets (N $>$ 5,000) was estimated from measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids using a constrained Markov Chain Monte Carlo (MCMC) approach.
 The MCMC algorithm minimized the error between the training data and model simulations starting from an initial parameter set assembled from literature and inspection.
-Parameter sets were selected for the ensemble based upon their error, and the Pearson correlation coefficient between the candidate and existing sets in the ensemble.
+Parameter sets were selected for the final ensemble (N $=$ 1,000) based upon their error and the Pearson correlation coefficient between the candidate and the existing sets in the ensemble; thus, an accurate yet diverse ensemble was created.
 The parameter set with the lowest error value was defined as the best-fit set.
-Central carbon metabolism (Fig.~\ref{fig:CarbonBoth}, top), energy species (Fig.~\ref{fig:Energy}), and amino acids (Fig.~\ref{fig:Amino}) were captured by the ensemble and the best-fit set.
-The constrained MCMC approach estimated parameter sets with a median error greater than two-order of magnitude less than random parameter sets generated within
-the same parameter bounds (Fig. \ref{fig:BoxPlot}); thus, we have confidence in the predictive capability of the estimated parameters.
+Central carbon metabolism (Fig.~\ref{fig:Carbon}, top), energy species (Fig.~\ref{fig:Energy}), and amino acids (Fig.~\ref{fig:Amino}) were captured by the ensemble and the best-fit set.
+The constrained MCMC approach estimated parameter sets with a median error greater than two-order of magnitude less than random parameter sets generated within the same parameter bounds (Fig.~\ref{fig:BoxPlot}); thus, we have confidence in the predictive capability of the estimated parameters.
 The model captured the biphasic CAT production: during the first hour glucose powers production, and CAT is produced at \texttildelow 10 $\mu$M/h; subsequently, pyruvate and lactate reserves are consumed to power metabolism, and CAT is produced less efficiently at \texttildelow 5 $\mu$M/h.
-Allosteric control was important to biphasic CAT production; without control, the CAT production rate increased and then ceased after 1.5 hr (Fig.~\ref{fig:CarbonBoth}, bottom).
-In addition, acetate no longer accumulated after 1.5 hours, in the absence of allosteric control.
-Interestingly, the simulated malate abundance tracked the experimental measurements during the glucose consumption phase, but increased sharply during the pyruvate consumption phase,
-without allosteric control. Taken together, we produced an ensemble of kinetic models that was consistent with time series measurements of the production of a model protein.
-However, while the ensemble described the experimental data, it was unclear which kinetic parameters most influenced CAT production, and whether the performance of the CFPS reaction was optimal.
+%Allosteric control was important to biphasic CAT production; without control, the CAT production rate increased and then ceased after 1.5 h (Fig.~\ref{fig:Carbon}, bottom).
+%In addition, acetate no longer accumulated after 1.5 hours, in the absence of allosteric control.
+%Interestingly, the simulated malate abundance tracked the experimental measurements during the glucose consumption phase, but increased sharply during the pyruvate consumption phase, without allosteric control.
+Allosteric control was important to central carbon metabolism, especially pyruvate and succinate, and to a lesser extent acetate and malate (Fig.~\ref{fig:Carbon}, bottom).
+Taken together, we produced an ensemble of kinetic models that was consistent with time series measurements of the production of a model protein.
+Although the ensemble described the experimental data, it was unclear which kinetic parameters most influenced CAT production, and whether the performance of the CFPS reaction was optimal.
+
+To better understand the effect of network reactions on system performance we conducted a group knockout analysis (Fig.~\ref{fig:GroupKO}).
+The network was divided into 19 groups of reactions, spanning central carbon metabolism, energetics, and amino aicd biosynthesis.
+The reactions in each of these groups were turned off, and the resulting change in cost function and system state were recorded.
+Then each pair of groups was knocked out to determine pairwise effects.
+These were summed with the first-order effect to obtain a total-order coefficient for each group for the change in cost function and system state.
+
+The two largest total-order effects on the cost function are cysteine/methionine biosynthesis and anaplerotic/glyoxylate reactions.
+This is likely because...
+The effect of knocking out both of these reaction groups is understandably very great.
+However, the largest pairwise effect is seen when knocking out the Entner-Doudoroff pathway and the TCA cycle, two of the reaction groups with the smallest total-order effects.
+This is true of both the cost function knockout array and the system state knockout array.
+This may be because...
 
@@
 All sets with a lower cost (and some with higher cost) were accepted into the ensemble.
-After generating greater that 10,000 sets, we selected N = 100 sets with minimal set to set correlation to avoid over-sampling any region of parameter space.
-% The original 12,437-set ensemble had a mean Pearson correlation coefficient of 0.94 between pairs of sets, while the 100-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.
+After generating XXX sets, we selected N = 1000 sets with minimal set to set correlation to avoid over-sampling any region of parameter space.
+% The original 12,437-set ensemble had a mean Pearson correlation coefficient of 0.94 between pairs of sets, while the 1000-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.
 
@@
 \subsubsection*{Quantification of uncertainty.}
-An ensemble of 100 sets of flux distributions was calculated for three different cases:  constrained by transcription/translation rates, constrained by transcription/translation rates without amino acid synthesis reactions, and constrained by transcription/translation rates and experimental measurements without amino acid synthesis reactions.
+An ensemble of 1000 sets of flux distributions was calculated for three different cases:  constrained by transcription/translation rates, constrained by transcription/translation rates without amino acid synthesis reactions, and constrained by transcription/translation rates and experimental measurements without amino acid synthesis reactions.
 For the first case, all rates were left unbounded, except the specific glucose uptake rate, transcription and translation rate.
@@
 % \centering
-% \includegraphics[width=1.00\textwidth]{./Figures/CarbonBoth.pdf}
+% \includegraphics[width=1.00\textwidth]{./Figures/Carbon.pdf}
 % \caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species.
 Best-fit parameter set (orange line) versus experimental data (points).
-95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
-% \label{fig:CarbonBoth}
+95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 1000 sets.}
+% \label{fig:Carbon}
 % \end{figure}
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/allostericControl_noindex.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/Carbon.pdf}
 \caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species.
 Best-fit parameter set (orange line) versus experimental data (points).
-95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
-\label{fig:CarbonBoth}
+95\% confidence interval (blue or gray shaded region) over the ensemble of 1000 sets.}
+\label{fig:Carbon}
 \end{figure}
@@
 Best-fit parameter set (orange line) versus experimental data (points).
-95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
+95\% confidence interval (blue shaded region) over the ensemble of 1000 sets.}
 \label{fig:Energy}
@@
 Best-fit parameter set (orange line) versus experimental data (points).
-95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
+95\% confidence interval (blue shaded region) over the ensemble of 1000 sets.}
 \label{fig:Amino}
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/BoxPlot-KCFM.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/CostArray-BoxPlot.pdf}
 \caption{Log of cost function across 37 datasets for data-trained ensemble (blue) and randomly generated ensemble (red, gray background).
@@
 \label{fig:BoxPlot}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.75\textwidth]{./Figures/GroupKO_vert.pdf}
+\caption{Effect of group knockouts on system.
+A.
+Log change in cost function when one (diagonal) or two (off-diagonal) reaction groups are turned off.
+B.
+Log change in system state (only species for which data exist) when one (diagonal) or two (off-diagonal) reaction groups are turned off.
+Total-order effect  for each group calculated as the sum of first-order effect and all pairwise effects.
+Larger and darker circles represent a greater effect.}
+\label{fig:GroupKO}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/GroupKO_horiz.pdf}
+\caption{Effect of group knockouts on system.
+A.
+Log change in cost function when one (diagonal) or two (off-diagonal) reaction groups are turned off.
+B.
+Log change in system state (only species for which data exist) when one (diagonal) or two (off-diagonal) reaction groups are turned off.
+Total-order effect  for each group calculated as the sum of first-order effect and all pairwise effects.
+Larger and darker circles represent a greater effect.}
+\label{fig:GroupKO_}
 \end{figure}
@@
 \centering
-\includegraphics[width=1\textwidth]{./Figures/oxKO_ssFBA_yields.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/oxKO_ssFBA_yields.pdf}
 \caption{The effects of oxidative phosphorylation and amino acid synthesis pathways on CAT production and carbon yield.

### 2017-05-02 Paper_v1.tex
Apr20
May2

@@
 All sets with a lower cost (and some with higher cost) were accepted into the ensemble.
-After generating XXX sets, we selected N = 1000 sets with minimal set to set correlation to avoid over-sampling any region of parameter space.
-% The original 12,437-set ensemble had a mean Pearson correlation coefficient of 0.94 between pairs of sets, while the 1000-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.
+After generating XXX sets, we selected N = 100 sets with minimal set to set correlation to avoid over-sampling any region of parameter space.
+% The original 12,437-set ensemble had a mean Pearson correlation coefficient of 0.94 between pairs of sets, while the 100-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.
 
@@
 \subsubsection*{Quantification of uncertainty.}
-An ensemble of 1000 sets of flux distributions was calculated for three different cases:  constrained by transcription/translation rates, constrained by transcription/translation rates without amino acid synthesis reactions, and constrained by transcription/translation rates and experimental measurements without amino acid synthesis reactions.
+An ensemble of 100 sets of flux distributions was calculated for three different cases:  constrained by transcription/translation rates, constrained by transcription/translation rates without amino acid synthesis reactions, and constrained by transcription/translation rates and experimental measurements without amino acid synthesis reactions.
 For the first case, all rates were left unbounded, except the specific glucose uptake rate, transcription and translation rate.
@@
 Best-fit parameter set (orange line) versus experimental data (points).
-95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 1000 sets.}
+95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
 % \label{fig:Carbon}
@@
 Best-fit parameter set (orange line) versus experimental data (points).
-95\% confidence interval (blue or gray shaded region) over the ensemble of 1000 sets.}
+95\% confidence interval (blue or gray shaded region) over the ensemble of 100 sets.}
 \label{fig:Carbon}
@@
 Best-fit parameter set (orange line) versus experimental data (points).
-95\% confidence interval (blue shaded region) over the ensemble of 1000 sets.}
+95\% confidence interval (blue shaded region) over the ensemble of 100 sets.}
 \label{fig:Energy}
@@
 Best-fit parameter set (orange line) versus experimental data (points).
-95\% confidence interval (blue shaded region) over the ensemble of 1000 sets.}
+95\% confidence interval (blue shaded region) over the ensemble of 100 sets.}
 \label{fig:Amino}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{./Figures/GroupKO_vert.pdf}
+\includegraphics[width=0.7\textwidth]{./Figures/GroupKO.pdf}
 \caption{Effect of group knockouts on system.
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/GroupKO_horiz.pdf}
-\caption{Effect of group knockouts on system.
-A.
-Log change in cost function when one (diagonal) or two (off-diagonal) reaction groups are turned off.
-B.
-Log change in system state (only species for which data exist) when one (diagonal) or two (off-diagonal) reaction groups are turned off.
-Total-order effect  for each group calculated as the sum of first-order effect and all pairwise effects.
-Larger and darker circles represent a greater effect.}
-\label{fig:GroupKO_}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=0.95\textwidth]{./Figures/Sensitivity-CAT.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/Sens_productivity.pdf}
 \caption{Sensitivity of CAT production to model parameters, decomposed into eigenmodes.
@@
 \centering
-\includegraphics[width=0.9\textwidth]{./Figures/Sensitivity-System.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/Sens_state.pdf}
 \caption{Sensitivity of system state to model parameters, decomposed into eigenmodes.

### 2017-05-04 Paper_v1.tex
May4

@@
 The CFPS model equations were formulated using the hybrid cell-free modeling framework of Wayman et al. \cite{pr3010138}.
-An initial ensemble of model parameter sets (N $>$ 5,000) was estimated from measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids using a constrained Markov Chain Monte Carlo (MCMC) approach.
+An initial ensemble of model parameter sets (N $\approx$ 30,000) was estimated from measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids using a constrained Markov Chain Monte Carlo (MCMC) approach.
 The MCMC algorithm minimized the error between the training data and model simulations starting from an initial parameter set assembled from literature and inspection.
-Parameter sets were selected for the final ensemble (N $=$ 1,000) based upon their error and the Pearson correlation coefficient between the candidate and the existing sets in the ensemble; thus, an accurate yet diverse ensemble was created.
-The parameter set with the lowest error value was defined as the best-fit set.
+A final ensemble of parameter sets (N $=$ 100) was constructed by selecting the sets with the lowest errors, the lowest of which was defined as the best-fit set.
+Parameter sets in the final ensemble had an average Pearson correlation coefficient of 0.77; thus, an accurate yet diverse ensemble was created.
+%had a median Pearson correlation coefficient of 0.79;
 Central carbon metabolism (Fig.~\ref{fig:Carbon}, top), energy species (Fig.~\ref{fig:Energy}), and amino acids (Fig.~\ref{fig:Amino}) were captured by the ensemble and the best-fit set.
-The constrained MCMC approach estimated parameter sets with a median error greater than two-order of magnitude less than random parameter sets generated within the same parameter bounds (Fig.~\ref{fig:BoxPlot}); thus, we have confidence in the predictive capability of the estimated parameters.
+The constrained MCMC approach estimated parameter sets with a median error an order of magnitude less than random parameter sets generated within the same parameter bounds (Fig.~\ref{fig:BoxPlot}); thus, we have confidence in the predictive capability of the estimated parameters.
 The model captured the biphasic CAT production: during the first hour glucose powers production, and CAT is produced at \texttildelow 10 $\mu$M/h; subsequently, pyruvate and lactate reserves are consumed to power metabolism, and CAT is produced less efficiently at \texttildelow 5 $\mu$M/h.
@@
 %Interestingly, the simulated malate abundance tracked the experimental measurements during the glucose consumption phase, but increased sharply during the pyruvate consumption phase, without allosteric control.
-Allosteric control was important to central carbon metabolism, especially pyruvate and succinate, and to a lesser extent acetate and malate (Fig.~\ref{fig:Carbon}, bottom).
+Allosteric control was important to central carbon metabolism, especially pyruvate, acetate, and succinate (Fig.~\ref{fig:Carbon}, bottom).
+The difference between the allosteric control and no-control cases is mostly seen in the second phase of CAT production, after glucose is exhausted.
 Taken together, we produced an ensemble of kinetic models that was consistent with time series measurements of the production of a model protein.
@@
 To better understand the effect of network reactions on system performance we conducted a group knockout analysis (Fig.~\ref{fig:GroupKO}).
-The network was divided into 19 groups of reactions, spanning central carbon metabolism, energetics, and amino aicd biosynthesis.
-The reactions in each of these groups were turned off, and the resulting change in cost function and system state were recorded.
+The network was divided into 19 groups of reactions, spanning central carbon metabolism, energetics, and amino acid biosynthesis.
+The reactions in each of these groups were turned off, and the resulting change in productivity and system state were recorded.
 Then each pair of groups was knocked out to determine pairwise effects.
-These were summed with the first-order effect to obtain a total-order coefficient for each group for the change in cost function and system state.
+These were summed with the first-order effect to obtain a total-order coefficient for each group for the change in productivity and system state.
+Glycolysis/gluconeogenesis and oxidative phosphorylation were seen to have the greatest effect on both productivity and system state.
+This is explained by their role in both central carbon metabolism and energy generation.
+In addition, CAT productivity is affected by two sectors of amino acid biosynthesis: alanine/aspartate/asparagine, and glutamate/glutamine.
+This is likely because aspartate, glutamate, and glutamine are key reactants in the biosynthesis of many other amino acids, all of which are required for CAT synthesis.
+%these amino acids are key reactants
+Meanwhile, the TCA cycle and the overflow metabolism, which includes acetyl-coA/acetate reactions
 
@@
 A.
-Log change in cost function when one (diagonal) or two (off-diagonal) reaction groups are turned off.
+Log change in CAT productivity when one (diagonal) or two (off-diagonal) reaction groups are turned off.
 B.
@@
 Total-order effect for each group calculated as the sum of first-order effect and all pairwise effects.
-Larger and darker circles represent a greater effect.}
+Larger and darker circles represent greater effects.}
 \label{fig:GroupKO}

### 2017-05-04 Paper_v1.tex
May4_2

@@
 A.
-Log change in CAT productivity when one (diagonal) or two (off-diagonal) reaction groups are turned off.
+Change in CAT productivity when one (diagonal) or two (off-diagonal) reaction groups are turned off.
 B.
-Log change in system state (only species for which data exist) when one (diagonal) or two (off-diagonal) reaction groups are turned off.
+Change in system state (only species for which data exist) when one (diagonal) or two (off-diagonal) reaction groups are turned off.
 Total-order effect for each group calculated as the sum of first-order effect and all pairwise effects.

### 2017-05-09 Paper_v1.tex
Apr9

@@
 \hyphenation{dec-ades} %Latex does not know how to hyphenate this by default, so we add it to the list
+\def\wl{\par \vspace{\baselineskip}}
 
@@
 The ensemble described all of the training data, especially the central carbon metabolism.
-The model predicted CAT production with a carbon yield equal to 16\%, and an energy efficiency equal to 9\%, of that of a physiologically realistic case calculated using sequence-specific flux balance analysis.
+XXX Talk about carbon and energy balances XXX
 This suggests that CAT production could be further optimized.
@@
 CAT production and other metabolites, specifically organic acid intermediates such as pyruvate, were sensitive to the presence of allosteric control mechanisms.
-Next, to gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield with the maximum theoretical CAT carbon yield calculated using sequence-specific flux balance analysis.
-The CAT yield estimated from the kinetic model was equal to 16\% of the theoretical yield when physiologically realistic constraints were used.
 Taken together, we have integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
 This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
-
-% In this study, we present an effective biochemical network modeling framework for building dynamic cell-free metabolic models.
-% The key innovation of our approach is the seamless integration of simple effective rules encoding complex regulation with traditional kinetic pathway modeling.
-% This integration allows the description of complex regulatory interactions in the absence of specific mechanistic information.
-% The regulatory rules are easy to understand, easy to formulate and do not rely on overarching theoretical abstractions or restrictive assumptions.
-% Using this approach, we have built an ensemble of models that predict an experimental dataset for CAT production in a cell-free \textit{E. coli} system.
-% We calculated the CAT yield for the ensemble at 45\% of the theoretical maximum according to constraint-based modeling, and determined through sensitivity analysis that CAT production was most sensitive to the parameters associated with CAT synthesis and the synthesis of GTP, GMP, and amino acids.
-% While only an initial proof-of-concept, the framework presented here could be an important first step toward genome-scale cell-free kinetic modeling of the biosynthetic capacity of industrially important organisms.
-% In the second case (ii), there was a high flux through fumarate dehydrogenase from aspartic acid uptake, whereas in cases i and iii, acetate and lactate accumulation occurred.
-% However, oxidative phosphorylation is not operating at its highest efficiency, since there is a mixture of acetate and lactate accumulation.
-% In addition, a handful of alanine synthesis pairwise sensitivities were also large.
 
@@
 This shows that there is much room for improvement of the efficiency of CFPS.
-Gene knockouts in the electron transport chain further reduced the performance of the CFPS reaction (Fig.~\ref{fig:oxKO_ssFBA_yields}).
 A key finding of both the CAT and overall system state sensitivity analysis was the importance of oxidative phosphorylation.
-To investigate this further, we knocked out key oxidative phosphorylation reactions in the ensemble of kinetic models to examine the effect on CAT production and carbon yield.
-A single \textit{cyd} knockout reduced the CAT carbon yield from 7.9\% to 2.6\% (Table~\ref{tbl:yield_breakdown}).
-On the other hand, a \textit{nuo} knockout showed a less dramatic decrease in yield,
-reducing the CAT carbon yield to 6.9\%.
-Knocking out \textit{app} increased CAT yield to 8.1\%, but this increase was not statistically significantly different from that of the control.
-Lastly, knocking out all three reactions reduced the CAT yield to 0.7\%, similar to knocking out the \textit{cyd} alone.
-Thus, the model suggested the key step in oxidative phosphorylation was catalyzed by the gene product of \textit{cyd}.
-However, while disruption of \textit{cyd} significantly reduced the CAT carbon yield, it did not completely eliminate the production of CAT.
-This suggested there was a mixture of energy sources supporting CAT production, with the most significant being oxidative phosphorylation.
-%The \textit{cyd} enzyme is responsible for transferring electrons with oxygen as an electron acceptor allowing the release of energy to form ATP.
-
-Sequence-specific flux balance analysis (ssFBA) predicted optimal CAT yields with no adjustable parameters (Fig.~\ref{fig:oxKO_ssFBA_yields}).
-Before exploring CFPS optimality, we first validated the ssFBA approach by comparing simulated and measured concentrations of CAT for the first hour of glucose consumption.
-We chose this time window (during the first phase of CAT production) because it was approximately linear both in glucose consumption and in the accumulation of most organic acids.
-As the ssFBA calculation had no adjustable parameters, bounds on transcription and translation rates and biochemical fluxes were either estimated from data or from mechanistic models parameterized from literature.
-Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates, or the upper bounds for oxygen and glucose consumption rates was addressed
-by sampling plausible ranges for these parameters.
-The ensemble of ssFBA simulations predicted CAT formation as a function of time during the first hour of production when constrained by the experimental metabolite data (Fig.~\ref{fig:oxKO_ssFBA_yields}C).
-Thus, the molecular description of transcription and translation were consistent with experimental measurements.
-Next, to gauge the performance of the CFPS reaction, we next calculated the CAT carbon yield for three classes of constraints:
-(i) theoretical maximum glucose, amino acid and oxygen upper bounds, and realistic transcriptional/translational constraints;
-(ii) theoretical maximum glucose, amino acid and oxygen upper bounds, realistic transcriptional/translational constraints and knockouts of amino acid synthesis reactions of amino acids supplemented in the \textit{E.
-coli} extract preparation.
-(iii) metabolite fluxes constrained by the CAT data, and realistic transcriptional/translational constraints and knockouts of amino acid synthesis reactions of amino acids supplemented in the \textit{E.
-coli} extract preparation (Fig.~\ref{fig:oxKO_ssFBA_yields}D).
-%The physiological theoretical maximum CAT carbon yield (case i) was 21.8\% $\pm$ 2.8\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, with bounded transcription and translation rates (96\% without glucose contribution in the carbon yield calculation).
-%For case ii, the optimal CAT carbon yield decreased to 18.2\% $\pm$ 3.0\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, middle).
-%Lastly, when metabolite constraints were applied with experimental measurements (case iii), the predicted carbon yield was 6.0\% $\pm$ 2.0\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, right).
-The physiological theoretical maximum CAT carbon yield (case i) was 49.3\% $\pm$ 3.5\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, left); this represents optimal network performance if glucose, oxygen and amino acids were produced or consumed at their upper bounds, with bounded transcription and translation rates (96\% without glucose contribution in the carbon yield calculation).
-For case ii, the optimal CAT carbon yield was 48.9\% $\pm$ 3.5\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, middle).
-Lastly, when metabolite constraints based on experimental measurements were applied (case iii), the predicted carbon yield was 6.4\% $\pm$ 2.9\% (Fig.~\ref{fig:oxKO_ssFBA_yields}D, right).
-Unsurprisingly, this range of carbon yield encompasses both the best-fit set (7.9\%) and the experimental dataset (8.2\%).
-For cases i and ii the energy efficiencies were 72.1\% $\pm$ 9.5\% and 71.2\% $\pm$ 9.6\%, respectively, while for case iii it was only 5.1\% $\pm$ 2.4\%.
-This dramatic decrease in efficiency when fluxes are constrained to data makes sense, as the network is forced toward a multitude of pathways that may not contribute to CAT production.
-This range of energy efficiency encompasses both the best-fit set and the experimental dataset (7\%).
-However, the model and the experimental dataset fall short of the optimum (cases i and ii).
-This suggests there is much room for improvement, primarily by reducing byproduct formation.
-%This is likely because of the steady-state assumption and the ability to choose the optimum from a variety of flux distributions; meanwhile, the kinetic model diverts flux toward the accumulation of sub-optimal metabolites.
-%Thus, the CFPS reaction was not optimal; the ssFBA calculations suggested that an approximately three-fold increase in carbon yield and an eleven-fold increase in energy efficiency were theoretically possible.
-
-To investigate the differences in carbon yield and energy efficiency, we compared the flux distributions predicted by ssFBA simulations for the different constraint cases (Fig.~\ref{fig:NetworkFlux}).
-%All cases heavily utilized the first step in the pentose phosphate pathway to generate NADPH; the carbon flux then continued through the Entner–Doudoroff pathway toward pyruvate.
-In cases i and ii, glycolysis was used for energy generation, and most of the carbon flux accumulated as acetate.
-The system produced NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to power oxidative phosphorylation.
-Case iii heavily utilized the first step in the pentose phosphate pathway to generate NADPH; the carbon flux then continued primarily through the Entner–Doudoroff pathway toward pyruvate.
-Case iii also predicted the accumulation of pyruvate, lacatate, acetate, and carbon dioxide, which contributed to the much lower carbon yield and energy efficiency.
-In all cases, the energy source was primarily oxidative phosphorylation converting cofactors generated throughout the network into ATP.
-%Oxidative phosphorylation lead to a high redox ratio contributing to the accumulation of acetate overflow.
-Taken together, this suggested CAT production could be increased by reducing the accumulation of acetate, lactate, and carbon dioxide.
 
@@
 A single knockout in oxidative phosphorylation reduced the CAT carbon yield \texttildelow 3-fold, as well as disrupting the system state, showing its crucial role in CFPS.
-In comparing flux distributions between low and high yield cases, carbon flux could be potentially diverted toward CAT by reducing acetate overflow and minimizing flux through the Entner-Doudoroff pathway.
 Taken together, these findings represent the first dynamic model of \textit{E. coli} cell-free protein synthesis, and an important step toward a functional genome scale description.
@@
 While it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, both of our modeling approaches suggest its importance to improving CFPS performance and protein yield.
-%Also important is the first step in pentose phosphate pathway, as it converts G6P to 6PG and produces NADPH.
-%The next most sensitive reaction is the conversion of NADPH to NADH, which helps to supply NADH to oxidative phosphorylation reactions.
-%Many of the insignificant reactions lead to the accumulation of unwanted byproducts such as ethanol and formate.
 
-%The ensemble of models captured the dynamic distribution of metabolic flux.
-%The ensemble of dynamic models predicted that most of the carbon flux going through glycolysis and the TCA cycle,
-%while the constraint-based approach predicted most of the carbon flux went through the Entner-Doudoroff pathway and only part of glycolysis.
-% We present the first dynamic, cell-free model of \textit{E. coli} metabolism and biosynthesis at this scope.
-% While the early models of Shuler and coworkers achieved large-scale, dynamic descriptions of single cells \citep{1984_domach_shuler_BiotechBioeng_01}, and the stoichiometric models associated with the FBA approach are computationally efficient amd widespread \citep{2009_feist_palsson_NatRevMicrobio}, none have yet been able to construct genome-scale kinetic models of cell-free \textit{E.
-coli} metabolism.
-% This study should provide a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} metabolism, toward industrial-scale biosynthetic production.
-%However, the low sensitivities to glucose and oxygen only apply in the specific case of the parameter sets that were studied.
-%When other species' initial conditions and associated parameters are different, glucose and oxygen could become more important to CAT production.
-
-A logical next step for this work would be sequence-specific dynamic modeling, as the kinetic modeling approach in this study used a single reaction to approximate CAT synthesis.
-Including specific transcription and translation steps for CAT would allow more accurate modeling of the complexity and the resource cost of protein synthesis.
-Sensitivity analysis could then be performed on these new parameters to determine the robustness of CAT synthesis to the processes of transcription and translation.
 %Another area for future work is to more thoroughly sample parameter space.
@@
 Specifically, CAT production performed in the absence of amino acids could inform the system's ability to manufacture them, while experimentation in the absence of glucose or oxygen could shed light on the importance of those substrates.
-%how important they are to protein synthesis, and under which conditions.
 Finally, the approach should be extended to other protein products.
@@
 After generating XXX sets, we selected N = 100 sets with minimal set to set correlation to avoid over-sampling any region of parameter space.
-% The original 12,437-set ensemble had a mean Pearson correlation coefficient of 0.94 between pairs of sets, while the 100-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.
 
@@
 Parameters and parameter combinations were stratified into five degrees of importance, from least to most sensitive.
-% This was achieved by assigning them a value of 0, 0.25, 0.5, 0.75, or 1.
-% The array of sensitivity indices was organized into a clustergram with both axes corresponding to the 163 maximum reaction rates
-% rearranged from most sensitive (top, right) to least sensitive (bottom, left).
 
@@
 The parameter sensitivities were stratified into five degrees of importance, from least to most sensitive, as above.
-
-% This array of sensitivity indices was also organized into a clustergram (Fig.~\ref{fig:Sensitivity-CAT}, bottom).
-% Theoretical maximum CAT carbon yields for three cases discussed below were calculated using flux balance anaylsis (FBA) with a sequence-specific description of CAT synthesis, where $\Delta m_{i}$ denotes the flux of the $i^{th}$ species.
-% The transcription and translation template reactions come from sequence-specific analysis \cite{Allen:2003aa}, and include transcription initiation, transcription, mRNA degradation, translation initiation, translation, and tRNA charging.
-% The rest of the network followed a pseudo steady-state assumption where metabolites were not allowed to accumulate; thus, the network could be solved by linear programming instead of solving differential equations.
 % \begin{eqnarray*}
@@
 % \end{eqnarray*}
-
-\subsection*{Sequence-specific calculation of carbon yield.}
-We estimated the theoretical maximum CAT carbon yield using sequence-specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
-The sequence-specific flux balance analysis problem was formulated as a linear program:
-\begin{equation}
- \begin{multlined}
-	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{TL} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
-	\mathrm{Subject \; to:}
-	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
-\alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
- \end{multlined}
-\end{equation}
-where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
-and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
-The stoichiometry of the kinetic model was used for the ssFBA calculations, with the execpetion of the transcription and translation rates.
-The transcription (TX) and translation (TL) stoichiometry was modeled using the template reactions taken from Allen and Palsson \cite{Allen:2003aa}:
-\begin{eqnarray*}
-\mathrm{{G}_{\mathcal{P}}}+\mathrm{{R}_{1}} &\longrightarrow& \mathrm{G_{\mathcal{P}}^{*}} \\
-\mathrm{G_{\mathcal{P}}^{*}}+\sum_{\mathrm{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \mathrm{\left\{k\right\}TP} &\overset{\rm TX}{\longrightarrow}& \mathrm{mRNA}+\mathrm{G_{\mathcal{P}}}+\mathrm{R_{1}}+ \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}2\eta_{k}\cdot Pi}\\
-\mathrm{mRNA} &\longrightarrow& \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}\eta_{k}\cdot \left\{k\right\}MP} \\
-\mathrm{mRNA}+\mathrm{R_{2}} &\longrightarrow& \mathrm{R_{2}^{*}} \\
-\mathrm{\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP} &\longrightarrow& \mathrm{\alpha_{j}\cdot AA_{j}-tRNA_{j}+}\\
-&& \mathrm{\qquad \alpha_{j}\cdot AMP+2\alpha_{j}\cdot Pi} \qquad{j=1,2,\hdots,20}\\
-\mathrm{R_{2}^{*}+\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot \Big(AA_{j}-tRNA_{j}+2\cdot GTP\Big)} &\stackrel{\rm TL}{\longrightarrow}& \mathcal{P}+\mathrm{R_{2}+mRNA}+\\
-&& \mathrm{\quad +\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot\Big(tRNA+2\cdot GDP+2\cdot Pi\Big)}
-\end{eqnarray*}
-where $G_{\mathcal{P}}$ denotes the gene encoding protein product $\mathcal{P}$,
-$\rm R_{1}$ denotes the concentration of RNA polymerase,
-$G_{\mathcal{P}}^{*}$ denotes the gene bounded by the RNA polymerase,
-$\eta_{i}$ and $ \alpha_{j}$ denote the stoichiometric coefficients for nucleotide and amino acid, respectively,
-$\rm Pi$ denotes inorganic phosphate,
-$\rm R_{2}$ denotes the ribosome concentration,
-$\rm R_{2}^{*}$ denotes bounded ribosome,
-and $AA_{j}$ denotes $j^{th}$ amino acid.
-
-The transcription rate ($w_{TX}$) was fixed in the ssFBA calculation at:
-\begin{equation}
-	w_{TX} = V_{TX}^{max}\left(\frac{G}{K_{TX}+G}\right)
-\end{equation}
-where $G$ denotes the gene concentration, and $K_{TX}$ denotes a transcription saturation coefficient.
-The maximum rate of transcription $V_{TX}^{max}$ was formulated as:
-\begin{equation}
-	V_{TX}^{max} \equiv \left[R_{1}\left(\frac{v_{TX}}{l_{G}}\right)\left(\frac{K_{T7}}{1 + K_{T7}}\right)\right]
-\end{equation}
-The term $R_{1}$ denotes the RNA polymerase abundance,
-$v_{TX}$ denotes the RNA polymerase elongation rate (nt/hr),
-$l_{G}$ denotes the gene length in nucleotides, and the last term describes T7 promoter activity,
-where $K_{T7}$ denotes a T7 RNA polymerase binding constant \cite{Moon:2012aa}.
-On the other hand, the translation rate ($w_{TL}$) was bounded by:
- \begin{equation}
-	0\leq w_{TL} \leq V_{TL}^{max}\left(\frac{\rm mRNA_{SS}}{K_{TL}+\rm mRNA_{SS}}\right)
-\end{equation}
-where $\rm mRNA_{SS}$ denotes the steady state mRNA abundance, and $K_{TL}$ denotes the translation saturation constant.
-The maximum translation rate $V_{TL}^{max}$ was formulated as:
-\begin{equation}
-	V_{TL}^{max} \equiv \left[K_{P} R_{2}\left(\frac{v_{TL}}{l_{P}}\right)\right]
-\end{equation}
-The term $K_{P}$ denotes the polysome amplification constant,
-$v_{TL}$ denotes the ribosome elongation rate (amino acids per hour),
-$l_{P}$ denotes the number of amino acids in the protein of interest,
-and $\rm mRNA_{SS}$ denotes the steady-state mRNA concentration:
-\begin{equation}
-	\rm mRNA_{SS}\simeq\frac{w_{TX}}{\lambda}
-\end{equation}where $\lambda$ denotes the rate constant controlling the mRNA degradation rate.
-
-The objective of the sequence-specific flux balance calculation was to maximize the rate of CAT translation, $w_{TL}$.
-The total glucose uptake rate was bounded by [0,40 mM/h] according to experimental data;
-while the amino acid uptake rates were bounded by [0,30 mM/h], but did not reach the maximum flux.
-The CAT gene and protein sequences were taken from literature.
-The sequence-specific flux balance linear program was solved using the GNU Linear Programming Kit (GLPK) v4.52 \cite{GLPK}.
 
@@
 J.V, J.W, and N.H developed the cell free protein synthesis mathematical model, and parameter ensemble.
-J.V and M.V performed the sequence-specific flux balance analysis calculations.
 The manuscript was prepared and edited for publication by J.S, N.H, M.V, J.W and J.V.
@@
 % Figures and captions go here ...
-% \begin{figure}[ht]
-% \centering
-% \includegraphics[width=1.00\textwidth]{./Figures/Carbon.pdf}
-% \caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species.
-Best-fit parameter set (orange line) versus experimental data (points).
-95\% confidence interval (blue shaded region) and 95\% confidence interval of the mean (gray shaded region) over the ensemble of 100 sets.}
-% \label{fig:Carbon}
-% \end{figure}
 
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/Carbon.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/Fig_1_Carbon.pdf}
 \caption{Central carbon metabolism in the presence (top) and absence (bottom) of allosteric control, including glucose (substrate), CAT (product), and intermediates, as well as total concentration of energy species.
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/Energy.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/Fig_2_Energy.pdf}
 \caption{Energy species and energy totals by base in the presence of allosteric control.
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/Amino.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/Fig_3_Amino.pdf}
 \caption{Amino acids in the presence of allosteric control.
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/CostArray-BoxPlot.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/Fig_4_EnsembleVsRandom.pdf}
 \caption{Log of cost function across 37 datasets for data-trained ensemble (blue) and randomly generated ensemble (red, gray background).
@@
 \centering
-\includegraphics[width=0.7\textwidth]{./Figures/GroupKO.pdf}
+\includegraphics[width=0.7\textwidth]{./Figures/Fig_5_GroupKO.pdf}
 \caption{Effect of group knockouts on system.
@@
 
-\begin{figure}[ht]
+\clearpage
+
+%\begin{table}
+%\centering
+% \caption{}
+% \renewcommand{\arraystretch}{1.3}
+% \begin{tabular}{llrrr} \toprule
+% & \multicolumn{1}{r}{\textbf{Original value}} & \multicolumn{2}{r}{\textbf{After dilution factor}} & \textbf{BN ID} \\ \hline
+%\smallskip
+%\rule{0pt}{3ex} \parbox{4.3 cm}{Characteristic Enzyme Concentration} & \multicolumn{1}{r}{5 $\mu$M} & \multicolumn{2}{r}{0.1667 $\mu$M} & 100735 \\ \toprule
+%\phantom{i}\textbf{Enzyme} & \textbf{Reaction name} & \textbf{Kcat (h$^-1$)} & 
+% \textbf{Vmax (mM/h)} & \textbf{BN ID} \\ \hline
+%\phantom{i}Serine dehydrase & R\_ser\_deg & 624000 & 104 & 101119 \\ \hline
+%\phantom{i}Isocitrate dehydrogenase & R\_icd & 714000 & 119 & 101152 \\ \hline
+%\phantom{i}Lactate dehydrogenase & R\_ldh & 348000 & 58 & 101036 \\ \hline
+%\phantom{i}Aspartate transaminase & R\_aspC, R\_tyr, R\_phe & 1548000 & 258 & 101108 \\ \hline
+%\phantom{i}Enolase & R\_eno & 792000 & 132 & 101028 \\ \hline
+%\smallskip
+%\rule{0pt}{3ex}
+%Pyruvate kinase & R\_pyk & 1500000 & 250 & \parbox{1.4 cm}{101029, 101030} \\ \hline
+%\phantom{i}Malic enzyme & R\_maeA, R\_maeB & 2124000 & 354 & 101167 \\ \hline
+%\phantom{i}Phosphofructokinase & R\_pfk & 33264000 & 5544 & 104955 \\ \hline
+%\phantom{i}Malate dehydrogenase & R\_mdh & 1980000 & 330 & 101163 \\ \hline
+%\phantom{i}Citrate Synthase & R\_gltA & 2520000 & 420 & 101149 \\ \hline
+%\phantom{i}6PG dehydrogenase & R\_zwf, R\_pgl, R\_gnd & 192000 & 32 & 101048 \\ \hline
+%\phantom{i}Succinate dehydrogenase & R\_sdh & 7260 & 1.21 & 101162 \\ \hline
+%\phantom{i}Succinyl-coA Synthetase & R\_sucCD & 282000 & 47 & 101158 \\ \hline
+%\phantom{i}3PGA dehydrogenase & R\_gpm & 66000 & 11 & 101135 \\ \hline
+%\phantom{i}PEP carboxylase & R\_ppc & 2124000 & 354 & 101139 \\ \hline
+%\phantom{i}3PGA kinase & R\_pgk & 258000 & 43 & 101016 \\ \midrule
+%\phantom{i}Geometric mean & & & 110 & \\ \toprule
+%\phantom{i}\textbf{Transcription/Translation} & \textbf{AA-tRNA complex (mM)} & \textbf{Kcat (h$^-1$)} & \textbf{Vmax (mM/h)} & \textbf{BN ID} \\ \hline
+%\phantom{i}tRNA charging & 0.03 & 14040 & 4212 & 104980 \\ \hline
+%% Serine dehydrase & Yield & 8.1\% & 6.9\% & 2.6\% \\ \bottomrule
+% \end{tabular}
+%\label{tbl:Literature_values}
+%\end{table}
+
+\begin{table}
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/Sens_productivity.pdf}
-\caption{Sensitivity of CAT production to model parameters, decomposed into eigenmodes.
-Sensitivity of top 10 eigenvectors  (columns, left to right) to top 20 most influential parameters (rows, bottom to top).
-Relative magnitudes of eigenvalues expressed as percentages.}
-\label{fig:Sensitivity-CAT}
-\end{figure}
+ \caption{}
+ \renewcommand{\arraystretch}{1.3}
+ \begin{tabular}{llrrr} \toprule
+\smallskip
+ & & \textbf{Original value} & \parbox{2.8 cm}{\textbf{After cell-free dilution factor}} & \textbf{BN ID} \\ \hline
+\smallskip
+\rule{0pt}{3ex} \parbox{4.3 cm}{Characteristic Enzyme Concentration} & & 5 $\mu$M & 167 nM & 100735 \\ \toprule
+\phantom{i}\textbf{Enzyme} & \textbf{Reaction} & \textbf{Kcat (min$^{-1}$)} & 
+ \textbf{Vmax (mM/h)} & \textbf{BN ID} \\ \hline
+\phantom{i}Serine dehydrase & R\_ser\_deg & 10400 & 104 & 101119 \\ \hline
+\phantom{i}Isocitrate dehydrogenase & R\_icd & 11900 & 119 & 101152 \\ \hline
+\phantom{i}Lactate dehydrogenase & R\_ldh & 5800 & 58 & 101036 \\ \hline
+\smallskip
+\rule{0pt}{4.5ex}
+Aspartate transaminase & \parbox{1 cm}{R\_aspC, R\_tyr, R\_phe} & 25800 & 258 & 101108 \\ \hline
+\phantom{i}Enolase & R\_eno & 13200 & 132 & 101028 \\ \hline
+\smallskip
+\rule{0pt}{3ex}
+Pyruvate kinase & R\_pyk & 25000 & 250 & \parbox{1.4 cm}{101029, 101030} \\ \hline
+\smallskip
+\rule{0pt}{3ex}
+Malic enzyme & \parbox{1 cm}{R\_maeA, R\_maeB} & 35400 & 354 & 101167 \\ \hline
+\phantom{i}Phosphofructokinase & R\_pfk & 554400 & 5544 & 104955 \\ \hline
+\phantom{i}Malate dehydrogenase & R\_mdh & 33000 & 330 & 101163 \\ \hline
+\phantom{i}Citrate Synthase & R\_gltA & 42000 & 420 & 101149 \\ \hline
+\smallskip
+\rule{0pt}{4.5ex}
+6PG dehydrogenase & \parbox{1 cm}{R\_zwf, R\_pgl, R\_gnd} &3200 & 32 & 101048 \\ \hline
+\phantom{i}Succinate dehydrogenase & R\_sdh & 121 & 1.21 & 101162 \\ \hline
+\phantom{i}Succinyl-coA synthetase & R\_sucCD & 4700 & 47 & 101158 \\ \hline
+\phantom{i}3PGA dehydrogenase & R\_gpm & 1100 & 11 & 101135 \\ \hline
+\phantom{i}PEP carboxylase & R\_ppc & 35400 & 354 & 101139 \\ \hline
+\phantom{i}3PGA kinase & R\_pgk & 4300 & 43 & 101016 \\ \midrule
+\phantom{i}Geometric mean & & & 110 & \\ \toprule
+\phantom{i}\textbf{Transcription/Translation} \\ \hline
+%\phantom{i}\textbf{Transcription/Translation} & \textbf{AA-tRNA (mM)} & \textbf{Kcat (h$^-1$)} & \textbf{Vmax (mM/h)} & \textbf{BN ID} \\ \hline
+\phantom{i}tRNA charging & 0.03 & 14040 & 4212 & 104980 \\ \hline
+% \bottomrule
+ \end{tabular}
+\label{tbl:Literature_values}
+\end{table}
 
-\begin{figure}[ht]
-\centering
-\includegraphics[width=1.00\textwidth]{./Figures/oxKO_ssFBA_yields.pdf}
-\caption{The effects of oxidative phosphorylation and amino acid synthesis pathways on CAT production and carbon yield.
-A.
-95\% confidence interval of the ensemble of kinetic models for CAT concentration versus time, for the best-fit set with no knockouts (black shaded region and dashed line), \textit{nuo} knockout (medium gray), and \textit{cyd} knockout (light gray).
-B.
-CAT carbon yield of the ensemble of kinetic models for no knockouts (black), \textit{app} knockout (dark gray), \textit{nuo} knockout (medium gray), \textit{cyd} knockout (light gray), and all three knockouts (white).
-Error bars represent standard deviation of the ensemble.
-C.
-95\% confidence interval of the ensemble of ssFBA simulations (light blue region) of CAT concentration over time, against experimental data (black).
-D.
-Theoretical maximum carbon yield of CAT production, calcualted by ssFBA for three different cases: constrained by transcription/translation (TX/TL) rates (black), same as previous but without amino acid synthesis reactions (medium gray), and same as previous but constrained by experimental measurements where available (light gray).
-Error bars represent standard deviation of the ensemble.}
-\label{fig:oxKO_ssFBA_yields}
-\end{figure}
+%\begin{table}
+%\centering
+% \caption{}
+% \renewcommand{\arraystretch}{1.3}
+% \begin{tabular}{llrrr} \toprule
+% & & \textbf{Original value} & \textbf{After dilution factor} & \textbf{BN ID} \\ \hline
+%\smallskip
+%\rule{0pt}{3ex} \parbox{4.3 cm}{Characteristic Enzyme Concentration} & & 5 $\mu$M & 0.1667 $\mu$M & 100735 \\ \toprule
+%\phantom{i}\textbf{Enzyme} & \textbf{Reaction} & \textbf{Kcat (h$^-1$)} & 
+% \textbf{Vmax (mM/h)} & \textbf{BN ID} \\ \hline
+%\phantom{i}Serine dehydrase & R\_ser\_deg & 624000 & 104 & 101119 \\ \hline
+%\phantom{i}Isocitrate dehydrogenase & R\_icd & 714000 & 119 & 101152 \\ \hline
+%\phantom{i}Lactate dehydrogenase & R\_ldh & 348000 & 58 & 101036 \\ \hline
+%\smallskip
+%\rule{0pt}{4.5ex}
+%Aspartate transaminase & \parbox{1 cm}{R\_aspC, R\_tyr, R\_phe} & 1548000 & 258 & 101108 \\ \hline
+%\phantom{i}Enolase & R\_eno & 792000 & 132 & 101028 \\ \hline
+%\smallskip
+%\rule{0pt}{3ex}
+%Pyruvate kinase & R\_pyk & 1500000 & 250 & \parbox{1.4 cm}{101029, 101030} \\ \hline
+%\smallskip
+%\rule{0pt}{3ex}
+%Malic enzyme & \parbox{1 cm}{R\_maeA, R\_maeB} & 2124000 & 354 & 101167 \\ \hline
+%\phantom{i}Phosphofructokinase & R\_pfk & 33264000 & 5544 & 104955 \\ \hline
+%\phantom{i}Malate dehydrogenase & R\_mdh & 1980000 & 330 & 101163 \\ \hline
+%\phantom{i}Citrate Synthase & R\_gltA & 2520000 & 420 & 101149 \\ \hline
+%\smallskip
+%\rule{0pt}{4.5ex}
+%6PG dehydrogenase & \parbox{1 cm}{R\_zwf, R\_pgl, R\_gnd} & 192000 & 32 & 101048 \\ \hline
+%\phantom{i}Succinate dehydrogenase & R\_sdh & 7260 & 1.21 & 101162 \\ \hline
+%\phantom{i}Succinyl-coA synthetase & R\_sucCD & 282000 & 47 & 101158 \\ \hline
+%\phantom{i}3PGA dehydrogenase & R\_gpm & 66000 & 11 & 101135 \\ \hline
+%\phantom{i}PEP carboxylase & R\_ppc & 2124000 & 354 & 101139 \\ \hline
+%\phantom{i}3PGA kinase & R\_pgk & 258000 & 43 & 101016 \\ \midrule
+%\phantom{i}Geometric mean & & & 110 & \\ \toprule
+%%\phantom{i}\textbf{Transcription/Translation} & \textbf{AA-tRNA (mM)} & \textbf{Kcat (h$^-1$)} & \textbf{Vmax (mM/h)} & \textbf{BN ID} \\ \hline
+%\phantom{i}tRNA charging & 0.03 & 14040 & 4212 & 104980 \\ \hline
+%% \bottomrule
+% \end{tabular}
+%\label{tbl:Literature_values}
+%\end{table}
+
+%\begin{table}
+%\centering
+%    \caption{CAT carbon yield breakdown for best-fit set, knockouts, and experimental data.
+Carbon produced as CAT, carbon consumed as glucose and each amino acid, sum of consumed species, and yield.
+Accumulation of alanine and glutamine (negative consumption terms) was not considered in yield calculation.}
+%    \renewcommand{\arraystretch}{1.3}
+%    \begin{tabular}{lrrrrrr} \toprule
+%        Carbon Produced (C-mM) & Best-fit & $\Delta$app & $\Delta$nuo & $\Delta$cyd & \parbox{1 cm}{$\Delta$app $\Delta$nuo $\Delta$cyd} & Data \\ \hline
+%        CAT & 20.9 & 21.4 & 18.1 & 6.5 & 5.1 & 21.6 \\ \midrule
+%        Carbon Consumed (C-mM) \\ \hline
+%        GLC & 215.4 & 215.4 & 215.4 & 215.4 & 159.8 & 215.4 \\ \hline
+%        ALA & -11.6 & -11.4 & 1.7 & -3.8 & -3.2 & -12.1 \\ \hline
+%%        ARG & 10.2 & 9.9 & 1.1 & 0.9 & 1.3 & - \\ \hline
+%        ASN & 6.2 & 6.2 & 6.2 & 6.3 & 6.3 & 6.3 \\ \hline
+%        ASP & 7.5 & 7.5 & 3.9 & 0.0 & 0.0 & 9.6 \\ \hline
+%        CYS & 3.0 & 3.1 & 3.0 & 2.9 & 2.9 & 3.7 \\ \hline
+%        GLN & -11.4 & -11.3 & -4.0 & 1.8 & 2.7 & -11.7 \\ \hline
+%%        GLU & 492.6 & 505.6 & 528.2 & 505.1 & 501.8 & - \\ \hline
+%        GLY & 3.1 & 3.1 & 2.6 & 1.1 & 0.9 & 1.5 \\ \hline
+%        HIS & 0.2 & 0.2 & 1.1 & 0.4 & 0.3 & 0.0 \\ \hline
+%        ILE & 1.0 & 1.0 & 0.8 & 0.3 & 0.2 & 1.7 \\ \hline
+%        LEU & 1.4 & 1.4 & 1.2 & 0.4 & 0.3 & 2.0 \\ \hline
+%        LYS & 10.7 & 10.7 & 13.1 & 13.2 & 13.2 & 8.3 \\ \hline
+%        MET & 0.8 & 0.8 & 0.7 & 0.2 & 0.2 & 2.9 \\ \hline
+%        PHE & 3.2 & 3.3 & 2.8 & 1.0 & 0.8 & 1.6 \\ \hline
+%        PRO & 2.4 & 2.4 & 0.7 & 0.2 & 0.2 & 1.9 \\ \hline
+%        SER & 2.5 & 2.5 & 2.4 & 2.1 & 2.1 & 1.8 \\ \hline
+%        THR & 3.4 & 3.4 & 3.3 & 2.9 & 2.8 & 2.8 \\ \hline
+%        TRP & 1.0 & 1.0 & 0.8 & 0.3 & 0.2 & 1.2 \\ \hline
+%        TYR & 1.1 & 1.1 & 1.1 & 0.4 & 0.4 & 0.6 \\ \hline
+%        VAL & 1.4 & 1.5 & 1.2 & 0.4 & 0.4 & 2.4 \\ \midrule
+%%        Sum & 767.1 & 780.1 & 791.3 & 755.3 & 696.8 & - \\ \hline
+%        Sum & 264.3 & 264.6 & 262.0 & 249.3 & 193.7 & 263.7 \\ \midrule
+%%        Yield & 2.7\% & 2.7\% & 2.3\% & 0.9\% & 0.7\% & - \\ \hline
+%        Yield & 7.9\% & 8.1\% & 6.9\% & 2.6\% & 2.7\% & 8.2\% \\ \bottomrule
+%    \end{tabular}
+%\label{tbl:yield_breakdown}
+%\end{table}
+
 \clearpage
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/NetworkFlux.pdf}
-\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, oxidative phosphorylation, and protein synthesis.
-Thicker and redder arrows signify larger flux values (see colorbar).
-A.
-ssFBA constrained by transcription and translation rates, with or without amino acid synthesis reactions.
-B.
-ssFBA constrained by transcription and translation rates and experimental measurements where available.}
-%Sequence-specific FBA flux value (mean ± standard deviation) across ensemble for 1 hr, normalized to glucose uptake flux.
-Flux distribution for three different cases: constrained by transcription and translation rates (top row), same as previous but without amino acid synthesis reactions  (second row), and same as previous but constrained by experimental measurements where available (bottom row).}
-\label{fig:NetworkFlux}
-\end{figure}
-
-\begin{table}
-\centering
-    \caption{CAT carbon yield breakdown for best-fit set, knockouts, and experimental data.
-Carbon produced as CAT, carbon consumed as glucose and each amino acid, sum of consumed species, and yield.
-Accumulation of alanine and glutamine (negative consumption terms) was not considered in yield calculation.}
-    \renewcommand{\arraystretch}{1.3}
-    \begin{tabular}{lrrrrrr} \toprule
-        Carbon Produced (C-mM) & Best-fit & $\Delta$app & $\Delta$nuo & $\Delta$cyd & \parbox{1 cm}{$\Delta$app $\Delta$nuo $\Delta$cyd} & Data \\ \hline
-        CAT & 20.9 & 21.4 & 18.1 & 6.5 & 5.1 & 21.6 \\ \midrule
-        Carbon Consumed (C-mM) \\ \hline
-        GLC & 215.4 & 215.4 & 215.4 & 215.4 & 159.8 & 215.4 \\ \hline
-        ALA & -11.6 & -11.4 & 1.7 & -3.8 & -3.2 & -12.1 \\ \hline
-%        ARG & 10.2 & 9.9 & 1.1 & 0.9 & 1.3 & - \\ \hline
-        ASN & 6.2 & 6.2 & 6.2 & 6.3 & 6.3 & 6.3 \\ \hline
-        ASP & 7.5 & 7.5 & 3.9 & 0.0 & 0.0 & 9.6 \\ \hline
-        CYS & 3.0 & 3.1 & 3.0 & 2.9 & 2.9 & 3.7 \\ \hline
-        GLN & -11.4 & -11.3 & -4.0 & 1.8 & 2.7 & -11.7 \\ \hline
-%        GLU & 492.6 & 505.6 & 528.2 & 505.1 & 501.8 & - \\ \hline
-        GLY & 3.1 & 3.1 & 2.6 & 1.1 & 0.9 & 1.5 \\ \hline
-        HIS & 0.2 & 0.2 & 1.1 & 0.4 & 0.3 & 0.0 \\ \hline
-        ILE & 1.0 & 1.0 & 0.8 & 0.3 & 0.2 & 1.7 \\ \hline
-        LEU & 1.4 & 1.4 & 1.2 & 0.4 & 0.3 & 2.0 \\ \hline
-        LYS & 10.7 & 10.7 & 13.1 & 13.2 & 13.2 & 8.3 \\ \hline
-        MET & 0.8 & 0.8 & 0.7 & 0.2 & 0.2 & 2.9 \\ \hline
-        PHE & 3.2 & 3.3 & 2.8 & 1.0 & 0.8 & 1.6 \\ \hline
-        PRO & 2.4 & 2.4 & 0.7 & 0.2 & 0.2 & 1.9 \\ \hline
-        SER & 2.5 & 2.5 & 2.4 & 2.1 & 2.1 & 1.8 \\ \hline
-        THR & 3.4 & 3.4 & 3.3 & 2.9 & 2.8 & 2.8 \\ \hline
-        TRP & 1.0 & 1.0 & 0.8 & 0.3 & 0.2 & 1.2 \\ \hline
-        TYR & 1.1 & 1.1 & 1.1 & 0.4 & 0.4 & 0.6 \\ \hline
-        VAL & 1.4 & 1.5 & 1.2 & 0.4 & 0.4 & 2.4 \\ \midrule
-%        Sum & 767.1 & 780.1 & 791.3 & 755.3 & 696.8 & - \\ \hline
-        Sum & 264.3 & 264.6 & 262.0 & 249.3 & 193.7 & 263.7 \\ \midrule
-%        Yield & 2.7\% & 2.7\% & 2.3\% & 0.9\% & 0.7\% & - \\ \hline
-        Yield & 7.9\% & 8.1\% & 6.9\% & 2.6\% & 2.7\% & 8.2\% \\ \bottomrule
-    \end{tabular}
-\label{tbl:yield_breakdown}
-\end{table}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=1.00\textwidth]{./Figures/CAT_balances.pdf}
-\caption{Carbon and energy balances for the best-fit set.
+\includegraphics[width=0.6\textwidth]{./Figures/Fig_6_CarbonEnergyBalances.pdf}
+%\caption{Carbon and energy balances for the best-fit set.
 A.
@@
 Energy costs calculated in terms of equivalent ATP molecules.}
+\caption{Carbon and energy balances for the best-fit set.
+A.
+Carbon moles produced as CAT, amino acids (alanine and glutamine), organic acids (lactate, acetate, succinate, and malate), and other byproducts including carbon dioxide, as percentages of total carbon consumption (glucose and all other amino acids).
+B.
+Energy cost of CAT production, accumulation of amino acids (alanine and glutamine), accumulation of organic acids (lactate, acetate, succinate, and malate), and other byproducts, as percentages of total energy utilization from glucose.
+Energy costs calculated in terms of equivalent ATP molecules.}
 \label{fig:CAT_balances}
 \end{figure}
-
-%\begin{table}
-%\centering
-%    \caption{Amino acid total net consumption, consumption toward CAT, and net consumption other than toward CAT for the best-fit set.
-Total net consumption defined as decrease in amino acid concentration over 3 hours.
-Consumption toward CAT defined as increase in CAT concentration over 3 hours, multiplied by stoichiometric coefficient for given amino acid in CAT synthesis.}
-%    \renewcommand{\arraystretch}{1.3}
-%    \begin{tabular}{lrrr} \toprule
-%        & \parbox{3.6cm}{Total Net \phantom{abcdefgh} Consumption (mM)} & \parbox{3.4cm}{Consumption \phantom{abc} Toward CAT (mM)} & \parbox{3.6cm}{Other \phantom{abcdefgh} Consumption (mM)} \\ \hline
-%        ALA  & -3.85 & 0.27 & -4.12 \\ \hline
-%        ARG  & 1.70  & 0.18  & 1.52 \\ \hline
-%        ASN  & 1.55  & 0.09  & 1.46 \\ \hline
-%        ASP  & 1.87  & 0.21  & 1.66 \\ \hline
-%        CYS  & 1.02  & 0.09  & 0.93 \\ \hline
-%        GLN  & -2.28 & 0.21 & -2.49 \\ \hline
-%        GLU  & 98.53  & 0.23  & 98.30 \\ \hline
-%        GLY  & 1.55  & 0.18  & 1.37 \\ \hline
-%        HIS  & 0.04  & 0.21  & -0.17 \\ \hline
-%        ILE  & 0.16  & 0.16  & 0.00 \\ \hline
-%        LEU  & 0.23  & 0.23  & 0.00 \\ \hline
-%        LYS  & 1.79  & 0.21  & 1.58 \\ \hline
-%        MET  & 0.16  & 0.16  & 0.00 \\ \hline
-%        PHE  & 0.36  & 0.36  & 0.00 \\ \hline
-%        PRO  & 0.48  & 0.12  & 0.36 \\ \hline
-%        SER  & 0.83  & 0.18  & 0.65 \\ \hline
-%        THR  & 0.86  & 0.23  & 0.63 \\ \hline
-%        TRP  & 0.09  & 0.09  & 0.00 \\ \hline
-%        TYR  & 0.12  & 0.20  & -0.08 \\ \hline
-%        VAL  & 0.29  & 0.29  & 0.00 \\ \bottomrule
-%    \end{tabular}
-%\label{tbl:AA_breakdown}
-%\end{table}
 
@@
 
-\begin{figure}[ht]
-\centering
-\includegraphics[width=1.00\textwidth]{./Figures/Sens_state.pdf}
-\caption{Sensitivity of system state to model parameters, decomposed into eigenmodes.
-Sensitivity of top 10 eigenvectors  (columns, left to right) to top 20 most influential parameters (rows, bottom to top).
-Relative magnitudes of eigenvalues expressed as percentages.}
-\label{fig:Sensitivity-System}
-\end{figure}
-
 \end{document}

### 2017-05-11 Paper_v1.tex
May9
May9_2
May11

@@
 Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids, and the protein product, chloramphenicol acetyltransferase (CAT).
-The ensemble described all of the training data, especially the central carbon metabolism.
-XXX Talk about carbon and energy balances XXX
-This suggests that CAT production could be further optimized.
-The dynamic modeling approach predicted that substrate consumption and oxidative phosphorylation were most important to both CAT production and the system as a whole, while CAT production alone depended heavily on the CAT synthesis reaction.
-Conversely, CAT production was robust to allosteric control, as was most of the network, with the exception of the organic acids in central carbon metabolism.
-This study is the first to model dynamic protein production in \textit{E. coli}, and should provide a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
+The ensemble captured all of the training data, especially the central carbon metabolism.
+CAT was produced with a carbon yield of 7\% and an energy efficiency of 5\%, suggesting that CAT production could be further optimized.
+Reaction group knockouts showed that protein productivity and the metabolism as a whole depend most on oxidative phosphorylation and glycolysis and gluconeogenesis.
+Amino acid biosynthesis is also important for productivity, while the overflow metabolism and TCA cycle affect the overall system state.
+%The dynamic modeling approach predicted that substrate consumption and oxidative phosphorylation were most important to both CAT production and the system as a whole, while CAT production alone depended heavily on the CAT synthesis reaction.
+In addition, CAT production was robust to allosteric control, as was most of the network, with the exception of the organic acids in central carbon metabolism.
+This study is the first to model dynamic protein production in \textit{E. coli}, and should provide a foundation for genome scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
 
@@
 
-% Interestingly, many of the challenges confronting genome-scale kinetic modeling can potentially be overcome in a cell-free system.
+% Interestingly, many of the challenges confronting genome scale kinetic modeling can potentially be overcome in a cell-free system.
 % For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth.
 % Thus, cell-free operation holds several significant advantages for model development, identification and validation.
-Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms,
+Theoretically, genome scale cell-free kinetic models may be possible for industrially important organisms,
 % such as \textit{E. coli} or \textit{B.~subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
@@
 Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
-Central amongst these, is direct access to metabolites and the biosynthetic machinery without the interference of a cell wall, or complications associated with cell growth.
+Central amongst these is direct access to metabolites and the biosynthetic machinery without the interference of a cell wall, or complications associated with cell growth.
 This allows us to interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
@@
 The single cell \textit{E.
-coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple, regulated catabolic and anabolic pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}.
+coli} models of Shuler and coworkers pioneered the construction of large-scale, dynamic metabolic models that incorporated multiple regulated catabolic and anabolic pathways constrained by experimentally determined kinetic parameters \citep{1984_domach_shuler_BiotechBioeng_01}.
 Shuler and coworkers generated many single cell kinetic models, including single cell models of eukaryotes \citep{1989_steinmeyer_shuler_ChemEngSci,1992_wu_shuler_AnnNYAcadSci}, minimal cell architectures \citep{2004_castellanos_shuler_PNAS}, as well as DNA sequence based whole-cell models of \textit{E.
@@
 In the post genomics world, large-scale stoichiometric reconstructions of microbial metabolism popularized by techniques such as flux balance analysis (FBA) have become a standard approach \citep{2012_lewis_palsson_NatRevMicrobio}.
-Since the first genome-scale stoichiometric model of \textit{E.
-coli}, developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, well over 100 organisms, including industrially important prokaryotes are now available \citep{2009_feist_palsson_NatRevMicrobio,Feist:2007aa,Oh:2007aa}.
-Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome-scale kinetic models to an underdetermined linear algebraic system, which can be solved efficiently even for large systems.
+Since the first genome scale stoichiometric model of \textit{E.
+coli} developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, well over 100 organisms including industrially important prokaryotes are now available \citep{2009_feist_palsson_NatRevMicrobio,Feist:2007aa,Oh:2007aa}.
+Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome scale kinetic models to an underdetermined linear algebraic system, which can be solved efficiently even for large systems.
 Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
@@
 Thus, modern constraint-based approaches have proven extremely useful in the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
-However, genome-scale kinetic models of industrial important organisms such as \textit{E. coli} have yet to be constructed.
+However, genome scale kinetic models of industrial important organisms such as \textit{E. coli} have yet to be constructed.
 
-In this study, we developed an ensemble of kinetic cell-free protein synthesis (CFPS) models using dynamic metabolite measurements in an \textit{E. coli} cell free extract.
+In this study, we developed an ensemble of kinetic cell-free protein synthesis (CFPS) models using dynamic metabolite measurements in an \textit{E. coli} cell-free extract.
 Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids, and the protein product, chloramphenicol acetyltransferase (CAT).
-Characteristic values for model parameters and initial conditions, estimated from literature, were used to constrain the parameter estimation problem.
-The ensemble of parameter sets described the training data with a median cost that was greater than two orders of magnitude smaller than random sets constructed using the literature parameter constraints.
+Characteristic values for model parameters and metabolite initial conditions, estimated from literature, were used to constrain the parameter estimation problem.
+The ensemble of parameter sets described the training data with a median cost that was more than two orders of magnitude smaller than random sets constructed using the literature parameter constraints.
 We then used the ensemble of kinetic models to analyze the CFPS reaction.
-First, sensitivity analysis of the dynamic model suggested that CAT production was most sensitive to CAT synthesis parameters, as well as reactions in oxidative phosphorylation and pyruvate consumption.
-Sensitivity analysis also showed that the system as a whole was most sensitive to these same parts of the network and glucose consumption.
-CAT production and other metabolites, specifically organic acid intermediates such as pyruvate, were sensitive to the presence of allosteric control mechanisms.
+By knocking out metabolic reactions in groups, we showed that CAT productivity depends most on oxidative phosphorylation, glycolysis/gluconeogenesis, and certain amino acid synthesis reactions, while the system state is most sensitive to glycolysis/gluconeogenesis, oxidative phosphorylation, the overflow metabolism, and the TCA cycle reactions.
+%First, sensitivity analysis of the dynamic model suggested that CAT production was most sensitive to CAT synthesis parameters, as well as reactions in oxidative phosphorylation and pyruvate consumption.
+%Sensitivity analysis also showed that the system as a whole was most sensitive to these same parts of the network and glucose consumption.
+%CAT production and other metabolites, specifically organic acid intermediates such as pyruvate, were sensitive to the presence of allosteric control mechanisms.
 Taken together, we have integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
-This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
+This study provides a foundation for genome scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
 
@@
 The CFPS model equations were formulated using the hybrid cell-free modeling framework of Wayman et al. \cite{pr3010138}.
-An initial ensemble of model parameter sets (N $\approx$ 30,000) was estimated from measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids using a constrained Markov Chain Monte Carlo (MCMC) approach.
+An initial ensemble of model parameter sets (N $>$ 30,000) was estimated from measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids using a constrained Markov Chain Monte Carlo (MCMC) approach.
+%$\approx$
 The MCMC algorithm minimized the error between the training data and model simulations starting from an initial parameter set assembled from literature and inspection.
 A final ensemble of parameter sets (N $=$ 100) was constructed by selecting the sets with the lowest errors, the lowest of which was defined as the best-fit set.
-Parameter sets in the final ensemble had an average Pearson correlation coefficient of 0.77; thus, an accurate yet diverse ensemble was created.
-%had a median Pearson correlation coefficient of 0.79;
+Parameter sets in the final ensemble had a mean Pearson correlation coefficient of 0.77; thus, an accurate yet diverse ensemble was created.
+
 Central carbon metabolism (Fig.~\ref{fig:Carbon}, top), energy species (Fig.~\ref{fig:Energy}), and amino acids (Fig.~\ref{fig:Amino}) were captured by the ensemble and the best-fit set.
-The constrained MCMC approach estimated parameter sets with a median error an order of magnitude less than random parameter sets generated within the same parameter bounds (Fig.~\ref{fig:BoxPlot}); thus, we have confidence in the predictive capability of the estimated parameters.
+The constrained MCMC approach estimated parameter sets with a median error more than two orders of magnitude less than random parameter sets generated within the same parameter bounds (Fig.~\ref{fig:BoxPlot}); thus, we have confidence in the predictive capability of the estimated parameters.
 The model captured the biphasic CAT production: during the first hour glucose powers production, and CAT is produced at \texttildelow 10 $\mu$M/h; subsequently, pyruvate and lactate reserves are consumed to power metabolism, and CAT is produced less efficiently at \texttildelow 5 $\mu$M/h.
@@
 This is likely because aspartate, glutamate, and glutamine are key reactants in the biosynthesis of many other amino acids, all of which are required for CAT synthesis.
-%these amino acids are key reactants
-Meanwhile, the TCA cycle and the overflow metabolism, which includes acetyl-coA/acetate reactions
+Meanwhile, the TCA cycle and the overflow metabolism (which includes acetyl-coA/acetate reactions and the interconversion of pyruvate and lactate) have a significant effect on the system state.
+These reactions directly impact key species in the system state: succinate and malate in the TCA cycle, and acetate, pyruvate, and lactate in the overflow metabolism.
 
-The two largest total-order effects on the cost function are cysteine/methionine biosynthesis and anaplerotic/glyoxylate reactions.
-This is likely because...
-The effect of knocking out both of these reaction groups is understandably very great.
-However, the largest pairwise effect is seen when knocking out the Entner-Doudoroff pathway and the TCA cycle, two of the reaction groups with the smallest total-order effects.
-This is true of both the cost function knockout array and the system state knockout array.
-This may be because...
-
-To better understand which parameters and parameter combinations influenced the performance of the kinetic model, we performed sensitivity analysis (Fig.~\ref{fig:Sensitivity-CAT}).
-We perturbed each $V^{max}$ parameter, either individually or in pairwise combinations and measured the change in either CAT production or the overall system state.
-The eigen decomposition of the sensitivity shows that CAT synthesis and oxidative phosphorylation are the most important to overall CAT production, followed by the pyruvate-consuming alanine synthesis reaction.
-Among the top 20 reactions, we saw a common theme of the cofactors ATP, NADH, NADPH, and coenzyme A, as well as the metabolites pyruvate, glutamate, and $\alpha$-ketoglutarate.
-ATP appears 7 times in these 20 reactions, while pyruvate, glutamate, and coenzyme A appear 6 times each.
-NADH, NADPH, and $\alpha$-ketoglutarate each appear 5 times.
-This result makes sense, as the high energy cost of protein synthesis means that energy cofactors played a crucial role.
-Also, pyruvate served as the primary substrate after glucose ran out, and pyruvate, glutamate, and $\alpha$-ketoglutarate were all important precursors for the synthesis of amino acids required by CAT production.
-%Meanwhile, the cofactors provided energy to power CAT synthesis, while pyruvate was important for energy generation following glucose depletion.
-We performed the same eigen decomposition on the sensitivity of the overall system state to network reactions (Fig.~\ref{fig:Sensitivity-System}).
-Cytochrome oxidase, part of oxidative phosphorylation, was seen to have the greatest effect on the system state.
-Next most influential was the forward reaction of lactate dehydrogenase, followed by NADH:ubiquinone oxidoreductase, another oxidative phosphorylation reaction.
-The overall system state was also sensitive to cofactors and substrates, specifically NADH and pyruvate.
-%The system was most sensitive to an oxidative phosphorylation reaction (cytochrome oxidase), which converted ubiquinol to ubiquinone while generating a proton gradient.
-%The next 4 most important reactions were all consumers of pyruvate: lactate dehydrogenase, pyruvate formate lyase, alanine synthesis, and PEP synthase.
-Among the top 20 reactions, NADH appears in 8 reactions, pyruvate in 6, coenzyme A in 5, and ATP in 4.
-Glutamate, $\alpha$-ketoglutarate, G3P, and ubiquinone/ubiquinol appear 3 times each.
-Taken together, sensitivity analysis identified that substrates and energy cofactors, specifically those around oxidative phosphorylation, most influenced model performance.
-
-To understand whether the CFPS performance was optimal, we calculated the carbon yield and energy efficiency of CAT production (Fig.~\ref{fig:CAT_balances}).
-The best-fit parameter set for the kinetic model predicted a CAT carbon yield of 7.9\%, while the experimental dataset had a CAT carbon yield of 8.2\%.
-This was calculated as the increase in CAT concentration times the CAT carbon number, divided by the sum of the consumption terms for glucose and all amino acids excpet arginine and glutamate, as no data were available for these, weighted by their respective carbon numbers.
-To explain where the remainder of carbon was going, we performed a carbon balance for the best-fit set (Fig.~\ref{fig:CAT_balances}A).
-Of the other 92\% of carbon, 35\% accumulated as organic acids (lactate, acetate, succinate and malate) and 9\% accumulated as amino acids (alanine and glutamine).
-The remaining 48\% went to the net accumulation of all other metabolites, particularly carbon dioxide.
-The best-fit set and the experimental dataset both produced CAT with an energy efficiency of 7\% (Fig.~\ref{fig:CAT_balances}B).
-This was calculated as the increase in CAT concentration times the CAT number of equivalent ATP molecules, divided by glucose consumption times times the number of equivalent ATP molecules for glucose, equal to 15 in the optimal case.
-%An additional 62\% of energy went to the accumulation of glycolysis metabolites, 6\% to TCA cycle metabolites, and 25\% to acetate and lactate.
-An additional 62\% of energy went to the accumulation of glycolysis metabolites, and 31\% to organic acids (lactate, acetate, succinate and malate).
-This shows that there is much room for improvement of the efficiency of CFPS.
-A key finding of both the CAT and overall system state sensitivity analysis was the importance of oxidative phosphorylation.
+To understand whether the CFPS performance was optimal, we calculated the carbon yield and energy efficiency of CAT production for the best-fit set (Fig.~\ref{fig:CAT_balances}).
+Of the carbon taken up as glucose and consumed as amino acids, 7\% went to CAT production.
+Another 7\% went to the accumulation of alanine and glutamine, in line with experimental data (these were not included in the amino acid consumption term).
+25\% of the carbon was accumulated as organic acids (lactate, acetate, succinate, and malate), leaving 61\% as accumulation of other byproducts, species for which data do not exist.
+CFPS also showed an energy efficiency of 5\%; this was formulated as the equivalent ATP value of CAT produced divided by the equivalent ATP value of glucose consumed.
+The remainder of the balance was very similar to that of the carbon balance: 8\% to alanine and glutamine, 26\% to organic acids, and 61\% to other byproducts.
+The equivalent ATP numbers for glucose, amino acids, and organic acids were calculated from the network stoichiometry; for example, 21 molecules of ATP shold be generated from one glucose molecule if the optimal path through glycolysis, the TCA cycle, and oxidative phosphorylation is taken.
+The ATP number for CAT was simply equal to the cost of transcription and translation.
+These results show that there is much room for improvement of the efficiency of CFPS.
+This demonstrates that CAT carbon yield could be improved by diverting flux away from some of the less efficient or altogether unnecessary metabolic pathways.
+%pathways of energy generation
+%A key finding of both the CAT and overall system state sensitivity analysis was the importance of oxidative phosphorylation.
 
@@
 We used the hybrid cell-free modeling approach of Wayman and coworkers, which integrates traditional kinetic modeling with a logic-based description of allosteric regulation.
-CFPS is seen to be biphasic relying on glucose during the first hour and pyruvate and lactate afterward.
-Allosteric control was essential to the maintenance of the network and production of CAT, as without it, central carbon metabolism is exhuasted within 1.5 hours leading to low CAT production.
+Our ensemble of models accurately predicts dynamic experimental measurements of central carbon metabolism, energy species, and amino acids over 100 times better than random sets in the same region of parameter space.
+%goodness-of-fit
+CFPS is seen to be biphasic, relying on glucose during the first hour and pyruvate and lactate afterward.
+Allosteric control was essential to the maintenance of the network, specifically the central carbon metabolism.
+% and production of CAT.
+% central carbon metabolism is exhuasted within 1.5 hours, leading to low CAT production.
+Without it, pyruvate, succinate, and malate are consumed more quickly following glucose exhaustion, presumably to power downstream reactions and ultimately CAT synthesis.
+Meanwhile, acetate accumulation during the increased after the first hour, suggesting a decrease in sytem efficiency.
+Interestingly, CAT production is virtually unaffected.
+This is because the amino acids and energy species used in CAT synthesis are not dependent on allosteric control.
 Having captured the experimental data, we investigated if CAT yield and CFPS performance could be further improved.
-We showed that the model produces CAT with a carbon yield equal to 16\% of that of a physiological case in which transcription and translation are constrained, and an energy efficiency equal to 9\% of that of the physiological case.
+We showed that the model predicts CAT production with a carbon yield of 7\% and energy efficiency of 5\%.
+%We showed that the model produces CAT with a carbon yield equal to 16\% of that of a physiological case in which transcription and translation are constrained, and an energy efficiency equal to 9\% of that of the physiological case.
 The accumulation of waste byproducts, especially acetate and carbon dioxide, is responsible for this sub-optimal performance.
-Sensitivity analysis showed that certain substrates and energy species are instrumental to CAT production and overall metabolism.
-The system heavily relied on oxidative phosphorylation for the system's energetic needs as well as for CAT synthesis.
-A single knockout in oxidative phosphorylation reduced the CAT carbon yield \texttildelow 3-fold, as well as disrupting the system state, showing its crucial role in CFPS.
+However, there is also some accumulation of useful metabolites in the central carbon metabolism; if these could be utilized fully, CAT production would likely increase.
+Knocking out sections of network metabolism revealed that glycolysis/gluconeogenesis and oxidative phosphorylation were the most important to CAT production and the system as a whole.
+Productivity was also heavily dependent on the synthesis reactions of alanine, aspartate, asparagine, glutamate, and glutamine, while TCA cycle and overflow reactions affected the system state.
 Taken together, these findings represent the first dynamic model of \textit{E. coli} cell-free protein synthesis, and an important step toward a functional genome scale description.
@@
 We present an ensemble of models that quantitatively describes the system behavior of cell-free metabolism and production of CAT.
-Experimental observations of the metabolites and cometabolites validate the structure of the model and the estimation of kinetic parameters.
-This is important in applying metabolic engineering principles to rationally design cell-free production processes and predict the redirection of carbon fluxes to product forming pathways.
-In analyzing the model parameters' effect on CAT production, CAT synthesis is the most important, followed by oxidative phosphorylation and the glutamate and pyruvate consuming reactions, as well as cofactor reactions which are necessary to drive CAT synthesis.
-For example, the conversion of ATP to GTP shows significance since it is necessary for CAT synthesis.
-While Jewett and coworkers have shown that ATP may be at saturation in CFPS \cite{Jewett:2008aa}, GTP is also required for CAT synthesis and may be a limiting reactant.
-Thus, supplementation with additional GTP may improve the efficiency of CAT production.
-A similar theme is seen in the sensitivity of overall model state, where the most important reactions are glucose and pyruvate consuming reactions and cofactor reactions which are vital to drive CFPS.
-This can be seen in the biphasic operation of CFPS, with the first phase operating on glucose and the second phase operating on pyruvate.
-During the first phase, there is an accumulation of byproducts from central carbon with the majority of flux going toward acetate and some toward pyruvate, lactate, and succinate; with the exception of acetate, these are all consumed in the second phase.
-This shows that CAT production can be sustained by pyruvate and glutamate in the absence of glucose, which provides alternative strategies to optimize CFPS performance.
-This is in accordance with literature, which showed pyruvate provided a relatively slow but continuous supply of ATP \cite{swartz_nature2001}.
-Taken together, this shows CFPS can be designed towards a specified application either requiring a slow stable energy source or faster production.
-This outstanding control on model performace was expected as these metabolites are responsible for driving CFPS and represent the first step in the model network.
-Nevertheless, there are further reactions with considerable impact on model performance.
-In examining oxidative phosphorylation activity, knockouts in the electron transport pathways disrupt metabolism across the network and drop CAT carbon yield from 7.9\% to 2.6\%; Jewett and coworkers also saw a similar decrease in CAT yield with pyruvate as the substrate, ranging from 1.5-fold to 4-fold, when knocking out oxidative phosphorylation reactions \cite{Jewett:2008aa}.
+Experimental observations of the metabolites validate the structure of the model and the estimation of kinetic parameters.
+% and cometabolites
+This is important in applying metabolic engineering principles to rationally design cell-free production processes and predicting the redirection of carbon fluxes to product-forming pathways.
+In analyzing the effect of reaction groups on CAT production and the system state, the regions of metabolism associated with substrate utilization and subsequent energy generation are the most important.
 Oxidative phosphorylation is vital, since it provides most of the energetic needs of CFPS.
-While it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, both of our modeling approaches suggest its importance to improving CFPS performance and protein yield.
+Jewett and coworkers observed a decrease in CAT yield, between 1.5-fold and 4-fold, when knocking out oxidative phosphorylation reactions \cite{Jewett:2008aa}.
+% with pyruvate as the substrate
+While it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, our modeling approach suggests its importance to CFPS performance and protein yield.
+However, the biphasic operation of CFPS highlights the ability of the system to respond to an absence of glucose.
+% with the first phase operating on glucose and the second phase operating on pyruvate.
+During the first phase, there is an accumulation of central carbon metabolites with the majority of flux going toward acetate and some toward pyruvate, lactate, succinate and malate.
+While acetate continues to accumulate as a byproduct, the other organic acids are consumed as secondary substrates after glucose is no longer available.
+Glutamate also serves as a substrate throughout both phases, powering amino acid synthesis.
+These results show that CAT production can be sustained by other substrates in the absence of glucose, providing alternative strategies to optimize CFPS performance.
+% sustained by pyruvate and glutamate
+While CAT synthesis can be powered by other substrates, the rate of production about half (\texttildelow 5 $\mu$M/h, as opposed to \texttildelow 10 $\mu$M/h).
+This is in accordance with literature, where pyruvate provided a relatively slow but continuous supply of ATP \cite{swartz_nature2001}.
+%This outstanding control on model performace was expected, as these metabolites are responsible for driving CFPS and represent the first step in the model network.
+Taken together, this shows CFPS can be designed towards a specified application, either requiring a slow stable energy source or faster production.
+
+%CAT synthesis is the most important, followed by oxidative phosphorylation and the glutamate and pyruvate consuming reactions, as well as cofactor reactions which are necessary to drive CAT synthesis.
+%For example, the conversion of ATP to GTP shows significance since it is necessary for CAT synthesis.
+%While Jewett and coworkers have shown that ATP may be at saturation in CFPS \cite{Jewett:2008aa}, GTP is also required for CAT synthesis and may be a limiting reactant.
+%Thus, supplementation with additional GTP may improve the efficiency of CAT production.
+%A similar theme is seen in the sensitivity of overall model state, where the most important reactions are glucose and pyruvate consuming reactions and cofactor reactions which are vital to drive CFPS.
+%This can be seen in the biphasic operation of CFPS, with the first phase operating on glucose and the second phase operating on pyruvate.
+%During the first phase, there is an accumulation of byproducts from central carbon with the majority of flux going toward acetate and some toward pyruvate, lactate, and succinate; with the exception of acetate, these are all consumed in the second phase.
+%This shows that CAT production can be sustained by pyruvate and glutamate in the absence of glucose, which provides alternative strategies to optimize CFPS performance.
+%This is in accordance with literature, which showed pyruvate provided a relatively slow but continuous supply of ATP \cite{swartz_nature2001}.
+%Taken together, this shows CFPS can be designed towards a specified application either requiring a slow stable energy source or faster production.
+%This outstanding control on model performace was expected as these metabolites are responsible for driving CFPS and represent the first step in the model network.
+%Nevertheless, there are further reactions with considerable impact on model performance.
+%In examining oxidative phosphorylation activity, knockouts in the electron transport pathways disrupt metabolism across the network and drop CAT carbon yield from 7.9\% to 2.6\%; Jewett and coworkers also saw a similar decrease in CAT yield with pyruvate as the substrate, ranging from 1.5-fold to 4-fold, when knocking out oxidative phosphorylation reactions \cite{Jewett:2008aa}.
+%Oxidative phosphorylation is vital, since it provides most of the energetic needs of CFPS.
+%While it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, both of our modeling approaches suggest its importance to improving CFPS performance and protein yield.
 
@@
 %Testing the model under a variety of conditions could strengthen or challenge the findings of this study.
-In addition, further experimentation could also be used to gain a deeper understanding of model performance under a variety of conditions.
+
+In this study we present an ensemble of \textit{E.
+coli} cell-free protein synthesis (CFPS) models that accurately predict a comprehensive CFPS dataset of glucose, CAT, central carbon metabolites, energy species, and amino acid measurements.
+
+Our ensemble of models accurately predicts dynamic experimental measurements of central carbon metabolism, energy species, and amino acids over 100 times better than random sets in the same region of parameter space.
+
+CFPS is seen to be biphasic, relying on glucose during the first hour and pyruvate and lactate afterward.
+Allosteric control was essential to the maintenance of the network, specifically the central carbon metabolism.
+% and production of CAT.
+% central carbon metabolism is exhuasted within 1.5 hours, leading to low CAT production.
+Without it, pyruvate, succinate, and malate are consumed more quickly following glucose exhaustion, presumably to power downstream reactions and ultimately CAT synthesis.
+Meanwhile, acetate accumulation during the increased after the first hour, suggesting a decrease in sytem efficiency.
+Interestingly, CAT production is virtually unaffected.
+This is because the amino acids and energy species used in CAT synthesis are not dependent on allosteric control.
+Having captured the experimental data, we investigated if CAT yield and CFPS performance could be further improved.
+We showed that the model predicts CAT production with a carbon yield of 7\% and energy efficiency of 5\%.
+%We showed that the model produces CAT with a carbon yield equal to 16\% of that of a physiological case in which transcription and translation are constrained, and an energy efficiency equal to 9\% of that of the physiological case.
+The accumulation of waste byproducts, especially acetate and carbon dioxide, is responsible for this sub-optimal performance.
+However, there is also some accumulation of useful metabolites in the central carbon metabolism; if these could be utilized fully, CAT production would likely increase.
+Knocking out sections of network metabolism revealed that glycolysis/gluconeogenesis and oxidative phosphorylation were the most important to CAT production and the system as a whole.
+Productivity was also heavily dependent on the synthesis reactions of alanine, aspartate, asparagine, glutamate, and glutamine, while TCA cycle and overflow reactions affected the system state.
+Taken together, these findings represent the first dynamic model of \textit{E. coli} cell-free protein synthesis, and an important step toward a functional genome scale description.
+
+This work represents the first dynamic model of \textit{E. coli} cell-free protein synthesis.
+We apply a hybrid modeling framework to capture an experimental dataset for production of a test protein, and identify system limitations and areas of improvement for production efficiency.
+This work could be extended through further experimentation to gain a deeper understanding of model performance under a variety of conditions.
 Specifically, CAT production performed in the absence of amino acids could inform the system's ability to manufacture them, while experimentation in the absence of glucose or oxygen could shed light on the importance of those substrates.
-Finally, the approach should be extended to other protein products.
+in addition, the approach should be extended to other protein products.
 CAT is only a test protein used for model identification; the modeling framework, and to some extent the parameter values, should be protein agnostic.
@@
 
-% The reaction rate was written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$), $r_{j}\left(\mathbf{x},\mathbf{k}\right)=\bar{r}_{j}v_{j}$.
-% In this study, we used either saturation or mass action kinetics.
-% The control term $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
-% For each rate, we used a rule-based approach to select from competing control factors.
-% If rate j was influenced by $1,\dots,m$ factors, we modeled this relationship as
-% $v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
-% where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
-% The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
-% variable. Each regulatory transfer function took the form:
-% \begin{equation}\label{eqn:control-factor}
-% 	f_{ij}\left(\mathcal{Z}_{i},k_{ij},\eta_{ij}\right)=k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}/\left({1 + k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}}\right)
-% \end{equation}where $\mathcal{Z}_{i}$ denotes the abundance factor $i$, $k_{ij}$ denotes a gain parameter, and $\eta_{ij}$ denotes a cooperativity parameter.
-% In this study, we used $\mathcal{I}_{j}\in\left\{mean\right\}$ \cite{pr3010138}. If a process has no modifying factors, $v_{j}=1$.
-% We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
-% \begin{equation}\label{eqn:rate-bar}
-% 	\bar{r}_{j}=k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
-% \end{equation}
-% where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
-% $K_{js}$ denotes the saturation constant for species $s$ in reaction $j$.
-% The product in Equation~\eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$).
-
 We included 17 allosteric regulation terms, taken from literature, in the CFPS model.
@@
 We estimated an ensemble of diverse parameter sets using a constrained Markov Chain Monte Carlo (MCMC) random walk strategy.
-Starting from a single best fit parameter set estimated by inspection and literature,
-we calculated the cost function, equal to the sum-squared-error between experimental data and model predictions:
+Starting from a single best-fit parameter set estimated by inspection and literature, we calculated the cost function, equal to the sum-squared-error between experimental data and model predictions:
 \begin{equation}\label{eqn:cost-function}
@@
 where $\mathcal{P}$ denotes the number of parameters ($\mathcal{P}$~=~815), which includes 163 maximum reaction rates ($V^{max}$), 163 enzyme activity decay constants, 455 saturation constants ($K_{js}$), and 34 control parameters, $k_i^{new}$ denotes the new value of the $i^{th}$ parameter, $k_i$ denotes the current value of the $i^{th}$ parameter, $a$ denotes a distribution variance, $r_i$ denotes a random sample from the normal distribution, $l_i$ denotes the lower bound for that parameter type, and $u_i$ denotes the upper bound for that parameter type.
-Maximum reaction rates were bounded between 0 and 500,000 mM/h \cite{Milo:2009aa}.
-Assuming a total enzyme concentration of 5.0 $\mu$M, this corresponds to catalytic rate bounds of 0 and 27,780 s$^{-1}$.
-These bounds resulted in a median catalytic rate of 0.16 s$^{-1}$ across the ensemble.
-Enzyme activity decay constants were bounded between 0 and 1 h$^{-1}$, corresponding to half lives of 42 minutes and infinity; median = 25 h.
-Saturation constants were bounded between 0.001 and 10 mM; median = 0.16 mM.
-Control parameters (gains and orders) were left unbounded; gain median = 0.076, order median = 0.69.
+
+%catalytic rate bounds of 0 and 27,780 s$^{-1}$.
+%\cite{Milo:2009aa}
+Model parameters were constrained by literature.
+Rate maxima were bounded within one order of magnitude of the literature value where available (Table~\ref{tbl:Rate_maxima}).
+All other rate maxima were bounded within two orders of magnitude of the geometric mean of the available values.
+The median maximum reaction rate was 7.8 mM/h; assuming a total cell-free enzyme concentration of 167 nM, this corresponds to a median catalytic rate of 0.08 s$^{-1}$ across the ensemble.
+Enzyme activity decay constants were bounded between 0 and 1 h$^{-1}$, corresponding to half lives of 42 minutes and infinity; median = 156 h.
+Saturation constants were bounded between 0.0001 and 10 mM; median = 1.0 mM.
+Control parameters (gains and orders) were bounded between 0.1 and 10 (dimensionless); median = 0.74.
 For each newly generated parameter set, we re-solved the balance equations and calculated the cost function.
 All sets with a lower cost (and some with higher cost) were accepted into the ensemble.
-After generating XXX sets, we selected N = 100 sets with minimal set to set correlation to avoid over-sampling any region of parameter space.
+After generating over 30,000 sets, N = 100 sets with minimal error were selected for the final ensemble.
+The final ensemble had a mean Pearson correlation coefficient of 0.77.
 
-\subsection*{Sensitivity analysis of the kinetic CFPS model.}
-We determined the reactions most important to protein production by computing the local sensitivity of CAT concentration (denoted as $\texttt{CAT}$) to each individual maximum reaction rate, and each pair of maximum reaction rates in the network.
-The sensitivity index was formulated as:
-\begin{equation}\label{eqn:CAT-sensitivity}
-    \mathcal{S}_{ij}^{\texttt{CAT}} = \norm{\texttt{CAT}(p_i,p_j,t)-\texttt{CAT}(\alpha \cdot p_i,\alpha \cdot p_j,t)}_{2}\qquad{i,j=1,2,\hdots\mathcal{P}}
-\end{equation}
-where $\mathcal{S}_{ij}^{\texttt{CAT}}$ denotes the sensitivity of CAT production to the $i^{th}$ and $j^{th}$ parameters,
-$\texttt{CAT}(p_i,p_j,t)$ denotes CAT concentration as a function of time and the $i^{th}$ and $j^{th}$ parameters,
-$\alpha$ denotes the perturbation factor, and $\mathcal{P}$ denotes the number of maximum reaction rates ($\mathcal{P}$ = 163).
-In calculating the pairwise sensitivities, each parameter was perturbed by 1\%; first-order sensitivities ($i$ = $j$) were subject to two 1\% perturbations.
-Parameters and parameter combinations were stratified into five degrees of importance, from least to most sensitive.
+\subsection*{Comparison against random ensemble.}
+A random ensemble of 100 parameter sets was generated from within the same parameter bounds as the trained ensemble.
+Sets were sampled using a Monte Carlo approach: each parameter was taken from a uniform distribution constructed between its upper and lower bounds.
+The model equations were then solved and the cost function was calculated in terms of the 37 separate experimental datasets.
+The random ensemble had a log median error of 0.80 across the datasets, as compared with a log median error of -1.43 for the trained ensemble (Fig.~\ref{fig:BoxPlot}).
+Thus, the trained ensemble fits the dataset over one hundred times better than a random ensemble generated within the same bounds.
 
-Likewise, we determined which reactions were most important to global system performance by computing the sensitivity of all species for which data exists (denoted as $\texttt{X}$)
-to each maximum reaction rate in the network.
-In this case, each sensitivity index was formulated as:
-\begin{equation}\label{eqn:system-sensitivity}
-    \mathcal{S}_{ij}^{\texttt{X}}=\norm{\texttt{X}(p_i,p_j,t)-\texttt{X}(\alpha \cdot p_i,\alpha \cdot p_j,t)}_{2}\qquad{i,j=1,2,\hdots\mathcal{P}}
-\end{equation}
-where $\mathcal{S}_{ij}^{\texttt{X}}$ denotes the sensitivity of the system state to the $i^{th}$ and $j^{th}$ parameters,
-and $\texttt{X}(p_i,p_j,t)$ denotes the system state, an array consisting of the concentration of every species for
-which data exists as a function of time and the $i^{th}$ and $j^{th}$ parameters.
-The parameter sensitivities were stratified into five degrees of importance, from least to most sensitive, as above.
-% \begin{eqnarray*}
-% G+R_{1} &\longrightarrow& G^{*}\\
-% G^{*}+151 \cdot gtp+144 \cdot ctp+189 \cdot utp+176 \cdot atp &\longrightarrow& mRNA+G+R_{1}+1320 \cdot Pi\\
-% mRNA &\longrightarrow& 151 \cdot gmp+144 \cdot cmp+189 \cdot ump+176 \cdot amp\\
-% \end{eqnarray*}
+\subsection*{Group knockouts.}
+The network was divided into 19 groups: glycolysis/gluconeogenesis, pentose phosphate, Entner-Doudoroff, TCA cycle, oxidative phosphorylation, cofactor reactions, anaplerotic/glyoxylate reactions, overflow metabolism, folate synthesis, purine/pyrimidine reactions, alanine/aspartate/asparagine synthesis, glutamate/glutamine synthesis, arginine/proline synthesis, glycine/serine synthesis, cysteine/methionine synthesis, threonine/lysine synthesis, histidine synthesis, tyrosine/tryptophan/phenylalanine synthesis, and valine/leucine/isoleucine synthesis.
+Each group of reactions was turned off individually, and then in pairs, and the model equations were re-solved.
+The CAT productivity was calculated and compared to that of the best-fit set (Fig.~\ref{fig:GroupKO}A).
+The absolute difference in productivity was recorded for each first-order knockout (diagonal elements) and each pairwise knockout, and a total-order coefficient was calculated by summing the first-order effect with all pairwise effects.
+Total-order coefficients were then normalized to fit within the same colorbar range as the first-order and pairwise effects.
+The system state was also calculated for each simulation, defined as the model predictions for all species for which data exist.
+The norm of the difference between the knockout system state and the best-fit system state is shown in (Fig.~\ref{fig:GroupKO}B).
 
-\subsubsection*{Quantification of uncertainty.}
-An ensemble of 100 sets of flux distributions was calculated for three different cases:  constrained by transcription/translation rates, constrained by transcription/translation rates without amino acid synthesis reactions, and constrained by transcription/translation rates and experimental measurements without amino acid synthesis reactions.
-For the first case, all rates were left unbounded, except the specific glucose uptake rate, transcription and translation rate.
-An ensemble of flux distributions was then calculated by randomly sampling the maximum specific glucose uptake rate from within a range of 30 to 40 mM/h, determined from experimental data and randomly sampling RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature..
-For the second case, an ensemble was generated by randomly sampling the same parameters as the first case, however certain amino acid synthesis reactions were removed from the network.
-This included all the amino acids that were present in the preparation of the \textit{E.
-coli} extract (alanine, arginine, aspartate, cysteine, glutamate, glutamine and serine were excluded from the media), thus reactions producing the excluded amino acids were left in the network.
-RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/sec \cite{2005_underwood_biotech, Garamella:2016aa}.
-For the third case, the ensemble was generated as in the second case, in addition to the lower and upper bounds on the fluxes for the data-informed metabolites were sampled within the range given by the experimental noise.
-This included the data for glucose, organic acids, energy species, and amino acids; CAT was not constrained by experimental data, but by the transcription/translation rates as stated above.
+%\subsection*{Sensitivity analysis of the kinetic CFPS model.}
+%We determined the reactions most important to protein production by computing the local sensitivity of CAT concentration (denoted as $\texttt{CAT}$) to each individual maximum reaction rate, and each pair of maximum reaction rates in the network.
+%The sensitivity index was formulated as:
+%\begin{equation}\label{eqn:CAT-sensitivity}
+%    \mathcal{S}_{ij}^{\texttt{CAT}} = \norm{\texttt{CAT}(p_i,p_j,t)-\texttt{CAT}(\alpha \cdot p_i,\alpha \cdot p_j,t)}_{2}\qquad{i,j=1,2,\hdots\mathcal{P}}
+%\end{equation}
+%where $\mathcal{S}_{ij}^{\texttt{CAT}}$ denotes the sensitivity of CAT production to the $i^{th}$ and $j^{th}$ parameters,
+%$\texttt{CAT}(p_i,p_j,t)$ denotes CAT concentration as a function of time and the $i^{th}$ and $j^{th}$ parameters,
+%$\alpha$ denotes the perturbation factor, and $\mathcal{P}$ denotes the number of maximum reaction rates ($\mathcal{P}$ = 163).
+%In calculating the pairwise sensitivities, each parameter was perturbed by 1\%; first-order sensitivities ($i$ = $j$) were subject to two 1\% perturbations.
+%Parameters and parameter combinations were stratified into five degrees of importance, from least to most sensitive.
+
+%Likewise, we determined which reactions were most important to global system performance by computing the sensitivity of all species for which data exist (denoted as $\texttt{X}$)
+%to each maximum reaction rate in the network.
+%In this case, each sensitivity index was formulated as:
+%\begin{equation}\label{eqn:system-sensitivity}
+%    \mathcal{S}_{ij}^{\texttt{X}}=\norm{\texttt{X}(p_i,p_j,t)-\texttt{X}(\alpha \cdot p_i,\alpha \cdot p_j,t)}_{2}\qquad{i,j=1,2,\hdots\mathcal{P}}
+%\end{equation}
+%where $\mathcal{S}_{ij}^{\texttt{X}}$ denotes the sensitivity of the system state to the $i^{th}$ and $j^{th}$ parameters,
+%and $\texttt{X}(p_i,p_j,t)$ denotes the system state, an array consisting of the concentration of every species for
+%which data exist as a function of time and the $i^{th}$ and $j^{th}$ parameters.
+%The parameter sensitivities were stratified into five degrees of importance, from least to most sensitive, as above.
+%% \begin{eqnarray*}
+%% G+R_{1} &\longrightarrow& G^{*}\\
+%% G^{*}+151 \cdot gtp+144 \cdot ctp+189 \cdot utp+176 \cdot atp &\longrightarrow& mRNA+G+R_{1}+1320 \cdot Pi\\
+%% mRNA &\longrightarrow& 151 \cdot gmp+144 \cdot cmp+189 \cdot ump+176 \cdot amp\\
+%% \end{eqnarray*}
 
@@
 $\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed (never allowed to be negative), and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
-Arginine and glutamate were excluded from the yield calculations, as no experimental measurements were available for these amino acids.
-Yield of the best-fit parameter set and the experimental data were calculated by setting $\Delta\texttt{CAT}$ equal to the final minus the initial CAT concentration,
-and setting $\Delta m_{i}$ equal to the initial minus the final reactant concentration.
-The individual CAT production and substrate consumption terms for the best-fit set, kinetic models with knockouts, and experimental data are shown in Table~\ref{tbl:yield_breakdown}.
-Total net consumption of amino acids and amino acid consumption via CAT synthesis were calculated for the best-fit set (Table~\ref{tbl:AA_breakdown}).
-Total net consumption was calculated as amino acid concentration at 0 hours minus concentration at 3 hours; it was negative if synthesis outweighed consumption.
-Consumption toward CAT was calculated as CAT concentration at 3 hours minus concentration at 0 hours, times the stoichiometric coefficient for that amino acid in the CAT synthesis reaction.
-The difference between these was defined as other consumption, equal to consumption from reactions other than CAT synthesis minus amino acid production.
+This analysis was extended to the accumulation of amino acids, organic acids, and other byproducts, to create a complete carbon balance through the network (Fig.~\ref{fig:CAT_balances}A).
+%Yield of the best-fit parameter set and the experimental data were calculated by setting $\Delta\texttt{CAT}$ equal to the final minus the initial CAT concentration,
+%and setting $\Delta m_{i}$ equal to the initial minus the final reactant concentration.
+%The individual CAT production and substrate consumption terms for the best-fit set, kinetic models with knockouts, and experimental data are shown in Table~\ref{tbl:yield_breakdown}.
+%Total net consumption of amino acids and amino acid consumption via CAT synthesis were calculated for the best-fit set (Table~\ref{tbl:AA_breakdown}).
+%Total net consumption was calculated as amino acid concentration at 0 hours minus concentration at 3 hours; it was negative if synthesis outweighed consumption.
+%Consumption toward CAT was calculated as CAT concentration at 3 hours minus concentration at 0 hours, times the stoichiometric coefficient for that amino acid in the CAT synthesis reaction.
+%The difference between these was defined as other consumption, equal to consumption from reactions other than CAT synthesis minus amino acid production.
 
@@
 $\rm {ATP}_{TX}$ = 176, $\rm {CTP}_{TX}$ = 144, $\rm {GTP}_{TX}$ = 151, $\rm {UTP}_{TX}$ = 189, $\rm {ATP}_{TL}$ = 219, $\rm {GTP}_{TL}$ = 438, $\rm {ATP}_{GLC}$ = 15.
+This analysis was also extended to the accumulation of amino acids, organic acids, and other byproducts to create a complete energy balance through the network (Fig.~\ref{fig:CAT_balances}B).
+Equivalent ATP numbers for glucose, amino acids, and organic acids were calculated from the network stoichiometry.
 
@@
 J.V directed the modeling study.
-K.C and J.S conducted the cell free protein synthesis experiments.
-J.V, J.W, and N.H developed the cell free protein synthesis mathematical model, and parameter ensemble.
+K.C and J.S conducted the cell-free protein synthesis experiments.
+J.V, J.W, and N.H developed the cell-free protein synthesis mathematical model, and parameter ensemble.
 The manuscript was prepared and edited for publication by J.S, N.H, M.V, J.W and J.V.
@@
 % \begin{tabular}{llrrr} \toprule
-% & \multicolumn{1}{r}{\textbf{Original value}} & \multicolumn{2}{r}{\textbf{After dilution factor}} & \textbf{BN ID} \\ \hline
+% & \multicolumn{1}{r}{\textbf{Original value}} & \multicolumn{2}{r}{\textbf{After dilution factor}} & \textbf{Reference} \\ \hline
 %\smallskip
@@
 %\phantom{i}\textbf{Enzyme} & \textbf{Reaction name} & \textbf{Kcat (h$^-1$)} & 
-% \textbf{Vmax (mM/h)} & \textbf{BN ID} \\ \hline
+% \textbf{Vmax (mM/h)} & \textbf{Reference} \\ \hline
 %\phantom{i}Serine dehydrase & R\_ser\_deg & 624000 & 104 & 101119 \\ \hline
@@
 %\phantom{i}Geometric mean & & & 110 & \\ \toprule
-%\phantom{i}\textbf{Transcription/Translation} & \textbf{AA-tRNA complex (mM)} & \textbf{Kcat (h$^-1$)} & \textbf{Vmax (mM/h)} & \textbf{BN ID} \\ \hline
+%\phantom{i}\textbf{Transcription/Translation} & \textbf{AA-tRNA complex (mM)} & \textbf{Kcat (h$^-1$)} & \textbf{Vmax (mM/h)} & \textbf{Reference} \\ \hline
 %\phantom{i}tRNA charging & 0.03 & 14040 & 4212 & 104980 \\ \hline
@@
 % \end{tabular}
-%\label{tbl:Literature_values}
+%\label{tbl:Rate_maxima}
 %\end{table}
+
+\begin{table}
+\centering
+ \caption{Reference values for reaction rate maxima (V$_{max}$) from literature.
+V$_{max}$ values calculated from original turnover numbers (K$_{cat}$) and characteristic enzyme concentration, extrapolated to cell-free using a dilution factor of 30.
+Characteristic rate maximum for all other reactions calculated as geometric mean of calculated rate maxima.}
+ \renewcommand{\arraystretch}{1.3}
+ \begin{tabular}{llrrr} \toprule
+\smallskip
+ & & \parbox{2.1 cm}{\textbf{Original value ($\mu$M)}} & \parbox{3.1 cm}{\textbf{Cell-free \phantom{abc} equivalent (nM)}} & \textbf{Reference} \\ \hline
+\smallskip
+\rule{0pt}{3ex} \parbox{4.2 cm}{Characteristic enzyme concentration} & & 5 & 167 & BNID 100735 \\ \toprule
+\phantom{i}\textbf{Enzyme} & \textbf{Reaction} & \textbf{K$_{cat}$ (min$^{-1}$)} & 
+ \textbf{V$_{max}$ (mM/h)} & \textbf{Reference} \\ \hline
+\phantom{i}Serine dehydrase & R\_ser\_deg & 10400 & 104 & BNID 101119 \\ \hline
+\phantom{i}Isocitrate dehydrogenase & R\_icd & 11900 & 119 & BNID 101152 \\ \hline
+\phantom{i}Lactate dehydrogenase & R\_ldh & 5800 & 58 & BNID 101036 \\ \hline
+\smallskip
+\rule{0pt}{4.5ex}
+Aspartate transaminase & \parbox{1 cm}{R\_aspC, R\_tyr, R\_phe} & 25800 & 258 & BNID 101108 \\ \hline
+\phantom{i}Enolase & R\_eno & 13200 & 132 & BNID 101028 \\ \hline
+\smallskip
+\rule{0pt}{3ex}
+Pyruvate kinase & R\_pyk & 25000 & 250 & \parbox{2.52 cm}{BNID 101029 BNID 101030} \\ \hline
+\smallskip
+\rule{0pt}{3ex}
+Malic enzyme & \parbox{1 cm}{R\_maeA, R\_maeB} & 35400 & 354 & BNID 101167 \\ \hline
+\phantom{i}Phosphofructokinase & R\_pfk & 554400 & 5544 & BNID 104955 \\ \hline
+\phantom{i}Malate dehydrogenase & R\_mdh & 33000 & 330 & BNID 101163 \\ \hline
+\phantom{i}Citrate Synthase & R\_gltA & 42000 & 420 & BNID 101149 \\ \hline
+\smallskip
+\rule{0pt}{4.5ex}
+6PG dehydrogenase & \parbox{1 cm}{R\_zwf, R\_pgl, R\_gnd} &3200 & 32 & BNID 101048 \\ \hline
+\phantom{i}Succinate dehydrogenase & R\_sdh & 121 & 1.21 & BNID 101162 \\ \hline
+\phantom{i}Succinyl-coA synthetase & R\_sucCD & 4700 & 47 & BNID 101158 \\ \hline
+\phantom{i}3PGA dehydrogenase & R\_gpm & 1100 & 11 & BNID 101135 \\ \hline
+\phantom{i}PEP carboxylase & R\_ppc & 35400 & 354 & BNID 101139 \\ \hline
+\phantom{i}3PGA kinase & R\_pgk & 4300 & 43 & BNID 101016 \\ \midrule
+\phantom{i}Characteristic rate maximum & & & 110 & \\ \toprule
+\phantom{i}\textbf{Transcription/Translation} & & & & \\ \hline
+%\phantom{i}\textbf{Transcription/Translation} & \textbf{AA-tRNA (mM)} & \textbf{Kcat (h$^-1$)} & \textbf{Vmax (mM/h)} & \textbf{Reference} \\ \hline
+\phantom{i}tRNA charging & 0.03 & 14040 & 4212 & BNID 104980 \\ \hline
+% \bottomrule
+ \end{tabular}
+\label{tbl:Rate_maxima}
+\end{table}
 
@@
  \renewcommand{\arraystretch}{1.3}
- \begin{tabular}{llrrr} \toprule
-\smallskip
- & & \textbf{Original value} & \parbox{2.8 cm}{\textbf{After cell-free dilution factor}} & \textbf{BN ID} \\ \hline
-\smallskip
-\rule{0pt}{3ex} \parbox{4.3 cm}{Characteristic Enzyme Concentration} & & 5 $\mu$M & 167 nM & 100735 \\ \toprule
-\phantom{i}\textbf{Enzyme} & \textbf{Reaction} & \textbf{Kcat (min$^{-1}$)} & 
- \textbf{Vmax (mM/h)} & \textbf{BN ID} \\ \hline
-\phantom{i}Serine dehydrase & R\_ser\_deg & 10400 & 104 & 101119 \\ \hline
-\phantom{i}Isocitrate dehydrogenase & R\_icd & 11900 & 119 & 101152 \\ \hline
-\phantom{i}Lactate dehydrogenase & R\_ldh & 5800 & 58 & 101036 \\ \hline
-\smallskip
-\rule{0pt}{4.5ex}
-Aspartate transaminase & \parbox{1 cm}{R\_aspC, R\_tyr, R\_phe} & 25800 & 258 & 101108 \\ \hline
-\phantom{i}Enolase & R\_eno & 13200 & 132 & 101028 \\ \hline
-\smallskip
-\rule{0pt}{3ex}
-Pyruvate kinase & R\_pyk & 25000 & 250 & \parbox{1.4 cm}{101029, 101030} \\ \hline
-\smallskip
-\rule{0pt}{3ex}
-Malic enzyme & \parbox{1 cm}{R\_maeA, R\_maeB} & 35400 & 354 & 101167 \\ \hline
-\phantom{i}Phosphofructokinase & R\_pfk & 554400 & 5544 & 104955 \\ \hline
-\phantom{i}Malate dehydrogenase & R\_mdh & 33000 & 330 & 101163 \\ \hline
-\phantom{i}Citrate Synthase & R\_gltA & 42000 & 420 & 101149 \\ \hline
-\smallskip
-\rule{0pt}{4.5ex}
-6PG dehydrogenase & \parbox{1 cm}{R\_zwf, R\_pgl, R\_gnd} &3200 & 32 & 101048 \\ \hline
-\phantom{i}Succinate dehydrogenase & R\_sdh & 121 & 1.21 & 101162 \\ \hline
-\phantom{i}Succinyl-coA synthetase & R\_sucCD & 4700 & 47 & 101158 \\ \hline
-\phantom{i}3PGA dehydrogenase & R\_gpm & 1100 & 11 & 101135 \\ \hline
-\phantom{i}PEP carboxylase & R\_ppc & 35400 & 354 & 101139 \\ \hline
-\phantom{i}3PGA kinase & R\_pgk & 4300 & 43 & 101016 \\ \midrule
-\phantom{i}Geometric mean & & & 110 & \\ \toprule
-\phantom{i}\textbf{Transcription/Translation} \\ \hline
-%\phantom{i}\textbf{Transcription/Translation} & \textbf{AA-tRNA (mM)} & \textbf{Kcat (h$^-1$)} & \textbf{Vmax (mM/h)} & \textbf{BN ID} \\ \hline
-\phantom{i}tRNA charging & 0.03 & 14040 & 4212 & 104980 \\ \hline
+ \begin{tabular}{lrr} \toprule
+& \textbf{Volume (L)} & \textbf{Reference} \\ \hline
+\phantom{i}Cellular volume & 3.881e-16 & XXX \\ \hline
+ \toprule
+ & \textbf{Concentration (mM)} & \textbf{Reference} \\ \hline
+\phantom{i}CAT gene & 5.19e-6 & XXX \\ \hline
+\phantom{i}RNA polymerase & 6.75e-5 & \cite{Garamella:2016aa} \\ \hline
+\phantom{i}Ribosome & 2e-3 & \cite{Garamella:2016aa} \\ \hline
+\phantom{i}tRNA & 1.6 & BNID 108611 \\ \hline
+\phantom{i}NAD & 1.47 & XXX \\ \hline
+\phantom{i}NADH & 0.1 & XXX \\ \hline
+\phantom{i}NADP & 0.195 & XXX \\ \hline
+\phantom{i}NADPH & 0.062 & XXX \\ \hline
+\phantom{i}G6P & 3.48 & XXX \\ \hline
+\phantom{i}F6P & 0.6 & XXX \\ \hline
+\phantom{i}FDP & 0.272 & XXX \\ \hline
+\phantom{i}DHAP & 0.167 & XXX \\ \hline
+\phantom{i}G3P & 0.218 & XXX \\ \hline
+\phantom{i}13DPG & 0.008 & XXX \\ \hline
+\phantom{i}3PG & 2.13 & XXX \\ \hline
+\phantom{i}2PG & 0.399 & XXX \\ \hline
+\phantom{i}6PGL & 0.808 & XXX \\ \hline
+\phantom{i}6PGC & 0.808 & XXX \\ \hline
+\phantom{i}RU5P & 0.111 & XXX \\ \hline
+\phantom{i}XU5P & 0.138 & XXX \\ \hline
+\phantom{i}R5P & 0.398 & XXX \\ \hline
+\phantom{i}S7P & 0.276 & XXX \\ \hline
+\phantom{i}E4P & 0.098 & XXX \\ \hline
+\phantom{i}NH3 & 0.052425 & XXX \\ \hline
+\phantom{i}ARG & 1.9326 & XXX \\ \hline
+\phantom{i}H2O2 & 0.052425 & XXX \\ \hline
+\phantom{i}HCO3 & 0.052425 & XXX \\ \hline
 % \bottomrule
  \end{tabular}
-\label{tbl:Literature_values}
+\label{tbl:Initial_conditions}
 \end{table}
@@
 % \begin{tabular}{llrrr} \toprule
-% & & \textbf{Original value} & \textbf{After dilution factor} & \textbf{BN ID} \\ \hline
+% & & \textbf{Original value} & \textbf{After dilution factor} & \textbf{Reference} \\ \hline
 %\smallskip
@@
 %\phantom{i}\textbf{Enzyme} & \textbf{Reaction} & \textbf{Kcat (h$^-1$)} & 
-% \textbf{Vmax (mM/h)} & \textbf{BN ID} \\ \hline
+% \textbf{Vmax (mM/h)} & \textbf{Reference} \\ \hline
 %\phantom{i}Serine dehydrase & R\_ser\_deg & 624000 & 104 & 101119 \\ \hline
@@
 %\phantom{i}Geometric mean & & & 110 & \\ \toprule
-%%\phantom{i}\textbf{Transcription/Translation} & \textbf{AA-tRNA (mM)} & \textbf{Kcat (h$^-1$)} & \textbf{Vmax (mM/h)} & \textbf{BN ID} \\ \hline
+%%\phantom{i}\textbf{Transcription/Translation} & \textbf{AA-tRNA (mM)} & \textbf{Kcat (h$^-1$)} & \textbf{Vmax (mM/h)} & \textbf{Reference} \\ \hline
 %\phantom{i}tRNA charging & 0.03 & 14040 & 4212 & 104980 \\ \hline
@@
 % \end{tabular}
-%\label{tbl:Literature_values}
+%\label{tbl:Rate_maxima}
 %\end{table}
@@
 \centering
-\includegraphics[width=0.6\textwidth]{./Figures/Fig_6_CarbonEnergyBalances.pdf}
+\includegraphics[width=0.55\textwidth]{./Figures/Fig_6_CarbonEnergyBalances.pdf}
 %\caption{Carbon and energy balances for the best-fit set.

### 2017-05-15 Paper_v1.tex
May15

@@
 In this study, we developed an ensemble of dynamic \textit{E. coli} cell-free protein synthesis (CFPS) models.
-Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids, and the protein product, chloramphenicol acetyltransferase (CAT).
-The ensemble captured all of the training data, especially the central carbon metabolism.
-CAT was produced with a carbon yield of 7\% and an energy efficiency of 5\%, suggesting that CAT production could be further optimized.
+Model parameters were estimated from a training dataset for the cell-free production of a protein product, chloramphenicol acetyltransferase (CAT).
+%This protein was used as a proof-of-concept
+The dataset consisted of measurements of glucose, organic acids, energy species, amino acids, and CAT.
+The ensemble accurately predicted these measurements, especially those of the central carbon metabolism.
+We then used the trained model to evaluate the optimality of protein production.
+CAT was produced with a carbon yield of 7\% and an energy efficiency of 5\%, suggesting that the process could be further optimized.
 Reaction group knockouts showed that protein productivity and the metabolism as a whole depend most on oxidative phosphorylation and glycolysis and gluconeogenesis.
@@
 In addition, CAT production was robust to allosteric control, as was most of the network, with the exception of the organic acids in central carbon metabolism.
-This study is the first to model dynamic protein production in \textit{E. coli}, and should provide a foundation for genome scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.
+This study is the first to use kinetic modeling to predict dynamic protein production in a cell-free \textit{E.
+coli} system, and should provide a foundation for genome scale, dynamic modeling of cell-free \textit{E.
+coli} protein synthesis.
 
@@
 However, if CFPS is to become a mainstream technology for applications such as point of care manufacturing, we must first understand the performance limits of these systems.
-One tool to address this question is mathematical modeling.
+%One tool to address this question is mathematical modeling.
+One tool we can use to achieve this understanding is mathematical modeling.
 
@@
 coli} \citep{2008_atlas_shuler_IETSysBio}.
-In the post genomics world, large-scale stoichiometric reconstructions of microbial metabolism popularized by techniques such as flux balance analysis (FBA) have become a standard approach \citep{2012_lewis_palsson_NatRevMicrobio}.
-Since the first genome scale stoichiometric model of \textit{E.
+%In the post genomics world, large-scale stoichiometric reconstructions of microbial metabolism popularized by techniques such as flux balance analysis (FBA) have become a standard approach \citep{2012_lewis_palsson_NatRevMicrobio}.
+%Since the first genome scale stoichiometric model of \textit{E.
 coli} developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, well over 100 organisms including industrially important prokaryotes are now available \citep{2009_feist_palsson_NatRevMicrobio,Feist:2007aa,Oh:2007aa}.
-Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome scale kinetic models to an underdetermined linear algebraic system, which can be solved efficiently even for large systems.
-Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
+%Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome scale kinetic models to an underdetermined linear algebraic system, which can be solved efficiently even for large systems.
+%Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
 Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \citep{2002_ibarra_edwards_palsson_Nat}, although other metabolic objectives have also been estimated \citep{2007_schuetz_sauer_MolSysBio}.
-Recent advances in constraint-based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \citep{2013_hyduke_lewis_palsson_MolBioSys}.
-Thus, modern constraint-based approaches have proven extremely useful in the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
+%Recent advances in constraint-based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \citep{2013_hyduke_lewis_palsson_MolBioSys}.
+%Thus, modern constraint-based approaches have proven extremely useful in the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
 However, genome scale kinetic models of industrial important organisms such as \textit{E. coli} have yet to be constructed.
@@
 Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids, and the protein product, chloramphenicol acetyltransferase (CAT).
-Characteristic values for model parameters and metabolite initial conditions, estimated from literature, were used to constrain the parameter estimation problem.
+%Characteristic values for model parameters and metabolite initial conditions, estimated from literature, were used to constrain the parameter estimation problem.
+The parameter estimation problem was constrained by characteristic values for model parameters and metabolite initial conditions estimated from literature.
 The ensemble of parameter sets described the training data with a median cost that was more than two orders of magnitude smaller than random sets constructed using the literature parameter constraints.
-We then used the ensemble of kinetic models to analyze the CFPS reaction.
-By knocking out metabolic reactions in groups, we showed that CAT productivity depends most on oxidative phosphorylation, glycolysis/gluconeogenesis, and certain amino acid synthesis reactions, while the system state is most sensitive to glycolysis/gluconeogenesis, oxidative phosphorylation, the overflow metabolism, and the TCA cycle reactions.
+We then used the ensemble of kinetic models to analyze the optimality of the CFPS reaction, and the pathways most important to CAT production.
+We calculated that CAT was produced with a carbon yield of 7\% and an energy efficiency of 5\%, suggesting that much of the resources for protein synthesis were diverted to non-productive pathways.
+By knocking out metabolic enzymes in groups, we showed that CAT productivity depends most on oxidative phosphorylation, glycolysis/gluconeogenesis, and certain amino acid synthesis reactions, while the system state is most sensitive to glycolysis/gluconeogenesis, oxidative phosphorylation, the overflow metabolism, and the TCA cycle reactions.
 %First, sensitivity analysis of the dynamic model suggested that CAT production was most sensitive to CAT synthesis parameters, as well as reactions in oxidative phosphorylation and pyruvate consumption.
@@
 
-The ensemble of kinetic CFPS models captured the time evolution of CAT biosynthesis (Fig.~\ref{fig:Carbon} - \ref{fig:Amino}).
 The cell-free \textit{E.~coli} metabolic network was constructed by removing growth associated reactions from the \textit{i}AF1260 reconstruction of K-12 MG1655 \textit{E.~coli} \cite{Feist:2007aa}, and by adding reactions describing chloramphenicol acetyltransferase (CAT) biosynthesis, a model protein for which there exists a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
@@
 
-Central carbon metabolism (Fig.~\ref{fig:Carbon}, top), energy species (Fig.~\ref{fig:Energy}), and amino acids (Fig.~\ref{fig:Amino}) were captured by the ensemble and the best-fit set.
+The ensemble of kinetic CFPS models captured the time evolution of CAT biosynthesis.
+%(Fig.~\ref{fig:Carbon} - \ref{fig:Amino}).
+Central carbon metabolites (Fig.~\ref{fig:Carbon}, top), energy species (Fig.~\ref{fig:Energy}), and amino acids (Fig.~\ref{fig:Amino}) were captured by the ensemble and the best-fit set.
 The constrained MCMC approach estimated parameter sets with a median error more than two orders of magnitude less than random parameter sets generated within the same parameter bounds (Fig.~\ref{fig:BoxPlot}); thus, we have confidence in the predictive capability of the estimated parameters.
@@
 The network was divided into 19 groups of reactions, spanning central carbon metabolism, energetics, and amino acid biosynthesis.
-The reactions in each of these groups were turned off, and the resulting change in productivity and system state were recorded.
+%The reactions in each of these groups were turned off, 
+The enzymes in each of these groups were knocked out, and the resulting change in productivity and system state were recorded.
 Then each pair of groups was knocked out to determine pairwise effects.
@@
 Another 7\% went to the accumulation of alanine and glutamine, in line with experimental data (these were not included in the amino acid consumption term).
-25\% of the carbon was accumulated as organic acids (lactate, acetate, succinate, and malate), leaving 61\% as accumulation of other byproducts, species for which data do not exist.
+25\% of the carbon was accounted for by the net accumulation of organic acids (only lactate, acetate, succinate, and malate showed positive net accumulation), leaving 61\% as accumulation of other byproducts, species for which data do not exist.
 CFPS also showed an energy efficiency of 5\%; this was formulated as the equivalent ATP value of CAT produced divided by the equivalent ATP value of glucose consumed.
@@
 The ATP number for CAT was simply equal to the cost of transcription and translation.
-These results show that there is much room for improvement of the efficiency of CFPS.
-This demonstrates that CAT carbon yield could be improved by diverting flux away from some of the less efficient or altogether unnecessary metabolic pathways.
+%These results show 
+%The fact that so much carbon and energy are spent on non-productive pathways
+The low carbon yield and energy efficiency show that there is much room for improvement; CFPS efficiency could be increased by diverting flux away from some of the less efficient or altogether unnecessary metabolic pathways.
 %pathways of energy generation
-%A key finding of both the CAT and overall system state sensitivity analysis was the importance of oxidative phosphorylation.
 
@@
 This is because the amino acids and energy species used in CAT synthesis are not dependent on allosteric control.
+
 Having captured the experimental data, we investigated if CAT yield and CFPS performance could be further improved.
@@
 %Testing the model under a variety of conditions could strengthen or challenge the findings of this study.
-
-In this study we present an ensemble of \textit{E.
-coli} cell-free protein synthesis (CFPS) models that accurately predict a comprehensive CFPS dataset of glucose, CAT, central carbon metabolites, energy species, and amino acid measurements.
-
-Our ensemble of models accurately predicts dynamic experimental measurements of central carbon metabolism, energy species, and amino acids over 100 times better than random sets in the same region of parameter space.
-
-CFPS is seen to be biphasic, relying on glucose during the first hour and pyruvate and lactate afterward.
-Allosteric control was essential to the maintenance of the network, specifically the central carbon metabolism.
-% and production of CAT.
-% central carbon metabolism is exhuasted within 1.5 hours, leading to low CAT production.
-Without it, pyruvate, succinate, and malate are consumed more quickly following glucose exhaustion, presumably to power downstream reactions and ultimately CAT synthesis.
-Meanwhile, acetate accumulation during the increased after the first hour, suggesting a decrease in sytem efficiency.
-Interestingly, CAT production is virtually unaffected.
-This is because the amino acids and energy species used in CAT synthesis are not dependent on allosteric control.
-Having captured the experimental data, we investigated if CAT yield and CFPS performance could be further improved.
-We showed that the model predicts CAT production with a carbon yield of 7\% and energy efficiency of 5\%.
-%We showed that the model produces CAT with a carbon yield equal to 16\% of that of a physiological case in which transcription and translation are constrained, and an energy efficiency equal to 9\% of that of the physiological case.
-The accumulation of waste byproducts, especially acetate and carbon dioxide, is responsible for this sub-optimal performance.
-However, there is also some accumulation of useful metabolites in the central carbon metabolism; if these could be utilized fully, CAT production would likely increase.
-Knocking out sections of network metabolism revealed that glycolysis/gluconeogenesis and oxidative phosphorylation were the most important to CAT production and the system as a whole.
-Productivity was also heavily dependent on the synthesis reactions of alanine, aspartate, asparagine, glutamate, and glutamine, while TCA cycle and overflow reactions affected the system state.
-Taken together, these findings represent the first dynamic model of \textit{E. coli} cell-free protein synthesis, and an important step toward a functional genome scale description.
 
@@
 %catalytic rate bounds of 0 and 27,780 s$^{-1}$.
-%\cite{Milo:2009aa}
-Model parameters were constrained by literature.
-Rate maxima were bounded within one order of magnitude of the literature value where available (Table~\ref{tbl:Rate_maxima}).
-All other rate maxima were bounded within two orders of magnitude of the geometric mean of the available values.
+Model parameters were constrained by literature \cite{Milo:2009aa}.
+A characteristic cell-free enzyme concentration of 167 nM was calculated by diluting the one-tenth maximal amount \textit{lacZ} (5 $\mu$M, BNID 100735) by  a cell-free dilution factor of 30.
+This enzyme level was then used to calculate rate maxima from turnover numbers for various enzymes from literature (Table~\ref{tbl:Rate_maxima}).
+Rate maxima were bounded within one order of magnitude of the calculated value where available; all other rate maxima were bounded within two orders of magnitude of the geometric mean of the available values.
 The median maximum reaction rate was 7.8 mM/h; assuming a total cell-free enzyme concentration of 167 nM, this corresponds to a median catalytic rate of 0.08 s$^{-1}$ across the ensemble.
@@
  \caption{Reference values for reaction rate maxima (V$_{max}$) from literature.
-V$_{max}$ values calculated from original turnover numbers (K$_{cat}$) and characteristic enzyme concentration, extrapolated to cell-free using a dilution factor of 30.
+V$_{max}$ values calculated from turnover numbers (K$_{cat}$) from literature, and a characteristic enzyme concentration of 167 nM.
 Characteristic rate maximum for all other reactions calculated as geometric mean of calculated rate maxima.}
@@
  \begin{tabular}{llrrr} \toprule
-\smallskip
- & & \parbox{2.1 cm}{\textbf{Original value ($\mu$M)}} & \parbox{3.1 cm}{\textbf{Cell-free \phantom{abc} equivalent (nM)}} & \textbf{Reference} \\ \hline
-\smallskip
-\rule{0pt}{3ex} \parbox{4.2 cm}{Characteristic enzyme concentration} & & 5 & 167 & BNID 100735 \\ \toprule
+%\smallskip
+% & & \parbox{2.1 cm}{\textbf{Original value ($\mu$M)}} & \parbox{3.1 cm}{\textbf{Cell-free \phantom{abc} equivalent (nM)}} & \textbf{Reference} \\ \hline
+%\smallskip
+%\rule{0pt}{3ex} \parbox{4.2 cm}{Characteristic enzyme concentration} & & 5 & 167 & BNID 100735 \\ \toprule
 \phantom{i}\textbf{Enzyme} & \textbf{Reaction} & \textbf{K$_{cat}$ (min$^{-1}$)} & 
@@
 \phantom{i}Characteristic rate maximum & & & 110 & \\ \toprule
-\phantom{i}\textbf{Transcription/Translation} & & & & \\ \hline
-%\phantom{i}\textbf{Transcription/Translation} & \textbf{AA-tRNA (mM)} & \textbf{Kcat (h$^-1$)} & \textbf{Vmax (mM/h)} & \textbf{Reference} \\ \hline
-\phantom{i}tRNA charging & 0.03 & 14040 & 4212 & BNID 104980 \\ \hline
+\phantom{i}\textbf{Transcription/Translation} & \textbf{Reaction} & \textbf{K$_{cat}$ (min$^{-1}$)} & \textbf{V$_{max}$ (mM/h)} & \textbf{Reference} \\ \hline
+\phantom{i}mRNA degradation & 5.2 (1/h) & 0.08667 & 0.00052 & BNID 104980 \\ \hline
+\phantom{i}tRNA charging & 0.03 & 2340 & 4212 & BNID 104980 \\ \hline
 % \bottomrule
@@
 
-\begin{table}
-\centering
- \caption{}
- \renewcommand{\arraystretch}{1.3}
- \begin{tabular}{lrr} \toprule
-& \textbf{Volume (L)} & \textbf{Reference} \\ \hline
-\phantom{i}Cellular volume & 3.881e-16 & XXX \\ \hline
- \toprule
- & \textbf{Concentration (mM)} & \textbf{Reference} \\ \hline
-\phantom{i}CAT gene & 5.19e-6 & XXX \\ \hline
-\phantom{i}RNA polymerase & 6.75e-5 & \cite{Garamella:2016aa} \\ \hline
-\phantom{i}Ribosome & 2e-3 & \cite{Garamella:2016aa} \\ \hline
-\phantom{i}tRNA & 1.6 & BNID 108611 \\ \hline
-\phantom{i}NAD & 1.47 & XXX \\ \hline
-\phantom{i}NADH & 0.1 & XXX \\ \hline
-\phantom{i}NADP & 0.195 & XXX \\ \hline
-\phantom{i}NADPH & 0.062 & XXX \\ \hline
-\phantom{i}G6P & 3.48 & XXX \\ \hline
-\phantom{i}F6P & 0.6 & XXX \\ \hline
-\phantom{i}FDP & 0.272 & XXX \\ \hline
-\phantom{i}DHAP & 0.167 & XXX \\ \hline
-\phantom{i}G3P & 0.218 & XXX \\ \hline
-\phantom{i}13DPG & 0.008 & XXX \\ \hline
-\phantom{i}3PG & 2.13 & XXX \\ \hline
-\phantom{i}2PG & 0.399 & XXX \\ \hline
-\phantom{i}6PGL & 0.808 & XXX \\ \hline
-\phantom{i}6PGC & 0.808 & XXX \\ \hline
-\phantom{i}RU5P & 0.111 & XXX \\ \hline
-\phantom{i}XU5P & 0.138 & XXX \\ \hline
-\phantom{i}R5P & 0.398 & XXX \\ \hline
-\phantom{i}S7P & 0.276 & XXX \\ \hline
-\phantom{i}E4P & 0.098 & XXX \\ \hline
-\phantom{i}NH3 & 0.052425 & XXX \\ \hline
-\phantom{i}ARG & 1.9326 & XXX \\ \hline
-\phantom{i}H2O2 & 0.052425 & XXX \\ \hline
-\phantom{i}HCO3 & 0.052425 & XXX \\ \hline
-% \bottomrule
- \end{tabular}
-\label{tbl:Initial_conditions}
-\end{table}
+%\begin{table}
+%\centering
+% \caption{}
+% \renewcommand{\arraystretch}{1.3}
+% \begin{tabular}{lrr} \toprule
+%& \textbf{Volume (L)} & \textbf{Reference} \\ \hline
+%\phantom{i}Cellular volume & 3.881e-16 & XX \\ \hline
+% \toprule
+% & \textbf{Concentration (mM)} & \textbf{Reference} \\ \hline
+%\phantom{i}CAT gene & 5.19e-6 & XX \\ \hline
+%\phantom{i}RNA polymerase & 6.75e-5 & \cite{Garamella:2016aa} \\ \hline
+%\phantom{i}Ribosome & 2e-3 & \cite{Garamella:2016aa} \\ \hline
+%\phantom{i}tRNA & 1.6 & BNID 108611 \\ \hline
+%\phantom{i}NAD & 1.47 & XX \\ \hline
+%\phantom{i}NADH & 0.1 & XX \\ \hline
+%\phantom{i}NADP & 0.195 & XX \\ \hline
+%\phantom{i}NADPH & 0.062 & XX \\ \hline
+%\phantom{i}G6P & 3.48 & XX \\ \hline
+%\phantom{i}F6P & 0.6 & XX \\ \hline
+%\phantom{i}FDP & 0.272 & XX \\ \hline
+%\phantom{i}DHAP & 0.167 & XX \\ \hline
+%\phantom{i}G3P & 0.218 & XX \\ \hline
+%\phantom{i}13DPG & 0.008 & XX \\ \hline
+%\phantom{i}3PG & 2.13 & XX \\ \hline
+%\phantom{i}2PG & 0.399 & XX \\ \hline
+%\phantom{i}6PGL & 0.808 & XX \\ \hline
+%\phantom{i}6PGC & 0.808 & XX \\ \hline
+%\phantom{i}RU5P & 0.111 & XX \\ \hline
+%\phantom{i}XU5P & 0.138 & XX \\ \hline
+%\phantom{i}R5P & 0.398 & XX \\ \hline
+%\phantom{i}S7P & 0.276 & XX \\ \hline
+%\phantom{i}E4P & 0.098 & XX \\ \hline
+%\phantom{i}NH3 & 0.052425 & XX \\ \hline
+%\phantom{i}ARG & 1.9326 & XX \\ \hline
+%\phantom{i}H2O2 & 0.052425 & XX \\ \hline
+%\phantom{i}HCO3 & 0.052425 & XX \\ \hline
+%% \bottomrule
+% \end{tabular}
+%\label{tbl:Initial_conditions}
+%\end{table}
 

### 2017-05-17 Paper_v1.tex
May17

@@
 \centering
-\includegraphics[width=0.55\textwidth]{./Figures/Fig_6_CarbonEnergyBalances.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/Fig_6_CarbonEnergyBalances.pdf}
 %\caption{Carbon and energy balances for the best-fit set.
-A.
-Carbon moles produced as CAT, amino acids (alanine and glutamine), organic acids (lactate, acetate, succinate, and malate), and other byproducts including carbon dioxide, as percentages of total carbon consumption (glucose and all other amino acids).
-B.
-Energy cost of CAT production, accumulation of organic acids (lactate, acetate, succinate, and malate), and glycolytic and pentose phosphate metabolites, as percentages of total energy utilization from glucose.
-Energy costs calculated in terms of equivalent ATP molecules.}
-\caption{Carbon and energy balances for the best-fit set.
 A.
@@
 Energy costs calculated in terms of equivalent ATP molecules.}
+\caption{Carbon and energy balances during the first and second phases of protein production for the best-fit set.
+Top left: Carbon moles produced as CAT, amino acids (alanine, isoleucine, glutamine, proline, and tyrosine), organic acids (pyruvate, lactate, acetate, succinate, and malate), and all other byproducts, as percentages of total carbon consumption (glucose and all other amino acids).
+Bottom left: Carbon moles produced as CAT, amino acids (alanine, glutamine, proline, and serine), organic acids (acetate only), and all other byproducts, as percentages of total carbon consumption (pyruvate, lactate, succinate, malate, and all other amino acids).
+Top right: Energy cost of CAT production, accumulation of amino acids (alanine, isoleucine, glutamine, proline, and tyrosine), accumulation of organic acids (pyruvate, lactate, acetate, succinate, and malate), and other byproducts, as percentages of total energy utilization from glucose.
+Bottom right: Energy cost of CAT production, accumulation of amino acids (alanine, glutamine, proline, and serine), accumulation of organic acids (acetate only), and other byproducts, as percentages of total energy utilization from pyruvate, lactate, succinate, and malate.
+Energy costs calculated in terms of equivalent ATP molecules.}
+% including carbon dioxide
 \label{fig:CAT_balances}

### 2017-05-18 Paper_v1.tex
May17_2

@@
 We calculated that CAT was produced with a carbon yield of 7\% and an energy efficiency of 5\%, suggesting that much of the resources for protein synthesis were diverted to non-productive pathways.
-By knocking out metabolic enzymes in groups, we showed that CAT productivity depends most on oxidative phosphorylation, glycolysis/gluconeogenesis, and certain amino acid synthesis reactions, while the system state is most sensitive to glycolysis/gluconeogenesis, oxidative phosphorylation, the overflow metabolism, and the TCA cycle reactions.
+By knocking out metabolic enzymes in groups, we showed that CAT productivity depends most on oxidative phosphorylation, glycolysis\slash gluconeogenesis, and certain amino acid synthesis reactions, while the system state is most sensitive to glycolysis\slash gluconeogenesis, oxidative phosphorylation, the overflow metabolism, and the TCA cycle reactions.
 %First, sensitivity analysis of the dynamic model suggested that CAT production was most sensitive to CAT synthesis parameters, as well as reactions in oxidative phosphorylation and pyruvate consumption.
@@
 The constrained MCMC approach estimated parameter sets with a median error more than two orders of magnitude less than random parameter sets generated within the same parameter bounds (Fig.~\ref{fig:BoxPlot}); thus, we have confidence in the predictive capability of the estimated parameters.
-The model captured the biphasic CAT production: during the first hour glucose powers production, and CAT is produced at \texttildelow 10 $\mu$M/h; subsequently, pyruvate and lactate reserves are consumed to power metabolism, and CAT is produced less efficiently at \texttildelow 5 $\mu$M/h.
+The model captured the biphasic CAT production: during the first hour glucose powers production, and CAT is produced at 10 $\mu$M/h; subsequently, pyruvate and lactate reserves are consumed to power metabolism, and CAT is produced less efficiently at 3 $\mu$M/h.
 %Allosteric control was important to biphasic CAT production; without control, the CAT production rate increased and then ceased after 1.5 h (Fig.~\ref{fig:Carbon}, bottom).
@@
 These were summed with the first-order effect to obtain a total-order coefficient for each group for the change in productivity and system state.
-Glycolysis/gluconeogenesis and oxidative phosphorylation were seen to have the greatest effect on both productivity and system state.
+Glycolysis\slash gluconeogenesis and oxidative phosphorylation were seen to have the greatest effect on both productivity and system state.
 This is explained by their role in both central carbon metabolism and energy generation.
-In addition, CAT productivity is affected by two sectors of amino acid biosynthesis: alanine/aspartate/asparagine, and glutamate/glutamine.
+In addition, CAT productivity is affected by two sectors of amino acid biosynthesis: alanine\slash aspartate\slash asparagine, and glutamate\slash glutamine.
 This is likely because aspartate, glutamate, and glutamine are key reactants in the biosynthesis of many other amino acids, all of which are required for CAT synthesis.
-Meanwhile, the TCA cycle and the overflow metabolism (which includes acetyl-coA/acetate reactions and the interconversion of pyruvate and lactate) have a significant effect on the system state.
+Meanwhile, the TCA cycle and the overflow metabolism (which includes acetyl-coA\slash acetate reactions and the interconversion of pyruvate and lactate) have a significant effect on the system state.
 These reactions directly impact key species in the system state: succinate and malate in the TCA cycle, and acetate, pyruvate, and lactate in the overflow metabolism.
 
-To understand whether the CFPS performance was optimal, we calculated the carbon yield and energy efficiency of CAT production for the best-fit set (Fig.~\ref{fig:CAT_balances}).
-Of the carbon taken up as glucose and consumed as amino acids, 7\% went to CAT production.
-Another 7\% went to the accumulation of alanine and glutamine, in line with experimental data (these were not included in the amino acid consumption term).
-25\% of the carbon was accounted for by the net accumulation of organic acids (only lactate, acetate, succinate, and malate showed positive net accumulation), leaving 61\% as accumulation of other byproducts, species for which data do not exist.
-CFPS also showed an energy efficiency of 5\%; this was formulated as the equivalent ATP value of CAT produced divided by the equivalent ATP value of glucose consumed.
-The remainder of the balance was very similar to that of the carbon balance: 8\% to alanine and glutamine, 26\% to organic acids, and 61\% to other byproducts.
-The equivalent ATP numbers for glucose, amino acids, and organic acids were calculated from the network stoichiometry; for example, 21 molecules of ATP shold be generated from one glucose molecule if the optimal path through glycolysis, the TCA cycle, and oxidative phosphorylation is taken.
-The ATP number for CAT was simply equal to the cost of transcription and translation.
-%These results show 
-%The fact that so much carbon and energy are spent on non-productive pathways
-The low carbon yield and energy efficiency show that there is much room for improvement; CFPS efficiency could be increased by diverting flux away from some of the less efficient or altogether unnecessary metabolic pathways.
-%pathways of energy generation
+To understand whether the CFPS performance was optimal, we calculated the carbon yield and energy efficiency of the two phases of CAT production for the best-fit set (Fig.~\ref{fig:CAT_balances}).
+During the first phase, with glucose as the substrate, CAT is produced with a carbon yield of 5\% and an energy efficiency of 3\%.
+Of the remaining carbon, 4\% is accounted for by the accumulation of amino acids (alanine, isoleucine, glutamine, proline, and tyrosine), 39\% by organic acid accumulation (pyruvate, lactate, acetate, succinate, and malate), and 52\% by the accumulation of other byproducts, primarily glycolytic intermediates and carbon dioxide.
+The breakdown is very similar with the energy efficiency: 4\% amino acids, 41\% organic acids, and 52\% other byproducts.
+This suggests that for glucose-driven production, the best ways to improve efficiency are by ensuring that the flux through glycolysis and the TCA cycle is constant throughout, preventing any accumulation of intermediates.
+However, the accumulated organic acids (except acetate) were then utilized as substrates in the second phase once glucose had run out.
+Although succinate and malate are consumed in the second phase, they only account for 14\% of the substrate consumption; thus, it may be reasonable to consider this as pyruvate-driven production.
+Interestingly, this mode of protein production showed higher carbon yield and energy efficiency: 6\% in each case.
+Of the remaining carbon, 11\% went to amino acids (alanine, glutamine, proline, and serine), 31\% went to organic acids (only acetate in this case), and 52\% went to other byproducts.
+The remainder of the energy is accounted for by 21\% amino acids, 49\% organic acids, and 24\% other byproducts.
+While efficiency appears to be higher for the pyruvate-driven phase, it must be noted that the productivity is not: 3 $\mu$M/h, as opposed to 10 $\mu$M/h under glucose.
+%Of the carbon taken up as glucose and consumed as amino acids, 7\% went to CAT production.
+%Another 7\% went to the accumulation of alanine and glutamine, in line with experimental data (these were not included in the amino acid consumption term).
+%25\% of the carbon was accounted for by the net accumulation of organic acids (only lactate, acetate, succinate, and malate showed positive net accumulation), leaving 61\% as accumulation of other byproducts, species for which data do not exist.
+%CFPS also showed an energy efficiency of 5\%; this was formulated as the equivalent ATP value of CAT produced divided by the equivalent ATP value of glucose consumed.
+%The remainder of the balance was very similar to that of the carbon balance: 8\% to alanine and glutamine, 26\% to organic acids, and 61\% to other byproducts.
+%The equivalent ATP numbers for glucose, amino acids, and organic acids were calculated from the network stoichiometry; for example, 21 molecules of ATP shold be generated from one glucose molecule if the optimal path through glycolysis, the TCA cycle, and oxidative phosphorylation is taken.
+%The ATP number for CAT was simply equal to the cost of transcription and translation.
+%%These results show 
+%%The fact that so much carbon and energy are spent on non-productive pathways
+%The low carbon yield and energy efficiency show that there is much room for improvement; CFPS efficiency could be increased by diverting flux away from some of the less efficient or altogether unnecessary metabolic pathways.
+%%pathways of energy generation
 
@@
 % central carbon metabolism is exhuasted within 1.5 hours, leading to low CAT production.
-Without it, pyruvate, succinate, and malate are consumed more quickly following glucose exhaustion, presumably to power downstream reactions and ultimately CAT synthesis.
-Meanwhile, acetate accumulation during the increased after the first hour, suggesting a decrease in sytem efficiency.
-Interestingly, CAT production is virtually unaffected.
-This is because the amino acids and energy species used in CAT synthesis are not dependent on allosteric control.
+Without it, pyruvate, succinate, and malate are consumed more quickly following glucose exhaustion to power downstream reactions and ultimately CAT synthesis.
+%Meanwhile, acetate accumulation increased after the first hour, suggesting a decrease in sytem efficiency.
+Interestingly, CAT production is virtually unaffected; this is because the amino acids and energy species that are reactants for CAT synthesis were also not affected by allosteric control.
 
 Having captured the experimental data, we investigated if CAT yield and CFPS performance could be further improved.
-We showed that the model predicts CAT production with a carbon yield of 7\% and energy efficiency of 5\%.
+We showed that the model predicts CAT production with a carbon yield of 5\% and energy efficiency of 3\% under glucose, and a carbon yield and energy efficiency of 6\% under pyruvate.
 %We showed that the model produces CAT with a carbon yield equal to 16\% of that of a physiological case in which transcription and translation are constrained, and an energy efficiency equal to 9\% of that of the physiological case.
-The accumulation of waste byproducts, especially acetate and carbon dioxide, is responsible for this sub-optimal performance.
-However, there is also some accumulation of useful metabolites in the central carbon metabolism; if these could be utilized fully, CAT production would likely increase.
-Knocking out sections of network metabolism revealed that glycolysis/gluconeogenesis and oxidative phosphorylation were the most important to CAT production and the system as a whole.
+The accumulation of glycolytic intermediates and byproducts such as acetate and carbon dioxide was responsible for this sub-optimal performance.
+If fluxes could be balanced such that intermediates were fully utilized, CAT production would increase.
+%However, there is also some accumulation of useful metabolites in the central carbon metabolism; if these could be utilized fully, CAT production would likely increase.
+Knocking out sections of network metabolism revealed that glycolysis\slash gluconeogenesis and oxidative phosphorylation were the most important to CAT production and the system as a whole.
 Productivity was also heavily dependent on the synthesis reactions of alanine, aspartate, asparagine, glutamate, and glutamine, while TCA cycle and overflow reactions affected the system state.
@@
 % sustained by pyruvate and glutamate
-While CAT synthesis can be powered by other substrates, the rate of production about half (\texttildelow 5 $\mu$M/h, as opposed to \texttildelow 10 $\mu$M/h).
+While CAT synthesis can be powered by other substrates, the productivity is significantly lower (3 $\mu$M/h, as opposed to 10 $\mu$M/h).
 This is in accordance with literature, where pyruvate provided a relatively slow but continuous supply of ATP \cite{swartz_nature2001}.
@@
 We included 17 allosteric regulation terms, taken from literature, in the CFPS model.
-PEP was modeled as an inhibitor for phosphofructokinase \cite{2010_kotte_MolSystBiol,2011_cabrera_JBiolChem}, PEP carboxykinase \cite{2010_kotte_MolSystBiol}, PEP synthetase \cite{2010_kotte_MolSystBiol,1973_chulavatnatol_JBiolChem}, isocitrate dehydrogenase \cite{2010_kotte_MolSystBiol,2007_ogawa_JBacteriol}, and isocitrate lyase/malate synthase \cite{2010_kotte_MolSystBiol,2007_ogawa_JBacteriol,1988_mackintosh_BiochemJ}, and as an activator for fructose-biphosphatase \cite{2010_kotte_MolSystBiol,2000_donahue_JBacteriol,2006_hines_JBiolChem,2007_hines_JBiolChem}.
-AKG was modeled as an inhibitor for citrate synthase \cite{2010_kotte_MolSystBiol,1994_pereira_JBiolChem,1983_robinson_FEBSLett} and isocitrate lyase/malate synthase \cite{2010_kotte_MolSystBiol,1988_mackintosh_BiochemJ}.
-3PG was modeled as an inhibitor for isocitrate lyase/malate synthase \cite{2010_kotte_MolSystBiol,1988_mackintosh_BiochemJ}.
+PEP was modeled as an inhibitor for phosphofructokinase \cite{2010_kotte_MolSystBiol,2011_cabrera_JBiolChem}, PEP carboxykinase \cite{2010_kotte_MolSystBiol}, PEP synthetase \cite{2010_kotte_MolSystBiol,1973_chulavatnatol_JBiolChem}, isocitrate dehydrogenase \cite{2010_kotte_MolSystBiol,2007_ogawa_JBacteriol}, and isocitrate lyase\slash malate synthase \cite{2010_kotte_MolSystBiol,2007_ogawa_JBacteriol,1988_mackintosh_BiochemJ}, and as an activator for fructose-biphosphatase \cite{2010_kotte_MolSystBiol,2000_donahue_JBacteriol,2006_hines_JBiolChem,2007_hines_JBiolChem}.
+AKG was modeled as an inhibitor for citrate synthase \cite{2010_kotte_MolSystBiol,1994_pereira_JBiolChem,1983_robinson_FEBSLett} and isocitrate lyase\slash malate synthase \cite{2010_kotte_MolSystBiol,1988_mackintosh_BiochemJ}.
+3PG was modeled as an inhibitor for isocitrate lyase\slash malate synthase \cite{2010_kotte_MolSystBiol,1988_mackintosh_BiochemJ}.
 FDP was modeled as an activator for pyruvate kinase \cite{2010_kotte_MolSystBiol,2010_zhu_Biochimie} and PEP carboxylase \cite{2010_kotte_MolSystBiol,1972_wohl_JBiolChem}.
@@
 \subsection*{Group knockouts.}
-The network was divided into 19 groups: glycolysis/gluconeogenesis, pentose phosphate, Entner-Doudoroff, TCA cycle, oxidative phosphorylation, cofactor reactions, anaplerotic/glyoxylate reactions, overflow metabolism, folate synthesis, purine/pyrimidine reactions, alanine/aspartate/asparagine synthesis, glutamate/glutamine synthesis, arginine/proline synthesis, glycine/serine synthesis, cysteine/methionine synthesis, threonine/lysine synthesis, histidine synthesis, tyrosine/tryptophan/phenylalanine synthesis, and valine/leucine/isoleucine synthesis.
+The network was divided into 19 groups: glycolysis\slash gluconeogenesis, pentose phosphate, Entner-Doudoroff, TCA cycle, oxidative phosphorylation, cofactor reactions, anaplerotic\slash glyoxylate reactions, overflow metabolism, folate synthesis, purine\slash pyrimidine reactions, alanine\slash aspartate\slash asparagine synthesis, glutamate\slash glutamine synthesis, arginine\slash proline synthesis, glycine\slash serine synthesis, cysteine\slash methionine synthesis, threonine\slash lysine synthesis, histidine synthesis, tyrosine\slash tryptophan\slash phenylalanine synthesis, and valine\slash leucine\slash isoleucine synthesis.
 Each group of reactions was turned off individually, and then in pairs, and the model equations were re-solved.
@@
 
-\subsection*{Calculation of the carbon yield.}
-The CAT carbon yield ($Y_{C}^{CAT}$) was calculated as the ratio of carbon produced as CAT divided by the carbon consumed as reactants (glucose and amino acids):
+\subsection*{Calculation of carbon yield.}
+The CAT carbon yield ($Y_{C}^{CAT}$) was calculated as the ratio of carbon produced as CAT divided by the carbon consumed as reactants:
+%\begin{equation}\label{eqn:yield-definition}
+%	Y_{C}^{CAT}=\frac{\Delta\texttt{CAT}\cdot C_{CAT}}{\displaystyle\sum_{i=1}^{\mathcal{R}}\max(\Delta m_{i},0)\cdot C_{m_i}}
+%\end{equation}
 \begin{equation}\label{eqn:yield-definition}
-	Y_{C}^{CAT}=\frac{\Delta\texttt{CAT}\cdot C_{CAT}}{\displaystyle\sum_{i=1}^{\mathcal{R}}\max(\Delta m_{i},0)\cdot C_{m_i}}
+	Y_{C}^{CAT}=\frac{\Delta\texttt{CAT}\cdot C_{CAT}}{\displaystyle\sum_{i=1}^{\mathcal{R}}\Delta m_{i}\cdot C_i}
 \end{equation}
-where $\Delta\texttt{CAT}$ denotes the abundance of CAT produced, $C_{CAT}$ denotes carbon number of CAT, $\mathcal{R}$ denotes the number of reactants,
-$\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed (never allowed to be negative), and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
-This analysis was extended to the accumulation of amino acids, organic acids, and other byproducts, to create a complete carbon balance through the network (Fig.~\ref{fig:CAT_balances}A).
+where $\Delta\texttt{CAT}$ denotes the abundance of CAT produced, $C_{CAT}$ denotes carbon number of CAT, $\mathcal{R}$ denotes the number of reactants, $\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed, and $C_i$ denotes the carbon number of the $i^{th}$ reactant.
+This analysis was extended to the accumulation of amino acids, organic acids, and other byproducts, to create a complete carbon balance through the network (Fig.~\ref{fig:CAT_balances}, left).
+The first phase of CAT production was defined as t = 0 h to t = 1.11 h, the time at which glucose concentration falls below 0.1 nM.
+In the first phase, amino acid accumulation consisted of alanine, isoleucine, glutamine, proline, and tyrosine, while organic acid accumulation consisted of pyruvate, lactate, acetate, succinate, and malate.
+Glucose and the amino acids that did not accumulate were considered reactants.
+In the second phase, amino acid accumulation consisted of alanine, glutamine, proline, and serine, while organic acid accumulation consisted of acetate only.
+Pyruvate, lactate, succinate, malate, and the amino acids that did not accumulate were considered reactants.
 %Yield of the best-fit parameter set and the experimental data were calculated by setting $\Delta\texttt{CAT}$ equal to the final minus the initial CAT concentration,
@@
 \subsection*{Calculation of energy efficiency.}
-Energy efficiency was calculated as the ratio of CAT production to glucose consumption, both in terms of equivalent ATP molecules:
+Energy efficiency was calculated as the ratio of CAT production to substrate consumption, both in terms of equivalent ATP molecules:
+%\begin{equation}\label{eqn:energy-efficiency-definition}
+%	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2\left({ATP}_{TX}+\rm {CTP}_{TX}+\rm {GTP}_{TX}+\rm {UTP}_{TX}\right)+\rm2\cdot\rm {ATP}_{TL}+\rm {GTP}_{TL}\right)}{\Delta\texttt{GLC}\cdot \rm {ATP}_{GLC}}
+%\end{equation}
 \begin{equation}\label{eqn:energy-efficiency-definition}
-	\rm Efficiency=\frac{\Delta\texttt{CAT}\cdot \left(2\left({ATP}_{TX}+\rm {CTP}_{TX}+\rm {GTP}_{TX}+\rm {UTP}_{TX}\right)+\rm2\cdot\rm {ATP}_{TL}+\rm {GTP}_{TL}\right)}{\Delta\texttt{GLC}\cdot \rm {ATP}_{GLC}}
+	{\rm Efficiency}=\frac{\Delta\texttt{CAT}\cdot \left(2\cdot\left(\rm {ATP}_{TX}+\rm {CTP}_{TX}+\rm {GTP}_{TX}+\rm {UTP}_{TX}\right)+\rm2\cdot\rm {ATP}_{TL}+\rm {GTP}_{TL}\right)}{\displaystyle\sum_{i=1}^{\mathcal{R}}\Delta m_{i}\cdot {\rm ATP}_i}
 \end{equation}
-where $\rm {ATP}_{TX}$, $\rm {CTP}_{TX}$, $\rm {GTP}_{TX}$, $\rm {UTP}_{TX}$ denote the stoichiometric coefficients of each energy species for CAT transcription, $\rm {ATP}_{TL}$, $\rm {GTP}_{TL}$ denote the stoichiometric coefficients of ATP and GTP for CAT translation, $\Delta\texttt{GLC}$ denotes the glucose consumption, equal to the initial minus the final glucose concentration, and $\rm {ATP}_{GLC}$ denotes the equivalent ATP number for glucose.
-$\rm {ATP}_{TX}$ = 176, $\rm {CTP}_{TX}$ = 144, $\rm {GTP}_{TX}$ = 151, $\rm {UTP}_{TX}$ = 189, $\rm {ATP}_{TL}$ = 219, $\rm {GTP}_{TL}$ = 438, $\rm {ATP}_{GLC}$ = 15.
-This analysis was also extended to the accumulation of amino acids, organic acids, and other byproducts to create a complete energy balance through the network (Fig.~\ref{fig:CAT_balances}B).
+where $\rm {ATP}_{TX}$, $\rm {CTP}_{TX}$, $\rm {GTP}_{TX}$, $\rm {UTP}_{TX}$ denote the stoichiometric coefficients of each energy species for CAT transcription, $\rm {ATP}_{TL}$, $\rm {GTP}_{TL}$ denote the stoichiometric coefficients of ATP and GTP for CAT translation, $\Delta m_{i}$ denotes the amount of the $i^{th}$ substrate consumed, and ${\rm ATP}_i$ denotes the equivalent ATP number of the $i^{th}$ substrate.
+%$\Delta\texttt{GLC}$ denotes the glucose consumption, equal to the initial minus the final glucose concentration, and $\rm {ATP}_{GLC}$ denotes the equivalent ATP number for glucose.
+$\rm {ATP}_{TX}$ = 176, $\rm {CTP}_{TX}$ = 144, $\rm {GTP}_{TX}$ = 151, $\rm {UTP}_{TX}$ = 189, $\rm {ATP}_{TL}$ = 219, $\rm {GTP}_{TL}$ = 438, $\rm {ATP}_{GLC}$ = 21, $\rm {ATP}_{PYR}$ = 8, $\rm {ATP}_{LAC}$ = 9.5, $\rm {ATP}_{SUCC}$ = 11.5, $\rm {ATP}_{MAL}$ = 10.5.
+This analysis was also extended to the accumulation of amino acids, organic acids, and other byproducts to create a complete energy balance through the network (Fig.~\ref{fig:CAT_balances}, right).
+In the first phase, amino acid accumulation consisted of alanine, isoleucine, glutamine, proline, and tyrosine, while organic acid accumulation consisted of pyruvate, lactate, acetate, succinate, and malate.
+Glucose was considered the substrate.
+In the second phase, amino acid accumulation consisted of alanine, glutamine, proline, and serine, while organic acid accumulation consisted of acetate only.
+Pyruvate, lactate, succinate, and malate were considered the substrates.
 Equivalent ATP numbers for glucose, amino acids, and organic acids were calculated from the network stoichiometry.
@@
 %\phantom{i}Geometric mean & & & 110 & \\ \toprule
-%\phantom{i}\textbf{Transcription/Translation} & \textbf{AA-tRNA complex (mM)} & \textbf{Kcat (h$^-1$)} & \textbf{Vmax (mM/h)} & \textbf{Reference} \\ \hline
+%\phantom{i}\textbf{Transcription\slash Translation} & \textbf{AA-tRNA complex (mM)} & \textbf{Kcat (h$^-1$)} & \textbf{Vmax (mM/h)} & \textbf{Reference} \\ \hline
 %\phantom{i}tRNA charging & 0.03 & 14040 & 4212 & 104980 \\ \hline
@@
 \phantom{i}Characteristic rate maximum & & & 110 & \\ \toprule
-\phantom{i}\textbf{Transcription/Translation} & \textbf{Reaction} & \textbf{K$_{cat}$ (min$^{-1}$)} & \textbf{V$_{max}$ (mM/h)} & \textbf{Reference} \\ \hline
+\phantom{i}\textbf{Transcription\slash Translation} & \textbf{Reaction} & \textbf{K$_{cat}$ (min$^{-1}$)} & \textbf{V$_{max}$ (mM/h)} & \textbf{Reference} \\ \hline
 \phantom{i}mRNA degradation & 5.2 (1/h) & 0.08667 & 0.00052 & BNID 104980 \\ \hline
@@
 %\phantom{i}Geometric mean & & & 110 & \\ \toprule
-%%\phantom{i}\textbf{Transcription/Translation} & \textbf{AA-tRNA (mM)} & \textbf{Kcat (h$^-1$)} & \textbf{Vmax (mM/h)} & \textbf{Reference} \\ \hline
+%%\phantom{i}\textbf{Transcription\slash Translation} & \textbf{AA-tRNA (mM)} & \textbf{Kcat (h$^-1$)} & \textbf{Vmax (mM/h)} & \textbf{Reference} \\ \hline
 %\phantom{i}tRNA charging & 0.03 & 14040 & 4212 & 104980 \\ \hline

### 2017-05-25 Paper_v1.tex
May25

@@
 \usepackage{booktabs}
+\usepackage{xfrac}
 \setlength\heavyrulewidth{0.4ex}
@@
 Model parameters were estimated from a training dataset for the cell-free production of a protein product, chloramphenicol acetyltransferase (CAT).
-%This protein was used as a proof-of-concept
 The dataset consisted of measurements of glucose, organic acids, energy species, amino acids, and CAT.
@@
 Amino acid biosynthesis is also important for productivity, while the overflow metabolism and TCA cycle affect the overall system state.
-%The dynamic modeling approach predicted that substrate consumption and oxidative phosphorylation were most important to both CAT production and the system as a whole, while CAT production alone depended heavily on the CAT synthesis reaction.
 In addition, CAT production was robust to allosteric control, as was most of the network, with the exception of the organic acids in central carbon metabolism.
@@
 \vspace{0.1in}
-{\noindent \textbf{Keywords:}~Biochemical engineering, systems biology, cell-free protein synthesis}
+{\noindent \textbf{Keywords:}~Biochemical engineering, systems biology, \textit{E. coli}, cell-free protein synthesis, kinetic modeling}
 
@@
 However, if CFPS is to become a mainstream technology for applications such as point of care manufacturing, we must first understand the performance limits of these systems.
-%One tool to address this question is mathematical modeling.
 One tool we can use to achieve this understanding is mathematical modeling.
@@
 %Thus, modern constraint-based approaches have proven extremely useful in the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
-However, genome scale kinetic models of industrial important organisms such as \textit{E. coli} have yet to be constructed.
+However, cell-free genome scale kinetic models of industrially important organisms such as \textit{E. coli} have yet to be constructed.
 
@@
 Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids, and the protein product, chloramphenicol acetyltransferase (CAT).
-%Characteristic values for model parameters and metabolite initial conditions, estimated from literature, were used to constrain the parameter estimation problem.
-The parameter estimation problem was constrained by characteristic values for model parameters and metabolite initial conditions estimated from literature.
+The parameter estimation problem was constrained by characteristic values for model parameters from literature.
 The ensemble of parameter sets described the training data with a median cost that was more than two orders of magnitude smaller than random sets constructed using the literature parameter constraints.
-We then used the ensemble of kinetic models to analyze the optimality of the CFPS reaction, and the pathways most important to CAT production.
+We then used the ensemble of kinetic models to analyze the optimality of the CFPS system, and the pathways most important to CAT production.
 We calculated that CAT was produced with a carbon yield of 7\% and an energy efficiency of 5\%, suggesting that much of the resources for protein synthesis were diverted to non-productive pathways.
-By knocking out metabolic enzymes in groups, we showed that CAT productivity depends most on oxidative phosphorylation, glycolysis\slash gluconeogenesis, and certain amino acid synthesis reactions, while the system state is most sensitive to glycolysis\slash gluconeogenesis, oxidative phosphorylation, the overflow metabolism, and the TCA cycle reactions.
+By knocking out metabolic enzymes in groups, we showed that the system metabolism and protein production depend most on oxidative phosphorylation and glycolysis\slash gluconeogenesis.
+%By knocking out metabolic enzymes in groups, we showed that CAT productivity depends most on oxidative phosphorylation, glycolysis\slash gluconeogenesis, and certain amino acid synthesis reactions, while the system state is most sensitive to glycolysis\slash gluconeogenesis, oxidative phosphorylation, the overflow metabolism, and the TCA cycle reactions.
 %First, sensitivity analysis of the dynamic model suggested that CAT production was most sensitive to CAT synthesis parameters, as well as reactions in oxidative phosphorylation and pyruvate consumption.
@@
 An initial ensemble of model parameter sets (N $>$ 30,000) was estimated from measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids using a constrained Markov Chain Monte Carlo (MCMC) approach.
-%$\approx$
-The MCMC algorithm minimized the error between the training data and model simulations starting from an initial parameter set assembled from literature and inspection.
+%The MCMC algorithm minimized the error between the training data and model simulations starting from an initial parameter set assembled from literature and inspection.
 A final ensemble of parameter sets (N $=$ 100) was constructed by selecting the sets with the lowest errors, the lowest of which was defined as the best-fit set.
@@
 The difference between the allosteric control and no-control cases is mostly seen in the second phase of CAT production, after glucose is exhausted.
+Specifically, there is an increase in the consumption of pyruvate, succinate, and malate, as well as the accumulation of acetate.
 Taken together, we produced an ensemble of kinetic models that was consistent with time series measurements of the production of a model protein.
@@
 The network was divided into 19 groups of reactions, spanning central carbon metabolism, energetics, and amino acid biosynthesis.
-%The reactions in each of these groups were turned off, 
 The enzymes in each of these groups were knocked out, and the resulting change in productivity and system state were recorded.
@@
 Glycolysis\slash gluconeogenesis and oxidative phosphorylation were seen to have the greatest effect on both productivity and system state.
-This is explained by their role in both central carbon metabolism and energy generation.
-In addition, CAT productivity is affected by two sectors of amino acid biosynthesis: alanine\slash aspartate\slash asparagine, and glutamate\slash glutamine.
+This supports previous findings that oxidative phosphorylation does occur in a cell-free system \cite{Jewett:2008aa}.
+Jewett and coworkers observed a decrease in CAT yield, ranging from 1.5-fold to 4-fold, when knocking out oxidative phosphorylation reactions in cell-free (the Cytomim platform) with both pyruvate and glutamate as substrates.
+CAT productivity was also affected by two sectors of amino acid biosynthesis: alanine\slash aspartate\slash asparagine, and glutamate\slash glutamine.
 This is likely because aspartate, glutamate, and glutamine are key reactants in the biosynthesis of many other amino acids, all of which are required for CAT synthesis.
-Meanwhile, the TCA cycle and the overflow metabolism (which includes acetyl-coA\slash acetate reactions and the interconversion of pyruvate and lactate) have a significant effect on the system state.
+Meanwhile, the TCA cycle and overflow metabolism (which includes acetyl-coA\slash acetate reactions and the interconversion of pyruvate and lactate) have a significant effect on the system state, defined as all species for which ata exist.
 These reactions directly impact key species in the system state: succinate and malate in the TCA cycle, and acetate, pyruvate, and lactate in the overflow metabolism.
@@
 To understand whether the CFPS performance was optimal, we calculated the carbon yield and energy efficiency of the two phases of CAT production for the best-fit set (Fig.~\ref{fig:CAT_balances}).
-During the first phase, with glucose as the substrate, CAT is produced with a carbon yield of 5\% and an energy efficiency of 3\%.
+The equivalent ATP numbers for glucose, amino acids, and organic acids were calculated from the network stoichiometry; for example, 21 molecules of ATP shold be generated from one glucose molecule if the optimal path through glycolysis, the TCA cycle, and oxidative phosphorylation is taken.
+The ATP number for CAT was simply equal to the cost of transcription and translation.
+During the first phase, with glucose as the substrate, CAT was produced with a carbon yield of 5\% and an energy efficiency of 3\%.
 Of the remaining carbon, 4\% is accounted for by the accumulation of amino acids (alanine, isoleucine, glutamine, proline, and tyrosine), 39\% by organic acid accumulation (pyruvate, lactate, acetate, succinate, and malate), and 52\% by the accumulation of other byproducts, primarily glycolytic intermediates and carbon dioxide.
@@
 This suggests that for glucose-driven production, the best ways to improve efficiency are by ensuring that the flux through glycolysis and the TCA cycle is constant throughout, preventing any accumulation of intermediates.
-However, the accumulated organic acids (except acetate) were then utilized as substrates in the second phase once glucose had run out.
-Although succinate and malate are consumed in the second phase, they only account for 14\% of the substrate consumption; thus, it may be reasonable to consider this as pyruvate-driven production.
+However, the accumulated organic acids (except acetate) were then utilized as substrates in the second phase once glucose was depleted.
+Although succinate and malate were consumed in the second phase, they only account for 14\% of the substrate consumption; thus, it may be reasonable to consider this as pyruvate-driven production.
 Interestingly, this mode of protein production showed higher carbon yield and energy efficiency: 6\% in each case.
 Of the remaining carbon, 11\% went to amino acids (alanine, glutamine, proline, and serine), 31\% went to organic acids (only acetate in this case), and 52\% went to other byproducts.
-The remainder of the energy is accounted for by 21\% amino acids, 49\% organic acids, and 24\% other byproducts.
-While efficiency appears to be higher for the pyruvate-driven phase, it must be noted that the productivity is not: 3 $\mu$M/h, as opposed to 10 $\mu$M/h under glucose.
+The remainder of the energy was accounted by 21\% in amino acids, 49\% in organic acids, and 24\% in other byproducts.
+While the efficiency of production is higher for the pyruvate-driven phase, it is still relatively low, and could be improved by utilizing more of the resources that are generated (amino aicds and glycoltyic\slash TCA intermediates).
 %Of the carbon taken up as glucose and consumed as amino acids, 7\% went to CAT production.
@@
 %The remainder of the balance was very similar to that of the carbon balance: 8\% to alanine and glutamine, 26\% to organic acids, and 61\% to other byproducts.
-%The equivalent ATP numbers for glucose, amino acids, and organic acids were calculated from the network stoichiometry; for example, 21 molecules of ATP shold be generated from one glucose molecule if the optimal path through glycolysis, the TCA cycle, and oxidative phosphorylation is taken.
-%The ATP number for CAT was simply equal to the cost of transcription and translation.
-%%These results show 
 %%The fact that so much carbon and energy are spent on non-productive pathways
 %The low carbon yield and energy efficiency show that there is much room for improvement; CFPS efficiency could be increased by diverting flux away from some of the less efficient or altogether unnecessary metabolic pathways.
-%%pathways of energy generation
 
@@
 Our ensemble of models accurately predicts dynamic experimental measurements of central carbon metabolism, energy species, and amino acids over 100 times better than random sets in the same region of parameter space.
-%goodness-of-fit
-CFPS is seen to be biphasic, relying on glucose during the first hour and pyruvate and lactate afterward.
-Allosteric control was essential to the maintenance of the network, specifically the central carbon metabolism.
-% and production of CAT.
-% central carbon metabolism is exhuasted within 1.5 hours, leading to low CAT production.
-Without it, pyruvate, succinate, and malate are consumed more quickly following glucose exhaustion to power downstream reactions and ultimately CAT synthesis.
-%Meanwhile, acetate accumulation increased after the first hour, suggesting a decrease in sytem efficiency.
+CFPS was seen to be biphasic, relying on glucose during the first hour and pyruvate and lactate afterward.
+Allosteric control was essential to the maintenance of the network, specifically central carbon metabolism.
+Without it, pyruvate, succinate, and malate were consumed more quickly following glucose exhaustion to power downstream reactions and ultimately CAT synthesis.
 Interestingly, CAT production is virtually unaffected; this is because the amino acids and energy species that are reactants for CAT synthesis were also not affected by allosteric control.
@@
 We showed that the model predicts CAT production with a carbon yield of 5\% and energy efficiency of 3\% under glucose, and a carbon yield and energy efficiency of 6\% under pyruvate.
-%We showed that the model produces CAT with a carbon yield equal to 16\% of that of a physiological case in which transcription and translation are constrained, and an energy efficiency equal to 9\% of that of the physiological case.
-The accumulation of glycolytic intermediates and byproducts such as acetate and carbon dioxide was responsible for this sub-optimal performance.
+The accumulation of glycolytic intermediates and byproducts such as acetate and carbon dioxide were responsible for this sub-optimal performance.
 If fluxes could be balanced such that intermediates were fully utilized, CAT production would increase.
-%However, there is also some accumulation of useful metabolites in the central carbon metabolism; if these could be utilized fully, CAT production would likely increase.
 Knocking out sections of network metabolism revealed that glycolysis\slash gluconeogenesis and oxidative phosphorylation were the most important to CAT production and the system as a whole.
@@
 
-We present an ensemble of models that quantitatively describes the system behavior of cell-free metabolism and production of CAT.
+We present an ensemble of models that quantitatively described the system behavior of cell-free metabolism and production of CAT.
 Experimental observations of the metabolites validate the structure of the model and the estimation of kinetic parameters.
-% and cometabolites
 This is important in applying metabolic engineering principles to rationally design cell-free production processes and predicting the redirection of carbon fluxes to product-forming pathways.
-In analyzing the effect of reaction groups on CAT production and the system state, the regions of metabolism associated with substrate utilization and subsequent energy generation are the most important.
+In analyzing the effect of reaction groups on CAT production and the system state, the regions of metabolism associated with substrate utilization and energy generation are the most important.
 Oxidative phosphorylation is vital, since it provides most of the energetic needs of CFPS.
-Jewett and coworkers observed a decrease in CAT yield, between 1.5-fold and 4-fold, when knocking out oxidative phosphorylation reactions \cite{Jewett:2008aa}.
-% with pyruvate as the substrate
 While it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, our modeling approach suggests its importance to CFPS performance and protein yield.
 However, the biphasic operation of CFPS highlights the ability of the system to respond to an absence of glucose.
-% with the first phase operating on glucose and the second phase operating on pyruvate.
 During the first phase, there is an accumulation of central carbon metabolites with the majority of flux going toward acetate and some toward pyruvate, lactate, succinate and malate.
-While acetate continues to accumulate as a byproduct, the other organic acids are consumed as secondary substrates after glucose is no longer available.
-Glutamate also serves as a substrate throughout both phases, powering amino acid synthesis.
-These results show that CAT production can be sustained by other substrates in the absence of glucose, providing alternative strategies to optimize CFPS performance.
-% sustained by pyruvate and glutamate
+While acetate continued to accumulate as a byproduct, the other organic acids were consumed as secondary substrates after glucose is no longer available.
+Glutamate also served as a substrate throughout both phases, powering amino acid synthesis.
+These results confirm experimental findings that CAT production can be sustained by other substrates in the absence of glucose, providing alternative strategies to optimize CFPS performance.
 While CAT synthesis can be powered by other substrates, the productivity is significantly lower (3 $\mu$M/h, as opposed to 10 $\mu$M/h).
@@
 Taken together, this shows CFPS can be designed towards a specified application, either requiring a slow stable energy source or faster production.
-
-%CAT synthesis is the most important, followed by oxidative phosphorylation and the glutamate and pyruvate consuming reactions, as well as cofactor reactions which are necessary to drive CAT synthesis.
-%For example, the conversion of ATP to GTP shows significance since it is necessary for CAT synthesis.
-%While Jewett and coworkers have shown that ATP may be at saturation in CFPS \cite{Jewett:2008aa}, GTP is also required for CAT synthesis and may be a limiting reactant.
-%Thus, supplementation with additional GTP may improve the efficiency of CAT production.
-%A similar theme is seen in the sensitivity of overall model state, where the most important reactions are glucose and pyruvate consuming reactions and cofactor reactions which are vital to drive CFPS.
-%This can be seen in the biphasic operation of CFPS, with the first phase operating on glucose and the second phase operating on pyruvate.
-%During the first phase, there is an accumulation of byproducts from central carbon with the majority of flux going toward acetate and some toward pyruvate, lactate, and succinate; with the exception of acetate, these are all consumed in the second phase.
-%This shows that CAT production can be sustained by pyruvate and glutamate in the absence of glucose, which provides alternative strategies to optimize CFPS performance.
-%This is in accordance with literature, which showed pyruvate provided a relatively slow but continuous supply of ATP \cite{swartz_nature2001}.
-%Taken together, this shows CFPS can be designed towards a specified application either requiring a slow stable energy source or faster production.
-%This outstanding control on model performace was expected as these metabolites are responsible for driving CFPS and represent the first step in the model network.
-%Nevertheless, there are further reactions with considerable impact on model performance.
-%In examining oxidative phosphorylation activity, knockouts in the electron transport pathways disrupt metabolism across the network and drop CAT carbon yield from 7.9\% to 2.6\%; Jewett and coworkers also saw a similar decrease in CAT yield with pyruvate as the substrate, ranging from 1.5-fold to 4-fold, when knocking out oxidative phosphorylation reactions \cite{Jewett:2008aa}.
-%Oxidative phosphorylation is vital, since it provides most of the energetic needs of CFPS.
-%While it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, both of our modeling approaches suggest its importance to improving CFPS performance and protein yield.
-
-%Another area for future work is to more thoroughly sample parameter space.
-%Parameters were varied so as to best fit the dataset; however, the resulting ensemble may not represent every biological possibility.
-%In a different region of parameter space, the system may behave differently but still fit the experimental data.
-%This could include the flux distribution through the network, the variation of predictions across the ensemble, and the relative sensitivity values.
-%Testing the model under a variety of conditions could strengthen or challenge the findings of this study.
+%\textcolor{red}{This paragraph should also compare other work done in literature, since this is mostly a modeling paper, maybe compare other cell-free models?
+Or say modeling efforts in CFPS are weak)}
 
@@
 Specifically, CAT production performed in the absence of amino acids could inform the system's ability to manufacture them, while experimentation in the absence of glucose or oxygen could shed light on the importance of those substrates.
-in addition, the approach should be extended to other protein products.
+In addition, the approach should be extended to other protein products.
 CAT is only a test protein used for model identification; the modeling framework, and to some extent the parameter values, should be protein agnostic.
@@
 \end{equation}
-where $V_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$,
-$K_{js}$ denotes the saturation constant for species $s$ in reaction $j$ and $m_{j}^{-}$ denotes the set of \textit{reactants} for reaction $j$.
+where $V_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, $K_{js}$ denotes the saturation constant for species $s$, in reaction $j$ and $m_{j}^{-}$ denotes the set of \textit{reactants} for reaction $j$.
 On the other hand, the control term $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
@@
 Acetyl CoA was modeled as an inhibitor for malate dehydrogenase \cite{2010_kotte_MolSystBiol}.
+
+Transcription was modeled as first-order in RNA polymerase, since it acts like an enzyme:
+\begin{equation}\label{eqn:transcription}
+    \bar{r}^{TX} = V_{max}^{TX}\cdot{\rm RNAP}\left(\frac{{\rm GENE}}{K_{GENE}^{TX}+{\rm GENE}}\right)\epsilon_{i}\prod_{s\in{m_{TX}^{-}}}\frac{x_{s}}{K_{s}^{TX}+x_{s}}
+\end{equation}
+where $V_{max}^{TX}$ denotes the maximum transcription rate, ${\rm RNAP}$ denotes the RNA polymerase concentration, $GENE$ denotes the gene concentration, $K_{GENE}^{TX}$ denotes the gene saturation constant, $\epsilon_{i}$ denotes the scaled enzyme activity, $K_{s}^{TX}$ denotes the saturation constant for species $s$, and $m_{TX}^{-}$ denotes the set of \textit{reactants} for transcription: ATP, GTP, CTP, UTP, and water.
+While transcription is modeled as saturating with respect to gene concentration, the gene is not considered a reactant in the stoichiometry as it is not consumed.
+
+Translation was modeled as first-order in ribosome, since it acts like an enzyme:
+\begin{equation}\label{eqn:translation}
+    \bar{r}^{TL} = V_{max}^{TL}\cdot{\rm RIBO}\left(\frac{{\rm mRNA}}{K_{mRNA}^{TL}+{\rm mRNA}}\right)\epsilon_{i}\prod_{s\in{m_{TL}^{-}}}\frac{x_{s}}{K_{s}^{TL}+x_{s}}
+\end{equation}
+where $V_{max}^{TL}$ denotes the maximum translation rate, ${\rm RIBO}$ denotes the ribosome concentration, ${\rm mRNA}$ denotes the transcript concentration, $K_{mRNA}^{TL}$ denotes the transcript saturation constant, $\epsilon_{i}$ denotes the scaled enzyme activity, $K_{s}^{TL}$ denotes the saturation constant for species $s$, and $m_{TL}^{-}$ denotes the set of \textit{reactants} for translation: GTP, water, and the 20 species representing tRNA charged with amino acids.
+While translation is modeled as saturating with respect to transcript concentration, the transcript is not considered a reactant in the stoichiometry as it is not consumed.
+
+Transcript degradation was modeled as first-order in transcript:
+\begin{equation}\label{eqn:mRNA_degradation}
+    \bar{r}_{deg}^{mRNA} = k_{deg}^{mRNA}\cdot{\rm mRNA}
+\end{equation}
+where $k_{deg}^{mRNA}$ denotes the transcript degradation rate constant.
 
@@
 
-%catalytic rate bounds of 0 and 27,780 s$^{-1}$.
 Model parameters were constrained by literature \cite{Milo:2009aa}.
+The rate maximum for CAT transcription was calculated as 123 h$^{-1}$:
+\begin{equation}\label{eqn:transcription}
+	V_{max}^{TX}=\left(\frac{k_{TX}}{l_{mRNA}}\right)\cdot P
+\end{equation}
+where $k_{TX}$ denotes the average mRNA elongation rate (25 nt/s, \cite{Garamella:2016aa}), $l_{mRNA}$ denotes the CAT mRNA length (660 nt), and $P$ denotes the level of promoter activity (estimated at 0.9).
+% $RNAP$ denotes the RNA polymerase concentration (67.5 nM, \cite{Garamella:2016aa}),
+The rate maximum for CAT translation was calculated as 247 h$^{-1}$:
+\begin{equation}\label{eqn:transcription}
+	V_{max}^{TL}=\left(\frac{k_{TL}}{l_{protein}}\right)\cdot K_P
+\end{equation}
+where $k_{TL}$ denotes the peptide chain elongation rate (1.5 aa/s, \cite{Garamella:2016aa}), $l_{protein}$ denotes the CAT protein length (219 aa), and $K_P$ denotes the polysome amplification constant (estimated at 10).
+% $Ribo$ denotes the ribosome concentration (2000 nM, \cite{Garamella:2016aa}),
+The transcript degradation rate constant was calculated as 5.2 h$^{-1}$:
+\begin{equation}\label{eqn:transcription}
+	k_{deg}^{mRNA}=\left(\frac{log(2)}{t_{\sfrac{1}{2}}}\right)\cdot{\rm mRNA}
+\end{equation}
+where $t_{\sfrac{1}{2}}$ denotes the mRNA degradation time (8 min, BNID 106253).
+%The rate maximum for tRNA charging was calculated as 4200 mM/h:
+%\begin{equation}\label{eqn:transcription}
+%	V_{max}^{tRNA}=k_{tRNA}\cdot{\rm tRNA}
+%\end{equation}
+%where $k_{tRNA}$ denotes the tRNA charging catalytic rate (39 s$^{-1}$, BNID 104980), and ${\rm tRNA}$ denotes the charged tRNA concentration (estimated at 30 $\mu$M, consistent with the kinetic model).
+Transcription, translation, mRNA degradation, and tRNA charging were bounded within two orders of magnitude of their calculated values.
 A characteristic cell-free enzyme concentration of 167 nM was calculated by diluting the one-tenth maximal amount \textit{lacZ} (5 $\mu$M, BNID 100735) by a cell-free dilution factor of 30.
@@
 Rate maxima were bounded within one order of magnitude of the calculated value where available; all other rate maxima were bounded within two orders of magnitude of the geometric mean of the available values.
+
 The median maximum reaction rate was 7.8 mM/h; assuming a total cell-free enzyme concentration of 167 nM, this corresponds to a median catalytic rate of 0.08 s$^{-1}$ across the ensemble.
@@
 This analysis was also extended to the accumulation of amino acids, organic acids, and other byproducts to create a complete energy balance through the network (Fig.~\ref{fig:CAT_balances}, right).
+These terms were calculated as accumulation times equivalent ATP number over total energy consumption.
+Equivalent ATP numbers for glucose, amino acids, and organic acids were calculated from the network stoichiometry.
 In the first phase, amino acid accumulation consisted of alanine, isoleucine, glutamine, proline, and tyrosine, while organic acid accumulation consisted of pyruvate, lactate, acetate, succinate, and malate.
@@
  \begin{tabular}{llrrr} \toprule
-%\smallskip
-% & & \parbox{2.1 cm}{\textbf{Original value ($\mu$M)}} & \parbox{3.1 cm}{\textbf{Cell-free \phantom{abc} equivalent (nM)}} & \textbf{Reference} \\ \hline
-%\smallskip
-%\rule{0pt}{3ex} \parbox{4.2 cm}{Characteristic enzyme concentration} & & 5 & 167 & BNID 100735 \\ \toprule
 \phantom{i}\textbf{Enzyme} & \textbf{Reaction} & \textbf{K$_{cat}$ (min$^{-1}$)} & 
- \textbf{V$_{max}$ (mM/h)} & \textbf{Reference} \\ \hline
-\phantom{i}Serine dehydrase & R\_ser\_deg & 10400 & 104 & BNID 101119 \\ \hline
-\phantom{i}Isocitrate dehydrogenase & R\_icd & 11900 & 119 & BNID 101152 \\ \hline
-\phantom{i}Lactate dehydrogenase & R\_ldh & 5800 & 58 & BNID 101036 \\ \hline
+ \textbf{V$_{max}$ (mM/h)} & \textbf{BNID \#} \\ \hline
+\phantom{i}Serine dehydrase & R\_ser\_deg & 10400 & 104 & 101119 \\ \hline
+\phantom{i}Isocitrate dehydrogenase & R\_icd & 11900 & 119 & 101152 \\ \hline
+\phantom{i}Lactate dehydrogenase & R\_ldh & 5800 & 58 & 101036 \\ \hline
 \smallskip
 \rule{0pt}{4.5ex}
-Aspartate transaminase & \parbox{1 cm}{R\_aspC, R\_tyr, R\_phe} & 25800 & 258 & BNID 101108 \\ \hline
-\phantom{i}Enolase & R\_eno & 13200 & 132 & BNID 101028 \\ \hline
+Aspartate transaminase & \parbox{1 cm}{R\_aspC, R\_tyr, R\_phe} & 25800 & 258 & 101108 \\ \hline
+\phantom{i}Enolase & R\_eno & 13200 & 132 & 101028 \\ \hline
 \smallskip
 \rule{0pt}{3ex}
-Pyruvate kinase & R\_pyk & 25000 & 250 & \parbox{2.52 cm}{BNID 101029 BNID 101030} \\ \hline
+Pyruvate kinase & R\_pyk & 25000 & 250 & \parbox{1.4 cm}{101029 101030} \\ \hline
 \smallskip
 \rule{0pt}{3ex}
-Malic enzyme & \parbox{1 cm}{R\_maeA, R\_maeB} & 35400 & 354 & BNID 101167 \\ \hline
-\phantom{i}Phosphofructokinase & R\_pfk & 554400 & 5544 & BNID 104955 \\ \hline
-\phantom{i}Malate dehydrogenase & R\_mdh & 33000 & 330 & BNID 101163 \\ \hline
-\phantom{i}Citrate Synthase & R\_gltA & 42000 & 420 & BNID 101149 \\ \hline
+Malic enzyme & \parbox{1 cm}{R\_maeA, R\_maeB} & 35400 & 354 & 101167 \\ \hline
+\phantom{i}Phosphofructokinase & R\_pfk & 554400 & 5544 & 104955 \\ \hline
+\phantom{i}Malate dehydrogenase & R\_mdh & 33000 & 330 & 101163 \\ \hline
+\phantom{i}Citrate Synthase & R\_gltA & 42000 & 420 & 101149 \\ \hline
 \smallskip
 \rule{0pt}{4.5ex}
-6PG dehydrogenase & \parbox{1 cm}{R\_zwf, R\_pgl, R\_gnd} &3200 & 32 & BNID 101048 \\ \hline
-\phantom{i}Succinate dehydrogenase & R\_sdh & 121 & 1.21 & BNID 101162 \\ \hline
-\phantom{i}Succinyl-coA synthetase & R\_sucCD & 4700 & 47 & BNID 101158 \\ \hline
-\phantom{i}3PGA dehydrogenase & R\_gpm & 1100 & 11 & BNID 101135 \\ \hline
-\phantom{i}PEP carboxylase & R\_ppc & 35400 & 354 & BNID 101139 \\ \hline
-\phantom{i}3PGA kinase & R\_pgk & 4300 & 43 & BNID 101016 \\ \midrule
-\phantom{i}Characteristic rate maximum & & & 110 & \\ \toprule
-\phantom{i}\textbf{Transcription\slash Translation} & \textbf{Reaction} & \textbf{K$_{cat}$ (min$^{-1}$)} & \textbf{V$_{max}$ (mM/h)} & \textbf{Reference} \\ \hline
-\phantom{i}mRNA degradation & 5.2 (1/h) & 0.08667 & 0.00052 & BNID 104980 \\ \hline
-\phantom{i}tRNA charging & 0.03 & 2340 & 4212 & BNID 104980 \\ \hline
-% \bottomrule
+6PG dehydrogenase & \parbox{1 cm}{R\_zwf, R\_pgl, R\_gnd} &3200 & 32 & 101048 \\ \hline
+\phantom{i}Succinate dehydrogenase & R\_sdh & 121 & 1.21 & 101162 \\ \hline
+\phantom{i}Succinyl-coA synthetase & R\_sucCD & 4700 & 47 & 101158 \\ \hline
+\phantom{i}3PGA dehydrogenase & R\_gpm & 1100 & 11 & 101135 \\ \hline
+\phantom{i}PEP carboxylase & R\_ppc & 35400 & 354 & 101139 \\ \hline
+\phantom{i}3PGA kinase & R\_pgk & 4300 & 43 & 101016 \\ \hline
+\phantom{i}Characteristic rate maximum & & & 110 & \\ \bottomrule
  \end{tabular}
@@
 %\phantom{i}Ribosome & 2e-3 & \cite{Garamella:2016aa} \\ \hline
-%\phantom{i}tRNA & 1.6 & BNID 108611 \\ \hline
+%\phantom{i}tRNA & 1.6 & 108611 \\ \hline
 %\phantom{i}NAD & 1.47 & XX \\ \hline

### 2017-05-26 Paper_v1.tex
May26

@@
 where $V_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, $K_{js}$ denotes the saturation constant for species $s$, in reaction $j$ and $m_{j}^{-}$ denotes the set of \textit{reactants} for reaction $j$.
-On the other hand, the control term $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
+Transcription was modeled as first-order in RNA polymerase, since it acts like an enzyme:
+\begin{equation}\label{eqn:transcription}
+    \bar{r}^{TX} = k_{cat}^{TX}\cdot{\rm RNAP}\left(\frac{{\rm GENE}}{K_{GENE}^{TX}+{\rm GENE}}\right)\prod_{s\in{m_{TX}^{-}}}\frac{x_{s}}{K_{s}^{TX}+x_{s}}
+\end{equation}
+where $k_{cat}^{TX}$ denotes the maximum transcription rate, ${\rm RNAP}$ denotes the RNA polymerase concentration, $GENE$ denotes the gene concentration, $K_{GENE}^{TX}$ denotes the gene saturation constant, $K_{s}^{TX}$ denotes the saturation constant for species $s$, and $m_{TX}^{-}$ denotes the set of \textit{reactants} for transcription: ATP, GTP, CTP, UTP, and water.
+While transcription is modeled as saturating with respect to gene concentration, the gene is not considered a reactant in the stoichiometry as it is not consumed.
+Translation was modeled as first-order in ribosome, since it acts like an enzyme:
+\begin{equation}\label{eqn:translation}
+    \bar{r}^{TL} = k_{cat}^{TL}\cdot{\rm RIBO}\left(\frac{{\rm mRNA}}{K_{mRNA}^{TL}+{\rm mRNA}}\right)\prod_{s\in{m_{TL}^{-}}}\frac{x_{s}}{K_{s}^{TL}+x_{s}}
+\end{equation}
+where $k_{cat}^{TL}$ denotes the maximum translation rate, ${\rm RIBO}$ denotes the ribosome concentration, ${\rm mRNA}$ denotes the transcript concentration, $K_{mRNA}^{TL}$ denotes the transcript saturation constant, $K_{s}^{TL}$ denotes the saturation constant for species $s$, and $m_{TL}^{-}$ denotes the set of \textit{reactants} for translation: GTP, water, and the 20 species representing tRNA charged with amino acids.
+While translation is modeled as saturating with respect to transcript concentration, the transcript is not considered a reactant in the stoichiometry as it is not consumed.
+Transcript degradation was modeled as first-order in transcript:
+\begin{equation}\label{eqn:mRNA_degradation}
+    \bar{r}_{deg}^{mRNA} = k_{deg}^{mRNA}\cdot{\rm mRNA}
+\end{equation}
+where $k_{deg}^{mRNA}$ denotes the transcript degradation rate constant.
+
+The control term $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
 For each rate, we used a rule-based approach to select from competing control factors.
@@
 Acetyl CoA was modeled as an inhibitor for malate dehydrogenase \cite{2010_kotte_MolSystBiol}.
-
-Transcription was modeled as first-order in RNA polymerase, since it acts like an enzyme:
-\begin{equation}\label{eqn:transcription}
-    \bar{r}^{TX} = V_{max}^{TX}\cdot{\rm RNAP}\left(\frac{{\rm GENE}}{K_{GENE}^{TX}+{\rm GENE}}\right)\epsilon_{i}\prod_{s\in{m_{TX}^{-}}}\frac{x_{s}}{K_{s}^{TX}+x_{s}}
-\end{equation}
-where $V_{max}^{TX}$ denotes the maximum transcription rate, ${\rm RNAP}$ denotes the RNA polymerase concentration, $GENE$ denotes the gene concentration, $K_{GENE}^{TX}$ denotes the gene saturation constant, $\epsilon_{i}$ denotes the scaled enzyme activity, $K_{s}^{TX}$ denotes the saturation constant for species $s$, and $m_{TX}^{-}$ denotes the set of \textit{reactants} for transcription: ATP, GTP, CTP, UTP, and water.
-While transcription is modeled as saturating with respect to gene concentration, the gene is not considered a reactant in the stoichiometry as it is not consumed.
-
-Translation was modeled as first-order in ribosome, since it acts like an enzyme:
-\begin{equation}\label{eqn:translation}
-    \bar{r}^{TL} = V_{max}^{TL}\cdot{\rm RIBO}\left(\frac{{\rm mRNA}}{K_{mRNA}^{TL}+{\rm mRNA}}\right)\epsilon_{i}\prod_{s\in{m_{TL}^{-}}}\frac{x_{s}}{K_{s}^{TL}+x_{s}}
-\end{equation}
-where $V_{max}^{TL}$ denotes the maximum translation rate, ${\rm RIBO}$ denotes the ribosome concentration, ${\rm mRNA}$ denotes the transcript concentration, $K_{mRNA}^{TL}$ denotes the transcript saturation constant, $\epsilon_{i}$ denotes the scaled enzyme activity, $K_{s}^{TL}$ denotes the saturation constant for species $s$, and $m_{TL}^{-}$ denotes the set of \textit{reactants} for translation: GTP, water, and the 20 species representing tRNA charged with amino acids.
-While translation is modeled as saturating with respect to transcript concentration, the transcript is not considered a reactant in the stoichiometry as it is not consumed.
-
-Transcript degradation was modeled as first-order in transcript:
-\begin{equation}\label{eqn:mRNA_degradation}
-    \bar{r}_{deg}^{mRNA} = k_{deg}^{mRNA}\cdot{\rm mRNA}
-\end{equation}
-where $k_{deg}^{mRNA}$ denotes the transcript degradation rate constant.
 
@@
 \begin{equation}\label{eqn:transcription}
-	V_{max}^{TX}=\left(\frac{k_{TX}}{l_{mRNA}}\right)\cdot P
+	k_{cat}^{TX}=\left(\frac{k_{TX}}{l_{mRNA}}\right)\cdot P
 \end{equation}
@@
 \begin{equation}\label{eqn:transcription}
-	V_{max}^{TL}=\left(\frac{k_{TL}}{l_{protein}}\right)\cdot K_P
+	k_{cat}^{TL}=\left(\frac{k_{TL}}{l_{protein}}\right)\cdot K_P
 \end{equation}

### 2017-05-26 Paper_v1.tex
May26

@@
 \begin{equation}\label{eqn:transcription}
-	k_{deg}^{mRNA}=\left(\frac{log(2)}{t_{\sfrac{1}{2}}}\right)\cdot{\rm mRNA}
+	k_{deg}^{mRNA}=\frac{log(2)}{t_{\sfrac{1}{2}}}
 \end{equation}
@@
 A characteristic cell-free enzyme concentration of 167 nM was calculated by diluting the one-tenth maximal amount \textit{lacZ} (5 $\mu$M, BNID 100735) by a cell-free dilution factor of 30.
-This enzyme level was then used to calculate rate maxima from turnover numbers for various enzymes from literature (Table~\ref{tbl:Rate_maxima}).
+This enzyme level was then used to calculate rate maxima from turnover numbers for various enzymes from literature.
 Rate maxima were bounded within one order of magnitude of the calculated value where available; all other rate maxima were bounded within two orders of magnitude of the geometric mean of the available values.
@@
 The model equations were then solved and the cost function was calculated in terms of the 37 separate experimental datasets.
-The random ensemble had a log median error of 0.80 across the datasets, as compared with a log median error of -1.43 for the trained ensemble (Fig.~\ref{fig:BoxPlot}).
+The random ensemble had a log median error of 0.80 across the datasets, as compared with a log median error of -1.43 for the trained ensemble.
 Thus, the trained ensemble fits the dataset over one hundred times better than a random ensemble generated within the same bounds.
@@
 Each group of reactions was turned off individually, and then in pairs, and the model equations were re-solved.
-The CAT productivity was calculated and compared to that of the best-fit set (Fig.~\ref{fig:GroupKO}A).
+The CAT productivity was calculated and compared to that of the best-fit set.
 The absolute difference in productivity was recorded for each first-order knockout (diagonal elements) and each pairwise knockout, and a total-order coefficient was calculated by summing the first-order effect with all pairwise effects.
@@
 The system state was also calculated for each simulation, defined as the model predictions for all species for which data exist.
-The norm of the difference between the knockout system state and the best-fit system state is shown in (Fig.~\ref{fig:GroupKO}B).
 
@@
 where $\Delta\texttt{CAT}$ denotes the abundance of CAT produced, $C_{CAT}$ denotes carbon number of CAT, $\mathcal{R}$ denotes the number of reactants, $\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed, and $C_i$ denotes the carbon number of the $i^{th}$ reactant.
-This analysis was extended to the accumulation of amino acids, organic acids, and other byproducts, to create a complete carbon balance through the network (Fig.~\ref{fig:CAT_balances}, left).
+This analysis was extended to the accumulation of amino acids, organic acids, and other byproducts, to create a complete carbon balance through the network.
 The first phase of CAT production was defined as t = 0 h to t = 1.11 h, the time at which glucose concentration falls below 0.1 nM.
@@
 %and setting $\Delta m_{i}$ equal to the initial minus the final reactant concentration.
-%The individual CAT production and substrate consumption terms for the best-fit set, kinetic models with knockouts, and experimental data are shown in Table~\ref{tbl:yield_breakdown}.
-%Total net consumption of amino acids and amino acid consumption via CAT synthesis were calculated for the best-fit set (Table~\ref{tbl:AA_breakdown}).
+%Total net consumption of amino acids and amino acid consumption via CAT synthesis were calculated for the best-fit set.
 %Total net consumption was calculated as amino acid concentration at 0 hours minus concentration at 3 hours; it was negative if synthesis outweighed consumption.
@@
 $\rm {ATP}_{TX}$ = 176, $\rm {CTP}_{TX}$ = 144, $\rm {GTP}_{TX}$ = 151, $\rm {UTP}_{TX}$ = 189, $\rm {ATP}_{TL}$ = 219, $\rm {GTP}_{TL}$ = 438, $\rm {ATP}_{GLC}$ = 21, $\rm {ATP}_{PYR}$ = 8, $\rm {ATP}_{LAC}$ = 9.5, $\rm {ATP}_{SUCC}$ = 11.5, $\rm {ATP}_{MAL}$ = 10.5.
-This analysis was also extended to the accumulation of amino acids, organic acids, and other byproducts to create a complete energy balance through the network (Fig.~\ref{fig:CAT_balances}, right).
+This analysis was also extended to the accumulation of amino acids, organic acids, and other byproducts to create a complete energy balance through the network.
 These terms were calculated as accumulation times equivalent ATP number over total energy consumption.
